### 2019-09-30 paper/ch_potentials.tex
Initial commit with LIPIcs template

\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
%This is a template for producing LIPIcs articles.
%See lipics-manual.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling cleveref support, use "autoref"

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Dummy title} %TODO Please add

\titlerunning{Dummy short title}%optional, please use if title is longer than one line

\author{John Q. Public}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional

\author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{J.\,Q. Public and J.\,R. Public}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{John Q. Public and Joan R. Public}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{General and reference~General literature}
\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm

\keywords{Dummy keyword}%TODO mandatory; please add comma-separated list of keywords

\category{}%optional, e.g. invited paper

\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversion{A full version of the paper is available at \url{...}.}

\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti.
\end{abstract}

\section{Typesetting instructions -- Summary}
\label{sec:typesetting-summary}

LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl.
In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others. The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.

Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume.
\paragraph*{Minimum requirements}

\begin{itemize}
\item Use pdflatex and an up-to-date \LaTeX{} system.
\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
\item Use a spellchecker to correct typos.
\end{itemize}

\paragraph*{Mandatory metadata macros}
Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
\begin{itemize}
\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
\item \verb+\keywords+ (a comma-separated list of keywords).
\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
\item \verb+\begin{abstract}...\end{abstract}+ .
\end{itemize}

\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
Generally speaking, please do not override the \texttt{lipics-v2019}-style defaults. To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2019} defaults}. In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
\begin{itemize}
\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
\item Do not alter the spacing of the \texttt{lipics-v2019.cls} style file.
\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
 \verb+\begin{enumerate}[(a)]+ or the like.)
\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions.
\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
\item Do not change the bibliography style. In particular, do not use author-year citations. (The
\texttt{natbib} package is not supported.)
\end{itemize}

\enlargethispage{\baselineskip}

This is only a summary containing the most relevant details. Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2019} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
\href{http://drops.dagstuhl.de/styles/lipics-v2019/lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2019/\newline lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}}

\section{Lorem ipsum dolor sit amet}

Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti. Donec eget odio et magna ullamcorper vehicula ut vitae libero. Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis. Pellentesque id ante erat. In hac habitasse platea dictumst. Curabitur a scelerisque odio. Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam. Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.

\begin{lemma}[Lorem ipsum]
\label{lemma:lorem}
Vestibulum sodales dolor et dui cursus iaculis. Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper. Proin facilisis gravida rutrum. Etiam sed sollicitudin lorem. Proin pellentesque risus at elit hendrerit pharetra. Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
\end{lemma}

\begin{proof}
Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.

\begin{claim}
content...
\end{claim}
\begin{claimproof}
content...
\end{claimproof}

\end{proof}

\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
\label{lemma:curabitur}
Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum. Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
\end{corollary}

\begin{proposition}\label{prop1}
This is a proposition
\end{proposition}

\autoref{prop1} and \cref{prop1} \ldots

\subsection{Curabitur dictum felis id sapien}

Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat. Curabitur sed velit diam. Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu. Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet. Duis rutrum dapibus ornare. Nulla vehicula vulputate iaculis. Proin a consequat neque. Donec ut rutrum urna. Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas. Aenean nec faucibus leo. Cras ut nisl odio, non tincidunt lorem. Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat. Fusce risus libero, convallis at fermentum in, dignissim sed sem. Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}. Donec non suscipit lorem. Nam sit amet enim vitae nisl accumsan pretium.

\begin{lstlisting}[caption={Useless code},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
for i:=maxint to 0 do
begin
    j:=square(root(i));
end;
\end{lstlisting}

\subsection{Proin ac fermentum augue}

Proin ac fermentum augue. Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis. Nulla facilisi. Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac. Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna. Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui. Praesent a pretium quam. Mauris non mauris sit amet eros rutrum aliquam id ut sapien. Nulla aliquet fringilla sagittis. Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor. Nulla cursus aliquet enim. Cras sapien risus, accumsan eu cursus ut, commodo vel velit. Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel. Integer faucibus faucibus felis.

\begin{itemize}
\item Ut vitae diam augue.
\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem.
\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit.
\end{itemize}

\begin{remark}
content...
\end{remark}

\section{Pellentesque quis tortor}

Nec urna malesuada sollicitudin. Nulla facilisi. Vivamus aliquam tempus ligula eget ornare. Praesent eget magna ut turpis mattis cursus. Aliquam vel condimentum orci. Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi. Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.

\begin{lemma} [Quisque blandit tempus nunc]
Sed interdum nisl pretium non. Mauris sodales consequat risus vel consectetur. Aliquam erat volutpat. Nunc sed sapien ligula. Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus. Nunc vestibulum nunc ac massa pretium pharetra. Nulla facilisis turpis id augue venenatis blandit. Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
\end{lemma}

Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.

\appendix
\section{Morbi eros magna}

Morbi eros magna, vestibulum non posuere non, porta eu quam. Maecenas vitae orci risus, eget imperdiet mauris. Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis. Sed condimentum convallis dolor, a dignissim est ultrices eu. Donec consectetur volutpat eros, et ornare dui ultricies id. Vivamus eu augue eget dolor euismod ultrices et sit amet nisi. Vivamus malesuada leo ac leo ullamcorper tempor. Donec justo mi, tempor vitae aliquet non, faucibus eu lacus. Donec dictum gravida neque, non porta turpis imperdiet eget. Curabitur quis euismod ligula.

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{references}

\appendix

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

List of different predefined enumeration styles:

\begin{itemize}
\item \verb|\begin{itemize}...\end{itemize}|
\item \dots
\item \dots
%\item \dots
\end{itemize}

\begin{enumerate}
\item \verb|\begin{enumerate}...\end{enumerate}|
\item \dots
\item \dots
%\item \dots
\end{enumerate}

\begin{alphaenumerate}
\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
\item \dots
\item \dots
%\item \dots
\end{alphaenumerate}

\begin{romanenumerate}
\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
\item \dots
\item \dots
%\item \dots
\end{romanenumerate}

\begin{bracketenumerate}
\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
\item \dots
\item \dots
%\item \dots
\end{bracketenumerate}

\begin{description}
\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
\item[Description 2] Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\item[Description 3]  \dots
%\item \dots
\end{description}

\section{Theorem-like environments}\label{sec:theorem-environments}

List of different predefined enumeration styles:

\begin{theorem}\label{testenv-theorem}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{theorem}

\begin{lemma}\label{testenv-lemma}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{lemma}

\begin{corollary}\label{testenv-corollary}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{corollary}

\begin{proposition}\label{testenv-proposition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proposition}

\begin{exercise}\label{testenv-exercise}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{exercise}

\begin{definition}\label{testenv-definition}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{definition}

\begin{example}\label{testenv-example}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{example}

\begin{note}\label{testenv-note}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note}

\begin{note*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{note*}

\begin{remark}\label{testenv-remark}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark}

\begin{remark*}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{remark*}

\begin{claim}\label{testenv-claim}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim}

\begin{claim*}\label{testenv-claim2}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claim*}

\begin{proof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{proof}

\begin{claimproof}
Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
\end{claimproof}

\end{document}

### 2019-09-30 paper/ch_potentials.tex
clean up template

@@
 
-\author{John Q.
-Public}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Ben Strasser}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+\author{Tim Zeitz}{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology, Germany}{tim.zeitz@kit.edu}{}{}
 
-\authorrunning{J.\,Q. Public and J.\,R. Public}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{B. Strasser and T.\, Zeitz}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
-\Copyright{John Q. Public and Joan R. Public}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\Copyright{Ben Strasser and Tim Zeitz}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\category{}%optional, e.g. invited paper
+% \category{}%optional, e.g. invited paper
 
-\relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+% \relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
 %\relatedversion{A full version of the paper is available at \url{...}.}
 
-\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+% \supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
 
@@
 
-\acknowledgements{I want to thank \dots}%optional
+% \acknowledgements{I want to thank \dots}%optional
 
@@
 
-\section{Typesetting instructions -- Summary}
-\label{sec:typesetting-summary}
-
-LIPIcs is a series of open access high-quality conference proceedings across all fields in informatics established in cooperation with Schloss Dagstuhl.
-In order to do justice to the high scientific quality of the conferences that publish their proceedings in the LIPIcs series, which is ensured by the thorough review process of the respective events, we believe that LIPIcs proceedings must have an attractive and consistent layout matching the standard of the series.
-Moreover, the quality of the metadata, the typesetting and the layout must also meet the requirements of other external parties such as indexing service, DOI registry, funding agencies, among others.
-The guidelines contained in this document serve as the baseline for the authors, editors, and the publisher to create documents that meet as many different requirements as possible.
-
-Please comply with the following instructions when preparing your article for a LIPIcs proceedings volume.
-\paragraph*{Minimum requirements}
-
-\begin{itemize}
-\item Use pdflatex and an up-to-date \LaTeX{} system.
-\item Use further \LaTeX{} packages and custom made macros carefully and only if required.
-\item Use the provided sectioning macros: \verb+\section+, \verb+\subsection+, \verb+\subsubsection+, \linebreak \verb+\paragraph+, \verb+\paragraph*+, and \verb+\subparagraph*+.
-\item Provide suitable graphics of at least 300dpi (preferably in PDF format).
-\item Use BibTeX and keep the standard style (\verb+plainurl+) for the bibliography.
-\item Please try to keep the warnings log as small as possible. Avoid overfull \verb+\hboxes+ and any kind of warnings/errors with the referenced BibTeX entries.
-\item Use a spellchecker to correct typos.
-\end{itemize}
-
-\paragraph*{Mandatory metadata macros}
-Please set the values of the metadata macros carefully since the information parsed from these macros will be passed to publication servers, catalogues and search engines.
-Avoid placing macros inside the metadata macros. The following metadata macros/environments are mandatory:
-\begin{itemize}
-\item \verb+\title+ and, in case of long titles, \verb+\titlerunning+.
-\item \verb+\author+, one for each author, even if two or more authors have the same affiliation.
-\item \verb+\authorrunning+ and \verb+\Copyright+ (concatenated author names)\\
-The \verb+\author+ macros and the \verb+\Copyright+ macro should contain full author names (especially with regard to the first name), while \verb+\authorrunning+ should contain abbreviated first names.
-\item \verb+\ccsdesc+ (ACM classification, see \url{https://www.acm.org/publications/class-2012}).
-\item \verb+\keywords+ (a comma-separated list of keywords).
-\item \verb+\relatedversion+ (if there is a related version, typically the ``full version''); please make sure to provide a persistent URL, e.\,g., at arXiv.
-\item \verb+\begin{abstract}...\end{abstract}+ .
-\end{itemize}
-
-\paragraph*{Please do not \ldots} %Do not override the \texttt{\seriesstyle}-defaults}
-Generally speaking, please do not override the \texttt{lipics-v2019}-style defaults.
-To be more specific, a short checklist also used by Dagstuhl Publishing during the final typesetting is given below.
-In case of \textbf{non-compliance} with these rules Dagstuhl Publishing will remove the corresponding parts of \LaTeX{} code and \textbf{replace it with the \texttt{lipics-v2019} defaults}.
-In serious cases, we may reject the LaTeX-source and expect the corresponding author to revise the relevant parts.
-\begin{itemize}
-\item Do not use a different main font. (For example, the \texttt{times} package is forbidden.)
-\item Do not alter the spacing of the \texttt{lipics-v2019.cls} style file.
-\item Do not use \verb+enumitem+ and \verb+paralist+. (The \texttt{enumerate} package is preloaded, so you can use
- \verb+\begin{enumerate}[(a)]+ or the like.)
-\item Do not use ``self-made'' sectioning commands (e.\,g., \verb+\noindent{\bf My+ \verb+Paragraph}+).
-\item Do not hide large text blocks using comments or \verb+\iffalse+ $\ldots$ \verb+\fi+ constructions.
-\item Do not use conditional structures to include/exclude content. Instead, please provide only the content that should be published -- in one file -- and nothing else.
-\item Do not wrap figures and tables with text. In particular, the package \texttt{wrapfig} is not supported.
-\item Do not change the bibliography style. In particular, do not use author-year citations. (The
-\texttt{natbib} package is not supported.)
-\end{itemize}
-
-\enlargethispage{\baselineskip}
-
-This is only a summary containing the most relevant details.
-Please read the complete document ``LIPIcs: Instructions for Authors and the \texttt{lipics-v2019} Class'' for all details and don't hesitate to contact Dagstuhl Publishing (\url{mailto:publishing@dagstuhl.de}) in case of questions or comments:
-\href{http://drops.dagstuhl.de/styles/lipics-v2019/lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}{\texttt{http://drops.dagstuhl.de/styles/lipics-v2019/\newline lipics-v2019-authors/lipics-v2019-authors-guidelines.pdf}}
-
-\section{Lorem ipsum dolor sit amet}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit \cite{DBLP:journals/cacm/Knuth74}.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
-Donec eget odio et magna ullamcorper vehicula ut vitae libero.
-Maecenas lectus nulla, auctor nec varius ac, ultricies et turpis.
-Pellentesque id ante erat.
-In hac habitasse platea dictumst.
-Curabitur a scelerisque odio.
-Pellentesque elit risus, posuere quis elementum at, pellentesque ut diam.
-Quisque aliquam libero id mi imperdiet quis convallis turpis eleifend.
-
-\begin{lemma}[Lorem ipsum]
-\label{lemma:lorem}
-Vestibulum sodales dolor et dui cursus iaculis.
-Nullam ullamcorper purus vel turpis lobortis eu tempus lorem semper.
-Proin facilisis gravida rutrum.
-Etiam sed sollicitudin lorem.
-Proin pellentesque risus at elit hendrerit pharetra.
-Integer at turpis varius libero rhoncus fermentum vitae vitae metus.
-\end{lemma}
-
-\begin{proof}
-Cras purus lorem, pulvinar et fermentum sagittis, suscipit quis magna.
-
-\begin{claim}
-content...
-\end{claim}
-\begin{claimproof}
-content...
-\end{claimproof}
-
-\end{proof}
-
-\begin{corollary}[Curabitur pulvinar, \cite{DBLP:books/mk/GrayR93}]
-\label{lemma:curabitur}
-Nam liber tempor cum soluta nobis eleifend option congue nihil imperdiet doming id quod mazim placerat facer possim assum.
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit, sed diam nonummy nibh euismod tincidunt ut laoreet dolore magna aliquam erat volutpat.
-\end{corollary}
-
-\begin{proposition}\label{prop1}
-This is a proposition
-\end{proposition}
-
-\autoref{prop1} and \cref{prop1} \ldots
-
-\subsection{Curabitur dictum felis id sapien}
-
-Curabitur dictum \cref{lemma:curabitur} felis id sapien \autoref{lemma:curabitur} mollis ut venenatis tortor feugiat.
-Curabitur sed velit diam.
-Integer aliquam, nunc ac egestas lacinia, nibh est vehicula nibh, ac auctor velit tellus non arcu.
-Vestibulum lacinia ipsum vitae nisi ultrices eget gravida turpis laoreet.
-Duis rutrum dapibus ornare.
-Nulla vehicula vulputate iaculis.
-Proin a consequat neque.
-Donec ut rutrum urna.
-Morbi scelerisque turpis sed elit sagittis eu scelerisque quam condimentum.
-Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.
-Aenean nec faucibus leo.
-Cras ut nisl odio, non tincidunt lorem.
-Integer purus ligula, venenatis et convallis lacinia, scelerisque at erat.
-Fusce risus libero, convallis at fermentum in, dignissim sed sem.
-Ut dapibus orci vitae nisl viverra nec adipiscing tortor condimentum \cite{DBLP:journals/cacm/Dijkstra68a}.
-Donec non suscipit lorem.
-Nam sit amet enim vitae nisl accumsan pretium.
-
-\begin{lstlisting}[caption={Useless code},label=list:8-6,captionpos=t,float,abovecaptionskip=-\medskipamount]
-for i:=maxint to 0 do
-begin
-    j:=square(root(i));
-end;
-\end{lstlisting}
-
-\subsection{Proin ac fermentum augue}
-
-Proin ac fermentum augue.
-Nullam bibendum enim sollicitudin tellus egestas lacinia euismod orci mollis.
-Nulla facilisi.
-Vivamus volutpat venenatis sapien, vitae feugiat arcu fringilla ac.
-Mauris sapien tortor, sagittis eget auctor at, vulputate pharetra magna.
-Sed congue, dui nec vulputate convallis, sem nunc adipiscing dui, vel venenatis mauris sem in dui.
-Praesent a pretium quam.
-Mauris non mauris sit amet eros rutrum aliquam id ut sapien.
-Nulla aliquet fringilla sagittis.
-Pellentesque eu metus posuere nunc tincidunt dignissim in tempor dolor.
-Nulla cursus aliquet enim.
-Cras sapien risus, accumsan eu cursus ut, commodo vel velit.
-Praesent aliquet consectetur ligula, vitae iaculis ligula interdum vel.
-Integer faucibus faucibus felis.
-
-\begin{itemize}
-\item Ut vitae diam augue.
-\item Integer lacus ante, pellentesque sed sollicitudin et, pulvinar adipiscing sem.
-\item Maecenas facilisis, leo quis tincidunt egestas, magna ipsum condimentum orci, vitae facilisis nibh turpis et elit.
-\end{itemize}
-
-\begin{remark}
-content...
-\end{remark}
-
-\section{Pellentesque quis tortor}
-
-Nec urna malesuada sollicitudin.
-Nulla facilisi.
-Vivamus aliquam tempus ligula eget ornare.
-Praesent eget magna ut turpis mattis cursus.
-Aliquam vel condimentum orci.
-Nunc congue, libero in gravida convallis \cite{DBLP:conf/focs/HopcroftPV75}, orci nibh sodales quam, id egestas felis mi nec nisi.
-Suspendisse tincidunt, est ac vestibulum posuere, justo odio bibendum urna, rutrum bibendum dolor sem nec tellus.
-
-\begin{lemma} [Quisque blandit tempus nunc]
-Sed interdum nisl pretium non.
-Mauris sodales consequat risus vel consectetur.
-Aliquam erat volutpat.
-Nunc sed sapien ligula.
-Proin faucibus sapien luctus nisl feugiat convallis faucibus elit cursus.
-Nunc vestibulum nunc ac massa pretium pharetra.
-Nulla facilisis turpis id augue venenatis blandit.
-Cum sociis natoque penatibus et magnis dis parturient montes, nascetur ridiculus mus.
-\end{lemma}
-
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-
-\appendix
-\section{Morbi eros magna}
-
-Morbi eros magna, vestibulum non posuere non, porta eu quam.
-Maecenas vitae orci risus, eget imperdiet mauris.
-Donec massa mauris, pellentesque vel lobortis eu, molestie ac turpis.
-Sed condimentum convallis dolor, a dignissim est ultrices eu.
-Donec consectetur volutpat eros, et ornare dui ultricies id.
-Vivamus eu augue eget dolor euismod ultrices et sit amet nisi.
-Vivamus malesuada leo ac leo ullamcorper tempor.
-Donec justo mi, tempor vitae aliquet non, faucibus eu lacus.
-Donec dictum gravida neque, non porta turpis imperdiet eget.
-Curabitur quis euismod ligula.
+\section{Introduction}
+\label{sec:intro}
 
@@
 
-\appendix
-
-\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}
-
-List of different predefined enumeration styles:
-
-\begin{itemize}
-\item \verb|\begin{itemize}...\end{itemize}|
-\item \dots
-\item \dots
-%\item \dots
-\end{itemize}
-
-\begin{enumerate}
-\item \verb|\begin{enumerate}...\end{enumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{enumerate}
-
-\begin{alphaenumerate}
-\item \verb|\begin{alphaenumerate}...\end{alphaenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{alphaenumerate}
-
-\begin{romanenumerate}
-\item \verb|\begin{romanenumerate}...\end{romanenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{romanenumerate}
-
-\begin{bracketenumerate}
-\item \verb|\begin{bracketenumerate}...\end{bracketenumerate}|
-\item \dots
-\item \dots
-%\item \dots
-\end{bracketenumerate}
-
-\begin{description}
-\item[Description 1] \verb|\begin{description} \item[Description 1]  ...\end{description}|
-\item[Description 2] Fusce eu leo nisi.
-Cras eget orci neque, eleifend dapibus felis.
-Duis et leo dui.
-Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\item[Description 3]  \dots
-%\item \dots
-\end{description}
-
-\section{Theorem-like environments}\label{sec:theorem-environments}
-
-List of different predefined enumeration styles:
-
-\begin{theorem}\label{testenv-theorem}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{theorem}
-
-\begin{lemma}\label{testenv-lemma}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{lemma}
-
-\begin{corollary}\label{testenv-corollary}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{corollary}
-
-\begin{proposition}\label{testenv-proposition}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{proposition}
-
-\begin{exercise}\label{testenv-exercise}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{exercise}
-
-\begin{definition}\label{testenv-definition}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{definition}
-
-\begin{example}\label{testenv-example}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{example}
-
-\begin{note}\label{testenv-note}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{note}
-
-\begin{note*}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{note*}
-
-\begin{remark}\label{testenv-remark}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{remark}
-
-\begin{remark*}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{remark*}
-
-\begin{claim}\label{testenv-claim}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claim}
-
-\begin{claim*}\label{testenv-claim2}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claim*}
-
-\begin{proof}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{proof}
-
-\begin{claimproof}
-Fusce eu leo nisi. Cras eget orci neque, eleifend dapibus felis. Duis et leo dui. Nam vulputate, velit et laoreet porttitor, quam arcu facilisis dui, sed malesuada risus massa sit amet neque.
-\end{claimproof}
+% \appendix
 

### 2019-10-06 paper/ch_potentials.tex
sketch out structure
classification foo
Initial text

@@
 %for enabling cleveref support, use "autoref"
+
+\usepackage[algo2e,vlined]{algorithm2e}
 
@@
 
-\ccsdesc[100]{General and reference~General literature}
-\ccsdesc[100]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
+\ccsdesc[500]{Theory of computation~Shortest paths}
+% \ccsdesc[500]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
 
@@
 
+\hideLIPIcs
+
 \begin{document}
@@
 
+Fast routing in huge street networks has seen a lot of research \cite{Survey}.
+For static, time- and user-invariant networks a lot of techniques exist. 
+One of the most successful ones are Contraction Hierarchies~\cite{CH}.
+These work in two phases.
+The first phase is called preprocessing phase and is a slow index-construction.
+In this phase the road graph and the edge weights are known.
+In the second phase, called query phase, minimum weight paths are computed.
+The query phase has access to the precomputed index and is very fast.
+Queries can be answered on the order of only a millisecond by CH.
+
+These techniques are used in online navigation services. 
+In these services, there is a single index.
+With this index a large number of user requests are answered.
+The preprocessing phases is rerun on a regular interval to incorporate map updates.
+
+While highly scalable, this setup has the disadvantage, that every user must have use the same weights.
+In this paper, we explore an novel algorithm, called CH-Potentials, that uses the two phase setup, however, allows the weights to be modified on a per query basis.
+Applications for this algorithm are personalized user settings such as for example:
+Avoiding highways, capping the maximum speed, or avoid user-configurable areas.
+
+\paragraph{Related Work}
+
+We are not the first to investigate this problem setting. 
+Two approaches exist in the litrature.
+
+The three phase approach splits the preprocessing phase into two.
+Techniques that use this setup are MLD/CRP \cite{?} and CCH \cite{?}.
+The first phase is a weight-independent preprocessing phase that analsyse the graph topology.
+The second phase is called customization phase and introduces the weights into the index.
+The third phase is the query phase that computes paths.
+The preprocessing phase is very slow.
+The customization phase runs on the order of seconds.
+The query phase runs within milliseconds.
+
+The three phase approach is ideal to incorporate weight changes that apply to all users.
+Live traffic information is a prime example.
+It has been sugested to use the three phase approach to handle user-specific requirements by storing a copy of the index per user.
+The problem with this approach is that it binds a lot of server-side resources per user.
+Further, if the user often changes his weights, the customization needs to run often.
+Every customization requires seconds of running time.
+Just applying Dijkstra's algorithm to the input graph results in similar running times.
+
+The second approach is Sabine's VLDB personalized routing.
+Also TopoCore~\cite{?}.
+
+A paragraph about A*~\cite{?}.
+
+A paragraph about PHAST~\cite{?}.
+
+\paragraph{CH-Potential Overview}
+
+Our novel algroithm CH-Potentials is at its core a combination of A* and PHAST.
+For an additional performance, we further combine it with TopoCore.
+
+The input of the preprocessing phase is a graph and lower-bound weights $w_{\ell}$.
+Denote by $d_{\ell}(x,y)$ the minimum weight $xy$-path distance with respect to $w_{\ell}$.
+The preprocessing phase consists of computing a CH.
+
+The input to the query phase is a source $s$ and a target $t$ node and user-specific weights $w_u$.
+These weights $w_u$ must not be smaller than the lower-bound weights $w_{\ell}$, i.e., $w_{\ell}(e) < w_u(e)$ for all edges $e$.
+Analogous to $d_{\ell}(x,y)$, $d_u(x,y)$ is the minimum-weight phase according to $w_u$.
+The core of the query phase is an A* search with respect to $d_u$ from $s$ to $t$.
+The distance $d_{\ell}(x,t)$ from a node $x$ to $t$ is used as potential.
+
+\begin{algorithm2e}
+\KwData{$B[x]$ distance from $s$ to $x$ compute in backward search, or $+\infty$ if unreachable}
+\KwData{$P[x]$ memoized potential for $x$, or $\bot$ if not yet computed. Equal to $d_{\ell}(x,t)$}
+\SetKwFunction{Pot}{Potential}
+\SetKwProg{Fn}{Function}{:}{}
+\Fn{\Pot{$x$}}{
+  \If{$P[x] = \bot$}{
+    $P[x]\leftarrow B[x]$\;
+    \For{$(x,y)$ is upward CH edge with weight $w_{xy}$}{
+      $P[x]\leftarrow \min\{P[x],w_{xy}+\Pot(y)\}$\;
+    }
+  }
+  \Return{$P[x]$}\;
+}
+
+\caption{Algorithm to compute Potentials.}
+\label{algo:pot}
+
+\end{algorithm2e}
+
+The core contribution of our algorithm consist in the way $d_{\ell}(x,t)$ is efficiently computed.
+For this, we use a PHAST variant.
+In the first step, we explore all nodes backward-reachable from $t$ in the CH.
+We store the computed distances in the array $B$.
+We then execute the A* search.
+Potentials are computed recursively, on-the-fly, with memoization as depicted in Algorithm \ref{algo:pot}.
+If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes in the CH backward search graph.
+These potentials are then increased by the distance in the CH backwards search graph.
+The potential of a node $x$ is the minimum over all these values and the distance found in the CH forward search.
+
+Our algorithm can be applied as described to the input graph.
+However, depending on the weights, the A* search can get stuck often in deadends.
+We therefore first coarsen the input graph using TopoCore and apply the above described algorithm to the core.
+In this setup, TopoCore handles all deadends.
+
 \section{Introduction}
 \label{sec:intro}
+
+\section{Preliminaries}
+\label{sec:prelim}
+
+\section{Perfect Potentials}
+\label{sec:algo}
+
+\section{Applications}
+\label{sec:applications}
+
+\subsection{Live Traffic}
+\subsection{Time-dependent Routing}
+
+\section{Experiments}
+\label{sec:exp}
+
+\section{Conclusion}
+\label{sec:conclusion}
 

### 2019-10-23 paper/ch_potentials.tex
add missing comma
write down stuff for arxiv version

--- 2019-09-30 paper/ch_potentials.tex
@@
 
+\usepackage[algo2e,vlined]{algorithm2e}
+
 %\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
@@
 
-\title{Dummy title} %TODO Please add
+\title{A* with perfect potentials} %TODO Please add
 
-\titlerunning{Dummy short title}%optional, please use if title is longer than one line
+% \titlerunning{Dummy short title}%optional, please use if title is longer than one line
 
@@
 
-\authorrunning{B. Strasser and T.\, Zeitz}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\authorrunning{B. Strasser and T. Zeitz}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
 
@@
 
-%TODO mandatory: add short abstract of the document
 \begin{abstract}
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Praesent convallis orci arcu, eu mollis dolor.
-Aliquam eleifend suscipit lacinia.
-Maecenas quam mi, porta ut lacinia sed, convallis ac dui.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Suspendisse potenti.
+A* is an extension to Dijkstra's algorithm.
+It uses a potential function to estimate the distance of each node to the target.
+By adding these estimates to the queue keys, the search is directed towards the target.
+The quality of this potential determines the performance of A*.
+We introduce a novel potential function which utilizes Contraction Hierarchies to efficiently calculate perfect potentials on road networks.
+Perfect potentials can be used to speed up shortest path queries for extended problem settings like for example frequently changing or time-dependent edge weights.
 \end{abstract}
@@
 \label{sec:intro}
+
+Fast routing in huge street networks has seen a lot of research \cite{Survey}.
+For static, time- and user-invariant networks a lot of techniques exist.
+They work in two phases.
+The first phase is called preprocessing phase and is a slow index-construction.
+In this phase the road graph and the edge weights are known.
+In the second phase, called query phase, minimum weight paths are computed.
+The query phase has access to the precomputed index and is very fast.
+
+One of the most successful ones are Contraction Hierarchies~(CH)~\cite{CH}.
+Contraction Hierarchies exploit inherent hierarchies in the shortest-path structure of road networks.
+Using CHs, Queries can be answered on the order of only a millisecond.
+However, CHs can not always be easily extended to more complicated problem settings.
+% TODO what problem settings
+While such extensions exist~\cite{CCH,KaTCH}, they often require substantial engineering effort.
+
+On the other hand, goal directed speed-up techniques based on A* are usually slower but more flexible.
+The performance of these approaches depends on the quality of the potential functions used for the A* search.
+In this paper, we propose an algorithm which utilizes Contraction Hierarchies to efficiently calculate perfect (best possible) potentials.
+This allows us to retain the flexibility of A* based approaches but use much more effective potentials than previously possible.
+% These techniques are used in online navigation services.
+% In these services, there is a single index.
+% With this index a large number of user requests are answered.
+% The preprocessing phases is rerun on a regular interval to incorporate map updates.
+
+% While highly scalable, this setup has the disadvantage, that every user must have use the same weights.
+% In this paper, we explore an novel algorithm, called CH-Potentials, that uses the two phase setup, however, allows the weights to be modified on a per query basis.
+% Applications for this algorithm are personalized user settings such as for example:
+% Avoiding highways, capping the maximum speed, or avoid user-configurable areas.
+
+\paragraph*{Related Work}
+
+% We are not the first to investigate this problem setting.
+% Two approaches exist in the literature.
+
+% The three phase approach splits the preprocessing phase into two.
+% Techniques that use this setup are MLD/CRP \cite{?} and CCH \cite{?}.
+% The first phase is a weight-independent preprocessing phase that analsyse the graph topology.
+% The second phase is called customization phase and introduces the weights into the index.
+% The third phase is the query phase that computes paths.
+% The preprocessing phase is very slow.
+% The customization phase runs on the order of seconds.
+% The query phase runs within milliseconds.
+
+% The three phase approach is ideal to incorporate weight changes that apply to all users.
+% Live traffic information is a prime example.
+% It has been suggested to use the three phase approach to handle user-specific requirements by storing a copy of the index per user.
+% The problem with this approach is that it binds a lot of server-side resources per user.
+% Further, if the user often changes his weights, the customization needs to run often.
+% Every customization requires seconds of running time.
+% Just applying Dijkstra's algorithm to the input graph results in similar running times.
+
+% The second approach is Sabine's VLDB personalized routing.
+% Also TopoCore~\cite{?}.
+
+A plethora of of research exists in the area of efficient routing in road networks \cite{Survey}.
+Here, we are going to focus on the algorithms relevant to this work.
+
+At the core is the algorithm of Dijkstra~\cite{Dijkstra}.
+This algorithm explores the graph starting at a source node $s$ and visiting nodes ordered by increasing distance from $s$.
+For each node $v$, a tentative distance $d_s(v)$ is maintained, which is initially set to $\infty$.
+In each iteration the node $u$ with the smallest remaining distance $d_s(u)$ is extracted from a priority queue and settled.
+Each outgoing edge $(u,v)$ of $u$ is relaxed, that is to check if $d_s(u) + w(u,v) < d_s(v)$ and if so update the distance and the queue position of $v$ or insert $v$ into the queue.
+Once the target node $t$ is settled, the shortest path and distance between $s$ and $t$ is known.
+
+A*~\cite{?} extends Dijkstra's algorithm by changing the order in which nodes are visited such that nodes in the direction of $t$ are visited earlier.
+This is achieved through a potential function $p$ which estimates for each node the distance to $t$.
+The priority queue is then ordered by $d_s(u) + p(u)$.
+
+ALT~\cite{ALT} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
+For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
+This achieves decent speed-ups and can be used in other problem settings but is not competitive to hierarchical approaches
+
+Contraction Hierarchies~\cite{CH} are a popular hierarchical speed-up technique.
+During preprocessing, all nodes are ranked by some measure of importance.
+Nodes which lie on many shortest paths (for example highway nodes) are important.
+Then, nodes are contracted iteratively by increasing rank.
+To contract a node $v$ means removing it form the graph, but preserving all shortest distances among all other nodes.
+This is achieved by inserting shortcut edges between the neighbors of $v$ which have the length of the path over $v$, if necessary.
+The query is a bidirectional Dijkstra starting from $s$ and $t$ where only edges to nodes with higher rank are relaxed.
+
+PHAST~\cite{?} is an algorithm for one-to-many queries based on Contraction Hierarchies.
+The preprocessing phase remains the same.
+For the query, the forward search $s$ also remains the same.
+Instead of running a backward upward search from all nodes, PHAST now relaxes all edges $(u,v)$ in the backward graph where $rank(u) > rank(v)$ ordered descending by $rank(v)$.
+To achieve good running times, it is important that nodes and edges are ordered such that relaxing the downward edges is just a linear sweep over the edge array.
+
+% \section{Preliminaries}
+% \label{sec:prelim}
+
+\section{Perfect Potentials}
+\label{sec:algo}
+
+Our new algorithm is at its core a combination of A* and PHAST.
+% For an additional performance, we further combine it with TopoCore.
+
+The input of the preprocessing phase is a graph and lower-bound weights $w_{\ell}$.
+Denote by $d_{\ell}(x,y)$ the minimum weight $xy$-path distance with respect to $w_{\ell}$.
+The preprocessing phase consists of computing a CH of the graph with lower-bound weights.
+
+The input to the query phase is a source $s$ and a target $t$ node and weights $w_u$.
+These weights $w_u$ must not be smaller than the lower-bound weights $w_{\ell}$, i.e., $w_{\ell}(e) < w_u(e)$ for all edges $e$.
+The weights may be time-dependent travel time predictions, incorporate live traffic or user specific preferences.
+Analogous to $d_{\ell}(x,y)$, $d_u(x,y)$ is the minimum-weight phase according to $w_u$.
+We obtain $d_u(x,y)$ through an A* search with respect to $d_u$ from $s$ to $t$.
+The distance $d_{\ell}(x,t)$ from a node $x$ to $t$ is used as potential.
+
+\begin{algorithm2e}
+\KwData{$B[x]$ distance from $s$ to $x$ compute in backward search, or $+\infty$ if unreachable}
+\KwData{$P[x]$ memoized potential for $x$, or $\bot$ if not yet computed. Equal to $d_{\ell}(x,t)$}
+\SetKwFunction{Pot}{Potential}
+\SetKwProg{Fn}{Function}{:}{}
+\Fn{\Pot{$x$}}{
+  \If{$P[x] = \bot$}{
+    $P[x]\leftarrow B[x]$\;
+    \For{$(x,y)$ is upward CH edge with weight $w_{xy}$}{
+      $P[x]\leftarrow \min\{P[x],w_{xy}+\Pot(y)\}$\;
+    }
+  }
+  \Return{$P[x]$}\;
+}
+
+\caption{Algorithm to compute Potentials.}
+\label{algo:pot}
+
+\end{algorithm2e}
+
+The core contribution of our algorithm consist in the way $d_{\ell}(x,t)$ is efficiently computed.
+For this, we use a PHAST variant.
+In the first step, we explore all nodes backward-reachable from $t$ in the CH.
+We store the computed distances in the array $B$.
+We then execute the A* search.
+Potentials are computed recursively, on-the-fly, with memoization as depicted in Algorithm \ref{algo:pot}.
+If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes in the CH backward search graph.
+These potentials are then increased by the distance in the CH backwards search graph.
+The potential of a node $x$ is the minimum over all these values and the distance found in the CH forward search.
+
+% Our algorithm can be applied as described to the input graph.
+% However, depending on the weights, the A* search can get stuck often in deadends.
+% We therefore first coarsen the input graph using TopoCore and apply the above described algorithm to the core.
+% In this setup, TopoCore handles all deadends.
+
+% \section{Applications}
+% \label{sec:applications}
+
+% \subsection{Live Traffic}
+% \subsection{Time-dependent Routing}
+
+% \section{Experiments}
+% \label{sec:exp}
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+We introduce a novel way to compute potentials for the A* algorithm.
+The potentials are calculated utilizing a CH on a lower bound graph and perfect with regards to that lower bound graph.
+With these potentials we can speed up A* search for extended route planning problems like route planning with live traffic, time-dependent travel time predictions or user preferences.
 

### 2019-10-24 paper/ch_potentials.tex
added Bens 'affiliation'

@@
 
-\author{Ben Strasser}{Dummy University Computing Laboratory, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.

### 2019-10-24 paper/ch_potentials.tex
add references

@@
 The quality of this potential determines the performance of A*.
-We introduce a novel potential function which utilizes Contraction Hierarchies to efficiently calculate perfect potentials on road networks.
+We introduce a novel potential function which utilizes Contraction Hierarchies to efficiently calculate perfect potentials on road networks. % TODO not the function is novel
 Perfect potentials can be used to speed up shortest path queries for extended problem settings like for example frequently changing or time-dependent edge weights.
@@
 
-Fast routing in huge street networks has seen a lot of research \cite{Survey}.
+Fast routing in huge street networks has seen a lot of research \cite{bdgmpsww-rptn-16}.
 For static, time- and user-invariant networks a lot of techniques exist.
@@
 
-One of the most successful ones are Contraction Hierarchies~(CH)~\cite{CH}.
+One of the most successful ones are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
 Contraction Hierarchies exploit inherent hierarchies in the shortest-path structure of road networks.
@@
 % TODO what problem settings
-While such extensions exist~\cite{CCH,KaTCH}, they often require substantial engineering effort.
+While such extensions exist~\cite{dsw-cch-15,bgsv-mtdtt-13}, they often require substantial engineering effort.
 
@@
 
-A plethora of of research exists in the area of efficient routing in road networks \cite{Survey}.
-Here, we are going to focus on the algorithms relevant to this work.
+A plethora of of research exists in the area of efficient routing in road networks \cite{bdgmpsww-rptn-16}.
+Here, we focus on the algorithms directly relevant to this work.
 
-At the core is the algorithm of Dijkstra~\cite{Dijkstra}.
+At the core is the algorithm of Dijkstra~\cite{d-ntpcg-59}.
 This algorithm explores the graph starting at a source node $s$ and visiting nodes ordered by increasing distance from $s$.
@@
 
-A*~\cite{?} extends Dijkstra's algorithm by changing the order in which nodes are visited such that nodes in the direction of $t$ are visited earlier.
+A*~\cite{hnr-afbhd-68} extends Dijkstra's algorithm by changing the order in which nodes are visited such that nodes in the direction of $t$ are visited earlier.
 This is achieved through a potential function $p$ which estimates for each node the distance to $t$.
@@
 
-ALT~\cite{ALT} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
+ALT~\cite{gh-cspas-05} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
 For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
@@
 
-Contraction Hierarchies~\cite{CH} are a popular hierarchical speed-up technique.
+Contraction Hierarchies~\cite{gssv-erlrn-12} are a popular hierarchical speed-up technique.
 During preprocessing, all nodes are ranked by some measure of importance.
@@
 
-PHAST~\cite{?} is an algorithm for one-to-many queries based on Contraction Hierarchies.
+PHAST~\cite{dgnw-phast-13} is an algorithm for one-to-many queries based on Contraction Hierarchies.
 The preprocessing phase remains the same.

### 2019-10-24 paper/ch_potentials.tex
wordsmithing

@@
 
-\keywords{Dummy keyword}%TODO mandatory; please add comma-separated list of keywords
+\keywords{Route planning, Contraction Hierarchies, A*}%TODO mandatory; please add comma-separated list of keywords
 
@@
 \begin{abstract}
+Quickly determining shortest paths in networks is an important ingredient for many routing applications.
+While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical problems.
 A* is an extension to Dijkstra's algorithm.
@@
 The quality of this potential determines the performance of A*.
-We introduce a novel potential function which utilizes Contraction Hierarchies to efficiently calculate perfect potentials on road networks. % TODO not the function is novel
-Perfect potentials can be used to speed up shortest path queries for extended problem settings like for example frequently changing or time-dependent edge weights.
+We introduce a novel way to efficiently calculate potentials through Contraction Hierarchies.
+These potentials are perfect with respect to a lower bound graph.
+They can be used to speed up shortest path queries for extended problem settings like frequently changing or time-dependent edge weights.
 \end{abstract}
@@
 At the core is the algorithm of Dijkstra~\cite{d-ntpcg-59}.
-This algorithm explores the graph starting at a source node $s$ and visiting nodes ordered by increasing distance from $s$.
+This algorithm explores the graph visiting nodes ordered by increasing distance from a source node $s$.
 For each node $v$, a tentative distance $d_s(v)$ is maintained, which is initially set to $\infty$.
 In each iteration the node $u$ with the smallest remaining distance $d_s(u)$ is extracted from a priority queue and settled.
-Each outgoing edge $(u,v)$ of $u$ is relaxed, that is to check if $d_s(u) + w(u,v) < d_s(v)$ and if so update the distance and the queue position of $v$ or insert $v$ into the queue.
+For outgoing edge $(u,v)$ of $u$ the algorithms checks if $d_s(u) + w(u,v) < d_s(v)$.
+If so, the distance and the queue position of $v$ is updated.
+This process is denoted as edge relaxation.
 Once the target node $t$ is settled, the shortest path and distance between $s$ and $t$ is known.
 
-A*~\cite{hnr-afbhd-68} extends Dijkstra's algorithm by changing the order in which nodes are visited such that nodes in the direction of $t$ are visited earlier.
-This is achieved through a potential function $p$ which estimates for each node the distance to $t$.
-The priority queue is then ordered by $d_s(u) + p(u)$.
+A*~\cite{hnr-afbhd-68} extends Dijkstra's algorithm by changing the order in which nodes are visited.
+Nodes in the direction of $t$ should be visited earlier.
+This is achieved through a potential function $p$ estimating the distance to $t$.
+The priority queue is ordered by $d_s(u) + p(u)$.
 
@@
 For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
-This achieves decent speed-ups and can be used in other problem settings but is not competitive to hierarchical approaches
+This achieves decent speed-ups and can be used in other problem settings.
+However, the performance is not competitive to hierarchical approaches.
 
@@
 During preprocessing, all nodes are ranked by some measure of importance.
-Nodes which lie on many shortest paths (for example highway nodes) are important.
+Nodes which lie on many shortest paths are important.
 Then, nodes are contracted iteratively by increasing rank.
-To contract a node $v$ means removing it form the graph, but preserving all shortest distances among all other nodes.
-This is achieved by inserting shortcut edges between the neighbors of $v$ which have the length of the path over $v$, if necessary.
-The query is a bidirectional Dijkstra starting from $s$ and $t$ where only edges to nodes with higher rank are relaxed.
+To contract a node $v$ means removing it form the graph, but preserving shortest distances among all other nodes.
+This is achieved by inserting shortcut edges between the neighbors of $v$ which have the length of the path over $v$.
+The query is a bidirectional Dijkstra starting from $s$ and $t$ where only edges to higher ranked nodes are relaxed.
 
@@
 The preprocessing phase remains the same.
-For the query, the forward search $s$ also remains the same.
-Instead of running a backward upward search from all nodes, PHAST now relaxes all edges $(u,v)$ in the backward graph where $rank(u) > rank(v)$ ordered descending by $rank(v)$.
-To achieve good running times, it is important that nodes and edges are ordered such that relaxing the downward edges is just a linear sweep over the edge array.
+The query begins with a forward upward search from $s$.
+The backward search now needs to determine distances to \emph{all} nodes.
+To achieve this, PHAST relaxes all edges $(u,v)$ in the backward graph where $rank(u) > rank(v)$ ordered descending by $rank(v)$.
+Having processed all edges, distances to all nodes are known.
+PHAST achieves good running times by heavily exploiting memory locality:
+Nodes and edges are reordered such that relaxing all downward edges is just a linear sweep over the edge array.
 
@@
 Our new algorithm is at its core a combination of A* and PHAST.
+We take the preprocessing from CHs and PHAST and apply it to a lower bound graph.
+The query is an A* variant.
+It uses exact distances on the lower bound graph as potentials and obtains these through a lazy PHAST-like approach.
 % For an additional performance, we further combine it with TopoCore.
@@
 Denote by $d_{\ell}(x,y)$ the minimum weight $xy$-path distance with respect to $w_{\ell}$.
+How exactly this lower bound graph is obtained depends on the problem setting.
+In the case of routing with life traffic it may be the free flow travel time.
+In the case of time-dependent routing it may be the lower bound of each travel time function.
 The preprocessing phase consists of computing a CH of the graph with lower-bound weights.
@@
 
-The core contribution of our algorithm consist in the way $d_{\ell}(x,t)$ is efficiently computed.
+Our core contribution consist in the way $d_{\ell}(x,t)$ is efficiently computed.
 For this, we use a PHAST variant.
-In the first step, we explore all nodes backward-reachable from $t$ in the CH.
+Before starting the A* search, we explore all nodes backward-reachable from $t$ in the CH.
 We store the computed distances in the array $B$.
@@
 
-We introduce a novel way to compute potentials for the A* algorithm.
+We introduce a novel way to compute potentials for A*.
 The potentials are calculated utilizing a CH on a lower bound graph and perfect with regards to that lower bound graph.

### 2019-10-24 paper/ch_potentials.tex
better keywords
more concrete abstract

@@
 
-\keywords{Route planning, Contraction Hierarchies, A*}%TODO mandatory; please add comma-separated list of keywords
+\keywords{shortest path, road graphs, goal-directed search, contraction hierarchy}%TODO mandatory; please add comma-separated list of keywords
 
@@
 The quality of this potential determines the performance of A*.
-We introduce a novel way to efficiently calculate potentials through Contraction Hierarchies.
-These potentials are perfect with respect to a lower bound graph.
-They can be used to speed up shortest path queries for extended problem settings like frequently changing or time-dependent edge weights.
+We introduce a novel way to efficiently calculate perfect potentials for extended problem settings where a lower bound graph is available.
+For example, in the case of routing with live traffic, this could be the free flow graph.
 \end{abstract}

### 2019-10-28 paper/submit_arxiv.sh
prepare for arxiv

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+if [[ $(git diff --shortstat 2> /dev/null | tail -n1) != "" ]] ; then
+  echo "Cant export - would possibly override stuff"
+  exit 0
+fi
+
+sed -i '/^%/ d' ch_potentials.tex
+sed -i 's/bibliography{references}/input{ch_potentials.bbl}/' ch_potentials.tex
+
+zip arxiv.zip \
+ch_potentials.tex \
+lipics-v2019.cls \
+ch_potentials.bbl
+
+git checkout ch_potentials.tex

### 2019-12-11 paper/ch_potentials.tex
capitalize title
prepare for arxiv
write stuff

@@
 
-\title{A* with perfect potentials} %TODO Please add
+\title{A* with Perfect Potentials} %TODO Please add
 
@@
 
-\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{TODO}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{John Q. Open and Joan R. Access}
-\EventNoEds{2}
-\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-\EventShortTitle{CVIT 2016}
-\EventAcronym{CVIT}
-\EventYear{2016}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
+% \EventEditors{John Q. Open and Joan R. Access}
+% \EventNoEds{2}
+% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
+% \EventShortTitle{CVIT 2016}
+% \EventAcronym{CVIT}
+% \EventYear{2016}
+% \EventDate{December 24--27, 2016}
+% \EventLocation{Little Whinging, United Kingdom}
+% \EventLogo{}
+% \SeriesVolume{42}
+% \ArticleNo{23}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \hideLIPIcs
+\nolinenumbers
 
@@
 \begin{abstract}
-Quickly determining shortest paths in networks is an important ingredient for many routing applications.
+Efficiently computing flexible and optimal routes in continental-sized road networks is a thought challenge.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical problems.
-A* is an extension to Dijkstra's algorithm.
-It uses a potential function to estimate the distance of each node to the target.
-By adding these estimates to the queue keys, the search is directed towards the target.
-The quality of this potential determines the performance of A*.
-We introduce a novel way to efficiently calculate perfect potentials for extended problem settings where a lower bound graph is available.
-For example, in the case of routing with live traffic, this could be the free flow graph.
+A* is the classical approach to accelerate Dijkstra's algorithm used in many fields such as Experimental Algorithms, Robotics, or Artificial Intelligence.
+However, A*'s performance depends on the availability of a good potential, also called heuristic, function.
+Computing a good potential is a challenge of its own.
+In the past decade, many hierachical speedup techniques have therefore been developped.
+They achieve speed and optimality but sacrifize flexibility.
+In this paper, we use hierarchical techniques to compute potential functions.
+With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
 \end{abstract}
@@
 \label{sec:intro}
+Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
+It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
+Efficiently computing flexible and optimal routes is a thought challenge.
+Optimal means that the computed path always has minimum cost.
+Being able to change the edge weights is flexibility.
+Efficiently means that the path computation is fast.
+
+Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately runs for several seconds on continental road networks, which is too slow for many applications.
+A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
+While flexible, A*'s performance crucially depends on the quality of the potential or heuristic function.
+Finding such a function is difficult.
+A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
+A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{?} and its Live-Traffic capable Customizable CH (CCH) \cite{?} extension.
+While undoubtedly fast, hierarchical approaches sacrifice flexibility to obtain speed.
+In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
+
+Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
+The novel idea is to compute optimal A* potentials lazily.
+The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
+As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
+With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
+
+TODO: Cite Sabine
+
+Our technique can be applied to a CH and a CCH.
+We describe our work mostly in CH terms.
+Fortunately, all results are also directly applicable to CCH.
+We describe the trade-off between the two variants in Section \ref{TODO}.
+
+\section{Applications - or why do we care about Flexibility?}
+
+\section{Outline}
+
+\section{Flexible Two-Phase Problem Formulation}
+
+Route planning algorithms usually operate in two or more phases.
+The setup details vary among papers as different problem formulations are considered.
+In this section, we describe the setup used in our work.
+
+We consider two phases: a preprocessing and a query phase.
+In a slow, offline \emph{preprocessing} phase, a graph $G$ with nodes $V$ and edges $E$ is given with \emph{lower bound edge weights} $w_\ell : E\rightarrow \mathbb{R}^+$.
+In this phase, auxilary data is computed which is used in a second, fast query phase.
+
+The input to the \emph{query phase} is a start or \emph{source node} $s$ and a destination or \emph{target node} $t$.
+Further, the input to the preprocessing phase and the computed auxilary data is available.
+Finally, a \emph{query weight function} $w:E\rightarrow \mathbb{R}^+$ is part of the query input.
+It must hold that $w_\ell(e) \le w(e)$ for all edges $e$.
+The output of the query phase is a path such that the sum over all $w$-weights is minimum.
+
+Variations of this setup exist.
+A very common assumption is that the query edge weights $w$ are known during the preprocessing phase.
+This allows for very fast query running times, however, changing $w$ requires rerunning the slow preprocessing phase.
+This setup is therefore very unflexible.
+Many hierarchical speedup techniques such as CH use this setup.
+Notable exceptions are \cite{MLD,CRP} and \cite{CCH} which use a third phase to change the query edge weights.
+
+\section{A*}
+
+A* star is a well-known extension to Dijkstra's algorithm.
+It uses a potential $\pi:V\rightarrow \mathbb{R}^+$.
+The potential is often called heuristic.
+A* is optimal, if the potential $\pi(v)$ is consistent \cite{Pearl, Judea (1984).
+Heuristics: Intelligent Search Strategies for Computer Problem Solving.
+Addison-Wesley.
+ISBN 0-201-05594-5.}, which essentially mean that $\pi(v)$ is a lower bound of the distance from $v$ to $t$.
+
+The best possible potential that can be derived from the auxilary data is the minimum path distance with respect to $w_\ell$.
+Every better, optimal potential requires knowledge of the query weights $w$.
+Our technique allows to compute these best possible potentials.
+Within this regard, CH-potentials are optimal.
+
+\section{CH}
+
+\begin{algorithm2e}
+\KwData{$B[x]$ tentative distance from $x$ to $t$, initially $+\infty$}
+\KwData{Mininum priority queue $Q$, also called open list, initially empty}
+Add $t$ to $Q$ with weight 0\;
+$B[t] \leftarrow 0$\;
+\While{not $Q$ empty}{
+	$y\leftarrow$ pop minimum element from $Q$\;
+	\For{$xy$ is down-edge in $G^+$}{
+		\If{$B[x] > w_\ell(xy) + B[y]$}{
+			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
+			\eIf{not $x$ in $Q$}{
+				Add $x$ to $Q$ with weight $w_\ell(xy) + B[y]$\;
+			}{
+				Decrease weight of $x$ in $Q$ to $w_\ell(xy) + B[y]$\;
+			}
+		}
+	}
+}
+\caption{CH backward search}
+\label{algo:ch-backward}
+\end{algorithm2e}
+
+A CH is a two phase technique to efficiently compute optimal, shortest paths.
+For a detailed exposition, we refer the interested reader to \cite{CCH, CH}.
+In this section, we give a high-level overview of the aspects relvant to this paper.
+
+A CH places nodes into levels.
+No edge must connect two nodes within one level.
+Levels are ordered by ``importance''.
+An edge goes up when it goes from a node in a lower level to higher level.
+Down edges are defined analogously.
+An up-down path is a path where only one node $v$ is more important than its neighbors.
+$v$ is called the \emph{mid} node.
+
+In the preprocessing phase, a CH adds shortcut edges to the input graph $G$ to obtain $G^+$.
+The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path.
+We may therefore restrict our search to up-down paths.
+This is done using a bidirectional search.
+The forward search starts from $s$ and only follows up-edges.
+Similarily, the backward search starts at $d$ and only follows down-edges in reversed direction.
+The two searches meet in the mid node.
+Pseudo-code for the backward search is presented in Listing~\ref{algo:ch-backward}.
+The forward search works analogously.
+
+A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
+On road networks, this is the case.
+
+Using a CH, we directly obtain a basic CH-potential implementation.
+We compute a CH with the lower bound edge weights $w_\ell$ in the preprocessing phase.
+In the query phase, we run a classical $A^*$ search and execute in each potential function evaluation a CH query.
+
+This works, however, while one CH query is fast, answering one for every node in the $A^*$ is slow.
+Fortunatly, we can do better.
+However, for this we need another component called Phast \cite{Phast}.
+
+\section{Phast}
+
+\begin{algorithm2e}
+\KwData{$P[x]$ tentative distance from $x$ to $t$}
+Execute Algorithm~\ref{algo:ch-backward}\;
+\For{all CH levels $L$ from most to least important}{
+	\For{all up edges $xy$ in $G^+$ with $x$ in $L$}{
+		\If{$P[x] < P[y] + w_\ell(xy)$}{
+			$P[x] \leftarrow P[y] + w_\ell(xy)$\;
+		}
+	}
+}
+\caption{Phast basic all-to-one search}
+\label{algo:phast}
+\end{algorithm2e}
+
+Phast \cite{Phast} is a CH extension that computes distances from all nodes to one target node.
+It works in two steps.
+The first step is analogously to the CH query:
+From the target node $t$, all reachable nodes via reversed down-edges are explored.
+The second step iterates over all CH level from top to bottom.
+In each iteration, all up-edges starting within the current level are followed in reverse.
+After all level are processed, the distances from all nodes towards $t$ is computed.
+Phast is faster than Dijkstra's algorithm on road graphs because Phast is a better fit for modern processor architecutres.
+We refer to \cite{Phast} for an in-depth experimental performance analysis.
+Pseudo-code is provided in Listing~\ref{algo:phast}.
+
+Using Phast, we can build another basic CH-potential version.
+In the query phase, we first use Phast to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
+In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.
+
+This setup also works.
+The potential evaluation and by extension the $A^*$ search is indeed fast.
+However, Phast step before the search is comparatively expensive.
+The reason for this is that the distance from all nodes towards $t$ are computed.
+Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.
+
+\section{CH-Potentials}
+
+\begin{algorithm2e}
+\KwData{$B[x]$ tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$P[x]$ memoized potential at $x$, $\bot$ if uncomputed}
+\SetKwFunction{Pot}{Potential}
+\SetKwProg{Fn}{Function}{:}{}
+\Fn{\Pot{$x$}}{
+	\If{$P[x] = \bot$}{
+		$P[x]\leftarrow B[x]$\;
+		\For{all up edges $xy$ in $G^+$}{
+			\If{$P[x] > w_\ell(xy)+\Pot(y)$}{
+				$P[x] \leftarrow w_\ell(xy)+\Pot(y)$\;
+			}
+		}
+	}
+	\Return{$P[x]$}\;
+}
+\caption{CH-Potentials Algorithm}
+\label{algo:pot}
+\end{algorithm2e}
+
+Fortunately, the Phast computation can be done lazily using memoization as depicted in Listing~\ref{algo:pot}.
+If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes reachable via up-edges in $G^+$.
+The computed potentials are then increased by the distance of the corresponding up-edge in $G^+$.
+The potential of a node $x$ is the minimum over all these values and the distance found in the CH backward search.
+
+\section{TopoCore}
+
+TODO: Below is old
 

### 2019-12-30 paper/ch_potentials.tex
foo

@@
 Efficiently computing flexible and optimal routes in continental-sized road networks is a thought challenge.
-While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical problems.
-A* is the classical approach to accelerate Dijkstra's algorithm used in many fields such as Experimental Algorithms, Robotics, or Artificial Intelligence.
+While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
+A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, or Artificial Intelligence.
 However, A*'s performance depends on the availability of a good potential, also called heuristic, function.
@@
 
-Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately runs for several seconds on continental road networks, which is too slow for many applications.
+Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
 A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
-While flexible, A*'s performance crucially depends on the quality of the potential or heuristic function.
+While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
 Finding such a function is difficult.
@@
 In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
-
 Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
@@
 With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
+Figure~\ref{img:search-space} depicts an example  A* search with CH-Potentials.
 
-TODO: Cite Sabine
+\begin{figure}
+TODO
+
+\caption{Example search space when TODO: what parametrization?}
+\label{img:search-space}
+\end{figure}
 
@@
 
-\section{Applications - or why do we care about Flexibility?}
+\section{Applications - Or why do we care about Flexibility?}
 
-\section{Outline}
+Modelling navigation as shortest path problem in graphs with scalar weights is a good but not perfect formalization.
+Several important effects are ignored.
+In this section, we list a few to motivate why flexibility is important.
 
-\section{Flexible Two-Phase Problem Formulation}
+The travel time across an edge is in general not constant over time.
+Live-traffic and historic traffic patterns are examples of how the travel time can change.
+Another example are road blockages.
+These are usually modelled by setting travel times to infinity.
+Some blockage types, such as accidents, are similar to live traffic.
+Other blockages such as reversible one-ways or time-dependent turn-restrictions or seasonal closures are more similar to historic traffic patterns.
+
+Flexibility is also necessary to account for driver preferences.
+Some drivers want to avoid highways whereas other drivers want to avoid city centers.
+Even other drivers are willing to make a detour to avoid certain high crime areas.
+These requirements are usally modelled by modifying the edge weights and are highly depend on the driver.
+Sometimes they even depend on his current mood and might change while driving.
+
+Trucks are another class of vehicles with diverse and specific needs.
+For example, trucks are not allowed in some areas depending on the time of the day.
+Some truck carry hazardous materials and are therefore not allowed in some areas because of environmental protection laws.
+Once the truck has delivered his goods, he may again pass the area.
+
+The reasons why flexibility is a necessary property of a routing engine are sheer endless.
+However, we need to make some assumptions to be able to design algorithms.
+We describe these in the next section.
+
+\section{Formalization}
+
+Throughout this paper, we assume that we are given a directed graph $G$ with nodes $V$ and edges $E$.
+Edge weights are given by a function $w_q:\mathbb{R}^+ \times E\rightarrow \mathbb{R}^+$.
+The first parameter to $w_q$ are the costs of the path when an edge is reached.
+This is necessary to support time-dependent effects such as historic traffic patterns or time-dependent restriction.
+We refer to the weights as \emph{query edge weights}.
+Troughout this paper, we assume that the weights fulfill the FIFO property.
+This is necessary to prevent the problem setting from becoming NP-hard \cite{Veit}.
+It states that $w_q(\tau,e)+h\le w_q(\tau+h,e)$ for all $\tau$, $h\in \mathbb{R}^+$, and $e\in E$.
+Informally, it states that waiting at a node is never beneficial.
+We further assume that \emph{lower bound edge weights} $w_\ell : E\rightarrow \mathbb{R}^+$ are available such that $w_\ell(e)\le w_q(\tau,e)$ for all $e$ and $\tau$.
+
+A path is sequence of connected edges from a \emph{source node} $s$ to a \emph{target node} $t$.
+Paths have cost that depends on the source costs $c\in \mathbb{R}^+$.
+We define the costs $C(P)$ of a path $e_1,e_2\ldots$ recursively as $C(e_1)=w_q(c,e_1)$ and $C(e_1\ldots c_k)=C(e_1\ldots c_{k-1})+w_q(C(e_1\ldots c_{k-1}),e_k)$.
+If the weights of an edge are independent of the other weights along the path and the source costs are zero, then the path costs boil down to the classical $C(P) = w_q(e_1)+w_q(e_2)+\ldots$ expression.
 
 Route planning algorithms usually operate in two or more phases.
-The setup details vary among papers as different problem formulations are considered.
+The exact details vary among papers as different problem formulations are considered.
 In this section, we describe the setup used in our work.
 
-We consider two phases: a preprocessing and a query phase.
-In a slow, offline \emph{preprocessing} phase, a graph $G$ with nodes $V$ and edges $E$ is given with \emph{lower bound edge weights} $w_\ell : E\rightarrow \mathbb{R}^+$.
+We consider two phases: a \emph{preprocessing} and a \emph{query} phase.
+In a slow, offline preprocessing phase, $G$ and $w_\ell$ are given.
+$s$, $t$, $c$, and $w_q$ are unknown.
 In this phase, auxilary data is computed which is used in a second, fast query phase.
 
-The input to the \emph{query phase} is a start or \emph{source node} $s$ and a destination or \emph{target node} $t$.
-Further, the input to the preprocessing phase and the computed auxilary data is available.
-Finally, a \emph{query weight function} $w:E\rightarrow \mathbb{R}^+$ is part of the query input.
-It must hold that $w_\ell(e) \le w(e)$ for all edges $e$.
-The output of the query phase is a path such that the sum over all $w$-weights is minimum.
+The input to the \emph{query phase} is a $s$, $t$, $c$, and $w_q$.
+Further, the input to the preprocessing phase and the computed auxilary data is also available to the query phase.
+The output of the query phase is a minimum cost path.
+
+The idea is that the preprocessing phase is performed once per map release.
+The computed auxilariy data is shared among many drivers and vehicles.
+
+The query phase is executed each time a driver requests a route.
+It accounts for the current road situation and the current driver preferences.
+In theory, it is possible that a driver changes his perferences for every query.
+
+\section{Related Work}
+
+TODO: Dijkstra's algorithm \cite{Dijkstra's} can be used to solve the stated problem formalization \cite{Veit?}.
+
+TODO: CRP was motivated by personalization.
 
 Variations of this setup exist.
-A very common assumption is that the query edge weights $w$ are known during the preprocessing phase.
+A very common assumption is that the query edge weights $w_q$ are known during the preprocessing phase.
+In this setup, $w_\ell$ does not exist.
 This allows for very fast query running times, however, changing $w$ requires rerunning the slow preprocessing phase.
 This setup is therefore very unflexible.
-Many hierarchical speedup techniques such as CH use this setup.
+Many hierarchical speedup techniques such as the original CH \cite{CH} use this setup.
 Notable exceptions are \cite{MLD,CRP} and \cite{CCH} which use a third phase to change the query edge weights.
+
+TODO: Cite Sabine VLDB and TopoCore
+
+TODO: Alt
+
+ALT~\cite{gh-cspas-05} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
+For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
+This achieves decent speed-ups and can be used in other problem settings.
+However, the performance is not competitive to hierarchical approaches.
 
@@
 A* star is a well-known extension to Dijkstra's algorithm.
-It uses a potential $\pi:V\rightarrow \mathbb{R}^+$.
+It uses a potential function $\pi:V\rightarrow \mathbb{R}^+$.
 The potential is often called heuristic.
@@
 The best possible potential that can be derived from the auxilary data is the minimum path distance with respect to $w_\ell$.
-Every better, optimal potential requires knowledge of the query weights $w$.
+Every better, consistent potential requires knowledge of the query weights $w_q$, which is unavailable during preprocessing.
 Our technique allows to compute these best possible potentials.
-Within this regard, CH-potentials are optimal.
+Within this regard, CH-potentials are \emph{optimal A* potentials}.
 
@@
 For a detailed exposition, we refer the interested reader to \cite{CCH, CH}.
-In this section, we give a high-level overview of the aspects relvant to this paper.
+In this section, we give a high-level overview of the aspects relevant to this paper.
 
@@
 Levels are ordered by ``importance''.
+The intuition is that dead-ends are unimportant while highway bridges are very important.
 An edge goes up when it goes from a node in a lower level to higher level.
 Down edges are defined analogously.
-An up-down path is a path where only one node $v$ is more important than its neighbors.
+An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
 $v$ is called the \emph{mid} node.
+An \emph{up path} is a path where the last node is the mid node.
+Similarily, the first node is the mid node of a \emph{down path}.
+Every up and down path is an up-down path.
 
-In the preprocessing phase, a CH adds shortcut edges to the input graph $G$ to obtain $G^+$.
+In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
 The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path.
 We may therefore restrict our search to up-down paths.
-This is done using a bidirectional search.
+This search is performed using a bidirectional search.
 The forward search starts from $s$ and only follows up-edges.
-Similarily, the backward search starts at $d$ and only follows down-edges in reversed direction.
+Similarily, the backward search starts at $t$ and only follows down-edges in reversed direction.
 The two searches meet in the mid node.
@@
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
-On road networks, this is the case.
+On road networks, this is the case as demonstrated in \cite{ch,punch,crp,cch,flowcutter}.
 
-Using a CH, we directly obtain a basic CH-potential implementation.
-We compute a CH with the lower bound edge weights $w_\ell$ in the preprocessing phase.
-In the query phase, we run a classical $A^*$ search and execute in each potential function evaluation a CH query.
+Using a CH, we can easily compute optimal A* potentials.
+In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
+The query phase consists of an A* search, where the evaluation of the potential $\pi(v)$ performs a CH-query from $v$ to $t$.
 
-This works, however, while one CH query is fast, answering one for every node in the $A^*$ is slow.
+This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
 Fortunatly, we can do better.
@@
 Phast \cite{Phast} is a CH extension that computes distances from all nodes to one target node.
-It works in two steps.
+First the preprocessing phase is executed analogously to the original CH.
+The query phase is divided into two steps.
 The first step is analogously to the CH query:
 From the target node $t$, all reachable nodes via reversed down-edges are explored.
+Listing~\ref{algo:ch-backward} shows this first step.
 The second step iterates over all CH level from top to bottom.
@@
 After all level are processed, the distances from all nodes towards $t$ is computed.
-Phast is faster than Dijkstra's algorithm on road graphs because Phast is a better fit for modern processor architecutres.
+Phast is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architecutres.
+Phast's main advantage is that is can be more easily parallelized than Dijkstra's algorithm.
+However, we will not consider parallelization in this paper.
 We refer to \cite{Phast} for an in-depth experimental performance analysis.
@@
 
-Using Phast, we can build another basic CH-potential version.
+Using Phast, we can compute optimal A* potentials in another way.
 In the query phase, we first use Phast to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
@@
 Fortunately, the Phast computation can be done lazily using memoization as depicted in Listing~\ref{algo:pot}.
-If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes reachable via up-edges in $G^+$.
-The computed potentials are then increased by the distance of the corresponding up-edge in $G^+$.
-The potential of a node $x$ is the minimum over all these values and the distance found in the CH backward search.
 
-\section{TopoCore}
+In a first step, we run the backward CH search from $t$ to obtain an array $B$.
+$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
+$B$ is computed as shown in Listing~\ref{algo:ch-backward}.
 
-TODO: Below is old
+Suppose, we need to compute the potential $\pi(v)$ of a node $v$.
+We do this by first recursively computing the potentials $\pi(u)$ of all nodes for which an up-edge $(v,u)$ exists.
+Afterwards, we compute using $\min_u\{w_\ell(v,u) + \pi(u)\}$ the minimum distance over all up-down paths that contain at least one up-edge.
+As the minimum distance up-down path does not necessarily have to contain an up-edge, we set $\pi(v) = \min \{ B[v], \min_u\{w_\ell(v,u) + \pi(u)\} \}$.
+This calculation is correct, as it computes the minimum up-down $vt$-path distance in $G^+$, which corresponds to the minimum $vt$-path distance in a CH.
 
-Fast routing in huge street networks has seen a lot of research \cite{bdgmpsww-rptn-16}.
-For static, time- and user-invariant networks a lot of techniques exist.
-They work in two phases.
-The first phase is called preprocessing phase and is a slow index-construction.
-In this phase the road graph and the edge weights are known.
-In the second phase, called query phase, minimum weight paths are computed.
-The query phase has access to the precomputed index and is very fast.
+The resulting algorithm is the basic CH-potential algorithm.
+It is flexible and optimal and slightly faster than Dijkstra's algorithm.
+However, we can still improve the performance.
+A quick investigation reveals, that most of the running time is spent in the potential evaluation.
+We therefore investigate methods to evaluate fewer potentials.
 
-One of the most successful ones are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
-Contraction Hierarchies exploit inherent hierarchies in the shortest-path structure of road networks.
-Using CHs, Queries can be answered on the order of only a millisecond.
-However, CHs can not always be easily extended to more complicated problem settings.
-% TODO what problem settings
-While such extensions exist~\cite{dsw-cch-15,bgsv-mtdtt-13}, they often require substantial engineering effort.
+\section{Largest Biconnected Component}
 
-On the other hand, goal directed speed-up techniques based on A* are usually slower but more flexible.
-The performance of these approaches depends on the quality of the potential functions used for the A* search.
-In this paper, we propose an algorithm which utilizes Contraction Hierarchies to efficiently calculate perfect (best possible) potentials.
-This allows us to retain the flexibility of A* based approaches but use much more effective potentials than previously possible.
-% These techniques are used in online navigation services.
-% In these services, there is a single index.
-% With this index a large number of user requests are answered.
-% The preprocessing phases is rerun on a regular interval to incorporate map updates.
+A lot of nodes in road networks lead to dead-ends.
+Unless the source or target is is in a dead-end, it is uncessary to explore these nodes.
+We employ an idea from \cite{TOPOCORE} to avoid these nodes entirely.
+This reduces the number of explored nodes and thus fewer potentials are evaluated.
 
-% While highly scalable, this setup has the disadvantage, that every user must have use the same weights.
-% In this paper, we explore an novel algorithm, called CH-Potentials, that uses the two phase setup, however, allows the weights to be modified on a per query basis.
-% Applications for this algorithm are personalized user settings such as for example:
-% Avoiding highways, capping the maximum speed, or avoid user-configurable areas.
+In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
+We do this using Tarjan's algorithm \cite{Tarjan's algorithm}.
+For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
+For nodes in the core, $a_v=v$.
 
-\paragraph*{Related Work}
+The query phase is divided into three steps.
+In the first step, we search a path from $s$ to $a_s$ using Dijkstra's algorithm restricted to $s$'s biconnected component.
+The second step, consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
+Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
+The final path is the concatenation of all three path parts.
 
-% We are not the first to investigate this problem setting.
-% Two approaches exist in the literature.
+The resulting algorithm is an improved CH-potential variant that visits fewer nodes and thus evaluates fewer potentials.
+An added advantage is that the A* search no longer explores unnecessary dead-ends.
+The resulting algorithm is fast, but can still be improved.
 
-% The three phase approach splits the preprocessing phase into two.
-% Techniques that use this setup are MLD/CRP \cite{?} and CCH \cite{?}.
-% The first phase is a weight-independent preprocessing phase that analsyse the graph topology.
-% The second phase is called customization phase and introduces the weights into the index.
-% The third phase is the query phase that computes paths.
-% The preprocessing phase is very slow.
-% The customization phase runs on the order of seconds.
-% The query phase runs within milliseconds.
-
-% The three phase approach is ideal to incorporate weight changes that apply to all users.
-% Live traffic information is a prime example.
-% It has been suggested to use the three phase approach to handle user-specific requirements by storing a copy of the index per user.
-% The problem with this approach is that it binds a lot of server-side resources per user.
-% Further, if the user often changes his weights, the customization needs to run often.
-% Every customization requires seconds of running time.
-% Just applying Dijkstra's algorithm to the input graph results in similar running times.
-
-% The second approach is Sabine's VLDB personalized routing.
-% Also TopoCore~\cite{?}.
-
-A plethora of of research exists in the area of efficient routing in road networks \cite{bdgmpsww-rptn-16}.
-Here, we focus on the algorithms directly relevant to this work.
-
-At the core is the algorithm of Dijkstra~\cite{d-ntpcg-59}.
-This algorithm explores the graph visiting nodes ordered by increasing distance from a source node $s$.
-For each node $v$, a tentative distance $d_s(v)$ is maintained, which is initially set to $\infty$.
-In each iteration the node $u$ with the smallest remaining distance $d_s(u)$ is extracted from a priority queue and settled.
-For outgoing edge $(u,v)$ of $u$ the algorithms checks if $d_s(u) + w(u,v) < d_s(v)$.
-If so, the distance and the queue position of $v$ is updated.
-This process is denoted as edge relaxation.
-Once the target node $t$ is settled, the shortest path and distance between $s$ and $t$ is known.
-
-A*~\cite{hnr-afbhd-68} extends Dijkstra's algorithm by changing the order in which nodes are visited.
-Nodes in the direction of $t$ should be visited earlier.
-This is achieved through a potential function $p$ estimating the distance to $t$.
-The priority queue is ordered by $d_s(u) + p(u)$.
-
-ALT~\cite{gh-cspas-05} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
-For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
-This achieves decent speed-ups and can be used in other problem settings.
-However, the performance is not competitive to hierarchical approaches.
-
-Contraction Hierarchies~\cite{gssv-erlrn-12} are a popular hierarchical speed-up technique.
-During preprocessing, all nodes are ranked by some measure of importance.
-Nodes which lie on many shortest paths are important.
-Then, nodes are contracted iteratively by increasing rank.
-To contract a node $v$ means removing it form the graph, but preserving shortest distances among all other nodes.
-This is achieved by inserting shortcut edges between the neighbors of $v$ which have the length of the path over $v$.
-The query is a bidirectional Dijkstra starting from $s$ and $t$ where only edges to higher ranked nodes are relaxed.
-
-PHAST~\cite{dgnw-phast-13} is an algorithm for one-to-many queries based on Contraction Hierarchies.
-The preprocessing phase remains the same.
-The query begins with a forward upward search from $s$.
-The backward search now needs to determine distances to \emph{all} nodes.
-To achieve this, PHAST relaxes all edges $(u,v)$ in the backward graph where $rank(u) > rank(v)$ ordered descending by $rank(v)$.
-Having processed all edges, distances to all nodes are known.
-PHAST achieves good running times by heavily exploiting memory locality:
-Nodes and edges are reordered such that relaxing all downward edges is just a linear sweep over the edge array.
-
-% \section{Preliminaries}
-% \label{sec:prelim}
-
-\section{Perfect Potentials}
-\label{sec:algo}
-
-Our new algorithm is at its core a combination of A* and PHAST.
-We take the preprocessing from CHs and PHAST and apply it to a lower bound graph.
-The query is an A* variant.
-It uses exact distances on the lower bound graph as potentials and obtains these through a lazy PHAST-like approach.
-% For an additional performance, we further combine it with TopoCore.
-
-The input of the preprocessing phase is a graph and lower-bound weights $w_{\ell}$.
-Denote by $d_{\ell}(x,y)$ the minimum weight $xy$-path distance with respect to $w_{\ell}$.
-How exactly this lower bound graph is obtained depends on the problem setting.
-In the case of routing with life traffic it may be the free flow travel time.
-In the case of time-dependent routing it may be the lower bound of each travel time function.
-The preprocessing phase consists of computing a CH of the graph with lower-bound weights.
-
-The input to the query phase is a source $s$ and a target $t$ node and weights $w_u$.
-These weights $w_u$ must not be smaller than the lower-bound weights $w_{\ell}$, i.e., $w_{\ell}(e) < w_u(e)$ for all edges $e$.
-The weights may be time-dependent travel time predictions, incorporate live traffic or user specific preferences.
-Analogous to $d_{\ell}(x,y)$, $d_u(x,y)$ is the minimum-weight phase according to $w_u$.
-We obtain $d_u(x,y)$ through an A* search with respect to $d_u$ from $s$ to $t$.
-The distance $d_{\ell}(x,t)$ from a node $x$ to $t$ is used as potential.
+\section{Skipping Degree Two Chains}
 
 \begin{algorithm2e}
-\KwData{$B[x]$ distance from $s$ to $x$ compute in backward search, or $+\infty$ if unreachable}
-\KwData{$P[x]$ memoized potential for $x$, or $\bot$ if not yet computed. Equal to $d_{\ell}(x,t)$}
-\SetKwFunction{Pot}{Potential}
-\SetKwProg{Fn}{Function}{:}{}
-\Fn{\Pot{$x$}}{
-  \If{$P[x] = \bot$}{
-    $P[x]\leftarrow B[x]$\;
-    \For{$(x,y)$ is upward CH edge with weight $w_{xy}$}{
-      $P[x]\leftarrow \min\{P[x],w_{xy}+\Pot(y)\}$\;
+\KwData{$D[x]$ tentative distance from $s$ to $x$, initially $+\infty$}
+\KwData{$Q$ minimum priority queue of nodes}
+$D[a_s]\leftarrow 0$\;
+Add $a_s$ to $Q$ with weight $\pi(a_s)$\;
+\While{not $Q$ empty}{
+  $x\leftarrow$ pop node from $Q$\;
+  \If{$x = a_t$}{finished\;}
+  \For{all edges $(x,y)$ in the core graph $G_C$}{
+    $\alpha\leftarrow x$;
+    $\beta\leftarrow y$;
+    $w\leftarrow w_q(\alpha,\beta)$\;
+    \While{undirected degree of $\beta$ is 2 and $z\neq a_t$}{
+      $\gamma\leftarrow$ neighbor of $\beta$ in $G_C$ different from $\alpha$\;
+      $w\leftarrow w+w_q(\beta,\gamma)$;
+      $\alpha\leftarrow \beta$;
+      $\beta\leftarrow \gamma$\;
+    }
+    \If{$D[x]+w < D[\beta]$}{
+        $D[\beta]\leftarrow D[x]+w$\;
+    	Add $\beta$ to Q with weight $D[\beta]+\pi(\beta)$\;
     }
   }
-  \Return{$P[x]$}\;
 }
-
-\caption{Algorithm to compute Potentials.}
-\label{algo:pot}
-
+\caption{A* star search in the core skipping degree two nodes}
+\label{algo:skip-a-star}
 \end{algorithm2e}
 
-Our core contribution consist in the way $d_{\ell}(x,t)$ is efficiently computed.
-For this, we use a PHAST variant.
-Before starting the A* search, we explore all nodes backward-reachable from $t$ in the CH.
-We store the computed distances in the array $B$.
-We then execute the A* search.
-Potentials are computed recursively, on-the-fly, with memoization as depicted in Algorithm \ref{algo:pot}.
-If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes in the CH backward search graph.
-These potentials are then increased by the distance in the CH backwards search graph.
-The potential of a node $x$ is the minimum over all these values and the distance found in the CH forward search.
+The core graph $G_C$ contains a lot of degree two nodes \cite{TOPOCORE}.
+At such nodes no directional decision is made.
+It is therefore unecessary to evaluate the potentials at these nodes.
 
-% Our algorithm can be applied as described to the input graph.
-% However, depending on the weights, the A* search can get stuck often in deadends.
-% We therefore first coarsen the input graph using TopoCore and apply the above described algorithm to the core.
-% In this setup, TopoCore handles all deadends.
+We exploit this intuition by modifying the A* search.
+When exploring the neighbors $y$ of node $x$, we follow chains of degree two nodes until a node $\beta$ with degree at least three is found.
+Instead of $y$, we add $\beta$ to the queue.
+No queue operations nor potential evaluations are required for the intermediate degree two nodes.
 
-% \section{Applications}
-% \label{sec:applications}
+\section{Skipping Trees}
 
-% \subsection{Live Traffic}
-% \subsection{Time-dependent Routing}
-
-% \section{Experiments}
-% \label{sec:exp}
+TODO @ Tim, bringt das wirklich was?
 

### 2020-04-08 paper/ch_potentials.tex
foo

@@
 \begin{abstract}
-Efficiently computing flexible and optimal routes in continental-sized road networks is a thought challenge.
+Efficiently computing flexible and optimal routes in continental-sized road networks is a tough challenge.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
@@
 Computing a good potential is a challenge of its own.
-In the past decade, many hierachical speedup techniques have therefore been developped.
+To avoid the need for a potential, many hierachical speedup techniques have been developped.
 They achieve speed and optimality but sacrifize flexibility.
@@
 Some blockage types, such as accidents, are similar to live traffic.
-Other blockages such as reversible one-ways or time-dependent turn-restrictions or seasonal closures are more similar to historic traffic patterns.
+Other blockages such as reversible one-ways, time-dependent turn-restrictions, or seasonal closures are more similar to historic traffic patterns.
 
 Flexibility is also necessary to account for driver preferences.
-Some drivers want to avoid highways whereas other drivers want to avoid city centers.
-Even other drivers are willing to make a detour to avoid certain high crime areas.
+For example, common options in navigation devices include avoiding highways or tunnels.
 These requirements are usally modelled by modifying the edge weights and are highly depend on the driver.
@@
 
-The reasons why flexibility is a necessary property of a routing engine are sheer endless.
-However, we need to make some assumptions to be able to design algorithms.
-We describe these in the next section.
+\section{Problem Setting}
 
-\section{Formalization}
+As detailed in the previous section, the reasons why flexibility is a necessary property of a routing engine are sheer endless.
+However, a formal problem specification that encompasses all these options is complex and well beyond the scope of this paper.
+For an example of such a formulation, we refer to \cite{sea2020paper} which covers a truck-specific application.
+The algorithm presented in \cite{sea2020paper} uses a preliminary version of CH-Potentials. 
+For simplicity, we limit our exposition to the classical shortest path formulation.
+To support algorithm extensions, we avoid two operations that are adverse to generalization.
+These are: 
+(1) traversing an input edge in backward direction, and
+(2) combining input weights.
+Backward traversal is problematic as a weight can depend on the moment that an edge is entered, which is unknown in a backward search.
+Combining input weights can be difficult, if the input weights are more complex objects than scalar values.
+For example, when considering historic traffic patterns, the input weights are usually functions \cite{citation needed}.
+
+Our algorithm has one major restriction, it requires the presence of an inflexible scalar weight lower bound for every input edge.
+The performance of our algorithm depends on this lower bound being close to the actual weight.
+Using the trivial lower bound of zero everywhere, CH-potentials degenerate to an engineered variant of Dijkstra's algorithm.
 
 Throughout this paper, we assume that we are given a directed graph $G$ with nodes $V$ and edges $E$.
-Edge weights are given by a function $w_q:\mathbb{R}^+ \times E\rightarrow \mathbb{R}^+$.
-The first parameter to $w_q$ are the costs of the path when an edge is reached.
-This is necessary to support time-dependent effects such as historic traffic patterns or time-dependent restriction.
-We refer to the weights as \emph{query edge weights}.
-Troughout this paper, we assume that the weights fulfill the FIFO property.
-This is necessary to prevent the problem setting from becoming NP-hard \cite{Veit}.
-It states that $w_q(\tau,e)+h\le w_q(\tau+h,e)$ for all $\tau$, $h\in \mathbb{R}^+$, and $e\in E$.
-Informally, it states that waiting at a node is never beneficial.
-We further assume that \emph{lower bound edge weights} $w_\ell : E\rightarrow \mathbb{R}^+$ are available such that $w_\ell(e)\le w_q(\tau,e)$ for all $e$ and $\tau$.
+Every edge has a lower bound weight $w_\ell:E\rightarrow \mathbb{R}^+$ and a query weight $w_q:E\rightarrow \mathbb{R}^+$.
+It must hold that $0\le w_\ell(e)\le w_q(e)$.
+We are further given a source node $s$ and a target node $t$.
 
-A path is sequence of connected edges from a \emph{source node} $s$ to a \emph{target node} $t$.
-Paths have cost that depends on the source costs $c\in \mathbb{R}^+$.
-We define the costs $C(P)$ of a path $e_1,e_2\ldots$ recursively as $C(e_1)=w_q(c,e_1)$ and $C(e_1\ldots c_k)=C(e_1\ldots c_{k-1})+w_q(C(e_1\ldots c_{k-1}),e_k)$.
-If the weights of an edge are independent of the other weights along the path and the source costs are zero, then the path costs boil down to the classical $C(P) = w_q(e_1)+w_q(e_2)+\ldots$ expression.
-
-Route planning algorithms usually operate in two or more phases.
-The exact details vary among papers as different problem formulations are considered.
-In this section, we describe the setup used in our work.
-
-We consider two phases: a \emph{preprocessing} and a \emph{query} phase.
-In a slow, offline preprocessing phase, $G$ and $w_\ell$ are given.
-$s$, $t$, $c$, and $w_q$ are unknown.
-In this phase, auxilary data is computed which is used in a second, fast query phase.
-
-The input to the \emph{query phase} is a $s$, $t$, $c$, and $w_q$.
-Further, the input to the preprocessing phase and the computed auxilary data is also available to the query phase.
-The output of the query phase is a minimum cost path.
-
-The idea is that the preprocessing phase is performed once per map release.
-The computed auxilariy data is shared among many drivers and vehicles.
-
-The query phase is executed each time a driver requests a route.
-It accounts for the current road situation and the current driver preferences.
-In theory, it is possible that a driver changes his perferences for every query.
+Our algorithm uses two phases.
+The first is a slow \emph{preprocessing phase}, where only $G$ and $w_\ell$ are known.
+It computes auxilary data.
+The second phase is the \emph{query phase}.
+It is given the auxilary data and $G$, $w_\ell$, $w_q$, $s$, and $t$.
+The output of the query phase, is a path $e_1,e_2\ldots e_k$ such that $\sum w_q(e_i)$ is minimized.
 
@@
 Addison-Wesley.
-ISBN 0-201-05594-5.}, which essentially mean that $\pi(v)$ is a lower bound of the distance from $v$ to $t$.
+ISBN 0-201-05594-5.}, which essentially means that $\pi(v)$ is a lower bound of the distance from $v$ to $t$.
 
@@
 In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
-The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path.
-We may therefore restrict our search to up-down paths.
+The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
+We may therefore restrict our search to up-down paths in $G^+$.
 This search is performed using a bidirectional search.
@@
 The two searches meet in the mid node.
-Pseudo-code for the backward search is presented in Listing~\ref{algo:ch-backward}.
+Pseudo-code for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
 The forward search works analogously.
@@
 From the target node $t$, all reachable nodes via reversed down-edges are explored.
-Listing~\ref{algo:ch-backward} shows this first step.
+Algorithm~\ref{algo:ch-backward} shows this first step.
 The second step iterates over all CH level from top to bottom.
@@
 After all level are processed, the distances from all nodes towards $t$ is computed.
-Phast is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architecutres.
-Phast's main advantage is that is can be more easily parallelized than Dijkstra's algorithm.
+Phast is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
+Phast's main advantage is that is easily be parallelized.
 However, we will not consider parallelization in this paper.
 We refer to \cite{Phast} for an in-depth experimental performance analysis.
-Pseudo-code is provided in Listing~\ref{algo:phast}.
+Pseudo-code is provided in Algorithm~\ref{algo:phast}.
 
@@
 The potential evaluation and by extension the $A^*$ search is indeed fast.
-However, Phast step before the search is comparatively expensive.
+However, the Phast step before the search is comparatively expensive.
 The reason for this is that the distance from all nodes towards $t$ are computed.
@@
 
-Fortunately, the Phast computation can be done lazily using memoization as depicted in Listing~\ref{algo:pot}.
+Fortunately, the Phast computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
 
@@
 $B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
-$B$ is computed as shown in Listing~\ref{algo:ch-backward}.
+$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.
 
@@
 
+\section{Avoid Placing Nodes in The Queue}
+
+We observe, that a potential evaluation is only necessary for nodes that are placed into the queue.
+We therefore avoid placing nodes into the queue.
+
+In this section, we slightly modify how A* works internally.
+Our modified version maintains the following entities:
+(1) a tentative distance $D$ for every node,
+(2) a queue of nodes $Q$ weighted by $D[v]+\pi(v)$.
+
+Initially, $Q$ only contains $s$.
+$D$ is $+\infty$ everywhere except for $D[s]=0$.
+
+Similarily to the basic A* algorithm, our algorithm continously removes nodes from the queue $Q$.
+Denote by $x$ such a node.
+Contrary to A*, our algorithm does not only consider the immediate outgoing edges of $x$.
+Instead, a queue-less backtracking search at $x$ is started.
+For every node $y$ visited in the backtracking search, the following steps are performed:
+
+If $D[y]$ is less than the distance along the path of the backtracking search, the search is pruned.
+Otherwise, $D[y]$ is reduced.
+
+If $y$ is already in $Q$, then the backtracking search is pruned and the key of $y$ is decreased, if appropriate.
+Otherwise, we do some magic and sometimes continue the search and sometimes prune.
+
+TODO: description incomplete
+
 \section{Largest Biconnected Component}
@@
 A lot of nodes in road networks lead to dead-ends.
-Unless the source or target is is in a dead-end, it is uncessary to explore these nodes.
+Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
 We employ an idea from \cite{TOPOCORE} to avoid these nodes entirely.
@@
 For nodes in the core, $a_v=v$.
+We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
+The basic CH-potential preprocessing step is only executed for $G_C$.
 
-The query phase is divided into three steps.
-In the first step, we search a path from $s$ to $a_s$ using Dijkstra's algorithm restricted to $s$'s biconnected component.
-The second step, consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
+The query phase is divided into four steps.
+In the first step, we use Dijkstra's algorithm to explore the component that contains $s$ until $a_s$ is reached.
+If this search finds $t$, then $s$ and $t$ are part of the same component.
+An optimal path was found and therefore the algorithm stops early.
+Otherwise, in the second step starts.
+It consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
 Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
@@
 
-The resulting algorithm is an improved CH-potential variant that visits fewer nodes and thus evaluates fewer potentials.
-An added advantage is that the A* search no longer explores unnecessary dead-ends.
-The resulting algorithm is fast, but can still be improved.
+TODO: Everything below is old.
 

### 2020-04-09 paper/ch_potentials.tex
foo

@@
 
-\section{A*}
+\section{Goal-Directed Search using A*}
 
@@
 
-\section{CH}
+\section{Contraction Hierarchy (CH)}
 
@@
 
-\section{Phast}
+\section{Potentials using PHAST}
 
@@
 
-TODO: Everything below is old.
-
-\section{Skipping Degree Two Chains}
-
-\begin{algorithm2e}
-\KwData{$D[x]$ tentative distance from $s$ to $x$, initially $+\infty$}
-\KwData{$Q$ minimum priority queue of nodes}
-$D[a_s]\leftarrow 0$\;
-Add $a_s$ to $Q$ with weight $\pi(a_s)$\;
-\While{not $Q$ empty}{
-  $x\leftarrow$ pop node from $Q$\;
-  \If{$x = a_t$}{finished\;}
-  \For{all edges $(x,y)$ in the core graph $G_C$}{
-    $\alpha\leftarrow x$;
-    $\beta\leftarrow y$;
-    $w\leftarrow w_q(\alpha,\beta)$\;
-    \While{undirected degree of $\beta$ is 2 and $z\neq a_t$}{
-      $\gamma\leftarrow$ neighbor of $\beta$ in $G_C$ different from $\alpha$\;
-      $w\leftarrow w+w_q(\beta,\gamma)$;
-      $\alpha\leftarrow \beta$;
-      $\beta\leftarrow \gamma$\;
-    }
-    \If{$D[x]+w < D[\beta]$}{
-        $D[\beta]\leftarrow D[x]+w$\;
-    	Add $\beta$ to Q with weight $D[\beta]+\pi(\beta)$\;
-    }
-  }
-}
-\caption{A* star search in the core skipping degree two nodes}
-\label{algo:skip-a-star}
-\end{algorithm2e}
-
-The core graph $G_C$ contains a lot of degree two nodes \cite{TOPOCORE}.
-At such nodes no directional decision is made.
-It is therefore unecessary to evaluate the potentials at these nodes.
-
-We exploit this intuition by modifying the A* search.
-When exploring the neighbors $y$ of node $x$, we follow chains of degree two nodes until a node $\beta$ with degree at least three is found.
-Instead of $y$, we add $\beta$ to the queue.
-No queue operations nor potential evaluations are required for the intermediate degree two nodes.
-
-\section{Skipping Trees}
-
-TODO @ Tim, bringt das wirklich was?
-
 \section{Conclusion}

### 2020-04-11 paper/ch_potentials.tex
foo

@@
 Efficiently computing flexible and optimal routes is a thought challenge.
-Optimal means that the computed path always has minimum cost.
-Being able to change the edge weights is flexibility.
+Optimal means that the computed path always has minimum weight.
 Efficiently means that the path computation is fast.
+Being able to change the edge weights between path computations is flexibility.
 
@@
 While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
-Finding such a function is difficult.
+Finding such a function that can be computed efficiently is difficult.
 A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
-A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{?} and its Live-Traffic capable Customizable CH (CCH) \cite{?} extension.
-While undoubtedly fast, hierarchical approaches sacrifice flexibility to obtain speed.
+A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{geisberger}.
+There is a lot of work that extends CH to handle more complex settings.
+For example, in \cite{sabine, kobitzsch} multi-criteria optimization is studied.
+In \cite{CCH}, CH is modified to support Live-Traffic.
+Historic traffic patterns are combined with CH in \cite{TD-CCH-preprint, TDCH, TDCRP}.
+Specialized variants for electric vehicle routing are developped in \cite{moritz baum, charge}.
+While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
+Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
+
+While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
 In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
+We are not the first to combine hierarchical approaches and A* \cite{core-alt, real, charge, bdsssw-chgds-08}.
+However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
+Our primary algorithm design objective is to retain A*'s flexibility.
+
 Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
@@
 
-We observe, that a potential evaluation is only necessary for nodes that are placed into the queue.
-We therefore avoid placing nodes into the queue.
+We observe, that a potential evaluation is always paired with a queue push operation.
+By avoiding push operations we can therefore also avoid potential evaluations.
 
@@
 Similarily to the basic A* algorithm, our algorithm continously removes nodes from the queue $Q$.
-Denote by $x$ such a node.
+Denote by $x$ a node removed with the queue.
 Contrary to A*, our algorithm does not only consider the immediate outgoing edges of $x$.
-Instead, a queue-less backtracking search at $x$ is started.
-For every node $y$ visited in the backtracking search, the following steps are performed:
+Instead, a limited, queue-less backtracking search at $x$ is started.
+The idea is to process local tree-like areas of the graph without queue interaction.
+We exploit that backtracking is fast on trees.
 
-If $D[y]$ is less than the distance along the path of the backtracking search, the search is pruned.
-Otherwise, $D[y]$ is reduced.
+The backtracking search continues until either the tentative distance at a node is lower than the length of the path found by the search or a stop node is encountered.
+Every node that was already pushed is a stop node.
+Nodes with degree 4 or more is a stop node (TODO: @Tim out-degree ?).
+Every degree 3 node that is reached via another degree 3 node from the backtracking search start is a stop node.
+If the search was stopped because of the tentative distance, then it is aborted.
+If the search was stopped because a stop node was encountered, the stop node is added to the queue.
+This implies a potential evaluation.
+If the node is already in the queue, a decrease key operation is executed.
+One can view this modifed A* algorithm as a lazy version of \cite{topocore} where the set of virtually contracted nodes is computed at query time.
 
-If $y$ is already in $Q$, then the backtracking search is pruned and the key of $y$ is decreased, if appropriate.
-Otherwise, we do some magic and sometimes continue the search and sometimes prune.
-
-TODO: description incomplete
-
-\section{Largest Biconnected Component}
+\subsection{Largest Biconnected Component}
 
@@
 
+\section{Experiments}
+
+\begin{itemize}
+\item Avoid nothing running times, i.e., what is if $w_q = q_\ell$?
+\item Multiply all weights by $\alpha$ for $\alpha$ in 1.1, 1.5, 2, 3
+\item Avoid highway
+\item Avoid tunnel
+\item Avoid highway + tunnel
+\item MapBox Live Traffic
+\item Ptv-TD traffic? Achtung: Das Fass würde ich nicht aufmachen. Am Ende will ein Reviewer, dass wir avoid highway mit TD kombinieren... das ist NP-schwer
+\item Running times if OSM degree 2 nodes are not extracted!
+\end{itemize}
+
 \section{Conclusion}

### 2020-04-11 paper/ch_potentials.tex
foo

@@
 
-\section{Avoid Placing Nodes in The Queue}
+\section{Avoiding Potential Function Evaluations}
 
@@
 
-\subsection{Largest Biconnected Component}
+\subsection{Avoid Dead-Ends using Largest Biconnected Component}
 
@@
 \begin{itemize}
-\item Avoid nothing running times, i.e., what is if $w_q = q_\ell$?
-\item Multiply all weights by $\alpha$ for $\alpha$ in 1.1, 1.5, 2, 3
+\item Avoid nothing running times, i.e., what is if $w_q = w_\ell$?
+\item Multiply all weights by $\alpha$, i.e., $w_q = \alpha w_\ell$ for $\alpha$ in 1.1, 1.5, 2, 3
 \item Avoid highway
@@
 \item Ptv-TD traffic? Achtung: Das Fass würde ich nicht aufmachen. Am Ende will ein Reviewer, dass wir avoid highway mit TD kombinieren... das ist NP-schwer
-\item Running times if OSM degree 2 nodes are not extracted!
+\item Running times if OSM degree 2 nodes are not extracted! Das einzige Argument pro Erhalten von Deg 2 das ich bisher kenne ist MapBox Live Traffic matchen.
 \end{itemize}

### 2020-04-15 paper/ch_potentials.tex
Rewrote Lazy Topocore

@@
 
-\section{Avoiding Potential Function Evaluations}
+\section{Avoiding Potential Function Evaluations at Degree Two Nodes}
 
 We observe, that a potential evaluation is always paired with a queue push operation.
-By avoiding push operations we can therefore also avoid potential evaluations.
+By avoiding push operations, we can therefore also avoid potential evaluations.
+In this section, we slightly modify how A* works internally.
+Our objective is process chains of low degree nodes sequentially without putting them into the queue.
 
-In this section, we slightly modify how A* works internally.
-Our modified version maintains the following entities:
-(1) a tentative distance $D$ for every node,
-(2) a queue of nodes $Q$ weighted by $D[v]+\pi(v)$.
+We denote the tentative distance array used in the A* algorithm by $D$.
+When removing a node $v_1$ from the A* queue, we iterate over all outgoing edges $(v_1,v_2)$ of $v_1$ and process the chain started by $(v_1, v_2)$.
+If $D[v_2] < D[v_1]+w_q(v_1,v_2)$, our algorithm stops processing the chain and continues with the next outgoing edge of $v_1$.
+Otherwise, $D[v_2]$ is set to $D[v_1]+w_q(v_1,v_2)$ and processing the chain continues.
+If $v_2$ has only one outgoing edge that leads to a node $v_3$ different from $v_1$, we process $v_3$ analogously.
+This is repeated until a $v_k$ is reached such that either $D[v_{k-1}]+w_q(v_{k-1},v_k)\ge D[v_k]$ or $v_k$ has more than one successor different from $v_{k-1}$.
+Our algorithm continues removing nodes from the queue until a node $x$ is removed for which $D[x]\ge D[t]$.
 
-Initially, $Q$ only contains $s$.
-$D$ is $+\infty$ everywhere except for $D[s]=0$.
+Our optimization can be described as node contraction performed lazily at query time.
+If $G$ is undirected, all nodes with a degree two in $G$ except $s$ are lazily contracted.
+As we only employ contraction during the query, we do not loose flexibility.
+This contrasts with most hierarchical approaches, that use contraction in the preprocessing step.
 
-Similarily to the basic A* algorithm, our algorithm continously removes nodes from the queue $Q$.
-Denote by $x$ a node removed with the queue.
-Contrary to A*, our algorithm does not only consider the immediate outgoing edges of $x$.
-Instead, a limited, queue-less backtracking search at $x$ is started.
-The idea is to process local tree-like areas of the graph without queue interaction.
-We exploit that backtracking is fast on trees.
+Not pushing degree two nodes into the queue gives a moderate to large speedup depending on how many degree two nodes are part of the graph.
+We can achieve a slightly larger speedup by also not pushing every degree three node into the queue.
 
-The backtracking search continues until either the tentative distance at a node is lower than the length of the path found by the search or a stop node is encountered.
-Every node that was already pushed is a stop node.
-Nodes with degree 4 or more is a stop node (TODO: @Tim out-degree ?).
-Every degree 3 node that is reached via another degree 3 node from the backtracking search start is a stop node.
-If the search was stopped because of the tentative distance, then it is aborted.
-If the search was stopped because a stop node was encountered, the stop node is added to the queue.
-This implies a potential evaluation.
-If the node is already in the queue, a decrease key operation is executed.
-One can view this modifed A* algorithm as a lazy version of \cite{topocore} where the set of virtually contracted nodes is computed at query time.
+\section{Avoiding Potential Function Evaluations at Degree Three Nodes}
 
-\subsection{Avoid Dead-Ends using Largest Biconnected Component}
+In the previous section, we described how a lazy degree two node contraction can be used to accelerate an A* search.
+Denote by $I$ the set of nodes that can be pushed into the A* queue.
+Using the $I$ from the previous section, $G\setminus I$ is a disjoint set of paths, if $G$ is undirected.
+For performance reasons, it is important that $G\setminus I$ results in a disjoint collection of simple graphs.
+Fortunately, we can allow slightly more complex graphs than just paths.
+In this section, we use stars where the center node has degree three.
+This allows us to further increase the size of $I$, which in turn reduces the number of A* queue operations and potential evaluations.
 
-A lot of nodes in road networks lead to dead-ends.
-Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
-We employ an idea from \cite{TOPOCORE} to avoid these nodes entirely.
-This reduces the number of explored nodes and thus fewer potentials are evaluated.
-
-In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
-We do this using Tarjan's algorithm \cite{Tarjan's algorithm}.
-For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
-For nodes in the core, $a_v=v$.
-We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
-The basic CH-potential preprocessing step is only executed for $G_C$.
-
-The query phase is divided into four steps.
-In the first step, we use Dijkstra's algorithm to explore the component that contains $s$ until $a_s$ is reached.
-If this search finds $t$, then $s$ and $t$ are part of the same component.
-An optimal path was found and therefore the algorithm stops early.
-Otherwise, in the second step starts.
-It consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
-Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
-The final path is the concatenation of all three path parts.
+Let $v_k$ be the last node of a chain as in the previous section.
+If $v_k$ has two successor nodes $u$ and $w$ that differ from $v_{k-1}$ and $v_k$ is not part of the queue, then we do not push $v_k$.
+Instead, we follow the chains started by $(v_k, u)$ and $(v_k, w)$ as described in the previous section.
 

### 2020-04-17 code/compute_ch/src/bin/compute_contraction_hierarchy_and_order.cpp
add routingkit ch computation

@@ -1,0 +1,83 @@
+#include <routingkit/vector_io.h>
+#include <routingkit/timer.h>
+#include <routingkit/contraction_hierarchy.h>
+#include <routingkit/min_max.h>
+#include <routingkit/inverse_vector.h>
+
+#include <iostream>
+#include <stdexcept>
+#include <vector>
+
+using namespace RoutingKit;
+using namespace std;
+
+int main(int argc, char*argv[]){
+
+  try{
+    string graph_first_out;
+    string graph_head;
+    string graph_weight;
+
+    string ch_order;
+    string ch_forward_first_out;
+    string ch_forward_head;
+    string ch_forward_weight;
+    string ch_backward_first_out;
+    string ch_backward_head;
+    string ch_backward_weight;
+
+    if(argc != 11){
+      cerr << argv[0] << " graph_first_out graph_head graph_weight ch_order ch_forward_first_out ch_forward_head ch_forward_weight ch_backward_first_out ch_backward_head ch_backward_weight" << endl;
+      return 1;
+    }else{
+      graph_first_out = argv[1];
+      graph_head = argv[2];
+      graph_weight = argv[3];
+      ch_order = argv[4];
+      ch_forward_first_out = argv[5];
+      ch_forward_head = argv[6];
+      ch_forward_weight = argv[7];
+      ch_backward_first_out = argv[5];
+      ch_backward_head = argv[6];
+      ch_backward_weight = argv[7];
+    }
+
+
+    cout << "Loading graph ... " << flush;
+
+    vector<unsigned>first_out = load_vector<unsigned>(graph_first_out);
+    vector<unsigned>head = load_vector<unsigned>(graph_head);
+    vector<unsigned>weight = load_vector<unsigned>(graph_weight);
+
+    cout << "done" << endl;
+
+    const unsigned node_count = first_out.size()-1;
+    const unsigned arc_count = head.size();
+
+    if(first_out.front() != 0)
+      throw runtime_error("The first element of first out must be 0.");
+    if(first_out.back() != arc_count)
+      throw runtime_error("The last element of first out must be the arc count.");
+    if(head.empty())
+      throw runtime_error("The head vector must not be empty.");
+    if(max_element_of(head) >= node_count)
+      throw runtime_error("The head vector contains an out-of-bounds node id.");
+    if(weight.size() != arc_count)
+      throw runtime_error("The weight vector must be as long as the number of arcs");
+
+
[…]

### 2020-04-17 paper/ch_potentials.tex
add sketch of experiments section

@@
 
+Tims plan:
+
+\begin{itemize}
+\item Effectiveness of A*
+	\begin{itemize}
+  \item Multiplied weights by 1 to 2 in steps of 0.05
+  \item On DIMACs Eur
+  \item Plot x factor, y running time
+  \end{itemize}
+
+\item Algo Building Blocks
+	\begin{itemize}
+  \item DIMACs Eur x1.1
+  \item Table
+  	\begin{itemize}
+    \item columns: running time, potential evals, relaxed arcs
+    \item rows: (Dijkstra, SCC, Deg2, Deg3) x (CH Pot or not)
+    \item or transposed?
+		\end{itemize}
+	\end{itemize}
+
+\item Applications
+	\begin{itemize}
+  \item Table
+  	\begin{itemize}
+    \item columns: running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
+    \item rows: OSM Ger Mapbox Live, forbidden highways, forbidden tunnels, TD Ger06, TD Eur17, DIMACs Eur 1000 {Random|Fastest} Arcs {x2, x10}
+		\end{itemize}
+	\end{itemize}
+\end{itemize}
+
 \section{Conclusion}

### 2020-04-20 code/routingkit_ch_pot/build.sh
add routingkit ch pot code

@@ -1,0 +1,2 @@
+#!/bin/sh
+g++ ch_pot.cpp -O3 -DNDEBUG -o ch_pot -lroutingkit

### 2020-04-20 code/routingkit_ch_pot/ch_pot.cpp

@@ -1,0 +1,484 @@
+#include <routingkit/timestamp_flag.h>
+#include <routingkit/contraction_hierarchy.h>
+#include <routingkit/id_queue.h>
+#include <routingkit/timer.h>
+#include <routingkit/vector_io.h>
+#include <routingkit/permutation.h>
+#include <routingkit/inverse_vector.h>
+#include <routingkit/graph_util.h>
+#include <routingkit/dijkstra.h>
+
+#include <iostream>
+#include <string>
+#include <vector>
+
+using namespace RoutingKit;
+using namespace std;
+
+std::vector<unsigned>compute_pseudo_dfs_node_order(const std::vector<unsigned>&first_out, const std::vector<unsigned>&head){
+        unsigned node_count = first_out.size()-1;
+        std::vector<bool>was_pushed(node_count, false);
+        std::vector<unsigned>order(node_count);
+        unsigned order_end = 0;
+        std::vector<unsigned>stack(node_count);
+        unsigned stack_end = 0;
+        for(unsigned s=0; s<node_count; ++s){
+                if(!was_pushed[s]){
+                        stack[stack_end++] = s;
+                        was_pushed[s] = true;
+
+                        while(stack_end != 0){
+                                unsigned x = stack[--stack_end];
+                                order[order_end++] = x;
+                                for(unsigned xy=first_out[x]; xy!=first_out[x+1]; ++xy){
+                                        unsigned y = head[x];
+                                        if(!was_pushed[y]){
+                                                stack[stack_end++] = y;
+                                                was_pushed[y] = true;                                        
+                                        }
+                                }
+                        }
+                }
+        }
+        return order;
+}
+
+struct ZeroPot{
+        unsigned eval(unsigned source_node){
+                return 0;
+        }
+
+        void set_target(unsigned target_node){
+
+        }
+
+        void preprocess(unsigned node_count, const std::vector<unsigned>&tail, const std::vector<unsigned>&head, const std::vector<unsigned>&lower_bound_weight, const ContractionHierarchy&ch){
+                
+        }
+};
+
+struct PotUsingCHQuery{
+        void preprocess(unsigned node_count, const std::vector<unsigned>&tail, const std::vector<unsigned>&head, const std::vector<unsigned>&lower_bound_weight, const ContractionHierarchy&ch){
+                ch_query.reset(ch);
+        }
+
+        void set_target(unsigned target_node){
+                this->target_node = target_node;
+        }
+
[…]

### 2020-04-20 paper/ch_potentials.tex
forgot reporting preprocessing
modified deg 2 and deg 3 sections

@@
 
+\usepackage{wrapfig}
 \usepackage[algo2e,vlined]{algorithm2e}
@@
 
-We observe, that a potential evaluation is always paired with a queue push operation.
-By avoiding push operations, we can therefore also avoid potential evaluations.
-In this section, we slightly modify how A* works internally.
-Our objective is process chains of low degree nodes sequentially without putting them into the queue.
+Potential evaluation is an expensive operation that we want to avoid.
+A potential evaluation is always paired with a queue push operation.
+By avoiding push operations, we can therefore avoid potential evaluations.
+This is the objective of this and the next section.
 
-We denote the tentative distance array used in the A* algorithm by $D$.
-When removing a node $v_1$ from the A* queue, we iterate over all outgoing edges $(v_1,v_2)$ of $v_1$ and process the chain started by $(v_1, v_2)$.
-If $D[v_2] < D[v_1]+w_q(v_1,v_2)$, our algorithm stops processing the chain and continues with the next outgoing edge of $v_1$.
-Otherwise, $D[v_2]$ is set to $D[v_1]+w_q(v_1,v_2)$ and processing the chain continues.
-If $v_2$ has only one outgoing edge that leads to a node $v_3$ different from $v_1$, we process $v_3$ analogously.
-This is repeated until a $v_k$ is reached such that either $D[v_{k-1}]+w_q(v_{k-1},v_k)\ge D[v_k]$ or $v_k$ has more than one successor different from $v_{k-1}$.
-Our algorithm continues removing nodes from the queue until a node $x$ is removed for which $D[x]\ge D[t]$.
+We modify how A* by processing low degree nodes consecutively without pushing them into the queue.
+Our algorithm makes use of the undirected degree $d(x)$ of a node $x$. 
+Formally, $d(x)$ is the number of nodes $y$ such that $(x,y)\in E$ or $(y,x)\in E$.
 
-Our optimization can be described as node contraction performed lazily at query time.
-If $G$ is undirected, all nodes with a degree two in $G$ except $s$ are lazily contracted.
-As we only employ contraction during the query, we do not loose flexibility.
-This contrasts with most hierarchical approaches, that use contraction in the preprocessing step.
+Analogous to A*, our algorithm stores for every node $x$ a tentative distance $D[x]$.
+Additionally, it maintains a minimum priority queue. 
+Diverging from A*, not all nodes can be pushed but every node has a tentative distance.
 
-Not pushing degree two nodes into the queue gives a moderate to large speedup depending on how many degree two nodes are part of the graph.
-We can achieve a slightly larger speedup by also not pushing every degree three node into the queue.
+Our algorithm differs from A* when removing a node $x$ from the queue.
+A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
+If A* succeeds, $y$'s weight in the queue is set to $D[y]+\pi(y)$.
+Our algorithm, however, behaves differently, if $d(y)\le 2$.
+Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z)\neq z$.
+If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
+Instead, it iteratively tries to reduce all $D[y_i]$.
+If it does not reach $z$, then only $D$ is modified but no queue action is performed.
+If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[y]+\pi(y)$.
+
+As the target node $t$ might have degree two, our algorithm cannot rely on stopping, when $t$ is removed from the queue.
+Instead, our algorithm stops as soon as $D[t]$ is less than the minimum weight in the queue.
 
@@
 
-In the previous section, we described how a lazy degree two node contraction can be used to accelerate an A* search.
-Denote by $I$ the set of nodes that can be pushed into the A* queue.
-Using the $I$ from the previous section, $G\setminus I$ is a disjoint set of paths, if $G$ is undirected.
-For performance reasons, it is important that $G\setminus I$ results in a disjoint collection of simple graphs.
-Fortunately, we can allow slightly more complex graphs than just paths.
-In this section, we use stars where the center node has degree three.
-This allows us to further increase the size of $I$, which in turn reduces the number of A* queue operations and potential evaluations.
+\begin{wrapfigure}{r}{3.5cm}
 
-Let $v_k$ be the last node of a chain as in the previous section.
-If $v_k$ has two successor nodes $u$ and $w$ that differ from $v_{k-1}$ and $v_k$ is not part of the queue, then we do not push $v_k$.
-Instead, we follow the chains started by $(v_k, u)$ and $(v_k, w)$ as described in the previous section.
+\begin{center}
+\includegraphics{push-deg3}
+\end{center}
+
+\caption{$x$ is removed from the queue. $y_i$, $z$, $a_i$, and $\alpha_i$ are never part of the queue. $\beta$ and $b$ are pushed into the queue.}
+\label{fig:push-deg3}
+
+\end{wrapfigure}
+
+In the previous section, we described an optimized A* variant that does not push degree two nodes.
+In this section, we also avoid degree three nodes.
+
+Denote by $x,y_1,\ldots, y_k, z$ a degree chain as described in the previous section.
+If $z$ is already in the queue, then $z$'s weight in the queue is adjusted as in the previous section.
+
+If $d(z)\neq 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
+Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
+Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
+If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+\pi(\beta)$.
+Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+\pi(b)$.
+If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
+The situation is depicted in Figure~\ref{fig:push-deg3}.
 
@@
   	\begin{itemize}
-    \item columns: running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
+    \item columns: preprocessing time, running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
     \item rows: OSM Ger Mapbox Live, forbidden highways, forbidden tunnels, TD Ger06, TD Eur17, DIMACs Eur 1000 {Random|Fastest} Arcs {x2, x10}

### 2020-04-20 paper/ch_potentials.tex
A* consistent text improved
restored section about deg 1

@@
 The potential is often called heuristic.
-A* is optimal, if the potential $\pi(v)$ is consistent \cite{Pearl, Judea (1984).
-Heuristics: Intelligent Search Strategies for Computer Problem Solving.
-Addison-Wesley.
-ISBN 0-201-05594-5.}, which essentially means that $\pi(v)$ is a lower bound of the distance from $v$ to $t$.
+A* is optimal, if the potential $\pi(v)$ is consistent \cite{Pearl, Judea (1984). Heuristics: Intelligent Search Strategies for Computer Problem Solving. Addison-Wesley. ISBN 0-201-05594-5.}.
+In this paper, all considered potentials are derived from exact shortest path distances and are thus consistent.
 
@@
 
+\subsection{Avoid Dead-Ends using Largest Biconnected Component}
+
+A lot of nodes in road networks lead to dead-ends.
+Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
+This reduces the number of explored nodes and thus fewer potentials are evaluated.
+
+In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
+We do this using Tarjan's algorithm \cite{Tarjan's algorithm}.
+For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
+For nodes in the core, $a_v=v$.
+We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
+The basic CH-potential preprocessing step is only executed for $G_C$.
+
+The query phase is divided into four steps.
+In the first step, we use Dijkstra's algorithm to explore the component that contains $s$ until $a_s$ is reached.
+If this search finds $t$, then $s$ and $t$ are part of the same component.
+An optimal path was found and therefore the algorithm stops early.
+Otherwise, in the second step starts.
+It consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
+Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
+The final path is the concatenation of all three path parts.
+
 \section{Experiments}

### 2020-05-05 paper/ch_potentials.tex
subsection -> section
start with experimental eval section

@@
 \usepackage{wrapfig}
+\usepackage{booktabs}
 \usepackage[algo2e,vlined]{algorithm2e}
@@
 
-\subsection{Avoid Dead-Ends using Largest Biconnected Component}
+\section{Avoid Dead-Ends using Largest Biconnected Component}
 
@@
 
-\begin{itemize}
-\item Avoid nothing running times, i.e., what is if $w_q = w_\ell$?
-\item Multiply all weights by $\alpha$, i.e., $w_q = \alpha w_\ell$ for $\alpha$ in 1.1, 1.5, 2, 3
-\item Avoid highway
-\item Avoid tunnel
-\item Avoid highway + tunnel
-\item MapBox Live Traffic
-\item Ptv-TD traffic? Achtung: Das Fass würde ich nicht aufmachen. Am Ende will ein Reviewer, dass wir avoid highway mit TD kombinieren... das ist NP-schwer
-\item Running times if OSM degree 2 nodes are not extracted! Das einzige Argument pro Erhalten von Deg 2 das ich bisher kenne ist MapBox Live Traffic matchen.
-\end{itemize}
+\label{sec:experiments}
+
+In this section, we present our experimental evaluation.
+Our benchmark machine runs openSUSE Leap 15.1 (kernel 4.12.14), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
+All experiments were performed sequentially.
+Our code is written in Rust and compiled with rustc 1.44.0-nightly in the release profile with the target-cpu=native option.
+
+\subparagraph{Inputs and Methodology}
+We perform experiments on several graphs with travel time weights which we modify in different ways.
+Preprocessing running times are averages over 10 runs.
+For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
+Our main benchmark instance is a graph of the road network of Western Europe made publicly available for the Ninth DIMACS implementation Challenge~\cite{DemetrescuGJ09} with 18M nodes and 42M edges.
+Additionally, we perform experiments on a recent OSM instance of Germany\footnote{\url{https://download.geofabrik.de/europe/germany.html}, obtained on August 2nd, 2019}.
+To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
+The graph has 11M nodes and 26M edges.
+For this instance, we have proprietary live traffic snapshot from Friday 2019/08/02 afternoon provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
+The live data comes in the form of 320K OSM node pairs and live speeds for the link between the nodes.
+We also have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
+One is an old instance of Germany from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges).
+
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
+\caption{Query running times for queries of DIMACs Europe with scaled edge weights.}\label{fig:scaled_weights}
+\end{figure}
+
+\begin{table}
+\centering
+\caption{Query performance with the different algorithmic features.}\label{tab:building_blocks}
+\begin{tabular}{lrrr}
+\toprule
+{} & Running Time &     Queue Pops &   Relaxed Arcs \\
+{} &         [ms] & [$\cdot 10^3$] & [$\cdot 10^6$] \\
+\midrule
+No Deg2, No Deg3, No SCC, No Pot. & 1765 & 7087.0 & 19.6 \\
+No Deg2, No Deg3, No Pot.         & 1308 & 6249.8 & 17.3 \\
+No Deg3, No Pot.                  & 1180 & 5233.7 & 16.5 \\
+No Pot.                           &  901 & 3035.5 & 15.1 \\
+No Deg2, No Deg3, No SCC          &  547 & 1006.4 &  2.7 \\
+No Deg2, No Deg3                  &  469 &  905.2 &  2.5 \\
+No Deg3                           &  410 &  767.5 &  2.4 \\
+                                  &  315 &  445.0 &  2.2 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 \begin{itemize}
-\item Effectiveness of A*
-	\begin{itemize}
-  \item Multiplied weights by 1 to 2 in steps of 0.05
-  \item On DIMACs Eur
-  \item Plot x factor, y running time
-  \end{itemize}
-
-\item Algo Building Blocks
-	\begin{itemize}
-  \item DIMACs Eur x1.1
-  \item Table
-  	\begin{itemize}
-    \item columns: running time, potential evals, relaxed arcs
-    \item rows: (Dijkstra, SCC, Deg2, Deg3) x (CH Pot or not)
-    \item or transposed?
-		\end{itemize}
-	\end{itemize}
-
 \item Applications

### 2020-05-06 paper/ch_potentials.tex
add preliminary applications table, fix a lot of cites

@@
 
-Dijkstra's algorithm \cite{?} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
+Dijkstra's algorithm \cite{d-ntpcg-59} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
 A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
@@
 Finding such a function that can be computed efficiently is difficult.
-A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{CH,CCH,CRP,HIGHWAYHIERACHIES,MLD} which do not require such a function.
-A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{geisberger}.
+A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{gssv-erlrn-12,dsw-cch-15,dgpw-crprn-13,ss-ehh-12} which do not require such a function.
+A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{gssv-erlrn-12}.
 There is a lot of work that extends CH to handle more complex settings.
-For example, in \cite{sabine, kobitzsch} multi-criteria optimization is studied.
-In \cite{CCH}, CH is modified to support Live-Traffic.
-Historic traffic patterns are combined with CH in \cite{TD-CCH-preprint, TDCH, TDCRP}.
-Specialized variants for electric vehicle routing are developped in \cite{moritz baum, charge}.
+For example, in \cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
+In \cite{dsw-cch-15}, CH is modified to support Live-Traffic.
+Historic traffic patterns are combined with CH in \cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
+Specialized variants for electric vehicle routing are developped in \cite{bdgwz-sfpcs-19}.
 While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
@@
 In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
-We are not the first to combine hierarchical approaches and A* \cite{core-alt, real, charge, bdsssw-chgds-08}.
+We are not the first to combine hierarchical approaches and A* \cite{bdsssw-chgds-10, gkw-blwr-07, bdgwz-sfpcs-19}.
 However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
@@
 
-Our algorithm is built upon ideas from Phast \cite{?}, a CH-extension.
+Our algorithm is built upon ideas from Phast \cite{dgnw-phast-13}, a CH-extension.
 The novel idea is to compute optimal A* potentials lazily.
@@
 However, a formal problem specification that encompasses all these options is complex and well beyond the scope of this paper.
-For an example of such a formulation, we refer to \cite{sea2020paper} which covers a truck-specific application.
-The algorithm presented in \cite{sea2020paper} uses a preliminary version of CH-Potentials.
+For an example of such a formulation, we refer to \cite{kswz-erptd-20} which covers a truck-specific application.
+The algorithm presented in \cite{kswz-erptd-20} uses a preliminary version of CH-Potentials.
 For simplicity, we limit our exposition to the classical shortest path formulation.
@@
 
-TODO: Dijkstra's algorithm \cite{Dijkstra's} can be used to solve the stated problem formalization \cite{Veit?}.
+TODO: Dijkstra's algorithm \cite{d-ntpcg-59} can be used to solve the stated problem formalization \cite{Veit?}.
 
@@
 This setup is therefore very unflexible.
-Many hierarchical speedup techniques such as the original CH \cite{CH} use this setup.
-Notable exceptions are \cite{MLD,CRP} and \cite{CCH} which use a third phase to change the query edge weights.
+Many hierarchical speedup techniques such as the original CH \cite{gssv-erlrn-12} use this setup.
+Notable exceptions are \cite{dgpw-crprn-13} and \cite{dsw-cch-15} which use a third phase to change the query edge weights.
 
@@
 The potential is often called heuristic.
-A* is optimal, if the potential $\pi(v)$ is consistent \cite{Pearl, Judea (1984). Heuristics: Intelligent Search Strategies for Computer Problem Solving. Addison-Wesley. ISBN 0-201-05594-5.}.
+A* is optimal, if the potential $\pi(v)$ is consistent \cite{p-hissc-84}.
 In this paper, all considered potentials are derived from exact shortest path distances and are thus consistent.
@@
 A CH is a two phase technique to efficiently compute optimal, shortest paths.
-For a detailed exposition, we refer the interested reader to \cite{CCH, CH}.
+For a detailed exposition, we refer the interested reader to \cite{dsw-cch-15,gssv-erlrn-12}.
 In this section, we give a high-level overview of the aspects relevant to this paper.
@@
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
-On road networks, this is the case as demonstrated in \cite{ch,punch,crp,cch,flowcutter}.
+On road networks, this is the case as demonstrated in \cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.
 
@@
 Fortunatly, we can do better.
-However, for this we need another component called Phast \cite{Phast}.
+However, for this we need another component called Phast \cite{dgnw-phast-13}.
 
@@
 
-Phast \cite{Phast} is a CH extension that computes distances from all nodes to one target node.
+Phast \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
 First the preprocessing phase is executed analogously to the original CH.
@@
 However, we will not consider parallelization in this paper.
-We refer to \cite{Phast} for an in-depth experimental performance analysis.
+We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
@@
 In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
-We do this using Tarjan's algorithm \cite{Tarjan's algorithm}.
+We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
 For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
+\caption{Query running times for queries of DIMACs Europe with scaled edge weights.}\label{fig:scaled_weights}
+\end{figure}
+
 In this section, we present our experimental evaluation.
@@
 One is an old instance of Germany from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges).
-
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
-\caption{Query running times for queries of DIMACs Europe with scaled edge weights.}\label{fig:scaled_weights}
-\end{figure}
 
@@
 
-Tims plan:
-
-\begin{itemize}
-\item Applications
-	\begin{itemize}
-  \item Table
-  	\begin{itemize}
-    \item columns: preprocessing time, running time, potential evals, relaxed arcs, shortest path distance / lower bound, dijkstra running time?
-    \item rows: OSM Ger Mapbox Live, forbidden highways, forbidden tunnels, TD Ger06, TD Eur17, DIMACs Eur 1000 {Random|Fastest} Arcs {x2, x10}
-		\end{itemize}
-	\end{itemize}
-\end{itemize}
+\begin{table}
+\centering
+% TODO dijkstra baseline, preprocessing
+\caption{Performance for different applications.}\label{tab:applications}
+\begin{tabular}{llrrrrr}
+\toprule
+        &                     &   Running &     Pot. Eval. &             Relaxed & Avg. length & Affected \\
+        &                     & time [ms] & $[\cdot 10^3]$ & arcs $[\cdot 10^3]$ &    increase &     [\%] \\
+\midrule
+OSM Ger & No Tunnels          &      39.4 &         3874.1 &               290.6 &        1.74 &     83.5 \\
+        & No Highways         &     338.5 &        30383.1 &              2555.9 &        6.24 &     95.0 \\
+        & Mapbox Live         &      96.5 &         9310.9 &               733.7 &        1.12 &     95.0 \\
+\addlinespace
+DIMACs  & Fastest $\times 10$ &       1.0 &           69.2 &                 2.7 &        1.00 &     16.5 \\
+Europe  & Fastest $\times 2$  &       1.0 &           68.9 &                 2.7 &        1.00 &     16.5 \\
+        & Random $\times 10$  &       1.1 &           80.9 &                 3.0 &        1.00 &      1.0 \\
+        & Random $\times 2$   &       1.0 &           76.7 &                 2.9 &        1.00 &      1.0 \\
+\addlinespace
+Ger06   & TD                  &       4.8 &          417.2 &                33.1 &        1.03 &     99.5 \\
+Eur17   & TD                  &      98.5 &         5143.8 &               553.3 &        1.04 &     98.5 \\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2020-05-06 Rakefile.rb
fix glob path of live data

@@ -43,7 +43,7 @@ td_graphs = [
 ]
 graphs = static_graphs + td_graphs
 
-live_dir = '/algoDaten/mapbox/live-speeds/2019-08-02-15:41/*'
+live_dir = '/algoDaten/mapbox/live-speeds/2019-08-02-15:41/'
 
 namespace "prep" do
   directory osm_graph
@@ -131,7 +131,7 @@ namespace "exp" do
       static_graphs.each do |graph|
         sh "cargo run --release --bin dijkstra -- #{graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       end
-      sh "cargo run --release --bin chpot_live -- #{osm_graph} '#{live_dir}' > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
+      sh "cargo run --release --bin chpot_live -- #{osm_graph} #{live_dir} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_blocked -- #{osm_graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_simulated_live -- #{dimacs_graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
     end

### 2020-05-06 paper/ch_potentials.tex
minor updates to exp intro

@@
 \subparagraph{Inputs and Methodology}
-We perform experiments on several graphs with travel time weights which we modify in different ways.
-Preprocessing running times are averages over 10 runs.
-For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
 Our main benchmark instance is a graph of the road network of Western Europe made publicly available for the Ninth DIMACS implementation Challenge~\cite{DemetrescuGJ09} with 18M nodes and 42M edges.
+We use this graph with the travel time as the lower bound and perform experiments with modified edge weights (scaling all travel times or reproducing an experiment from~\cite{dw-lbrdg-07} where 1\,000 random edges are perturbed).
 Additionally, we perform experiments on a recent OSM instance of Germany\footnote{\url{https://download.geofabrik.de/europe/germany.html}, obtained on August 2nd, 2019}.
@@
 We also have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
-One is an old instance of Germany from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges).
+One is an old instance of Germany with traffic predictions from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges, from 2017).
+Preprocessing running times are averages over 10 runs.
+For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
 

### 2020-05-15 paper/ch_potentials.tex
switch to acm class

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref]{lipics-v2019}
+\documentclass[sigconf]{acmart}
 %This is a template for producing LIPIcs articles.
@@
 
-%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
-
-\title{A* with Perfect Potentials} %TODO Please add
-
-% \titlerunning{Dummy short title}%optional, please use if title is longer than one line
-
-\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{TODO}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional
-
-\author{Tim Zeitz}{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology, Germany}{tim.zeitz@kit.edu}{}{}
-
-\authorrunning{B. Strasser and T. Zeitz}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Ben Strasser and Tim Zeitz}%TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\ccsdesc[500]{Theory of computation~Shortest paths}
-% \ccsdesc[500]{General and reference}%TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm
-
-\keywords{shortest path, road graphs, goal-directed search, contraction hierarchy}%TODO mandatory; please add comma-separated list of keywords
-
-% \category{}%optional, e.g. invited paper
-
-% \relatedversion{}%optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversion{A full version of the paper is available at \url{...}.}
-
-% \supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-% \acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \EventEditors{John Q. Open and Joan R. Access}
-% \EventNoEds{2}
-% \EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
-% \EventShortTitle{CVIT 2016}
-% \EventAcronym{CVIT}
-% \EventYear{2016}
-% \EventDate{December 24--27, 2016}
-% \EventLocation{Little Whinging, United Kingdom}
-% \EventLogo{}
-% \SeriesVolume{42}
-% \ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\hideLIPIcs
-\nolinenumbers
-
 \begin{document}
 
-\maketitle
+%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
+\title{A* with Perfect Potentials}
 
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research.
+\author{Ben Strasser}
+\email{academia@ben-strasser.net}
+\orcid{1234-5678-9012}
+\affiliation{}
+
+\author{Tim Zeitz}
+\email{tim.zeitz@kit.edu}
+\affiliation{%
+  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
+  \city{Karlsruhe}
+  \country{Germany}
+}
+
+%%
+%% By default, the full list of authors will be used in the page
+%% headers. Often, this list is too long, and will overlap
+%% other information printed in the page headers. This command allows
+%% the author to define a more concise list
+%% of authors' names for this purpose.
+% \renewcommand{\shortauthors}{Strasser and Zeitz}
+
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
 \begin{abstract}
@@
 \end{abstract}
+
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+\begin{CCSXML}
+<ccs2012>
+ <concept>
+  <concept_id>10010520.10010553.10010562</concept_id>
+  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+  <concept_significance>500</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010575.10010755</concept_id>
+  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+  <concept_significance>300</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010553.10010554</concept_id>
+  <concept_desc>Computer systems organization~Robotics</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10003033.10003083.10003095</concept_id>
+  <concept_desc>Networks~Network reliability</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+</ccs2012>
+\end{CCSXML}
+
+% \ccsdesc[500]{Computer systems organization~Embedded systems}
+% \ccsdesc[300]{Computer systems organization~Redundancy}
+% \ccsdesc{Computer systems organization~Robotics}
+% \ccsdesc[100]{Networks~Network reliability}
+\ccsdesc[500]{Theory of computation~Shortest paths}
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{shortest path, road graphs, goal-directed search, contraction hierarchy}
+
+%% A "teaser" image appears between the author and affiliation
+%% information and the body of the document, and typically spans the
+%% page.
+% \begin{teaserfigure}
+%   \includegraphics[width=\textwidth]{sampleteaser}
+%   \caption{Seattle Mariners at Spring Training, 2010.}
+%   \Description{Enjoying the baseball game from the third-base
+%   seats. Ichiro Suzuki preparing to bat.}
+%   \label{fig:teaser}
+% \end{teaserfigure}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
 
@@
 
-\begin{wrapfigure}{r}{3.5cm}
+\begin{figure}
 
@@
 
-\end{wrapfigure}
+\end{figure}
 
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
@@
 \caption{Query running times for queries of DIMACs Europe with scaled edge weights.}\label{fig:scaled_weights}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{table}
+\begin{table*}
 \centering
@@
 \end{tabular}
-\end{table}
+\end{table*}
 
-\begin{table}
+\begin{table*}
 \centering
@@
 \end{tabular}
-\end{table}
+\end{table*}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{references}

### 2020-05-15 paper/ch_potentials.tex
add pipeline for tables and figures

@@
 \caption{Query performance with the different algorithmic features.}\label{tab:building_blocks}
-\begin{tabular}{lrrr}
-\toprule
-{} & Running Time &     Queue Pops &   Relaxed Arcs \\
-{} &         [ms] & [$\cdot 10^3$] & [$\cdot 10^6$] \\
-\midrule
-No Deg2, No Deg3, No SCC, No Pot. & 1765 & 7087.0 & 19.6 \\
-No Deg2, No Deg3, No Pot.         & 1308 & 6249.8 & 17.3 \\
-No Deg3, No Pot.                  & 1180 & 5233.7 & 16.5 \\
-No Pot.                           &  901 & 3035.5 & 15.1 \\
-No Deg2, No Deg3, No SCC          &  547 & 1006.4 &  2.7 \\
-No Deg2, No Deg3                  &  469 &  905.2 &  2.5 \\
-No Deg3                           &  410 &  767.5 &  2.4 \\
-                                  &  315 &  445.0 &  2.2 \\
-\bottomrule
-\end{tabular}
+\input{table/building_blocks.tex}
 \end{table*}
@@
 \caption{Performance for different applications.}\label{tab:applications}
-\begin{tabular}{llrrrrr}
-\toprule
-        &                     &   Running &     Pot. Eval. &             Relaxed & Avg. length & Affected \\
-        &                     & time [ms] & $[\cdot 10^3]$ & arcs $[\cdot 10^3]$ &    increase &     [\%] \\
-\midrule
-OSM Ger & No Tunnels          &      39.4 &         3874.1 &               290.6 &        1.74 &     83.5 \\
-        & No Highways         &     338.5 &        30383.1 &              2555.9 &        6.24 &     95.0 \\
-        & Mapbox Live         &      96.5 &         9310.9 &               733.7 &        1.12 &     95.0 \\
-\addlinespace
-DIMACs  & Fastest $\times 10$ &       1.0 &           69.2 &                 2.7 &        1.00 &     16.5 \\
-Europe  & Fastest $\times 2$  &       1.0 &           68.9 &                 2.7 &        1.00 &     16.5 \\
-        & Random $\times 10$  &       1.1 &           80.9 &                 3.0 &        1.00 &      1.0 \\
-        & Random $\times 2$   &       1.0 &           76.7 &                 2.9 &        1.00 &      1.0 \\
-\addlinespace
-Ger06   & TD                  &       4.8 &          417.2 &                33.1 &        1.03 &     99.5 \\
-Eur17   & TD                  &      98.5 &         5143.8 &               553.3 &        1.04 &     98.5 \\
-\bottomrule
-\end{tabular}
+\input{table/applications.tex}
 \end{table*}

### 2020-05-15 Rakefile.rb

@@ -12,25 +12,25 @@ end
 task default: "paper/ch_potentials.pdf"
 
 namespace "fig" do
-  # file "paper/fig/ranks.pdf" => FileList["#{exp_dir}/rank_queries_all_graphs/*.json"] + FileList["#{exp_dir}/rank_query_stats_no_astar_all_graphs/*.json"] + ['eval/rank_plot.py'] do
-  #   sh "eval/rank_plot.py"
-  # end
+  file "paper/fig/scaled_weights.pdf" => FileList["#{exp_dir}/scaled_weights/*.json"] + ["eval/scaled_weights.py"] do
+    sh "eval/scaled_weights.py"
+  end
 end
 
 namespace "table" do
-  # file "paper/table/related_work.tex" => FileList[
-  #   "#{exp_dir}/katch/contraction/*.out",
-  #   "#{exp_dir}/katch/uniform_queries/*.out",
-  #   "#{exp_dir}/tds/contraction/*.out",
-  #   "#{exp_dir}/tds/uniform_queries/*.out",
-  #   "#{exp_dir}/queries_all_graphs/*.json",
-  #   "#{exp_dir}/customization_all_graphs/*.json",
-  #   "#{exp_dir}/partition/*.out",
-  #   "#{exp_dir}/contraction/*.json",
-  #   ] + ["eval/related_work_cmp.py"] do
+  directory "paper/table"
 
-  #   sh "eval/related_work_cmp.py"
-  # end
+  file "paper/table/applications.tex" => FileList[
+    "#{exp_dir}/preprocessing/*.out",
+    "#{exp_dir}/applications/*.json",
+    ] + ["eval/applications.py", "paper/table"] do
+
+    sh "eval/applications.py"
+  end
+
+  file "paper/table/building_blocks.tex" => FileList["#{exp_dir}/building_blocks/*.json"] + ["eval/building_blocks.py", "paper/table"] do
+    sh "eval/building_blocks.py"
+  end
 end
 
 osm_graph = "/algoDaten/zeitz/roadgraphs/osm_ger_rel/"

### 2020-05-15 eval/applications.py

@@ -1,0 +1,78 @@
+#!/usr/bin/env python3
+
+import pandas as pd
+
+import json
+import glob
+import os
+import re
+
+base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+paths = glob.glob(base + "applications/*.json")
+data = [json.load(open(path)) for path in paths]
+
+
+queries = pd.DataFrame.from_records([{
+    'exp': exp['experiment'],
+    'graph': run['args'][1],
+    **algo }
+    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
+
+queries = queries.append(pd.DataFrame.from_records([{
+    'exp': run['program'],
+    'graph': run['args'][1],
+    **algo }
+    for run in data for algo in run.get('algo_runs', []) if algo.get('algo') in ['CH Potentials Query', 'CH Potentials TD Query']]))
+
+
+dijkstra_queries = pd.DataFrame.from_records([{
+    'exp': exp['experiment'],
+    'graph': run['args'][1],
+    **algo }
+    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') == 'Dijkstra Query'])
+
+dijkstra_queries = dijkstra_queries.append(pd.DataFrame.from_records([{
+    'exp': run['program'],
+    'graph': run['args'][1],
+    **algo }
+    for run in data for algo in run.get('algo_runs', []) if algo.get('algo') in ['Dijkstra Query', 'TD Dijkstra Query']]))
+
+
+runtime_pattern = re.compile(".*Needed (\\d+)musec\\..*")
+
+def parse_contraction_output(path):
+    stats = { 'preprocessing_running_time_s': 0.0 }
+
+    with open(path, 'r') as f:
+        for line in f:
+            if not 'graph' in stats:
+                stats['graph'] = line.strip()
+            else:
+                match = runtime_pattern.match(line)
+                if match:
+                    stats['preprocessing_running_time_s'] += int(match[1]) / 1000000
+
+    return stats
+
+preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for path in glob.glob(base + "preprocessing/*.out")])
+
+queries['affected'] = queries['lower_bound'] != queries['result']
+queries['increase'] = queries['result'] / queries['lower_bound']
+
[…]

### 2020-05-15 eval/building_blocks.py

@@ -1,0 +1,23 @@
+#!/usr/bin/env python3
+
+import pandas as pd
+
+import json
+import glob
+import os
+
+base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+paths = glob.glob(base + "building_blocks/*.json")
+data = [json.load(open(path)) for path in paths]
+
+queries = pd.DataFrame.from_records([{
+    'features': run['feature_flags'].replace('DEFAULT, TDCCH_APPROX, TDCCH_POSTCUSTOMIZATION, TDCCH_PRECUSTOMIZATION, TDCCH_QUERY_ASTAR, TDCCH_TRIANGLE_SORTING', ''),
+    **algo }
+    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
+
+table = queries.groupby('features')[['running_time_ms', 'num_queue_pops', 'num_relaxed_arcs']].mean()
+table['num_relaxed_arcs'] = table['num_relaxed_arcs'] / 1000000
+table['num_queue_pops'] = table['num_queue_pops'] / 1000
+
+with open("paper/table/building_blocks.tex", 'w') as f:
+  f.write(table.to_latex())

### 2020-05-15 eval/scaled_weights.py

@@ -1,0 +1,30 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import matplotlib as mpl
+from matplotlib import pyplot as plt
+import seaborn as sns
+sns.set()
+
+import pandas as pd
+
+import json
+import glob
+import os
+
+base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+paths = glob.glob(base + "scaled_weights/*.json")
+data = [json.load(open(path)) for path in paths]
+
+queries = pd.DataFrame.from_records([{
+    'weight_factor': exp['factor'],
+    'graph': run['args'][1],
+    **algo }
+    for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
+
+plt.figure(figsize=(11,4))
+g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8)
+g.set_ylabel('Running Time [ms]')
+g.set_xlabel('Weight Factor')
+plt.tight_layout()
+g.get_figure().savefig('paper/fig/scaled_weights.pdf')

### 2020-05-20 Rakefile.rb
table updates

@@ -2,7 +2,10 @@ raise "No or wrong experiments output dir configured - please set CHPOT_EXP_OUTP
 exp_dir = ENV['CHPOT_EXP_OUTPUT_DIR']
 
 file "paper/ch_potentials.pdf" => [
-  "paper/ch_potentials.tex"] do
+  "paper/ch_potentials.tex",
+  "paper/fig/scaled_weights.pdf",
+  "paper/table/applications.tex",
+  "paper/table/building_blocks.tex"] do
 
   Dir.chdir "paper" do
     sh "latexmk -pdf ch_potentials.tex"

### 2020-05-20 eval/applications.py

@@ -57,7 +57,7 @@ def parse_contraction_output(path):
 preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for path in glob.glob(base + "preprocessing/*.out")])
 
 queries['affected'] = queries['lower_bound'] != queries['result']
-queries['increase'] = queries['result'] / queries['lower_bound']
+queries['increase'] = (queries['result'] / queries['lower_bound'] - 1) * 100
 
 table = queries.groupby(['graph', 'exp']).agg(
     running_time_ms=('running_time_ms', 'mean'),
@@ -66,13 +66,54 @@ table = queries.groupby(['graph', 'exp']).agg(
     increase=('increase', 'mean'),
     affected=('affected', 'sum'),
     size=('affected', 'count'))
+
+table = table.reindex(['/algoDaten/zeitz/roadgraphs/europe/',
+    '/algoDaten/zeitz/roadgraphs/osm_ger_rel/',
+    '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/',
+    '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/'], level=0) \
+    .reindex(['chpot_td',
+    'random_times_2',
+    'random_times_10',
+    'fastest_times_2',
+    'fastest_times_10',
+    'chpot_live',
+    'no_tunnels',
+    'no_highways',
+    'chpot_turns'], level=1)
+
 table['affected'] = table['affected'] * 100 / table['size']
 table = table[table.columns[0:-1]]
 table['num_relaxed_arcs'] = table['num_relaxed_arcs'] / 1000
 table['num_pot_evals'] = table['num_pot_evals'] / 1000
 table = table.join(preprocessing.groupby('graph').mean())
 table = table.join(dijkstra_queries.groupby(['graph', 'exp']).agg(dijkstra_running_time_ms=('running_time_ms', 'mean')))
-#table = table.round(1)
+table = table.round(1)
+
+table = table.rename(index={
+    '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/': 'TDGer06',
+    '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/': 'TDEur17',
+    '/algoDaten/zeitz/roadgraphs/osm_ger_rel/': 'OSM Germany',
[…]

### 2020-05-20 eval/building_blocks.py

@@ -5,6 +5,7 @@ import pandas as pd
 import json
 import glob
 import os
+import re
 
 base = os.environ['CHPOT_EXP_OUTPUT_DIR']
 paths = glob.glob(base + "building_blocks/*.json")
@@ -16,8 +17,32 @@ queries = pd.DataFrame.from_records([{
     for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 
 table = queries.groupby('features')[['running_time_ms', 'num_queue_pops', 'num_relaxed_arcs']].mean()
-table['num_relaxed_arcs'] = table['num_relaxed_arcs'] / 1000000
+table['num_relaxed_arcs'] = table['num_relaxed_arcs'] / 1000
 table['num_queue_pops'] = table['num_queue_pops'] / 1000
+table = table.round(1)
+
+table = table.rename(index={
+    'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_NO_SCC, CHPOT_ONLY_TOPO, ': 'No Deg2, No Deg3, No SCC, No Pot.',
+    'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ':               'No Deg2, No Deg3, No Pot.',
+    'CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ':                              'No Deg3, No Pot.',
+    'CHPOT_ONLY_TOPO, ':                                             'No Pot.',
+    'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_NO_SCC, ':                  'No Deg2, No Deg3, No SCC',
+    'CHPOT_NO_DEG2, CHPOT_NO_DEG3, ':                                'No Deg2, No Deg3',
+    'CHPOT_NO_DEG3, ':                                               'No Deg3',
+})
+
+table = table.reindex(['No Deg2, No Deg3, No SCC, No Pot.', 'No Deg2, No Deg3, No Pot.', 'No Deg3, No Pot.', 'No Pot.',
+    'No Deg2, No Deg3, No SCC', 'No Deg2, No Deg3', 'No Deg3', ''])
+
+lines = table.to_latex(escape=False).split("\n")
+
+lines = lines[:2] + [
+    R"{} & Running Time &     Queue Pops &   Relaxed Arcs \\",
[…]

### 2020-05-20 eval/scaled_weights.py

@@ -18,13 +18,14 @@ data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
     'weight_factor': exp['factor'],
-    'graph': run['args'][1],
+    'graph': { '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe', '/algoDaten/zeitz/roadgraphs/osm_ger_rel/': 'OSM Germany' }[run['args'][1]],
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 
 plt.figure(figsize=(11,4))
-g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8)
+g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8, showmeans=True)
 g.set_ylabel('Running Time [ms]')
 g.set_xlabel('Weight Factor')
+g.legend().set_title('')
 plt.tight_layout()
 g.get_figure().savefig('paper/fig/scaled_weights.pdf')

### 2020-05-20 code/osm_import/build.sh
fixed build.sh

@@ -1,2 +1,2 @@
 #!/bin/sh
-g++ import_osm.cpp -O3 -o import_osm -lroutingkit
+g++ src/bin/import_osm.cpp -O3 -o import_osm -lroutingkit

### 2020-06-08 paper/ch_potentials.tex
foo

@@
 
-\section{Applications - Or why do we care about Flexibility?}
+\subsection{Applications need Flexibility}
 
@@
 
-\section{Problem Setting}
+\subsection{Problem Setting}
 
@@
 Throughout this paper, we assume that we are given a directed graph $G$ with nodes $V$ and edges $E$.
-Every edge has a lower bound weight $w_\ell:E\rightarrow \mathbb{R}^+$ and a query weight $w_q:E\rightarrow \mathbb{R}^+$.
+Every edge $(x,y)\in E$ has a \emph{tail} node $x$ and a \emph{head} node $y$.
+Further, every edge has a lower bound weight $w_\ell:E\rightarrow \mathbb{R}^+$ and a query weight $w_q:E\rightarrow \mathbb{R}^+ \cup \{+\infty\}$.
 It must hold that $0\le w_\ell(e)\le w_q(e)$.
@@
 
-\section{Related Work}
+\subsection{Related Work}
 
@@
 
-\section{Goal-Directed Search using A*}
+\section{Algorithm Description}
 
@@
 
-\section{Contraction Hierarchy (CH)}
+\subsection{Contraction Hierarchy (CH)}
 
 \begin{algorithm2e}
-\KwData{$B[x]$ tentative distance from $x$ to $t$, initially $+\infty$}
-\KwData{Mininum priority queue $Q$, also called open list, initially empty}
-Add $t$ to $Q$ with weight 0\;
+\KwData{$B[x]$ tentative distance from $x$ to $t$}
+\KwData{Min. priority queue $Q$, also called open list}
+$B[x] \leftarrow +\infty$ for all $x\neq t$\;
 $B[t] \leftarrow 0$\;
+Make $Q$ only contain $t$ with weight $0$\;
 \While{not $Q$ empty}{
@@
 Levels are ordered by ``importance''.
-The intuition is that dead-ends are unimportant while highway bridges are very important.
-An edge goes up when it goes from a node in a lower level to higher level.
-Down edges are defined analogously.
+The intuition is that dead-ends are unimportant and low while highway bridges are very important and high.
+An edge goes \emph{up} when it goes from a node in a lower level to higher level.
+\emph{Down} edges are defined analogously.
 An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
@@
 We may therefore restrict our search to up-down paths in $G^+$.
-This search is performed using a bidirectional search.
+This search is bidirectional.
 The forward search starts from $s$ and only follows up-edges.
@@
 
-\section{Potentials using PHAST}
+\subsection{Potentials using PHAST}
 
@@
 
-\section{CH-Potentials}
+\subsection{CH-Potentials}
 
@@
 
-\section{Avoiding Potential Function Evaluations at Degree Two Nodes}
+\subsection{Skip Degree Two Nodes}
 
@@
 
-\section{Avoiding Potential Function Evaluations at Degree Three Nodes}
+\subsection{Skip Degree Three Nodes}
 
@@
 
-\section{Avoid Dead-Ends using Largest Biconnected Component}
+\subsection{Skip Nodes Outside of Largest Biconnected Component}
+
+\label{sec:largested-biconnected-component}
 
@@
 The final path is the concatenation of all three path parts.
+
+\section{Extensions}
+
+A* is a very flexible algorithm. 
+As such there exist many extensions of CH-Potentials to solve extended problem settings.
+While extending with CH-Potentials is comparatively easy, this is usually not the case with competitor algorithms.
+
+\subsection{Forbidden Turns}
+
+The classical shortest path problem allows to turn from every edge to every other incident edge at every node.
+However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
+Handling turn restrictions correctly is crucial for many applications.
+Such restrictions are usually modelled using turn weights.
+We extend CH-potentials by using zero as lower bound for every turn weight in the potential.
+
+In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{related-turn-litrature,see-geisberger-turn-table-paper}.
+$w_t$ maps a pair of incident edges onto the time required to perform the turn.
+Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup\{+\infty\}$.
+A forbidden turn is modelled by setting the corresponding $w_t$ to $+\infty$.
+
+A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware path weight}:\[
+w_q(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1})  + w_q(v_i,v_{i+1}) 
+\] The input to the extended problem consists of a source edge and a target edge.
+The objective is to find a path connecting these edges such that the turn-aware path weight is minimized.
+The first term $w_q(v_1, v_2)$ only depends on the source edge.
+As all considered paths share the same source edge, the term is the same for all paths.
+We can therefore ignore it during the optimization.
+
+To account for turns, many algorithms construct a \emph{turn-expanded} graph $G'$.
+Edges in the input graph $G$ correspond to \emph{expanded nodes} in $G'$. 
+For every pair of successive edges $(x,y)$ and $(y,z)$ in $G$, there is an \emph{expanded edge} in $G'$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
+Notice how an expanded edge weight looks similar to a term of the turn-aware path weight.
+
+A sequence of expanded nodes in the expanded graph $G'$ corresponds to a sequence of edges in the input graph $G$.
+The weight of a path in $G'$ is equal to the turn-aware weight of the corresponding path in $G$ minus the irrelevant $w_q(v_1,v_2)$ term.
+Turn-aware shortest paths can thus be computed by searching for shortest paths in the expanded graph.
+
+The most straight-forward combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
+However, this implies computing a CH on the expanded graph.
+This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{geisberger-turn-table-paper} which we want to avoid.
+
+We therefore only apply the A* search to the expanded graph $G'$. 
+The potentials are computed with respect to the input graph $G$.
+
+Let $(p,q)$ be the target edge and $(x,y)$ the edge for which the A* search wants to compute a potential.
+We construct non-turn-aware CH-Potentials towards $q$.
+The potential for the edge $(x,y)$ is equal to the potential of $y$.
+
+To prove correctness, consider the turn-aware path weight of an optimal path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
+We can lower bound the expression as follows:
+\[
+\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1}) \le \sum_{i=2}^{k-1} \underbrace{w_t(v_{i-1},v_i,v_{i+1})}_{0\le} + \sum_{i=2}^{k-1} \underbrace{w_q(v_i,v_{i+1})}_{w_\ell(v_i,v_{i+1})\le}
+\]
+The $\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ expression is the length of some, not necessarly shortest, $yq$-path in $G$ with $w_\ell$.
+It is by definition not smaller than the shortest $yq$-path length in $G$ with $w_\ell$.
+The shortest $yq$-path length is exactly what CH-Potentials compute.
+The potential is therefore a lower bound.
+It is consistent because it is derived from exact shortest path in $G$ with $w_\ell$.
+
+Unfortunately, the undirected graph underlying the expanded graph is always biconnected.
+The optimization described in \ref{sec:largested-biconnected-component} is therefore ineffective in this setting.
+
+With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH and the preprocessing.
+This allows for a comparatively easy setup and turn-aware routing can be activate or deactivate for every query individually.
+
+\subsection{Time-Dependent Routing}
 

### 2020-06-09 paper/ch_potentials.tex
foo

@@
 The potential is therefore a lower bound.
-It is consistent because it is derived from exact shortest path in $G$ with $w_\ell$.
+It is consistent because it is derived from exact shortest paths in $G$ with $w_\ell$.
 
@@
 
-With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH and the preprocessing.
+With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH and its preprocessing.
 This allows for a comparatively easy setup and turn-aware routing can be activate or deactivate for every query individually.
 
-\subsection{Time-Dependent Routing}
+\subsection{Historic Traffic Patterns}
+
+The classical shortest path problem assumes that edge weights are scalars. 
+However in practice, travel times vary along an edge.
+One source of variance are recurring congestion patterns.
+
+These are modelled using functions as edge weights~\cite{lots-of-td-papers}.
+They map the time of entry into an edge onto the travel time through it.
+Formally, $w_q$ is now a \emph{travel time function} from $E\times \mathbb{R}$ to $\mathbb{R}^+$. 
+$w_q(e, \tau)$ is the travel time through edge $e$ when entering it at $\tau$.
+
+The input to the problem consists of a source node $s$ and a target node $t$ as in the classical problem formulation.
+Additionally, the input contains a source time $\tau_s$.
+A path with edges $e_1,e_2\ldots e_k$ is weighted using $\alpha_k$, which is defined recursively as follows:\[
+\begin{split}
+\alpha_{1} & = w_q(e_1, \tau_s) \\
+\alpha_{k} & = \alpha_{k-1} + w_q(e_1, \alpha_{k-1})
+\end{split}
+\]
+The objective is to find a path that minimizes $\alpha_k$.
+
+If all travel time functions fulfill the \emph{no-waiting property}, this problem can be solved using a straight forward extension of Dijkstra's algorithm \cite{td-alt?}.
+The necessary modification to A* is analogous.
+Without the no-waiting property the problem is in general NP-hard \cite{veit}.
+The no-waiting property states that it is never benficial to wait at a node before entering an edge.
+Formally stated, the following must hold:\[
+\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta
+\]
+
+Usually, travel time functions are stored as piece-wise linear functions.
+Our implementation follows this setup.
+However in priciple, our algorithm works with any representation that enables an efficient function evaluation.
+
+Following \cite{td-alt}, we set $w_\ell(e)$ to the minimum travel time for every edge, i.e., $w_\ell(e) = \min_\tau w_q(e,\tau)$.
+This setup gives us a lower bound weight that is independent of the edge entry time. 
+Howeve, for every entry time, $w_\ell(e)$ is a valid lower bound.
+
+The modifications to make CH-Potentials work are straight forward.
+We compute the CH with respect to $w_\ell$ which maps edges onto scalars.
+With this CH, we can compute CH-Potentials.
+These are then combined with the travel-time-function-aware A* variant of \cite{td-alt?}.
+
+With this setup we extended CH-Potentials in a way that keeps all travel-time-function-related information out of the CH and its preprocessing.
+This way, we avoid the main problems of \cite{tdch,catchup}, which propagate the travel time functions into the CH construction.
+
+\subsection{Live Traffic}
 

### 2020-06-10 paper/ch_potentials.tex
foo

@@
 
-TODO: Cite Sabine VLDB and TopoCore
+Simple Routes: https://dl.acm.org/doi/10.1145/2525314.2525362 example of flexibility need
+
+EV: https://dl.acm.org/doi/10.1145/2666310.2666382
+EV: https://dl.acm.org/doi/10.1145/2820783.2820826
+
+Traffic Planning: https://dl.acm.org/doi/10.1145/2666310.2666406 uses an A* search with adjusted costs
+
+Learning Cost functions with CRP: https://dl.acm.org/doi/10.1145/2820783.2820808 strong argument for flexibility. Must Cite.
+Also Learning Cost functions: https://dl.acm.org/doi/10.1145/2996913.2997004 Must Cite.
+
+Personalized Route Planning: https://dl.acm.org/doi/10.1145/2820783.2820830 exact same motivation argument as our paper. Must Cite. 
+TopoCore: https://dl.acm.org/doi/10.1145/2820783.2820856 Must Cite.
+
+Compute Travel Time Probability Per arc: https://dl.acm.org/doi/10.1145/2820783.2820836 Bullshit contentwise. However, could be cited in TD context.
+
+Travel Time Probability with A*: https://dl.acm.org/doi/10.1145/2996913.2996964
+
+Historic Travel Time Function Computation : https://dl.acm.org/doi/10.1145/2996913.2996988 not really relevant but could be cited.
+
+Most preferred path: https://dl.acm.org/doi/10.1145/3139958.3140029
+
+Off-Topic Important for Map-Matching: https://dl.acm.org/doi/10.1145/2525314.2525466
 
@@
 
+Historic travel time : $w_h(e,\tau)$
+
+Live travel time : $w_c(e)$
+
+Lower bound travel time : $w_\ell(e)$
+
+$\tau_{\mathrm{now}}$ : the current moment in time
+
+$\tau_{\mathrm{soon}}$ : now plus one hour
+
+A* works on $w_q(e,\tau)$ which is defined as follows:
+
+$w_q(e,\tau) = w_c(e)$ for $\tau \le \tau_{\mathrm{soon}}$
+
+If $w_h(e,\tau_{\mathrm{soon}}) < w_c(e)$, then $w_q(e,\tau) = \max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_h(e,\tau)\}$ for $\tau \ge \tau_{\mathrm{soon}}$.
+
+If $w_h(e,\tau_{\mathrm{soon}}) = w_c(e)$, then $w_q(e,\tau) = w_h(e,\tau)$ for $\tau \ge \tau_{\mathrm{soon}}$.
+
+If $w_h(e,\tau_{\mathrm{soon}}) > w_c(e)$, then $w_q(e,\tau) = \min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_h(e,\tau)\}$ for $\tau \ge \tau_{\mathrm{soon}}$.
+
 \section{Experiments}

### 2020-06-11 paper/ch_potentials.tex
foo

--- 2020-06-09 paper/ch_potentials.tex
@@
 Computing a good potential is a challenge of its own.
-To avoid the need for a potential, many hierachical speedup techniques have been developped.
+To avoid the need for a potential, many hierarchical speedup techniques have been developped.
 They achieve speed and optimality but sacrifize flexibility.
@@
 A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{gssv-erlrn-12,dsw-cch-15,dgpw-crprn-13,ss-ehh-12} which do not require such a function.
-A popular hierachical algorithm is Contraction Hierarchy (CH) \cite{gssv-erlrn-12}.
+A popular hierarchical algorithm is Contraction Hierarchy (CH) \cite{gssv-erlrn-12}.
 There is a lot of work that extends CH to handle more complex settings.
@@
 While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
-In this paper, we explore CH-Potentials, a novel combination of A* and the hierachical techniques CH and CCH.
+In this paper, we explore CH-Potentials, a novel combination of A* and the hierarchical techniques CH and CCH.
 We are not the first to combine hierarchical approaches and A* \cite{bdsssw-chgds-10, gkw-blwr-07, bdgwz-sfpcs-19}.
@@
 In this section, we list a few to motivate why flexibility is important.
+In Section~\ref{sec:extensions}, we describe how to handle some with CH-Potentials in detail.
 
 The travel time across an edge is in general not constant over time.
-Live-traffic and historic traffic patterns are examples of how the travel time can change.
+Live and predicted traffic are examples of how the travel time can change.
 Another example are road blockages.
@@
 Some blockage types, such as accidents, are similar to live traffic.
-Other blockages such as reversible one-ways, time-dependent turn-restrictions, or seasonal closures are more similar to historic traffic patterns.
+Other blockages such as reversible one-ways, time-dependent turn-restrictions, or seasonal closures are more similar to predicted traffic.
 
@@
 For example, common options in navigation devices include avoiding highways or tunnels.
+A user preference for simple routes~\cite{https://dl.acm.org/doi/10.1145/2525314.2525362}, i.e., routes with few road changes, may also be considered.  
 These requirements are usally modelled by modifying the edge weights and are highly depend on the driver.
 Sometimes they even depend on his current mood and might change while driving.
+In this paper, we assume that the driver preference is known.
+We refer to \cite{https://dl.acm.org/doi/10.1145/2820783.2820808,https://dl.acm.org/doi/10.1145/2996913.2997004} for approaches to learn these requirements.
 
@@
 Once the truck has delivered his goods, he may again pass the area.
+
+Flexibility can also be used to limit the slope and speed of the roads used during routing.
+This can be useful in the context electric cars to limit energy consumption, increasing the effective range.
+We refer to \cite{https://dl.acm.org/doi/10.1145/2666310.2666382,https://dl.acm.org/doi/10.1145/2820783.2820826} for an introduction to the EV topic.
+
+A necessity for flexibility can also arise as part of a larger optimization.
+For example, \cite{https://dl.acm.org/doi/10.1145/2666310.2666406} adjusts the edge weights to coordinate the routing of large number of vehicle to avoid congestions.
 
@@
 
-TODO: Cite Sabine VLDB and TopoCore
+Personalized Route Planning: https://dl.acm.org/doi/10.1145/2820783.2820830 exact same motivation argument as our paper. Must Cite. 
+TopoCore: https://dl.acm.org/doi/10.1145/2820783.2820856 Must Cite.
 
@@
 \section{Extensions}
+\label{sec:extensions}
 
 A* is a very flexible algorithm. 
-As such there exist many extensions of CH-Potentials to solve extended problem settings.
-While extending with CH-Potentials is comparatively easy, this is usually not the case with competitor algorithms.
+Many extended problems can therefore be solved using the CH-Potentials algorithm.
+In this section, we describe some extended problems.
+
+\subsection{Avoiding Tunnels and/or Highways}
+\label{sec:no-tunnel-highway}
+
+Avoiding tunnels and/or highways is a common feature of navigation devices.
+Implementing this feature with CH-Potentials is easy.
+We set $w_\ell$ to the freeflow travel time. 
+If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
+Otherwise, $w_q$ is also to the freeflow travel time.
+This is all that is necessary to make our CH-Potentials algorithm avoid tunnels and/or highways.
 
 \subsection{Forbidden Turns}
+\label{sec:no-turns}
 
@@
 However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
-Handling turn restrictions correctly is crucial for many applications.
 Such restrictions are usually modelled using turn weights.
 We extend CH-potentials by using zero as lower bound for every turn weight in the potential.
+In this section, we first present the extended problem setting.
+Afterwards, we describe how it can be solved with CH-Potentials.
 
@@
 $w_t$ maps a pair of incident edges onto the time required to perform the turn.
-Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup\{+\infty\}$.
+Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup \{+\infty\}$.
 A forbidden turn is modelled by setting the corresponding $w_t$ to $+\infty$.
@@
 For every pair of successive edges $(x,y)$ and $(y,z)$ in $G$, there is an \emph{expanded edge} in $G'$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
-Notice how an expanded edge weight looks similar to a term of the turn-aware path weight.
-
 A sequence of expanded nodes in the expanded graph $G'$ corresponds to a sequence of edges in the input graph $G$.
@@
 
-The most straight-forward combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
+The simplest combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
 However, this implies computing a CH on the expanded graph.
-This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{geisberger-turn-table-paper} which we want to avoid.
+This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{geisberger-turn-table-paper}.
+We want to avoid these complications.
 
@@
 The potentials are computed with respect to the input graph $G$.
-
 Let $(p,q)$ be the target edge and $(x,y)$ the edge for which the A* search wants to compute a potential.
@@
 
-With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH and its preprocessing.
-This allows for a comparatively easy setup and turn-aware routing can be activate or deactivate for every query individually.
+With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH.
+This allows for a comparatively simple implementation.
 
-\subsection{Historic Traffic Patterns}
+\subsection{Predicted Traffic}
+\label{sec:predicted-traffic}
 
@@
 However in practice, travel times vary along an edge.
-One source of variance are recurring congestion patterns.
+The primary reason is traffic.
+Recurring traffic can be predicted by observing the traffic in the past.
+It is common \cite{lots-of-td-papers} to represent these predictions as \emph{travel time function}.
+An edge weight is no longer a scalar value but a function that maps the entry time onto the traversal time.
+Performing this prediction is an very interesting research topic~\cite{https://dl.acm.org/doi/10.1145/2820783.2820836,https://dl.acm.org/doi/10.1145/2996913.2996988}, but outside of the scope of this paper.
+It is common to refer to routing with predicted traffic as \emph{time-dependent routing}.
+Again, we first formalize the extended problem setting and then describe our solution with CH-Potentials.
 
-These are modelled using functions as edge weights~\cite{lots-of-td-papers}.
-They map the time of entry into an edge onto the travel time through it.
-Formally, $w_q$ is now a \emph{travel time function} from $E\times \mathbb{R}$ to $\mathbb{R}^+$. 
-$w_q(e, \tau)$ is the travel time through edge $e$ when entering it at $\tau$.
-
-The input to the problem consists of a source node $s$ and a target node $t$ as in the classical problem formulation.
+Formally, $w_q$ is a function from $E\times \mathbb{R}$ to $\mathbb{R}^+$. 
+$w_q(e, \tau)$ is the travel time through edge $e$ when entering it at moment $\tau$.
+The input to the extended problem consists of a source node $s$ and a target node $t$, as in the classical problem formulation.
 Additionally, the input contains a source time $\tau_s$.
@@
 
-Following \cite{td-alt}, we set $w_\ell(e)$ to the minimum travel time for every edge, i.e., $w_\ell(e) = \min_\tau w_q(e,\tau)$.
-This setup gives us a lower bound weight that is independent of the edge entry time. 
-Howeve, for every entry time, $w_\ell(e)$ is a valid lower bound.
+To solve the time-dependent routing problem, we employ a stratedgy very similar to TD-ALT~\cite{td-alt}.
+The main difference is that we use CH-Potentials instead of landmarks to guide the A* search.
+We set $w_\ell(e)$ to the minimum travel time for every edge, i.e., $w_\ell(e) = \min_\tau w_q(e,\tau)$.
 
-The modifications to make CH-Potentials work are straight forward.
-We compute the CH with respect to $w_\ell$ which maps edges onto scalars.
-With this CH, we can compute CH-Potentials.
-These are then combined with the travel-time-function-aware A* variant of \cite{td-alt?}.
+We modify the A* search to use the tentative distance at a node $x$ as $\tau$ when evaluating the weight of an outgoing edge $(x,y)$.
+This modification is straight-forward, as our optimizations do not use shortcuts and only travers edges in forward direction.
+For the correctness proof, we refer to~\cite{td-alt?}.
 
-With this setup we extended CH-Potentials in a way that keeps all travel-time-function-related information out of the CH and its preprocessing.
-This way, we avoid the main problems of \cite{tdch,catchup}, which propagate the travel time functions into the CH construction.
+With this setup, we extended CH-Potentials to support time-dependent routing.
+As we were able to keep all travel time functions out of the CH.
+This avoids a lot of algorithmic complications compared to~\cite{TD-CH,CatchUp,TD-CRP}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
 
-\subsection{Live Traffic}
+\subsection{Live and Predicted Traffic}
+\label{sec:live-predicted-traffic}
+
+Beside predicted traffic, we also consider live traffic.
+Live traffic refers to the current traffic situation.
+It is important to distinguish between predicted and live traffic.
+The predicted travel time along an edge was estimated in the past.
+It is possible that it differes from the current travel time significantly, if unexpected events happen.
+An accident is an example of an unexpected event.
+Live traffic data is more accurate for the current moment than predicted data.
+However, just using live traffic data is problematic for long routes as traffic changes while driving.
+At some point, one wants to switch from live traffic back to the predicted traffic.
+In this section, we first describe how to use only live traffic and then describe how to combine it with predicted traffic.
+
+To support live traffic, we set $w_\ell$ to the travel time without any congestions.
+$w_q$ is set to the travel time accounting for the current traffic.
+As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
+With this setup, the CH-Potentials algorithms can make use of live traffic.
+
+To combine live traffic with predicted traffic, we define a modified travel time function.
+Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
+Further, $w_c(e)$ is the travel time according to the live traffic.
+Finally, we denote by $\tau_{\mathrm{soon}}$ the moment up to which we trust the live traffic.
+In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
+We need to make sure that the modified travel time function fulfills the no-waiting property.
+For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
+Our modified travel time function linearly approaches the predicted travel time with a bounded slope.
+
+Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau_{\mathrm{soon}}$.
+Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
+If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
+Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
+We set $w_\ell$ again to the freeflow travel time.
+Finally, we run the modified time-dependent A* described in the previous section on this modified travel time function. 
+In our implementation, we to not modify the representation of $w_p$.
+Instead, we evaluate the formulas above each time that the modified travel time function is evaluated.
+
+With this setup, we extended CH-Potentials to support a combination of live and predicted traffic.
+We did not make any modification to our algorithm, that would make a combination with tunnel and/or highway avoidance or turn-aware routing difficult.
+This straight-forward integration of complex routing problems is the strengh of the CH-Potential algorithm.
 

### 2020-06-11 paper/ch_potentials.tex
foo

@@
 For simplicity, we limit our exposition to the classical shortest path formulation.
-To support algorithm extensions, we avoid two operations that are adverse to generalization.
+To support algorithm extensions, such as time-dependent routing as descibed in Section \ref{sec:predicted-traffic}, we avoid two operations that are adverse to generalization.
 These are:
@@
 The performance of our algorithm depends on this lower bound being close to the actual weight.
-Using the trivial lower bound of zero everywhere, CH-potentials degenerate to an engineered variant of Dijkstra's algorithm.
+Using the trivial lower bound of zero everywhere, CH-potentials degenerate to a variant of Dijkstra's algorithm.
 

### 2020-07-27 paper/ch_potentials.tex
foo
worked on related work

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{A* with Perfect Potentials}
+\title{A* with Perfect Heuristic/Potentials in Road Networks}
 
@@
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
-A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, or Artificial Intelligence.
-However, A*'s performance depends on the availability of a good potential, also called heuristic, function.
-Computing a good potential is a challenge of its own.
-To avoid the need for a potential, many hierarchical speedup techniques have been developped.
-They achieve speed and optimality but sacrifize flexibility.
-In this paper, we use hierarchical techniques to compute potential functions.
+A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
+However, A*'s performance depends on the availability of a good heuristic, also called potential, function.
+Computing a good heuristic/potential is a challenge of its own.
+To avoid the need for a heuristic/potential, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
+These techniques are not based on A* and therefore do not need a heuristic/potential.
+They achieve speed and optimality but sacrifize A*'s flexibility.
+In this paper, we use hierarchical techniques to compute heuristic/potential functions.
+We call our technique CH-Potential.
+Additionally, we describe some A* optimizations to accelerate the processing of low degree nodes, which occur in road networks.
 With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{shortest path, road graphs, goal-directed search, contraction hierarchy}
+\keywords{shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search}
 
@@
 
-Our technique can be applied to a CH and a CCH.
-We describe our work mostly in CH terms.
-Fortunately, all results are also directly applicable to CCH.
-We describe the trade-off between the two variants in Section \ref{TODO}.
-
 \subsection{Applications need Flexibility}
@@
 In Section~\ref{sec:extensions}, we describe how to handle some with CH-Potentials in detail.
+Our argument is a variation of the one made by~\cite{CRP,CCH,DBLP:conf/gis/FunkeS15}.
 
@@
 For example, common options in navigation devices include avoiding highways or tunnels. 
-A user preference for simple routes~\cite{https://dl.acm.org/doi/10.1145/2525314.2525362}, i.e., routes with few road changes, may also be considered.  
 These requirements are usally modelled by modifying the edge weights and are highly depend on the driver.
@@
 In this paper, we assume that the driver preference is known.
-We refer to \cite{https://dl.acm.org/doi/10.1145/2820783.2820808,https://dl.acm.org/doi/10.1145/2996913.2997004} for approaches to learn these requirements.
+We refer to \cite{DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16} for approaches to learn these requirements.
 
@@
 Once the truck has delivered his goods, he may again pass the area.
-
 Flexibility can also be used to limit the slope and speed of the roads used during routing.
-This can be useful in the context electric cars to limit energy consumption, increasing the effective range.
-We refer to \cite{https://dl.acm.org/doi/10.1145/2666310.2666382,https://dl.acm.org/doi/10.1145/2820783.2820826} for an introduction to the EV topic.
-
-A necessity for flexibility can also arise as part of a larger optimization.
-For example, \cite{https://dl.acm.org/doi/10.1145/2666310.2666406} adjusts the edge weights to coordinate the routing of large number of vehicle to avoid congestions.
 
@@
 The performance of our algorithm depends on this lower bound being close to the actual weight.
-Using the trivial lower bound of zero everywhere, CH-potentials degenerate to a variant of Dijkstra's algorithm.
+Using the trivial lower bound of zero everywhere, CH-Potentials degenerate to a variant of Dijkstra's algorithm.
 
@@
 
-TODO: Dijkstra's algorithm \cite{d-ntpcg-59} can be used to solve the stated problem formalization \cite{Veit?}.
+The closest CH-Potential related works are ALT~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07} and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19}.
+Just as our approach, ALT is A*-based.
+ALT computes the potential using landmarks.
+The computed potential values are not optimal and thus of lower quality than the optimal ones computed by CH-Potentials.
+ALT is therefore slower.
 
-TODO: CRP was motivated by personalization.
+CPD-Heursitics make use of compressed path databases (CPD). 
+A CPD can quickly compute the first edge of a shortest path between any two nodes.
+In \cite{DBLP:conf/ijcai/BonoGHS19}, a CPD~\cite{DBLP:conf/socs/StrasserHB14} is used to create a potential.
+In every potential evaluation, a path to the target is compute, whose length is used as potential.
+On a conceptual level, this is very similar to CH-Potentials. 
+Fortunately, CH-Potentials do not have to compute a path per evaluation, which makes CH-Potentials faster.
+Further, the CPD used has quadratic preprocessing running times.
+A CH does not.
+This makes CPDs too expensive on large continental road networks.
+The computed potential values are also optimal.
 
-Variations of this setup exist.
-A very common assumption is that the query edge weights $w_q$ are known during the preprocessing phase.
-In this setup, $w_\ell$ does not exist.
-This allows for very fast query running times, however, changing $w$ requires rerunning the slow preprocessing phase.
-This setup is therefore very unflexible.
-Many hierarchical speedup techniques such as the original CH \cite{gssv-erlrn-12} use this setup.
-Notable exceptions are \cite{dgpw-crprn-13} and \cite{dsw-cch-15} which use a third phase to change the query edge weights.
+As ALT and CPD-Heursitics have a similar structure than CH-Potentials, the extensions described in Section~\ref{sec:extensions} can also be applied to them.
+For example, in~\cite{td-alt} it is extended to time-dependent routing.
 
-Personalized Route Planning: https://dl.acm.org/doi/10.1145/2820783.2820830 exact same motivation argument as our paper. Must Cite. 
-TopoCore: https://dl.acm.org/doi/10.1145/2820783.2820856 Must Cite.
+The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.
 
-TODO: Alt
+TODO: cite \cite{DBLP:journals/tciaig/BaierBHH15} https://ieeexplore.ieee.org/document/6851877
 
-ALT~\cite{gh-cspas-05} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
-For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
-This achieves decent speed-ups and can be used in other problem settings.
-However, the performance is not competitive to hierarchical approaches.
+TODO: cite \cite{DBLP:journals/ai/SharonSFS15} https://www.aaai.org/ocs/index.php/AAAI/AAAI12/paper/viewFile/5062/5239
+
+TODO: cite \cite{DBLP:conf/ijcai/0002UJAKK18} https://arxiv.org/abs/1706.02792
 
@@
 Our technique allows to compute these best possible potentials.
-Within this regard, CH-potentials are \emph{optimal A* potentials}.
+Within this regard, CH-Potentials are \emph{optimal A* potentials}.
 
@@
 Such restrictions are usually modelled using turn weights.
-We extend CH-potentials by using zero as lower bound for every turn weight in the potential.
+We extend CH-Potentials by using zero as lower bound for every turn weight in the potential.
 In this section, we first present the extended problem setting.
@@
 
-Unfortunately, the undirected graph underlying the expanded graph is always biconnected.
+Unfortunately, the undirected graph underlying the expanded graph is always biconnected, if the input graph is strongly connected.
 The optimization described in \ref{sec:largested-biconnected-component} is therefore ineffective in this setting.
@@
 An edge weight is no longer a scalar value but a function that maps the entry time onto the traversal time.
-Performing this prediction is an very interesting research topic~\cite{https://dl.acm.org/doi/10.1145/2820783.2820836,https://dl.acm.org/doi/10.1145/2996913.2996988}, but outside of the scope of this paper.
+Performing this prediction is outside of the scope of this paper.
 It is common to refer to routing with predicted traffic as \emph{time-dependent routing}.

### 2020-07-27 paper/ch_potentials.tex
changed to AAAI 2020 style

@@
-\documentclass[sigconf]{acmart}
+\def\year{2021}\relax
+\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
+\usepackage{aaai20}  % DO NOT CHANGE THIS
+\usepackage{times}  % DO NOT CHANGE THIS
+\usepackage{helvet} % DO NOT CHANGE THIS
+\usepackage{courier}  % DO NOT CHANGE THIS
+\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
+\usepackage{graphicx} % DO NOT CHANGE THIS
+\urlstyle{rm} % DO NOT CHANGE THIS
+\def\UrlFont{\rm}  % DO NOT CHANGE THIS
+\usepackage{graphicx}  % DO NOT CHANGE THIS
+\frenchspacing  % DO NOT CHANGE THIS
+\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
+\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
+
+\usepackage{amssymb}
+\usepackage{amsmath}
+
+\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
+
 %This is a template for producing LIPIcs articles.
@@
 
-\usepackage{wrapfig}
 \usepackage{booktabs}
@@
 
-\begin{document}
-
-%
-%% The "title" command has an optional parameter,
-%% allowing the author to define a "short title" to be used in page headers.
+\pdfinfo{
+/Title (A* with Perfect Heuristic/Potentials in Road Networks)
+/Author (Ben Strasser, Tim Zeitz)
+/Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
+}
 \title{A* with Perfect Heuristic/Potentials in Road Networks}
+\author{Ben Strasser\\
+academia@ben-strasser.net\\
+\And
+Tim Zeitz\\
+tim.zeitz@kit.edu\\
+Karlsruhe Institute of Technology
+}
 
@@
 %% used to denote shared contribution to the research.
-\author{Ben Strasser}
-\email{academia@ben-strasser.net}
-\orcid{1234-5678-9012}
-\affiliation{}
+%\author{
+%Ben Strasser
 
-\author{Tim Zeitz}
-\email{tim.zeitz@kit.edu}
-\affiliation{%
-  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
-  \city{Karlsruhe}
-  \country{Germany}
-}
+%\author{Tim Zeitz}
+%tim.zeitz@kit.edu}
+%\affiliation{%
+%  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
+%  \city{Karlsruhe}
+%  \country{Germany}
+%}
+
+\begin{document}
+
+\maketitle
 
@@
 \end{abstract}
-
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-\begin{CCSXML}
-<ccs2012>
- <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-  <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
-</ccs2012>
-\end{CCSXML}
-
-% \ccsdesc[500]{Computer systems organization~Embedded systems}
-% \ccsdesc[300]{Computer systems organization~Redundancy}
-% \ccsdesc{Computer systems organization~Robotics}
-% \ccsdesc[100]{Networks~Network reliability}
-\ccsdesc[500]{Theory of computation~Shortest paths}
-
-%%
-%% Keywords. The author(s) should pick words that accurately describe
-%% the work being presented. Separate the keywords with commas.
-\keywords{shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search}
-
-%% A "teaser" image appears between the author and affiliation
-%% information and the body of the document, and typically spans the
-%% page.
-% \begin{teaserfigure}
-%   \includegraphics[width=\textwidth]{sampleteaser}
-%   \caption{Seattle Mariners at Spring Training, 2010.}
-%   \Description{Enjoying the baseball game from the third-base
-%   seats. Ichiro Suzuki preparing to bat.}
-%   \label{fig:teaser}
-% \end{teaserfigure}
-
-%%
-%% This command processes the author and affiliation and title
-%% information and builds the first part of the formatted document.
-\maketitle
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliography{dblp}
 \bibliography{references}
-
-% \appendix
-
+\bibliographystyle{aaai}
 \end{document}

### 2020-07-27 paper/formatting-instructions-latex-2020.tex
changed to AAAI 2020 style

\def\year{2020}\relax
%File: formatting-instruction.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai20}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet} % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{graphicx}  % DO NOT CHANGE THIS
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
%\nocopyright
%PDF Info Is REQUIRED.
% For /Author, add all authors within the parentheses, separated by commas. No accents or commands.
% For /Title, add Title in Mixed Case. No accents or commands. Retain the parentheses.
 \pdfinfo{
/Title (AAAI Press Formatting Instructions for Authors Using LaTeX -- A Guide)
/Author (AAAI Press Staff, Pater Patel Schneider, Sunil Issar, J. Scott Penberthy, George Ferguson, Hans Guesgen)
} %Leave this	
% /Title ()
% Put your actual complete title (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case
% Leave the space between \Title and the beginning parenthesis alone
% /Author ()
% Put your actual complete list of authors (no codes, scripts, shortcuts, or LaTeX commands) within the parentheses in mixed case. 
% Each author should be only by a comma. If the name contains accents, remove them. If there are any LaTeX commands, 
% remove them.

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{caption} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \natbib} -- This package is specifically forbidden -- use the following workaround:
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai20.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\setlength\titlebox{2.5in} % If your paper contains an overfull \vbox too high warning at the beginning of the document, use this
% command to correct it. You may not alter the value below 2.5 in
\title{AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide }
%Your title must be in mixed case, not sentence case. 
% That means all verbs (including short verbs like be, is, using,and go), 
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
\author{Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{Primarily Mike Hamilton of the Live Oak Press, LLC, with help from the AAAI Publications Committee}\\ \Large \textbf{AAAI Style Contributions by
Pater Patel Schneider,} \\ \Large \textbf{Sunil Issar, J. Scott Penberthy, George Ferguson, Hans Guesgen}\\ % All authors must be in the same font size and format. Use \Large and \textbf to achieve this result when breaking a line
\textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\ %If you have multiple authors and multiple affiliations
% use superscripts in text and roman font to identify them. For example, Sunil Issar,\textsuperscript{\rm 2} J. Scott Penberthy\textsuperscript{\rm 3} George Ferguson,\textsuperscript{\rm 4} Hans Guesgen\textsuperscript{\rm 5}. Note that the comma should be placed BEFORE the superscript for optimum readability
2275 East Bayshore Road, Suite 160\\
Palo Alto, California 94303\\
publications20@aaai.org % email address must be in roman text type, not monospace or sans serif
}
 \begin{document}

\maketitle

\begin{abstract}
AAAI creates proceedings, working notes, and technical reports directly from electronic source furnished by the authors. To ensure that all papers in the publication have a uniform appearance, authors must adhere to the following instructions. 
\end{abstract}

\noindent Congratulations on having a paper selected for inclusion in an AAAI Press proceedings or technical report! This document details the requirements necessary to get your accepted paper published using PDF\LaTeX{}. If you are using Microsoft Word, instructions are provided in a different document. AAAI Press does not support any other formatting software.

The instructions herein are provided as a general guide for experienced \LaTeX{} users. If you do not know how to use \LaTeX{}, please obtain assistance locally. AAAI cannot provide you with support and the accompanying style files are \textbf{not} guaranteed to work. If the results you obtain are not in accordance with the specifications you received, you must correct your source file to achieve the correct result.

These instructions are generic. Consequently, they do not include specific dates, page charges, and so forth. Please consult your specific written conference instructions for details regarding your submission. Please review the entire document for specific instructions that might apply to your particular situation. All authors must comply with the following:

\begin{itemize}
\item You must use the 2020 AAAI Press \LaTeX{} style file and the aaai.bst bibliography style file, which are located in the 2020 AAAI Author Kit (aaai20.sty and aaai.bst).
\item You must complete, sign, and return by the deadline the AAAI copyright form (unless directed by AAAI Press to use the AAAI Distribution License instead).
\item You must read and format your paper source and PDF according to the formatting instructions for authors.
\item You must submit your electronic files and abstract using our electronic submission form \textbf{on time.}
\item You must pay any required page or formatting charges to AAAI Press so that they are received by the deadline.
\item You must check your paper before submitting it, ensuring that it compiles without error, and complies with the guidelines found in the AAAI Author Kit.
\end{itemize}

\section{Copyright}
All papers submitted for publication by AAAI Press must be accompanied by a valid signed copyright form. There are no exceptions to this requirement. You must send us the original version of this form. However, to meet the deadline, you may fax (1-650-321-4457) or scan and e-mail the form (pubforms20@aaai.org) to AAAI by the submission deadline, and then mail the original via postal mail to the AAAI office. If you fail to send in a signed copyright or permission form, we will be unable to publish your paper. There are \textbf{no exceptions} to this policy.You will find PDF versions of the AAAI copyright and permission to distribute forms in the AAAI AuthorKit.

\section{Formatting Requirements in Brief}
We need source and PDF files that can be used in a variety of ways and can be output on a variety of devices. The design and appearance of the paper is strictly governed by the aaai style file (aaai20.sty). 
\textbf{You must not make any changes to the aaai style file, nor use any commands, packages, style files, or macros within your own paper that alter that design, including, but not limited to spacing, floats, margins, fonts, font size, and appearance.} AAAI imposes requirements on your source and PDF files that must be followed. Most of these requirements are based on our efforts to standardize conference manuscript properties and layout. All papers submitted to AAAI for publication will be recompiled for standardization purposes. Consequently, every paper submission must comply with the following requirements:

\begin{quote}
\begin{itemize}
\item Your .tex file must compile in PDF\LaTeX{} --- ( you may not include .ps or .eps figure files.)
\item All fonts must be embedded in the PDF file --- including includes your figures.
\item Modifications to the style file, whether directly or via commands in your document may not ever be made, most especially when made in an effort to avoid extra page charges or make your paper fit in a specific number of pages.
\item No type 3 fonts may be used (even in illustrations).
\item You may not alter the spacing above and below captions, figures, headings, and subheadings.
\item You may not alter the font sizes of text elements, footnotes, heading elements, captions, or title information (for references and mathematics, please see the the limited exceptions provided herein).
\item You may not alter the line spacing of text.
\item Your title must follow Title Case capitalization rules (not sentence case).
\item Your .tex file must include completed metadata to pass-through to the PDF (see PDFINFO below)
\item \LaTeX{} documents must use the Times or Nimbus font package (you may not use Computer Modern for the text of your paper).
\item No \LaTeX{} 209 documents may be used or submitted.
\item Your source must not require use of fonts for non-Roman alphabets within the text itself. If your paper includes symbols in other languages (such as, but not limited to, Arabic, Chinese, Hebrew, Japanese, Thai, Russian and other Cyrillic languages), you must restrict their use to bit-mapped figures. Fonts that require non-English language support (CID and Identity-H) must be converted to outlines or 300 dpi bitmap or removed from the document (even if they are in a graphics file embedded in the document). 
\item Two-column format in AAAI style is required for all papers.
\item The paper size for final submission must be US letter without exception.
\item The source file must exactly match the PDF.
\item The document margins may not be exceeded (no overfull boxes).
\item The number of pages and the file size must be as specified for your event.
\item No document may be password protected.
\item Neither the PDFs nor the source may contain any embedded links or bookmarks (no hyperref or navigator packages).
\item Your source and PDF must not have any page numbers, footers, or headers (no pagestyle commands).
\item Your PDF must be compatible with Acrobat 5 or higher.
\item Your \LaTeX{} source file (excluding references) must consist of a \textbf{single} file (use of the ``input" command is not allowed.
\item Your graphics must be sized appropriately outside of \LaTeX{} (do not use the ``clip" or ``trim'' command) .
\end{itemize}
\end{quote}

If you do not follow these requirements, you will be required to correct the deficiencies and resubmit the paper. A resubmission fee will apply.

\section{What Files to Submit}
You must submit the following items to ensure that your paper is published:
\begin{itemize}
\item A fully-compliant PDF file that includes PDF metadata.
\item Your \LaTeX{} source file submitted as a \textbf{single} .tex file (do not use the ``input" command to include sections of your paper --- every section must be in the single source file). (The only allowable exception is .bib file, which should be included separately). 
\item The bibliography (.bib) file(s).
\item Your source must compile on our system, which includes only standard \LaTeX{} 2019 TeXLive support files. 
\item Only the graphics files used in compiling paper.
\item The \LaTeX{}-generated files (e.g. .aux,  .bbl file, PDF, etc.).
\end{itemize}

Your \LaTeX{} source will be reviewed and recompiled on our system (if it does not compile, you will be required to resubmit, which will incur fees). \textbf{Do not submit your source in multiple text files.} Your single \LaTeX{} source file must include all your text, your bibliography (formatted using aaai.bst), and any custom macros.

Your files should work without any supporting files (other than the program itself) on any computer with a standard \LaTeX{} distribution.

\textbf{Do not send files that are not actually used in the paper.} We don't want you to send us any files not needed for compiling your paper, including, for example, this instructions file, unused graphics files, style files, additional material sent for the purpose of the paper review, and so forth.

\textbf{Do not send supporting files that are not actually used in the paper.} We don't want you to send us any files not needed for compiling your paper, including, for example, this instructions file, unused graphics files, style files, additional material sent for the purpose of the paper review, and so forth.

\textbf{Obsolete style files.} The commands for some common packages (such as some used for algorithms), may have changed. Please be certain that you are not compiling your paper using old or obsolete style files. 
\textbf{Final Archive.} Place your PDF and source files in a single archive which should be compressed using .zip. The final file size may not exceed 10 MB.
Name your source file with the last (family) name of the first author, even if that is not you.

\section{Using \LaTeX{} to Format Your Paper}

The latest version of the AAAI style file is available on AAAI's website. Download this file and place it in the \TeX\ search path. Placing it in the same directory as the paper should also work. You must download the latest version of the complete AAAI Author Kit so that you will have the latest instruction set and style file.

\subsection{Document Preamble}

In the \LaTeX{} source for your paper, you \textbf{must} place the following lines as shown in the example in this subsection. This command set-up is for three authors. Add or subtract author and address lines as necessary, and uncomment the portions that apply to you. In most instances, this is all you need to do to format your paper in the Times font. The helvet package will cause Helvetica to be used for sans serif. These files are part of the PSNFSS2e package, which is freely available from many Internet sites (and is often part of a standard installation).

Leave the setcounter for section number depth commented out and set at 0 unless you want to add section numbers to your paper. If you do add section numbers, you must uncomment this line and change the number to 1 (for section numbers), or 2 (for section and subsection numbers). The style file will not work properly with numbering of subsubsections, so do not use a number higher than 2.

If (and only if) your author title information will not fit within the specified height allowed, put \textbackslash setlength \textbackslash titlebox{2.5in} in your preamble. Increase the height until the height error disappears from your log. You may not use the \textbackslash setlength command elsewhere in your paper, and it may not be used to reduce the height of the author-title box.

\subsubsection{The Following Must Appear in Your Preamble}
\begin{quote}
\begin{scriptsize}\begin{verbatim}
\documentclass[letterpaper]{article}
\usepackage{aaai20}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage[hyphens]{url} 
\usepackage{graphicx}
\urlstyle{rm}
\def\UrlFont{\rm}
\usepackage{graphicx}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}
% Add additional packages here, but check 
% the list of disallowed packages 
% (including, but not limited to
% authblk, caption, CJK, float, fullpage, geometry, 
% hyperref, layout, nameref, natbib, savetrees, 
% setspace, titlesec, tocbibind, ulem)
% and illegal commands provided in the 
% common formatting errors document
% included in the  Author Kit before doing so. 
%
% PDFINFO
% You are required to complete the following
% for pass-through to the PDF. 
% No LaTeX commands of any kind may be
% entered. The parentheses and spaces 
% are an integral part of the 
% pdfinfo script and must not be removed.
%
\pdfinfo{
/Title (Type Your Paper Title Here in Mixed Case)
/Author (John Doe, Jane Doe)
/Keywords (Input your keywords in this optional area)
}
%
% Section Numbers
% Uncomment if you want to use section numbers
% and change the 0 to a 1 or 2
% \setcounter{secnumdepth}{0}

% Title and Author Information Must Immediately Follow
% the pdfinfo within the preamble
%
\title{Title}\\
\author\{Author 1 \ and Author 2\\
Address line\\
Address line\\
\ And\\
Author 3\\
Address line\\
Address line
}\\
\end{verbatim}\end{scriptsize}
\end{quote}

\subsection{Preparing Your Paper}

After the preamble above, you should prepare your paper as follows:
\begin{quote}
\begin{scriptsize}\begin{verbatim}
%
\begin{document}
\maketitle
\begin{abstract}
%...
\end{abstract}\end{verbatim}\end{scriptsize}
\end{quote}

\subsubsection{The Following Must Conclude Your Document}
\begin{quote}
\begin{scriptsize}\begin{verbatim}
% References and End of Paper
% These lines must be placed at the end of your paper
\bibliography{Bibliography-File}
\bibliographystyle{aaai}
\end{document}
\end{verbatim}\end{scriptsize}
\end{quote}

\subsection{Inserting Document Metadata with \LaTeX{}}
PDF files contain document summary information that enables us to create an Acrobat index (pdx) file, and also allows search engines to locate and present your paper more accurately. \textit{Document metadata for author and title are REQUIRED.} You may not apply any script or macro to implementation of the title, author, and metadata information in your paper.

\textit{Important:} Do not include \textit{any} \LaTeX{} code or nonascii characters (including accented characters) in the metadata. The data in the metadata must be completely plain ascii. It may not include slashes, accents, linebreaks, unicode, or any \LaTeX{} commands. Type the title exactly as it appears on the paper (minus all formatting). Input the author names in the order in which they appear on the paper (minus all accents), separating each author by a comma. You may also include keywords in the optional Keywords field.

\begin{quote}
\begin{scriptsize}\begin{verbatim}
\begin{document}\\
\maketitle\\
...\\
\bibliography{Bibliography-File}\\
\bibliographystyle{aaai}\\
\end{document}\\
\end{verbatim}\end{scriptsize}
\end{quote}

\subsection{Commands and Packages That May Not Be Used}
\begin{table*}[t]
\centering
\caption{Commands that must not be used}\smallskip
\begin{tabular}{l|l|l|l}
\textbackslash abovecaption & 
\textbackslash abovedisplay & 
\textbackslash addevensidemargin & 
\textbackslash addsidemargin \\ 
\textbackslash addtolength & 
\textbackslash baselinestretch & 
\textbackslash belowcaption & 
\textbackslash belowdisplay \\ 
\textbackslash break &
\textbackslash clearpage & 
\textbackslash clip & 
\textbackslash columnsep \\ 
\textbackslash float & 
\textbackslash input & 
\textbackslash input & 
\textbackslash linespread \\ 
\textbackslash newpage &
\textbackslash pagebreak & 
\textbackslash renewcommand & 
\textbackslash setlength \\ 
\textbackslash text height & 
\textbackslash tiny & 
\textbackslash top margin & 
\textbackslash trim \\ 
\textbackslash vskip\{- & 
\textbackslash vspace\{- \\ 
\end{tabular}
%}
\label{table1}
\end{table*}

\begin{table}[t]
\caption{LaTeX style packages that must not be used.}\smallskip
\centering
%\resizebox{.95\columnwidth}{!}{
\smallskip\begin{tabular}{l|l|l|l}
authblk & babel & caption & cjk\\
dvips & epsf & epsfig & euler\\
float & fullpage & geometry & graphics\\
hyperref & layout & linespread & lmodern\\
maltepaper & natbib & navigator & pdfcomment\\
pgfplots & psfig & pstricks & t1enc \\
titlesec & tocbind & ulem\\
\end{tabular}
\label{table2}
\end{table}

There are a number of packages, commands, scripts, and macros that are incompatable with aaai20.sty. The common ones are listed in tables \ref{table1} and \ref{table2}. Generally, if a command, package, script, or macro alters floats, margins, fonts, sizing, linespacing, or the presentation of the references and citations, it is unacceptable. Note that negative vskip and vspace may not be used except in certain rare occurances, and may never be used around tables, figures, captions, sections, subsections, subsections, or references.

\subsection{Page Breaks}
For your final camera ready copy, you must not use any page break commands. References must flow directly after the text without breaks. Note that some conferences require references to be on a separate page during the review process. AAAI Press, however, does not require this condition for the final paper.

\subsection{Paper Size, Margins, and Column Width}
Papers must be formatted to print in two-column format on 8.5 x 11 inch US letter-sized paper. The margins must be exactly as follows: 
\begin{itemize}
\item Top margin: .75 inches
\item Left margin: .75 inches
\item Right margin: .75 inches
\item Bottom margin: 1.25 inches
\end{itemize}

The default paper size in most installations of \LaTeX{} is A4. However, because we require that your electronic paper be formatted in US letter size, the preamble we have provided includes commands that alter the default to US letter size. Please note that using any other package to alter page size (such as, but not limited to the Geometry package) will result in your final paper being returned to you for correction and payment of a resubmission fee.

\subsubsection{Column Width and Margins.}
To ensure maximum readability, your paper must include two columns. Each column should be 3.3 inches wide (slightly more than 3.25 inches), with a .375 inch (.952 cm) gutter of white space between the two columns. The aaai20.sty file will automatically create these columns for you.

\subsection{Overlength Papers}
If your paper is too long, turn on \textbackslash frenchspacing, which will reduce the space after periods. Next, shrink the size of your graphics. Use \textbackslash centering instead of \textbackslash begin\{center\} in your figure environment. For mathematical environments, you may reduce fontsize {\bf but not below 6.5 point}. You may also alter the size of your bibliography and tables by inserting \textbackslash fontsize\{9.5pt\}\{10.5pt\} \textbackslash selectfont
right before the bibliography (the minimum size is \textbackslash fontsize\{9.0pt\}\{10.0pt\}. You may not use any commands that further reduce point size. Be careful when scaling tables and figures. Tables that do not fit in a single column, even after the font size reduction to 9 point, must be placed across double columns. If your table won't fit within the margins even in double columns, you must split it. We will not accept single (or double) column tables that have been scaled to anything below 9 point.

Commands that alter page layout are forbidden. These include \textbackslash columnsep, \textbackslash topmargin, \textbackslash topskip, \textbackslash textheight, \textbackslash textwidth, \textbackslash oddsidemargin, and \textbackslash evensizemargin (this list is not exhaustive). If you alter page layout, you will be required to pay the page fee \textit{plus} a reformatting fee. Other commands that are questionable and may cause your paper to be rejected include \textbackslash parindent, and \textbackslash parskip. Commands that alter the space between sections are forbidden. The title sec package is not allowed. Regardless of the above, if your paper is obviously ``squeezed" it is not going to to be accepted. Options for reducing the length of a paper include reducing the size of your graphics, cutting text, or paying the extra page charge (if it is offered).

\subsection{Figures}
Your paper must compile in PDF\LaTeX{}. Consequently, all your figures must be .jpg, .png, or .pdf. You may not use the .gif (the resolution is too low), .ps, or .eps file format for your figures.

If you normally create your figures using pgfplots, please create the figures first, and then import them as pdfs with proper bounding boxes, as the bounding and trim boxes created by pfgplots are fragile, and not valid.

When you include your figures, you must crop them \textbf{outside} of \LaTeX{}. The command \textbackslash includegraphics*[clip=true, viewport 0 0 10 10]{...} might result in a PDF that looks great, but the image is \textbf{not really cropped.} The full image can reappear (and obscure whatever it is overlapping) when page numbers are applied or color space is standardized. Figures \ref{fig1}, and \ref{fig2} display some unwanted results that often occur.

Do not use minipage to group figures. Additionally, the font and size of figure captions must be 10 point roman. Do not make them smaller, bold, or italic. (Individual words may be italicized if the context requires differentiation.)

\begin{figure}[t]
\centering
\includegraphics[width=0.9\columnwidth]{figure1} % Reduce the figure size so that it is slightly narrower than the column. Don't use precise values for figure width.This setup will avoid overfull boxes. 
\caption{Using the trim and clip commands produces fragile layers that can result in disasters (like this one from an actual paper) when the color space is corrected or the PDF combined with others for the final proceedings. Crop your figures properly in a graphics program -- not in LaTeX}
\label{fig1}
\end{figure}

\begin{figure*}[t]
\centering
\includegraphics[width=0.8\textwidth]{figure2} % Reduce the figure size so that it is slightly narrower than the column.
\caption{Adjusting the bounding box instead of actually removing the unwanted data resulted multiple layers in this paper. It also needlessly increased the PDF size. In this case, the size of the unwanted layer doubled the paper's size, and produced the following surprising results in final production. Crop your figures properly in a graphics program. Don't just alter the bounding box.}
\label{fig2}
\end{figure*}

% Using the \centering command instead of \begin{center} ... \end{center} will save space
% Positioning your figure at the top of the page will save space and make the paper more readable
% Using 0.95\columnwidth in conjunction with the

\subsection{Type Font and Size}
Your paper must be formatted in Times Roman or Nimbus. We will not accept papers formatted using Computer Modern or Palatino or some other font as the text or heading typeface. Sans serif, when used, should be Courier. Use Symbol or Lucida or Computer Modern for \textit{mathematics only. }

Do not use type 3 fonts for any portion of your paper, including graphics. Type 3 bitmapped fonts are designed for fixed resolution printers. Most print at 300 dpi even if the printer resolution is 1200 dpi or higher. They also often cause high resolution imagesetter devices and our PDF indexing software to crash. Consequently, AAAI will not accept electronic files containing obsolete type 3 fonts. Files containing those fonts (even in graphics) will be rejected.

Fortunately, there are effective workarounds that will prevent your file from embedding type 3 bitmapped fonts. The easiest workaround is to use the required times, helvet, and courier packages with \LaTeX{}2e. (Note that papers formatted in this way will still use Computer Modern for the mathematics. To make the math look good, you'll either have to use Symbol or Lucida, or you will need to install type 1 Computer Modern fonts --- for more on these fonts, see the section ``Obtaining Type 1 Computer Modern.")

If you are unsure if your paper contains type 3 fonts, view the PDF in Acrobat Reader. The Properties/Fonts window will display the font name, font type, and encoding properties of all the fonts in the document. If you are unsure if your graphics contain type 3 fonts (and they are PostScript or encapsulated PostScript documents), create PDF versions of them, and consult the properties window in Acrobat Reader.

The default size for your type must be ten-point with twelve-point leading (line spacing). Start all pages (except the first) directly under the top margin. (See the next section for instructions on formatting the title page.) Indent ten points when beginning a new paragraph, unless the paragraph begins directly below a heading or subheading.

\subsubsection{Obtaining Type 1 Computer Modern for \LaTeX{}.}

If you use Computer Modern for the mathematics in your paper (you cannot use it for the text) you may need to download type 1 Computer fonts. They are available without charge from the American Mathematical Society:
http://www.ams.org/tex/type1-fonts.html.

\subsubsection{Nonroman Fonts}
If your paper includes symbols in other languages (such as, but not limited to, Arabic, Chinese, Hebrew, Japanese, Thai, Russian and other Cyrillic languages), you must restrict their use to bit-mapped figures.

\subsection{Title and Authors}
Your title must appear in mixed case (nouns, pronouns, and verbs are capitalized) near the top of the first page, centered over both columns in sixteen-point bold type (twenty-four point leading). This style is called ``mixed case," which means that means all verbs (including short verbs like be, is, using,and go), nouns, adverbs, adjectives, and pronouns should be capitalized, (including both words in hyphenated terms), while articles, conjunctions, and prepositions are lower case unless they directly follow a colon or long dash. Author's names should appear below the title of the paper, centered in twelve-point type (with fifteen point leading), along with affiliation(s) and complete address(es) (including electronic mail address if available) in nine-point roman type (the twelve point leading). (If the title is long, or you have many authors, you may reduce the specified point sizes by up to two points.) You should begin the two-column format when you come to the abstract.

\subsubsection{Formatting Author Information}
Author information can be set in a number of different styles, depending on the number of authors and the number of affiliations you need to display. In formatting your author information, however, you may not use a table nor may you employ the \textbackslash authorblk.sty package. For several authors from the same institution, please just separate with commas:

\begin{quote}\begin{scriptsize}\begin{verbatim}
\author{Author 1, ... Author n\\ 
Address line \\ ... \\ Address line}
\end{verbatim}\end{scriptsize}\end{quote}

\noindent If the names do not fit well on one line use:

\begin{quote}\begin{scriptsize}\begin{verbatim}
\author{Author 1} ... \\
{\bf \Large Author ... Author}\\ 
Address line \\ ... \\ Address line
}
\end{verbatim}\end{scriptsize}\end{quote}

\noindent For two (or three) authors from different institutions, use \textbackslash And:

\begin{quote}\begin{scriptsize}\begin{verbatim}
\author{Author 1\\ Address line \\ ... \\ Address line 
\And ... \And Author n\\
Address line\\ ... \\ Address line}
\end{verbatim}\end{scriptsize}\end{quote}

\noindent To start a separate ``row" of authors, use \textbackslash AND:

If the title and author information does not fit in the area allocated, place
\textbackslash setlength\textbackslash titlebox\{\textit{height}\}
after the \textbackslash documentclass line where \{\textit{height}\} is 2.5in or greater. (This one of the only allowable uses of the setlength command. Check with AAAI Press before using it elsewhere.)

\subsubsection{Formatting Author Information --- Alternative Method}
If your paper has a large number of authors from different institutions, you may use the following alternative method for displaying the author information.

\begin{quote}\begin{scriptsize}\begin{verbatim}
\author{AuthorOne},\textsuperscript{\rm 1} 
AuthorTwo,\textsuperscript{\rm 2}
AuthorThree,\textsuperscript{\rm 3}
\\bf \Large AuthorFour,\textsuperscript{\rm 4} 
AuthorFive, \textsuperscript{\rm 5}\\
\textsuperscript{\rm 1}AffiliationOne, 
\textsuperscript{\rm 2}AffiliationTwo, 
\textsuperscript{\rm 3}AffiliationThree\\
\textsuperscript{\rm 4}AffiliationFour, 
\textsuperscript{\rm 5}AffiliationFive 
\{email, email\}@affiliation.com,
email@affiliation.com,
email@affiliation.com, 
email@affiliation.com
\end{verbatim}\end{scriptsize}\end{quote}

Note that you should break the author list before it extends into the right column margin. Put a line break, followed by   \textbackslash bf  \textbackslash Large to put the second line of authors in the same font and size as the first line (you may not make authors names smaller to save space.) Affiliations can be broken with a simple line break (\textbackslash  \textbackslash).
\subsection{\LaTeX{} Copyright Notice}
The copyright notice automatically appears if you use aaai20.sty. It has been hardcoded and may not be disabled.

\subsection{Credits}
Any credits to a sponsoring agency should appear in the acknowledgments section, unless the agency requires different placement. If it is necessary to include this information on the front page, use
\textbackslash thanks in either the \textbackslash author or \textbackslash title commands.
For example:
\begin{quote}
\begin{small}
\textbackslash title\{Very Important Results in AI\textbackslash thanks\{This work is
 supported by everybody.\}\}
\end{small}
\end{quote}
Multiple \textbackslash thanks commands can be given. Each will result in a separate footnote indication in the author or title with the corresponding text at the botton of the first column of the document. Note that the \textbackslash thanks command is fragile. You will need to use \textbackslash protect.

Please do not include \textbackslash pubnote commands in your document.

\subsection{Abstract}
Follow the example commands in this document for creation of your abstract. The command \textbackslash begin\{abstract\} will automatically indent the text block. Please do not indent it further. {Do not include references in your abstract!}

\subsection{Page Numbers}

Do not \textbf{ever} print any page numbers on your paper. The use of \textbackslash pagestyle is forbidden.

\subsection{Text }
The main body of the paper must be formatted in black, ten-point Times Roman with twelve-point leading (line spacing). You may not reduce font size or the linespacing. Commands that alter font size or line spacing (including, but not limited to baselinestretch, baselineshift, linespread, and others) are expressly forbidden. In addition, you may not use color in the text.

\subsection{Citations}
Citations within the text should include the author's last name and year, for example (Newell 1980). Append lower-case letters to the year in cases of ambiguity. Multiple authors should be treated as follows: (Feigenbaum and Engelmore 1988) or (Ford, Hayes, and Glymour 1992). In the case of four or more authors, list only the first author, followed by et al. (Ford et al. 1997).

\subsection{Extracts}
Long quotations and extracts should be indented ten points from the left and right margins.

\begin{quote}
This is an example of an extract or quotation. Note the indent on both sides. Quotation marks are not necessary if you offset the text in a block like this, and properly identify and cite the quotation in the text.

\end{quote}

\subsection{Footnotes}
Avoid footnotes as much as possible; they interrupt the reading of the text. When essential, they should be consecutively numbered throughout with superscript Arabic numbers. Footnotes should appear at the bottom of the page, separated from the text by a blank line space and a thin, half-point rule.

\subsection{Headings and Sections}
When necessary, headings should be used to separate major sections of your paper. Remember, you are writing a short paper, not a lengthy book! An overabundance of headings will tend to make your paper look more like an outline than a paper. The aaai.sty package will create headings for you. Do not alter their size nor their spacing above or below.

\subsubsection{Section Numbers}
The use of section numbers in AAAI Press papers is optional. To use section numbers in \LaTeX{}, uncomment the setcounter line in your document preamble and change the 0 to a 1 or 2. Section numbers should not be used in short poster papers.

\subsubsection{Section Headings.}
Sections should be arranged and headed as follows:

\subsubsection{Acknowledgments.}
The acknowledgments section, if included, appears after the main body of text and is headed ``Acknowledgments." This section includes acknowledgments of help from associates and colleagues, credits to sponsoring agencies, financial support, and permission to publish. Please acknowledge other contributors, grant support, and so forth, in this section. Do not put acknowledgments in a footnote on the first page. If your grant agency requires acknowledgment of the grant on page 1, limit the footnote to the required statement, and put the remaining acknowledgments at the back. Please try to limit acknowledgments to no more than three sentences.

\subsubsection{Appendices.}
Any appendices follow the acknowledgments, if included, or after the main body of text if no acknowledgments appear.

\subsubsection{References}
The references section should be labeled ``References" and should appear at the very end of the paper (don't end the paper with references, and then put a figure by itself on the last page). A sample list of references is given later on in these instructions. Please use a consistent format for references. Poorly prepared or sloppy references reflect badly on the quality of your paper and your research. Please prepare complete and accurate citations.

\subsection{Illustrations and Figures}
Figures, drawings, tables, and photographs should be placed throughout the paper near the place where they are first discussed. Do not group them together at the end of the paper. If placed at the top or bottom of the paper, illustrations may run across both columns. Figures must not invade the top, bottom, or side margin areas. Figures must be inserted using the \textbackslash usepackage\{graphicx\}. Number figures sequentially, for example, figure 1, and so on.

The illustration number and caption should appear under the illustration. Labels, and other text with the actual illustration must be at least nine-point type.

If your paper includes illustrations that are not compatible with PDF\TeX{} (such as .eps or .ps documents), you will need to convert them. The epstopdf package will usually work for eps files. You will need to convert your ps files to PDF however.

\subsubsection{Low-Resolution Bitmaps.}
You may not use low-resolution (such as 72 dpi) screen-dumps and GIF files---these files contain so few pixels that they are always blurry, and illegible when printed. If they are color, they will become an indecipherable mess when converted to black and white. This is always the case with gif files, which should never be used. The resolution of screen dumps can be increased by reducing the print size of the original file while retaining the same number of pixels. You can also enlarge files by manipulating them in software such as PhotoShop. Your figures should be 300 dpi when incorporated into your document.

\subsubsection{\LaTeX{} Overflow.}
\LaTeX{} users please beware: \LaTeX{} will sometimes put portions of the figure or table or an equation in the margin. If this happens, you need to scale the figure or table down, or reformat the equation.{ \bf Check your log file!} You must fix any overflow into the margin (that means no overfull boxes in \LaTeX{}). \textbf{Nothing is permitted to intrude into the margin or gutter.}

The most efficient and trouble-free way to fix overfull boxes in graphics is with the following command:

\begin{quote}
\begin{small}
\textbackslash resizebox\{.95\textbackslash columnwidth\}{!}\{
\}
\end{small}\end{quote}

Be certain, however, that your figures remain legible without magnification.

\subsubsection{Using Color.}
Use of color is restricted to figures only. It must be WACG 2.0 compliant. (That is, the contrast ratio must be greater than 4.5:1 no matter the font size.) It must be CMYK, NOT RGB. It may never be used for any portion of the text of your paper. The archival version of your paper will be printed in black and white and grayscale.The web version must be readable by persons with disabilities. Consequently, because conversion to grayscale can cause undesirable effects (red changes to black, yellow can disappear, and so forth), we strongly suggest you avoid placing color figures in your document. If you do include color figures, you must (1) use the CMYK (not RGB) colorspace and (2) be mindful of readers who may happen to have trouble distinguishing colors. Your paper must be decipherable without using color for distinction.

\subsubsection{Drawings.}
We suggest you use computer drawing software (such as Adobe Illustrator or, (if unavoidable), the drawing tools in Microsoft Word) to create your illustrations. Do not use Microsoft Publisher. These illustrations will look best if all line widths are uniform (half- to two-point in size), and you do not create labels over shaded areas. Shading should be 133 lines per inch if possible. Use Times Roman or Helvetica for all figure call-outs. \textbf{Do not use hairline width lines} --- be sure that the stroke width of all lines is at least .5 pt. Zero point lines will print on a laser printer, but will completely disappear on the high-resolution devices used by our printers.

\subsubsection{Photographs and Images.}
Photographs and other images should be in grayscale (color photographs will not reproduce well; for example, red tones will reproduce as black, yellow may turn to white, and so forth) and set to a minimum of 300 dpi. Do not prescreen images.

\subsubsection{Resizing Graphics.}
Resize your graphics \textbf{before} you include them with LaTeX. You may \textbf{not} use trim or clip options as part of your \textbackslash includegraphics command. Resize the media box of your PDF using a graphics program instead.

\subsubsection{Fonts in Your Illustrations}
You must embed all fonts in your graphics before including them in your LaTeX document.

\subsection{References} 
The AAAI style includes a set of definitions for use in formatting references with BibTeX. These definitions make the bibliography style fairly close to the one specified below. To use these definitions, you also need the BibTeX style file ``aaai.bst," available in the AAAI Author Kit on the AAAI web site. Then, at the end of your paper but before \textbackslash end{document}, you need to put the following lines:

\begin{quote}
\begin{small}
\textbackslash bibliographystyle\{aaai\}
\textbackslash bibliography\{bibfile1,bibfile2,...\}
\end{small}
\end{quote}

Please note that you are required to use \textbackslash bibliographystyle\{aaai\} for your references. You may not use named, plain, apalike, acm, ieeetr, siam, chicago, or any other style. Use of natbib is also not acceptable. (In addition to natbib, the aaai20.sty file is also incompatible with the hyperref and navigator packages. If you use either, your references will be garbled and your paper will be returned to you.) If you used natbib commands, an imprecise workaround is available (although it does not always work). You may put the following in your preamble (after removing \textbackslash usepackage\{natbib\}

\begin{quote}\small
\textbackslash newcommand\{\textbackslash citet\}[1]\{\textbackslash citeauthor\{\#1\}~\textbackslash shortcite\{\#1\}\}
\textbackslash newcommand\{\textbackslash citep\}\{\textbackslash cite\}
\textbackslash newcommand\{\textbackslash citealp\}[1]\{\textbackslash citeauthor\{\#1\}~\textbackslash citeyear\{\#1\}\}
\end{quote}

References may be the same size as surrounding text. However, in this section (only), you may reduce the size to \textbackslash small if your paper exceeds the allowable number of pages. Making it any smaller than 9 point with 10 point linespacing, however, is not allowed. A more precise and exact method of reducing the size of your references minimally is by means of the following command: \begin{quote}
\textbackslash fontsize\{9.8pt\}\{10.8pt\}
\textbackslash selectfont\end{quote}

\noindent You must reduce the size equally for both font size and line spacing, and may not reduce the size beyond \{9.0pt\}\{10.0pt\}.

The list of files in the \textbackslash bibliography command should be the names of your BibTeX source files (that is, the .bib files referenced in your paper).

The following commands are available for your use in citing references:
\begin{quote}
{\em \textbackslash cite:} Cites the given reference(s) with a full citation. This appears as ``(Author Year)'' for one reference, or ``(Author Year; Author Year)'' for multiple references.\smallskip\\
{\em \textbackslash shortcite:} Cites the given reference(s) with just the year. This appears as ``(Year)'' for one reference, or ``(Year; Year)'' for multiple references.\smallskip\\
{\em \textbackslash citeauthor:} Cites the given reference(s) with just the author name(s) and no parentheses.\smallskip\\
{\em \textbackslash citeyear:} Cites the given reference(s) with just the date(s) and no parentheses.
\end{quote}

Formatted bibliographies should look like the following examples.

\smallskip \noindent \textit{Book with Multiple Authors}\\
Engelmore, R., and Morgan, A. eds. 1986. \textit{Blackboard Systems.} Reading, Mass.: Addison-Wesley.

\smallskip \noindent \textit{Journal Article}\\
Robinson, A. L. 1980a. New Ways to Make Microcircuits Smaller. \textit{Science} 208: 1019--1026.

\smallskip \noindent \textit{Magazine Article}\\
Hasling, D. W.; Clancey, W. J.; and Rennels, G. R. 1983. Strategic Explanations in Consultation. \textit{The International Journal of Man-Machine Studies} 20(1): 3--19.

\smallskip \noindent \textit{Proceedings Paper Published by a Society}\\
Clancey, W. J. 1983. Communication, Simulation, and Intelligent Agents: Implications of Personal Intelligent Machines for Medical Education. In \textit{Proceedings of the Eighth International Joint Conference on Artificial Intelligence,} 556--560. Menlo Park, Calif.: International Joint Conferences on Artificial Intelligence, Inc.

\smallskip \noindent \textit{Proceedings Paper Published by a Press or Publisher}\\
Clancey, W. J. 1984. Classification Problem Solving. In \textit{Proceedings of the Fourth National Conference on Artificial Intelligence,} 49--54. Menlo Park, Calif.: AAAI Press.

\smallskip \noindent \textit{University Technical Report}\\
Rice, J. 1986. Poligon: A System for Parallel Problem Solving, Technical Report, KSL-86-19, Dept. of Computer Science, Stanford Univ.

\smallskip \noindent \textit{Dissertation or Thesis}\\
Clancey, W. J. 1979. Transfer of Rule-Based Expertise through a Tutorial Dialogue. Ph.D. diss., Dept. of Computer Science, Stanford Univ., Stanford, Calif.

\smallskip \noindent \textit{Forthcoming Publication}\\
Clancey, W. J. 2021. The Engineering of Qualitative Models. Forthcoming.

For the most up to date version of the AAAI reference style, please consult the \textit{AI Magazine} Author Guidelines at \url{https://aaai.org/ojs/index.php/aimagazine/about/submissions#authorGuidelines}

\section{Proofreading Your PDF}
Please check all the pages of your PDF file. The most commonly forgotten element is the acknowledgements --- especially the correct grant number. Authors also commonly forget to add the metadata to the source, use the wrong reference style file, or don't follow the capitalization rules or comma placement for their author-title information properly. A final common problem is text (expecially equations) that runs into the margin. You will need to fix these common errors before submitting your file.

\section{Improperly Formatted Files }
In the past, AAAI has corrected improperly formatted files submitted by the authors. Unfortunately, this has become an increasingly burdensome expense that we can no longer absorb (we are charged double for papers that require reformatting). Consequently, if your file is improperly formatted, it will probably be returned to you by the outside Production agency. If that happens, you will be required to fix your file and pay a resubmission fee.

\subsection{\LaTeX{} 209 Warning}
If you use \LaTeX{} 209 your paper will be returned to you unpublished. Convert your paper to \LaTeX{}2e.

\section{Naming Your Electronic File}
We require that you name your \LaTeX{} source file with the last name (family name) of the first author so that it can easily be differentiated from other submissions. Complete file-naming instructions will be provided to you in the submission instructions.

\section{Submitting Your Electronic Files to AAAI}
Instructions on paper submittal will be provided to you in your acceptance letter.

\section{Inquiries} 
If you have any questions about the preparation or submission of your paper as instructed in this document, please contact AAAI Press at the address given below. If you have technical questions about implementation of the aaai style file, please contact an expert at your site. We do not provide technical support for \LaTeX{} or any other software package. To avoid problems, please keep your paper simple, and do not incorporate complicated macros and style files.

\begin{quote}
\noindent AAAI Press\\
2275 East Bayshore Road, Suite 160\\
Palo Alto, California 94303\\ 
\textit{Telephone:} (650) 328-3123\\ 
\textit{E-mail:} See the submission instructions for your particular conference or event.
\end{quote}

\section{Additional Resources}
\LaTeX{} is a difficult program to master. If you've used that software, and this document didn't help or some items were not explained clearly, we recommend you read Michael Shell's excellent document (testflow doc.txt V1.0a 2002/08/13) about obtaining correct PS/PDF output on \LaTeX{} systems. (It was written for another purpose, but it has general application as well). It is available at www.ctan.org in the tex-archive.

\section{ Acknowledgments}
AAAI is especially grateful to Peter Patel Schneider for his work in implementing the aaai.sty file, liberally using the ideas of other style hackers, including Barbara Beeton. We also acknowledge with thanks the work of George Ferguson for his guide to using the style and BibTeX files --- which has been incorporated into this document --- and Hans Guesgen, who provided several timely modifications, as well as the many others who have, from time to time, sent in suggestions on improvements to the AAAI style.

The preparation of the \LaTeX{} and Bib\TeX{} files that implement these instructions was supported by Schlumberger Palo Alto Research, AT\&T Bell Laboratories, Morgan Kaufmann Publishers, The Live Oak Press, LLC, and AAAI Press. Bibliography style changes were added by Sunil Issar. \verb+\+pubnote was added by J. Scott Penberthy. George Ferguson added support for printing the AAAI copyright slug. Additional changes to aaai.sty and aaai.bst have been made by the AAAI staff.

\bigskip
\noindent Thank you for reading these instructions carefully. We look forward to receiving your electronic files!

\end{document}

### 2020-07-30 paper/ch_potentials.tex
fixed bib
foo

@@
 
-\setcounter{secnumdepth}{1} %May be changed to 1 or 2 if section numbers are desired.
+\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.
 
@@
 Efficiently computing flexible and optimal routes is a thought challenge.
-Optimal means that the computed path always has minimum weight.
-Efficiently means that the path computation is fast.
-Being able to change the edge weights between path computations is flexibility.
+\emph{Optimal} means that the computed path always has minimum weight.
+\emph{Efficiently} means that the path computation is fast.
+Being able to change the edge weights between path computations is \emph{flexibility}.
 
-Dijkstra's algorithm \cite{d-ntpcg-59} achieves flexiblity and optimality but unfortunately rrequires several seconds to compute paths across continental road networks, which is too slow for many applications.
-A classic speedup approach is the well-known A* algorithm \cite{hnr-afbhd-68}.
-While flexible, A*'s performance crucially depends on the quality of the potential, also called heuristic, function.
+\begin{figure}
+
+\centering
+\includegraphics[width=\columnwidth,height=200px]{fig/searchspace}
+
+\caption{Example A* search space with CH-Potentials TODO: what parametrization?}
+\label{img:search-space}
+\end{figure}
+
+Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexiblity and optimality but is comparatively slow.
+A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
+While flexible, A*'s performance crucially depends on the quality of the heuristic, also called potential, function.
 Finding such a function that can be computed efficiently is difficult.
-A lot of research of the last decade has therefore focused on mostly hierarchical approaches \cite{gssv-erlrn-12,dsw-cch-15,dgpw-crprn-13,ss-ehh-12} which do not require such a function.
-A popular hierarchical algorithm is Contraction Hierarchy (CH) \cite{gssv-erlrn-12}.
+
+In this paper, we consider a setup with a preprocessing phase and a query phase.
+The preprocessing phase can be slow and computes auxilary data that can be used in the query phase that should be fast.
+The input to the preprocessing phase is the graph weighted with lower bound weights $w_\ell$.
+The source/origin and target/goal of the path and the actual edge weights $w_q$ are the input to the query phase.
+We require that $w_q(e)\ge w_\ell(e)$ for all edges $e$.
+In the query phase, the path is computed.
+
+Street networks of continental scale has many millions of nodes.
+While the preprocessing phase may be slow, it must eventually finish.
+On graphs of this size, quadratic preprocessing time is thus prohibitive.
+
+In this paper, we consider an algorithm, where the query phase consists of an A* search.
+The preprocessing phase computes auxilary data to efficiently evaluate the heuristic/potential.
+Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
+Following, their terminology, we refer to the heuristic as potential function $\pi$.
+Further, paths are computed from a source node $s$ to a target node $t$.
+
+We say that a potential $\pi$ is \emph{perfect}, if no tighter potential can be computed in the preprocessing phase.
+The shortest path distances with respect to $w_\ell$ are perfect.
+CH-Potentials compute $w_\ell$ shortest paths and are thus perfect.
+
+\section{Related Work}
+
+A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
+In this setting, mostly hierarchical approaches dominate in termes of running time~\cite{dgpw-crprn-13,ss-ehh-12}. 
+A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12,dsw-cch-15}.
+
 There is a lot of work that extends CH to handle more complex settings.
@@
 Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
+Combining all these hierarchical extensions is an unsolved problem.
 
 While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
-In this paper, we explore CH-Potentials, a novel combination of A* and the hierarchical techniques CH and CCH.
-We are not the first to combine hierarchical approaches and A* \cite{bdsssw-chgds-10, gkw-blwr-07, bdgwz-sfpcs-19}.
+In this paper, we explore CH-Potentials, a novel combination of A* and the hierarchical techniques CH using some ideas from \cite{dgnw-phast-13}.
+We are not the first to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
 However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
@@
 
-Our algorithm is built upon ideas from Phast \cite{dgnw-phast-13}, a CH-extension.
-The novel idea is to compute optimal A* potentials lazily.
-The hierarchical approach is applied to the lower bound graph with respect to the edge costs.
-As the hierarchical approaches are optimal, the resulting A* potential is the tightest possible.
-With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
-Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
+The most similar approaches to CH-Potentials are ALT and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19}.
+Just as our approach, ALT is A*-based.
+However, ALT's potentials are not perfect. 
+%
+CPD-Heursitics are a combination of A* and Compressed Path Databases (CPD). 
+A CPD can quickly compute the first edge of a shortest path between any two nodes.
+In \cite{DBLP:conf/ijcai/BonoGHS19}, a CPD~\cite{DBLP:conf/socs/StrasserHB14} is used to create a potential.
+In every potential evaluation, a path to the target is compute, whose length is used as potential.
+The resulting potential is perfect.
+Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
+%
+MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
+%
+In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
+The Euclidean distance is then used as A* heuristic.
+%
+The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.
 
-\begin{figure}
-TODO
+\subsection{Oracle-A* \& Comparison with Related Work}
 
-\caption{Example search space when TODO: what parametrization?}
-\label{img:search-space}
-\end{figure}
+To compare against related work, we introduce the concept of Oracle-A*.
+It has access to a perfect shortest distance array as potential.
+The array computation time is unaccounted for, i.e., it is magically filled instantly.
+Oracle-A* is not achievable in practice.
+However, is a lower bound for every algorithm following out setup.
+We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.
 
@@
 
-Modelling navigation as shortest path problem in graphs with scalar weights is a good but not perfect formalization.
-Several important effects are ignored.
-In this section, we list a few to motivate why flexibility is important.
-In Section~\ref{sec:extensions}, we describe how to handle some with CH-Potentials in detail.
-Our argument is a variation of the one made by~\cite{CRP,CCH,DBLP:conf/gis/FunkeS15}.
+Many applications need flexibility.
+In road routing, the most common ones are traffic avoidance, user preferences, and vehicle restrictions.
+Traffic avoidance comes in two forms: Live~\cite{CRP,CCH} and predicted traffic~\cite{td-alt,td-ch}.
+Predicted traffic is often called time-dependent routing.
+%
+User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}
+For example, for some users shorter routes are more important fast routes.
+Others want to avoid tunnels or highways.
+%
+Truck routing differs from car routing because, depending on their load, trucks are not allowed everywhere~\cite{kswz-erptd-20}.
+%
+Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
 
-The travel time across an edge is in general not constant over time.
-Live and predicted traffic are examples of how the travel time can change.
-Another example are road blockages.
-These are usually modelled by setting travel times to infinity.
-Some blockage types, such as accidents, are similar to live traffic.
-Other blockages such as reversible one-ways, time-dependent turn-restrictions, or seasonal closures are more similar to predicted traffic.
+\section{Contribution \& Outline}
 
-Flexibility is also necessary to account for driver preferences.
-For example, common options in navigation devices include avoiding highways or tunnels. 
-These requirements are usally modelled by modifying the edge weights and are highly depend on the driver.
-Sometimes they even depend on his current mood and might change while driving.
-In this paper, we assume that the driver preference is known.
-We refer to \cite{DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16} for approaches to learn these requirements.
-
-Trucks are another class of vehicles with diverse and specific needs.
-For example, trucks are not allowed in some areas depending on the time of the day.
-Some truck carry hazardous materials and are therefore not allowed in some areas because of environmental protection laws.
-Once the truck has delivered his goods, he may again pass the area.
-Flexibility can also be used to limit the slope and speed of the roads used during routing.
-
-\subsection{Problem Setting}
-
-As detailed in the previous section, the reasons why flexibility is a necessary property of a routing engine are sheer endless.
-However, a formal problem specification that encompasses all these options is complex and well beyond the scope of this paper.
-For an example of such a formulation, we refer to \cite{kswz-erptd-20} which covers a truck-specific application.
-The algorithm presented in \cite{kswz-erptd-20} uses a preliminary version of CH-Potentials.
-For simplicity, we limit our exposition to the classical shortest path formulation.
-To support algorithm extensions, such as time-dependent routing as descibed in Section \ref{sec:predicted-traffic}, we avoid two operations that are adverse to generalization.
-These are:
-(1) traversing an input edge in backward direction, and
-(2) combining input weights.
-Backward traversal is problematic as a weight can depend on the moment that an edge is entered, which is unknown in a backward search.
-Combining input weights can be difficult, if the input weights are more complex objects than scalar values.
-For example, when considering historic traffic patterns, the input weights are usually functions \cite{citation needed}.
-
-Our algorithm has one major restriction, it requires the presence of an inflexible scalar weight lower bound for every input edge.
-The performance of our algorithm depends on this lower bound being close to the actual weight.
-Using the trivial lower bound of zero everywhere, CH-Potentials degenerate to a variant of Dijkstra's algorithm.
-
-Throughout this paper, we assume that we are given a directed graph $G$ with nodes $V$ and edges $E$.
-Every edge $(x,y)\in E$ has a \emph{tail} node $x$ and a \emph{head} node $y$.
-Further, every edge has a lower bound weight $w_\ell:E\rightarrow \mathbb{R}^+$ and a query weight $w_q:E\rightarrow \mathbb{R}^+ \cup \{+\infty\}$.
-It must hold that $0\le w_\ell(e)\le w_q(e)$.
-We are further given a source node $s$ and a target node $t$.
-
-Our algorithm uses two phases.
-The first is a slow \emph{preprocessing phase}, where only $G$ and $w_\ell$ are known.
-It computes auxilary data.
-The second phase is the \emph{query phase}.
-It is given the auxilary data and $G$, $w_\ell$, $w_q$, $s$, and $t$.
-The output of the query phase, is a path $e_1,e_2\ldots e_k$ such that $\sum w_q(e_i)$ is minimized.
-
-\subsection{Related Work}
-
-The closest CH-Potential related works are ALT~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07} and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19}.
-Just as our approach, ALT is A*-based.
-ALT computes the potential using landmarks.
-The computed potential values are not optimal and thus of lower quality than the optimal ones computed by CH-Potentials.
-ALT is therefore slower.
-
-CPD-Heursitics make use of compressed path databases (CPD). 
-A CPD can quickly compute the first edge of a shortest path between any two nodes.
-In \cite{DBLP:conf/ijcai/BonoGHS19}, a CPD~\cite{DBLP:conf/socs/StrasserHB14} is used to create a potential.
-In every potential evaluation, a path to the target is compute, whose length is used as potential.
-On a conceptual level, this is very similar to CH-Potentials. 
-Fortunately, CH-Potentials do not have to compute a path per evaluation, which makes CH-Potentials faster.
-Further, the CPD used has quadratic preprocessing running times.
-A CH does not.
-This makes CPDs too expensive on large continental road networks.
-The computed potential values are also optimal.
-
-As ALT and CPD-Heursitics have a similar structure than CH-Potentials, the extensions described in Section~\ref{sec:extensions} can also be applied to them.
-For example, in~\cite{td-alt} it is extended to time-dependent routing.
-
-The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.
-
-TODO: cite \cite{DBLP:journals/tciaig/BaierBHH15} https://ieeexplore.ieee.org/document/6851877
-
-TODO: cite \cite{DBLP:journals/ai/SharonSFS15} https://www.aaai.org/ocs/index.php/AAAI/AAAI12/paper/viewFile/5062/5239
-
-TODO: cite \cite{DBLP:conf/ijcai/0002UJAKK18} https://arxiv.org/abs/1706.02792
+We introduce CH-Potentials in Section~\ref{sec:main-algo}. 
+A two-phase technique to efficiently compute perfect potentials.
+Combined with A*, this yields an optimal, efficient, and flexibile route planning algorithms.
+Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
+Beside the potentials, we describe, in Section\ref{sec:low-deg-improvment}, some improvments to A* to more efficiently process low degree nodes, common in road networks.
+In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and showing how they can be solved using CH-Potentials.
+Finally in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach. 
+By comparing against Oracle-A*, we show that no alternative approach can be significantly faster.
 
@@
 
-A* star is a well-known extension to Dijkstra's algorithm.
-It uses a potential function $\pi:V\rightarrow \mathbb{R}^+$.
-The potential is often called heuristic.
-A* is optimal, if the potential $\pi(v)$ is consistent \cite{p-hissc-84}.
-In this paper, all considered potentials are derived from exact shortest path distances and are thus consistent.
-
-The best possible potential that can be derived from the auxilary data is the minimum path distance with respect to $w_\ell$.
-Every better, consistent potential requires knowledge of the query weights $w_q$, which is unavailable during preprocessing.
-Our technique allows to compute these best possible potentials.
-Within this regard, CH-Potentials are \emph{optimal A* potentials}.
-
-\subsection{Contraction Hierarchy (CH)}
+\label{sec:main-algo}
 
 \begin{algorithm2e}
-\KwData{$B[x]$ tentative distance from $x$ to $t$}
+\KwData{$B[x]$: tentative distance from $x$ to target $t$}
 \KwData{Min. priority queue $Q$, also called open list}
-$B[x] \leftarrow +\infty$ for all $x\neq t$\;
+$B[x] \leftarrow +\infty$ for all $x\neq t$;
 $B[t] \leftarrow 0$\;
@@
 			$B[x]\leftarrow w_\ell(xy) + B[y]$\;
-			\eIf{not $x$ in $Q$}{
-				Add $x$ to $Q$ with weight $w_\ell(xy) + B[y]$\;
-			}{
-				Decrease weight of $x$ in $Q$ to $w_\ell(xy) + B[y]$\;
-			}
+                        Add $x$ or decrease $x$'s key in $Q$ to $B[x]$\;
 		}
@@
 
-A CH is a two phase technique to efficiently compute optimal, shortest paths.
-For a detailed exposition, we refer the interested reader to \cite{dsw-cch-15,gssv-erlrn-12}.
-In this section, we give a high-level overview of the aspects relevant to this paper.
+\begin{figure}
+\centering
+\includegraphics{fig/ch}
+\caption{
+Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is mid node.
+}
+\label{fig:ch}
+\end{figure}
+
+A Contraction Hierarchy (CH) is a two phase technique to efficiently compute optimal, shortest paths.
+For a detailed exposition, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
+In this section, we give a high-level overview.
 
@@
 Every up and down path is an up-down path.
-
+%
 In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
+This is done by repeatedly contracting unimportant nodes. 
+See \cite{gssv-erlrn-12} for the details.
 The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
+See Figure~\ref{fig:ch} for an illustration.
 We may therefore restrict our search to up-down paths in $G^+$.
@@
 The two searches meet in the mid node.
-Pseudo-code for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
+Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
 The forward search works analogously.
-
+%
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
@@
 
-Using a CH, we can easily compute optimal A* potentials.
+Using a CH, we can easily compute perfect A* potentials.
 In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
 The query phase consists of an A* search, where the evaluation of the potential $\pi(v)$ performs a CH-query from $v$ to $t$.
-
+%
 This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
 Fortunatly, we can do better.
-However, for this we need another component called Phast \cite{dgnw-phast-13}.
+However, for this we need another component called PHAST \cite{dgnw-phast-13}.
 
@@
 \begin{algorithm2e}
-\KwData{$P[x]$ tentative distance from $x$ to $t$}
+\KwData{$P[x]$: tentative distance from $x$ to $t$}
 Execute Algorithm~\ref{algo:ch-backward}\;
@@
 }
-\caption{Phast basic all-to-one search}
+\caption{PHAST basic all-to-one search}
 \label{algo:phast}
@@
 
-Phast \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
+PHAST \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
 First the preprocessing phase is executed analogously to the original CH.
@@
 After all level are processed, the distances from all nodes towards $t$ is computed.
-Phast is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
-Phast's main advantage is that is easily be parallelized.
+PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
+PHAST's main advantage is that is easily be parallelized.
 However, we will not consider parallelization in this paper.
@@
 
-Using Phast, we can compute optimal A* potentials in another way.
-In the query phase, we first use Phast to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
+Using PHAST, we can compute optimal A* potentials in another way.
+In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
 In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.
@@
 The potential evaluation and by extension the $A^*$ search is indeed fast.
-However, the Phast step before the search is comparatively expensive.
+However, the PHAST step before the search is comparatively expensive.
 The reason for this is that the distance from all nodes towards $t$ are computed.
@@
 \begin{algorithm2e}
-\KwData{$B[x]$ tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$P[x]$ memoized potential at $x$, $\bot$ if uncomputed}
-\SetKwFunction{Pot}{Potential}
+\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$P[x]$: memoized potential at $x$, $\bot$ initially}
+\SetKwFunction{Pot}{Pot}
 \SetKwProg{Fn}{Function}{:}{}
@@
 		\For{all up edges $xy$ in $G^+$}{
-			\If{$P[x] > w_\ell(xy)+\Pot(y)$}{
-				$P[x] \leftarrow w_\ell(xy)+\Pot(y)$\;
-			}
+                        $P[x]\leftarrow\min\{P[x],w_\ell(xy)+\Pot(y)\}$\;
 		}
@@
 
-Fortunately, the Phast computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
+Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
 
@@
 
-\subsection{Skip Degree Two Nodes}
+\section{Low Degree A* Improvments}
+
+\label{sec:low-deg-improvment}
 
@@
 
-\begin{figure}
+%\begin{figure}
 
-\begin{center}
-\includegraphics{push-deg3}
-\end{center}
+%\begin{center}
+%\includegraphics{push-deg3}
+%\end{center}
 
-\caption{$x$ is removed from the queue. $y_i$, $z$, $a_i$, and $\alpha_i$ are never part of the queue. $\beta$ and $b$ are pushed into the queue.}
-\label{fig:push-deg3}
+%\caption{$x$ is removed from the queue. $y_i$, $z$, $a_i$, and $\alpha_i$ are never part of the queue. $\beta$ and $b$ are pushed into the queue.}
+%\label{fig:push-deg3}
 
-\end{figure}
+%\end{figure}
 
@@
 If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
-The situation is depicted in Figure~\ref{fig:push-deg3}.
+%The situation is depicted in Figure~\ref{fig:push-deg3}.
 
@@
 
-\bibliography{dblp}
-\bibliography{references}
+\pagebreak
+
+\bibliography{dblp,references}
 \bibliographystyle{aaai}

### 2020-07-30 paper/ch_potentials.tex
foo

@@
 
-\caption{Example A* search space with CH-Potentials TODO: what parametrization?}
+\caption{Example A* search space with CH-Potentials}
 \label{img:search-space}
@@
 
+In this section, we first describe Contraction Hierarchies, then PHAST, and finally CH-Potentials.
+
 \label{sec:main-algo}
@@
 
-A Contraction Hierarchy (CH) is a two phase technique to efficiently compute optimal, shortest paths.
-For a detailed exposition, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
-In this section, we give a high-level overview.
+\subsection{Contraction Hierarchy (CH)}
+
+A CH is a two phase technique to efficiently compute optimal, shortest paths.
+It is not flexibile, i.e., $w_q=w_\ell$.
+For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
+In this section, we give an overview.
 
@@
 Levels are ordered by ``importance''.
-The intuition is that dead-ends are unimportant and low while highway bridges are very important and high.
+The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
 An edge goes \emph{up} when it goes from a node in a lower level to higher level.
@@
 In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
-This is done by repeatedly contracting unimportant nodes. 
-See \cite{gssv-erlrn-12} for the details.
+This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
+See~\cite{gssv-erlrn-12} for the details.
 The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
-See Figure~\ref{fig:ch} for an illustration.
+See Figure~\ref{fig:ch} for a proof scetch.
+From every shortest path (red) in $G$, a up-down path of equal length in $G+$ (blue) can be constructed.
 We may therefore restrict our search to up-down paths in $G^+$.
@@
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-
 Using PHAST, we can compute optimal A* potentials in another way.
@@
 
-This setup also works.
+This PHAST-based algorithm works.
 The potential evaluation and by extension the $A^*$ search is indeed fast.
@@
 Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
-
 In a first step, we run the backward CH search from $t$ to obtain an array $B$.
@@
 By avoiding push operations, we can therefore avoid potential evaluations.
-This is the objective of this and the next section.
+This is the objective of this section.
+
+\subsection{Skip Degree Two Nodes}
 
@@
 
-%\begin{figure}
-
-%\begin{center}
-%\includegraphics{push-deg3}
-%\end{center}
-
-%\caption{$x$ is removed from the queue. $y_i$, $z$, $a_i$, and $\alpha_i$ are never part of the queue. $\beta$ and $b$ are pushed into the queue.}
-%\label{fig:push-deg3}
-
-%\end{figure}
-
 In the previous section, we described an optimized A* variant that does not push degree two nodes.
@@
 If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
-%The situation is depicted in Figure~\ref{fig:push-deg3}.
 
-\subsection{Skip Nodes Outside of Largest Biconnected Component}
+\subsection{Stay in Largest Biconnected Component}
 
@@
 
-\section{Extensions}
+\section{Algorithm Extensions}
 \label{sec:extensions}
@@
 
-\subsection{Forbidden Turns}
+\subsection{Forbidden Turns and Turn Costs}
 \label{sec:no-turns}
@@
 
-\subsection{Predicted Traffic}
+\subsection{Predicted Traffic or Time-Dependent Routing}
 \label{sec:predicted-traffic}

### 2020-08-05 Rakefile.rb
fix osm import

@@ -77,7 +77,7 @@ namespace "prep" do
   file osm_ger => ["code/osm_import/build/import_osm", osm_ger_src_file] do
     wd = Dir.pwd
     Dir.chdir osm_ger do
-      sh "#{wd}/code/osm_import/build/import_osm /algoDaten/zeitz/roadgraphs/germany-latest.osm.pbf #{Dir[live_dir + '*'].join(' ')} #{typical_file}"
+      sh "#{wd}/code/osm_import/build/import_osm #{osm_ger_src_file} #{Dir[live_dir + '*'].join(' ')} #{typical_file}"
     end
   end
 
@@ -94,7 +94,7 @@ namespace "prep" do
   file osm_europe => ["code/osm_import/build/import_osm", osm_eur_src_file] do
     wd = Dir.pwd
     Dir.chdir osm_europe do
-      sh "#{wd}/code/osm_import/build/import_osm /algoDaten/zeitz/roadgraphs/europe-latest.osm.pbf"
+      sh "#{wd}/code/osm_import/build/import_osm #{osm_eur_src_file}"
     end
   end
 

### 2020-08-05 Rakefile.rb
fix var name

@@ -38,7 +38,7 @@ end
 
 osm_ger_src = 'https://download.geofabrik.de/europe/germany-200101.osm.pbf'
 osm_ger_src_file = '/algoDaten/zeitz/roadgraphs/germany-200101.osm.pbf'
-osgm_eur_src = 'https://download.geofabrik.de/europe-200101.osm.pbf'
+osm_eur_src = 'https://download.geofabrik.de/europe-200101.osm.pbf'
 osm_eur_src_file = '/algoDaten/zeitz/roadgraphs/europe-200101.osm.pbf'
 
 osm_ger = "/algoDaten/zeitz/roadgraphs/osm_ger/"

### 2020-08-05 Rakefile.rb
more rake fixes

@@ -91,6 +91,7 @@ namespace "prep" do
     end
   end
 
+  directory osm_europe
   file osm_europe => ["code/osm_import/build/import_osm", osm_eur_src_file] do
     wd = Dir.pwd
     Dir.chdir osm_europe do
@@ -115,7 +116,7 @@ namespace "prep" do
   end
 
   td_graphs.each do |graph|
-    file graph + "lower_bound" do
+    file graph + "lower_bound" => graph do
       Dir.chdir "code/bmw_routing_engine/engine" do
         sh "cargo run --release --bin td_lower_bound -- #{graph}"
       end

### 2020-08-05 eval/building_blocks.py
start updating eval scripts

@@ -22,17 +22,17 @@ table['num_queue_pops'] = table['num_queue_pops'] / 1000
 table = table.round(1)
 
 table = table.rename(index={
-    'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_NO_SCC, CHPOT_ONLY_TOPO, ': 'No Deg2, No Deg3, No SCC, No Pot.',
+    'CHPOT_NO_BCC, CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ': 'No Deg2, No Deg3, No BCC, No Pot.',
     'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ':               'No Deg2, No Deg3, No Pot.',
     'CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ':                              'No Deg3, No Pot.',
     'CHPOT_ONLY_TOPO, ':                                             'No Pot.',
-    'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_NO_SCC, ':                  'No Deg2, No Deg3, No SCC',
+    'CHPOT_NO_BCC, CHPOT_NO_DEG2, CHPOT_NO_DEG3, ':                  'No Deg2, No Deg3, No BCC',
     'CHPOT_NO_DEG2, CHPOT_NO_DEG3, ':                                'No Deg2, No Deg3',
     'CHPOT_NO_DEG3, ':                                               'No Deg3',
 })
 
-table = table.reindex(['No Deg2, No Deg3, No SCC, No Pot.', 'No Deg2, No Deg3, No Pot.', 'No Deg3, No Pot.', 'No Pot.',
-    'No Deg2, No Deg3, No SCC', 'No Deg2, No Deg3', 'No Deg3', ''])
+table = table.reindex(['No Deg2, No Deg3, No BCC, No Pot.', 'No Deg2, No Deg3, No Pot.', 'No Deg3, No Pot.', 'No Pot.',
+    'No Deg2, No Deg3, No BCC', 'No Deg2, No Deg3', 'No Deg3', ''])
 
 lines = table.to_latex(escape=False).split("\n")
 

### 2020-08-05 eval/scaled_weights.py

@@ -18,7 +18,7 @@ data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
     'weight_factor': exp['factor'],
-    'graph': { '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe', '/algoDaten/zeitz/roadgraphs/osm_ger_rel/': 'OSM Germany' }[run['args'][1]],
+    'graph': { '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe', '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Germany', '/algoDaten/zeitz/roadgraphs/osm_eur/': 'OSM Europe' }[run['args'][1]],
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 

### 2020-08-05 paper/ch_potentials.tex
foo

@@
 To compare against related work, we introduce the concept of Oracle-A*.
-It has access to a perfect shortest distance array as potential.
+It has access to a perfect shortest distance array with respect to $w_\ell$ as potential.
+This is the tightest potential achievable during preprocessing. 
 The array computation time is unaccounted for, i.e., it is magically filled instantly.
 Oracle-A* is not achievable in practice.
-However, is a lower bound for every algorithm following out setup.
+However, is a lower bound for every algorithm following our setup.
 We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.

### 2020-08-05 paper/ch_potentials.tex
foo

@@
 \pdfinfo{
-/Title (A* with Perfect Heuristic/Potentials in Road Networks)
+/Title (A* with Fast and Tight Heuristic/Potentials in Road Networks)
 /Author (Ben Strasser, Tim Zeitz)
@@
 }
-\title{A* with Perfect Heuristic/Potentials in Road Networks}
+\title{A* with Fast and Tight Heuristic/Potentials in Road Networks}
 \author{Ben Strasser\\
@@
 However, A*'s performance depends on the availability of a good heuristic, also called potential, function.
-Computing a good heuristic/potential is a challenge of its own.
-To avoid the need for a heuristic/potential, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
-These techniques are not based on A* and therefore do not need a heuristic/potential.
+Computing a good heuristic function is a challenge of its own.
+To avoid the need for a heuristic function, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
+These techniques are not based on A* and therefore do not need a heuristic function.
 They achieve speed and optimality but sacrifize A*'s flexibility.
-In this paper, we use hierarchical techniques to compute heuristic/potential functions.
+In this paper, we use hierarchical techniques to compute heuristic functions.
 We call our technique CH-Potential.
-Additionally, we describe some A* optimizations to accelerate the processing of low degree nodes, which occur in road networks.
+Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
 With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
@@
 Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-Following, their terminology, we refer to the heuristic as potential function $\pi$.
+Following, their terminology, we refer to the heuristic as \emph{potential} function $\pi$.
 Further, paths are computed from a source node $s$ to a target node $t$.
 
-We say that a potential $\pi$ is \emph{perfect}, if no tighter potential can be computed in the preprocessing phase.
-The shortest path distances with respect to $w_\ell$ are perfect.
-CH-Potentials compute $w_\ell$ shortest paths and are thus perfect.
+We say that a potential $\pi$ is \emph{tight}, if no tighter potential can be computed in the preprocessing phase.
+The shortest path distances with respect to $w_\ell$ are tight.
+CH-Potentials compute $w_\ell$ shortest paths and are thus tight.
 
@@
 Just as our approach, ALT is A*-based.
-However, ALT's potentials are not perfect. 
+However, ALT's potentials are not tight. 
 %
@@
 In every potential evaluation, a path to the target is compute, whose length is used as potential.
-The resulting potential is perfect.
+The resulting potential is tight.
 Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
@@
 To compare against related work, we introduce the concept of Oracle-A*.
-It has access to a perfect shortest distance array with respect to $w_\ell$ as potential.
+It has access to a shortest distance array with respect to $w_\ell$ as potential, i.e., it is tight.
 This is the tightest potential achievable during preprocessing. 
@@
 We introduce CH-Potentials in Section~\ref{sec:main-algo}. 
-A two-phase technique to efficiently compute perfect potentials.
+A two-phase technique to efficiently compute tight potentials.
 Combined with A*, this yields an optimal, efficient, and flexibile route planning algorithms.
@@
 
-Using a CH, we can easily compute perfect A* potentials.
+Using a CH, we can easily compute tight A* potentials.
 In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
@@
 We introduce a novel way to compute potentials for A*.
-The potentials are calculated utilizing a CH on a lower bound graph and perfect with regards to that lower bound graph.
+The potentials are calculated utilizing a CH on a lower bound graph and tight with regards to that lower bound graph.
 With these potentials we can speed up A* search for extended route planning problems like route planning with live traffic, time-dependent travel time predictions or user preferences.

### 2020-08-05 Rakefile.rb
new routing engine repo

@@ -60,7 +60,7 @@ typical_file = '/algoDaten/mapbox/typical-tuesday-cleaned.csv'
 
 namespace "prep" do
   file typical_file do
-    Dir.chdir "code/bmw_routing_engine/engine" do
+    Dir.chdir "code/rust_road_router/engine" do
       sh "cat #{typical_glob} | cargo run --release --bin scrub_td_mapbox -- 576 864 > #{typical_file}"
     end
   end
@@ -86,7 +86,7 @@ namespace "prep" do
     ['first_out', 'head', 'travel_time', 'geo_distance', 'osm_node_ids', 'arc_category', 'forbidden_turn_from_arc', 'forbidden_turn_to_arc', 'latitude', 'longitude', 'tail'].each do |file|
       sh "ln -s #{osm_ger}/#{file} #{osm_ger_td}/#{file}"
     end
-    Dir.chdir "code/bmw_routing_engine/engine" do
+    Dir.chdir "code/rust_road_router/engine" do
       sh "cargo run --release --bin import_td_mapbox -- #{osm_ger_td} #{typical_file}"
     end
   end
@@ -117,7 +117,7 @@ namespace "prep" do
 
   td_graphs.each do |graph|
     file graph + "lower_bound" => graph do
-      Dir.chdir "code/bmw_routing_engine/engine" do
+      Dir.chdir "code/rust_road_router/engine" do
         sh "cargo run --release --bin td_lower_bound -- #{graph}"
       end
     end
@@ -150,7 +150,7 @@ namespace "exp" do
   end
 
   task scaled_weights: [dimacs_graph + 'lower_bound_ch', osm_ger + 'lower_bound_ch', osm_europe + 'lower_bound_ch'] + ["#{exp_dir}/scaled_weights"] do
-    Dir.chdir "code/bmw_routing_engine/engine" do
+    Dir.chdir "code/rust_road_router/engine" do
       sh "cargo run --release --bin chpot_weight_scaling -- #{dimacs_graph} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_weight_scaling -- #{osm_ger} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_weight_scaling -- #{osm_europe} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
[…]

### 2020-08-05 paper/ch_potentials.tex
foo

@@
 \pdfinfo{
-/Title (A* with Fast and Tight Heuristic/Potentials in Road Networks)
+/Title (A* with Fast and Tight Heuristic in Road Networks)
 /Author (Ben Strasser, Tim Zeitz)
@@
 }
-\title{A* with Fast and Tight Heuristic/Potentials in Road Networks}
+\title{A* with Fast and Tight Heuristic in Road Networks}
 \author{Ben Strasser\\
@@
 \begin{abstract}
-Efficiently computing flexible and optimal routes in continental-sized road networks is a tough challenge.
+Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
 A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
-However, A*'s performance depends on the availability of a good heuristic, also called potential, function.
-Computing a good heuristic function is a challenge of its own.
-To avoid the need for a heuristic function, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
-These techniques are not based on A* and therefore do not need a heuristic function.
-They achieve speed and optimality but sacrifize A*'s flexibility.
-In this paper, we use hierarchical techniques to compute heuristic functions.
+However, A*'s performance depends on the availability of a good heuristic.
+Computing a good heuristic is a challenge of its own.
+To avoid the need for a heuristic, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
+These techniques are not based on A* and therefore do not need a heuristic.
+They achieve speed and exactness but sacrifize A*'s flexibility.
+In this paper, we use hierarchical techniques to compute tight A* heuristics.
 We call our technique CH-Potential.
 Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
-With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining optimality.
+With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining exactness.
 \end{abstract}
@@
 It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
-Efficiently computing flexible and optimal routes is a thought challenge.
-\emph{Optimal} means that the computed path always has minimum weight.
+Efficiently computing flexible and exact routes is a thought challenge.
+\emph{Exact} means that the computed path always has minimum weight.
 \emph{Efficiently} means that the path computation is fast.
@@
 
-Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexiblity and optimality but is comparatively slow.
+Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexiblity and exactness but is comparatively slow.
 A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
-While flexible, A*'s performance crucially depends on the quality of the heuristic, also called potential, function.
-Finding such a function that can be computed efficiently is difficult.
+While flexible, A*'s performance crucially depends on the quality of the employed heuristic.
+Finding a good heuristic that can be evaluated efficiently is difficult.
 
 In this paper, we consider a setup with a preprocessing phase and a query phase.
-The preprocessing phase can be slow and computes auxilary data that can be used in the query phase that should be fast.
+The preprocessing phase can be slow and computes auxilary data that can be used in the query phase, which should be fast.
 The input to the preprocessing phase is the graph weighted with lower bound weights $w_\ell$.
-The source/origin and target/goal of the path and the actual edge weights $w_q$ are the input to the query phase.
+The source/origin $s$ and target/goal $t$ of the path and the actual edge weights $w_q$ are the input to the query phase.
 We require that $w_q(e)\ge w_\ell(e)$ for all edges $e$.
 In the query phase, the path is computed.
-
-Street networks of continental scale has many millions of nodes.
+%
+Street networks of continental scale have many millions of nodes.
 While the preprocessing phase may be slow, it must eventually finish.
@@
 
-In this paper, we consider an algorithm, where the query phase consists of an A* search.
-The preprocessing phase computes auxilary data to efficiently evaluate the heuristic/potential.
+In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
+The preprocessing phase computes auxilary data to efficiently evaluate the heuristic.
 Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-Following, their terminology, we refer to the heuristic as \emph{potential} function $\pi$.
-Further, paths are computed from a source node $s$ to a target node $t$.
+We call our algorithm CH-Potentials, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, which are called potentials in~\cite{gh-cspas-05}.
 
-We say that a potential $\pi$ is \emph{tight}, if no tighter potential can be computed in the preprocessing phase.
-The shortest path distances with respect to $w_\ell$ are tight.
-CH-Potentials compute $w_\ell$ shortest paths and are thus tight.
+Formally, a heuristic $h$ is a function that maps nodes onto numbers.
+$h$ depends on the target $t$ of the search but not one the source $s$.
+Without loose of generality, we assume that $h(t)=0$.
+The \emph{tight} heuristic assigns to every node $v$, the length of a shortest $vt$-path.
+All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
+CH-Potentials compute the tight heursitc with respect to $w_\ell$.
+Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
+CH-Potentials are optimal in this sense.
 
@@
 A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
-In this setting, mostly hierarchical approaches dominate in termes of running time~\cite{dgpw-crprn-13,ss-ehh-12}. 
+In this setting, mostly hierarchical approaches dominate in termes of running time~\cite{bdgmpsww-rptn-16}. 
 A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12,dsw-cch-15}.
@@
 Combining all these hierarchical extensions is an unsolved problem.
-
-While undoubtedly fast, hierarchical approaches tend to sacrifice flexibility to obtain speed.
-In this paper, we explore CH-Potentials, a novel combination of A* and the hierarchical techniques CH using some ideas from \cite{dgnw-phast-13}.
-We are not the first to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, past work mostly uses A* to accelerate hierarchical approaches and sacrifzies A*'s felxibility in the process.
+%
+CH-Potentials is not the first work to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
+However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifzies A*'s felxibility in the process.
 Our primary algorithm design objective is to retain A*'s flexibility.
 
-The most similar approaches to CH-Potentials are ALT and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19}.
+ALT~\cite{gh-cspas-05} and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
 Just as our approach, ALT is A*-based.
-However, ALT's potentials are not tight. 
+However, ALT's heuristic is not tight. 
 %
@@
 A CPD can quickly compute the first edge of a shortest path between any two nodes.
-In \cite{DBLP:conf/ijcai/BonoGHS19}, a CPD~\cite{DBLP:conf/socs/StrasserHB14} is used to create a potential.
-In every potential evaluation, a path to the target is compute, whose length is used as potential.
-The resulting potential is tight.
+In \cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
+In every potential evaluation, a shortest path to the target is compute, whose length is used as tight heuristic.
 Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
@@
 Oracle-A* is not achievable in practice.
-However, is a lower bound for every algorithm following our setup.
+However, it is a lower bound for every algorithm following our setup.
 We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.
@@
 A two-phase technique to efficiently compute tight potentials.
-Combined with A*, this yields an optimal, efficient, and flexibile route planning algorithms.
+Combined with A*, this yields an exact, efficient, and flexibile route planning algorithms.
 Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
@@
 
-A CH is a two phase technique to efficiently compute optimal, shortest paths.
+A CH is a two phase technique to efficiently compute exact, shortest paths.
 It is not flexibile, i.e., $w_q=w_\ell$.
@@
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-Using PHAST, we can compute optimal A* potentials in another way.
+Using PHAST, we can compute tight A* potentials in another way.
 In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
@@
 The resulting algorithm is the basic CH-potential algorithm.
-It is flexible and optimal and slightly faster than Dijkstra's algorithm.
+It is flexible and exact.
 However, we can still improve the performance.
@@
 Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
-This reduces the number of explored nodes and thus fewer potentials are evaluated.
+This reduces the number of explored nodes and thus fewer heuristic evaluations.
 
@@
 If this search finds $t$, then $s$ and $t$ are part of the same component.
-An optimal path was found and therefore the algorithm stops early.
+A shortest path was found and therefore the algorithm stops early.
 Otherwise, in the second step starts.
@@
 Such restrictions are usually modelled using turn weights.
-We extend CH-Potentials by using zero as lower bound for every turn weight in the potential.
+We extend CH-Potentials by using zero as lower bound for every turn weight in the heuristic.
 In this section, we first present the extended problem setting.
@@
 We therefore only apply the A* search to the expanded graph $G'$. 
-The potentials are computed with respect to the input graph $G$.
-Let $(p,q)$ be the target edge and $(x,y)$ the edge for which the A* search wants to compute a potential.
+The heuristic is computed with respect to the input graph $G$.
+Let $(p,q)$ be the target edge and $(x,y)$ the edge for which the A* search wants to compute a heuristic.
 We construct non-turn-aware CH-Potentials towards $q$.
-The potential for the edge $(x,y)$ is equal to the potential of $y$.
+The heuristic for the edge $(x,y)$ is equal to the heuristic of $y$.
 
-To prove correctness, consider the turn-aware path weight of an optimal path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
+To prove correctness, consider the turn-aware path weight of a shortest path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
 We can lower bound the expression as follows:
@@
 The shortest $yq$-path length is exactly what CH-Potentials compute.
-The potential is therefore a lower bound.
+The heuristic is therefore a lower bound.
 It is consistent because it is derived from exact shortest paths in $G$ with $w_\ell$.
@@
 
-We introduce a novel way to compute potentials for A*.
-The potentials are calculated utilizing a CH on a lower bound graph and tight with regards to that lower bound graph.
-With these potentials we can speed up A* search for extended route planning problems like route planning with live traffic, time-dependent travel time predictions or user preferences.
+TODO
 

### 2020-08-05 Rakefile.rb
update engine and rake script to include perfect exp

@@ -179,6 +179,7 @@ namespace "exp" do
       td_graphs.each do |graph|
         sh "cargo run --release --bin chpot_td -- #{graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       end
+      sh "cargo run --release --bin chpot_unmodified -- #{osm_eur} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_turns -- #{osm_eur} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_live -- #{osm_ger} #{live_dir} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_td_live -- #{osm_ger_td} #{live_dir} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"

### 2020-08-05 Rakefile.rb
fix var name

@@ -43,10 +43,10 @@ osm_eur_src_file = '/algoDaten/zeitz/roadgraphs/europe-200101.osm.pbf'
 
 osm_ger = "/algoDaten/zeitz/roadgraphs/osm_ger/"
 osm_ger_td = "/algoDaten/zeitz/roadgraphs/osm_ger_td/"
-osm_europe = "/algoDaten/zeitz/roadgraphs/osm_europe/"
+osm_eur = "/algoDaten/zeitz/roadgraphs/osm_europe/"
 dimacs_graph = "/algoDaten/zeitz/roadgraphs/europe/"
 
-static_graphs = [osm_ger, osm_europe, dimacs_graph]
+static_graphs = [osm_ger, osm_eur, dimacs_graph]
 td_graphs = [
   "/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/",
   "/algoDaten/graphs/cleaned_td_road_data/de/day/dido/",
@@ -91,10 +91,10 @@ namespace "prep" do
     end
   end
 
-  directory osm_europe
-  file osm_europe => ["code/osm_import/build/import_osm", osm_eur_src_file] do
+  directory osm_eur
+  file osm_eur => ["code/osm_import/build/import_osm", osm_eur_src_file] do
     wd = Dir.pwd
-    Dir.chdir osm_europe do
+    Dir.chdir osm_eur do
       sh "#{wd}/code/osm_import/build/import_osm #{osm_eur_src_file}"
     end
   end
@@ -149,28 +149,28 @@ namespace "exp" do
     end
   end
 
-  task scaled_weights: [dimacs_graph + 'lower_bound_ch', osm_ger + 'lower_bound_ch', osm_europe + 'lower_bound_ch'] + ["#{exp_dir}/scaled_weights"] do
+  task scaled_weights: [dimacs_graph + 'lower_bound_ch', osm_ger + 'lower_bound_ch', osm_eur + 'lower_bound_ch'] + ["#{exp_dir}/scaled_weights"] do
     Dir.chdir "code/rust_road_router/engine" do
       sh "cargo run --release --bin chpot_weight_scaling -- #{dimacs_graph} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin chpot_weight_scaling -- #{osm_ger} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
[…]

### 2020-08-07 eval/applications.py
new numbers, finally

@@ -68,7 +68,7 @@ table = queries.groupby(['graph', 'exp']).agg(
     size=('affected', 'count'))
 
 table = table.reindex(['/algoDaten/zeitz/roadgraphs/europe/',
-    '/algoDaten/zeitz/roadgraphs/osm_eur/',
+    '/algoDaten/zeitz/roadgraphs/osm_europe/',
     '/algoDaten/zeitz/roadgraphs/osm_ger/',
     '/algoDaten/zeitz/roadgraphs/osm_ger_td/',
     '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/',
@@ -78,6 +78,7 @@ table = table.reindex(['/algoDaten/zeitz/roadgraphs/europe/',
     'random_times_10',
     'fastest_times_2',
     'fastest_times_10',
+    'perfect',
     'chpot_live',
     'chpot_td_live',
     'chpot_turns_td_live',
@@ -96,13 +97,15 @@ table = table.round(1)
 table = table.rename(index={
     '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/': 'TDGer06',
     '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/': 'TDEur17',
+    '/algoDaten/zeitz/roadgraphs/osm_europe/': 'OSM Europe',
     '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Germany',
     '/algoDaten/zeitz/roadgraphs/osm_ger_td/': 'OSM Germany',
     '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe',
+    'perfect': 'Unmodified',
     'chpot_td': 'TD',
-    'chpot_live': 'Mapbox Live',
-    'chpot_td_live': 'Mapbox TD + Live',
-    'chpot_turns_td_live': 'Mapbox TD + Live + Turns',
+    'chpot_live': 'Live',
+    'chpot_td_live': 'TD + Live',
+    'chpot_turns_td_live': 'TD + Live + Turns',
     'chpot_turns': 'Turns',
     'no_highways': 'No Highways',
     'no_tunnels': 'No Tunnels',

### 2020-08-07 eval/building_blocks.py

@@ -12,33 +12,47 @@ paths = glob.glob(base + "building_blocks/*.json")
 data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
-    'features': run['feature_flags'].replace('DEFAULT, TDCCH_APPROX, TDCCH_POSTCUSTOMIZATION, TDCCH_PRECUSTOMIZATION, TDCCH_QUERY_ASTAR, TDCCH_TRIANGLE_SORTING', ''),
+    'features': run['feature_flags'],
     **algo }
-    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
+    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') in ['CH Potentials Query', 'Dijkstra Query']])
 
-table = queries.groupby('features')[['running_time_ms', 'num_queue_pops', 'num_relaxed_arcs']].mean()
+queries['BCC'] = ~queries['features'].str.contains('CHPOT_NO_BCC')
+queries['Skip Deg. 2'] = ~queries['features'].str.contains('CHPOT_NO_DEG2')
+queries['Skip Deg. 3'] = ~queries['features'].str.contains('CHPOT_NO_DEG3')
+
+def heuristic_name(feature_flags):
+    if 'CHPOT_ONLY_TOPO' in feature_flags:
+        return '-'
+    elif 'CHPOT_ORACLE' in feature_flags:
+        return 'Oracle'
+    else:
+        return 'CH'
+
+queries['Heuristic'] = queries['features'].apply(heuristic_name)
+
+table = queries.groupby(['algo', 'Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3'])[['running_time_ms', 'num_queue_pops', 'num_relaxed_arcs']].mean()
 table['num_relaxed_arcs'] = table['num_relaxed_arcs'] / 1000
 table['num_queue_pops'] = table['num_queue_pops'] / 1000
 table = table.round(1)
 
-table = table.rename(index={
-    'CHPOT_NO_BCC, CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ': 'No Deg2, No Deg3, No BCC, No Pot.',
[…]

### 2020-08-07 eval/scaled_weights.py

@@ -18,7 +18,7 @@ data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
     'weight_factor': exp['factor'],
-    'graph': { '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe', '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Germany', '/algoDaten/zeitz/roadgraphs/osm_eur/': 'OSM Europe' }[run['args'][1]],
+    'graph': { '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe', '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Germany', '/algoDaten/zeitz/roadgraphs/osm_europe/': 'OSM Europe' }[run['args'][1]],
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 

### 2020-08-07 Rakefile.rb
full scale exp

@@ -137,7 +137,7 @@ namespace "exp" do
 
   task preprocessing: ["code/compute_ch/build/compute_ch", "#{exp_dir}/preprocessing"] + graphs.map { |g| g + 'lower_bound' } do
     graphs.each do |graph|
-      5.times do
+      10.times do
         filename = "#{exp_dir}/preprocessing/" + `date --iso-8601=seconds`.strip + '.out'
         sh "echo '#{graph}' >> #{filename}"
         sh("code/compute_ch/build/compute_ch #{graph}/first_out #{graph}/head #{graph}/lower_bound " +
@@ -149,28 +149,30 @@ namespace "exp" do
     end
   end
 
-  task scaled_weights: [dimacs_graph + 'lower_bound_ch', osm_ger + 'lower_bound_ch', osm_eur + 'lower_bound_ch'] + ["#{exp_dir}/scaled_weights"] do
+  task scaled_weights: static_graphs.map { |g| g + 'lower_bound_ch' } + ["#{exp_dir}/scaled_weights"] do
     Dir.chdir "code/rust_road_router/engine" do
-      sh "cargo run --release --bin chpot_weight_scaling -- #{dimacs_graph} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
-      sh "cargo run --release --bin chpot_weight_scaling -- #{osm_ger} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
-      sh "cargo run --release --bin chpot_weight_scaling -- #{osm_eur} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
+      static_graphs.each do |graph|
+        sh "cargo run --release --bin chpot_weight_scaling -- #{graph} > #{exp_dir}/scaled_weights/$(date --iso-8601=seconds).json"
+      end
     end
   end
 
-  task building_blocks: [osm_eur + 'lower_bound_ch'] + ["#{exp_dir}/building_blocks"] do
+  task building_blocks: static_graphs.map { |g| g + 'lower_bound_ch' } + ["#{exp_dir}/building_blocks"] do
     Dir.chdir "code/rust_road_router/engine" do
[…]

### 2020-08-12 paper/ch_potentials.tex
ev routing citation
evaluation, finally

@@
 \usepackage{amsmath}
+% checkmarks
+\usepackage{pifont}
+\newcommand{\cmark}{\ding{51}}
+\newcommand{\xmark}{\ding{55}}
 
@@
 It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
-Efficiently computing flexible and exact routes is a thought challenge.
+Efficiently computing flexible and exact routes is a though challenge.
 \emph{Exact} means that the computed path always has minimum weight.
@@
 CH-Potentials is not the first work to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifzies A*'s felxibility in the process.
+However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
 Our primary algorithm design objective is to retain A*'s flexibility.
@@
 Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
+%
+Electric Vehicle Routing also requires flexibility~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
 
@@
 
-Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau_{\mathrm{soon}}$.
+Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
 Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
@@
 
-\section{Experiments}
+\section{Evaluation}
 
 \label{sec:experiments}
+
+\begin{table}
+\centering
+\caption{Instances used in the evaluation.}\label{tab:graphs}
+\input{table/graphs.tex}
+\end{table}
 
@@
 \includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
-\caption{Query running times for queries of DIMACs Europe with scaled edge weights.}\label{fig:scaled_weights}
+\caption{
+Running times on a logarithmic scale for queries on OSM Ger with scaled edge weights $w_q = \alpha \cdot w_\ell$.
+The boxes cover the range between the first and third quartile.
+The band in the box indicates the median, the diamond the mean.
+The whiskers cover 1.5 times the inter quartile range.
+All other running times are marked as outliers.
+}\label{fig:scaled_weights}
 \end{figure*}
@@
 All experiments were performed sequentially.
-Our code is written in Rust and compiled with rustc 1.44.0-nightly in the release profile with the target-cpu=native option.
+Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile with the target-cpu=native option.
 
 \subparagraph{Inputs and Methodology}
-Our main benchmark instance is a graph of the road network of Western Europe made publicly available for the Ninth DIMACS implementation Challenge~\cite{DemetrescuGJ09} with 18M nodes and 42M edges.
-We use this graph with the travel time as the lower bound and perform experiments with modified edge weights (scaling all travel times or reproducing an experiment from~\cite{dw-lbrdg-07} where 1\,000 random edges are perturbed).
-Additionally, we perform experiments on a recent OSM instance of Germany\footnote{\url{https://download.geofabrik.de/europe/germany.html}, obtained on August 2nd, 2019}.
+Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
 To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
-The graph has 11M nodes and 26M edges.
-For this instance, we have proprietary live traffic snapshot from Friday 2019/08/02 afternoon provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
-The live data comes in the form of 320K OSM node pairs and live speeds for the link between the nodes.
-We also have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
+The graph has 16M nodes and 35M edges.
+For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
+The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the link between the nodes.
+It also includes traffic predictions for 38\% of the links as predicted speeds for all five minute periods over the course of a week.
+Additionally, have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
 One is an old instance of Germany with traffic predictions from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges, from 2017).
-Preprocessing running times are averages over 10 runs.
+Table~\ref{tab:graphs} contains an overview over our instances.
+We report preprocessing running times as averages over 10 runs.
 For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
+
+\subparagraph{Experiments}
+We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
+For that, we take our main graph, use the unmodified travel time metric as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
+When $\alpha = 1$, the heuristic is as tight as possible.
+With growing $\alpha$ the heuristic degrades.
+Figure~\ref{fig:scaled_weights} depicts the results.
+Clearly, $\alpha$ has significant influence on the running time.
+Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
+Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
+We observe a significant amount of heavy outliers sometimes more than an order of magnitude slower than the median.
+Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
+The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
+When $\alpha$ grows, more and more nodes in the area around the shortest paths are included in the search space.
+So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when the shortest path goes only through sparse areas.
+% Conclusions?
+
+\begin{table}
+\centering
+\caption{Query performance with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
+\input{table/building_blocks.tex}
+\end{table}
+
+In Table~\ref{tab:building_blocks} we investigate the impact of our optimizations.
+We run queries with different heuristic functions: the \emph{Zero} heuristic which turns A* back into the classical algorithm of Dijkstra, our new CH-Potential heuristic, and an \emph{Oracle} heuristic where all distances to $t$ with regards to $w_\ell$ can be obtained immediately.
+We first observe that each one of our optimizations gains useful speedups regardless of the heuristic used.
+Our optimizations do not only reduce heuristic evaluations but also queue operations which always accelerates the algorithm.
+The number of queue pushs roughly corresponds to the running time for all heuristics.
+When comparing CH-Potentials to the Oracle heuristic we note that CH-Potentials are only a factor of 1.6 slower.
+This clearly shows that CH-Potentials have very little overhead.
+Assuming that $w_\ell$ is the tightest possible lower bound that can be known in advance, no other heuristic can be more than a factor of 1.6 faster than CH-Potentials without sacrificing exactness.
 
@@
 \centering
-\caption{Query performance with the different algorithmic features.}\label{tab:building_blocks}
-\input{table/building_blocks.tex}
+\caption{
+CH-Potentials performance for different route planning applications.
+We report average running times and number of queue pushs.
+We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
+Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup of CH-Potentials over the baseline.
+}\label{tab:applications}
+\input{table/applications.tex}
 \end{table*}
 
-\begin{table*}
-\centering
-% TODO dijkstra baseline, preprocessing
-\caption{Performance for different applications.}\label{tab:applications}
-\input{table/applications.tex}
-\end{table*}
+Table~\ref{tab:applications} depicts results when applying our algorithm to different route planning problems.
+The greatest speedup is, of course, achieved when applying CH-Potentials to an unmodified graph with $w_q = w_\ell$.
+In that case the performance is comparable to what other state-of-the-art route planning techniques like CH achieve.
+When considering extended problem settings, the performance of CH-Potentials strongly depends on the tightness of $w_\ell$, as previously observed in Figure~\ref{fig:scaled_weights}.
+On one hand, forbidding highways leads to over 40\% longer paths and the smallest speedup over Dijkstra (only 5.3) among all settings.
+On the other hand, incorporating turn restrictions increase path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity.
+% other techniques need crazy shit, cite turn papers
+Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 125\,ms on average.
+The Mapbox traffic predictions increase shortest path lengths even stronger and running times increase correspondingly.
+The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to much better running times and speedups.
+Assessing which of these predictions is more realistic is not possible within the scope of this paper.
+Combining Mapbox live traffic and traffic predictions yields running times only slightly slower than traffic predictions alone.
+To the best of our knowledge, no other exact technique so far could combine these two scenarios.
+Additionally adding turn restrictions slows running times down to around 850\,ms which is still a speedup of 7.3 over the baseline despite the length increase only growing by 1.5 percentage points.
+The reason for this is that the turn expanded graph is three times bigger and consists of only one biconnected component, which disables our optimization from Section~\ref{sec:largested-biconnected-component}.
+To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.
+While the achieved speedup in our case is not particularly large, the reason is not the complexity of the scenario but the length increase in the time-dependent data.
 
@@
 
-TODO
+In this paper, we CH-Potentials introduced a fast and exact new A* heuristic for finding shortest paths in road networks.
+The approach is flexible and can handle a multitude of routing scenarios with very small implementation complexity.
+However, the performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.
+Nevertheless, our experiments show that CH-Potentials are efficient in extracting distance estimates from a lower bound graph and only around 1.6 times slower than extracting lower bound distances from a precomputed matrix.
 

### 2020-08-12 paper/table/graphs.tex
evaluation, finally

\begin{tabular}{lrrr}
\toprule
 &          Nodes &          Edges & Preprocessing \\ & $[\cdot 10^6]$ & $[\cdot 10^6]$ &           [s] \\
\midrule
OSM Ger &       16.2 &       35.4 &                         295.2 \\
TDEur17 &       25.8 &       55.5 &                         292.6 \\
TDGer06 &        4.7 &       10.8 &                          58.9 \\
\bottomrule
\end{tabular}

### 2020-08-12 Rakefile.rb

@@ -5,6 +5,7 @@ file "paper/ch_potentials.pdf" => [
   "paper/ch_potentials.tex",
   "paper/fig/scaled_weights.pdf",
   "paper/table/applications.tex",
+  "paper/table/graphs.tex",
   "paper/table/building_blocks.tex"] do
 
   Dir.chdir "paper" do
@@ -29,6 +30,14 @@ namespace "table" do
     ] + ["eval/applications.py", "paper/table"] do
 
     sh "eval/applications.py"
+  end
+
+  file "paper/table/graphs.tex" => FileList[
+    "#{exp_dir}/preprocessing/*.out",
+    "#{exp_dir}/applications/*.json",
+    ] + ["eval/graphs.py", "paper/table"] do
+
+    sh "eval/graphs.py"
   end
 
   file "paper/table/building_blocks.tex" => FileList["#{exp_dir}/building_blocks/*.json"] + ["eval/building_blocks.py", "paper/table"] do

### 2020-08-12 eval/applications.py

@@ -38,34 +38,16 @@ dijkstra_queries = dijkstra_queries.append(pd.DataFrame.from_records([{
     for run in data for algo in run.get('algo_runs', []) if algo.get('algo') == 'Dijkstra Query']))
 
 
-runtime_pattern = re.compile(".*Needed (\\d+)musec\\..*")
-
-def parse_contraction_output(path):
-    stats = { 'preprocessing_running_time_s': 0.0 }
-
-    with open(path, 'r') as f:
-        for line in f:
-            if not 'graph' in stats:
-                stats['graph'] = line.strip()
-            else:
-                match = runtime_pattern.match(line)
-                if match:
-                    stats['preprocessing_running_time_s'] += int(match[1]) / 1000000
-
-    return stats
-
-preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for path in glob.glob(base + "preprocessing/*.out")])
-
 queries['affected'] = queries['lower_bound'] != queries['result']
 queries['increase'] = (queries['result'] / queries['lower_bound'] - 1) * 100
 
 table = queries.groupby(['graph', 'exp']).agg(
     running_time_ms=('running_time_ms', 'mean'),
-    num_pot_evals=('num_pot_evals', 'mean'),
-    num_relaxed_arcs=('num_relaxed_arcs', 'mean'),
+    num_queue_pushs=('num_queue_pushs', 'mean'),
     increase=('increase', 'mean'),
-    affected=('affected', 'sum'),
-    size=('affected', 'count'))
+    # affected=('affected', 'sum'),
+    # size=('affected', 'count')
+    )
 
 table = table.reindex(['/algoDaten/zeitz/roadgraphs/europe/',
     '/algoDaten/zeitz/roadgraphs/osm_europe/',
@@ -79,36 +61,36 @@ table = table.reindex(['/algoDaten/zeitz/roadgraphs/europe/',
     'fastest_times_2',
     'fastest_times_10',
     'perfect',
+    'chpot_turns',
+    'no_tunnels',
+    'no_highways',
     'chpot_live',
     'chpot_td_live',
     'chpot_turns_td_live',
-    'no_tunnels',
-    'no_highways',
-    'chpot_turns'], level=1)
+    ], level=1)
 
[…]

### 2020-08-12 eval/building_blocks.py

@@ -22,7 +22,7 @@ queries['Skip Deg. 3'] = ~queries['features'].str.contains('CHPOT_NO_DEG3')
 
 def heuristic_name(feature_flags):
     if 'CHPOT_ONLY_TOPO' in feature_flags:
-        return '-'
+        return 'Zero'
     elif 'CHPOT_ORACLE' in feature_flags:
         return 'Oracle'
     else:
@@ -30,32 +30,25 @@ def heuristic_name(feature_flags):
 
 queries['Heuristic'] = queries['features'].apply(heuristic_name)
 
-table = queries.groupby(['algo', 'Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3'])[['running_time_ms', 'num_queue_pops', 'num_relaxed_arcs']].mean()
-table['num_relaxed_arcs'] = table['num_relaxed_arcs'] / 1000
-table['num_queue_pops'] = table['num_queue_pops'] / 1000
+table = queries.groupby(['algo', 'Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3'])[['running_time_ms', 'num_queue_pushs']].mean()
+table['num_queue_pushs'] = table['num_queue_pushs'] / 1000
 table = table.round(1)
 
-# table = table.rename(index={
-#     'CHPOT_NO_BCC, CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ': 'No Deg2, No Deg3, No BCC, No Pot.',
-#     'CHPOT_NO_DEG2, CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ':               'No Deg2, No Deg3, No Pot.',
-#     'CHPOT_NO_DEG3, CHPOT_ONLY_TOPO, ':                              'No Deg3, No Pot.',
-#     'CHPOT_ONLY_TOPO, ':                                             'No Pot.',
-#     'CHPOT_NO_BCC, CHPOT_NO_DEG2, CHPOT_NO_DEG3, ':                  'No Deg2, No Deg3, No BCC',
-#     'CHPOT_NO_DEG2, CHPOT_NO_DEG3, ':                                'No Deg2, No Deg3',
-#     'CHPOT_NO_DEG3, ':                                               'No Deg3',
-# })
+table = table.reindex(['Zero', 'CH', 'Oracle'], level=1)
 
[…]

### 2020-08-12 eval/graphs.py

@@ -1,0 +1,71 @@
+#!/usr/bin/env python3
+
+import pandas as pd
+
+import json
+import glob
+import os
+import re
+
+base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+paths = glob.glob(base + "applications/*.json")
+data = [json.load(open(path)) for path in paths]
+
+
+queries = pd.DataFrame.from_records([{
+    'graph': run['args'][1],
+    'num_nodes': (run['graph'] if 'graph' in run else run['experiments'][0]['graph'])['num_nodes'],
+    'num_edges': (run['graph'] if 'graph' in run else run['experiments'][0]['graph'])['num_arcs'],
+    } for run in data if 'graph' in run or 'experiments' in run])
+
+runtime_pattern = re.compile(".*Needed (\\d+)musec\\..*")
+
+def parse_contraction_output(path):
+    stats = { 'preprocessing_running_time_s': 0.0 }
+
+    with open(path, 'r') as f:
+        for line in f:
+            if not 'graph' in stats:
+                stats['graph'] = line.strip()
+            else:
+                match = runtime_pattern.match(line)
+                if match:
+                    stats['preprocessing_running_time_s'] += int(match[1]) / 1000000
+
+    return stats
+
+preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for path in glob.glob(base + "preprocessing/*.out")])
+
+table = queries.groupby(['graph'])[['num_nodes', 'num_edges']].mean()
+
+table = table.reindex([
+    '/algoDaten/zeitz/roadgraphs/osm_ger/',
+    '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/',
+    '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/'])
+
+table = table.join(preprocessing.groupby('graph').mean())
+table['num_nodes'] = table['num_nodes'] / 1000000.0
+table['num_edges'] = table['num_edges'] / 1000000.0
+table = table.round(1)
+
+table = table.rename(index={
+    '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/': 'TDGer06',
+    '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/': 'TDEur17',
[…]

### 2020-08-12 eval/scaled_weights.py

@@ -22,10 +22,11 @@ queries = pd.DataFrame.from_records([{
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 
-plt.figure(figsize=(11,4))
+plt.figure(figsize=(10.5,4))
 g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8, showmeans=True)
+g.set_yscale('log')
 g.set_ylabel('Running Time [ms]')
-g.set_xlabel('Weight Factor')
-g.legend().set_title('')
+g.set_xlabel('Weight Factor $\\alpha$')
+g.legend().remove()
 plt.tight_layout()
 g.get_figure().savefig('paper/fig/scaled_weights.pdf')

### 2020-08-13 eval/building_blocks.py
final numbers (mapbox td significantly improved by filtering outliers in profiles)

@@ -44,7 +44,7 @@ lines = lines[:2] + [
     R" & & & & Running &     Queue \\",
     R" Heur. & BCC & Deg2 & Deg3 & time [ms] & pushs [$\cdot 10^3$] \\"
 # ] + lines[4:]
-] + [lines[4], lines[17]] + lines[6:14] + [lines[16]] + lines[20:]
+] + [lines[4], lines[17]] + lines[6:9] + ["\\addlinespace"] + lines[9:13] + ["\\addlinespace"] + [lines[13], lines[16]] + lines[20:]
 
 output = "\n".join(lines) + "\n"
 output = output.replace('True', '\\cmark')

### 2020-08-13 paper/ch_potentials.tex
add shiny search space figure
updates and improvements to abstract and Sections 1-3

@@
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
-A* is the classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
+A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
 However, A*'s performance depends on the availability of a good heuristic.
 Computing a good heuristic is a challenge of its own.
-To avoid the need for a heuristic, many hierarchical speedup techniques, such as Contraction Hierarchies (CH), have been developped.
-These techniques are not based on A* and therefore do not need a heuristic.
-They achieve speed and exactness but sacrifize A*'s flexibility.
-In this paper, we use hierarchical techniques to compute tight A* heuristics.
+In contrast, for road networks there also exist hierarchical speedup techniques, which do not need a heuristic.
+They achieve speed and exactness but sacrifice A*'s flexibility.
+In this paper, we use such as Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
 We call our technique CH-Potential.
+With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
 Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
-With this approach, we combine some of the hierarchical speed with the flexibility offered by A* - the best of both worlds - while retaining exactness.
 \end{abstract}
@@
 It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
-Efficiently computing flexible and exact routes is a though challenge.
+Developing flexible techniques to efficiently compute exact routes is a tough challenge.
 \emph{Exact} means that the computed path always has minimum weight.
 \emph{Efficiently} means that the path computation is fast.
-Being able to change the edge weights between path computations is \emph{flexibility}.
+Being able to change the edge weights between path computations or handle more complex edge weights (for example travel times which change over the course of the day obtained through traffic predictions) is \emph{flexibility}.
 
@@
 \centering
-\includegraphics[width=\columnwidth,height=200px]{fig/searchspace}
+\includegraphics[width=\columnwidth]{fig/searchspace.png}
 
-\caption{Example A* search space with CH-Potentials}
+\caption{Nodes explored by A* with CH-Potentials. Nodes were explored from blue through green to red.}
 \label{img:search-space}
@@
 
-Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexiblity and exactness but is comparatively slow.
+Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
+Dijkstra's algorithm iteratively explores nodes of the graph with ascending distance from the origin node until the target is reached.
 A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
-While flexible, A*'s performance crucially depends on the quality of the employed heuristic.
+It changes the order in which nodes are explored by using the distance from the origin plus an estimate of the remaining distance to the target as the ordering criterion.
+Figure~\ref{img:search-space} depicts an example A* search.
+While flexible, A*'s performance crucially depends on the quality of the employed estimating heuristic.
 Finding a good heuristic that can be evaluated efficiently is difficult.
@@
 In this paper, we consider a setup with a preprocessing phase and a query phase.
-The preprocessing phase can be slow and computes auxilary data that can be used in the query phase, which should be fast.
+The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
 The input to the preprocessing phase is the graph weighted with lower bound weights $w_\ell$.
@@
 Street networks of continental scale have many millions of nodes.
-While the preprocessing phase may be slow, it must eventually finish.
-On graphs of this size, quadratic preprocessing time is thus prohibitive.
+% While the preprocessing phase may be slow, it must eventually finish.
+On graphs of this size, quadratic preprocessing time is prohibitive.
 
 In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
-The preprocessing phase computes auxilary data to efficiently evaluate the heuristic.
+The preprocessing phase computes auxiliary data to efficiently evaluate the heuristic.
 Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-We call our algorithm CH-Potentials, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, which are called potentials in~\cite{gh-cspas-05}.
+We call our algorithm CH-Potentials, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, also called potentials in~\cite{gh-cspas-05}.
 
@@
 $h$ depends on the target $t$ of the search but not one the source $s$.
-Without loose of generality, we assume that $h(t)=0$.
+Without loss of generality, we assume that $h(t)=0$.
 The \emph{tight} heuristic assigns to every node $v$, the length of a shortest $vt$-path.
 All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
-CH-Potentials compute the tight heursitc with respect to $w_\ell$.
+CH-Potentials compute the tight heuristic with respect to $w_\ell$.
 Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
-CH-Potentials are optimal in this sense.
+In this sense, CH-Potentials are as tight as possible.
 
@@
 A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
-In this setting, mostly hierarchical approaches dominate in termes of running time~\cite{bdgmpsww-rptn-16}.
-A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12,dsw-cch-15}.
+In this setting, mostly hierarchical approaches dominate in terms of running time~\cite{bdgmpsww-rptn-16}.
+A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12}.
 
 There is a lot of work that extends CH to handle more complex settings.
-For example, in \cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
-In \cite{dsw-cch-15}, CH is modified to support Live-Traffic.
-Historic traffic patterns are combined with CH in \cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
-Specialized variants for electric vehicle routing are developped in \cite{bdgwz-sfpcs-19}.
+For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
+In~\cite{dsw-cch-15}, CH is modified to support Live-Traffic.
+Historic traffic patterns are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
+Specialized variants for electric vehicle routing are developed in~\cite{bdgwz-sfpcs-19}.
 While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
@@
 %
-CH-Potentials is not the first work to combine hierarchical approaches and A* \cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
+CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
 However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
@@
 
-ALT~\cite{gh-cspas-05} and CPD-Heursitics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
+ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
 Just as our approach, ALT is A*-based.
@@
 %
-CPD-Heursitics are a combination of A* and Compressed Path Databases (CPD).
+CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
 A CPD can quickly compute the first edge of a shortest path between any two nodes.
-In \cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
-In every potential evaluation, a shortest path to the target is compute, whose length is used as tight heuristic.
+In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
+For every distance estimation, a shortest path to the target is computed, whose length is used as the tight heuristic.
 Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
@@
 To compare against related work, we introduce the concept of Oracle-A*.
-It has access to a shortest distance array with respect to $w_\ell$ as potential, i.e., it is tight.
+It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
 This is the tightest potential achievable during preprocessing.
@@
 Many applications need flexibility.
-In road routing, the most common ones are traffic avoidance, user preferences, and vehicle restrictions.
-Traffic avoidance comes in two forms: Live~\cite{CRP,CCH} and predicted traffic~\cite{td-alt,td-ch}.
+The most common ones are traffic avoidance, user preferences, and vehicle restrictions.
+Traffic avoidance comes in two forms: Live~\cite{dgpw-crprn-13,dsw-cch-15} and predicted traffic~\cite{ndls-bastd-12,bgsv-mtdtt-13}.
 Predicted traffic is often called time-dependent routing.
 %
-User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}
-For example, for some users shorter routes are more important fast routes.
+User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}.
+For example, for some users shorter routes are more important than fast routes.
 Others want to avoid tunnels or highways.
 %
-Truck routing differs from car routing because, depending on their load, trucks are not allowed everywhere~\cite{kswz-erptd-20}.
+Truck routing differs from car routing due to night driving bans and other restrictions~\cite{kswz-erptd-20}.
 %
@@
 
-We introduce CH-Potentials in Section~\ref{sec:main-algo}.
-A two-phase technique to efficiently compute tight potentials.
-Combined with A*, this yields an exact, efficient, and flexibile route planning algorithms.
-Figure~\ref{img:search-space} depicts an example A* search with CH-Potentials.
-Beside the potentials, we describe, in Section\ref{sec:low-deg-improvment}, some improvments to A* to more efficiently process low degree nodes, common in road networks.
-In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and showing how they can be solved using CH-Potentials.
-Finally in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
-By comparing against Oracle-A*, we show that no alternative approach can be significantly faster.
+We introduce CH-Potentials, a two-phase technique to efficiently compute tight heuristics, in Section~\ref{sec:main-algo}.
+Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
+Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we describe some improvements to A* to more efficiently process low degree nodes, common in road networks.
+In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and how they can be solved using CH-Potentials.
+Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
+By comparing against Oracle-A*, we show that no alternative approach can be significantly faster, assuming that $w_\ell$ is the tightest lower bound that can be known in advance.
 
@@
 
-PHAST \cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
+PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
 First the preprocessing phase is executed analogously to the original CH.
@@
 The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
-When $\alpha$ grows, more and more nodes in the area around the shortest paths are included in the search space.
+When $\alpha$ grows, more and more nodes in the area around the shortest paths are explored.
 So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when the shortest path goes only through sparse areas.

### 2020-08-13 paper/ch_potentials.tex
updates and improvements to sections 4-6

@@
 For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
-In this section, we give an overview.
+In this section, we give an introduction.
 
@@
 \emph{Down} edges are defined analogously.
-An \emph{up-down path} is a path where only one node $v$ is more important than its neighbors.
+An \emph{up-down path} is a path where only one node $v$ is more important than both its neighbors.
 $v$ is called the \emph{mid} node.
 An \emph{up path} is a path where the last node is the mid node.
-Similarily, the first node is the mid node of a \emph{down path}.
+Similarly, the first node is the mid node of a \emph{down path}.
 Every up and down path is an up-down path.
@@
 See~\cite{gssv-erlrn-12} for the details.
-The trick is that in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
-See Figure~\ref{fig:ch} for a proof scetch.
+After the preprocessing, in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
+See Figure~\ref{fig:ch} for a proof sketch.
 From every shortest path (red) in $G$, a up-down path of equal length in $G+$ (blue) can be constructed.
@@
 The forward search starts from $s$ and only follows up-edges.
-Similarily, the backward search starts at $t$ and only follows down-edges in reversed direction.
-The two searches meet in the mid node.
+Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
+The two searches meet at the mid node.
 Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
@@
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
-On road networks, this is the case as demonstrated in \cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.
+On road networks, this is the case as demonstrated in~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.
 
-Using a CH, we can easily compute tight A* potentials.
+Using a CH, we can compute a tight A* heuristic.
 In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
-The query phase consists of an A* search, where the evaluation of the potential $\pi(v)$ performs a CH-query from $v$ to $t$.
+The query phase consists of an A* search, where the evaluation of the heuristic $\pi(v)$ performs a CH-query from $v$ to $t$.
 %
-This works, however, while one CH query is fast, answering one for every node in the $A^*$ search is slow.
-Fortunatly, we can do better.
-However, for this we need another component called PHAST \cite{dgnw-phast-13}.
+This works, however, while one CH query is fast, answering one for every node explored in the $A^*$ search is slow.
+Fortunately, we can do better.
+However, for this we need another component called PHAST~\cite{dgnw-phast-13}.
 
@@
 Algorithm~\ref{algo:ch-backward} shows this first step.
-The second step iterates over all CH level from top to bottom.
+The second step iterates over all CH levels from top to bottom.
 In each iteration, all up-edges starting within the current level are followed in reverse.
-After all level are processed, the distances from all nodes towards $t$ is computed.
+After all levels are processed, the distances from all nodes towards $t$ is computed.
 PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
-PHAST's main advantage is that is easily be parallelized.
+PHAST's main advantage is that can easily be parallelized.
 However, we will not consider parallelization in this paper.
-We refer to \cite{dgnw-phast-13} for an in-depth experimental performance analysis.
+We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-Using PHAST, we can compute tight A* potentials in another way.
+Using PHAST, we can compute a tight A* heuristic in another way.
 In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
-In the next step, we run $A^*$ and implement the potential function as lookup in the array $H$.
+In the next step, we run $A^*$ and implement the heuristic function as lookup in the array $H$.
 
 This PHAST-based algorithm works.
-The potential evaluation and by extension the $A^*$ search is indeed fast.
+The heuristic evaluation and by extension the $A^*$ search is indeed fast.
 However, the PHAST step before the search is comparatively expensive.
-The reason for this is that the distance from all nodes towards $t$ are computed.
+The reason for this is that the distances from all nodes towards $t$ are computed.
 Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.
@@
 However, we can still improve the performance.
-A quick investigation reveals, that most of the running time is spent in the potential evaluation.
-We therefore investigate methods to evaluate fewer potentials.
+A quick investigation reveals, that a significant amount of running time is spent in the heuristic evaluation.
+We therefore investigate methods to evaluate the heuristic less often.
 
-\section{Low Degree A* Improvments}
+\section{Low Degree A* Improvements}
 
@@
 
-Potential evaluation is an expensive operation that we want to avoid.
-A potential evaluation is always paired with a queue push operation.
-By avoiding push operations, we can therefore avoid potential evaluations.
+Heuristic evaluation is an expensive operation that we want to avoid.
+The heuristic needs to be evaluated when a node is pushed into the queue.
+By avoiding push operations, we can therefore avoid heuristic evaluations.
 This is the objective of this section.
@@
 
-We modify how A* by processing low degree nodes consecutively without pushing them into the queue.
+We modify A* by processing low degree nodes consecutively without pushing them into the queue.
 Our algorithm makes use of the undirected degree $d(x)$ of a node $x$.
@@
 Our algorithm, however, behaves differently, if $d(y)\le 2$.
-Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z)\neq z$.
+Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
 If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
@@
 If it does not reach $z$, then only $D$ is modified but no queue action is performed.
-If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[y]+\pi(y)$.
+If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+\pi(z)$.
 
@@
 
-Denote by $x,y_1,\ldots, y_k, z$ a degree chain as described in the previous section.
-If $z$ is already in the queue, then $z$'s weight in the queue is adjusted as in the previous section.
-
-If $d(z)\neq 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
+Denote by $x,y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
+If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
 Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
@@
 Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
-This reduces the number of explored nodes and thus fewer heuristic evaluations.
+This reduces the number of explored nodes and thus leads to fewer heuristic evaluations.
 
@@
 We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
-The basic CH-potential preprocessing step is only executed for $G_C$.
+The CH-Potentials preprocessing step is only executed for $G_C$.
 
@@
 Otherwise, in the second step starts.
-It consits of applying the basic CH-potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
+It consists of applying the CH-Potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
 Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
@@
 If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
-Otherwise, $w_q$ is also to the freeflow travel time.
+Otherwise, $w_q$ is set to the freeflow travel time.
 This is all that is necessary to make our CH-Potentials algorithm avoid tunnels and/or highways.
@@
 However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
-Such restrictions are usually modelled using turn weights.
+Such restrictions are usually modeled using turn weights.
 We extend CH-Potentials by using zero as lower bound for every turn weight in the heuristic.
@@
 
-In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{related-turn-litrature,see-geisberger-turn-table-paper}.
+In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
 $w_t$ maps a pair of incident edges onto the time required to perform the turn.
 Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup \{+\infty\}$.
-A forbidden turn is modelled by setting the corresponding $w_t$ to $+\infty$.
+A forbidden turn is modeled by setting the corresponding $w_t$ to $+\infty$.
 
@@
 However, this implies computing a CH on the expanded graph.
-This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{geisberger-turn-table-paper}.
+This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{gv-errnt-11}.
 We want to avoid these complications.
@@
 \]
-The $\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ expression is the length of some, not necessarly shortest, $yq$-path in $G$ with $w_\ell$.
+The $\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ expression is the length of some, not necessarily shortest, $yq$-path in $G$ with $w_\ell$.
 It is by definition not smaller than the shortest $yq$-path length in $G$ with $w_\ell$.
@@
 Unfortunately, the undirected graph underlying the expanded graph is always biconnected, if the input graph is strongly connected.
-The optimization described in \ref{sec:largested-biconnected-component} is therefore ineffective in this setting.
+The optimization described in~\ref{sec:largested-biconnected-component} is therefore ineffective in this setting.
 
-With this setup we extended CH-Potentials in a way that keeps all turn-related information out of the CH.
+With this setup we extend CH-Potentials in a way that keeps all turn-related information out of the CH.
 This allows for a comparatively simple implementation.
@@
 Recurring traffic can be predicted by observing the traffic in the past.
-It is common \cite{lots-of-td-papers} to represent these predictions as \emph{travel time function}.
+It is common \cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-19} to represent these predictions as \emph{travel time functions}.
 An edge weight is no longer a scalar value but a function that maps the entry time onto the traversal time.
@@
 
-If all travel time functions fulfill the \emph{no-waiting property}, this problem can be solved using a straight forward extension of Dijkstra's algorithm \cite{td-alt?}.
+If all travel time functions fulfill the \emph{no-waiting property}, this problem can be solved using a straight forward extension of Dijkstra's algorithm \cite{d-aassp-69}.
 The necessary modification to A* is analogous.
-Without the no-waiting property the problem is in general NP-hard \cite{veit}.
-The no-waiting property states that it is never benficial to wait at a node before entering an edge.
+Without the no-waiting property the problem is in general NP-hard \cite{or-tnp-89}.
+The no-waiting property states that it is never beneficial to wait at a node before entering an edge.
 Formally stated, the following must hold:\[
@@
 Our implementation follows this setup.
-However in priciple, our algorithm works with any representation that enables an efficient function evaluation.
+However in principle, our algorithm works with any representation that enables an efficient function evaluation.
 
-To solve the time-dependent routing problem, we employ a stratedgy very similar to TD-ALT~\cite{td-alt}.
+To solve the time-dependent routing problem, we employ a strategy very similar to TD-ALT~\cite{ndls-bastd-12}.
 The main difference is that we use CH-Potentials instead of landmarks to guide the A* search.
@@
 We modify the A* search to use the tentative distance at a node $x$ as $\tau$ when evaluating the weight of an outgoing edge $(x,y)$.
-This modification is straight-forward, as our optimizations do not use shortcuts and only travers edges in forward direction.
-For the correctness proof, we refer to~\cite{td-alt?}.
+This modification is straight-forward, as our optimizations do not use shortcuts and only traverse edges in forward direction.
+For the correctness proof, we refer to~\cite{dw-lbrdg-07}.
 
@@
 As we were able to keep all travel time functions out of the CH.
-This avoids a lot of algorithmic complications compared to~\cite{TD-CH,CatchUp,TD-CRP}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
+This avoids a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-19}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
 
@@
 The predicted travel time along an edge was estimated in the past.
-It is possible that it differes from the current travel time significantly, if unexpected events happen.
-An accident is an example of an unexpected event.
+It is possible that it differs from the current travel time significantly, if unexpected events happen.
+Accidents are examples of such unexpected events.
 Live traffic data is more accurate for the current moment than predicted data.
@@
 We did not make any modification to our algorithm, that would make a combination with tunnel and/or highway avoidance or turn-aware routing difficult.
-This straight-forward integration of complex routing problems is the strengh of the CH-Potential algorithm.
+This straight-forward integration of complex routing problems is the strength of the CH-Potential algorithm.
 

### 2020-08-13 paper/ch_potentials.tex
some polishing to exp section

@@
 For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
-The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the link between the nodes.
-It also includes traffic predictions for 38\% of the links as predicted speeds for all five minute periods over the course of a week.
+The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
+It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
+We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
 Additionally, have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
-One is an old instance of Germany with traffic predictions from 2006 (4.7M nodes, 10.7M edges) and the other a newer instance of Europe (25.8M nodes, 55.5M edges, from 2017).
+One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
 Table~\ref{tab:graphs} contains an overview over our instances.
 We report preprocessing running times as averages over 10 runs.
-For the queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
+For queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
 
@@
 We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
-For that, we take our main graph, use the unmodified travel time metric as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
+For that, we take our main graph, use the unmodified travel time as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
 When $\alpha = 1$, the heuristic is as tight as possible.
@@
 Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
-We observe a significant amount of heavy outliers sometimes more than an order of magnitude slower than the median.
+We observe a significant amount of heavy outliers -- sometimes more than an order of magnitude slower than the median.
 Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
 The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
-When $\alpha$ grows, more and more nodes in the area around the shortest paths are explored.
-So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when the shortest path goes only through sparse areas.
+As $\alpha$ grows, more and more nodes in the area around the shortest paths are explored.
+So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, then when the shortest path goes only through sparse areas.
 % Conclusions?
@@
 On the other hand, incorporating turn restrictions increase path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity.
-% other techniques need crazy shit, cite turn papers
-Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 125\,ms on average.
-The Mapbox traffic predictions increase shortest path lengths even stronger and running times increase correspondingly.
-The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to much better running times and speedups.
+Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 127\,ms on average.
+The Mapbox traffic predictions increase shortest path lengths by about 18\%.
+Combined with the additional work from function evaluations this leads to running times around 200\,ms.
+The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to better running times and speedups.
 Assessing which of these predictions is more realistic is not possible within the scope of this paper.
@@
 To the best of our knowledge, no other exact technique so far could combine these two scenarios.
-Additionally adding turn restrictions slows running times down to around 850\,ms which is still a speedup of 7.3 over the baseline despite the length increase only growing by 1.5 percentage points.
-The reason for this is that the turn expanded graph is three times bigger and consists of only one biconnected component, which disables our optimization from Section~\ref{sec:largested-biconnected-component}.
+Additionally adding turn restrictions slows running times down to around 510\,ms despite the length increase only growing by 1.3 percentage points.
+This is still a speedup of 12.2 over the baseline.
+The reason for this is that the turn expanded graph is three times bigger and that the BCC optimization is not applicable for turn-expanded graphs.
 To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.
-While the achieved speedup in our case is not particularly large, the reason is not the complexity of the scenario but the length increase in the time-dependent data.
 
@@
 
-In this paper, we CH-Potentials introduced a fast and exact new A* heuristic for finding shortest paths in road networks.
+In this paper, we introduced CH-Potentials, a fast and exact new A* heuristic for finding shortest paths in road networks.
 The approach is flexible and can handle a multitude of routing scenarios with very small implementation complexity.
-However, the performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.
-Nevertheless, our experiments show that CH-Potentials are efficient in extracting distance estimates from a lower bound graph and only around 1.6 times slower than extracting lower bound distances from a precomputed matrix.
+The performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.
+Our experiments show that CH-Potentials are efficient in extracting distance estimates from a lower bound graph and only around 1.6 times slower than immediately obtaining these estimates from an hypothetical oracle.
 

### 2020-08-13 paper/ch_potentials.tex
move truck routing from related work to applications

@@
 Others want to avoid tunnels or highways.
-%
-Truck routing differs from car routing due to night driving bans and other restrictions~\cite{kswz-erptd-20}.
 %
@@
 
+\subsection{Temporary Driving Bans}
+
+Truck routing differs from car routing due to night driving bans and other restrictions.
+In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is successfully employed to efficiently find routes for trucks when edges may be blocked during certain times, drivers may need to wait at parking locations, and several routes with different trade-offs between arrival time and route quality may need to be considered.
+
 \section{Evaluation}

### 2020-08-13 eval/scaled_weights.py
make boxplot thinner

@@ -22,7 +22,7 @@ queries = pd.DataFrame.from_records([{
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 
-plt.figure(figsize=(10.5,4))
+plt.figure(figsize=(10.5,3))
 g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8, showmeans=True)
 g.set_yscale('log')
 g.set_ylabel('Running Time [ms]')

### 2020-08-13 paper/ch_potentials.tex
improve pic
some cleanup in story and context

@@
 In this paper, we use such as Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
-We call our technique CH-Potential.
+We call our technique CH-Potentials.
 With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
@@
 \centering
-\includegraphics[width=\columnwidth]{fig/searchspace.png}
+\includegraphics[width=\columnwidth]{fig/searchspace_st.png}
 
@@
 It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
-This is the tightest potential achievable during preprocessing.
+This is the tightest heuristic achievable during preprocessing.
 The array computation time is unaccounted for, i.e., it is magically filled instantly.
@@
 
-In this section, we first describe Contraction Hierarchies, then PHAST, and finally CH-Potentials.
+In this section, we first describe Contraction Hierarchies, then PHAST, a CH extension, and finally CH-Potentials.
 
@@
 A CH is a two phase technique to efficiently compute exact, shortest paths.
-It is not flexibile, i.e., $w_q=w_\ell$.
+It is not flexible, i.e., $w_q=w_\ell$.
 For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
@@
 
-\subsection{Potentials using PHAST}
+\subsection{PHAST based heuristic}
 
@@
 It is flexible and exact.
-However, we can still improve the performance.
-A quick investigation reveals, that a significant amount of running time is spent in the heuristic evaluation.
-We therefore investigate methods to evaluate the heuristic less often.
+However, we can still improve the performance by reducing the number of queue operations and heuristic evaluations.
 
@@
 Heuristic evaluation is an expensive operation that we want to avoid.
-The heuristic needs to be evaluated when a node is pushed into the queue.
-By avoiding push operations, we can therefore avoid heuristic evaluations.
-This is the objective of this section.
+Together with the queue operations, it makes up a significant amount of the running time of A*.
+The heuristic needs to be evaluated whenever a node is pushed into the queue.
+As our experiments show, the total running time roughly corresponds to the number of queue pushs.
+In this section, we propose optimizations to reduce the number of nodes pushed into the queue.
 

### 2020-08-13 paper/ch_potentials.tex
graph preliminaries

@@
 The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
-The input to the preprocessing phase is the graph weighted with lower bound weights $w_\ell$.
+The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with lower bound weights $w_\ell$.
 The source/origin $s$ and target/goal $t$ of the path and the actual edge weights $w_q$ are the input to the query phase.
-We require that $w_q(e)\ge w_\ell(e)$ for all edges $e$.
+We require that $w_q(e)\ge w_\ell(e)$ for all edges $e \in E$.
 In the query phase, the path is computed.
@@
 Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-We call our algorithm CH-Potentials, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, also called potentials in~\cite{gh-cspas-05}.
+We call our algorithm \emph{CH-Potentials}, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, also called potentials in~\cite{gh-cspas-05}.
 

### 2020-08-13 paper/ch_potentials.tex
change queue column label to fix tab 2 width

@@
 \centering
-\caption{Query performance with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
+\caption{Average query running times and number of queue pushs with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
 \input{table/building_blocks.tex}

### 2020-08-13 eval/applications.py

@@ -101,7 +101,7 @@ lines = table.to_latex(escape=False).split("\n")
 
 lines = lines[:2] + [
   R" & &   Running &                Queue &     Length & Dijkstra & Speedup \\"
-  R" & & time [ms] & pushs $[\cdot 10^3]$ & incr. [\%] &     [ms] &         \\"
+  R" & & time [ms] & $[\cdot 10^3]$ & incr. [\%] &     [ms] &         \\"
 ] + lines[4:13] + ["\\addlinespace"] + lines[13:]
 
 output = "\n".join(lines) + "\n"

### 2020-08-13 eval/building_blocks.py

@@ -42,7 +42,7 @@ lines = table.to_latex(escape=False).split("\n")
 
 lines = lines[:2] + [
     R" & & & & Running &     Queue \\",
-    R" Heur. & BCC & Deg2 & Deg3 & time [ms] & pushs [$\cdot 10^3$] \\"
+    R" Heur. & BCC & Deg2 & Deg3 & time [ms] & [$\cdot 10^3$] \\"
 # ] + lines[4:]
 ] + [lines[4], lines[17]] + lines[6:9] + ["\\addlinespace"] + lines[9:13] + ["\\addlinespace"] + [lines[13], lines[16]] + lines[20:]
 

### 2020-08-13 paper/ch_potentials.tex
minor title change, low hanging fruits for shortening

@@
 \pdfinfo{
-/Title (A* with Fast and Tight Heuristic in Road Networks)
+/Title (A Fast and Tight Heuristic for A* in Road Networks)
 /Author (Ben Strasser, Tim Zeitz)
@@
 }
-\title{A* with Fast and Tight Heuristic in Road Networks}
+\title{A Fast and Tight Heuristic for A* in Road Networks}
 \author{Ben Strasser\\
@@
 Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
-Dijkstra's algorithm iteratively explores nodes of the graph with ascending distance from the origin node until the target is reached.
+It iteratively explores nodes of the graph in ascending distance from the source node until the target is reached.
 A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
-It changes the order in which nodes are explored by using the distance from the origin plus an estimate of the remaining distance to the target as the ordering criterion.
+It changes the order in which nodes are explored by using the distance from the source plus an estimate of the remaining distance to the target as the ordering criterion.
 Figure~\ref{img:search-space} depicts an example A* search.
 While flexible, A*'s performance crucially depends on the quality of the employed estimating heuristic.
-Finding a good heuristic that can be evaluated efficiently is difficult.
+Finding good heuristics that can be evaluated efficiently is difficult.
 
@@
 The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with lower bound weights $w_\ell$.
-The source/origin $s$ and target/goal $t$ of the path and the actual edge weights $w_q$ are the input to the query phase.
+The source $s$ and target $t$ of and the actual edge weights $w_q$ are the input to the query phase in which shortest paths are computed.
 We require that $w_q(e)\ge w_\ell(e)$ for all edges $e \in E$.
-In the query phase, the path is computed.
 %
@@
 \caption{
-Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is mid node.
+Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is the mid node.
 }
@@
 It is not flexible, i.e., $w_q=w_\ell$.
-For a details, we refer the interested reader to \cite{gssv-erlrn-12,dsw-cch-15}.
+For a details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
 In this section, we give an introduction.
@@
 Similarly, the first node is the mid node of a \emph{down path}.
-Every up and down path is an up-down path.
+% Every up and down path is an up-down path.
 %
@@
 See~\cite{gssv-erlrn-12} for the details.
-After the preprocessing, in $G^+$ for pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path with the same length as a shortest path in $G$.
+After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
 See Figure~\ref{fig:ch} for a proof sketch.
-From every shortest path (red) in $G$, a up-down path of equal length in $G+$ (blue) can be constructed.
-We may therefore restrict our search to up-down paths in $G^+$.
-This search is bidirectional.
+From every shortest path (red) in $G$, an up-down path of equal length in $G+$ (blue) exists.
+Thus, we can restrict our search to up-down paths in $G^+$.
+The search is bidirectional.
 The forward search starts from $s$ and only follows up-edges.
@@
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
-On road networks, this is the case as demonstrated in~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.
+On road networks, this is the case~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.
 
@@
 The query phase is divided into two steps.
-The first step is analogously to the CH query:
+The first step is analogue to the CH query:
 From the target node $t$, all reachable nodes via reversed down-edges are explored.
@@
 Using PHAST, we can compute a tight A* heuristic in another way.
-In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the query edge weights $w_\ell$ and store the result in an array $H$.
-In the next step, we run $A^*$ and implement the heuristic function as lookup in the array $H$.
+In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the lower bound edge weights $w_\ell$ and store the result in an array $H$.
+Then, we run $A^*$ and implement the heuristic as a lookup in the array $H$.
 
-This PHAST-based algorithm works.
-The heuristic evaluation and by extension the $A^*$ search is indeed fast.
+% This PHAST-based algorithm works.
+The PHAST-based heuristic evaluation and by extension the $A^*$ search is indeed fast.
 However, the PHAST step before the search is comparatively expensive.
-The reason for this is that the distances from all nodes towards $t$ are computed.
+The reason is that the distances towards $t$ are computed for \emph{all} nodes.
 Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.
@@
 We do this by first recursively computing the potentials $\pi(u)$ of all nodes for which an up-edge $(v,u)$ exists.
-Afterwards, we compute using $\min_u\{w_\ell(v,u) + \pi(u)\}$ the minimum distance over all up-down paths that contain at least one up-edge.
-As the minimum distance up-down path does not necessarily have to contain an up-edge, we set $\pi(v) = \min \{ B[v], \min_u\{w_\ell(v,u) + \pi(u)\} \}$.
+Afterwards, we compute $d = \min_u\{w_\ell(v,u) + \pi(u)\}$ the minimum distance over all up-down paths that contain at least one up-edge.
+As the shortest up-down path does not necessarily have to contain an up-edge, we set $\pi(v) = \min \{ B[v], d \}$.
 This calculation is correct, as it computes the minimum up-down $vt$-path distance in $G^+$, which corresponds to the minimum $vt$-path distance in a CH.
@@
 We modify A* by processing low degree nodes consecutively without pushing them into the queue.
-Our algorithm makes use of the undirected degree $d(x)$ of a node $x$.
+Our algorithm uses the undirected degree $d(x)$ of a node $x$.
 Formally, $d(x)$ is the number of nodes $y$ such that $(x,y)\in E$ or $(y,x)\in E$.
@@
 Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
-The final path is the concatenation of all three path parts.
+The final path is the concatenation of all three paths.
 
@@
 It is by definition not smaller than the shortest $yq$-path length in $G$ with $w_\ell$.
-The shortest $yq$-path length is exactly what CH-Potentials compute.
+The shortest $yq$-path length is what CH-Potentials compute.
 The heuristic is therefore a lower bound.
@@
 The main difference is that we use CH-Potentials instead of landmarks to guide the A* search.
-We set $w_\ell(e)$ to the minimum travel time for every edge, i.e., $w_\ell(e) = \min_\tau w_q(e,\tau)$.
+We set $w_\ell(e) = \min_\tau w_q(e,\tau)$, that is the minimum travel time.
 
@@
 To support live traffic, we set $w_\ell$ to the travel time without any congestions.
-$w_q$ is set to the travel time accounting for the current traffic.
+$w_q$ is set to the travel time accounting for current traffic.
 As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
@@
 Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
-Further, $w_c(e)$ is the travel time according to the live traffic.
+Further, $w_c(e)$ is the travel time according to live traffic.
 Finally, we denote by $\tau_{\mathrm{soon}}$ the moment up to which we trust the live traffic.
@@
 For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
-Our modified travel time function linearly approaches the predicted travel time with a bounded slope.
+Our modified travel time function linearly approaches the predicted travel time. % with a bounded slope.
 
@@
 
-With this setup, we extended CH-Potentials to support a combination of live and predicted traffic.
-We did not make any modification to our algorithm, that would make a combination with tunnel and/or highway avoidance or turn-aware routing difficult.
+With this setup, CH-Potentials support a combination of live and predicted traffic.
+We did not make any modification to our algorithm, that would hinder a combination with tunnel and/or highway avoidance or turn-aware routing.
 This straight-forward integration of complex routing problems is the strength of the CH-Potential algorithm.
@@
 We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
-Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup of CH-Potentials over the baseline.
+Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup over this baseline.
 }\label{tab:applications}
@@
 In this paper, we introduced CH-Potentials, a fast and exact new A* heuristic for finding shortest paths in road networks.
-The approach is flexible and can handle a multitude of routing scenarios with very small implementation complexity.
+The approach is flexible and can handle a multitude of routing scenarios with very little implementation complexity.
 The performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.

### 2020-08-14 paper/ch_potentials.tex
updates to BCC part

@@
 
-In the preprocessing phase, before the CH is created, we compute the subgraph $G_C$, called \emph{core}, induced by the largested biconnected component of the undirected graph underlying $G$.
+In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}, induced by the largest biconnected component of the undirected graph underlying $G$.
 We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
@@
 We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
-The CH-Potentials preprocessing step is only executed for $G_C$.
 
-The query phase is divided into four steps.
-In the first step, we use Dijkstra's algorithm to explore the component that contains $s$ until $a_s$ is reached.
-If this search finds $t$, then $s$ and $t$ are part of the same component.
-A shortest path was found and therefore the algorithm stops early.
-Otherwise, in the second step starts.
-It consists of applying the CH-Potential algorithm restricted to $G_C$ from $a_s$ to $a_t$.
-Finally, in the third step, we search a path from $a_t$ towards $t$ using Dijkstra's algorithm restricted to $t$'s biconnected component.
-The final path is the concatenation of all three paths.
+The query phase is divided into two steps.
+In the first step, we apply A* with CH-Potentials to $G_C$ combined with the component that contains $s$.
+This can be achieved implicitly by removing edges from $G_C$ into other components during preprocessing.
+If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
+Otherwise we find $a_t$.
+In that case, we continue by searching a path from $a_t$ towards $t$ restricted to $t$'s biconnected component.
+The final result is the concatenation of both paths.
 

### 2020-08-14 paper/ch_potentials.tex
some shortening

@@
 They achieve speed and exactness but sacrifice A*'s flexibility.
-In this paper, we use such as Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
+In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
 We call our technique CH-Potentials.
@@
 Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
-It is usually formalized as a variant of the classical shortest path problem in weighted graphs.
+It is often formalized as a variant of the classical shortest path problem in weighted graphs.
 Developing flexible techniques to efficiently compute exact routes is a tough challenge.
@@
 \emph{Efficiently} means that the path computation is fast.
-Being able to change the edge weights between path computations or handle more complex edge weights (for example travel times which change over the course of the day obtained through traffic predictions) is \emph{flexibility}.
+Being able to change the edge weights between path computations or handle more complex weights (e.g. travel times depending on the time of the day) is \emph{flexibility}.
 
@@
 
-The resulting algorithm is the basic CH-potential algorithm.
+The resulting algorithm is the basic CH-Potentials algorithm.
 It is flexible and exact.
@@
 Many extended problems can therefore be solved using the CH-Potentials algorithm.
-In this section, we describe some extended problems.
+In this section, we describe some extended problems and how they can be solved with CH-Potentials.
 
@@
 However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
-Such restrictions are usually modeled using turn weights.
+Such restrictions can be modeled using turn weights~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
+% In this section, we first present the extended problem setting.
+% Afterwards, we describe how it can be solved with CH-Potentials.
 We extend CH-Potentials by using zero as lower bound for every turn weight in the heuristic.
-In this section, we first present the extended problem setting.
-Afterwards, we describe how it can be solved with CH-Potentials.
 
-In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
-$w_t$ maps a pair of incident edges onto the time required to perform the turn.
+% In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
+A \emph{turn weight} $w_t$ maps a pair of incident edges onto the time required to perform the turn.
 Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup \{+\infty\}$.
@@
 
-The simplest combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
-However, this implies computing a CH on the expanded graph.
-This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{gv-errnt-11}.
-We want to avoid these complications.
+% The simplest combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
+% However, this implies computing a CH on the expanded graph.
+% This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{gv-errnt-11}.
+% We want to avoid these complications.
 
-We therefore only apply the A* search to the expanded graph $G'$.
+% We therefore only apply the A* search to the expanded graph $G'$.
+We only apply the A* search to the expanded graph $G'$.
 The heuristic is computed with respect to the input graph $G$.
@@
 An edge weight is no longer a scalar value but a function that maps the entry time onto the traversal time.
-Performing this prediction is outside of the scope of this paper.
+% Performing this prediction is outside of the scope of this paper.
 It is common to refer to routing with predicted traffic as \emph{time-dependent routing}.
-Again, we first formalize the extended problem setting and then describe our solution with CH-Potentials.
+% Again, we first formalize the extended problem setting and then describe our solution with CH-Potentials.
 
@@
 With this setup, we extended CH-Potentials to support time-dependent routing.
-As we were able to keep all travel time functions out of the CH.
-This avoids a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-19}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
+As we were able to keep all travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-19}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
 
@@
 We did not make any modification to our algorithm, that would hinder a combination with tunnel and/or highway avoidance or turn-aware routing.
-This straight-forward integration of complex routing problems is the strength of the CH-Potential algorithm.
+This straight-forward integration of complex routing problems is the strength of the CH-Potentials algorithm.
 
@@
 All experiments were performed sequentially.
-Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile with the target-cpu=native option.
+Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile. % with the target-cpu=native option.
 
@@
 The graph has 16M nodes and 35M edges.
-For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com/}}.
+For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com}}.
 The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
@@
 Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
-The reason is that different queries (even with similar shortest path lengths) are affected differently by growing $\alpha$ depending on the geographic regions the shortest path traverses.
-As $\alpha$ grows, more and more nodes in the area around the shortest paths are explored.
-So when the shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, then when the shortest path goes only through sparse areas.
+The reason is that different queries (even with similar shortest path lengths) are affected differently by $\alpha$ depending on the geographic regions the shortest path traverses.
+As $\alpha$ grows, more nodes in the area around the shortest paths are explored.
+So when a shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when it passes sparse areas.
 % Conclusions?
@@
 In Table~\ref{tab:building_blocks} we investigate the impact of our optimizations.
-We run queries with different heuristic functions: the \emph{Zero} heuristic which turns A* back into the classical algorithm of Dijkstra, our new CH-Potential heuristic, and an \emph{Oracle} heuristic where all distances to $t$ with regards to $w_\ell$ can be obtained immediately.
+We run queries with different heuristic functions: the \emph{Zero} heuristic which turns A* back into the classical algorithm of Dijkstra, our new CH-Potentials heuristic, and an \emph{Oracle} heuristic where all distances to $t$ with regards to $w_\ell$ can be obtained immediately.
 We first observe that each one of our optimizations gains useful speedups regardless of the heuristic used.
@@
 The greatest speedup is, of course, achieved when applying CH-Potentials to an unmodified graph with $w_q = w_\ell$.
-In that case the performance is comparable to what other state-of-the-art route planning techniques like CH achieve.
-When considering extended problem settings, the performance of CH-Potentials strongly depends on the tightness of $w_\ell$, as previously observed in Figure~\ref{fig:scaled_weights}.
-On one hand, forbidding highways leads to over 40\% longer paths and the smallest speedup over Dijkstra (only 5.3) among all settings.
-On the other hand, incorporating turn restrictions increase path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity.
+In that case, the performance is within an order of magnitude to what other state-of-the-art techniques achieve.
+When considering extended problems, the performance of CH-Potentials strongly depends on the tightness of $w_\ell$.
+On one hand, forbidding highways leads to over 40\% longer paths and the smallest speedup (only 5.3) among all settings.
+On the other hand, incorporating turn restrictions increases path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity.
 Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 127\,ms on average.
 The Mapbox traffic predictions increase shortest path lengths by about 18\%.
-Combined with the additional work from function evaluations this leads to running times around 200\,ms.
+Combined with the additional work from function evaluations, this leads to running times around 200\,ms.
 The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to better running times and speedups.
-Assessing which of these predictions is more realistic is not possible within the scope of this paper.
+% Assessing which of these predictions is more realistic is not possible within the scope of this paper.
 Combining Mapbox live traffic and traffic predictions yields running times only slightly slower than traffic predictions alone.
-To the best of our knowledge, no other exact technique so far could combine these two scenarios.
+% To the best of our knowledge, no other exact technique so far could combine these two scenarios.
 Additionally adding turn restrictions slows running times down to around 510\,ms despite the length increase only growing by 1.3 percentage points.
+The reason for this is that the turn expanded graph is three times bigger and that the BCC optimization is not applicable for turn-expanded graphs.
 This is still a speedup of 12.2 over the baseline.
-The reason for this is that the turn expanded graph is three times bigger and that the BCC optimization is not applicable for turn-expanded graphs.
-To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.
+% To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.
 

### 2020-08-14 eval/scaled_weights.py
shrink boxplot

@@ -22,8 +22,10 @@ queries = pd.DataFrame.from_records([{
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 
-plt.figure(figsize=(10.5,3))
+plt.figure(figsize=(6,3))
 g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8, showmeans=True)
+g.xaxis.set_major_locator(mpl.ticker.MultipleLocator(5))
+# g.xaxis.set_major_formatter(mpl.ticker.ScalarFormatter())
 g.set_yscale('log')
 g.set_ylabel('Running Time [ms]')
 g.set_xlabel('Weight Factor $\\alpha$')

### 2020-08-21 paper/Formatting-Instructions-LaTeX-2021.tex
stuff

--- 2020-07-27 paper/formatting-instructions-latex-2020.tex
@@
-\def\year{2020}\relax
-%File: formatting-instruction.tex
+\def\year{2021}\relax
+%File: formatting-instructions-latex-2021.tex
+%release 2021.1
 \documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage{aaai20}  % DO NOT CHANGE THIS
+\usepackage{aaai21}  % DO NOT CHANGE THIS
 \usepackage{times}  % DO NOT CHANGE THIS
@@
 \def\UrlFont{\rm}  % DO NOT CHANGE THIS
-\usepackage{graphicx}  % DO NOT CHANGE THIS
+\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
+\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
 \frenchspacing  % DO NOT CHANGE THIS
@@
 /Title (AAAI Press Formatting Instructions for Authors Using LaTeX -- A Guide)
-/Author (AAAI Press Staff, Pater Patel Schneider, Sunil Issar, J. Scott Penberthy, George Ferguson, Hans Guesgen)
+/Author (AAAI Press Staff, Pater Patel Schneider, Sunil Issar, J. Scott Penberthy, George Ferguson, Hans Guesgen, Francisco Cruz, Marc Pujol-Gonzalez)
+/TemplateVersion (2021.1)
 } %Leave this
@@
 % \usepackage{balance} -- This package is specifically forbidden
-% \usepackage{caption} -- This package is specifically forbidden
 % \usepackage{color (if used in text)
@@
 % \nameref} -- This package is specifically forbidden
-% \natbib} -- This package is specifically forbidden -- use the following workaround:
 % \usepackage{savetrees} -- This package is specifically forbidden
@@
 
-% The file aaai20.sty is the style file for AAAI Press 
+% The file aaai21.sty is the style file for AAAI Press
 % proceedings, working notes, and technical reports.
 %
-\setlength\titlebox{2.5in} % If your paper contains an overfull \vbox too high warning at the beginning of the document, use this
-% command to correct it. You may not alter the value below 2.5 in
-\title{AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide }
-%Your title must be in mixed case, not sentence case. 
+
+% Title
+
+% Your title must be in mixed case, not sentence case.
 % That means all verbs (including short verbs like be, is, using,and go),
@@
 % directly follow a colon or long dash
-\author{Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{Primarily Mike Hamilton of the Live Oak Press, LLC, with help from the AAAI Publications Committee}\\ \Large \textbf{AAAI Style Contributions by
-Pater Patel Schneider,} \\ \Large \textbf{Sunil Issar, J.
-Scott Penberthy, George Ferguson, Hans Guesgen}\\ % All authors must be in the same font size and format.
-Use \Large and \textbf to achieve this result when breaking a line
-\textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\ %If you have multiple authors and multiple affiliations
+
+\title{AAAI Press Formatting Instructions \\for Authors Using \LaTeX{} --- A Guide }
+\author{
+
+    %Authors
+    % All authors must be in the same font size and format.
+    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
+    AAAI Style Contributions by Pater Patel Schneider,
+    Sunil Issar,  \\
+    J. Scott Penberthy,
+    George Ferguson,
+    Hans Guesgen,
+    Francisco Cruz,
+    Marc Pujol-Gonzalez
+    \\
+}
+\affiliations{
+    %Afiliations
+
+    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
+    %If you have multiple authors and multiple affiliations
     % use superscripts in text and roman font to identify them.
-For example, Sunil Issar,\textsuperscript{\rm 2} J.
-Scott Penberthy\textsuperscript{\rm 3} George Ferguson,\textsuperscript{\rm 4} Hans Guesgen\textsuperscript{\rm 5}.
-Note that the comma should be placed BEFORE the superscript for optimum readability
+    %For example,
+
+    % Sunil Issar, \textsuperscript{\rm 2}
+    % J. Scott Penberthy, \textsuperscript{\rm 3}
+    % George Ferguson,\textsuperscript{\rm 4}
+    % Hans Guesgen, \textsuperscript{\rm 5}.
+    % Note that the comma should be placed BEFORE the superscript for optimum readability
+
     2275 East Bayshore Road, Suite 160\\
     Palo Alto, California 94303\\
-publications20@aaai.org % email address must be in roman text type, not monospace or sans serif
+    % email address must be in roman text type, not monospace or sans serif
+    publications21@aaai.org
+
+    % See more examples next
 }
+\iffalse
+%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
+\title{My Publication Title --- Single Author}
+\author {
+    % Author
+    Author Name \\
+}
+
+\affiliations{
+    Affiliation \\
+    Affiliation Line 2 \\
+    name@example.com
+}
+\fi
+
+\iffalse
+%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
+\title{My Publication Title --- Multiple Authors}
+\author {
+    % Authors
+
+        First Author Name,\textsuperscript{\rm 1}
+        Second Author Name, \textsuperscript{\rm 2}
+        Third Author Name \textsuperscript{\rm 1} \\
+}
+\affiliations {
+    % Affiliations
+    \textsuperscript{\rm 1} Affiliation 1 \\
+    \textsuperscript{\rm 2} Affiliation 2 \\
+    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
+}
+\fi
 \begin{document}
@@
 \begin{itemize}
-\item You must use the 2020 AAAI Press \LaTeX{} style file and the aaai.bst bibliography style file, which are located in the 2020 AAAI Author Kit (aaai20.sty and aaai.bst).
+\item You must use the 2021 AAAI Press \LaTeX{} style file and the aaai21.bst bibliography style files, which are located in the 2021 AAAI Author Kit (aaai21.sty, aaai21.bst).
 \item You must complete, sign, and return by the deadline the AAAI copyright form (unless directed by AAAI Press to use the AAAI Distribution License instead).
@@
 All papers submitted for publication by AAAI Press must be accompanied by a valid signed copyright form.
-There are no exceptions to this requirement.
-You must send us the original version of this form.
-However, to meet the deadline, you may fax (1-650-321-4457) or scan and e-mail the form (pubforms20@aaai.org) to AAAI by the submission deadline, and then mail the original via postal mail to the AAAI office.
-If you fail to send in a signed copyright or permission form, we will be unable to publish your paper.
-There are \textbf{no exceptions} to this policy.You will find PDF versions of the AAAI copyright and permission to distribute forms in the AAAI AuthorKit.
+They must also contain the AAAI copyright notice at the bottom of the first page of the paper.
+There are no exceptions to these requirements.
+If you fail to provide us with a signed copyright form or disable the copyright notice, we will be unable to publish your paper.
+There are \textbf{no exceptions} to this policy.
+You will find a PDF version of the AAAI copyright form in the AAAI AuthorKit.
+Please see the specific instructions for your conference for submission details.
 
@@
 We need source and PDF files that can be used in a variety of ways and can be output on a variety of devices.
-The design and appearance of the paper is strictly governed by the aaai style file (aaai20.sty).
+The design and appearance of the paper is strictly governed by the aaai style file (aaai21.sty).
 \textbf{You must not make any changes to the aaai style file, nor use any commands, packages, style files, or macros within your own paper that alter that design, including, but not limited to spacing, floats, margins, fonts, font size, and appearance.} AAAI imposes requirements on your source and PDF files that must be followed.
@@
 \begin{itemize}
-\item Your .tex file must compile in PDF\LaTeX{} --- ( you may not include .ps or .eps figure files.)
+\item Your .tex file must compile in PDF\LaTeX{} --- (you may not include .ps or .eps figure files.)
 \item All fonts must be embedded in the PDF file --- including includes your figures.
@@
 \item You may not alter the spacing above and below captions, figures, headings, and subheadings.
-\item You may not alter the font sizes of text elements, footnotes, heading elements, captions, or title information (for references and mathematics, please see the the limited exceptions provided herein).
+\item You may not alter the font sizes of text elements, footnotes, heading elements, captions, or title information (for references and mathematics, please see the limited exceptions provided herein).
 \item You may not alter the line spacing of text.
 \item Your title must follow Title Case capitalization rules (not sentence case).
-\item Your .tex file must include completed metadata to pass-through to the PDF (see PDFINFO below)
+\item Your .tex file must include completed metadata to pass-through to the PDF (see PDFINFO below).
 \item \LaTeX{} documents must use the Times or Nimbus font package (you may not use Computer Modern for the text of your paper).
@@
 
-If you do not follow these requirements, you will be required to correct the deficiencies and resubmit the paper. A resubmission fee will apply.
+If you do not follow these requirements, your paper will be returned to you to correct the deficiencies.
 
@@
 \item The bibliography (.bib) file(s).
-\item Your source must compile on our system, which includes only standard \LaTeX{} 2019 TeXLive support files. 
+\item Your source must compile on our system, which includes only standard \LaTeX{} 2020 TeXLive support files.
 \item Only the graphics files used in compiling paper.
@@
 
-Your \LaTeX{} source will be reviewed and recompiled on our system (if it does not compile, you will be required to resubmit, which will incur fees).
-\textbf{Do not submit your source in multiple text files.} Your single \LaTeX{} source file must include all your text, your bibliography (formatted using aaai.bst), and any custom macros.
+Your \LaTeX{} source will be reviewed and recompiled on our system (if it does not compile, your paper will be returned to you.
+\textbf{Do not submit your source in multiple text files.} Your single \LaTeX{} source file must include all your text, your bibliography (formatted using aaai21.bst), and any custom macros.
 
@@
 Please be certain that you are not compiling your paper using old or obsolete style files.
+
 \textbf{Final Archive.} Place your PDF and source files in a single archive which should be compressed using .zip. The final file size may not exceed 10 MB.
@@
 
-If (and only if) your author title information will not fit within the specified height allowed, put \textbackslash setlength \textbackslash titlebox{2.5in} in your preamble.
-Increase the height until the height error disappears from your log.
-You may not use the \textbackslash setlength command elsewhere in your paper, and it may not be used to reduce the height of the author-title box.
-
 \subsubsection{The Following Must Appear in Your Preamble}
@@
 \begin{scriptsize}\begin{verbatim}
-\documentclass[letterpaper]{article}
-\usepackage{aaai20}
-\usepackage{times}
-\usepackage{helvet}
-\usepackage{courier}
-\usepackage[hyphens]{url} 
-\usepackage{graphicx}
-\urlstyle{rm}
-\def\UrlFont{\rm}
-\usepackage{graphicx}
-\frenchspacing
-\setlength{\pdfpagewidth}{8.5in}
-\setlength{\pdfpageheight}{11in}
-% Add additional packages here, but check 
-% the list of disallowed packages 
-% (including, but not limited to
-% authblk, caption, CJK, float, fullpage, geometry, 
-% hyperref, layout, nameref, natbib, savetrees, 
-% setspace, titlesec, tocbibind, ulem)
-% and illegal commands provided in the 
-% common formatting errors document
-% included in the  Author Kit before doing so. 
+\def\year{2021}\relax
+% File: formatting-instruction.tex
+\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
+\usepackage{aaai21} % DO NOT CHANGE THIS
+\usepackage{times} % DO NOT CHANGE THIS
+\usepackage{helvet} % DO NOT CHANGE THIS
+\usepackage{courier} % DO NOT CHANGE THIS
+\usepackage[hyphens]{url} % DO NOT CHANGE THIS
+\usepackage{graphicx} % DO NOT CHANGE THIS
+\urlstyle{rm} % DO NOT CHANGE THIS
+\def\UrlFont{\rm} % DO NOT CHANGE THIS
+\usepackage{graphicx}  % DO NOT CHANGE THIS
+\usepackage{natbib} % DO NOT CHANGE THIS OR ADD OPTIONS
+\usepackage{caption} % DO NOT CHANGE THIS OR ADD OPTIONS
+\frenchspacing % DO NOT CHANGE THIS
+\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
+\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
 %
-% PDFINFO
-% You are required to complete the following
-% for pass-through to the PDF. 
-% No LaTeX commands of any kind may be
-% entered. The parentheses and spaces 
-% are an integral part of the 
-% pdfinfo script and must not be removed.
-%
+% PDF Info Is REQUIRED.
+% For /Author, add all authors within the parentheses,
+% separated by commas. No accents or commands.
+% For /Title, add Title in Mixed Case.
+% No accents or commands. Retain the parentheses.
 \pdfinfo{
-/Title (Type Your Paper Title Here in Mixed Case)
-/Author (John Doe, Jane Doe)
-/Keywords (Input your keywords in this optional area)
+/Title (AAAI Press Formatting Instructions for Authors
+Using LaTeX -- A Guide)
+/Author (AAAI Press Staff, Pater Patel Schneider,
+Sunil Issar, J. Scott Penberthy, George Ferguson,
+Hans Guesgen, Francisco Cruz, Marc Pujol-Gonzalez)
+/TemplateVersion (2021.1)
 }
-%
-% Section Numbers
-% Uncomment if you want to use section numbers
-% and change the 0 to a 1 or 2
-% \setcounter{secnumdepth}{0}
-
-% Title and Author Information Must Immediately Follow
-% the pdfinfo within the preamble
-%
-\title{Title}\\
-\author\{Author 1 \ and Author 2\\
-Address line\\
-Address line\\
-\ And\\
-Author 3\\
-Address line\\
-Address line
-}\\
 \end{verbatim}\end{scriptsize}
@@
 
-\subsubsection{The Following Must Conclude Your Document}
+\noindent You should then continue with the body of your paper. Your paper must conclude with the references, which should be inserted as follows:
 \begin{quote}
@@
 \bibliography{Bibliography-File}
-\bibliographystyle{aaai}
 \end{document}
@@
 \bibliography{Bibliography-File}\\
-\bibliographystyle{aaai}\\
 \end{document}\\
@@
 \centering
-\caption{Commands that must not be used}\smallskip
+
 \begin{tabular}{l|l|l|l}
@@
 %}
+\caption{Commands that must not be used}
 \label{table1}
@@
 \begin{table}[t]
-\caption{LaTeX style packages that must not be used.}\smallskip
 \centering
 %\resizebox{.95\columnwidth}{!}{
-\smallskip\begin{tabular}{l|l|l|l}
-authblk & babel & caption & cjk\\
-dvips & epsf & epsfig & euler\\
-float & fullpage & geometry & graphics\\
-hyperref & layout & linespread & lmodern\\
-maltepaper & natbib & navigator & pdfcomment\\
-pgfplots & psfig & pstricks & t1enc \\
-titlesec & tocbind & ulem\\
+\begin{tabular}{l|l|l|l}
+    authblk & babel & cjk & dvips \\
+    epsf & epsfig & euler & float \\
+    fullpage & geometry & graphics & hyperref \\
+    layout & linespread & lmodern & maltepaper \\
+    navigator & pdfcomment & pgfplots & psfig \\
+    pstricks & t1enc & titlesec & tocbind \\
+    ulem
 \end{tabular}
+\caption{LaTeX style packages that must not be used.}
 \label{table2}
@@
 
-There are a number of packages, commands, scripts, and macros that are incompatable with aaai20.sty.
+There are a number of packages, commands, scripts, and macros that are incompatable with aaai21.sty.
 The common ones are listed in tables \ref{table1} and \ref{table2}.
@@
 However, because we require that your electronic paper be formatted in US letter size, the preamble we have provided includes commands that alter the default to US letter size.
-Please note that using any other package to alter page size (such as, but not limited to the Geometry package) will result in your final paper being returned to you for correction and payment of a resubmission fee.
+Please note that using any other package to alter page size (such as, but not limited to the Geometry package) will result in your final paper being returned to you for correction.
 
@@
 Each column should be 3.3 inches wide (slightly more than 3.25 inches), with a .375 inch (.952 cm) gutter of white space between the two columns.
-The aaai20.sty file will automatically create these columns for you.
+The aaai21.sty file will automatically create these columns for you.
 
 \subsection{Overlength Papers}
-If your paper is too long, turn on \textbackslash frenchspacing, which will reduce the space after periods.
-Next, shrink the size of your graphics.
-Use \textbackslash centering instead of \textbackslash begin\{center\} in your figure environment.
+If your paper is too long and you resort to formatting tricks to make it fit, it is quite likely that it will be returned to you.
+The best way to retain readability if the paper is overlength is to cut text, figures, or tables.
+There are, a few acceptable ways to reduce paper size that don't affect readability.
+First, turn on \textbackslash frenchspacing, which will reduce the space after periods.
+Next, move all your figures and tables to the top of the page.
+Consider removing less important portions of a figure.
+If you use \textbackslash centering instead of \textbackslash begin\{center\} in your figure environment, you can also buy some space.
 For mathematical environments, you may reduce fontsize {\bf but not below 6.5 point}.
-You may also alter the size of your bibliography and tables by inserting \textbackslash fontsize\{9.5pt\}\{10.5pt\} \textbackslash selectfont
-right before the bibliography (the minimum size is \textbackslash fontsize\{9.0pt\}\{10.0pt\}.
-You may not use any commands that further reduce point size.
-Be careful when scaling tables and figures.
-Tables that do not fit in a single column, even after the font size reduction to 9 point, must be placed across double columns.
-If your table won't fit within the margins even in double columns, you must split it.
-We will not accept single (or double) column tables that have been scaled to anything below 9 point.
 
 Commands that alter page layout are forbidden.
-These include \textbackslash columnsep, \textbackslash topmargin, \textbackslash topskip, \textbackslash textheight, \textbackslash textwidth, \textbackslash oddsidemargin, and \textbackslash evensizemargin (this list is not exhaustive).
-If you alter page layout, you will be required to pay the page fee \textit{plus} a reformatting fee.
+These include \textbackslash columnsep,  \textbackslash float, \textbackslash topmargin, \textbackslash topskip, \textbackslash textheight, \textbackslash textwidth, \textbackslash oddsidemargin, and \textbackslash evensizemargin (this list is not exhaustive).
+If you alter page layout, you will be required to pay the page fee.
 Other commands that are questionable and may cause your paper to be rejected include \textbackslash parindent, and \textbackslash parskip.
@@
 Options for reducing the length of a paper include reducing the size of your graphics, cutting text, or paying the extra page charge (if it is offered).
-
-\subsection{Figures}
-Your paper must compile in PDF\LaTeX{}. Consequently, all your figures must be .jpg, .png, or .pdf. You may not use the .gif (the resolution is too low), .ps, or .eps file format for your figures.
-
-If you normally create your figures using pgfplots, please create the figures first, and then import them as pdfs with proper bounding boxes, as the bounding and trim boxes created by pfgplots are fragile, and not valid.
-
-When you include your figures, you must crop them \textbf{outside} of \LaTeX{}.
-The command \textbackslash includegraphics*[clip=true, viewport 0 0 10 10]{...} might result in a PDF that looks great, but the image is \textbf{not really cropped.} The full image can reappear (and obscure whatever it is overlapping) when page numbers are applied or color space is standardized.
-Figures \ref{fig1}, and \ref{fig2} display some unwanted results that often occur.
-
-Do not use minipage to group figures.
-Additionally, the font and size of figure captions must be 10 point roman.
-Do not make them smaller, bold, or italic.
-(Individual words may be italicized if the context requires differentiation.)
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=0.9\columnwidth]{figure1} % Reduce the figure size so that it is slightly narrower than the column.
-Don't use precise values for figure width.This setup will avoid overfull boxes.
-\caption{Using the trim and clip commands produces fragile layers that can result in disasters (like this one from an actual paper) when the color space is corrected or the PDF combined with others for the final proceedings.
-Crop your figures properly in a graphics program -- not in LaTeX}
-\label{fig1}
-\end{figure}
-
-\begin{figure*}[t]
-\centering
-\includegraphics[width=0.8\textwidth]{figure2} % Reduce the figure size so that it is slightly narrower than the column.
-\caption{Adjusting the bounding box instead of actually removing the unwanted data resulted multiple layers in this paper.
-It also needlessly increased the PDF size.
-In this case, the size of the unwanted layer doubled the paper's size, and produced the following surprising results in final production.
-Crop your figures properly in a graphics program.
-Don't just alter the bounding box.}
-\label{fig2}
-\end{figure*}
-
-% Using the \centering command instead of \begin{center} ... \end{center} will save space
-% Positioning your figure at the top of the page will save space and make the paper more readable
-% Using 0.95\columnwidth in conjunction with the
 
@@
 Most print at 300 dpi even if the printer resolution is 1200 dpi or higher.
-They also often cause high resolution imagesetter devices and our PDF indexing software to crash.
+They also often cause high resolution imagesetter devices to crash.
 Consequently, AAAI will not accept electronic files containing obsolete type 3 fonts.
 Files containing those fonts (even in graphics) will be rejected.
+(Authors using blackboard symbols must be avoid those packages that use type 3 fonts.)
 
@@
 
-\subsubsection{Nonroman Fonts}
+\subsubsection{Nonroman Fonts.}
 If your paper includes symbols in other languages (such as, but not limited to, Arabic, Chinese, Hebrew, Japanese, Thai, Russian and other Cyrillic languages), you must restrict their use to bit-mapped figures.
@@
 Your title must appear in mixed case (nouns, pronouns, and verbs are capitalized) near the top of the first page, centered over both columns in sixteen-point bold type (twenty-four point leading).
-This style is called ``mixed case," which means that means all verbs (including short verbs like be, is, using,and go), nouns, adverbs, adjectives, and pronouns should be capitalized, (including both words in hyphenated terms), while articles, conjunctions, and prepositions are lower case unless they directly follow a colon or long dash.
+This style is called ``mixed case," which means that means all verbs (including short verbs like be, is, using, and go), nouns, adverbs, adjectives, and pronouns should be capitalized, (including both words in hyphenated terms), while articles, conjunctions, and prepositions are lower case unless they directly follow a colon or long dash.
 Author's names should appear below the title of the paper, centered in twelve-point type (with fifteen point leading), along with affiliation(s) and complete address(es) (including electronic mail address if available) in nine-point roman type (the twelve point leading).
@@
 
-\subsubsection{Formatting Author Information}
-Author information can be set in a number of different styles, depending on the number of authors and the number of affiliations you need to display.
-In formatting your author information, however, you may not use a table nor may you employ the \textbackslash authorblk.sty package.
-For several authors from the same institution, please just separate with commas:
+\subsubsection{Formatting Author Information.}
+Author information has to be set according the following specification depending if you have one or more than one affiliation.
+You may not use a table nor may you employ the \textbackslash authorblk.sty package.
+For one or several authors from the same institution, please just separate with commas and write the affiliation directly below using the macros \textbackslash author and \textbackslash affiliations:
 
 \begin{quote}\begin{scriptsize}\begin{verbatim}
-\author{Author 1, ... Author n\\ 
-Address line \\ ... \\ Address line}
-\end{verbatim}\end{scriptsize}\end{quote}
-
-\noindent If the names do not fit well on one line use:
-
-\begin{quote}\begin{scriptsize}\begin{verbatim}
-\author{Author 1} ... \\
-{\bf \Large Author ... Author}\\ 
-Address line \\ ... \\ Address line
+\author{
+    Author 1, ..., Author n \\
+}
+\affiliations {
+    Address line \\
+    ... \\
+    Address line
 }
@@
 
-\noindent For two (or three) authors from different institutions, use \textbackslash And:
+\noindent For authors from different institutions, use \textbackslash textsuperscript \{\textbackslash rm x \} to match authors and affiliations.
 
 \begin{quote}\begin{scriptsize}\begin{verbatim}
-\author{Author 1\\ Address line \\ ... \\ Address line 
-\And ... \And Author n\\
-Address line\\ ... \\ Address line}
-\end{verbatim}\end{scriptsize}\end{quote}
-
-\noindent To start a separate ``row" of authors, use \textbackslash AND:
-
-If the title and author information does not fit in the area allocated, place
-\textbackslash setlength\textbackslash titlebox\{\textit{height}\}
-after the \textbackslash documentclass line where \{\textit{height}\} is 2.5in or greater.
-(This one of the only allowable uses of the setlength command.
-Check with AAAI Press before using it elsewhere.)
-
-\subsubsection{Formatting Author Information --- Alternative Method}
-If your paper has a large number of authors from different institutions, you may use the following alternative method for displaying the author information.
-
-\begin{quote}\begin{scriptsize}\begin{verbatim}
-\author{AuthorOne},\textsuperscript{\rm 1} 
+\author{
+    AuthorOne,\textsuperscript{\rm 1}
     AuthorTwo,\textsuperscript{\rm 2}
     AuthorThree,\textsuperscript{\rm 3}
-\\bf \Large AuthorFour,\textsuperscript{\rm 4} 
-AuthorFive, \textsuperscript{\rm 5}\\
-\textsuperscript{\rm 1}AffiliationOne, 
-\textsuperscript{\rm 2}AffiliationTwo, 
-\textsuperscript{\rm 3}AffiliationThree\\
-\textsuperscript{\rm 4}AffiliationFour, 
-\textsuperscript{\rm 5}AffiliationFive 
+    AuthorFour,\textsuperscript{\rm 4}
+    AuthorFive \textsuperscript{\rm 5}}\\
+}
+\affiliations {
+    \textsuperscript{\rm 1}AffiliationOne, \\
+    \textsuperscript{\rm 2}AffiliationTwo, \\
+    \textsuperscript{\rm 3}AffiliationThree, \\
+    \textsuperscript{\rm 4}AffiliationFour, \\
+    \textsuperscript{\rm 5}AffiliationFive \\
     \{email, email\}@affiliation.com,
@@
     email@affiliation.com
+}
 \end{verbatim}\end{scriptsize}\end{quote}
 
-Note that you should break the author list before it extends into the right column margin.
-Put a line break, followed by   \textbackslash bf  \textbackslash Large to put the second line of authors in the same font and size as the first line (you may not make authors names smaller to save space.) Affiliations can be broken with a simple line break (\textbackslash  \textbackslash).
+Note that you may want to  break the author list for better visualization. You can achieve this using a simple line break (\textbackslash  \textbackslash).
+
 \subsection{\LaTeX{} Copyright Notice}
-The copyright notice automatically appears if you use aaai20.sty. It has been hardcoded and may not be disabled.
+The copyright notice automatically appears if you use aaai21.sty. It has been hardcoded and may not be disabled.
 
@@
 An overabundance of headings will tend to make your paper look more like an outline than a paper.
-The aaai.sty package will create headings for you.
+The aaai21.sty package will create headings for you.
 Do not alter their size nor their spacing above or below.
 
-\subsubsection{Section Numbers}
+\subsubsection{Section Numbers.}
 The use of section numbers in AAAI Press papers is optional.
@@
 
-\subsubsection{References}
+\subsubsection{References.}
 The references section should be labeled ``References" and should appear at the very end of the paper (don't end the paper with references, and then put a figure by itself on the last page).
@@
 \subsection{Illustrations and  Figures}
-Figures, drawings, tables, and photographs should be placed throughout the paper near the place where they are first discussed.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=0.9\columnwidth]{figure1} % Reduce the figure size so that it is slightly narrower than the column.
+Don't use precise values for figure width.This setup will avoid overfull boxes.
+\caption{Using the trim and clip commands produces fragile layers that can result in disasters (like this one from an actual paper) when the color space is corrected or the PDF combined with others for the final proceedings.
+Crop your figures properly in a graphics program -- not in LaTeX}.
+\label{fig1}
+\end{figure}
+
+\begin{figure*}[t]
+\centering
+\includegraphics[width=0.8\textwidth]{figure2} % Reduce the figure size so that it is slightly narrower than the column.
+\caption{Adjusting the bounding box instead of actually removing the unwanted data resulted multiple layers in this paper.
+It also needlessly increased the PDF size.
+In this case, the size of the unwanted layer doubled the paper's size, and produced the following surprising results in final production.
+Crop your figures properly in a graphics program.
+Don't just alter the bounding box.}
+\label{fig2}
+\end{figure*}
+
+% Using the \centering command instead of \begin{center} ... \end{center} will save space
+% Positioning your figure at the top of the page will save space and make the paper more readable
+% Using 0.95\columnwidth in conjunction with the
+
+Your paper must compile in PDF\LaTeX{}. Consequently, all your figures must be .jpg, .png, or .pdf. You may not use the .gif (the resolution is too low), .ps, or .eps file format for your figures.
+
+Figures, drawings, tables, and photographs should be placed throughout the paper on the page (or the subsequent page) where they are first discussed.
 Do not group them together at the end of the paper.
-If placed at the top or bottom of the paper, illustrations may run across both columns.
+If placed at the top of the paper, illustrations may run across both columns.
 Figures must not invade the top, bottom, or side margin areas.
@@
 Number figures sequentially, for example, figure 1, and so on.
+Do not use minipage to group figures.
 
-The illustration number and caption should appear under the illustration. Labels, and other text with the actual illustration must be at least nine-point type.
+If you normally create your figures using pgfplots, please create the figures first, and then import them as pdfs with proper bounding boxes, as the bounding and trim boxes created by pfgplots are fragile and not valid.
+
+When you include your figures, you must crop them \textbf{outside} of \LaTeX{}.
+The command \textbackslash includegraphics*[clip=true, viewport 0 0 10 10]{...} might result in a PDF that looks great, but the image is \textbf{not really cropped.} The full image can reappear (and obscure whatever it is overlapping) when page numbers are applied or color space is standardized.
+Figures \ref{fig1}, and \ref{fig2} display some unwanted results that often occur.
 
@@
 You will need to convert your ps files to PDF however.
+
+\subsubsection {Figure Captions.}The illustration number and caption must appear \textit{under} the illustration.
+Labels, and other text with the actual illustration must be at least nine-point type.
+However, the font and size of figure captions must be 10 point roman.
+Do not make them smaller, bold, or italic.
+(Individual words may be italicized if the context requires differentiation.)
+
+\subsection{Tables}
+
+Tables should be presented in 10 point roman type.
+If necessary, they may be altered to 9 point type.
+You may not use any commands that further reduce point size below nine points.
+Tables that do not fit in a single column must be placed across double columns.
+If your table won't fit within the margins even when spanning both columns, you must split it.
+Do not use minipage to group tables.
+
+\subsubsection {Table Captions.} The number and caption for your table must appear \textit{under} (not above) the table.
+Additionally, the font and size of table captions must be 10 point roman and must be placed beneath the figure.
+Do not make them smaller, bold, or italic.
+(Individual words may be italicized if the context requires differentiation.)
 
@@
 \LaTeX{} users please beware: \LaTeX{} will sometimes put portions of the figure or table or an equation in the margin.
-If this happens, you need to scale the figure or table down, or reformat the equation.{ \bf Check your log file!} You must fix any overflow into the margin (that means no overfull boxes in \LaTeX{}).
+If this happens, you need to make the figure or table span both columns.
+If absolutely necessary, you may reduce the figure, or reformat the equation, or reconfigure the table.{ \bf Check your log file!} You must fix any overflow into the margin (that means no overfull boxes in \LaTeX{}).
 \textbf{Nothing is permitted to intrude into the margin or gutter.}
-
-The most efficient and trouble-free way to fix overfull boxes in graphics is with the following command:
-
-\begin{quote}
-\begin{small}
-\textbackslash resizebox\{.95\textbackslash columnwidth\}{!}\{
-\}
-\end{small}\end{quote}
-
-Be certain, however, that your figures remain legible without magnification.
 
@@
 It may never be used for any portion of the text of your paper.
-The archival version of your paper will be printed in black and white and grayscale.The web version must be readable by persons with disabilities.
+The archival version of your paper will be printed in black and white and grayscale.
+The web version must be readable by persons with disabilities.
 Consequently, because conversion to grayscale can cause undesirable effects (red changes to black, yellow can disappear, and so forth), we strongly suggest you avoid placing color figures in your document.
@@
 
-\subsubsection{Fonts in Your Illustrations}
+\subsubsection{Fonts in Your Illustrations.}
 You must embed all fonts in your graphics before including them in your LaTeX document.
@@
 These definitions make the bibliography style fairly close to the one specified below.
-To use these definitions, you also need the BibTeX style file ``aaai.bst," available in the AAAI Author Kit on the AAAI web site.
+To use these definitions, you also need the BibTeX style file ``aaai21.bst," available in the AAAI Author Kit on the AAAI web site.
 Then, at the end of your paper but before \textbackslash end{document}, you need to put the following lines:
@@
 \begin{small}
-\textbackslash bibliographystyle\{aaai\}
 \textbackslash bibliography\{bibfile1,bibfile2,...\}
@@
 
-Please note that you are required to use \textbackslash bibliographystyle\{aaai\} for your references.
-You may not use named, plain, apalike, acm, ieeetr, siam, chicago, or any other style.
-Use of natbib is also not acceptable.
-(In addition to natbib, the aaai20.sty file is also incompatible with the hyperref and navigator packages.
-If you use either, your references will be garbled and your paper will be returned to you.) If you used natbib commands, an imprecise workaround is available (although it does not always work).
-You may put the following in your preamble (after removing \textbackslash usepackage\{natbib\}
-
-\begin{quote}\small
-\textbackslash newcommand\{\textbackslash citet\}[1]\{\textbackslash citeauthor\{\#1\}~\textbackslash shortcite\{\#1\}\}
-\textbackslash newcommand\{\textbackslash citep\}\{\textbackslash cite\}
-\textbackslash newcommand\{\textbackslash citealp\}[1]\{\textbackslash citeauthor\{\#1\}~\textbackslash citeyear\{\#1\}\}
-\end{quote}
+Please note that the aaai21.sty class already sets the bibliographystyle for you, so you do not have to place any \textbackslash bibliographystyle command in the document yourselves.
+The aaai21.sty file is incompatible with the hyperref and navigator packages.
+If you use either, your references will be garbled and your paper will be returned to you.
 
@@
 In the past, AAAI has corrected improperly formatted files submitted by the authors.
-Unfortunately, this has become an increasingly burdensome expense that we can no longer absorb (we are charged double for papers that require reformatting).
-Consequently, if your file is improperly formatted, it will probably be returned to you by the outside Production agency.
-If that happens, you will be required to fix your file and pay a resubmission fee.
+Unfortunately, this has become an increasingly burdensome expense that we can no longer absorb).
+Consequently, if your file is improperly formatted, it will be returned to you for correction.
 
@@
 \section{ Acknowledgments}
-AAAI is especially grateful to Peter Patel Schneider for his work in implementing the aaai.sty file, liberally using the ideas of other style hackers, including Barbara Beeton.
+AAAI is especially grateful to Peter Patel Schneider for his work in implementing the original aaai.sty file, liberally using the ideas of other style hackers, including Barbara Beeton.
 We also acknowledge with thanks the work of George Ferguson for his guide to using the style and BibTeX files --- which has been incorporated into this document --- and Hans Guesgen, who provided several timely modifications, as well as the many others who have, from time to time, sent in suggestions on improvements to the AAAI style.
+We are especially grateful to Francisco Cruz, Marc Pujol-Gonzalez, and Mico Loretan for the improvements to the Bib\TeX{} and \LaTeX{} files made in 2020.
 
@@
 George Ferguson added support for printing the AAAI copyright slug.
-Additional changes to aaai.sty and aaai.bst have been made by the AAAI staff.
+Additional changes to aaai21.sty and aaai21.bst have been made by Francisco Cruz, Marc Pujol-Gonzalez, and Mico Loretan.
 

### 2020-08-21 paper/ch_potentials.tex
shrink boxplot
stuff

@@
 \def\year{2021}\relax
+%File: formatting-instructions-latex-2021.tex
+%release 2021.1
 \documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage{aaai20}  % DO NOT CHANGE THIS
+\usepackage{aaai21}  % DO NOT CHANGE THIS
 \usepackage{times}  % DO NOT CHANGE THIS
@@
 /Title (A Fast and Tight Heuristic for A* in Road Networks)
-/Author (Ben Strasser, Tim Zeitz)
+/Author (Foo Bar)
 /Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
@@
 \title{A Fast and Tight Heuristic for A* in Road Networks}
-\author{Ben Strasser\\
-academia@ben-strasser.net\\
-\And
-Tim Zeitz\\
-tim.zeitz@kit.edu\\
-Karlsruhe Institute of Technology
-}
+\author{Foo Bar}
+
+% \author{Ben Strasser\\
+% academia@ben-strasser.net\\
+% \And
+% Tim Zeitz\\
+% tim.zeitz@kit.edu\\
+% Karlsruhe Institute of Technology
+% }
 
@@
 \begin{abstract}
-Efficiently computing flexible and exact routes in continental-sized road networks is a tough challenge.
+Efficiently computing flexible and exact routes in large road networks is a tough challenge.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
@@
 Computing a good heuristic is a challenge of its own.
-In contrast, for road networks there also exist hierarchical speedup techniques, which do not need a heuristic.
+Shortest paths can also be quickly computed using hierarchical speedup techniques.
+They are not based on A* and do not use a heuristic.
+Hierarchical techniques are known to work well on road networks.
 They achieve speed and exactness but sacrifice A*'s flexibility.
@@
 \label{sec:intro}
-Route planning in street networks of continental scale is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
+Route planning in large street networks is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
 It is often formalized as a variant of the classical shortest path problem in weighted graphs.
 Developing flexible techniques to efficiently compute exact routes is a tough challenge.
-\emph{Exact} means that the computed path always has minimum weight.
+\emph{Exact} means that a path of provable minimum weight is computed.
 \emph{Efficiently} means that the path computation is fast.
-Being able to change the edge weights between path computations or handle more complex weights (e.g. travel times depending on the time of the day) is \emph{flexibility}.
+Being able to change the edge weights between path computations is \emph{flexibility}.
+Handling complex weights, such as edge weight functions, is another aspect of flexibility.
 
@@
 
-\caption{Nodes explored by A* with CH-Potentials. Nodes were explored from blue through green to red.}
+\caption{Nodes explored by A* with CH-Potentials. Color indicates the node removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
 \label{img:search-space}
@@
 Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
-It iteratively explores nodes of the graph in ascending distance from the source node until the target is reached.
+It iteratively explores nodes ordered by their distance from the start.
 A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
-It changes the order in which nodes are explored by using the distance from the source plus an estimate of the remaining distance to the target as the ordering criterion.
-Figure~\ref{img:search-space} depicts an example A* search.
-While flexible, A*'s performance crucially depends on the quality of the employed estimating heuristic.
+It processes nodes in a different order, preferring nodes closer to the destination.
+Which nodes are considered close is encoded using a heuristic function.
+While flexible, A*'s performance crucially depends on the quality of this function.
 Finding good heuristics that can be evaluated efficiently is difficult.
+Constructing good heuristics on road networks, such as the one depicted in Figure~\ref{img:search-space}, is the focus of this paper.
 
-In this paper, we consider a setup with a preprocessing phase and a query phase.
+In this paper, we consider a setup with an offline preprocessing phase and an online query phase.
 The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
-The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with lower bound weights $w_\ell$.
-The source $s$ and target $t$ of and the actual edge weights $w_q$ are the input to the query phase in which shortest paths are computed.
-We require that $w_q(e)\ge w_\ell(e)$ for all edges $e \in E$.
+The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with \emph{lower bound weights} $w_\ell$.
+The query phase is given the \emph{source} $s$ and the \emph{target} $t$ as input.
+$s$ is also called start and synonyms for $t$ are goal or destination.
+It is further given the \emph{query edge weights} $w_q$.
+The task is to compute a minimum length path with respect to $w_q$.
+Of $w_q$, it is only known that $w_\ell$ is a lower bound, i.e., $w_\ell \le w_q$.
 %
-Street networks of continental scale have many millions of nodes.
-% While the preprocessing phase may be slow, it must eventually finish.
-On graphs of this size, quadratic preprocessing time is prohibitive.
+While the preprocessing phase may be slow, it must eventually finish.
+On graphs with millions of nodes, quadratic preprocessing time is prohibitive.
 
 In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
-The preprocessing phase computes auxiliary data to efficiently evaluate the heuristic.
+In the preprocessing phase, it computes a Contraction Hierarchy (CH)~\cite{gssv-erlrn-12}.
+The CH is used to efficiently evaluate the heuristic.
 Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-We call our algorithm \emph{CH-Potentials}, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics, also called potentials in~\cite{gh-cspas-05}.
+We call our algorithm \emph{CH-Potentials}, following the ALT terminology.
 
-Formally, a heuristic $h$ is a function that maps nodes onto numbers.
-$h$ depends on the target $t$ of the search but not one the source $s$.
+A heuristic $h$ is a function that maps nodes onto numbers.
+$h$ depends on the target $t$ but not on the source $s$.
 Without loss of generality, we assume that $h(t)=0$.
-The \emph{tight} heuristic assigns to every node $v$, the length of a shortest $vt$-path.
+The \emph{tight} heuristic assigns to every node $x$, the length of a shortest $xt$-path.
 All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
@@
 Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
-In this sense, CH-Potentials are as tight as possible.
+In this sense, CH-Potentials are as tight as possible and thus optimal.
 
@@
 Historic traffic patterns are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
-Specialized variants for electric vehicle routing are developed in~\cite{bdgwz-sfpcs-19}.
+Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
 While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
-Further, in every setting the flexibility available at query time is fairly limited and usually encompasses only one very specific setting.
+Further, in every setting the flexibility available at query time is fairly limited.
 Combining all these hierarchical extensions is an unsolved problem.
@@
 CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, past work mostly uses A* to accelerate hierarchical approaches even further and sacrifices A*'s flexibility in the process.
-Our primary algorithm design objective is to retain A*'s flexibility.
+However, a lot of past work uses A* to accelerate hierarchical approaches even further and loose A*'s' flexibility.
 
-ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with the highest conceptual similarity to CH-Potentials.
+ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
 Just as our approach, ALT is A*-based.
@@
 In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
-For every distance estimation, a shortest path to the target is computed, whose length is used as the tight heuristic.
-Unfortunately, the CPD used has quadratic preprocessing running times, which is prohibitively expensive on large street networks.
+For every distance estimation, a shortest path to the target is computed, whose length is used as tight heuristic.
+Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
 %
@@
 The Euclidean distance is then used as A* heuristic.
+This results in a good but not a tight heuristic.
 %
@@
 It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
-This is the tightest heuristic achievable during preprocessing.
 The array computation time is unaccounted for, i.e., it is magically filled instantly.
 Oracle-A* is not achievable in practice.
-However, it is a lower bound for every algorithm following our setup.
+However, it is a lower bound for every algorithm following our two phase setup.
 We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.
@@
 %
-Electric Vehicle Routing also requires flexibility~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
-
 \section{Contribution \& Outline}
@@
 Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
-Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we describe some improvements to A* to more efficiently process low degree nodes, common in road networks.
-In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing various extended route planning applications and how they can be solved using CH-Potentials.
+Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we improve A*'s handling of low degree nodes, common in road networks.
+In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing algorithms for various extended route planning applications.
 Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
-By comparing against Oracle-A*, we show that no alternative approach can be significantly faster, assuming that $w_\ell$ is the tightest lower bound that can be known in advance.
+By comparing against Oracle-A*, we show that no two-phase alternative exact A* approach can be significantly faster.
 
@@
 A CH is a two phase technique to efficiently compute exact, shortest paths.
-It is not flexible, i.e., $w_q=w_\ell$.
+It is not flexible, i.e., requires $w_q=w_\ell$.
 For a details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
@@
 \emph{Down} edges are defined analogously.
-An \emph{up-down path} is a path where only one node $v$ is more important than both its neighbors.
-$v$ is called the \emph{mid} node.
+An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
+$m$ is called the \emph{mid} node.
 An \emph{up path} is a path where the last node is the mid node.
@@
 This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
-See~\cite{gssv-erlrn-12} for the details.
+%See~\cite{gssv-erlrn-12} for the details.
 After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
@@
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
-On road networks, this is the case~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13,dsw-cch-15,hs-gbpo-18}.
+On road networks, this is the case~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13}.
+On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.
 
-Using a CH, we can compute a tight A* heuristic.
+Using a CH, we can compute a simple and tight heuristic.
 In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
-The query phase consists of an A* search, where the evaluation of the heuristic $\pi(v)$ performs a CH-query from $v$ to $t$.
+The heuristic evaluation $h(x)$ performs a CH-query from $x$ to $t$.
 %
@@
 
-\subsection{PHAST based heuristic}
+\subsection{PHAST based Heuristic}
 
@@
 The first step is analogue to the CH query:
-From the target node $t$, all reachable nodes via reversed down-edges are explored.
+From $t$, all reachable nodes via reversed down-edges are explored.
 Algorithm~\ref{algo:ch-backward} shows this first step.
@@
 After all levels are processed, the distances from all nodes towards $t$ is computed.
-PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
-PHAST's main advantage is that can easily be parallelized.
-However, we will not consider parallelization in this paper.
-We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
+%PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
+%PHAST's main advantage is that can easily be parallelized.
+%However, we will not consider parallelization in this paper.
+%We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-Using PHAST, we can compute a tight A* heuristic in another way.
-In the query phase, we first use PHAST to compute the distances from every node to $t$ with respect to the lower bound edge weights $w_\ell$ and store the result in an array $H$.
-Then, we run $A^*$ and implement the heuristic as a lookup in the array $H$.
+Using PHAST, we can also compute a tight A* heuristic.
+In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
+Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.
 
 % This PHAST-based algorithm works.
-The PHAST-based heuristic evaluation and by extension the $A^*$ search is indeed fast.
+The $H$ lookup and by extension the $A^*$ search is indeed fast.
 However, the PHAST step before the search is comparatively expensive.
 The reason is that the distances towards $t$ are computed for \emph{all} nodes.
-Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search towards $t$.
+Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.
 
@@
 In a first step, we run the backward CH search from $t$ to obtain an array $B$.
-$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
+$B[x]$ is the minimum down $xt$-path distance or $+\infty$, if there is no such path.
 $B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.
 
-Suppose, we need to compute the potential $\pi(v)$ of a node $v$.
-We do this by first recursively computing the potentials $\pi(u)$ of all nodes for which an up-edge $(v,u)$ exists.
-Afterwards, we compute $d = \min_u\{w_\ell(v,u) + \pi(u)\}$ the minimum distance over all up-down paths that contain at least one up-edge.
-As the shortest up-down path does not necessarily have to contain an up-edge, we set $\pi(v) = \min \{ B[v], d \}$.
-This calculation is correct, as it computes the minimum up-down $vt$-path distance in $G^+$, which corresponds to the minimum $vt$-path distance in a CH.
-
-The resulting algorithm is the basic CH-Potentials algorithm.
-It is flexible and exact.
-However, we can still improve the performance by reducing the number of queue operations and heuristic evaluations.
+To compute the heuristic $h(x)$, we recursively compute for all up-edges $(x,y)$ the heuristic $h(y)$.
+Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_y\{w_\ell(x,y) + h(y)\}$.
+As not all shortest up-down paths contain an up-edge, we set $h(x) = \min \{ B[x], d \}$.
+This calculation is correct, as it computes the minimum up-down $xt$-path distance in $G^+$, which corresponds to the minimum $xt$-path distance in a CH.
+A* with this heuristic is the basic CH-Potentials algorithm.
 
@@
 
-Heuristic evaluation is an expensive operation that we want to avoid.
-Together with the queue operations, it makes up a significant amount of the running time of A*.
-The heuristic needs to be evaluated whenever a node is pushed into the queue.
-As our experiments show, the total running time roughly corresponds to the number of queue pushs.
-In this section, we propose optimizations to reduce the number of nodes pushed into the queue.
+Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue push operations.
+As both appear in pairs, reducing the number of push operations, is beneficial for both.
+Avoiding push operations on low degree nodes is the focus of this section.
 
@@
 A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
-If A* succeeds, $y$'s weight in the queue is set to $D[y]+\pi(y)$.
+If A* succeeds, $y$'s weight in the queue is set to $D[y]+h(y)$.
 Our algorithm, however, behaves differently, if $d(y)\le 2$.
@@
 If it does not reach $z$, then only $D$ is modified but no queue action is performed.
-If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+\pi(z)$.
+If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h(z)$.
 
@@
 Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
-If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+\pi(\beta)$.
-Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+\pi(b)$.
+If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h(\beta)$.
+Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h(b)$.
 If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
@@
 Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
-This reduces the number of explored nodes and thus leads to fewer heuristic evaluations.
 
@@
 
-A* is a very flexible algorithm.
+A* is a flexible algorithm.
 Many extended problems can therefore be solved using the CH-Potentials algorithm.
@@
 Otherwise, $w_q$ is set to the freeflow travel time.
-This is all that is necessary to make our CH-Potentials algorithm avoid tunnels and/or highways.
 
@@
 Truck routing differs from car routing due to night driving bans and other restrictions.
-In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is successfully employed to efficiently find routes for trucks when edges may be blocked during certain times, drivers may need to wait at parking locations, and several routes with different trade-offs between arrival time and route quality may need to be considered.
+In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is used is such scenario.
+The work considers time-dependent blocked edges and waiting at parking locations.
+Further, a trade-off between arrival time and route quality is considered.
 
@@
 
-\begin{figure*}
+\begin{figure}
 \centering
-\includegraphics[width=\textwidth]{fig/scaled_weights.pdf}
+\includegraphics[width=\linewidth]{fig/scaled_weights.pdf}
 \caption{
@@
 }\label{fig:scaled_weights}
-\end{figure*}
+\end{figure}
+
+\begin{table}
+\centering
+\caption{Average query running times and number of queue pushs with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
+\input{table/building_blocks.tex}
+\end{table}
+
+\begin{table*}
+\centering
+\caption{
+CH-Potentials performance for different route planning applications.
+We report average running times and number of queue pushs.
+We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
+Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup over this baseline.
+}\label{tab:applications}
+\input{table/applications.tex}
+\end{table*}
 
@@
 
-\begin{table}
-\centering
-\caption{Average query running times and number of queue pushs with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
-\input{table/building_blocks.tex}
-\end{table}
-
 In Table~\ref{tab:building_blocks} we investigate the impact of our optimizations.
@@
 Assuming that $w_\ell$ is the tightest possible lower bound that can be known in advance, no other heuristic can be more than a factor of 1.6 faster than CH-Potentials without sacrificing exactness.
-
-\begin{table*}
-\centering
-\caption{
-CH-Potentials performance for different route planning applications.
-We report average running times and number of queue pushs.
-We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
-Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup over this baseline.
-}\label{tab:applications}
-\input{table/applications.tex}
-\end{table*}
 

### 2020-08-22 paper/ch_potentials.tex
foo

@@
 \usepackage{booktabs}
+\usepackage{multirow}
 \usepackage[algo2e,vlined]{algorithm2e}
@@
 \subsection{Oracle-A* \& Comparison with Related Work}
+\label{sec:oracle}
 
@@
 
-In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}, induced by the largest biconnected component of the undirected graph underlying $G$.
+In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}.
+$G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
 We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
@@
 \subparagraph{Inputs and Methodology}
-Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
-To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
+Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{download.geofabrik.de/europe/germany-200101.osm.pbf}}.
+To obtain the routing graph, we use the import from RoutingKit\footnote{\url{github.com/RoutingKit/RoutingKit}}.
 The graph has 16M nodes and 35M edges.
-For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://www.mapbox.com}}.
+For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{www.mapbox.com}}.
 The  data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
@@
 We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
-Additionally, have two graphs with traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
+Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{ptvgroup.com}}.
+The PTV instances are not OSM-based.
 One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
 Table~\ref{tab:graphs} contains an overview over our instances.
+In this table, we further include the sequential running time necessary to construct the CH.
 We report preprocessing running times as averages over 10 runs.
@@
 \subparagraph{Experiments}
-We first investigate the effectiveness of the heuristic depending on how tight the lower bound is.
-For that, we take our main graph, use the unmodified travel time as the lower bound $w_\ell$, and perform queries on $w_q = \alpha \cdot w_\ell$ where $\alpha$ is a factor $\geq 1$.
-When $\alpha = 1$, the heuristic is as tight as possible.
-With growing $\alpha$ the heuristic degrades.
+The first experiment varies the tightness of the heuristic.
+The lower bound $w_\ell$ is set to the freeflow travel time.
+The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
+When $\alpha = 1$, the heuristic is tight.
+Increasing $\alpha$ degrades the heuristic's quality.
 Figure~\ref{fig:scaled_weights} depicts the results.
@@
 Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
-Up to around $\alpha = 1.1$ the running time growth is roughly quadratic before leveling off to something more resembling linear growth.
-We observe a significant amount of heavy outliers -- sometimes more than an order of magnitude slower than the median.
-Generally, running times vary quite strongly even though most queries are long range queries as we draw source and target node uniformly at random.
-The reason is that different queries (even with similar shortest path lengths) are affected differently by $\alpha$ depending on the geographic regions the shortest path traverses.
-As $\alpha$ grows, more nodes in the area around the shortest paths are explored.
-So when a shortest path traverses an area where many nodes are reachable in short distance, the search space grows stronger, than when it passes sparse areas.
-% Conclusions?
+Up to around $\alpha = 1.1$ the running time grows quickly.
+After $\alpha = 1.1$ the growth slows down.
+We observe that the running times for a fixed $\alpha$ vary strongly.
+This is an interesting observation, as with uniform source and target sampling, nearly all queries are long-distance.
+The query distance is thus not the reason.
+After some investigation, we concluded that this is due to non-uniform road graph density.
+Some regions have more roads per area than others.
+The search space of A* is an ellipsoid.
+The number explored A* nodes depends on the density of the area covered by this shape.
+As the density varies, the running times vary.
 
-In Table~\ref{tab:building_blocks} we investigate the impact of our optimizations.
-We run queries with different heuristic functions: the \emph{Zero} heuristic which turns A* back into the classical algorithm of Dijkstra, our new CH-Potentials heuristic, and an \emph{Oracle} heuristic where all distances to $t$ with regards to $w_\ell$ can be obtained immediately.
-We first observe that each one of our optimizations gains useful speedups regardless of the heuristic used.
-Our optimizations do not only reduce heuristic evaluations but also queue operations which always accelerates the algorithm.
-The number of queue pushs roughly corresponds to the running time for all heuristics.
-When comparing CH-Potentials to the Oracle heuristic we note that CH-Potentials are only a factor of 1.6 slower.
-This clearly shows that CH-Potentials have very little overhead.
-Assuming that $w_\ell$ is the tightest possible lower bound that can be known in advance, no other heuristic can be more than a factor of 1.6 faster than CH-Potentials without sacrificing exactness.
+Table~\ref{tab:building_blocks} shows the impact of our optimizations.
+We run queries with three heuristic.
+The Zero heuristic is simply $h(x)=0$ for all $x$.
+This corresponds to using Dijkstra's algorithm.
+Further, we run queries with our novel CH-Potential heuristic.
+Finally, we use the unattainable lower bound Oracle-A* as described in Section~\ref{sec:oracle}.
+Oracle-A* is unbeatable theoretic lower bound.
+The reported running time of Oracle-A* do \emph{not} account for the heuristic evaluation.
+We include Oracle-A* as it is a lower bound not only for our algorithm but every related work.
 
-Table~\ref{tab:applications} depicts results when applying our algorithm to different route planning problems.
-The greatest speedup is, of course, achieved when applying CH-Potentials to an unmodified graph with $w_q = w_\ell$.
-In that case, the performance is within an order of magnitude to what other state-of-the-art techniques achieve.
-When considering extended problems, the performance of CH-Potentials strongly depends on the tightness of $w_\ell$.
-On one hand, forbidding highways leads to over 40\% longer paths and the smallest speedup (only 5.3) among all settings.
-On the other hand, incorporating turn restrictions increases path lengths only marginally and CH-Potentials yield a speedup of around 1500 over Dijkstra on the turn expanded graph without any additional implementation complexity.
-Live traffic increases shortest paths lengths by around 15\% longer which yields running times of 127\,ms on average.
-The Mapbox traffic predictions increase shortest path lengths by about 18\%.
-Combined with the additional work from function evaluations, this leads to running times around 200\,ms.
-The PTV traffic predictions yield much smaller length increases (less than 4\%) which leads to better running times and speedups.
-% Assessing which of these predictions is more realistic is not possible within the scope of this paper.
-Combining Mapbox live traffic and traffic predictions yields running times only slightly slower than traffic predictions alone.
-% To the best of our knowledge, no other exact technique so far could combine these two scenarios.
-Additionally adding turn restrictions slows running times down to around 510\,ms despite the length increase only growing by 1.3 percentage points.
-The reason for this is that the turn expanded graph is three times bigger and that the BCC optimization is not applicable for turn-expanded graphs.
-This is still a speedup of 12.2 over the baseline.
-% To the best of our knowledge no other approach so far did even attempt handling this combination of scenarios.
+We observe that the number of queue pushes roughly correlates with running time.
+Further, each of our optimizations reduces the number of queue pushes.
+Very interesting is the comparison with Oracle-A*.
+Our algorithm is only 1.6 times slower than an unachievable theoretic lower bound.
+This means that no competitor algorithm such as ALT or CPD-Heuristics can be significantly faster than CH-Potentials.
+
+Table~\ref{tab:applications} depicts the running times of CH-Potentials in the various applications, such as those described in Section~\ref{sec:extensions}.
+Speedups compared to extensions of Dijkstra's algorithm for the respective application are depicted.
+We start with the inflexible routing problem where $w_q = w_\ell$.
+This is the only problem variant solved by the basic CH algorithm.
+In this setting, there is a huge speedup of 3243.
+Such large speedups are typical for CH.
+This shows that CH-Potentials gracefully converges to a CH in the $w_q = w_\ell$ special case.
+
+In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the tight heuristic.
+We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
+Forbidding highways results in the largest length increase and in the smallest speedup.
+The other extreme are turn restrictions.
+They have only a small impact on the length increase.
+The achieved speedups are therefore comparable to CH speedups.
+%
+Mapbox live traffic has a length increase of around 15\%, which yields running times of 127\,ms.
+The length increase of Mapbox traffic predictions are about 18\%, and results in a running time of 200\,ms.
+The speedup in the predicted scenario is larger than in the live setting, as the travel time function evaluations slow down Dijkstra's algorithm.
+Combining predicted and live traffic results in a running time only slightly higher than for the predicted scenario.
+Further adding turn restrictions, increases the running times.
+This increase is mostly due to the BCC optimization becoming ineffective when considering turns.
+It is not due to the length increase of using turns.
+With everything activated, our algorithm still has a speedup of 12.2 over the baseline.
+Interestingly, the PTV traffic predictions have a much smaller length increase that the Mapbox predictions.
+This results in smaller running times of our algorithm.
 
@@
 
-In this paper, we introduced CH-Potentials, a fast and exact new A* heuristic for finding shortest paths in road networks.
-The approach is flexible and can handle a multitude of routing scenarios with very little implementation complexity.
-The performance of CH-Potentials depends on the tightness of the lower bound, so whether CH-Potentials are fit for a certain application depends on whether a tight lower bound can be derived.
-Our experiments show that CH-Potentials are efficient in extracting distance estimates from a lower bound graph and only around 1.6 times slower than immediately obtaining these estimates from an hypothetical oracle.
+In this paper, we introduced CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and CH for finding shortest paths in road networks.
+The approach can handle a multitude of routing scenarios with very little implementation complexity.
+The performance of CH-Potentials crucially depends on the availability of good weight lower bounds in preprocessing phase.
+Our experiments show, that this availability is highly application dependent.
+We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the heuristic.
+We conclude from this, that to achieve smaller running times with A* based algorithms variations of the problem setting are necessary.
+This leads to multiple avenues for future research.
+
+Dropping the provable exactness requirement using a setup similar to anytime A* \cite{DBLP:conf/aaai/ZhouH02,DBLP:conf/nips/LikhachevGT03} would be interesting.
+Another promising research avenue would be to investigate other graphs than road networks.
+A lot of research into grid maps exists including a series of competitions called GPPC \cite{DBLP:conf/socs/SturtevantTTUKS15}.
+Hierarchical techniques have been shown to work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
 

### 2020-08-22 paper/ch_potentials.tex
foo

@@
 
-The classical shortest path problem allows to turn from every edge to every other incident edge at every node.
+The classical shortest path problem allows to freely change edges at nodes.
 However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
@@
 
-% In the extended problem, a \emph{turn weight} $w_t$ is introduced~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
-A \emph{turn weight} $w_t$ maps a pair of incident edges onto the time required to perform the turn.
-Formally, $w_t$ is a function from $V^3$ to $\mathbb{R}^+ \cup \{+\infty\}$.
-A forbidden turn is modeled by setting the corresponding $w_t$ to $+\infty$.
+% Ich bleibe bei v_i hier, damit keine Verwechselung mit den freien Variablen x,y,z gibt.
+A \emph{turn weight} $w_t$  maps a pair of incident edges onto the turning time or $+\infty$.
+A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}:\[
+w_q(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1})  + w_q(v_i,v_{i+1})
+\] The objective is to find a path between two given edges with minimum turn-aware weight.
+The first term $w_q(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
+It can thus be ignored during optimization.
 
-A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware path weight}:\[
-w_q(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1})  + w_q(v_i,v_{i+1})
-\] The input to the extended problem consists of a source edge and a target edge.
-The objective is to find a path connecting these edges such that the turn-aware path weight is minimized.
-The first term $w_q(v_1, v_2)$ only depends on the source edge.
-As all considered paths share the same source edge, the term is the same for all paths.
-We can therefore ignore it during the optimization.
-
-To account for turns, many algorithms construct a \emph{turn-expanded} graph $G'$.
+A common strategy is to construct a \emph{turn-expanded} graph $G'$.
 Edges in the input graph $G$ correspond to \emph{expanded nodes} in $G'$.
-For every pair of successive edges $(x,y)$ and $(y,z)$ in $G$, there is an \emph{expanded edge} in $G'$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
+For every pair of incident edges $(x,y)$ and $(y,z)$ in $G$, there is an \emph{expanded edge} in $G'$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
 A sequence of expanded nodes in the expanded graph $G'$ corresponds to a sequence of edges in the input graph $G$.
 The weight of a path in $G'$ is equal to the turn-aware weight of the corresponding path in $G$ minus the irrelevant $w_q(v_1,v_2)$ term.
-Turn-aware shortest paths can thus be computed by searching for shortest paths in the expanded graph.
+The turn-aware routing problem can be solved by searching for paths in $G'$.
 
-% The simplest combination of turn-aware routing with CH-Potentials consists of using the expanded graph as input to the CH-Potential algorithm.
-% However, this implies computing a CH on the expanded graph.
-% This is feasible but has a number of problems such as higher preprocessing running time, increased memory consumption, or more complex code~\cite{gv-errnt-11}.
-% We want to avoid these complications.
+The A* search runs in the expanded $G'$ and uses a heuristic $h'$.
+The CH is constructed for $G$.
+With CH-Potentials, we thus get the tight heuristic $h$.
+We set $h'(x,y) = h(y)$.
 
-% We therefore only apply the A* search to the expanded graph $G'$.
-We only apply the A* search to the expanded graph $G'$.
-The heuristic is computed with respect to the input graph $G$.
-Let $(p,q)$ be the target edge and $(x,y)$ the edge for which the A* search wants to compute a heuristic.
-We construct non-turn-aware CH-Potentials towards $q$.
-The heuristic for the edge $(x,y)$ is equal to the heuristic of $y$.
-
-To prove correctness, consider the turn-aware path weight of a shortest path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
+To prove that $h'(x,y)$ is consistent, consider the turn-aware weight of a path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
 We can lower bound the expression as follows:
@@
 \]
-The $\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ expression is the length of some, not necessarily shortest, $yq$-path in $G$ with $w_\ell$.
-It is by definition not smaller than the shortest $yq$-path length in $G$ with $w_\ell$.
-The shortest $yq$-path length is what CH-Potentials compute.
+$\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ is a $yq$-path length in $G$.
+It is no shorter than the shortest $yq$-path length in $G$, which is equal to $h(y)$.
 The heuristic is therefore a lower bound.
@@
 
-Unfortunately, the undirected graph underlying the expanded graph is always biconnected, if the input graph is strongly connected.
-The optimization described in~\ref{sec:largested-biconnected-component} is therefore ineffective in this setting.
+Sadly, the undirected graph underlying $G'$ is always biconnected, if the input graph is strongly connected.
+The optimization described in Section~\ref{sec:largested-biconnected-component} is therefore ineffective.
 
-With this setup we extend CH-Potentials in a way that keeps all turn-related information out of the CH.
-This allows for a comparatively simple implementation.
+With this setup CH-Potentials support turns without requiring turn information in the CH.
 
@@
 
-Formally, $w_q$ is a function from $E\times \mathbb{R}$ to $\mathbb{R}^+$.
-$w_q(e, \tau)$ is the travel time through edge $e$ when entering it at moment $\tau$.
+Formally, the time-dependent query weight $w_p$ is a function from $E\times \mathbb{R}$ to $\mathbb{R}^+$.
+$w_p(e, \tau)$ is the travel time through edge $e$ when entering it at moment $\tau$.
 The input to the extended problem consists of a source node $s$ and a target node $t$, as in the classical problem formulation.
@@
 \begin{split}
-\alpha_{1} & = w_q(e_1, \tau_s) \\
-\alpha_{k} & = \alpha_{k-1} + w_q(e_1, \alpha_{k-1})
+\alpha_{1} & = w_p(e_1, \tau_s) \\
+\alpha_{k} & = \alpha_{k-1} + w_p(e_1, \alpha_{k-1})
 \end{split}
@@
 Formally stated, the following must hold:\[
-\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta
+\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_p(e, \tau) \le w_p(e, \tau+\delta) + \delta
 \]
+Our implementation stores edge travel times using piece-wise linear functions.
 
-Usually, travel time functions are stored as piece-wise linear functions.
-Our implementation follows this setup.
-However in principle, our algorithm works with any representation that enables an efficient function evaluation.
-
-To solve the time-dependent routing problem, we employ a strategy very similar to TD-ALT~\cite{ndls-bastd-12}.
+To solve the time-dependent routing problem, we employ a strategy very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
 The main difference is that we use CH-Potentials instead of landmarks to guide the A* search.
-We set $w_\ell(e) = \min_\tau w_q(e,\tau)$, that is the minimum travel time.
-
-We modify the A* search to use the tentative distance at a node $x$ as $\tau$ when evaluating the weight of an outgoing edge $(x,y)$.
-This modification is straight-forward, as our optimizations do not use shortcuts and only traverse edges in forward direction.
-For the correctness proof, we refer to~\cite{dw-lbrdg-07}.
+We set $w_\ell(e) = \min_\tau w_p(e,\tau)$, that is the minimum travel time.
+Following TD-ALT, we modify the A* search to use the tentative distance at a node $x$ as $\tau$ when evaluating the weight of an outgoing edge $(x,y)$.
 
@@
 However, just using live traffic data is problematic for long routes as traffic changes while driving.
-At some point, one wants to switch from live traffic back to the predicted traffic.
-In this section, we first describe how to use only live traffic and then describe how to combine it with predicted traffic.
+At some point, one wants to switch from live traffic to the predicted traffic.
+In this section, we first describe a setup with only live traffic and then extend it with predicted traffic.
 
-To support live traffic, we set $w_\ell$ to the travel time without any congestions.
+To support live traffic, we set $w_\ell$ to the freeflow travel time.
 $w_q$ is set to the travel time accounting for current traffic.
 As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
-With this setup, the CH-Potentials algorithms can make use of live traffic.
+This is all that is necessary to combine live traffic and CH-Potentials.ic.
 
-To combine live traffic with predicted traffic, we define a modified travel time function.
+To combine live traffic with predicted traffic, we define a modified travel time function $w_q$ that is then used as query weights.
 Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
-Further, $w_c(e)$ is the travel time according to live traffic.
-Finally, we denote by $\tau_{\mathrm{soon}}$ the moment up to which we trust the live traffic.
+Further, $w_c(e)$ is the travel time according to current live traffic.
+Finally, we denote by $\tau_{\mathrm{soon}}$ the moment when we switch to predicted traffic.
 In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
@@
 Our modified travel time function linearly approaches the predicted travel time. % with a bounded slope.
-
+%
 Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
@@
 We set $w_\ell$ again to the freeflow travel time.
-Finally, we run the modified time-dependent A* described in the previous section on this modified travel time function.
-In our implementation, we to not modify the representation of $w_p$.
-Instead, we evaluate the formulas above each time that the modified travel time function is evaluated.
+We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
+In our implementation, we to not modify the representation of $w_p$ and evaluate the formulas above at each heuristic evaluation.
 
 With this setup, CH-Potentials support a combination of live and predicted traffic.
-We did not make any modification to our algorithm, that would hinder a combination with tunnel and/or highway avoidance or turn-aware routing.
+Further adding tunnel and/or highway avoidance or turn-aware routing is simple.
 This straight-forward integration of complex routing problems is the strength of the CH-Potentials algorithm.
@@
 Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{download.geofabrik.de/europe/germany-200101.osm.pbf}}.
-To obtain the routing graph, we use the import from RoutingKit\footnote{\url{github.com/RoutingKit/RoutingKit}}.
+To obtain the routing graph, we use the import from RoutingKit\footnote{\url{www.github.com/RoutingKit/RoutingKit}}.
 The graph has 16M nodes and 35M edges.
@@
 We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
-Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{ptvgroup.com}}.
+Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{www.ptvgroup.com}}.
 The PTV instances are not OSM-based.
@@
 Further adding turn restrictions, increases the running times.
-This increase is mostly due to the BCC optimization becoming ineffective when considering turns.
+This increase is mostly due to the BCC optimization of Section~\ref{sec:largested-biconnected-component} becoming ineffective when considering turns.
 It is not due to the length increase of using turns.
@@
 In this paper, we introduced CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and CH for finding shortest paths in road networks.
-The approach can handle a multitude of routing scenarios with very little implementation complexity.
+The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
 The performance of CH-Potentials crucially depends on the availability of good weight lower bounds in preprocessing phase.
-Our experiments show, that this availability is highly application dependent.
-We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the heuristic.
-We conclude from this, that to achieve smaller running times with A* based algorithms variations of the problem setting are necessary.
+Our experiments show, that this availability highly depends on the application.
+We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the tight heuristic.
+We conclude from this, that to achieve significantly smaller running times with A* based algorithms in road networks, variations of the problem setting are necessary.
 This leads to multiple avenues for future research.

### 2020-08-25 paper/ch_potentials.tex
typos etc

@@
 As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
-This is all that is necessary to combine live traffic and CH-Potentials.ic.
+This is all that is necessary to combine live traffic and CH-Potentials.
 
@@
 We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
-In our implementation, we to not modify the representation of $w_p$ and evaluate the formulas above at each heuristic evaluation.
+In our implementation, we to not modify the representation of $w_p$ and evaluate the formulas above at each travel time evaluation.
 
@@
 Truck routing differs from car routing due to night driving bans and other restrictions.
-In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is used is such scenario.
+In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is used for such a scenario.
 The work considers time-dependent blocked edges and waiting at parking locations.
@@
 Table~\ref{tab:building_blocks} shows the impact of our optimizations.
-We run queries with three heuristic.
+We run queries with three heuristics.
 The Zero heuristic is simply $h(x)=0$ for all $x$.
@@
 Finally, we use the unattainable lower bound Oracle-A* as described in Section~\ref{sec:oracle}.
-Oracle-A* is unbeatable theoretic lower bound.
+Oracle-A* is an unbeatable theoretic lower bound.
 The reported running time of Oracle-A* do \emph{not} account for the heuristic evaluation.

### 2020-08-29 paper/ch_potentials.tex
use preprocessing tight

@@
 \pdfinfo{
-/Title (A Fast and Tight Heuristic for A* in Road Networks)
+/Title (A Fast and Preprocessing-Tight Heuristic for A* in Road Networks)
 /Author (Foo Bar)
@@
 }
-\title{A Fast and Tight Heuristic for A* in Road Networks}
+\title{A Fast and Preprocessing-Tight Heuristic for A* in Road Networks}
 \author{Foo Bar}
@@
 They achieve speed and exactness but sacrifice A*'s flexibility.
-In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute tight A* heuristics.
+In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute preprocessing-tight A* heuristics.
 We call our technique CH-Potentials.
@@
 In this sense, CH-Potentials are as tight as possible and thus optimal.
+We say that this heuristic that is tight with respect to $w_\ell$ is \emph{preprocessing-tight}.
 
@@
 Just as our approach, ALT is A*-based.
-However, ALT's heuristic is not tight.
+However, ALT's heuristic is not preprocessing-tight.
 %
@@
 The Euclidean distance is then used as A* heuristic.
-This results in a good but not a tight heuristic.
+This results in a good but not a preprocessing-tight heuristic.
 %
@@
 To compare against related work, we introduce the concept of Oracle-A*.
-It has access to a shortest distance array with respect to $w_\ell$, i.e., it is tight.
+It has access to a shortest distance array with respect to $w_\ell$, i.e., it is preprocessing-tight.
 The array computation time is unaccounted for, i.e., it is magically filled instantly.
@@
 
-We introduce CH-Potentials, a two-phase technique to efficiently compute tight heuristics, in Section~\ref{sec:main-algo}.
+We introduce CH-Potentials, a two-phase technique to efficiently compute preprocessing-tight heuristics, in Section~\ref{sec:main-algo}.
 Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
@@
 
-Using a CH, we can compute a simple and tight heuristic.
+Using a CH, we can compute a simple and preprocessing-tight heuristic.
 In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
@@
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-Using PHAST, we can also compute a tight A* heuristic.
+Using PHAST, we can also compute a preprocessing-tight A* heuristic.
 In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
@@
 The CH is constructed for $G$.
-With CH-Potentials, we thus get the tight heuristic $h$.
+With CH-Potentials, we get the preprocessing-tight heuristic $h$.
 We set $h'(x,y) = h(y)$.
@@
 
-In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the tight heuristic.
+In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the preprocessing-tight heuristic.
 We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
@@
 Our experiments show, that this availability highly depends on the application.
-We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the tight heuristic.
+We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
 We conclude from this, that to achieve significantly smaller running times with A* based algorithms in road networks, variations of the problem setting are necessary.

### 2020-09-07 paper/ch_potentials.tex
more polishing

@@
 \pdfinfo{
-/Title (A Fast and Preprocessing-Tight Heuristic for A* in Road Networks)
+/Title (A Fast and Tight Heuristic for A* in Road Networks)
 /Author (Foo Bar)
@@
 }
-\title{A Fast and Preprocessing-Tight Heuristic for A* in Road Networks}
+\title{A Fast and Tight Heuristic for A* in Road Networks}
 \author{Foo Bar}
@@
 Computing a good heuristic is a challenge of its own.
-Shortest paths can also be quickly computed using hierarchical speedup techniques.
-They are not based on A* and do not use a heuristic.
-Hierarchical techniques are known to work well on road networks.
+On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
+They are not based on A* and do not use a heuristic but compute auxiliary data during a preprocessing phase.
+% Hierarchical techniques are known to work well on road networks.
 They achieve speed and exactness but sacrifice A*'s flexibility.
-In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute preprocessing-tight A* heuristics.
+In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute A* heuristics.
 We call our technique CH-Potentials.
@@
 \emph{Efficiently} means that the path computation is fast.
-Being able to change the edge weights between path computations is \emph{flexibility}.
-Handling complex weights, such as edge weight functions, is another aspect of flexibility.
+Being able to change the edge weights between path computations or handling complex weights, such as edge weight functions, is \emph{flexibility}.
+% finde es klingt stringenter, wenn jede Eigenschaft genau einen Satz hat.
 
@@
 Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-We call our algorithm \emph{CH-Potentials}, following the ALT terminology.
+Following the ALT terminology, we call our algorithm \emph{CH-Potentials}, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics.
 
@@
 In this sense, CH-Potentials are as tight as possible and thus optimal.
-We say that this heuristic that is tight with respect to $w_\ell$ is \emph{preprocessing-tight}.
+We say that a heuristic that is tight with respect to $w_\ell$ is \emph{preprocessing-tight}.
 
@@
 CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, a lot of past work uses A* to accelerate hierarchical approaches even further and loose A*'s' flexibility.
+However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s' flexibility.
 
@@
 In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
-For every distance estimation, a shortest path to the target is computed, whose length is used as tight heuristic.
+For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
+% du hast mein "the" wieder rausgenommen, aber ohne Artikel klingt das fuer mich einfach falsch...
+Lasse mich aber gerne eines besseren belehren.
 Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
@@
 Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
-By comparing against Oracle-A*, we show that no two-phase alternative exact A* approach can be significantly faster.
+By comparing against Oracle-A*, we show that no other exact two-phase A*-based approach can be significantly faster.
 
@@
 
-Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue push operations.
-As both appear in pairs, reducing the number of push operations, is beneficial for both.
-Avoiding push operations on low degree nodes is the focus of this section.
+Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue operations. % queue pushs stimmt hier nicht...
+We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
+Avoiding pushing low degree nodes into the queue is the focus of this section.
 
@@
 We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
-In our implementation, we to not modify the representation of $w_p$ and evaluate the formulas above at each travel time evaluation.
+In our implementation, we to not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
 
 With this setup, CH-Potentials support a combination of live and predicted traffic.
+We did not make any modification, that would hinder a combination with other extensions.
 Further adding tunnel and/or highway avoidance or turn-aware routing is simple.
@@
 Some regions have more roads per area than others.
-The search space of A* is an ellipsoid.
-The number explored A* nodes depends on the density of the area covered by this shape.
+% glaub das mit dem ellispoid lenkt eher ab und braeuchte wenn dann noch eine ganze menge extraerklaerungen...
+The number explored A* nodes depends on the density of the search space area.
 As the density varies, the running times vary.
@@
 The reported running time of Oracle-A* do \emph{not} account for the heuristic evaluation.
-We include Oracle-A* as it is a lower bound not only for our algorithm but every related work.
+We include Oracle-A* as it is a lower bound not only for our algorithm but also related work.
 
 We observe that the number of queue pushes roughly correlates with running time.
-Further, each of our optimizations reduces the number of queue pushes.
+Each optimization reduces both queue pushes and running times.
+% Further, each of our optimizations reduces the number of queue pushes.
 Very interesting is the comparison with Oracle-A*.
-Our algorithm is only 1.6 times slower than an unachievable theoretic lower bound.
-This means that no competitor algorithm such as ALT or CPD-Heuristics can be significantly faster than CH-Potentials.
+Our algorithm is only 1.6 times slower than the unachievable theoretic lower bound.
+This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster than CH-Potentials.
 
@@
 We start with the inflexible routing problem where $w_q = w_\ell$.
-This is the only problem variant solved by the basic CH algorithm.
+This is the problem variant solved by the basic CH algorithm.
 In this setting, there is a huge speedup of 3243.
@@
 The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
-The performance of CH-Potentials crucially depends on the availability of good weight lower bounds in preprocessing phase.
+The performance of CH-Potentials crucially depends on the availability of good lower bounds in the preprocessing phase.
 Our experiments show, that this availability highly depends on the application.
 We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
-We conclude from this, that to achieve significantly smaller running times with A* based algorithms in road networks, variations of the problem setting are necessary.
+% We conclude from this, that to achieve significantly smaller running times with A* based algorithms in road networks, variations of the problem setting are necessary.
+Achieving significantly smaller running times could still be possible in variations of the problem setting. % hab das mal ein bisschen mehr auf future work hin getrimmt
 This leads to multiple avenues for future research.

### 2020-09-07 eval/applications.py
update scripts so they match the adjusted tables

@@ -80,10 +80,10 @@ table = table.rename(index={
     '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/': 'TDGer06',
     '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/': 'TDEur17',
     '/algoDaten/zeitz/roadgraphs/osm_europe/': 'OSM Europe',
-    '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Ger',
-    '/algoDaten/zeitz/roadgraphs/osm_ger_td/': 'OSM Ger',
+    '/algoDaten/zeitz/roadgraphs/osm_ger/': R'\multirow{8}{*}{OSM Ger}',
+    '/algoDaten/zeitz/roadgraphs/osm_ger_td/': R'\multirow{8}{*}{OSM Ger}',
     '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe',
-    'perfect': 'Unmodified',
+    'perfect': 'Unmodified ($w_q=w_\\ell$)',
     'no_highways': 'No Highways',
     'no_tunnels': 'No Tunnels',
     'chpot_turns': 'Turns',

### 2020-09-07 eval/building_blocks.py

@@ -35,12 +35,17 @@ table['num_queue_pushs'] = table['num_queue_pushs'] / 1000
 table = table.round(1)
 
 table = table.reindex(['Zero', 'CH', 'Oracle'], level=1)
+table = table.rename(index={
+    'Zero': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{Zero}}',
+    'CH': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{CH-\\Potentials}}}',
+    'Oracle': R'\multirow{2}{*}{Oracle}',
+})
 
 table = table.reset_index(level=0,drop=True).reset_index(level=[2,3])
 
 lines = table.to_latex(escape=False).split("\n")
 
-lines = lines[:2] + [
+lines = [R'\begin{tabular}{clllrr}'] + [lines[1]] + [
     R" & & & & Running &     Queue \\",
     R" Heur. & BCC & Deg2 & Deg3 & time [ms] & [$\cdot 10^3$] \\"
 # ] + lines[4:]

### 2020-09-08 paper/ch_potentials.tex
final changes by Ben

@@
 Computing a good heuristic is a challenge of its own.
-On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
-They are not based on A* and do not use a heuristic but compute auxiliary data during a preprocessing phase.
+In road networks, shortest paths can alterantively be quickly computed using hierarchical techniques.
+They are not based on A* and do not use a heuristic.
+Hierarchical techniques use a slow preprocessing and fast query phase.
+The preprocessing phase computes auxiliary data that is used in the query phase.
+The auxiliary is independent of origin or destination of the search.
 % Hierarchical techniques are known to work well on road networks.
-They achieve speed and exactness but sacrifice A*'s flexibility.
-In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute A* heuristics.
+Hierarchical techniques achieve speed and exactness but sacrifice A*'s flexibility.
+In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute preprocessing-tight A* heuristics.
 We call our technique CH-Potentials.
@@
 For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
-% du hast mein "the" wieder rausgenommen, aber ohne Artikel klingt das fuer mich einfach falsch...
-Lasse mich aber gerne eines besseren belehren.
 Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
@@
 
-Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue operations. % queue pushs stimmt hier nicht...
+Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue operations.
 We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
@@
 Some regions have more roads per area than others.
-% glaub das mit dem ellispoid lenkt eher ab und braeuchte wenn dann noch eine ganze menge extraerklaerungen...
 The number explored A* nodes depends on the density of the search space area.
@@
 We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
-% We conclude from this, that to achieve significantly smaller running times with A* based algorithms in road networks, variations of the problem setting are necessary.
-Achieving significantly smaller running times could still be possible in variations of the problem setting. % hab das mal ein bisschen mehr auf future work hin getrimmt
+Achieving significantly smaller running times could still be possible in variations of the problem setting.
 This leads to multiple avenues for future research.

### 2020-09-08 paper/ch_potentials.tex
revert abstract

@@
 Computing a good heuristic is a challenge of its own.
-In road networks, shortest paths can alterantively be quickly computed using hierarchical techniques.
+Shortest paths can also be quickly computed using hierarchical speedup techniques.
 They are not based on A* and do not use a heuristic.
-Hierarchical techniques use a slow preprocessing and fast query phase.
-The preprocessing phase computes auxiliary data that is used in the query phase.
-The auxiliary is independent of origin or destination of the search.
-% Hierarchical techniques are known to work well on road networks.
-Hierarchical techniques achieve speed and exactness but sacrifice A*'s flexibility.
-In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute preprocessing-tight A* heuristics.
+Hierarchical techniques are known to work well on road networks.
+They achieve speed and exactness but sacrifice A*'s flexibility.
+In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute a A* heuristics.
 We call our technique CH-Potentials.

### 2020-09-08 Rakefile.rb
prepare reproducability packge

@@ -1,5 +1,7 @@
-raise "No or wrong experiments output dir configured - please set CHPOT_EXP_OUTPUT_DIR env var" unless Dir.exists? ENV['CHPOT_EXP_OUTPUT_DIR'] || ""
-exp_dir = ENV['CHPOT_EXP_OUTPUT_DIR']
+exp_dir = 'exp'
+data_dir = 'data'
+
+only_public = !ENV['ONLY_PUBLIC'].nil?
 
 file "paper/ch_potentials.pdf" => [
   "paper/ch_potentials.tex",
@@ -46,27 +48,30 @@ namespace "table" do
 end
 
 osm_ger_src = 'https://download.geofabrik.de/europe/germany-200101.osm.pbf'
-osm_ger_src_file = '/algoDaten/zeitz/roadgraphs/germany-200101.osm.pbf'
+osm_ger_src_file = "#{data_dir}/germany-200101.osm.pbf"
 osm_eur_src = 'https://download.geofabrik.de/europe-200101.osm.pbf'
-osm_eur_src_file = '/algoDaten/zeitz/roadgraphs/europe-200101.osm.pbf'
+osm_eur_src_file = "#{data_dir}/europe-200101.osm.pbf"
 
-osm_ger = "/algoDaten/zeitz/roadgraphs/osm_ger/"
-osm_ger_td = "/algoDaten/zeitz/roadgraphs/osm_ger_td/"
-osm_eur = "/algoDaten/zeitz/roadgraphs/osm_europe/"
-dimacs_graph = "/algoDaten/zeitz/roadgraphs/europe/"
+osm_ger = "#{data_dir}/osm_ger/"
+osm_ger_td = "#{data_dir}/osm_ger_td/"
+osm_eur = "#{data_dir}/osm_europe/"
+dimacs_graph = "#{data_dir}/europe/"
 
 # static_graphs = [osm_ger, osm_eur, dimacs_graph]
 static_graphs = [osm_ger]
-td_graphs = [
-  "/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/",
-  "/algoDaten/graphs/cleaned_td_road_data/de/day/dido/",
-  osm_ger_td
-]
+td_graphs = []
+unless only_public
+  td_graphs += [
+    "#{data_dir}/ptv17/",
+    "#{data_dir}/ger06/",
+    osm_ger_td
+  ]
+end
 graphs = static_graphs + td_graphs
 
[…]

### 2020-09-08 eval/building_blocks.py

@@ -7,7 +7,7 @@ import glob
 import os
 import re
 
-base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+base = "exp/"
 paths = glob.glob(base + "building_blocks/*.json")
 data = [json.load(open(path)) for path in paths]
 

### 2020-09-08 eval/graphs.py

@@ -7,7 +7,7 @@ import glob
 import os
 import re
 
-base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+base = "exp/"
 paths = glob.glob(base + "applications/*.json")
 data = [json.load(open(path)) for path in paths]
 
@@ -39,9 +39,9 @@ preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for pa
 table = queries.groupby(['graph'])[['num_nodes', 'num_edges']].mean()
 
 table = table.reindex([
-    '/algoDaten/zeitz/roadgraphs/osm_ger/',
-    '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/',
-    '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/'])
+    'data/osm_ger/',
+    'data/ptv17',
+    'data/ger06'])
 
 table = table.join(preprocessing.groupby('graph').mean())
 table['num_nodes'] = table['num_nodes'] / 1000000.0
@@ -49,12 +49,12 @@ table['num_edges'] = table['num_edges'] / 1000000.0
 table = table.round(1)
 
 table = table.rename(index={
-    '/algoDaten/graphs/cleaned_td_road_data/de/day/dido/': 'TDGer06',
-    '/algoDaten/graphs/cleaned_td_road_data/ptv17-eur-car/day/di/': 'TDEur17',
-    '/algoDaten/zeitz/roadgraphs/osm_europe/': 'OSM Europe',
-    '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Ger',
-    '/algoDaten/zeitz/roadgraphs/osm_ger_td/': 'TD OSM Ger',
-    '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe',
+    'data/ger06': 'TDGer06',
+    'data/ptv17': 'TDEur17',
+    'data/osm_europe/': 'OSM Europe',
+    'data/osm_ger/': 'OSM Ger',
+    'data/osm_ger_td/': 'TD OSM Ger',
+    'data/europe/': 'DIMACs Europe',
 })
 
 lines = table.to_latex(escape=False).split("\n")

### 2020-09-08 eval/scaled_weights.py

@@ -12,13 +12,13 @@ import json
 import glob
 import os
 
-base = os.environ['CHPOT_EXP_OUTPUT_DIR']
+base = "exp/"
 paths = glob.glob(base + "scaled_weights/*.json")
 data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
     'weight_factor': exp['factor'],
-    'graph': { '/algoDaten/zeitz/roadgraphs/europe/': 'DIMACs Europe', '/algoDaten/zeitz/roadgraphs/osm_ger/': 'OSM Germany', '/algoDaten/zeitz/roadgraphs/osm_europe/': 'OSM Europe' }[run['args'][1]],
+    'graph': { 'data/europe/': 'DIMACs Europe', 'data/osm_ger/': 'OSM Germany', 'data/osm_europe/': 'OSM Europe' }[run['args'][1]],
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 

### 2020-09-08 Rakefile.rb
reproducibility path fixes

@@ -1,5 +1,5 @@
-exp_dir = 'exp'
-data_dir = 'data'
+exp_dir = Dir.pwd + '/exp'
+data_dir = Dir.pwd + '/data'
 
 only_public = !ENV['ONLY_PUBLIC'].nil?
 

### 2020-09-08 eval/applications.py

@@ -8,6 +8,7 @@ import os
 import re
 
 base = "exp/"
+cwd = os.getcwd()
 paths = glob.glob(base + "applications/*.json")
 data = [json.load(open(path)) for path in paths]
 
@@ -50,11 +51,11 @@ table = queries.groupby(['graph', 'exp']).agg(
     )
 
 table = table.reindex(['data/europe/',
-    'data/osm_europe/',
-    'data/osm_ger/',
-    'data/osm_ger_td/',
-    'data/ptv17',
-    'data/ger06'], level=0) \
+    cwd + '/data/osm_europe/',
+    cwd + '/data/osm_ger/',
+    cwd + '/data/osm_ger_td/',
+    cwd + '/data/ptv17',
+    cwd + '/data/ger06'], level=0) \
     .reindex(['chpot_td',
     'random_times_2',
     'random_times_10',
@@ -77,12 +78,12 @@ table['speedup'] = table['dijkstra_running_time_ms'] / table['running_time_ms']
 table = table.round(1)
 
 table = table.rename(index={
-    'data/ger06': 'TDGer06',
-    'data/ptv17': 'TDEur17',
-    'data/osm_europe/': 'OSM Europe',
-    'data/osm_ger/': R'\multirow{8}{*}{OSM Ger}',
-    'data/osm_ger_td/': R'\multirow{8}{*}{OSM Ger}',
-    'data/europe/': 'DIMACs Europe',
+    cwd + '/data/ger06': 'TDGer06',
+    cwd + '/data/ptv17': 'TDEur17',
+    cwd + '/data/osm_europe/': 'OSM Europe',
+    cwd + '/data/osm_ger/': R'\multirow{8}{*}{OSM Ger}',
+    cwd + '/data/osm_ger_td/': R'\multirow{8}{*}{OSM Ger}',
+    cwd + '/data/europe/': 'DIMACs Europe',
     'perfect': 'Unmodified ($w_q=w_\\ell$)',
     'no_highways': 'No Highways',
     'no_tunnels': 'No Tunnels',

### 2020-09-08 eval/graphs.py

@@ -8,6 +8,7 @@ import os
 import re
 
 base = "exp/"
+cwd = os.getcwd()
 paths = glob.glob(base + "applications/*.json")
 data = [json.load(open(path)) for path in paths]
 
@@ -39,9 +40,9 @@ preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for pa
 table = queries.groupby(['graph'])[['num_nodes', 'num_edges']].mean()
 
 table = table.reindex([
-    'data/osm_ger/',
-    'data/ptv17',
-    'data/ger06'])
+    cwd + '/data/osm_ger/',
+    cwd + '/data/ptv17',
+    cwd + '/data/ger06'])
 
 table = table.join(preprocessing.groupby('graph').mean())
 table['num_nodes'] = table['num_nodes'] / 1000000.0
@@ -49,12 +50,12 @@ table['num_edges'] = table['num_edges'] / 1000000.0
 table = table.round(1)
 
 table = table.rename(index={
-    'data/ger06': 'TDGer06',
-    'data/ptv17': 'TDEur17',
-    'data/osm_europe/': 'OSM Europe',
-    'data/osm_ger/': 'OSM Ger',
-    'data/osm_ger_td/': 'TD OSM Ger',
-    'data/europe/': 'DIMACs Europe',
+    cwd + '/data/ger06': 'TDGer06',
+    cwd + '/data/ptv17': 'TDEur17',
+    cwd + '/data/osm_europe/': 'OSM Europe',
+    cwd + '/data/osm_ger/': 'OSM Ger',
+    cwd + '/data/osm_ger_td/': 'TD OSM Ger',
+    cwd + '/data/europe/': 'DIMACs Europe',
 })
 
 lines = table.to_latex(escape=False).split("\n")

### 2020-09-08 eval/scaled_weights.py

@@ -13,12 +13,13 @@ import glob
 import os
 
 base = "exp/"
+cwd = os.getcwd()
 paths = glob.glob(base + "scaled_weights/*.json")
 data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
     'weight_factor': exp['factor'],
-    'graph': { 'data/europe/': 'DIMACs Europe', 'data/osm_ger/': 'OSM Germany', 'data/osm_europe/': 'OSM Europe' }[run['args'][1]],
+    'graph': { cwd + '/data/europe/': 'DIMACs Europe', cwd + '/data/osm_ger/': 'OSM Germany', cwd + '/data/osm_europe/': 'OSM Europe' }[run['args'][1]],
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 

### 2020-09-08 Rakefile.rb
more path fixes

@@ -160,9 +160,9 @@ namespace "exp" do
         filename = "#{exp_dir}/preprocessing/" + `date --iso-8601=seconds`.strip + '.out'
         sh "echo '#{graph}' >> #{filename}"
         sh("code/compute_ch/build/compute_ch #{graph}/first_out #{graph}/head #{graph}/lower_bound " +
-          "#{graph}/lower_bound_ch/order " +
-          "#{graph}/lower_bound_ch/forward_first_out #{graph}/lower_bound_ch/forward_head #{graph}/lower_bound_ch/forward_weight " +
-          "#{graph}/lower_bound_ch/backward_first_out #{graph}/lower_bound_ch/backward_head #{graph}/lower_bound_ch/backward_weight " +
+          "/dev/null " +
+          "/dev/null /dev/null /dev/null " +
+          "/dev/null /dev/null /dev/null " +
           ">> #{filename}")
       end
     end

### 2020-09-09 eval/applications.py
fix application table for publically available data

@@ -12,6 +12,7 @@ cwd = os.getcwd()
 paths = glob.glob(base + "applications/*.json")
 data = [json.load(open(path)) for path in paths]
 
+only_public = 'ONLY_PUBLIC' in os.environ
 
 queries = pd.DataFrame.from_records([{
     'exp': exp['experiment'],
@@ -77,12 +78,18 @@ table = table.join(dijkstra_queries.groupby(['graph', 'exp']).agg(dijkstra_runni
 table['speedup'] = table['dijkstra_running_time_ms'] / table['running_time_ms']
 table = table.round(1)
 
+if only_public:
+    ger_name = R'\multirow{4}{*}{OSM Ger}'
+else:
+    ger_name = R'\multirow{8}{*}{OSM Ger}'
+
+
 table = table.rename(index={
     cwd + '/data/ger06': 'TDGer06',
     cwd + '/data/ptv17': 'TDEur17',
     cwd + '/data/osm_europe/': 'OSM Europe',
-    cwd + '/data/osm_ger/': R'\multirow{8}{*}{OSM Ger}',
-    cwd + '/data/osm_ger_td/': R'\multirow{8}{*}{OSM Ger}',
+    cwd + '/data/osm_ger/': ger_name,
+    cwd + '/data/osm_ger_td/': ger_name,
     cwd + '/data/europe/': 'DIMACs Europe',
     'perfect': 'Unmodified ($w_q=w_\\ell$)',
     'no_highways': 'No Highways',
@@ -103,7 +110,12 @@ lines = table.to_latex(escape=False).split("\n")
 lines = lines[:2] + [
   R" & &   Running &                Queue &     Length & Dijkstra & Speedup \\"
   R" & & time [ms] & $[\cdot 10^3]$ & incr. [\%] &     [ms] &         \\"
-] + lines[4:13] + ["\\addlinespace"] + lines[13:]
+]
+
+if only_public:
+    lines += lines[4:]
+else:
+    lines[4:13] + ["\\addlinespace"] + lines[13:]
 
 output = "\n".join(lines) + "\n"
 output = re.sub(re.compile('([0-9]{3}(?=[0-9]))'), '\\g<0>,\\\\', output[::-1])[::-1]

### 2020-09-09 eval/applications.py
fix lines

@@ -105,17 +105,17 @@ table = table.rename(index={
     'random_times_2': 'Random $\\times 2$',
 })
 
-lines = table.to_latex(escape=False).split("\n")
+orig_lines = table.to_latex(escape=False).split("\n")
 
-lines = lines[:2] + [
+lines = orig_lines[:2] + [
   R" & &   Running &                Queue &     Length & Dijkstra & Speedup \\"
   R" & & time [ms] & $[\cdot 10^3]$ & incr. [\%] &     [ms] &         \\"
 ]
 
 if only_public:
-    lines += lines[4:]
+    lines += orig_lines[4:]
 else:
-    lines[4:13] + ["\\addlinespace"] + lines[13:]
+    lines += orig_lines[4:13] + ["\\addlinespace"] + orig_lines[13:]
 
 output = "\n".join(lines) + "\n"
 output = re.sub(re.compile('([0-9]{3}(?=[0-9]))'), '\\g<0>,\\\\', output[::-1])[::-1]

### 2020-09-09 paper/ch_potentials.tex
add line numbers
better anonymous author

@@
 
+\usepackage[switch]{lineno}
+
 \pdfinfo{
 /Title (A Fast and Tight Heuristic for A* in Road Networks)
-/Author (Foo Bar)
+/Author (Marvin the Paranoid Android)
 /Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
@@
 \title{A Fast and Tight Heuristic for A* in Road Networks}
-\author{Foo Bar}
+\author{Marvin the Paranoid Android}
 
@@
 \begin{document}
+\linenumbers
 

### 2020-12-23 eval/building_blocks.py
add ALT exp

@@ -25,6 +25,8 @@ def heuristic_name(feature_flags):
         return 'Zero'
     elif 'CHPOT_ORACLE' in feature_flags:
         return 'Oracle'
+    elif 'CHPOT_ALT' in feature_flags:
+        return 'ALT'
     else:
         return 'CH'
 
@@ -34,9 +36,10 @@ table = queries.groupby(['algo', 'Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3
 table['num_queue_pushs'] = table['num_queue_pushs'] / 1000
 table = table.round(1)
 
-table = table.reindex(['Zero', 'CH', 'Oracle'], level=1)
+table = table.reindex(['Zero', 'ALT', 'CH', 'Oracle'], level=1)
 table = table.rename(index={
     'Zero': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{Zero}}',
+    'ALT': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{ALT}}',
     'CH': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{CH-\\Potentials}}}',
     'Oracle': R'\multirow{2}{*}{Oracle}',
 })
@@ -49,7 +52,7 @@ lines = [R'\begin{tabular}{clllrr}'] + [lines[1]] + [
     R" & & & & Running &     Queue \\",
     R" Heur. & BCC & Deg2 & Deg3 & time [ms] & [$\cdot 10^3$] \\"
 # ] + lines[4:]
-] + [lines[4], lines[17]] + lines[6:9] + ["\\addlinespace"] + lines[9:13] + ["\\addlinespace"] + [lines[13], lines[16]] + lines[20:]
+] + [lines[4], lines[21]] + lines[6:9] + ["\\addlinespace"] + lines[9:13] + ["\\addlinespace"] + lines[13:17] + ["\\addlinespace"] + [lines[17], lines[20]] + lines[24:]
 
 output = "\n".join(lines) + "\n"
 output = output.replace('True', '\\cmark')

### 2020-12-23 eval/building_blocks.py
more alt updates

@@ -52,7 +52,7 @@ lines = [R'\begin{tabular}{clllrr}'] + [lines[1]] + [
     R" & & & & Running &     Queue \\",
     R" Heur. & BCC & Deg2 & Deg3 & time [ms] & [$\cdot 10^3$] \\"
 # ] + lines[4:]
-] + [lines[4], lines[21]] + lines[6:9] + ["\\addlinespace"] + lines[9:13] + ["\\addlinespace"] + lines[13:17] + ["\\addlinespace"] + [lines[17], lines[20]] + lines[24:]
+] + [lines[4], lines[21]] + lines[6:9] + ["\\addlinespace"] + lines[9:13] + ["\\addlinespace"] + lines[13:17] + ["\\addlinespace"] + [lines[17], lines[20]] + lines[25:]
 
 output = "\n".join(lines) + "\n"
 output = output.replace('True', '\\cmark')

### 2020-12-23 paper/rebuttal.txt
rebuttal

First of all, we would like to thank all reviewers for their constructive feedback.
We hope that we get the chance to incorporate it into our paper.

Reviewer specific responses:

Reviewer #1
In 11. Reasons to Reject you asked for a comparison with Contraction Hierarchies running times.
We had hoped that lines 531 to 539 would address this sufficiently.
However, we did not provide any CH query times.
This is definitely something that we should improve.
We conducted the experiment and CH queries take on average 0.163ms in our experiments.
This confirms our statement that CH-Potentials converge gracefully to a CH.
If we get the opportunity, we would add this information to the paragraph.

In 13. Questions for the Authors you asked how we can prove the flexibility of our algorithm.
The flexibility of our approach is shown by the numerous problem settings and extensions that our approach can handle.
For each of these problem settings there are often entire lines of research with several papers to extend algorithms like CH to these settings.
CH-Potentials can handle all of these basically without modification and with very little implementation complexity.

Reviewer #2
In 13. Questions for the Authors you ask how the query weights are provided.
This very much depends on the application.
We choose the w_q formulation as a somewhat simplified formalization for the paper.
But the core strength of CH-Potentials is that w_q can be basically anything (a modified weight, a time-dependent function, a weight from an expanded graph, ...) and and can be provided in a number of different ways depending on the application.

For example, in the case of turn costs, w_q really is two edges from the turn-expanded graph (which is in fact known at preprocessing time).
But instead of reengineering CH/CCH preprocessing to mitigate the slowdown for preprocessing on the expanded graph [cite, cite] we just compute a CH on the original graph and run A* on the expanded graph.

In the time-dependent (traffic predictions) case, w_q is a function (also known at preprocessing time).
However adapting CH [cite]/CCH [cite]/CRP [cite] preprocessing to this setting proved to be very challenging (computing the travel time functions for shortcuts is quite involved and expensive. This is also a reason, why we skip over nodes instead of using shortcuts like Goldberg for Core-ALT).
For CH-Potentials, we just need to compute a CH on the lower bound graph and run A* on the time-dependent graph.

In the live-traffic scenario there are two ways, that w_q could be provided.
When w_q is an array in memory which would be constantly modified between queries, CRP/CCH would not fare very well, because one would need to run the customization for every query - which would take a couple of seconds per query - even with parallelization.
However, when w_q is provided as a completely new set of weights between a bigger number of queries, than CRP/CCH would indeed be a better choice for a practical application.
But this is not a weakness of CH-Potentials.
In fact, it is completely possible to replace the CH in CH-Potentials with a CCH to combine the benefits of customization with the flexibility of CH-Potentials.
This is actually implemented in our code (see Supplementary Material code/rust_road_router/engine/src/algo/ch_potentials.rs) and we also performed experiments with CCH-Potentials, but did not report the results due to space constraints and to keep the paper simpler.

Reviewer #3
Preprocessing

Reviewer #4
In 11. Reasons to Reject you asked for using more/different road networks.
In fact, we also performed experiments on other networks (including the standard DIMACs Europe benchmark instance) but the results did not provide any additional insights.
Thus, and due to the constrained space, we omitted them.
Adding them again would be easily possible, if this helps.

Thank you for the feedback in 12.
In case we get the opportunity, we will try make it clearer where our contribution starts.
Regarding citations for section 5: We are only aware of the work in [Topocore].
Which papers were we missing?

Regarding the question from In 13. Questions for the Authors:
1. Yes, the approach can be generalized to degree N
2. We do not think that its worthwhile due to the reasons state in [Topocore]
3. Yes, the benefit is graph specific: we can only skip low degree nodes if there are any. However, due to the reasons discussed in [Topocore], skipping nodes of degree higher than 3 should never pay off.

General
  Reproducibility

Loading graph ... done [9040ms]
Undirecting graph ... done [2980ms]
Computing order ... done [1685ms]
Checking if order valid ... done [84ms]
Saving order ... done [1152ms]
Testing if correctly saved ... done [49ms]

### 2020-12-23 paper/ch_potentials.tex
switch to lipics

--- 2020-09-08 paper/ch_potentials.tex
@@
-\def\year{2021}\relax
-%File: formatting-instructions-latex-2021.tex
-%release 2021.1
-\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
-\usepackage{aaai21}  % DO NOT CHANGE THIS
-\usepackage{times}  % DO NOT CHANGE THIS
-\usepackage{helvet} % DO NOT CHANGE THIS
-\usepackage{courier}  % DO NOT CHANGE THIS
-\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
-\usepackage{graphicx} % DO NOT CHANGE THIS
-\urlstyle{rm} % DO NOT CHANGE THIS
-\def\UrlFont{\rm}  % DO NOT CHANGE THIS
-\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
-\frenchspacing  % DO NOT CHANGE THIS
-\setlength{\pdfpagewidth}{8.5in}  % DO NOT CHANGE THIS
-\setlength{\pdfpageheight}{11in}  % DO NOT CHANGE THIS
+\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
 
@@
 
-\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.
+% \setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.
 
@@
 
-\pdfinfo{
-/Title (A Fast and Tight Heuristic for A* in Road Networks)
-/Author (Foo Bar)
-/Keywords (shortest path, road graphs, goal-directed search, contraction hierarchy, heuristic search)
-}
+\bibliographystyle{plainurl} % the mandatory bibstyle
+
 \title{A Fast and Tight Heuristic for A* in Road Networks}
-\author{Foo Bar}
+\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{}{}
+\author{Tim Zeitz}{Karlsruhe Institute of Technology, Germany}{tim.zeitz@kit.edu}{https://orcid.org/0000-0003-4746-3582}{}
 
-% \author{Ben Strasser\\
-% academia@ben-strasser.net\\
-% \And
-% Tim Zeitz\\
-% tim.zeitz@kit.edu\\
-% Karlsruhe Institute of Technology
-% }
+\authorrunning{B. Strasser and T. Zeitz}
+\Copyright{Ben Strasser and Tim Zeitz}
 
-%%
-%% The "author" command and its associated commands are used to define
-%% the authors and their affiliations.
-%% Of note is the shared affiliation of the first two authors, and the
-%% "authornote" and "authornotemark" commands
-%% used to denote shared contribution to the research.
-%\author{
-%Ben Strasser
+\ccsdesc[500]{Theory of computation~Shortest paths}
+\ccsdesc[300]{Mathematics of computing~Graph algorithms}
+\ccsdesc[500]{Applied computing~Transportation}
 
-%\author{Tim Zeitz}
-%tim.zeitz@kit.edu}
-%\affiliation{%
-%  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
-%  \city{Karlsruhe}
-%  \country{Germany}
-%}
+\keywords{route planning, shortest paths}
 
@@
 \bibliography{dblp,references}
-\bibliographystyle{aaai}
 \end{document}

### 2021-01-15 paper/ch_potentials.tex
new story for sea

@@
 
+\usepackage{todonotes}
 \usepackage{booktabs}
@@
 
-\keywords{route planning, shortest paths}
+\keywords{route planning, shortest paths, realistic road networks}
 
@@
 \begin{abstract}
-Efficiently computing flexible and exact routes in large road networks is a tough challenge.
+We study exact, efficient and practical algorithms for route planning in large road networks.
+Realistic routing often requires integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
-A* is a classical approach to accelerate Dijkstra's algorithm and is used in many fields such as Experimental Algorithms, Robotics, and Artificial Intelligence.
-However, A*'s performance depends on the availability of a good heuristic.
-Computing a good heuristic is a challenge of its own.
-Shortest paths can also be quickly computed using hierarchical speedup techniques.
-They are not based on A* and do not use a heuristic.
-Hierarchical techniques are known to work well on road networks.
-They achieve speed and exactness but sacrifice A*'s flexibility.
-In this paper, we use Contraction Hierarchies (CH), a hierarchical technique, to compute a A* heuristics.
-We call our technique CH-Potentials.
-With this approach, we combine some of the speed of hierarchical techniques with the flexibility offered by A* - the best of both worlds - while retaining exactness.
+A* is a classical approach to accelerate Dijkstra's algorithm.
+A* can support many extended scenarios without much additional implementation complexity.
+However, A*'s performance depends on the availability of a good heuristic to estimate distances.
+Computing tight distance estimates is a challenge on its own.
+\todo{tight}
+On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
+They achieve speed and exactness but sacrifice A*'s flexibility and are not as extensible.
+In this paper, we present an algorithm to efficiently extract distance estimates for A* from Contraction Hierarchies (CH), a hierarchical technique.
+We call our heuristic CH-Potentials.
+Our approach allows decoupling the supported extensions from the hierarchical speed-up technique.
 Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
@@
 
+\newpage
+
 \section{Introduction}
 \label{sec:intro}
-Route planning in large street networks is a well-researched topic with many applications~\cite{bdgmpsww-rptn-16}.
-It is often formalized as a variant of the classical shortest path problem in weighted graphs.
-Developing flexible techniques to efficiently compute exact routes is a tough challenge.
-\emph{Exact} means that a path of provable minimum weight is computed.
-\emph{Efficiently} means that the path computation is fast.
-Being able to change the edge weights between path computations or handling complex weights, such as edge weight functions, is \emph{flexibility}.
-% finde es klingt stringenter, wenn jede Eigenschaft genau einen Satz hat.
+The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
+Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
+Road intersections become nodes.
+Road segments become edges.
+The weight of an edge is the time a car needs to traverse the road segment.
+The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
+However, on continental sized networks, it is too slow for many applications.
+Thus, speed-up techniques have been developed.
+A popular example for such a technique, which has been successfully used in many real world applications, is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
+CH exploit the inherent hierarchy of road networks.
+In a preprocessing step, additional shortcut edges are inserted, which allow skipping over unimportant parts of the network during query time.
+Another technique, also successfully employed in practice, is Multi-Level-Dijkstra~(MLD)~\cite{dgpw-crprn-13} better known as CRP.
+It also exploits the hierarchy of road networks using shortcut edges.
+Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
+Their implementation complexity is also quite manageable.
+This makes them very attractive in practice.
 
 \begin{figure}
-
 \centering
-\includegraphics[width=\columnwidth]{fig/searchspace_st.png}
-
-\caption{Nodes explored by A* with CH-Potentials. Color indicates the node removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
+\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
+\caption{Nodes explored by A*. Color indicates the node removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
 \label{img:search-space}
@@
 
-Dijkstra's algorithm~\cite{d-ntpcg-59} achieves flexibility and exactness but is comparatively slow.
-It iteratively explores nodes ordered by their distance from the start.
-A classic speedup approach is the well-known A* algorithm~\cite{hnr-afbhd-68}.
-It processes nodes in a different order, preferring nodes closer to the destination.
-Which nodes are considered close is encoded using a heuristic function.
-While flexible, A*'s performance crucially depends on the quality of this function.
-Finding good heuristics that can be evaluated efficiently is difficult.
-Constructing good heuristics on road networks, such as the one depicted in Figure~\ref{img:search-space}, is the focus of this paper.
+However, for many real world routing applications, this basic model turns out to be too simple.
+For realistic routing, many additional features need to be considered.
+This includes but is not limited to turn costs and restrictions, live traffic, user preferences, traffic predictions, temporary driving bans for certain types of vehicles and many more.
+Hierarchical speed-up techniques can be extended to support these features.
+However, each feature requires significant effort.
+Many combinations of specific techniques with certain extensions have their own dedicated research paper(s).
+Adding these features introduces significant additional implementation complexity.
+Supporting combinations of extensions is at least non-trivial, oftentimes downright impossible.
 
-In this paper, we consider a setup with an offline preprocessing phase and an online query phase.
-The preprocessing phase can be slow and computes auxiliary data that can be used in the query phase, which should be fast.
-The input to the preprocessing phase is a directed graph $G=(V,E)$ weighted with \emph{lower bound weights} $w_\ell$.
-The query phase is given the \emph{source} $s$ and the \emph{target} $t$ as input.
-$s$ is also called start and synonyms for $t$ are goal or destination.
-It is further given the \emph{query edge weights} $w_q$.
-The task is to compute a minimum length path with respect to $w_q$.
-Of $w_q$, it is only known that $w_\ell$ is a lower bound, i.e., $w_\ell \le w_q$.
-%
-While the preprocessing phase may be slow, it must eventually finish.
-On graphs with millions of nodes, quadratic preprocessing time is prohibitive.
+In this work, we propose a novel building block for routing algorithms which drastically improves on this extensibility problem with very little implementation complexity.
+For this, we revisit goal-directed search (A*).
+Goal directed search is a simple extension of Dijkstra's algorithm.
+An estimate of the remaining distance to the target is used, to change the order in which nodes are traversed during the search.
+This guides the search towards the target.
+See Figure~\ref{img:search-space} for an example of nodes traversed during a goal-directed search.
+This approach is far easier to extend.
+Many extensions can be supported without modifying the distance estimation at all.
+The downside is that the effectiveness of this approach highly depends on the tightness of the estimate.
+There has been extensive research on goal-directed search in the past~\cite{}.
+However, on road networks, previous approaches could not deliver performance on par with hierarchical techniques.
+We introduce CH-Potentials, a new distance estimation function which can efficiently compute tight distance estimates.
+\todo{tight}
+CH-Potentials are a simple combination of existing approaches.
+CH-Potentials use hierarchical speed-up techniques but only indirectly through goal directed search.
+This new building block allows handling extensions orthogonal to the speed-up technique.
+\todo{contribution - no better heuristic possible}
+% By comparing against Oracle-A*, we show that no other exact two-phase A*-based approach can be significantly faster.
+In addition, we propose optimizations for Dijkstra's/A*'s handling of low-degree nodes, common in road networks.
 
-In this paper, we consider an algorithm, where the query phase mostly consists of an A* search.
-In the preprocessing phase, it computes a Contraction Hierarchy (CH)~\cite{gssv-erlrn-12}.
-The CH is used to efficiently evaluate the heuristic.
-Our setup is very similar to the well-known ALT technique~\cite{gh-cspas-05,DBLP:conf/wea/DellingW07}.
-Following the ALT terminology, we call our algorithm \emph{CH-Potentials}, as it combines ideas from Contraction Hierarchies (CH)~\cite{gssv-erlrn-12} with A* heuristics.
+The rest of the paper is organized as follows.
+In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
+CH-Potentials, our new distance estimation function is introduced in Section~\ref{sec:main-algo}.
+Section~\ref{sec:low-deg-improvment} discusses our improvements for the handling of low-degree nodes.
+In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply our technique to different practical applications.
+Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
 
-A heuristic $h$ is a function that maps nodes onto numbers.
-$h$ depends on the target $t$ but not on the source $s$.
-Without loss of generality, we assume that $h(t)=0$.
-The \emph{tight} heuristic assigns to every node $x$, the length of a shortest $xt$-path.
-All heuristics, for which A* is exact, are lower bounds of the tight heuristic.
-CH-Potentials compute the tight heuristic with respect to $w_\ell$.
-Without access to the query weights $w_q$, which are unavailable during preprocessing, no value of the heuristic can be increased, without loosing A*'s exactness.
-In this sense, CH-Potentials are as tight as possible and thus optimal.
-We say that a heuristic that is tight with respect to $w_\ell$ is \emph{preprocessing-tight}.
+\section{Related Work}\label{sec:related_work}
 
-\section{Related Work}
+% A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
+% In this setting, mostly hierarchical approaches dominate in terms of running time~\cite{bdgmpsww-rptn-16}.
+% A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12}.
 
-A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
-In this setting, mostly hierarchical approaches dominate in terms of running time~\cite{bdgmpsww-rptn-16}.
-A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12}.
-
-There is a lot of work that extends CH to handle more complex settings.
-For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
-In~\cite{dsw-cch-15}, CH is modified to support Live-Traffic.
-Historic traffic patterns are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13,bdpw-dtdrp-16}.
+There is a lot of work that extends hierarchical speed-up techniques to more complex settings.
+\todo{cite overview}
+% For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
+MLD/CRP~\cite{dgpw-crprn-13} was developed to allow exchanging weights in reasonable time.
+This enable user preferences and live traffic.
+MLD/CRP also supports turn costs.
+Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
+In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH), which also supports exchanging weights and thus supporting user preferences and live traffic.
+CH is extended to support turn costs in~\cite{}, CCH in~\cite{bwzz-cchtc-20}.
+Traffic predictions are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13}.
+\todo{cite correct CATCHUp Version}
+\todo{improve! and discuss hierarchical extensions and what they achieve and their difficulties in more details}
 Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
@@
 CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s' flexibility.
+However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s flexibility.
+\todo{cite traffic aware routing}
+%
 
@@
 Just as our approach, ALT is A*-based.
-However, ALT's heuristic is not preprocessing-tight.
+% However, ALT's heuristic is not preprocessing-tight.
 %
@@
 The Euclidean distance is then used as A* heuristic.
-This results in a good but not a preprocessing-tight heuristic.
+% This results in a good but not a preprocessing-tight heuristic.
 %
@@
 
-\subsection{Oracle-A* \& Comparison with Related Work}
-\label{sec:oracle}
+\section{Algorithm}\label{sec:main-algo}
 
-To compare against related work, we introduce the concept of Oracle-A*.
-It has access to a shortest distance array with respect to $w_\ell$, i.e., it is preprocessing-tight.
-The array computation time is unaccounted for, i.e., it is magically filled instantly.
-Oracle-A* is not achievable in practice.
-However, it is a lower bound for every algorithm following our two phase setup.
-We experimentally show that CH-Potentials are within a factor two of Oracle-A* and thus every related work.
+In this section, we first discuss the framework in which CH-Potentials can be used.
+Then, we describe the building blocks of CH-Potentials: Contraction Hierarchies and PHAST, a CH extension.
+Finally, we introduce the CH-Potentials heuristic.
 
-\subsection{Applications need Flexibility}
+\subsection{Framework}
 
-Many applications need flexibility.
-The most common ones are traffic avoidance, user preferences, and vehicle restrictions.
-Traffic avoidance comes in two forms: Live~\cite{dgpw-crprn-13,dsw-cch-15} and predicted traffic~\cite{ndls-bastd-12,bgsv-mtdtt-13}.
-Predicted traffic is often called time-dependent routing.
-%
-User preferences can vary significantly~\cite{DBLP:conf/gis/FunkeS15,DBLP:conf/gis/DellingGGKTW15,DBLP:conf/gis/FunkeLS16}.
-For example, for some users shorter routes are more important than fast routes.
-Others want to avoid tunnels or highways.
-%
-Flexibility is also often used as building block in multi-agent planning settings~\cite{DBLP:journals/ai/SharonSFS15,DBLP:journals/tciaig/BaierBHH15}.
-%
-\section{Contribution \& Outline}
+We consider a two-phase setup.
+In the preprocessing phase, we are given a lower bound graph $G_\ell$ with lower bound weights $w_\ell$.
+The lower bound weights are single scalar numbers for each edge.
+We now perform some precomputation, i.e. apply the Contraction Hierarchy preprocessing to this graph.
+This generates some auxiliary data which we will use in the query phase to efficiently compute distance estimates.
+We will do so with the heuristic function $h_t(x)$ where $x$ and $t$ are nodes from $G_\ell$.
+The heuristic function returns the shortest distance from $x$ to $t$ with respect to $w_\ell$.
+Before each query, the heuristic has to be initialized with the target node $t$.
+We will discuss the implementation of this heuristic function and the initialization later in this section.
 
-We introduce CH-Potentials, a two-phase technique to efficiently compute preprocessing-tight heuristics, in Section~\ref{sec:main-algo}.
-Combined with A*, this yields an exact, efficient, and flexible route planning algorithm.
-Beside the heuristic, in Section\ref{sec:low-deg-improvment}, we improve A*'s handling of low degree nodes, common in road networks.
-In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing algorithms for various extended route planning applications.
-Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
-By comparing against Oracle-A*, we show that no other exact two-phase A*-based approach can be significantly faster.
+In the query phase, we have a query graph $G_q$ with query weights $w_q$ and a mapping $T$ which maps nodes from $G_q$ to $G_\ell$.
+For most applications, $G_q = G_\ell$ and $T$ will be the identity.
+The query weights $w_q$ are the main mechanism to implement different applications with CH-Potentials.
+In some cases, $w_q$ will be simple scalar values like the lower bound weights.
+However, for $w_q$ this is not a requirement.
+It can be any function that computes a weight for an edge.
+This function can also take additional parameters from the state of the search.
 
-\section{Algorithm Description}
+The query phase then is to execute A* on $G_q$.
+Before running the search, the heuristic will be initialized with the target $T(t)$.
+The distance estimate for a node $x$ from $G_q$ can be computed as $h_{T(t)}(T(x))$.
+For CH-Potentials to find correct distances, $w_q \geq w_\ell$ has to hold for all arcs.
+\todo{Correctness Argument?}
+\todo{Describe A*?}
 
-In this section, we first describe Contraction Hierarchies, then PHAST, a CH extension, and finally CH-Potentials.
-
-\label{sec:main-algo}
+\subsection{Contraction Hierarchy (CH)}
 
@@
 
-\subsection{Contraction Hierarchy (CH)}
-
 A CH is a two phase technique to efficiently compute exact, shortest paths.
-It is not flexible, i.e., requires $w_q=w_\ell$.
 For a details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
@@
 
-Using a CH, we can compute a simple and preprocessing-tight heuristic.
-In the preprocessing phase, a CH with respect to $w_\ell$ is computed.
-The heuristic evaluation $h(x)$ performs a CH-query from $x$ to $t$.
+Using the CH query algorithm, we can already give a simple heuristic.
+The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
 %
@@
 Using PHAST, we can also compute a preprocessing-tight A* heuristic.
+\todo{preprocessing tight}
 In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
@@
 
-\section{Algorithm Extensions}
+\section{Applications}
 \label{sec:extensions}
@@
 A* is a flexible algorithm.
-Many extended problems can therefore be solved using the CH-Potentials algorithm.
+Many extended problems can therefore be solved using the CH-Potentials framework.
 In this section, we describe some extended problems and how they can be solved with CH-Potentials.
@@
 With CH-Potentials, we get the preprocessing-tight heuristic $h$.
+\todo{prepro tight}
 We set $h'(x,y) = h(y)$.
@@
 
+\todo{talk about CCH-Pot}
+
 Beside predicted traffic, we also consider live traffic.
@@
 Further, a trade-off between arrival time and route quality is considered.
+\todo{cite correct Closures Version}
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{fig/scaled_weights.pdf}
-\caption{
-Running times on a logarithmic scale for queries on OSM Ger with scaled edge weights $w_q = \alpha \cdot w_\ell$.
-The boxes cover the range between the first and third quartile.
-The band in the box indicates the median, the diamond the mean.
-The whiskers cover 1.5 times the inter quartile range.
-All other running times are marked as outliers.
-}\label{fig:scaled_weights}
-\end{figure}
-
-\begin{table}
-\centering
-\caption{Average query running times and number of queue pushs with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
-\input{table/building_blocks.tex}
-\end{table}
-
-\begin{table*}
-\centering
-\caption{
-CH-Potentials performance for different route planning applications.
-We report average running times and number of queue pushs.
-We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
-Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup over this baseline.
-}\label{tab:applications}
-\input{table/applications.tex}
-\end{table*}
-
 In this section, we present our experimental evaluation.
@@
 All experiments were performed sequentially.
-Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile. % with the target-cpu=native option.
+Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile with the target-cpu=native option.
 
@@
 \subparagraph{Experiments}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/scaled_weights.pdf}
+\caption{
+Running times on a logarithmic scale for queries on OSM Ger with scaled edge weights $w_q = \alpha \cdot w_\ell$.
+The boxes cover the range between the first and third quartile.
+The band in the box indicates the median, the diamond the mean.
+The whiskers cover 1.5 times the inter quartile range.
+All other running times are marked as outliers.
+}\label{fig:scaled_weights}
+\end{figure}
+
 The first experiment varies the tightness of the heuristic.
@@
 The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
-When $\alpha = 1$, the heuristic is tight.
+% When $\alpha = 1$, the heuristic is tight.
 Increasing $\alpha$ degrades the heuristic's quality.
 Figure~\ref{fig:scaled_weights} depicts the results.
+\todo{decrease figure height}
 Clearly, $\alpha$ has significant influence on the running time.
@@
 After $\alpha = 1.1$ the growth slows down.
+This illustrates both the strengths and limits of our approach and goal directed search in general.
+Goal directed search (and thus CH-Potentials) can only achieve competitive running times if the application allows for a sufficiently tight heuristic.
+
 We observe that the running times for a fixed $\alpha$ vary strongly.
@@
 
-Table~\ref{tab:building_blocks} shows the impact of our optimizations.
-We run queries with three heuristics.
-The Zero heuristic is simply $h(x)=0$ for all $x$.
+\begin{table}
+\centering
+\caption{Average query running times and number of queue pushs with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
+\input{table/building_blocks.tex}
+\end{table}
+
+Table~\ref{tab:building_blocks} depicts the performance of A* with different heuristics and optimizations.
+We compare CH-Potentials to three other heuristics.
+First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
 This corresponds to using Dijkstra's algorithm.
-Further, we run queries with our novel CH-Potential heuristic.
-Finally, we use the unattainable lower bound Oracle-A* as described in Section~\ref{sec:oracle}.
-Oracle-A* is an unbeatable theoretic lower bound.
-The reported running time of Oracle-A* do \emph{not} account for the heuristic evaluation.
-We include Oracle-A* as it is a lower bound not only for our algorithm but also related work.
+Second, we compare against our own implementation\footnote{
+We use 16 landmarks generate with the avoid strategy.
+We store all landmark distances consecutive for each node.
+This increases cache efficiency when reading many distances for the same node.
+We do not disable some landmarks during the query, but use all landmarks all the time.
+Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
+} of ALT~\cite{}.
+Finally we compare against a hypothetical \emph{Oracle-A*} heuristic.
+This heuristic has instant access to a shortest distance array with respect to $w_\ell$, that is it is the most efficient heuristic possible in our model.
+We fill this array before each query using a reverse Dijkstra search from the target node.
+Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
+Comparing against Oracle-A* allows us to measure the overhead of the heuristic evaluation of CH-Potentials.
+Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.
 
@@
 Each optimization reduces both queue pushes and running times.
-% Further, each of our optimizations reduces the number of queue pushes.
-Very interesting is the comparison with Oracle-A*.
-Our algorithm is only 1.6 times slower than the unachievable theoretic lower bound.
-This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster than CH-Potentials.
+All optimizations yield a combined speed-up of around 3.
+% Skipping degree three nodes has the smallest impact.
+CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly less nodes.
+This is not surprising, since ALT computes worse distance estimates.
+In contrast, CH-Potentials already compute exact distances with respect to $w_\ell$.
+The number of popped nodes is the same for CH-Potentials and Oracle-A*.
+The only difference between CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
+This overhead leads to a slowdown of around 1.6.
+Thus, CH-Potentials are already very close to the best possible heuristic in this model.
+% Very interesting is the comparison with Oracle-A*.
+% Our algorithm is only 1.6 times slower than accessing the exact $w_\ell$ distances instantly from a precomputed array.
+This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
+
+\begin{table*}
+\centering
+\caption{
+CH-Potentials performance for different route planning applications.
+We report average running times and number of queue pushs.
+We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
+Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup over this baseline.
+}\label{tab:applications}
+\input{table/applications.tex}
+\end{table*}
 
@@
 Speedups compared to extensions of Dijkstra's algorithm for the respective application are depicted.
-We start with the inflexible routing problem where $w_q = w_\ell$.
-This is the problem variant solved by the basic CH algorithm.
-In this setting, there is a huge speedup of 3243.
+We start with the base case where $w_q = w_\ell$.
+This is the problem variant solved by the basic CH algorithm (a CH query takes 0.163\,ms on average on this graph).
+CH achieves query running times of 0.16\,ms on average.
+CH-Potentials are roughly four times slower but still achieve a huge speedup over Dijkstra of 3243.
 Such large speedups are typical for CH.
-This shows that CH-Potentials gracefully converges to a CH in the $w_q = w_\ell$ special case.
+This shows that CH-Potentials gracefully converges toward a CH in the $w_q = w_\ell$ special case.
 
-In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the preprocessing-tight heuristic.
+In the other scenarios, the performance of CH-Potentials strongly depends on the quality of the heuristic.
 We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
@@
 We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
+\todo{preprocessing tight}
 Achieving significantly smaller running times could still be possible in variations of the problem setting.
@@
 Hierarchical techniques have been shown to work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
+\todo{der paragraph war eher so fuer die AAAI, oder?}
 

### 2021-01-16 paper/ch_potentials.tex
intro changes

@@
 Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
-Road intersections become nodes.
-Road segments become edges.
-The weight of an edge is the time a car needs to traverse the road segment.
+Nodes represent intersections.
+Roads are modeled using edges.
+Edges are weighted by their traversal times.
 The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
-However, on continental sized networks, it is too slow for many applications.
+Unfortunately, on continental sized networks, it is too slow for many applications.
 Thus, speed-up techniques have been developed.
-A popular example for such a technique, which has been successfully used in many real world applications, is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
-CH exploit the inherent hierarchy of road networks.
-In a preprocessing step, additional shortcut edges are inserted, which allow skipping over unimportant parts of the network during query time.
-Another technique, also successfully employed in practice, is Multi-Level-Dijkstra~(MLD)~\cite{dgpw-crprn-13} better known as CRP.
-It also exploits the hierarchy of road networks using shortcut edges.
+One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
+They have been used successfully in many real world applications\todo{citation?}.
+A CH exploits the inherent hierarchy of road networks.
+In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant roads during query time.
+A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
+It is also used in practice~\todo{citation, bing blog?}.
+MLD also also uses shortcut edges.
 Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
-Their implementation complexity is also quite manageable.
-This makes them very attractive in practice.
+% --- Unser Argument ist, dass die zu komplex zum erweitern sind... da passen diese 
+%     Sätze nicht in die Story
+% Their implementation complexity is also quite manageable.    
+% This makes them very attractive in practice.
 
@@
 
-However, for many real world routing applications, this basic model turns out to be too simple.
-For realistic routing, many additional features need to be considered.
-This includes but is not limited to turn costs and restrictions, live traffic, user preferences, traffic predictions, temporary driving bans for certain types of vehicles and many more.
-Hierarchical speed-up techniques can be extended to support these features.
-However, each feature requires significant effort.
-Many combinations of specific techniques with certain extensions have their own dedicated research paper(s).
-Adding these features introduces significant additional implementation complexity.
-Supporting combinations of extensions is at least non-trivial, oftentimes downright impossible.
+Unfortunately, for many real world applications, this basic graph model is too simplistic.
+Many additional features need to be considered.
+% --- ich habe turn costs weg gelassen, weil wir die nicht in den experimenten haben, ich will keiner Reviewer auf blöde Ideeen bringen
+% --- ich habe temporary driving bans weg gelassen, weil wir das nicht in den Experimenten haben und ich "we can do all of the above features" claimen möchte
+This includes turn restrictions, live traffic, user preferences, traffic predictions.%, temporary driving bans for certain types of vehicles.
+Specific applications often have additional application-specific requirements.
+Extending Dijkstra's algorithm to support these features is usually easy.
+On the other hand, while extending hierarchical speed-up techniques is possible, the algorithm development is vastly more complex.
+For every feature, dedicated research paper(s) exist that extend CH.
+Supporting the combination of several features is even harder. 
+For example, we are not aware of any work combining all features mentioned above.
+In this paper, we describe an algorithm that handles the combination of all mentioned features.
 
-In this work, we propose a novel building block for routing algorithms which drastically improves on this extensibility problem with very little implementation complexity.
-For this, we revisit goal-directed search (A*).
-Goal directed search is a simple extension of Dijkstra's algorithm.
-An estimate of the remaining distance to the target is used, to change the order in which nodes are traversed during the search.
-This guides the search towards the target.
-See Figure~\ref{img:search-space} for an example of nodes traversed during a goal-directed search.
-This approach is far easier to extend.
-Many extensions can be supported without modifying the distance estimation at all.
-The downside is that the effectiveness of this approach highly depends on the tightness of the estimate.
-There has been extensive research on goal-directed search in the past~\cite{}.
-However, on road networks, previous approaches could not deliver performance on par with hierarchical techniques.
-We introduce CH-Potentials, a new distance estimation function which can efficiently compute tight distance estimates.
-\todo{tight}
-CH-Potentials are a simple combination of existing approaches.
-CH-Potentials use hierarchical speed-up techniques but only indirectly through goal directed search.
-This new building block allows handling extensions orthogonal to the speed-up technique.
-\todo{contribution - no better heuristic possible}
-% By comparing against Oracle-A*, we show that no other exact two-phase A*-based approach can be significantly faster.
-In addition, we propose optimizations for Dijkstra's/A*'s handling of low-degree nodes, common in road networks.
+% --- mit "novel" und "propose" muss man vorsichtig sein, wegen SEA paper von letztem Jahr
+We describe an algorithmic building block that mostly decouples feature development from the hierarchical speed-up algorithm.
+Our goal is to simplify incorporating new features.
+We build upon the A* algorithm \cite{A* literature}.
+A* is a goal-directed variant of Dijkstra's algorithm.
+It uses a heuristic\todo{heuristic vs potential} to guide the search towards the goal.
+A heuristic is function that maps a node $v$ onto a lower bound of the distance from $v$ to the goal.
+A*'s running time crucially depends on how tight this lower bound is.
+Further, evaluating the heuristic must be fast.
+The free flow travel time to the goal is a valid heuristic for most features.
+In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
+However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
+In this paper, we describe CH-Potentials, a technique that achieves both goals.
+Internally, the heuristic uses a CH.
+Fortunately, this is an implementation detail from the perspective of the A*.
+To support a new feature, we only need to modify the A* algorithm.
+The heuristic core containing the CH remains untouched.
+Extending A* is vastly easier than extending a CH.
+This enables us to design an algorithms that supports a multitude of features.
+In addition, we describe optimizations for handling of low-degree nodes, common in road networks.
+These low degree optimizations are applicable to Dijkstra's algorithm and A*.
 

### 2021-01-17 paper/ch_potentials.tex
readding turn costs
Consistent heuristic
formal setup

@@
 Many additional features need to be considered.
-% --- ich habe turn costs weg gelassen, weil wir die nicht in den experimenten haben, ich will keiner Reviewer auf blöde Ideeen bringen
 % --- ich habe temporary driving bans weg gelassen, weil wir das nicht in den Experimenten haben und ich "we can do all of the above features" claimen möchte
-This includes turn restrictions, live traffic, user preferences, traffic predictions.%, temporary driving bans for certain types of vehicles.
+This includes turn costs and restrictions, live traffic, user preferences, traffic predictions.%, temporary driving bans for certain types of vehicles.
 Specific applications often have additional application-specific requirements.
@@
 
+\subsection{Formal Setup: Inputs, Outputs, and Phases}
+
+Throughout this paper, we consider a two phase setup with an off-line preprocessing phase and an on-line query phase.
+The input to the preprocessing phase is data that is static, such as the road network and freeflow travel times.
+On the other hand, the input to the query phase changes a lot.
+For example, the source and the target nodes are query inputs.
+Further, the user preferences and live traffic are query inputs.
+The preprocessing phase may be slow and computes auxiliary data.
+The query phase obtains this auxiliary data as input and performs the path computation.
+
+Formally, the input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$.
+$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
+In this paper, $w_\ell$ is always the freeflow travel time.
+The output is auxiliary data that enables an efficient heuristic function $h_t(x)$.
+$h_t(x)$ is the exact distance from $x$ to $t$ in $G_\ell$.
+
+The formal inputs of the query phase depend on the considered features.
+In the simplest case, the input is a query weight $w_q$, and source and target nodes $s$ and $t$.
+It is guaranteed that $w_\ell(e) \le w_q(e)$ for all edges $e$.
+The output is a shortest $st$-path with respect to $w_q$.
+The query has access to the $h_t$ heuristic function.
+An example of the simple case is live-traffic.
+In this case, $w_q$ is an array of live travel times.
+
+With more complex features, the inputs become more complex.
+The extended query weight $w_q(e,\tau,p)$ has three parameters: the edge $e$, the edge entry time $\tau$, and the user preferences $p$.
+Beside $s$ and $t$ the input contains a departure time $\tau_s$.
+The distance of a path $e_1,e_2\ldots$ is defined as $w_q(e_1,\tau_s,p)+w_q(e_2,\tau_s+w_q(e_1,\tau_s,p),p)+\cdots$
+Further, the query input contains a graph $G_q$ with nodes and edges differing from $G_\ell$.
+To be able to use the $h$ heuristic, the query implementation need to have two node mapping functions $\phi$ and $\psi$ that relate nodes in $G_\ell$ and $G_q$ as follows:
+$h_{\phi(t)}(\psi(x))$ must lower bound the distance from $x$ to $t$ in $G_q$.
+
+Applications requiring many features have an inherent complexity.
+The departure and entry times are needed for prediction traffic.
+The user preference parameter is needed for tunnel and highway avoidance.
+The differentiation between $G_\ell$ and $G_q$ is needed to support turn costs and restrictions.
+
+This application complexity translates into the complex query inputs. 
+Fortunately, the preprocessing inputs are unaffected.
+A highlight of our technique is that all features can be supported without changing the preprocessing at all.
+
 \subsection{Framework}
@@
 For CH-Potentials to find correct distances, $w_q \geq w_\ell$ has to hold for all arcs.
-\todo{Correctness Argument?}
+\todo{Correctness Argument? Stichwort: Consistent/Montone heuristic}
 \todo{Describe A*?}

### 2021-01-18 paper/ch_potentials.tex
merge updates from Ben

--- 2021-01-16 paper/ch_potentials.tex
@@
 A CH exploits the inherent hierarchy of road networks.
-In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant roads during query time.
+In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network during query time.
 A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
@@
 Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
-% --- Unser Argument ist, dass die zu komplex zum erweitern sind... da passen diese 
-%     Sätze nicht in die Story
-% Their implementation complexity is also quite manageable.    
-% This makes them very attractive in practice.
 
@@
 Unfortunately, for many real world applications, this basic graph model is too simplistic.
-Many additional features need to be considered.
-% --- ich habe turn costs weg gelassen, weil wir die nicht in den experimenten haben, ich will keiner Reviewer auf blöde Ideeen bringen
-% --- ich habe temporary driving bans weg gelassen, weil wir das nicht in den Experimenten haben und ich "we can do all of the above features" claimen möchte
-This includes turn restrictions, live traffic, user preferences, traffic predictions.%, temporary driving bans for certain types of vehicles.
-Specific applications often have additional application-specific requirements.
+For realistic routing, many additional features need to be considered.
+This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions.%, temporary driving bans for certain types of vehicles.
+Some applications may have additional application-specific requirements.
 Extending Dijkstra's algorithm to support these features is usually easy.
-On the other hand, while extending hierarchical speed-up techniques is possible, the algorithm development is vastly more complex.
+Extending hierarchical speed-up techniques is also possible.
+However, the algorithm development is vastly more complex.
 For every feature, dedicated research paper(s) exist that extend CH.
@@
 For example, we are not aware of any work combining all features mentioned above.
-In this paper, we describe an algorithm that handles the combination of all mentioned features.
+In this paper, we describe an algorithmic building block, that allows handling the combination of all above mentioned features -- and probably more.
 
-% --- mit "novel" und "propose" muss man vorsichtig sein, wegen SEA paper von letztem Jahr
-We describe an algorithmic building block that mostly decouples feature development from the hierarchical speed-up algorithm.
-Our goal is to simplify incorporating new features.
-We build upon the A* algorithm \cite{A* literature}.
+Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{}.
 A* is a goal-directed variant of Dijkstra's algorithm.
-It uses a heuristic\todo{heuristic vs potential} to guide the search towards the goal.
-A heuristic is function that maps a node $v$ onto a lower bound of the distance from $v$ to the goal.
-A*'s running time crucially depends on how tight this lower bound is.
+See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
+A* uses a \emph{heuristic} to guide the search towards the goal.
+A heuristic is function that maps a node $v$ onto an estimate of the distance from $v$ to the goal.
+A*'s running time crucially depends on how tight this estimate is.
 Further, evaluating the heuristic must be fast.
-The free flow travel time to the goal is a valid heuristic for most features.
-In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
-However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
-In this paper, we describe CH-Potentials, a technique that achieves both goals.
+% The free flow travel time to the goal is a valid heuristic for most features.
+% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
+% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
+In this paper, we describe CH-Potentials, a heuristic that achieves both goals.
 Internally, the heuristic uses a CH.
@@
 Extending A* is vastly easier than extending a CH.
-This enables us to design an algorithms that supports a multitude of features.
-In addition, we describe optimizations for handling of low-degree nodes, common in road networks.
+This enables us to design algorithms that support a multitude of features.
+In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
 These low degree optimizations are applicable to Dijkstra's algorithm and A*.
@@
 
-\subsection{Framework}
+\subsection{Formal Setup: Inputs, Outputs, and Phases}
 
-We consider a two-phase setup.
-In the preprocessing phase, we are given a lower bound graph $G_\ell$ with lower bound weights $w_\ell$.
-The lower bound weights are single scalar numbers for each edge.
-We now perform some precomputation, i.e. apply the Contraction Hierarchy preprocessing to this graph.
-This generates some auxiliary data which we will use in the query phase to efficiently compute distance estimates.
-We will do so with the heuristic function $h_t(x)$ where $x$ and $t$ are nodes from $G_\ell$.
-The heuristic function returns the shortest distance from $x$ to $t$ with respect to $w_\ell$.
-Before each query, the heuristic has to be initialized with the target node $t$.
-We will discuss the implementation of this heuristic function and the initialization later in this section.
-
-In the query phase, we have a query graph $G_q$ with query weights $w_q$ and a mapping $T$ which maps nodes from $G_q$ to $G_\ell$.
-For most applications, $G_q = G_\ell$ and $T$ will be the identity.
-The query weights $w_q$ are the main mechanism to implement different applications with CH-Potentials.
-In some cases, $w_q$ will be simple scalar values like the lower bound weights.
-However, for $w_q$ this is not a requirement.
+In this paper, we consider different applications, with slightly different problem models.
+The goal is always to quickly answer many shortest path queries.
+For the purpose of describing our framework, we establish a shared notation:
+Input to each query are nodes $s$ and $t$, and a graph $G$ with query weights $w_q$.
+However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
+In the simplest case, $w_q$ will be scalar edge weights.
+However, this is not a requirement.
 It can be any function that computes a weight for an edge.
 This function can also take additional parameters from the state of the search.
+For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
+However, values of $w_q$ might change between queries.
+In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
 
-The query phase then is to execute A* on $G_q$.
-Before running the search, the heuristic will be initialized with the target $T(t)$.
-The distance estimate for a node $x$ from $G_q$ can be computed as $h_{T(t)}(T(x))$.
-For CH-Potentials to find correct distances, $w_q \geq w_\ell$ has to hold for all arcs.
-\todo{Correctness Argument?}
-\todo{Describe A*?}
+To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
+The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
+$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
+We require that $w_q(uv)$ is greater or equal to the shortest distance from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
+The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
+$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
+In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
+
+The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
+The exact implementation of this A* search depend on the application.
+Our approach only provides the heuristic $h_t$ for the A* search.
+In contrast, the preprocessing phase remains the same for all applications.
 

### 2021-01-18 paper/ch_potentials.tex
some exp improvement

@@
 
-With this setup CH-Potentials support turns without requiring turn information in the CH.
+With this setup, CH-Potentials support turns without requiring turn information in the CH.
 
@@
 
-The first experiment varies the tightness of the heuristic.
+The performance of A* depends on the tightness of the heuristic.
+CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
+However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise we could use an unmodified CH).
+We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
 The lower bound $w_\ell$ is set to the freeflow travel time.
@@
 This illustrates both the strengths and limits of our approach and goal directed search in general.
-Goal directed search (and thus CH-Potentials) can only achieve competitive running times if the application allows for a sufficiently tight heuristic.
+CH-Potentials can only achieve competitive running times if the application allows for a sufficiently tight lower bounds at preprocessing time.
 

### 2021-01-19 paper/ch_potentials.tex
write on CCH-Pot

@@
 
-\todo{talk about CCH-Pot}
-
 Beside predicted traffic, we also consider live traffic.
@@
 This straight-forward integration of complex routing problems is the strength of the CH-Potentials algorithm.
+
+\subsubsection{Three-Phase Setups}
+
+Supporting live traffic is also possible with a three-phase setup:
+A slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
+The customization phase is run regularly and incorporates updates to the weights into the auxiliary data.
+CRP~\cite{dgpw-crprn-13} and Customizable CH~(CCH)~\cite{dsw-cch-15} follow this setup.
+Luckily, a CCH is just a CH with some additional properties.
+The CH in CH-Potentials can be replaced by a CCH without further modifications.
+Thus, CCH-Potentials could also support a three-phase setup.
+For applications which allow for a three-phase setup, CCH-Potentials are likely a better choice for supporting live traffic.
+However, evaluating CCH-Potentials is beyond the scope of this paper.
+We focus on evaluating CH-Potentials as a building block in the two-phase setup.
 

### 2021-01-19 eval/scaled_weights.py
improve boxplot

@@ -23,8 +23,8 @@ queries = pd.DataFrame.from_records([{
     **algo }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
 
-plt.figure(figsize=(6,3))
-g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', linewidth=0.8, showmeans=True)
+plt.figure(figsize=(11,4))
+g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
 g.xaxis.set_major_locator(mpl.ticker.MultipleLocator(5))
 # g.xaxis.set_major_formatter(mpl.ticker.ScalarFormatter())
 g.set_yscale('log')

### 2021-01-19 paper/ch_potentials.tex
improve citing situation for temporary road closures

@@
 Truck routing differs from car routing due to night driving bans and other restrictions.
-In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials is used for such a scenario.
+In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials\footnote{
+In~\cite{kswz-erptd-20}, CH-Potentials are only used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
+The paper does not describe any of the contribution of this paper.
+} is used for such a scenario.
 The work considers time-dependent blocked edges and waiting at parking locations.
@@
 The whiskers cover 1.5 times the inter quartile range.
-All other running times are marked as outliers.
+All other running times are indicated as outliers.
 }\label{fig:scaled_weights}

### 2021-01-20 paper/ch_potentials.tex
fix citations

@@
 CH is extended to support turn costs in~\cite{}, CCH in~\cite{bwzz-cchtc-20}.
-Traffic predictions are combined with CH in~\cite{swz-sfert-19,bgsv-mtdtt-13}.
+Traffic predictions are combined with CH in~\cite{swz-sfert-20,bgsv-mtdtt-13}.
 \todo{cite correct CATCHUp Version}
@@
 Recurring traffic can be predicted by observing the traffic in the past.
-It is common \cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-19} to represent these predictions as \emph{travel time functions}.
+It is common \cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-20} to represent these predictions as \emph{travel time functions}.
 An edge weight is no longer a scalar value but a function that maps the entry time onto the traversal time.
@@
 With this setup, we extended CH-Potentials to support time-dependent routing.
-As we were able to keep all travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-19}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
+As we were able to keep all travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-20}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
 
@@
 Truck routing differs from car routing due to night driving bans and other restrictions.
-In~\cite{kswz-erptd-20}, a preliminary version of CH-Potentials\footnote{
-In~\cite{kswz-erptd-20}, CH-Potentials are only used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
+In~\cite{kswz-erptd-p-20}, a preliminary version of CH-Potentials\footnote{
+In~\cite{kswz-erptd-p-20}, CH-Potentials are used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
 The paper does not describe any of the contribution of this paper.
@@
 Further, a trade-off between arrival time and route quality is considered.
-\todo{cite correct Closures Version}
 

### 2021-01-21 eval/building_blocks.py
improve building block table

@@ -32,27 +32,24 @@ def heuristic_name(feature_flags):
 
 queries['Heuristic'] = queries['features'].apply(heuristic_name)
 
-table = queries.groupby(['algo', 'Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3'])[['running_time_ms', 'num_queue_pushs']].mean()
+queries_subset = queries.loc[lambda x: ((x['algo'] == 'Dijkstra Query') & (x['Heuristic'] == 'Zero')) | ((x['algo'] != 'Dijkstra Query') & ~((x['Heuristic'] == 'Zero') & ~x['BCC'] & ~x['Skip Deg. 2'] & ~x['Skip Deg. 3']))]
+
+table = queries_subset.groupby(['Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3'])[['running_time_ms', 'num_queue_pushs']].mean()
 table['num_queue_pushs'] = table['num_queue_pushs'] / 1000
 table = table.round(1)
-
-table = table.reindex(['Zero', 'ALT', 'CH', 'Oracle'], level=1)
-table = table.rename(index={
-    'Zero': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{Zero}}',
-    'ALT': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{ALT}}',
-    'CH': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{CH-\\Potentials}}}',
-    'Oracle': R'\multirow{2}{*}{Oracle}',
-})
-
-table = table.reset_index(level=0,drop=True).reset_index(level=[2,3])
+table = table.unstack(0).stack(0) \
+    .swaplevel(2, 3).swaplevel(1,2).swaplevel(0,1) \
+    .loc[['running_time_ms', 'num_queue_pushs']][['Zero', 'ALT', 'CH', 'Oracle']].reset_index(level=[2,3]) \
+    .rename(index={
+        'running_time_ms': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{Running\\time [ms]}}}',
+        'num_queue_pushs': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{Queue\\pushs [$\cdot 10^3$]}}}',
+    })
 
 lines = table.to_latex(escape=False).split("\n")
 
[…]

### 2021-01-21 paper/ch_potentials.tex
improve references

@@
 In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network during query time.
-A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
+A further popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
 It is also used in practice~\todo{citation, bing blog?}.
@@
 
-Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{}.
+Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
 A* is a goal-directed variant of Dijkstra's algorithm.
@@
 In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH), which also supports exchanging weights and thus supporting user preferences and live traffic.
-CH is extended to support turn costs in~\cite{}, CCH in~\cite{bwzz-cchtc-20}.
+CH is extended to support turn costs in~\cite{gv-errnt-11}, CCH in~\cite{bwzz-cchtc-20}.
 Traffic predictions are combined with CH in~\cite{swz-sfert-20,bgsv-mtdtt-13}.
-\todo{cite correct CATCHUp Version}
 \todo{improve! and discuss hierarchical extensions and what they achieve and their difficulties in more details}
+\todo{cite traffic aware routing}
 Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
@@
 However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s flexibility.
-\todo{cite traffic aware routing}
 %
 
-ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
+ALT~\cite{gh-cspas-05,gw-cppsp-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
 Just as our approach, ALT is A*-based.
@@
 Figure~\ref{fig:scaled_weights} depicts the results.
-\todo{decrease figure height}
 Clearly, $\alpha$ has significant influence on the running time.
@@
 Second, we compare against our own implementation\footnote{
-We use 16 landmarks generate with the avoid strategy.
+We use 16 landmarks generate with the avoid~\cite{gw-cppsp-05} strategy.
 We store all landmark distances consecutive for each node.
@@
 Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
-} of ALT~\cite{}.
+} of ALT~\cite{gw-cppsp-05}.
 Finally we compare against a hypothetical \emph{Oracle-A*} heuristic.

### 2021-01-21 paper/ch_potentials.tex
adapt applications to new framework description

@@
 For the purpose of describing our framework, we establish a shared notation:
-Input to each query are nodes $s$ and $t$, and a graph $G$ with query weights $w_q$.
+Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
 However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
@@
 	$y\leftarrow$ pop minimum element from $Q$\;
-	\For{$xy$ is down-edge in $G^+$}{
+	\For{$xy$ is down-edge in $G^+_\ell$}{
 		\If{$B[x] > w_\ell(xy) + B[y]$}{
@@
 The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
-%
-This works, however, while one CH query is fast, answering one for every node explored in the $A^*$ search is slow.
+This yields tight estimates but a high overhead for the heuristic evaluation.
+While a single CH query is fast, answering one for every node explored in the $A^*$ search is slow.
 Fortunately, we can do better.
-However, for this we need another component called PHAST~\cite{dgnw-phast-13}.
+% For this we need another component called PHAST~\cite{dgnw-phast-13}.
 
@@
 \For{all CH levels $L$ from most to least important}{
-	\For{all up edges $xy$ in $G^+$ with $x$ in $L$}{
+	\For{all up edges $xy$ in $G^+_\ell$ with $x$ in $L$}{
 		\If{$P[x] < P[y] + w_\ell(xy)$}{
@@
 PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
-First the preprocessing phase is executed analogously to the original CH.
-The query phase is divided into two steps.
+The preprocessing phase remains unchanged.
+The query phase is split into two steps.
 The first step is analogue to the CH query:
@@
 In each iteration, all up-edges starting within the current level are followed in reverse.
-After all levels are processed, the distances from all nodes towards $t$ is computed.
+After all levels are processed, the shortest distances from all nodes to $t$ were computed.
 %PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
@@
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-Using PHAST, we can also compute a preprocessing-tight A* heuristic.
-\todo{preprocessing tight}
+Using PHAST, we can also compute a tight A* heuristic.
 In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
@@
 		$P[x]\leftarrow B[x]$\;
-		\For{all up edges $xy$ in $G^+$}{
+		\For{all up edges $xy$ in $G^+_\ell$}{
                         $P[x]\leftarrow\min\{P[x],w_\ell(xy)+\Pot(y)\}$\;
@@
 
-To compute the heuristic $h(x)$, we recursively compute for all up-edges $(x,y)$ the heuristic $h(y)$.
-Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_y\{w_\ell(x,y) + h(y)\}$.
-As not all shortest up-down paths contain an up-edge, we set $h(x) = \min \{ B[x], d \}$.
-This calculation is correct, as it computes the minimum up-down $xt$-path distance in $G^+$, which corresponds to the minimum $xt$-path distance in a CH.
+To compute the heuristic $h_t(x)$, we recursively compute for all up-edges $(x,y)$ the heuristic $h_t(y)$.
+Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_y\{w_\ell(x,y) + h_t(y)\}$.
+As not all shortest up-down paths contain an up-edge, we set $h_t(x) = \min \{ B[x], d \}$.
+This calculation is correct, as it computes the minimum up-down $xt$-path distance in $G^+_\ell$, which corresponds to the minimum $xt$-path distance in a CH.
 A* with this heuristic is the basic CH-Potentials algorithm.
@@
 
-Preliminary experiments showed, that most A* running time is spent in heuristic evaluations and queue operations.
+Preliminary experiments showed, that a significant amount of query running time is spent in heuristic evaluations and queue operations.
 We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
 Avoiding pushing low degree nodes into the queue is the focus of this section.
-
-\subsection{Skip Degree Two Nodes}
 
@@
 
+\subsection{Skip Degree Two Nodes}
+
 Our algorithm differs from A* when removing a node $x$ from the queue.
 A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
-If A* succeeds, $y$'s weight in the queue is set to $D[y]+h(y)$.
+If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
 Our algorithm, however, behaves differently, if $d(y)\le 2$.
@@
 If it does not reach $z$, then only $D$ is modified but no queue action is performed.
-If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h(z)$.
+If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h_t(z)$.
 
@@
 
-In the previous section, we described an optimized A* variant that does not push degree two nodes.
-In this section, we also avoid degree three nodes.
-
+We can also skip some degree three nodes.
 Denote by $x,y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
@@
 Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
-If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h(\beta)$.
-Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h(b)$.
+If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h_t(\beta)$.
+Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h_t(b)$.
 If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
@@
 \subsection{Stay in Largest Biconnected Component}
-
 \label{sec:largested-biconnected-component}
@@
 A lot of nodes in road networks lead to dead-ends.
-Unless the source or target is is in a dead-end, it is unnecessary to explore these nodes.
+Unless the source or target is is in this dead-end, it is unnecessary to explore these nodes.
 
@@
 
-A* is a flexible algorithm.
-Many extended problems can therefore be solved using the CH-Potentials framework.
-In this section, we describe some extended problems and how they can be solved with CH-Potentials.
+% A* is a flexible algorithm.
+% Many extended problems can therefore be solved using the CH-Potentials framework.
+We describe some extended routing problems and how to apply CH-Potentials to them.
+Unless statet otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
 
@@
 However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
-Such restrictions can be modeled using turn weights~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
+Also, taking a left turn might take longer than going straight.
+This can be modeled using turn weights~\cite{gv-errnt-11,dgpw-crprn-13,bwzz-cchtc-20}.
 % In this section, we first present the extended problem setting.
 % Afterwards, we describe how it can be solved with CH-Potentials.
-We extend CH-Potentials by using zero as lower bound for every turn weight in the heuristic.
+A \emph{turn weight} $w_t$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns.
+For CH-Potentials, we use zero as lower bound for every turn weight in the heuristic.
+Thus, the graph $G_\ell$ and weights $w_\ell$ for preprocessing is the unmodified input graph without turn weights.
 
 % Ich bleibe bei v_i hier, damit keine Verwechselung mit den freien Variablen x,y,z gibt.
-A \emph{turn weight} $w_t$  maps a pair of incident edges onto the turning time or $+\infty$.
-A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}:\[
-w_q(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1})  + w_q(v_i,v_{i+1})
-\] The objective is to find a path between two given edges with minimum turn-aware weight.
-The first term $w_q(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
+A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
+w_\ell(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i,v_{i+1})
+\]
+The objective is to find a path between two given edges with minimum turn-aware weight.
+The first term $w_\ell(v_1, v_2)$ is the same for all paths, as it only depends on the source edge.
 It can thus be ignored during optimization.
 
-A common strategy is to construct a \emph{turn-expanded} graph $G'$.
-Edges in the input graph $G$ correspond to \emph{expanded nodes} in $G'$.
-For every pair of incident edges $(x,y)$ and $(y,z)$ in $G$, there is an \emph{expanded edge} in $G'$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
-A sequence of expanded nodes in the expanded graph $G'$ corresponds to a sequence of edges in the input graph $G$.
-The weight of a path in $G'$ is equal to the turn-aware weight of the corresponding path in $G$ minus the irrelevant $w_q(v_1,v_2)$ term.
-The turn-aware routing problem can be solved by searching for paths in $G'$.
+We solve this problem by constructing a \emph{turn-expanded} graph as $G_q$.
+Edges in the input graph $G_\ell$ correspond to \emph{expanded nodes} in $G_q$.
+For every pair of incident edges $(x,y)$ and $(y,z)$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
+A sequence of expanded nodes in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
+The weight of a path in $G_q$ is equal to the turn-aware weight of the corresponding path in $G_\ell$ minus the irrelevant $w_\ell(v_1,v_2)$ term.
+Thus, the turn-aware routing problem can be solved by searching for shortest paths in $G_q$.
 
-The A* search runs in the expanded $G'$ and uses a heuristic $h'$.
-The CH is constructed for $G$.
-With CH-Potentials, we get the preprocessing-tight heuristic $h$.
-\todo{prepro tight}
-We set $h'(x,y) = h(y)$.
+In this scenario, preprocessing and query use different graphs $G_\ell$ and $G_q$.
+We define the node mapping function $\phi$ as $\phi(x,y) = y$.
+Sadly, the undirected graph underlying $G_q$ is always biconnected, if the input graph is strongly connected.
+The optimization described in Section~\ref{sec:largested-biconnected-component} is therefore ineffective.
+With this setup, CH-Potentials support turns without requiring turn information in the CH.
 
-To prove that $h'(x,y)$ is consistent, consider the turn-aware weight of a path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
-We can lower bound the expression as follows:
-\[
-\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1}) \le \sum_{i=2}^{k-1} \underbrace{w_t(v_{i-1},v_i,v_{i+1})}_{0\le} + \sum_{i=2}^{k-1} \underbrace{w_q(v_i,v_{i+1})}_{w_\ell(v_i,v_{i+1})\le}
-\]
-$\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ is a $yq$-path length in $G$.
-It is no shorter than the shortest $yq$-path length in $G$, which is equal to $h(y)$.
-The heuristic is therefore a lower bound.
-It is consistent because it is derived from exact shortest paths in $G$ with $w_\ell$.
-
-Sadly, the undirected graph underlying $G'$ is always biconnected, if the input graph is strongly connected.
-The optimization described in Section~\ref{sec:largested-biconnected-component} is therefore ineffective.
-
-With this setup, CH-Potentials support turns without requiring turn information in the CH.
+% To prove that $h'(x,y)$ is consistent, consider the turn-aware weight of a path $v_1,v_2\ldots v_k$ with $v_1=x$, $v_2=y$, $v_{k-1}=p$, and $v_k=q$.
+% We can lower bound the expression as follows:
+% \[
+% \sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1}) \le \sum_{i=2}^{k-1} \underbrace{w_t(v_{i-1},v_i,v_{i+1})}_{0\le} + \sum_{i=2}^{k-1} \underbrace{w_q(v_i,v_{i+1})}_{w_\ell(v_i,v_{i+1})\le}
+% \]
+% $\sum_{i=2}^{k-1} w_\ell(v_i,v_{i+1})$ is a $yq$-path length in $G$.
+% It is no shorter than the shortest $yq$-path length in $G$, which is equal to $h(y)$.
+% The heuristic is therefore a lower bound.
+% It is consistent because it is derived from exact shortest paths in $G$ with $w_\ell$.
 
@@
 The classical shortest path problem assumes that edge weights are scalars.
-However in practice, travel times vary along an edge.
-The primary reason is traffic.
+However, in practice, travel times vary along an edge due to the traffic situation.
+% The primary reason is traffic.
 Recurring traffic can be predicted by observing the traffic in the past.
@@
 % Performing this prediction is outside of the scope of this paper.
-It is common to refer to routing with predicted traffic as \emph{time-dependent routing}.
+% It is common to refer to routing with predicted traffic as \emph{time-dependent routing}.
 % Again, we first formalize the extended problem setting and then describe our solution with CH-Potentials.
 
-Formally, the time-dependent query weight $w_p$ is a function from $E\times \mathbb{R}$ to $\mathbb{R}^+$.
-$w_p(e, \tau)$ is the travel time through edge $e$ when entering it at moment $\tau$.
+In this setting, the query weight $w_q$ is a function from $E\times \mathbb{R}$ to $\mathbb{R}^+$.
+$w_q(e, \tau)$ is the travel time through edge $e$ when entering it at moment $\tau$.
 The input to the extended problem consists of a source node $s$ and a target node $t$, as in the classical problem formulation.
@@
 \begin{split}
-\alpha_{1} & = w_p(e_1, \tau_s) \\
-\alpha_{k} & = \alpha_{k-1} + w_p(e_1, \alpha_{k-1})
+\alpha_{1} & = w_q(e_1, \tau_s) \\
+\alpha_{k} & = \alpha_{k-1} + w_q(e_1, \alpha_{k-1})
 \end{split}
 \]
-The objective is to find a path that minimizes $\alpha_k$.
+The objective is to find a path to $t$ that minimizes $\alpha_k$.
 
-If all travel time functions fulfill the \emph{no-waiting property}, this problem can be solved using a straight forward extension of Dijkstra's algorithm \cite{d-aassp-69}.
+If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straight forward extension of Dijkstra's algorithm \cite{d-aassp-69}.
 The necessary modification to A* is analogous.
-Without the no-waiting property the problem is in general NP-hard \cite{or-tnp-89}.
-The no-waiting property states that it is never beneficial to wait at a node before entering an edge.
-Formally stated, the following must hold:\[
-\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_p(e, \tau) \le w_p(e, \tau+\delta) + \delta
-\]
+Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
+% The FIFO property states that it is never beneficial to wait at a node before entering an edge.
+The FIFO property states that it is not possible to arrive earlier by departing later.
+Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta$.
 Our implementation stores edge travel times using piece-wise linear functions.
+The A* search uses the tentative distance $\tau$ at a node $x$ when to evaluating the travel time of outgoing edges $(x,y)$.
+This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
 
-To solve the time-dependent routing problem, we employ a strategy very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
-The main difference is that we use CH-Potentials instead of landmarks to guide the A* search.
-We set $w_\ell(e) = \min_\tau w_p(e,\tau)$, that is the minimum travel time.
-Following TD-ALT, we modify the A* search to use the tentative distance at a node $x$ as $\tau$ when evaluating the weight of an outgoing edge $(x,y)$.
-
-With this setup, we extended CH-Potentials to support time-dependent routing.
-As we were able to keep all travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-20}, which create shortcuts of travel time functions to combine hierarchical speedup techniques with time-dependent routing.
+For the preprocessing, we set $w_\ell(e) = \min_\tau w_q(e,\tau)$, that is the minimum travel time.
+% With this setup, we extended CH-Potentials to support time-dependent routing.
+By keeping travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-20,dn-crdtd-12} which have to create shortcuts of travel time functions.
 
@@
 It is important to distinguish between predicted and live traffic.
-The predicted travel time along an edge was estimated in the past.
-It is possible that it differs from the current travel time significantly, if unexpected events happen.
-Accidents are examples of such unexpected events.
 Live traffic data is more accurate for the current moment than predicted data.
+It is possible that it differs significantly from predicted traffic, if unexpected events like accidents happen.
+% The predicted travel time along an edge was estimated in the past.
+% Accidents are examples of such unexpected events.
 However, just using live traffic data is problematic for long routes as traffic changes while driving.
 At some point, one wants to switch from live traffic to the predicted traffic.
-In this section, we first describe a setup with only live traffic and then extend it with predicted traffic.
+In this section, we first describe a setup with only live traffic and then combine it with predicted traffic.
 
-To support live traffic, we set $w_\ell$ to the freeflow travel time.
+To support only live traffic, we set $w_\ell$ to the freeflow travel time.
 $w_q$ is set to the travel time accounting for current traffic.
 As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
-This is all that is necessary to combine live traffic and CH-Potentials.
+In a real world application, values from $w_q$ could be updated between queries.
+This is all that is necessary to apply CH-Potentials in a live traffic scenario.
 
@@
 Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
+% We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
+In our implementation, we to not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
 We set $w_\ell$ again to the freeflow travel time.
-We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
-In our implementation, we to not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
 
@@
 Further adding tunnel and/or highway avoidance or turn-aware routing is simple.
-This straight-forward integration of complex routing problems is the strength of the CH-Potentials algorithm.
+This straight-forward integration of complex routing problems is the strength of the CH-Potentials.
 
@@
 A slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
-The customization phase is run regularly and incorporates updates to the weights into the auxiliary data.
+The customization phase is run regularly and incorporates updates to the weights into the auxiliary preprocessing data.
 CRP~\cite{dgpw-crprn-13} and Customizable CH~(CCH)~\cite{dsw-cch-15} follow this setup.
@@
 Thus, CCH-Potentials could also support a three-phase setup.
-For applications which allow for a three-phase setup, CCH-Potentials are likely a better choice for supporting live traffic.
+% For applications which allow for a three-phase setup, CCH-Potentials are likely a better choice for supporting live traffic.
 However, evaluating CCH-Potentials is beyond the scope of this paper.
-We focus on evaluating CH-Potentials as a building block in the two-phase setup.
+We focus on evaluating CH-Potentials as a simple building block in the two-phase setup.
 

### 2021-01-21 paper/ch_potentials.tex
correctness argument

@@
 $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
-We require that $w_q(uv)$ is greater or equal to the shortest distance from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
+We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
 The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
@@
 In contrast, the preprocessing phase remains the same for all applications.
+
+Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v)$ holds for all edges.
+By requirement and because of the triangle inequality the following holds:
+\[
+w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
+\]
+Thus, A* will always determine the correct shortest distances.
 
@@
 Edges in the input graph $G_\ell$ correspond to \emph{expanded nodes} in $G_q$.
-For every pair of incident edges $(x,y)$ and $(y,z)$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_q(y,z)$.
+For every pair of incident edges $(x,y)$ and $(y,z)$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_\ell(y,z)$.
 A sequence of expanded nodes in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
@@
 We define the node mapping function $\phi$ as $\phi(x,y) = y$.
+Obviously, $w_q(xy, yz) = w_t(x,y,z) + w_\ell(y,z) \geq \operatorname{dist}_\ell(\phi(x,y), \phi(y,z))$ and this approach yields a feasible heuristic.
 Sadly, the undirected graph underlying $G_q$ is always biconnected, if the input graph is strongly connected.

### 2021-01-21 paper/ch_potentials.tex
related work

@@
 
-% A lot of research focuses on the inflexible setting, i.e., the special case where $w_q = w_\ell$.
-% In this setting, mostly hierarchical approaches dominate in terms of running time~\cite{bdgmpsww-rptn-16}.
-% A well-known hierarchical technique are Contraction Hierarchies (CH)~\cite{gssv-erlrn-12}.
-
-There is a lot of work that extends hierarchical speed-up techniques to more complex settings.
-\todo{cite overview}
-% For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
-MLD/CRP~\cite{dgpw-crprn-13} was developed to allow exchanging weights in reasonable time.
-This enable user preferences and live traffic.
-MLD/CRP also supports turn costs.
+There is a lot of work that extends hierarchical speed-up techniques to more complex settings~\cite{bdgmpsww-rptn-16}.
+For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
+The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
+A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
+Several papers~\cite{} and an entire dissertation~\cite{} have been published on the subject.
+\todo{cite td stuff}
+Different techniques with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
+Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves all three goals but only at the cost of considerable implementation complexity.
+%
+CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of Multi-Level Dijkstra~(MLD)~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
+For this, a faster, second preprocessing phase is introduced which can be run regularly to update weights.
+In theory, this enables the integration of live traffic and user preferences.
+In practice, live traffic feed data is imperfect.
+Computing ``good'' routes without undesired detours due to artifacts in the data requires additional algorithmic extensions~\cite{}.
+\todo{cite traffic aware routing}
+CRP also supports turn costs.
 Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
-In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH), which also supports exchanging weights and thus supporting user preferences and live traffic.
-CH is extended to support turn costs in~\cite{gv-errnt-11}, CCH in~\cite{bwzz-cchtc-20}.
-Traffic predictions are combined with CH in~\cite{swz-sfert-20,bgsv-mtdtt-13}.
-\todo{improve! and discuss hierarchical extensions and what they achieve and their difficulties in more details}
-\todo{cite traffic aware routing}
-Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
+However, on continental sized networks, the proposed approach can only compute approximate shortest distances (rather than paths).
+%
+In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH).
+CCH also has a second preprocessing phase where weights can be altered.
+Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
+Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
+Other extensions studied include electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11} or multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10}.
+%
 While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
 Further, in every setting the flexibility available at query time is fairly limited.
-Combining all these hierarchical extensions is an unsolved problem.
-%
+Combining these hierarchical extensions is an unsolved problem.
+
 CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.

### 2021-01-25 paper/ch_potentials.tex
more citations and cleanup

@@
 Computing tight distance estimates is a challenge on its own.
-\todo{tight}
 On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
@@
 A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
-Several papers~\cite{} and an entire dissertation~\cite{} have been published on the subject.
-\todo{cite td stuff}
+Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
 Different techniques with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
@@
 In practice, live traffic feed data is imperfect.
-Computing ``good'' routes without undesired detours due to artifacts in the data requires additional algorithmic extensions~\cite{}.
-\todo{cite traffic aware routing}
+Computing ``good'' routes without undesired detours due to artifacts in the data requires additional algorithmic extensions~\cite{dss-tarrn-18}.
 CRP also supports turn costs.

### 2021-01-25 paper/ch_potentials.tex
make related work less beefy and shorter

@@
 Section~\ref{sec:low-deg-improvment} discusses our improvements for the handling of low-degree nodes.
-In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply our technique to different practical applications.
+In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
 Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
@@
 For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
-The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
+% The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
 A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
 Several papers~\cite{bdsv-tdch-09,bgns-tdcha-10,klsv-dtdch-10,bgsv-mtdtt-13} and an entire dissertation~\cite{b-tdrpc-14} have been published on the subject.
-Different techniques with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
-Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves all three goals but only at the cost of considerable implementation complexity.
+Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
+Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
 %
 CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of Multi-Level Dijkstra~(MLD)~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
-For this, a faster, second preprocessing phase is introduced which can be run regularly to update weights.
+For this, a faster, second preprocessing phase is introduced.
+It can be run regularly to update weights.
 In theory, this enables the integration of live traffic and user preferences.
@@
 Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
-However, on continental sized networks, the proposed approach can only compute approximate shortest distances (rather than paths).
+On continental sized networks, TD-CRP can only compute approximate shortest distances (rather than paths).
 %
@@
 Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
-Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
+% Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
 Other extensions studied include electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11} or multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10}.
@@
 While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
-Further, in every setting the flexibility available at query time is fairly limited.
+Further, in every extension the flexibility available at query time is fairly limited.
 Combining these hierarchical extensions is an unsolved problem.
@@
 CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s flexibility.
-%
+However, previous works mostly focused on accelerating hierarchical approaches further rather than exploiting A*'s flexibility.
 
 ALT~\cite{gh-cspas-05,gw-cppsp-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
-Just as our approach, ALT is A*-based.
+% Just as our approach, ALT is A*-based.
+ALT has been combined with shortcuts~\cite{bdsssw-chgds-10} and also extended for dynamic graphs~\cite{dw-lbrdg-07} and time-dependent routing~\cite{ndls-bastd-12,dn-crdtd-12}.
 % However, ALT's heuristic is not preprocessing-tight.
@@
 For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
-Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
+Unfortunately, the employed CPD's quadratic preprocessing running time is problematic on large street networks.
 %
-MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
+% MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
 %
-In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
-The Euclidean distance is then used as A* heuristic.
+In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap. % such that distances in space and distances in the graph roughly correspond.
+The Euclidean distance is then used as a distance estimate for A*.
 % This results in a good but not a preprocessing-tight heuristic.
-%
-The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.
 
@@
 Avoiding pushing low degree nodes into the queue is the focus of this section.
+The techniques discussed here are a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
 
@@
 Second, we compare against our own implementation\footnote{
-We use 16 landmarks generate with the avoid~\cite{gw-cppsp-05} strategy.
+We use 16 landmarks generate with the avoid strategy~\cite{gw-cppsp-05}.
 We store all landmark distances consecutive for each node.

### 2021-01-25 paper/ch_potentials.tex
various fixes, update conclusion, shorten references

@@
 A CH exploits the inherent hierarchy of road networks.
-In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network during query time.
-A further popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
-It is also used in practice~\todo{citation, bing blog?}.
-MLD also also uses shortcut edges.
+In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
+Another popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
+It is also used in practice~\cite{bingblog}.
+MLD also uses shortcut edges.
 Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
@@
 % However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
-In this paper, we describe CH-Potentials, a heuristic that achieves both goals.
+In this paper, we describe CH-Potentials, a heuristic that achieves both.
 Internally, the heuristic uses a CH.
@@
 Extending A* is vastly easier than extending a CH.
-This enables us to design algorithms that support a multitude of features.
+This enables us to design algorithms for a multitude of features.
 In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
@@
 %
-CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of Multi-Level Dijkstra~(MLD)~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
+CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
 For this, a faster, second preprocessing phase is introduced.
@@
 
-Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v)$ holds for all edges.
+Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
 By requirement and because of the triangle inequality the following holds:
@@
 A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
-On road networks, this is the case~\cite{gssv-erlrn-12,dgrw-gpnc-11,dgpw-crprn-13}.
+On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
 On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.
@@
 The customization phase is run regularly and incorporates updates to the weights into the auxiliary preprocessing data.
-CRP~\cite{dgpw-crprn-13} and Customizable CH~(CCH)~\cite{dsw-cch-15} follow this setup.
+CRP~\cite{dgpw-crprn-13} and CCH~\cite{dsw-cch-15} follow this setup.
 Luckily, a CCH is just a CH with some additional properties.
@@
 CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
-However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise we could use an unmodified CH).
+However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
 We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
@@
 The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
-The performance of CH-Potentials crucially depends on the availability of good lower bounds in the preprocessing phase.
+CH-Potentials provides \emph{exact} distances with respect to lower bound weights known at preprocessing time as an A* heuristic.
+Thus, the query performance of CH-Potentials crucially depends on the availability of good lower bounds in the preprocessing phase.
 Our experiments show, that this availability highly depends on the application.
-We experimentally show that the running time of our algorithm is within a factor 1.6 of a hypothetical A* variant that can instantly evaluate the preprocessing-tight heuristic.
-\todo{preprocessing tight}
+We also show that the overhead of our heuristic is within a factor 1.6 of a hypothetical A*-heuristic that can instantly access lower bound distances.
 Achieving significantly smaller running times could still be possible in variations of the problem setting.
-This leads to multiple avenues for future research.
+% This leads to multiple avenues for future research.
 
@@
 Hierarchical techniques have been shown to work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
-\todo{der paragraph war eher so fuer die AAAI, oder?}
 

### 2021-01-25 paper/ch_potentials.tex
language polishing

@@
 % However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
-In this paper, we describe CH-Potentials, a heuristic that achieves both.
+In this paper, we describe CH-Potentials, a fast heuristic with tight estimates.
 Internally, the heuristic uses a CH.
@@
 A CH is a two phase technique to efficiently compute exact, shortest paths.
-For a details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
+For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
 In this section, we give an introduction.
@@
 A lot of nodes in road networks lead to dead-ends.
-Unless the source or target is is in this dead-end, it is unnecessary to explore these nodes.
+Unless the source or target is in this dead-end, it is unnecessary to explore these nodes.
 
@@
 If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
-Otherwise we find $a_t$.
+Otherwise, we find $a_t$.
 In that case, we continue by searching a path from $a_t$ towards $t$ restricted to $t$'s biconnected component.
@@
 We describe some extended routing problems and how to apply CH-Potentials to them.
-Unless statet otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
+Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
 
@@
 The classical shortest path problem allows to freely change edges at nodes.
-However in the real world, turn restrictions, such as a forbidden left or right turn, exist.
+However, in the real world, turn restrictions, such as a forbidden left or right turn, exist.
 Also, taking a left turn might take longer than going straight.
@@
 The band in the box indicates the median, the diamond the mean.
-The whiskers cover 1.5 times the inter quartile range.
+The whiskers cover 1.5 times the interquartile range.
 All other running times are indicated as outliers.
@@
 Up to around $\alpha = 1.1$ the running time grows quickly.
-After $\alpha = 1.1$ the growth slows down.
+For $\alpha > 1.1$, the growth slows down.
 This illustrates both the strengths and limits of our approach and goal directed search in general.
@@
 } of ALT~\cite{gw-cppsp-05}.
-Finally we compare against a hypothetical \emph{Oracle-A*} heuristic.
-This heuristic has instant access to a shortest distance array with respect to $w_\ell$, that is it is the most efficient heuristic possible in our model.
+Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
+This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is the most efficient heuristic possible in our model.
 We fill this array before each query using a reverse Dijkstra search from the target node.
@@
 % Skipping degree three nodes has the smallest impact.
-CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly less nodes.
+CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly fewer nodes.
 This is not surprising, since ALT computes worse distance estimates.
@@
 CH-Potentials performance for different route planning applications.
-We report average running times and number of queue pushs.
+We report average running times and number of queue pushes.
 We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
-Finally, we report the average running time of Dijkstras algorithm as a baseline and the speedup over this baseline.
+Finally, we report the average running time of Dijkstra's algorithm as a baseline and the speedup over this baseline.
 }\label{tab:applications}
@@
 We also show that the overhead of our heuristic is within a factor 1.6 of a hypothetical A*-heuristic that can instantly access lower bound distances.
-Achieving significantly smaller running times could still be possible in variations of the problem setting.
+Achieving significantly faster running times could still be possible in variations of the problem setting.
 % This leads to multiple avenues for future research.

### 2021-01-26 paper/ch_potentials.tex
remove final todo
small changes

@@
 We study exact, efficient and practical algorithms for route planning in large road networks.
-Realistic routing often requires integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
-While Dijkstra's algorithm can be used to solve these problems, it is too slow for many practical applications.
+Routing applications often requires integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
+While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
 A* is a classical approach to accelerate Dijkstra's algorithm.
 A* can support many extended scenarios without much additional implementation complexity.
-However, A*'s performance depends on the availability of a good heuristic to estimate distances.
+However, A*'s performance depends on the availability of a good heuristic that estimates distances.
 Computing tight distance estimates is a challenge on its own.
 On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
-They achieve speed and exactness but sacrifice A*'s flexibility and are not as extensible.
+They achieve speed and exactness but sacrifice A*'s flexibility.
+Extending them to certain practical applications can be hard.
 In this paper, we present an algorithm to efficiently extract distance estimates for A* from Contraction Hierarchies (CH), a hierarchical technique.
@@
 One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
-They have been used successfully in many real world applications\todo{citation?}.
+They have been used successfully in many real world applications.
 A CH exploits the inherent hierarchy of road networks.
@@
 MLD also uses shortcut edges.
-Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
+Both approaches achieve speed-ups of three orders of magnitude over Dijkstra's algorithm.
 
@@
 In~\cite{kswz-erptd-p-20}, CH-Potentials are used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
-The paper does not describe any of the contribution of this paper.
+Our submitted paper is the finished version of the ArXiv preprint.
+The submitted paper does not describe any of the contribution of \cite{kswz-erptd-p-20}.
 } is used for such a scenario.
@@
 Second, we compare against our own implementation\footnote{
-We use 16 landmarks generate with the avoid strategy~\cite{gw-cppsp-05}.
-We store all landmark distances consecutive for each node.
-This increases cache efficiency when reading many distances for the same node.
-We do not disable some landmarks during the query, but use all landmarks all the time.
-Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
+We use 16 landmarks generate with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
+%We store all landmark distances consecutive for each node.
+%This increases cache efficiency when reading many distances for the same node.
+%We do not disable some landmarks during the query, but use all landmarks all the time.
+Our ALT implementation is uni-directional, as a bidirectional search creates problems for some settings, such as predicted traffic.
+%Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
 } of ALT~\cite{gw-cppsp-05}.
@@
 With everything activated, our algorithm still has a speedup of 12.2 over the baseline.
-Interestingly, the PTV traffic predictions have a much smaller length increase that the Mapbox predictions.
+Interestingly, the PTV traffic predictions have a much smaller length increase than the Mapbox predictions.
 This results in smaller running times of our algorithm.

### 2021-01-27 paper/ch_potentials.tex
grammar fix, comments, wordsmithing ALT footnote

@@
 We study exact, efficient and practical algorithms for route planning in large road networks.
-Routing applications often requires integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
+Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
@@
 MLD also uses shortcut edges.
+% ich haette das at least eigentlich schon gerne drin. CH erreicht wesentlich mehr als 3 groessenordnungen
 Both approaches achieve speed-ups of three orders of magnitude over Dijkstra's algorithm.
@@
 Our submitted paper is the finished version of the ArXiv preprint.
+% der punkt macht fuer mich keinen Sinn.
+% Dass wir keine der Contributions von kswz-erptd-p-20 beschreiben ist ja offensichtlich, sieht ja jeder in dem Abschnitt.
+% Wichtiger finde ich, dass kswz-erptd-p-20 nix von uns beschreibt.
+% Das meinte ich zumindest mit dem Satz den du geloescht hast.
+% Aber der jetzt folgende Satz sagt nur, dass unser paper hier nix von kswz-erptd-p-20 beschreibt.
 The submitted paper does not describe any of the contribution of \cite{kswz-erptd-p-20}.
@@
 Second, we compare against our own implementation\footnote{
-We use 16 landmarks generate with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
-%We store all landmark distances consecutive for each node.
-%This increases cache efficiency when reading many distances for the same node.
-%We do not disable some landmarks during the query, but use all landmarks all the time.
-Our ALT implementation is uni-directional, as a bidirectional search creates problems for some settings, such as predicted traffic.
-%Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
+We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
+Our ALT implementation is uni-directional.
+In this work, we do not consider bidirectional search as it creates problems for some settings, such as predicted traffic.
 } of ALT~\cite{gw-cppsp-05}.

### 2021-01-27 Rakefile.rb
reproducability: make sure all dijkstra queries get the same tuples

@@ -182,7 +182,8 @@ namespace "exp" do
         sh "cargo run --release --bin chpot_simple_scale_dijkstra --features 'chpot-oracle' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
         sh "cargo run --release --bin chpot_simple_scale_dijkstra --features 'chpot-alt' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
         sh "cargo run --release --bin chpot_simple_scale_dijkstra --features '' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "cargo run --release --bin chpot_simple_scale --features 'chpot-only-topo chpot-no-bcc chpot-no-deg2 chpot-no-deg3' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
+        sh "cargo run --release --bin chpot_simple_scale_dijkstra --features 'chpot-only-topo' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
+        sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale --features 'chpot-only-topo chpot-no-bcc chpot-no-deg2 chpot-no-deg3' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
         sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale --features 'chpot-only-topo chpot-no-deg2 chpot-no-deg3' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
         sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale --features 'chpot-only-topo chpot-no-deg3' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
         sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale --features 'chpot-only-topo' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"

### 2021-01-27 eval/building_blocks.py

@@ -16,9 +16,9 @@ queries = pd.DataFrame.from_records([{
     **algo }
     for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') in ['CH Potentials Query', 'Dijkstra Query']])
 
-queries['BCC'] = ~queries['features'].str.contains('CHPOT_NO_BCC')
-queries['Skip Deg. 2'] = ~queries['features'].str.contains('CHPOT_NO_DEG2')
-queries['Skip Deg. 3'] = ~queries['features'].str.contains('CHPOT_NO_DEG3')
+queries['BCC'] = (queries['algo'] == 'CH Potentials Query') & ~queries['features'].str.contains('CHPOT_NO_BCC')
+queries['Skip Deg. 2'] = (queries['algo'] == 'CH Potentials Query') & ~queries['features'].str.contains('CHPOT_NO_DEG2')
+queries['Skip Deg. 3'] = (queries['algo'] == 'CH Potentials Query') & ~queries['features'].str.contains('CHPOT_NO_DEG3')
 
 def heuristic_name(feature_flags):
     if 'CHPOT_ONLY_TOPO' in feature_flags:

### 2021-01-29 paper/ch_potentials.tex
suggestions for fixes of the things I commented on

@@
 % ich haette das at least eigentlich schon gerne drin. CH erreicht wesentlich mehr als 3 groessenordnungen
-Both approaches achieve speed-ups of three orders of magnitude over Dijkstra's algorithm.
+Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
 
@@
 % Aber der jetzt folgende Satz sagt nur, dass unser paper hier nix von kswz-erptd-p-20 beschreibt.
-The submitted paper does not describe any of the contribution of \cite{kswz-erptd-p-20}.
+\cite{kswz-erptd-p-20} does not describe any of the contribution of this paper.
 } is used for such a scenario.

### 2021-01-31 paper/ch_potentials.tex
mini stuff

@@
 For realistic routing, many additional features need to be considered.
-This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions.%, temporary driving bans for certain types of vehicles.
+This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions. %, temporary driving bans for certain types of vehicles.
 Some applications may have additional application-specific requirements.
@@
 The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
-The exact implementation of this A* search depend on the application.
+The exact implementation of this A* search depends on the application.
 Our approach only provides the heuristic $h_t$ for the A* search.
@@
 % Aber der jetzt folgende Satz sagt nur, dass unser paper hier nix von kswz-erptd-p-20 beschreibt.
-\cite{kswz-erptd-p-20} does not describe any of the contribution of this paper.
+\cite{kswz-erptd-p-20} does not describe any of the contributions of this paper.
 } is used for such a scenario.
@@
 Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
-This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is the most efficient heuristic possible in our model.
+This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
 We fill this array before each query using a reverse Dijkstra search from the target node.

### 2021-01-31 paper/ch_potentials.tex
add basic CH to table

@@
 \caption{
-CH-Potentials performance for different route planning applications.
+CH and CH-Potentials performance for different route planning applications.
 We report average running times and number of queue pushes.
@@
 
-Table~\ref{tab:applications} depicts the running times of CH-Potentials in the various applications, such as those described in Section~\ref{sec:extensions}.
+Table~\ref{tab:applications} mainly depicts the running times and CH-Potentials in the various applications, such as those described in Section~\ref{sec:extensions}.
+For comparison, we also report CH running times.
+A basic CH can only handle the $w_q = w_\ell$ case.
 Speedups compared to extensions of Dijkstra's algorithm for the respective application are depicted.

### 2021-01-31 paper/ch_potentials.tex
compress conclusion to fit page limit

@@
 % This leads to multiple avenues for future research.
-
+%
 Dropping the provable exactness requirement using a setup similar to anytime A* \cite{DBLP:conf/aaai/ZhouH02,DBLP:conf/nips/LikhachevGT03} would be interesting.
-Another promising research avenue would be to investigate other graphs than road networks.
-A lot of research into grid maps exists including a series of competitions called GPPC \cite{DBLP:conf/socs/SturtevantTTUKS15}.
-Hierarchical techniques have been shown to work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
+Another research avenue is investigating non-road graph types.
+Much grid map research exists with series of competitions called GPPC \cite{DBLP:conf/socs/SturtevantTTUKS15}.
+Hierarchical approaches work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
 

### 2021-02-01 paper/ch_potentials.tex
revert app table updates and conclusion combined with some additional wordsmithing

--- 2021-01-31 paper/ch_potentials.tex
@@
 
-Table~\ref{tab:applications} depicts the running times of CH-Potentials in the various applications, such as those described in Section~\ref{sec:extensions}.
-Speedups compared to extensions of Dijkstra's algorithm for the respective application are depicted.
+Table~\ref{tab:applications} depicts the running times of CH-Potentials in various applications, such as those described in Section~\ref{sec:extensions}.
+We report speedups compared to extensions of Dijkstra's algorithm for each application respectively.
 We start with the base case where $w_q = w_\ell$.
-This is the problem variant solved by the basic CH algorithm (a CH query takes 0.163\,ms on average on this graph).
-CH achieves query running times of 0.16\,ms on average.
-CH-Potentials are roughly four times slower but still achieve a huge speedup over Dijkstra of 3243.
+This is the problem variant solved by the basic CH algorithm.
+CH achieves average query running times of 0.16\,ms on OSM Ger.
+CH-Potentials are roughly four times slower but still achieve a huge speedup of 3243 over Dijkstra.
 Such large speedups are typical for CH.
@@
 
-Dropping the provable exactness requirement using a setup similar to anytime A* \cite{DBLP:conf/aaai/ZhouH02,DBLP:conf/nips/LikhachevGT03} would be interesting.
-Another promising research avenue would be to investigate other graphs than road networks.
-A lot of research into grid maps exists including a series of competitions called GPPC \cite{DBLP:conf/socs/SturtevantTTUKS15}.
-Hierarchical techniques have been shown to work on these graphs \cite{DBLP:conf/aaai/UrasK14}.
+Dropping the provable exactness requirement using a setup similar to anytime A*~\cite{DBLP:conf/aaai/ZhouH02,DBLP:conf/nips/LikhachevGT03} would be interesting.
+Another promising research avenue would be to investigate graphs other than road networks.
+A lot of research into grid maps exists including a series of competitions called GPPC~\cite{DBLP:conf/socs/SturtevantTTUKS15}.
+Hierarchical techniques have been shown to work well on these graphs~\cite{DBLP:conf/aaai/UrasK14}.
 

### 2021-03-12 paper/ch_potentials.tex
cleanup styles and use lipics 2021
address reviewer feedback

@@
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
+\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
 
@@
 \newcommand{\xmark}{\ding{55}}
-
-% \setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.
-
-%This is a template for producing LIPIcs articles.
-%See lipics-manual.pdf for further information.
-%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-%for section-numbered lemmas etc., use "numberwithinsect"
-%for enabling cleveref support, use "cleveref"
-%for enabling cleveref support, use "autoref"
 
@@
 
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\EventEditors{David Coudert and Emanuele Natale}
+\EventNoEds{2}
+\EventLongTitle{19th International Symposium on Experimental Algorithms (SEA 2021)}
+\EventShortTitle{SEA 2021}
+\EventAcronym{SEA}
+\EventYear{2021}
+\EventDate{June 7--9, 2021}
+\EventLocation{Nice, France}
+\EventLogo{}
+\SeriesVolume{190}
+\ArticleNo{6}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{document}
@@
 
-%%
-%% By default, the full list of authors will be used in the page
-%% headers. Often, this list is too long, and will overlap
-%% other information printed in the page headers. This command allows
-%% the author to define a more concise list
-%% of authors' names for this purpose.
-% \renewcommand{\shortauthors}{Strasser and Zeitz}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
@@
 MLD also uses shortcut edges.
-% ich haette das at least eigentlich schon gerne drin. CH erreicht wesentlich mehr als 3 groessenordnungen
 Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
@@
 Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
-Comparing against Oracle-A* allows us to measure the overhead of the heuristic evaluation of CH-Potentials.
+CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
+Comparing against Oracle-A* allows us to measure this overhead.
 Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.
@@
 
-\bibliography{dblp,references}
+\bibliography{references}
 \end{document}

### 2021-03-12 paper/ch_potentials.tex
link to github and make sure we dont need another page, url cleanup

@@
 Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile with the target-cpu=native option.
+The source code of our implementation and the experimental evaluation can be found on Github\footnote{\url{https://github.com/kit-algo/ch_potentials}}.
 
 \subparagraph{Inputs and Methodology}
-Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{download.geofabrik.de/europe/germany-200101.osm.pbf}}.
-To obtain the routing graph, we use the import from RoutingKit\footnote{\url{www.github.com/RoutingKit/RoutingKit}}.
+Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
+To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
 The graph has 16M nodes and 35M edges.
-For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{www.mapbox.com}}.
+For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://mapbox.com}}.
 The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
@@
 We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
-Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{www.ptvgroup.com}}.
+Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
 The PTV instances are not OSM-based.
@@
 This corresponds to using Dijkstra's algorithm.
-Second, we compare against our own implementation\footnote{
+Second, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
 We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
@@
 In this work, we do not consider bidirectional search as it creates problems for some settings, such as predicted traffic.
-} of ALT~\cite{gw-cppsp-05}.
 Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.

### 2021-03-12 paper/export_dagstuhl.sh
ready export

@@ -1,0 +1,17 @@
+#!/bin/bash
+
+if [[ $(git diff --shortstat 2> /dev/null | tail -n1) != "" ]] ; then
+  echo "Cant export - would possibly override stuff"
+  exit 0
+fi
+
+sed -i '/^%/ d' ch_potentials.tex
+
+zip dagstuhl.zip \
+ch_potentials.tex \
+references.bib \
+lipics-v2021.cls \
+fig/* \
+table/*
+
+git checkout ch_potentials.tex

### 2021-03-26 paper/submit_arxiv.sh
prepare for arxiv

@@ -10,7 +10,9 @@ sed -i 's/bibliography{references}/input{ch_potentials.bbl}/' ch_potentials.tex
 
 zip arxiv.zip \
 ch_potentials.tex \
-lipics-v2019.cls \
-ch_potentials.bbl
+ch_potentials.bbl \
+lipics-v2021.cls \
+fig/* \
+table/*
 
 git checkout ch_potentials.tex

### 2021-08-09 Rakefile.rb
add figures for many-to-{one, many}

@@ -6,6 +6,8 @@ only_public = !ENV['ONLY_PUBLIC'].nil?
 file "paper/ch_potentials.pdf" => [
   "paper/ch_potentials.tex",
   "paper/fig/scaled_weights.pdf",
+  "paper/fig/lazy_rphast_many_to_one.pdf",
+  "paper/fig/lazy_rphast_many_to_many.pdf",
   "paper/table/applications.tex",
   "paper/table/graphs.tex",
   "paper/table/building_blocks.tex"] do
@@ -20,6 +22,14 @@ task default: "paper/ch_potentials.pdf"
 namespace "fig" do
   file "paper/fig/scaled_weights.pdf" => FileList["#{exp_dir}/scaled_weights/*.json"] + ["eval/scaled_weights.py"] do
     sh "eval/scaled_weights.py"
+  end
+
+  file "paper/fig/lazy_rphast_many_to_one.pdf" => FileList["#{exp_dir}/rphast/*.json"] + ["eval/rphast_comparison.py"] do
+    sh "eval/rphast_comparison.py"
+  end
+
+  file "paper/fig/lazy_rphast_many_to_many.pdf" => FileList["#{exp_dir}/rphast/*.json"] + ["eval/rphast_comparison.py"] do
+    sh "eval/rphast_comparison.py"
   end
 end
 

### 2021-08-09 eval/rphast_comparison.py

@@ -1,0 +1,46 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import matplotlib as mpl
+from matplotlib import pyplot as plt
+import seaborn as sns
+sns.set()
+
+import pandas as pd
+
+import json
+import glob
+import os
+
+base = "exp/"
+cwd = os.getcwd()
+paths = glob.glob(base + "rphast/*.json")
+data = [json.load(open(path)) for path in paths]
+
+queries = pd.DataFrame.from_records([{
+    **algo,
+    'experiment': exp['experiment'],
+    'target_set_size_exp': exp['target_set_size_exp'],
+    'ball_size_exp': exp['ball_size_exp'],
+    'graph': { cwd + '/data/europe/': 'DIMACs Europe', cwd + '/data/osm_ger/': 'OSM Germany', cwd + '/data/osm_europe/': 'OSM Europe' }[run['args'][1]],
+    }
+    for run in data for exp in run['experiments'] for algo in exp['algo_runs']])
+
+queries['running_time_s'] = queries['running_time_ms'] / 1000.0
+
+plt.figure(figsize=(11,4))
+g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_one'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_ms', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
+g.set_ylabel('Running Time [ms]')
+g.set_xlabel('Ball Size $|B|$')
+g.legend().remove()
+plt.tight_layout()
+g.get_figure().savefig('paper/fig/lazy_rphast_many_to_one.pdf')
+
+
+plt.figure(figsize=(11,4))
+g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_many'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_s', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
+g.set_ylabel('Running Time [s]')
+g.set_xlabel('Ball Size $|B|$')
+g.legend().remove()
+plt.tight_layout()
+g.get_figure().savefig('paper/fig/lazy_rphast_many_to_many.pdf')

### 2021-08-09 paper/ch_potentials.tex
remove pagebreaks
remove line numbers
prepare for arxiv
add some very dirty writing on many-to-{one,many} lazy rphast

@@
 \usepackage[algo2e,vlined]{algorithm2e}
+
+\nolinenumbers
+\hideLIPIcs
 
@@
 \end{abstract}
-
-\newpage
 
@@
 \KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$P[x]$: memoized potential at $x$, $\bot$ initially}
-\SetKwFunction{Pot}{Pot}
+\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
+\SetKwFunction{Dist}{ComputeAndMemoizeDist}
 \SetKwProg{Fn}{Function}{:}{}
-\Fn{\Pot{$x$}}{
-	\If{$P[x] = \bot$}{
-		$P[x]\leftarrow B[x]$\;
+\Fn{\Dist{$x$}}{
+	\If{$D[x] = \bot$}{
+		$D[x]\leftarrow B[x]$\;
 		\For{all up edges $xy$ in $G^+_\ell$}{
-                        $P[x]\leftarrow\min\{P[x],w_\ell(xy)+\Pot(y)\}$\;
+                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
 		}
@@
 }
-\caption{CH-Potentials Algorithm}
+\caption{Lazy RPHAST Algorithm}
 \label{algo:pot}
@@
 A* with this heuristic is the basic CH-Potentials algorithm.
+
+\subsection{CCH-Potentials}
+
+\begin{algorithm2e}
+\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
+\KwData{$P[x]$: parent node of $x$ in the elimination tree}
+\SetKwFunction{Dist}{ComputeAndMemoizeDist}
+\SetKwProg{Fn}{Function}{:}{}
+\Fn{\Dist{$x$}}{
+	\If{$D[x] = \bot$}{
+		\If{$D[P[x]] = \bot$}{
+			$\Dist(P[x])$\;
+		}
+		$D[x]\leftarrow B[x]$\;
+		\For{all up edges $xy$ in $G^+_\ell$}{
+                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
+		}
+	}
+	\Return{$P[x]$}\;
+}
+\caption{Elimination Tree Lazy RPHAST Algorithm}
+\label{algo:cch_pot}
+\end{algorithm2e}
+
+Algorithm~\ref{algo:pot} can be applied to CCH without any modifications.
+However, similarly to the elimination tree query algorithm, we can also exploit the elimination tree for the Lazy RPHAST algorithm.
+Computing the distance $D[x]$ requires knowing the final distance of all upward neighbors.
+In Algorithm~\ref{algo:pot} we have to compute these distances recursively.
+In~\cite{}, it was proven that the set of the ancestors in the elimination tree of a node $x$ equals the CH search space of $x$. % TODO cite
+Since all upward neighbors of $x$ are by definition included in the search space, they are also ancestors in the elimination tree.
+If the distance of the parent in the elimination tree has already been computed, the distances of all other upward neighbors have also been computed.
+Thus, it is sufficient to recursively (or iteratively) ensure that the distance of the parent in the elimination tree has been computed.
+As soon as a node with final distance is encountered, all ancestors are known to have their final distance.
+This allows for the more efficient implementation depicted in Algorithm~\ref{algo:cch_pot}.
+
+\subsection{Many-to-Many with Lazy RPHAST}
+
+In this section, we discuss how the Lazy RPHAST schema can be applied in a Many-to-Many context.
+To the best of our knowledge, the fastest way to compute a full distance table between a set of sources and a set of targets is RPHAST utilizing SSE instructions to process several sources at once.
+First, the selection phase is performed for the full target set.
+Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
+This second step can be improved by processing $k$ (typically 16) sources at once using SIMD vector instructions (SSE RPHAST).
+
+Since our algorithm is a variation RPHAST, we can also apply it to the Many-to-Many problem.
+% TODO explain
+However, a naive approach would explore the upward search space of all the source over and over again.
+This naive approach would not be competitive.
+Luckily, we can do better by using a different One-to-Many CH variant -- the Bucket-CH algorithm -- to select all targets at once.
+The Bucket CH selection phase works as follows:
+For each node $x$, we maintain a list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
+These buckets can be populated by running a backward CH search from each $t_i$.
+An even faster way to compute the buckets is to initialize a backward CH search with all nodes at once, order the queue by node rank and use the buckets as labels.
+A bucket can be propagated along an edge by increasing each entry by the edge weight, and merged with another tentative bucket with a parallel sweep over both buckets.
+% TODO way too dense
+Once the buckets have been computed, we can compute distances to all targets once with Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}).
+The only difference is that $D[x]$ is now an array of distances to each $t_i$.
+We evaluate this algorithm in Section~\ref{}.
 
@@
 
-\pagebreak
-
 \bibliography{references}

### 2021-08-09 paper/ch_potentials.tex
reordering, some a* and preliminary writing

@@
 
-\section{Algorithm}\label{sec:main-algo}
+\section{Preliminaries}\label{sec:preliminaries}
 
-In this section, we first discuss the framework in which CH-Potentials can be used.
-Then, we describe the building blocks of CH-Potentials: Contraction Hierarchies and PHAST, a CH extension.
-Finally, we introduce the CH-Potentials heuristic.
+We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
+Given nodes $s$ and $t$ we want to obtain the path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
+We denote this shortest path weight as the distance $\operatorname{dist}_w(s,t)$ between $s$ and $t$.
 
-\subsection{Formal Setup: Inputs, Outputs, and Phases}
+\subsection{Dijkstra's Algorithm}
 
-In this paper, we consider different applications, with slightly different problem models.
-The goal is always to quickly answer many shortest path queries.
-For the purpose of describing our framework, we establish a shared notation:
-Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
-However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
-In the simplest case, $w_q$ will be scalar edge weights.
-However, this is not a requirement.
-It can be any function that computes a weight for an edge.
-This function can also take additional parameters from the state of the search.
-For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
-However, values of $w_q$ might change between queries.
-In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
+Dijkstra's algorithm can be used to solve this point to point shortest path problem.
+It maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
+Initially, all distances are set to $\infty$.
+The start node distance $d[s]$ is set to zero and the queue initialized with $s$.
+Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
+Outgoing edges $uv$ are relaxed, i.e. $d[v] = \min(d[v], d[u] + w(uv))$.
+If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
+Once $t$ was settled, $d[t]$ contains the distance between $s$ and $t$.
 
-To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
-The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
-$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
-We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
-The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
-$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
-In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
+\subsection{A*}
 
-The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
-The exact implementation of this A* search depends on the application.
-Our approach only provides the heuristic $h_t$ for the A* search.
-In contrast, the preprocessing phase remains the same for all applications.
+A* is a goal-directed variant of Dijkstra's algorithm.
+% A* uses a \emph{heuristic} to guide the search towards the goal.
+It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target.
+The priority queue is then ordered by $d[v] + h_t(v)$.
+Thus, nodes closer to the target a explored earlier.
+See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
+A*'s running time crucially depends on how tight this estimate is.
 
-Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
-By requirement and because of the triangle inequality the following holds:
-\[
-w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
-\]
-Thus, A* will always determine the correct shortest distances.
+A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights} $w_{h_t}$.
+This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
+A heuristic function is \emph{feasible} iff $w_{h_t} \geq 0$.
 
@@
 See Figure~\ref{fig:ch} for a proof sketch.
-From every shortest path (red) in $G$, an up-down path of equal length in $G+$ (blue) exists.
+From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) exists.
 Thus, we can restrict our search to up-down paths in $G^+$.
@@
 
-Using the CH query algorithm, we can already give a simple heuristic.
-The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
-This yields tight estimates but a high overhead for the heuristic evaluation.
-While a single CH query is fast, answering one for every node explored in the $A^*$ search is slow.
-Fortunately, we can do better.
+% Using the CH query algorithm, we can already give a simple heuristic.
+% The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
+% This yields tight estimates but a high overhead for the heuristic evaluation.
+% While a single CH query is fast, answering one for every node explored in the $A^*$ search is slow.
+% Fortunately, we can do better.
 % For this we need another component called PHAST~\cite{dgnw-phast-13}.
 
-\subsection{PHAST based Heuristic}
+\subsection{Customizable Contraction Hierarchy (CCH)}
+
+% need elimination tree
+
+\subsection{PHAST}
 
@@
 
-PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node.
-The preprocessing phase remains unchanged.
+PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the one-to-all case works analogously).
+The preprocessing phase remains the same as for CH.
 The query phase is split into two steps.
@@
 After all levels are processed, the shortest distances from all nodes to $t$ were computed.
-%PHAST is slightly faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
-%PHAST's main advantage is that can easily be parallelized.
-%However, we will not consider parallelization in this paper.
-%We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
 Pseudo-code is provided in Algorithm~\ref{algo:phast}.
-Using PHAST, we can also compute a tight A* heuristic.
-In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
-Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.
+PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
+It can also easily be parallelized.
+% However, we will not consider parallelization in this paper.
+We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
+% Using PHAST, we can also compute a tight A* heuristic.
+% In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
+% Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.
 
 % This PHAST-based algorithm works.
-The $H$ lookup and by extension the $A^*$ search is indeed fast.
-However, the PHAST step before the search is comparatively expensive.
-The reason is that the distances towards $t$ are computed for \emph{all} nodes.
-Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.
+% The $H$ lookup and by extension the $A^*$ search is indeed fast.
+% However, the PHAST step before the search is comparatively expensive.
+% The reason is that the distances towards $t$ are computed for \emph{all} nodes.
+% Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.
 
-\subsection{CH-Potentials}
+\subsection{RPHAST}
+
+RPHAST, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node.
+Given a set of source nodes $S$ and a target node $t$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
+This step is called \emph{selection}.
+Then, the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the new restricted subgraph.
+RPHAST is particularly effective when many targets are queried for the same source set $S$.
+If the source set changes often, selection times may become problematic.
+
+\section{The Incremental Many-to-One Problem}\label{sec:lazy-rphast}
+
+% TODO not so great intro directly after PHAST/RPHAST
+Route planning algorithms typically focus on the point-to-point shortest path problem.
+However, certain applications need distances between sets $S$ and $T$ of nodes.
+This is variant is known as the \emph{many-to-many} problem variant.
+In some cases on of the sets is a single node.
+This is then called a \emph{one-to-many} or \emph{many-to-one} problem.
+In this section, we study (discuss) a very specific setting arises naturally from A* heuristics.
+We denote this problem as the \emph{incremental many-to-one} problem.
+The first step is the \emph{target selection} where the target node is provided.
+Then, an arbitrary number of source nodes $s_i$ are given one after another and for each source the distance to the target has to be computed.
+
+We consider the combined running time of the target selection and each incremental query as the running time to answer such an incremental many-to-one query.
+Since the target selection time is included, computing the distances to all nodes with Dijkstra or PHAST is too slow.
+Fortunately, we can do better.
+
+\subsection{Lazy RPHAST}
 
@@
 
-Fortunately, the PHAST computation can be done lazily using memoization as depicted in Algorithm~\ref{algo:pot}.
-In a first step, we run the backward CH search from $t$ to obtain an array $B$.
+The core idea of our algorithm is to do the PHAST computation lazily and using memoization as depicted in Algorithm~\ref{algo:pot}.
+In the target selection, we first run the backward CH search from $t$ to obtain an array $B$.
 $B[x]$ is the minimum down $xt$-path distance or $+\infty$, if there is no such path.
@@
 
-To compute the heuristic $h_t(x)$, we recursively compute for all up-edges $(x,y)$ the heuristic $h_t(y)$.
-Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_y\{w_\ell(x,y) + h_t(y)\}$.
-As not all shortest up-down paths contain an up-edge, we set $h_t(x) = \min \{ B[x], d \}$.
-This calculation is correct, as it computes the minimum up-down $xt$-path distance in $G^+_\ell$, which corresponds to the minimum $xt$-path distance in a CH.
-A* with this heuristic is the basic CH-Potentials algorithm.
+To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i,x)$ the distance $D[x]$.
+Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i,x) + D[x]\}$.
+As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
+This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
+% TODO consistent uv vs u,v
+% A* with this heuristic is the basic CH-Potentials algorithm.
 
-\subsection{CCH-Potentials}
+\subsection{Lazy RPHAST on CCH}
 
@@
 
-\section{Low Degree A* Improvements}
+\section{A* with CH-Potentials}\label{sec:main-algo}
 
-\label{sec:low-deg-improvment}
+In this section, we first discuss the framework in which CH-Potentials can be used.
+Then, we describe the building blocks of CH-Potentials: Contraction Hierarchies and PHAST, a CH extension.
+Finally, we introduce the CH-Potentials heuristic.
+
+\subsection{Formal Setup: Inputs, Outputs, and Phases}
+
+In this paper, we consider different applications, with slightly different problem models.
+The goal is always to quickly answer many shortest path queries.
+For the purpose of describing our framework, we establish a shared notation:
+Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
+However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
+In the simplest case, $w_q$ will be scalar edge weights.
+However, this is not a requirement.
+It can be any function that computes a weight for an edge.
+This function can also take additional parameters from the state of the search.
+For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
+However, values of $w_q$ might change between queries.
+In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
+
+To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
+The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
+$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
+We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
+The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
+$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
+In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
+
+The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
+The exact implementation of this A* search depends on the application.
+Our approach only provides the heuristic $h_t$ for the A* search.
+In contrast, the preprocessing phase remains the same for all applications.
+
+Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
+By requirement and because of the triangle inequality the following holds:
+\[
+w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
+\]
+Thus, A* will always determine the correct shortest distances.
+
+\subsection{CH-Potentials Heuristic}
+
+CH-Pot is A* with Lazy RPHAST Pot.
+Given $s$ and $t$ upward search from $t$
+Then compute $h_t(x)$ for with ComputeAndMemoizeDist
+
+\subsection{Bidirectional A*}
+
+Bidir A* not as straightforward which one might expect.
+As pointed out in~\cite{gh-cspas-05}, one can not simply use a forward potential $h_t$ for the forward search and a backward potential $h_s$ for the reverse search.
+Since these searches would operate on different reduced graphs, it is unclear when the searches can terminate.
+To the best of our knowledge, no better stopping criterion exists than running both searches until the minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
+That is the same stopping crit as for a unidir search..
+symmetric
+Thus, the symmetric bidirectional algorithm performs double the work than a simple unidir search.
+can be slightly improved through pruning $\overrightarrow{D[x]} + h_t(x) > \mu$.
+Still not competitive
+
+average potential - reduced graph with average of reduced edge weights
+but average pot in each dir worse than orig pot for dir
+also overhead
+
+we propose an improved pruning crit, one which takes the search progress into account (through the min queue keys), which makes the simpler symmetric approach competitive or even faster than the average pot
+If $x$ was not yet settled by reverse search with minimum queue key $\overleftarrow{k}$ then
+$\operatorname{dist}(x,t) \geq \overleftarrow{k} - h_s(x)$.
+Thus, if $\overrightarrow{D[x]} + \overleftarrow{k} - h_s(x) > \mu$ we can discard $x$.
+
+% direction switch strats!
+
+\subsection{Low Degree A* Improvements}\label{sec:low-deg-improvment}
 
@@
 
-\subsection{Skip Degree Two Nodes}
+\subsubsection{Skip Degree Two Nodes}
 
@@
 
-\subsection{Skip Degree Three Nodes}
+\subsubsection{Skip Degree Three Nodes}
 
@@
 
-\subsection{Stay in Largest Biconnected Component}
-\label{sec:largested-biconnected-component}
+\subsubsection{Stay in Largest Biconnected Component}\label{sec:largested-biconnected-component}
 
@@
 
-\section{Applications}
-\label{sec:extensions}
+\subsection{Applications}\label{sec:extensions}
 
@@
 
-\subsection{Avoiding Tunnels and/or Highways}
+\subsubsection{Avoiding Tunnels and/or Highways}
 \label{sec:no-tunnel-highway}
@@
 
-\subsection{Forbidden Turns and Turn Costs}
+\subsubsection{Forbidden Turns and Turn Costs}
 \label{sec:no-turns}
@@
 
-\subsection{Predicted Traffic or Time-Dependent Routing}
+\subsubsection{Predicted Traffic or Time-Dependent Routing}
 \label{sec:predicted-traffic}
@@
 
-\subsection{Live and Predicted Traffic}
+\subsubsection{Live and Predicted Traffic}
 \label{sec:live-predicted-traffic}
@@
 
-\subsubsection{Three-Phase Setups}
+\subsubsection{CCH Potentials}
+% TODO CCH-Pot
 
@@
 
-\subsection{Temporary Driving Bans}
+\subsubsection{Temporary Driving Bans}
 
@@
 Further, a trade-off between arrival time and route quality is considered.
+
+\subsubsection{Alternative Routes with the Penalty Method}
+
+% TODO
 

### 2021-08-11 paper/ch_potentials.tex
writing

@@
 
+Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
+We call this a \emph{backward search}.
+Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
+The searches can terminate when the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$.
+
 \subsection{A*}
@@
 This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
-A heuristic function is \emph{feasible} iff $w_{h_t} \geq 0$.
+A heuristic function is called \emph{feasible} if $w_{h_t} \geq 0$ for all edges.
 
@@
 %
-In the preprocessing phase, a CH adds \emph{shortcut} edges to the input graph $G$ to obtain $G^+$.
+In the preprocessing phase, first, a node ordering is computed heuristically.
+Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
 This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
@@
 
-% need elimination tree
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} are a special variant of Contraction Hierarchies.
+They use a node importance order independent of the weight function.
+This allows splitting the preprocessing in two subphases.
+The first one is slow but independent of the weight function.
+The second one, called \emph{customization}, is fast (less than a second) and applies a weight function to the preprocessed data.
+This allows regular updates to the weight function, for example to incorporate live traffic.
+
+As for CH, the result of the preprocessing is an augmented graph $G^+$ fulfilling the same properties as for CH.
+Thus, the CH query algorithms can be applied without modifications.
+However, the CCH preprocessing grants some additional useful properties.
+During preprocessing, an \emph{elimination tree} can be constructed.
+The root of the elimination tree is the most important node.
+For every other node its parent is its least important upward neighbor.
+In~\cite{bcrw-s-16}, it was proven that set of ancestors of a node in the elimination tree is equal to the nodes reachable in a forward/backward CH search.
+This makes it possible to explore the CCH search space with a more efficient shortest path algorithm for DAGs:
+Traverse elimination tree from source and target to root node and for each node relax all up-edges (reversed down-edges for the backward search).
+Because this algorithm does not require a queue, it is faster than a Dijkstra based CH query.
 
@@
 
-RPHAST, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node.
-Given a set of source nodes $S$ and a target node $t$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
+RPHAST, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously).
+Given a set of source nodes $S$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
 This step is called \emph{selection}.
-Then, the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the new restricted subgraph.
+Then, in the query step a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
 RPHAST is particularly effective when many targets are queried for the same source set $S$.
@@
 
-% TODO not so great intro directly after PHAST/RPHAST
-Route planning algorithms typically focus on the point-to-point shortest path problem.
+Most route planning algorithms focus on the point-to-point shortest path problem.
 However, certain applications need distances between sets $S$ and $T$ of nodes.
 This is variant is known as the \emph{many-to-many} problem variant.
-In some cases on of the sets is a single node.
+In some cases one of the sets is a single node.
 This is then called a \emph{one-to-many} or \emph{many-to-one} problem.
-In this section, we study (discuss) a very specific setting arises naturally from A* heuristics.
+RPHAST, as discussed in the previous paragraph, can be applied to these problems.
+
+In this section, we discuss a very specific setting arises naturally from A* heuristics.
+Here, the target node is known in the selection step but the source nodes are only queried one after another.
 We denote this problem as the \emph{incremental many-to-one} problem.
@@
 
-We consider the combined running time of the target selection and each incremental query as the running time to answer such an incremental many-to-one query.
+We consider the combined running time of the target selection and each incremental query as the total running time to answer an incremental many-to-one query.
 Since the target selection time is included, computing the distances to all nodes with Dijkstra or PHAST is too slow.
+Also, since the source set is only provided incrementally, RPHAST in its basic form is not well suited to our problem.
 Fortunately, we can do better.
@@
 }
-\caption{Lazy RPHAST Algorithm}
+\caption{Lazy RPHAST algorithm}
 \label{algo:pot}
@@
 
-The core idea of our algorithm is to do the PHAST computation lazily and using memoization as depicted in Algorithm~\ref{algo:pot}.
+The core idea of our algorithm is to do the (R)PHAST computation lazily and using memoization as depicted in Algorithm~\ref{algo:pot}.
 In the target selection, we first run the backward CH search from $t$ to obtain an array $B$.
@@
 
+Algorithm~\ref{algo:pot} can be applied to CCH without any modifications.
+However, we can also exploit the elimination tree for the Lazy RPHAST algorithm.
+Computing the distance $D[x]$ requires knowing the final distance of all upward neighbors.
+In Algorithm~\ref{algo:pot} we have to compute these distances recursively.
+All upward neighbors of $x$ are by definition included in the search space, thus they are also ancestors in the elimination tree.
+If the distance of the parent in the elimination tree has already been computed, the distances of all other upward neighbors also has to have been computed.
+Thus, it is sufficient to recursively (or iteratively) ensure that the distance of the parent in the elimination tree has been computed.
+As soon as a node with final distance is encountered, all ancestors are known to have their final distance.
+This allows for the more efficient implementation depicted in Algorithm~\ref{algo:cch_pot}.
+
 \begin{algorithm2e}
@@
 	\If{$D[x] = \bot$}{
-		\If{$D[P[x]] = \bot$}{
+		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
 			$\Dist(P[x])$\;
@@
 }
-\caption{Elimination Tree Lazy RPHAST Algorithm}
+\caption{Elimination tree based Lazy RPHAST algorithm}
 \label{algo:cch_pot}
@@
 
-Algorithm~\ref{algo:pot} can be applied to CCH without any modifications.
-However, similarly to the elimination tree query algorithm, we can also exploit the elimination tree for the Lazy RPHAST algorithm.
-Computing the distance $D[x]$ requires knowing the final distance of all upward neighbors.
-In Algorithm~\ref{algo:pot} we have to compute these distances recursively.
-In~\cite{}, it was proven that the set of the ancestors in the elimination tree of a node $x$ equals the CH search space of $x$. % TODO cite
-Since all upward neighbors of $x$ are by definition included in the search space, they are also ancestors in the elimination tree.
-If the distance of the parent in the elimination tree has already been computed, the distances of all other upward neighbors have also been computed.
-Thus, it is sufficient to recursively (or iteratively) ensure that the distance of the parent in the elimination tree has been computed.
-As soon as a node with final distance is encountered, all ancestors are known to have their final distance.
-This allows for the more efficient implementation depicted in Algorithm~\ref{algo:cch_pot}.
-
 \subsection{Many-to-Many with Lazy RPHAST}
 
-In this section, we discuss how the Lazy RPHAST schema can be applied in a Many-to-Many context.
+In this section, we discuss how the Lazy RPHAST algorithm can be applied in a many-to-many context.
 To the best of our knowledge, the fastest way to compute a full distance table between a set of sources and a set of targets is RPHAST utilizing SSE instructions to process several sources at once.
@@
 
-Since our algorithm is a variation RPHAST, we can also apply it to the Many-to-Many problem.
-% TODO explain
-However, a naive approach would explore the upward search space of all the source over and over again.
-This naive approach would not be competitive.
-Luckily, we can do better by using a different One-to-Many CH variant -- the Bucket-CH algorithm -- to select all targets at once.
+Since RPHAST is the fastest known algorithm for many-to-many computations, investing how well Lazy RPHAST performs in this setting appears to be an interesting question.
+The naive approach would be to select one target after another and for each one doing the full lazy exploration from all sources.
+However, this approach would explore the upward search space of all the sources over and over again.
+This would not be competitive.
+
+Luckily, we can do better by using a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
 The Bucket CH selection phase works as follows:
-For each node $x$, we maintain a list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
-These buckets can be populated by running a backward CH search from each $t_i$.
-An even faster way to compute the buckets is to initialize a backward CH search with all nodes at once, order the queue by node rank and use the buckets as labels.
-A bucket can be propagated along an edge by increasing each entry by the edge weight, and merged with another tentative bucket with a parallel sweep over both buckets.
-% TODO way too dense
-Once the buckets have been computed, we can compute distances to all targets once with Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}).
-The only difference is that $D[x]$ is now an array of distances to each $t_i$.
-We evaluate this algorithm in Section~\ref{}.
+For each node $x$, we maintain an initially empty list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
+These buckets can be populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
+Now we can utilize Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) to compute distances from a given source to \emph{all} targets at once.
+We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
+% TODO formatting
+After calling the adjusted ComputeAndMemoizeDist for each source, we know the shortest distances between each pair of source and target node.
+This way, we explore the search space of the sources only once.
+
+The Bucket-CH selection phase still explores the shared parts of the search space of the targets several times.
+This can also be improved with the following modified CH backward search:
+All target nodes are inserted into the queue at once.
+The priority queue is ordered by importance instead of distance.
+Distance labels are buckets instead if scalar values.
+The buckets of each target node are initialized with a single entry for the respective node with distance zero.
+When relaxing an edge, the bucket of the head nodes is merged with the bucket of the tail node with each distance increased by the weight of the edge.
+By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
+We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{todo}.
 
@@
 
-In this section, we first discuss the framework in which CH-Potentials can be used.
-Then, we describe the building blocks of CH-Potentials: Contraction Hierarchies and PHAST, a CH extension.
-Finally, we introduce the CH-Potentials heuristic.
+In this section, we describe how to use the Lazy RPHAST algorithm to build a fast and tight heuristic for A*.
+We also discuss CH-Potentials for the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
+Further, we present several low degree node optimization which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
+Finally, we introduce an algorithmic framework to apply A* with CH-Potentials to a variety of practical route planning problems.
 
-\subsection{Formal Setup: Inputs, Outputs, and Phases}
+The CH-Potentials heuristic is a straightforward application of the Lazy RPHAST algorithm.
+In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
+Then, the heuristic function $h_t(x)$ is implemented by call the ComputeAndMemoizeDist for node $x$.
 
-In this paper, we consider different applications, with slightly different problem models.
-The goal is always to quickly answer many shortest path queries.
-For the purpose of describing our framework, we establish a shared notation:
-Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
-However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
-In the simplest case, $w_q$ will be scalar edge weights.
-However, this is not a requirement.
-It can be any function that computes a weight for an edge.
-This function can also take additional parameters from the state of the search.
-For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
-However, values of $w_q$ might change between queries.
-In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
-
-To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
-The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
-$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
-We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
-The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
-$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
-In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
-
-The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
-The exact implementation of this A* search depends on the application.
-Our approach only provides the heuristic $h_t$ for the A* search.
-In contrast, the preprocessing phase remains the same for all applications.
-
-Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
-By requirement and because of the triangle inequality the following holds:
-\[
-w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
-\]
-Thus, A* will always determine the correct shortest distances.
-
-\subsection{CH-Potentials Heuristic}
-
-CH-Pot is A* with Lazy RPHAST Pot.
-Given $s$ and $t$ upward search from $t$
-Then compute $h_t(x)$ for with ComputeAndMemoizeDist
+This yields a \emph{perfect} heuristic, at least as long as the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function.
+This case is of course not particularly interesting.
+One could just answer the shortest path query directly with a CH query.
+This approach becomes useful when the query runs on a \emph{different but related} graph or weight function than the preprocessing.
+We will discuss such applications in Section~\ref{sec:extensions}.
+But before that, we discuss two extensions/optimizations to A* which we will utilize.
 
@@
 Thus, the symmetric bidirectional algorithm performs double the work than a simple unidir search.
-can be slightly improved through pruning $\overrightarrow{D[x]} + h_t(x) > \mu$.
+can be slightly improved through pruning $\overrightarrow{D}[x] + h_t(x) > \mu$.
 Still not competitive
@@
 $\operatorname{dist}(x,t) \geq \overleftarrow{k} - h_s(x)$.
-Thus, if $\overrightarrow{D[x]} + \overleftarrow{k} - h_s(x) > \mu$ we can discard $x$.
+Thus, if $\overrightarrow{D}[x] + \overleftarrow{k} - h_s(x) > \mu$ we can discard $x$.
 
@@
 % Many extended problems can therefore be solved using the CH-Potentials framework.
+We start by establishing a common formal framework for the use of CH-Potentials.
 We describe some extended routing problems and how to apply CH-Potentials to them.
 Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
+
+\subsubsection{Formal Setup: Inputs, Outputs, and Phases}
+
+In this paper, we consider different applications, with slightly different problem models.
+The goal is always to quickly answer many shortest path queries.
+For the purpose of describing our framework, we establish a shared notation:
+Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
+However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
+In the simplest case, $w_q$ will be scalar edge weights.
+However, this is not a requirement.
+It can be any function that computes a weight for an edge.
+This function can also take additional parameters from the state of the search.
+For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
+However, values of $w_q$ might change between queries.
+In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
+
+To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
+The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
+$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
+We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
+The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
+$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
+In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
+
+The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
+The exact implementation of this A* search depends on the application.
+Our approach only provides the heuristic $h_t$ for the A* search.
+In contrast, the preprocessing phase remains the same for all applications.
+
+Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
+By requirement and because of the triangle inequality the following holds:
+\[
+w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
+\]
+Thus, A* will always determine the correct shortest distances.
 
@@
 
-% TODO
+Important feature alternative routes
+One popular framework is the penalty method
+It computes an alternative graph instead of a set of routes
+the graph is constructed iteratively
+in each iteration, first, a shortest path between s and t is computed
+if the path or certain parts of it meet certain quality criteria add to alternative graph
+finally penalize all edge weights of current path and possibly adjoined edges and continue until a certain stopping criterion is fulfilled
+
+running time dominated by shortest path computations
+since edge weights only increased natural fit for A* with CH-Pot
+lots of details to implement and define
+we take config of kobitzsch diss
+quality eval over there
+here only performance
+
+for sake of reproducibility we briefly discuss the details of our configuration
+code from authors
+eps/psi = 1.25
+penalization schema
+	exp psi
+	asymmetric additive
+stopping crit
+	saturation
+	max it
+	mas new parts
+
+feasibility
+
+perf
+	pruning by orig
+	pruning by penalized
+	parallelization
 

### 2021-08-20 Rakefile.rb
add cch-pot to some exps

@@ -137,6 +137,12 @@ namespace "prep" do
           "#{graph}/lower_bound_ch/forward_first_out #{graph}/lower_bound_ch/forward_head #{graph}/lower_bound_ch/forward_weight " +
           "#{graph}/lower_bound_ch/backward_first_out #{graph}/lower_bound_ch/backward_head #{graph}/lower_bound_ch/backward_weight")
     end
+
+    file graph + "cch_perm" => ["code/rust_road_router/lib/InertialFlowCutter/build/console"] do
+      Dir.chdir "code/rust_road_router" do
+        sh "./flow_cutter_cch_order.sh #{graph} #{Etc.nprocessors}"
+      end
+    end
   end
 
   static_graphs.each do |graph|
@@ -189,29 +195,14 @@ namespace "exp" do
     end
   end
 
-  task building_blocks: static_graphs.map { |g| g + 'lower_bound_ch' } + ["#{exp_dir}/building_blocks"] do
+  task building_blocks: ["#{osm_ger}lower_bound_ch", "#{osm_ger}cch_perm", "#{exp_dir}/building_blocks"] do
     Dir.chdir "code/rust_road_router" do
-      static_graphs.each do |graph|
-        sh "cargo run --release --bin chpot_simple_scale_dijkstra --features 'chpot-oracle' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "cargo run --release --bin chpot_simple_scale_dijkstra --features 'chpot-alt' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "cargo run --release --bin chpot_simple_scale_dijkstra --features '' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "cargo run --release --bin chpot_simple_scale_dijkstra --features 'chpot-only-topo' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale --features 'chpot-only-topo chpot-no-bcc chpot-no-deg2 chpot-no-deg3' -- #{graph} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
[…]

### 2021-08-20 eval/rphast_comparison.py
preliminary experiments

@@ -13,7 +13,6 @@ import glob
 import os
 
 base = "exp/"
-cwd = os.getcwd()
 paths = glob.glob(base + "rphast/*.json")
 data = [json.load(open(path)) for path in paths]
 
@@ -22,7 +21,7 @@ queries = pd.DataFrame.from_records([{
     'experiment': exp['experiment'],
     'target_set_size_exp': exp['target_set_size_exp'],
     'ball_size_exp': exp['ball_size_exp'],
-    'graph': { cwd + '/data/europe/': 'DIMACs Europe', cwd + '/data/osm_ger/': 'OSM Germany', cwd + '/data/osm_europe/': 'OSM Europe' }[run['args'][1]],
+    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
     }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs']])
 
@@ -30,17 +29,18 @@ queries['running_time_s'] = queries['running_time_ms'] / 1000.0
 
 plt.figure(figsize=(11,4))
 g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_one'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_ms', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
-g.set_ylabel('Running Time [ms]')
 g.set_xlabel('Ball Size $|B|$')
-g.legend().remove()
+g.set_ylabel('Running Time [ms]')
+g.set_yscale('log')
+g.legend().set_title('')
 plt.tight_layout()
 g.get_figure().savefig('paper/fig/lazy_rphast_many_to_one.pdf')
 
 
 plt.figure(figsize=(11,4))
 g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_many'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_s', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
-g.set_ylabel('Running Time [s]')
 g.set_xlabel('Ball Size $|B|$')
[…]

### 2021-08-20 paper/ch_potentials.tex
writing on penalty

@@
 \usepackage[algo2e,vlined]{algorithm2e}
+
+\newcommand*{\dist}{\mathcal{D}}
 
@@
 A* uses a \emph{heuristic} to guide the search towards the goal.
-A heuristic is function that maps a node $v$ onto an estimate of the distance from $v$ to the goal.
+A heuristic is function that maps a node to an estimate of the remaining distance to the goal.
 A*'s running time crucially depends on how tight this estimate is.
@@
 For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
-% The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
+The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
+
 A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
@@
 Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
-%
+
 CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
@@
 On continental sized networks, TD-CRP can only compute approximate shortest distances (rather than paths).
-%
+
 In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH).
@@
 % Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
+
 Other extensions studied include electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11} or multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10}.
-%
+
+Another feature which has received quite some attention is the computation alternative routes~\cite{bdgs-argrn-11,adgw-arrn-13}.
+Even formally defining what makes a good alternative route is non-trivial~\cite{bdgs-argrn-11}.
+A variety of paradigms exists.
+Competitive approaches require careful adjustment of their respective hierarchical speedup technique for the targeted paradigm.
+In~\cite{krs-eepma-13}, the CRP customization is utilized to repeatedly compute shortest paths on graphs with slightly modified weights.
+In~\cite{k-hdara-13}, partial shortest path trees are extracted from CH search spaces to find via-node candidates for alternative routes.
+In~\cite{barth2019alternative}, a multi-criteria CH is extended to compute alternative routes with regards to multiple edge weight functions.
+
 While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
@@
 Given nodes $s$ and $t$ we want to obtain the path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
-We denote this shortest path weight as the distance $\operatorname{dist}_w(s,t)$ between $s$ and $t$.
+We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
 
@@
 
-Dijkstra's algorithm can be used to solve this point to point shortest path problem.
+Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this point to point shortest path problem.
 It maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
@@
 Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
-The searches can terminate when the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$.
-
-\subsection{A*}
+Typically, the searches are interleaved by alternating settling a node from each direction.
+Another common approach is to always settle a node from the direction with the smaller minimal queue key.
+Theoretically any direction interleaving strategy is possible.
+The searches can also be run in parallel in two threads.
+When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.
+
+\subsection{A*}\label{sec:a_star}
 
@@
 It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target.
-The priority queue is then ordered by $d[v] + h_t(v)$.
+The priority queue is ordered by $d[v] + h_t(v)$.
 Thus, nodes closer to the target a explored earlier.
-See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
-A*'s running time crucially depends on how tight this estimate is.
-
-A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights} $w_{h_t}$.
+
+A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
 This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
@@
 	}
-	\Return{$P[x]$}\;
+	\Return{$D[x]$}\;
 }
@@
 
-To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i,x)$ the distance $D[x]$.
-Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i,x) + D[x]\}$.
+To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i x)$ the distance $D[x]$.
+Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i x) + D[x]\}$.
 As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
 This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
-% TODO consistent uv vs u,v
 % A* with this heuristic is the basic CH-Potentials algorithm.
@@
 We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
-% TODO formatting
-After calling the adjusted ComputeAndMemoizeDist for each source, we know the shortest distances between each pair of source and target node.
+After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target node.
 This way, we explore the search space of the sources only once.
@@
 In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
-Then, the heuristic function $h_t(x)$ is implemented by call the ComputeAndMemoizeDist for node $x$.
+The heuristic function $h_t(x)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for node $x$.
 
@@
 
-Bidir A* not as straightforward which one might expect.
-As pointed out in~\cite{gh-cspas-05}, one can not simply use a forward potential $h_t$ for the forward search and a backward potential $h_s$ for the reverse search.
-Since these searches would operate on different reduced graphs, it is unclear when the searches can terminate.
-To the best of our knowledge, no better stopping criterion exists than running both searches until the minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
-That is the same stopping crit as for a unidir search..
-symmetric
-Thus, the symmetric bidirectional algorithm performs double the work than a simple unidir search.
-can be slightly improved through pruning $\overrightarrow{D}[x] + h_t(x) > \mu$.
-Still not competitive
-
-average potential - reduced graph with average of reduced edge weights
-but average pot in each dir worse than orig pot for dir
-also overhead
-
-we propose an improved pruning crit, one which takes the search progress into account (through the min queue keys), which makes the simpler symmetric approach competitive or even faster than the average pot
-If $x$ was not yet settled by reverse search with minimum queue key $\overleftarrow{k}$ then
-$\operatorname{dist}(x,t) \geq \overleftarrow{k} - h_s(x)$.
-Thus, if $\overrightarrow{D}[x] + \overleftarrow{k} - h_s(x) > \mu$ we can discard $x$.
-
-% direction switch strats!
+Since bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm, a bidirectional variant of A* also seems desirable.
+However, as pointed out in~\cite{gh-cspas-05}, this is not as straightforward as one might expect.
+One would like to use a forward potential $\overrightarrow{h}_t$ in the forward search and a backward potential $\overleftarrow{h}_s$ (computing distances in the reversed graph) in the reverse search.
+The problem is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
+Thus, in the equivalent bidirectional Dijkstra search each direction would use different graph.
+This would break the bidirectional Dijkstra stopping criterion.
+To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until their minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
+But this is the same stopping criterion that a unidirectional A* search has.
+Thus, this straightforward bidirectional A* performs about twice the work than a unidirectional A*.
+This can be slightly improved by introducing pruning.
+Nodes, for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e.
+their distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, can not contribute to a shorter path and need not to be added to the queue.
+This approach is denoted as the \emph{symmetric} bidirectional A*.
+
+To overcome this problem the authors of propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
+The reduced graph of the average potential is the graph with the average weights of the reduced graphs of the forward and backward potentials.
+The potential function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$ (and $-h_{\overline{st}}(x)$ for the backward search).
+This potential allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
+Additionally, the pruning from the symmetric approach can be utilized with the original potentials.
+To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
+Still, it has two downsides.
+First, evaluating the average potential requires evaluating the original heuristics for both directions.
+This can cause some overhead depending on the heuristics.
+Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
+
+We propose an improved pruning criterion for the symmetric approach.
+It takes the search progress (in form of the growing minimal queue keys) into account to tighten the pruning with increasing search progress and compensates for the worse stopping criterion.
+As our experiments show, this makes the simpler symmetric approach at least competitive and sometimes even faster than the average potential.
+Recall that the queue is ordered by $D[x] + h(x)$.
+We use the minimal queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$.
+If a node $x$ was not yet settled by backward search then $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
+Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
+The pruning for the backwards search works analogously.
 
@@
 Our algorithm uses the undirected degree $d(x)$ of a node $x$.
-Formally, $d(x)$ is the number of nodes $y$ such that $(x,y)\in E$ or $(y,x)\in E$.
+Formally, $d(x)$ is the number of nodes $y$ such that $(xy)\in E$ or $(yx)\in E$.
 
@@
 Our algorithm differs from A* when removing a node $x$ from the queue.
-A* iterates over the outgoing arcs $(x,y)$ of $x$ and tries to reduce $D[y]$ by relaxing $(x,y)$.
+A* iterates over the outgoing arcs $(xy)$ of $x$ and tries to reduce $D[y]$ by relaxing $(xy)$.
 If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
 Our algorithm, however, behaves differently, if $d(y)\le 2$.
-Our algorithm determines the longest degree two chain of nodes $x,y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
+Our algorithm determines the longest degree two chain of nodes $x y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
 If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
@@
 We can also skip some degree three nodes.
-Denote by $x,y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
+Denote by $x y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
 If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
-Otherwise, there exist up to two degree chains $z,a_1,\ldots,a_p,b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
+Otherwise, there exist up to two degree chains $z a_1,\ldots,a_p b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
 Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
@@
 We start by establishing a common formal framework for the use of CH-Potentials.
-We describe some extended routing problems and how to apply CH-Potentials to them.
-Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
+Then, we exemplary describe some extended routing problems and how to apply CH-Potentials to them.
 
@@
 
-In this paper, we consider different applications, with slightly different problem models.
+We consider a variety different applications, with slightly different problem models.
 The goal is always to quickly answer many shortest path queries.
@@
 $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
-We require that $w_q(u,v)$ is greater or equal to the shortest distance $\operatorname{dist}_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
+We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
 The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
@@
 The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
+Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
 The exact implementation of this A* search depends on the application.
@@
 
-Our heuristic is always \emph{feasible}~\cite{hnr-afbhd-68}, i.e. $w_q(u,v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
+Our heuristic is always \emph{feasible}, i.e. $w_q(u v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
 By requirement and because of the triangle inequality the following holds:
 \[
-w_q(u,v) - h_t(u) + h_t(v) \geq \operatorname{dist}_\ell(\phi(u), \phi(v)) - \operatorname{dist}_\ell(\phi(u), \phi(t)) + \operatorname{dist}_\ell(\phi(v), \phi(t)) \geq 0
+w_q(u v) - h_t(u) + h_t(v) \geq \dist_\ell(\phi(u), \phi(v)) - \dist_\ell(\phi(u), \phi(t)) + \dist_\ell(\phi(v), \phi(t)) \geq 0
 \]
@@
 A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
-w_\ell(v_1, v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i,v_{i+1})
+w_\ell(v_1 v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i v_{i+1})
 \]
@@
 Edges in the input graph $G_\ell$ correspond to \emph{expanded nodes} in $G_q$.
-For every pair of incident edges $(x,y)$ and $(y,z)$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_\ell(y,z)$.
+For every pair of incident edges $x y$ and $y z$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_\ell(y z)$.
 A sequence of expanded nodes in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
@@
 In this scenario, preprocessing and query use different graphs $G_\ell$ and $G_q$.
-We define the node mapping function $\phi$ as $\phi(x,y) = y$.
-Obviously, $w_q(xy, yz) = w_t(x,y,z) + w_\ell(y,z) \geq \operatorname{dist}_\ell(\phi(x,y), \phi(y,z))$ and this approach yields a feasible heuristic.
+We define the node mapping function $\phi$ as $\phi(x y) = y$.
+Obviously, $w_q(xy, yz) = w_t(x,y,z) + w_\ell(y z) \geq \dist_\ell(\phi(x y), \phi(y z))$ and this approach yields a feasible heuristic.
 Sadly, the undirected graph underlying $G_q$ is always biconnected, if the input graph is strongly connected.
@@
 
-If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straight forward extension of Dijkstra's algorithm \cite{d-aassp-69}.
+If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
 The necessary modification to A* is analogous.
@@
 Our implementation stores edge travel times using piece-wise linear functions.
-The A* search uses the tentative distance $\tau$ at a node $x$ when to evaluating the travel time of outgoing edges $(x,y)$.
+The A* search uses the tentative distance $\tau$ at a node $x$ when to evaluating the travel time of outgoing edges $x y$.
 This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
@@
 \subsubsection{CCH Potentials}
-% TODO CCH-Pot
 
@@
 CRP~\cite{dgpw-crprn-13} and CCH~\cite{dsw-cch-15} follow this setup.
-Luckily, a CCH is just a CH with some additional properties.
-The CH in CH-Potentials can be replaced by a CCH without further modifications.
-Thus, CCH-Potentials could also support a three-phase setup.
+As discussed in Section~\ref{sec:lazy-rphast}, the Lazy RPHAST algorithm and thus CH-Potentials can also be used with a CCH.
+In terms of our framework this would mean having a second preprocessing phase which allows updating $w_\ell$ to the current traffic situation.
+Thus, with CCH-Potentials, we could also support a three-phase setup.
+This could, for example, be used to combine live traffic with turn costs and restrictions.
+One could also apply live traffic updates to $w_q$ immediately and only update $w_\ell$ once the difference between $w_\ell$ and $w_q$ grows too big or when some edge weight update would set $w_q < w_\ell$.
 % For applications which allow for a three-phase setup, CCH-Potentials are likely a better choice for supporting live traffic.
-However, evaluating CCH-Potentials is beyond the scope of this paper.
-We focus on evaluating CH-Potentials as a simple building block in the two-phase setup.
+% However, evaluating CCH-Potentials is beyond the scope of this paper.
+% We focus on evaluating CH-Potentials as a simple building block in the two-phase setup.
+
+\subsubsection{Alternative Routes with the Penalty Method}
+
+Providing users with multiple alternative good routes to choose from is another relevant feature for practical routing applications.
+The penalty method is an established framework for computing such alternative routes~\cite{bdgs-argrn-11,krs-eepma-13,pz-iarp-13,kobitzsch2015alternative}.
+It computes an alternative graph instead of a set of alternative routes.
+This graph is constructed iteratively.
+In each iteration, a shortest path between $s$ and $t$ is computed.
+This path (or certain parts of it) is then added to the alternative graph depending on certain quality criteria.
+Finally, all edge weights of the path and possibly adjoined edges are penalized.
+This is repeated until the current path is more than $1 + \epsilon$ times longer than the original distance or until certain other stopping criteria are fulfilled.
+
+The running time of the penalty method is usually dominated by shortest path computations.
+Since this boils down to computing shortest paths on $G_{\ell}$ with a few increased edge weights this seems like a natural fit for A* with CH-Potentials.
+While the schema outlined above is conceptually simple, it leaves open quite a few details which may have a severe impact on performance.
+In this work, we do not develop our own variant of the penalty method but aim to reproduce the configuration reported by Kobitzsch in his dissertation~\cite{kobitzsch2015alternative}.
+To the best of our knowledge, this is the latest, most thoroughly engineered and evaluated iteration of the penalty method which also achieves the qualitatively best results.
+While some details remain unclear from the description in~\cite{kobitzsch2015alternative}, luckily we could obtain the original implementation from the authors to fill in the remaining details.
+Since we reproduce this exact configuration except for the shortest path algorithm, qualitative results should carry over and we can focus only on running times in our work.
+
+For sake of keeping this section self-contained, we now briefly reiterate the details of our configuration.
+The stretch limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \dist(u,v)$.
+Shortest path edges are penalized multiplicative with a factor of $\psi = 1.25$\footnote{Kobitzsch reports using $\psi = 1.1$ but the code uses $\psi=1.25$ by default.
+As our qualitative results match Kobitzsch's results when we use $\psi = 1.25$ but not with $\psi = 1.1$, we choose $\psi = 1.25$.}, i.e.
+when an edges was $k$ times on a shortest path its weight will be $w_q = w_{\ell}\cdot\psi^k$.
+We also penalize edges incident to shortest paths with an asymmetric additive rejoin penalty.
+An edge $uv$ where $u$ is on a shortest path but $v$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(s,u)}{\dist_{w_q}(s,t)}$.
+An edge $uv$ where $v$ is on a shortest path but $u$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(v,t)}{\dist_{w_q}(s,t)}$.
+$\psi_r$ is set to 0.01.
+To avoid over-penalization, we count the times a \emph{node} was on a shortest path and do not apply penalties, when this number exceeds $k_{\max} := \lceil\log_{\psi}(1+\epsilon)\rceil + 2$.
+Shortest paths are split into segments not yet contained in the alternative graph.
+A segment is only added if it is long enough (at least $0.1 \cdot \dist_{w_{\ell}}(s,t)$) and does not violate the maximum stretch condition.
+We terminate the algorithm when any of the following conditions becomes true:
+\begin{itemize}
+	\item All nodes on the current shortest path have been penalized $k_{\max}$ times.
+	\item Segments from 10 different iterations have been added to the alternative graph.
+	\item The shortest path is longer than $(1+\epsilon) \cdot \dist_{w_{\ell}}(s,t)$ with respect to the original metric.
+	\item The shortest path is longer than $(1+\epsilon) \cdot \psi + 2\psi_r \cdot \dist_{w_{\ell}}(s,t)$ with respect to the penalized metric.
+\end{itemize}
+The last two stopping criteria can both be utilized for additional pruning in the A*.
+As a final optimization, we run the forward and backward search in parallel in two threads.
 
@@
 Further, a trade-off between arrival time and route quality is considered.
-
-\subsubsection{Alternative Routes with the Penalty Method}
-
-Important feature alternative routes
-One popular framework is the penalty method
-It computes an alternative graph instead of a set of routes
-the graph is constructed iteratively
-in each iteration, first, a shortest path between s and t is computed
-if the path or certain parts of it meet certain quality criteria add to alternative graph
-finally penalize all edge weights of current path and possibly adjoined edges and continue until a certain stopping criterion is fulfilled
-
-running time dominated by shortest path computations
-since edge weights only increased natural fit for A* with CH-Pot
-lots of details to implement and define
-we take config of kobitzsch diss
-quality eval over there
-here only performance
-
-for sake of reproducibility we briefly discuss the details of our configuration
-code from authors
-eps/psi = 1.25
-penalization schema
-	exp psi
-	asymmetric additive
-stopping crit
-	saturation
-	max it
-	mas new parts
-
-feasibility
-
-perf
-	pruning by orig
-	pruning by penalized
-	parallelization
 

### 2021-08-20 Rakefile.rb
update preprocessing experiments

@@ -173,7 +173,7 @@ namespace "exp" do
   directory "#{exp_dir}/bidir_features"
   directory "#{exp_dir}/applications"
 
-  task preprocessing: ["code/compute_ch/build/compute_ch", "#{exp_dir}/preprocessing"] + graphs.map { |g| g + 'lower_bound' } do
+  task preprocessing: ["code/compute_ch/build/compute_ch", "#{exp_dir}/preprocessing", "code/rust_road_router/lib/InertialFlowCutter/build/console"] + graphs.map { |g| g + 'lower_bound' } do
     graphs.each do |graph|
       10.times do
         filename = "#{exp_dir}/preprocessing/" + `date --iso-8601=seconds`.strip + '.out'
@@ -183,6 +183,13 @@ namespace "exp" do
           "/dev/null /dev/null /dev/null " +
           "/dev/null /dev/null /dev/null " +
           ">> #{filename}")
+
+        Dir.chdir "code/rust_road_router" do
+          filename = "#{exp_dir}/preprocessing/" + `date --iso-8601=seconds`.strip + '.out'
+          sh "./flow_cutter_cch_order.sh #{graph} #{Etc.nprocessors} >> #{filename}"
+          filename = "#{exp_dir}/preprocessing/" + `date --iso-8601=seconds`.strip + '.json'
+          sh "cargo run --release --bin cch_preprocessing -- #{graph} >> #{filename}"
+        end
       end
     end
   end

### 2021-08-27 Rakefile.rb
update engine and experiments

@@ -257,6 +257,8 @@ namespace "exp" do
       sh "cargo run --release --bin chpot_blocked -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin bidir_chpot_blocked -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin bidir_chpot_turns -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
+      sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
+      sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{dimacs_graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
     end
   end
 end

### 2021-08-27 eval/bidir_features.py
start with bidir tables

@@ -1,0 +1,92 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import pandas as pd
+pd.options.display.float_format = '{:.2f}'.format
+
+import json
+import glob
+import os
+import re
+
+paths = glob.glob("exp/bidir_features/*.json")
+data = [json.load(open(path)) for path in paths]
+
+queries = pd.DataFrame.from_records([{
+    **algo,
+    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
+    'experiment': exp['experiment'],
+    'factor': exp['factor'],
+    'probability': exp.get('probability', 1),
+    'speed': exp.get('speed', 0),
+    'potential': run['potential'],
+    'num_queue_pops': algo['num_queue_pops'] + algo.get('core_search', {}).get('num_queue_pops', 0),
+    'num_queue_pushs': algo['num_queue_pushs'] + algo.get('core_search', {}).get('num_queue_pushs', 0),
+    'num_relaxed_arcs': algo['num_relaxed_arcs'] + algo.get('core_search', {}).get('num_relaxed_arcs', 0),
+    }
+    for run in data for exp in run['experiments'] for algo in exp['algo_runs']])
+
+queries['choose_direction_strategy'] = queries['choose_direction_strategy'].fillna('unidir')
+queries['bidir_pot'] = queries['bidir_pot'].fillna('unidir')
+queries['improved_pruning'] = queries['improved_pruning'].fillna('unidir')
+queries['num_queue_pushs_k'] = queries['num_queue_pushs'] / 1000
+
+queries = queries.loc[lambda x: x['graph'] == 'OSM Germany']
+
+pruning = queries.loc[lambda x: x['choose_direction_strategy'] != 'min_key'] \
+    .loc[lambda x: x['bidir_pot'] != 'unidir'] \
+    .groupby(['potential', 'experiment', 'factor', 'probability', 'algo', 'bidir_pot', 'improved_pruning'])[['running_time_ms', 'num_queue_pushs_k']].mean().unstack(0) \
+    .loc[('weight_scale', 1.05, 1.00)] \
+    .rename(columns={
[…]

### 2021-08-27 paper/ch_potentials.tex
start with bidir tables
update text with actual penalty config

@@
 The stretch limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \dist(u,v)$.
-Shortest path edges are penalized multiplicative with a factor of $\psi = 1.25$\footnote{Kobitzsch reports using $\psi = 1.1$ but the code uses $\psi=1.25$ by default.
-As our qualitative results match Kobitzsch's results when we use $\psi = 1.25$ but not with $\psi = 1.1$, we choose $\psi = 1.25$.}, i.e.
-when an edges was $k$ times on a shortest path its weight will be $w_q = w_{\ell}\cdot\psi^k$.
+Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was $k$ times on a shortest path its weight will be $w_q = w_{\ell}\cdot\psi^k$.
 We also penalize edges incident to shortest paths with an asymmetric additive rejoin penalty.
@@
 Shortest paths are split into segments not yet contained in the alternative graph.
-A segment is only added if it is long enough (at least $0.1 \cdot \dist_{w_{\ell}}(s,t)$) and does not violate the maximum stretch condition.
+A segment is only added if it is long enough (at least $0.05 \cdot \dist_{w_{\ell}}(s,t)$) and does not violate the maximum stretch condition.
 We terminate the algorithm when any of the following conditions becomes true:
@@
 	\item All nodes on the current shortest path have been penalized $k_{\max}$ times.
-	\item Segments from 10 different iterations have been added to the alternative graph.
+	\item 10 ore more segments have been added to the alternative graph.
 	\item The shortest path is longer than $(1+\epsilon) \cdot \dist_{w_{\ell}}(s,t)$ with respect to the original metric.
@@
 This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
+
+\begin{table}
+\centering
+\caption{
+TODO
+}\label{tab:bidir_pruning}
+\input{table/bidir_pruning.tex}
+\end{table}
 

### 2021-09-28 paper/ch_potentials.tex
math formatting
switch to acm template for jea

@@
-\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
+% \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
+\documentclass[manuscript,review]{acmart}
+
+% \usepackage{amssymb}
+% \usepackage{amsmath}
 % checkmarks
-\usepackage{pifont}
+\usepackage{pifont} % TODO
 \newcommand{\cmark}{\ding{51}}
@@
 \usepackage{todonotes}
-\usepackage{booktabs}
+% \usepackage{booktabs}
 \usepackage{multirow}
@@
 
-\nolinenumbers
-\hideLIPIcs
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{acmcopyright}
+\copyrightyear{2018}
+\acmYear{2018}
+\acmDOI{10.1145/1122445.1122456}
+
+\begin{document}
 
@@
 \title{A Fast and Tight Heuristic for A* in Road Networks}
-\author{Ben Strasser}{Germany}{academia@ben-strasser.net}{}{}
-\author{Tim Zeitz}{Karlsruhe Institute of Technology, Germany}{tim.zeitz@kit.edu}{https://orcid.org/0000-0003-4746-3582}{}
-
-\authorrunning{B. Strasser and T. Zeitz}
-\Copyright{Ben Strasser and Tim Zeitz}
-
-\ccsdesc[500]{Theory of computation~Shortest paths}
-\ccsdesc[300]{Mathematics of computing~Graph algorithms}
-\ccsdesc[500]{Applied computing~Transportation}
-
-\keywords{route planning, shortest paths, realistic road networks}
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\EventEditors{David Coudert and Emanuele Natale}
-\EventNoEds{2}
-\EventLongTitle{19th International Symposium on Experimental Algorithms (SEA 2021)}
-\EventShortTitle{SEA 2021}
-\EventAcronym{SEA}
-\EventYear{2021}
-\EventDate{June 7--9, 2021}
-\EventLocation{Nice, France}
-\EventLogo{}
-\SeriesVolume{190}
-\ArticleNo{6}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
-
-\maketitle
+
+\author{Ben Strasser}
+\email{academia@ben-strasser.net}
+\affiliation{}
+
+\author{Tim Zeitz}
+\email{tim.zeitz@kit.edu}
+\orcid{0000-0003-4746-3582}
+\affiliation{%
+  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
+  \city{Karlsruhe}
+  \country{Germany}
+}
+
+% \author{Ben Strasser}{Germany}{academia@ben-strasser.net}{}{}
+% \author{Tim Zeitz}{Karlsruhe Institute of Technology, Germany}{tim.zeitz@kit.edu}{https://orcid.org/0000-0003-4746-3582}{}
+
+% \authorrunning{B. Strasser and T. Zeitz}
+\renewcommand{\shortauthors}{B. Strasser and T. Zeitz}
+% \Copyright{Ben Strasser and Tim Zeitz}
 
@@
 \end{abstract}
+
+\ccsdesc[500]{Theory of computation~Shortest paths}
+\ccsdesc[300]{Mathematics of computing~Graph algorithms}
+\ccsdesc[500]{Applied computing~Transportation}
+
+\keywords{route planning, shortest paths, realistic road networks}
+
+\maketitle
 
@@
 For sake of keeping this section self-contained, we now briefly reiterate the details of our configuration.
-The stretch limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \dist(u,v)$.
+The stretch limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \cdot \dist(u,v)$.
 Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was $k$ times on a shortest path its weight will be $w_q = w_{\ell}\cdot\psi^k$.

### 2021-09-28 Rakefile.rb
update lazy rphast figures

@@ -7,6 +7,7 @@ file "paper/ch_potentials.pdf" => [
   "paper/ch_potentials.tex",
   "paper/fig/scaled_weights.pdf",
   "paper/fig/lazy_rphast_many_to_one.pdf",
+  "paper/fig/lazy_rphast_cch_many_to_one.pdf",
   "paper/fig/lazy_rphast_many_to_many.pdf",
   "paper/table/applications.tex",
   "paper/table/graphs.tex",
@@ -25,6 +26,10 @@ namespace "fig" do
   end
 
   file "paper/fig/lazy_rphast_many_to_one.pdf" => FileList["#{exp_dir}/rphast/*.json"] + ["eval/rphast_comparison.py"] do
+    sh "eval/rphast_comparison.py"
+  end
+
+  file "paper/fig/lazy_rphast_cch_many_to_one.pdf" => FileList["#{exp_dir}/rphast/*.json"] + ["eval/rphast_comparison.py"] do
     sh "eval/rphast_comparison.py"
   end
 

### 2021-09-28 eval/rphast_comparison.py

@@ -32,10 +32,22 @@ g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_one
 g.set_xlabel('Ball Size $|B|$')
 g.set_ylabel('Running Time [ms]')
 g.set_yscale('log')
+g.yaxis.set_major_formatter(mpl.ticker.LogFormatter())
+plt.grid(True, which="minor", linewidth=0.6)
 g.legend().set_title('')
 plt.tight_layout()
 g.get_figure().savefig('paper/fig/lazy_rphast_many_to_one.pdf')
 
+plt.figure(figsize=(11,4))
+g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_cch_many_to_one'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_ms', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
+g.set_xlabel('Ball Size $|B|$')
+g.set_ylabel('Running Time [ms]')
+g.set_yscale('log')
+g.yaxis.set_major_formatter(mpl.ticker.LogFormatter())
+plt.grid(True, which="minor", linewidth=0.6)
+g.legend().set_title('')
+plt.tight_layout()
+g.get_figure().savefig('paper/fig/lazy_rphast_cch_many_to_one.pdf')
 
 plt.figure(figsize=(11,4))
 g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_many'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_s', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))

### 2021-09-28 Rakefile.rb
update rakefile to use new chpot features bin for building block exp

@@ -210,11 +210,11 @@ namespace "exp" do
   task building_blocks: ["#{osm_ger}lower_bound_ch", "#{osm_ger}cch_perm", "#{exp_dir}/building_blocks"] do
     Dir.chdir "code/rust_road_router" do
       [[], ['chpot-alt'], ['chpot-cch'], ['chpot-oracle'], ['chpot-only-topo']].each do |pot_feats|
-        sh "cargo run --release --bin chpot_simple_scale_dijkstra #{features(pot_feats)} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale #{features(pot_feats + ['chpot-no-bcc', 'chpot-no-deg2', 'chpot-no-deg3'])} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale #{features(pot_feats + ['chpot-no-deg2', 'chpot-no-deg3'])} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale #{features(pot_feats + ['chpot-no-deg3'])} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
-        sh "NUM_DIJKSTRA_QUERIES=0 cargo run --release --bin chpot_simple_scale #{features(pot_feats)} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
+        sh "cargo run --release --bin chpot_features #{features(pot_feats + ['chpot-no-reorder'])} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
+        sh "cargo run --release --bin chpot_features #{features(pot_feats + ['chpot-no-bcc', 'chpot-no-deg2', 'chpot-no-deg3'])} -- #{osm_ger} > #{exp_dir}/building_blocks/$(date --iso-8601=seconds).json"
[…]

### 2021-09-30 paper/ch_potentials.tex
add bidir tables and scripting around them

@@
 % \usepackage{amsmath}
-% checkmarks
-\usepackage{pifont} % TODO
-\newcommand{\cmark}{\ding{51}}
-\newcommand{\xmark}{\ding{55}}
+\newcommand{\cmark}{$\bullet$}
+\newcommand{\xmark}{$\circ$}
 
@@
 
-\subparagraph{Inputs and Methodology}
+\paragraph{Inputs and Methodology}
 Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
@@
 
-\subparagraph{Experiments}
+\paragraph{Experiments}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/lazy_rphast_many_to_one.pdf}
+\caption{
+Many-to-One CH
+}\label{fig:many_to_one}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/lazy_rphast_cch_many_to_one.pdf}
+\caption{
+Many-to-One CCH
+}\label{fig:many_to_one_cch}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/lazy_rphast_many_to_many.pdf}
+\caption{
+Many-to-Many
+}\label{fig:many_to_many}
+\end{figure}
 
@@
 TODO
+OSM Ger, Weight Scaled uniformly by 1.05, alternating dirs
 }\label{tab:bidir_pruning}
 \input{table/bidir_pruning.tex}
+\end{table}
+
+\begin{table}
+\centering
+\caption{
+TODO
+OSM Ger, Weight Scaled 1.5, sym with improved pruning, avg without, with topo opt
+}\label{tab:bidir_switching}
+\input{table/bidir_switching.tex}
+\end{table}
+
+\begin{table}
+\centering
+\caption{
+TODO
+OSM Ger
+}\label{tab:bidir}
+\input{table/bidir.tex}
 \end{table}

### 2021-09-30 Rakefile.rb

@@ -10,6 +10,9 @@ file "paper/ch_potentials.pdf" => [
   "paper/fig/lazy_rphast_cch_many_to_one.pdf",
   "paper/fig/lazy_rphast_many_to_many.pdf",
   "paper/table/applications.tex",
+  "paper/table/bidir_pruning.tex",
+  "paper/table/bidir_switching.tex",
+  "paper/table/bidir.tex",
   "paper/table/graphs.tex",
   "paper/table/building_blocks.tex"] do
 
@@ -55,6 +58,18 @@ namespace "table" do
     ] + ["eval/graphs.py", "paper/table"] do
 
     sh "eval/graphs.py"
+  end
+
+  file "paper/table/bidir.tex" => FileList["#{exp_dir}/bidir_features/*.out"] + ["eval/bidir_features.py", "paper/table"] do
+    sh "eval/bidir_features.py"
+  end
+
+  file "paper/table/bidir_pruning.tex" => FileList["#{exp_dir}/bidir_features/*.out"] + ["eval/bidir_features.py", "paper/table"] do
+    sh "eval/bidir_features.py"
+  end
+
+  file "paper/table/bidir_switching.tex" => FileList["#{exp_dir}/bidir_features/*.out"] + ["eval/bidir_features.py", "paper/table"] do
+    sh "eval/bidir_features.py"
   end
 
   file "paper/table/building_blocks.tex" => FileList["#{exp_dir}/building_blocks/*.json"] + ["eval/building_blocks.py", "paper/table"] do

### 2021-09-30 eval/bidir_features.py

@@ -35,58 +35,120 @@ queries = queries.loc[lambda x: x['graph'] == 'OSM Germany']
 
 pruning = queries.loc[lambda x: x['choose_direction_strategy'] != 'min_key'] \
     .loc[lambda x: x['bidir_pot'] != 'unidir'] \
-    .groupby(['potential', 'experiment', 'factor', 'probability', 'algo', 'bidir_pot', 'improved_pruning'])[['running_time_ms', 'num_queue_pushs_k']].mean().unstack(0) \
+    .groupby(['potential', 'experiment', 'factor', 'probability', 'algo', 'bidir_pot', 'improved_pruning'])[['running_time_ms', 'num_queue_pushs_k']].mean() \
+    .unstack(0) \
     .loc[('weight_scale', 1.05, 1.00)] \
-    .rename(columns={
+    .reindex(columns=['Zero', 'ALT', 'CH', 'CCH', 'Oracle'], level=1)
+
+pruning[('num_queue_pushs_k', '(C)CH/Oracle')] = pruning[('num_queue_pushs_k', 'CH')]
+pruning = pruning.drop([('num_queue_pushs_k', 'CH'), ('num_queue_pushs_k', 'CCH'), ('num_queue_pushs_k', 'Oracle')], axis = 'columns')
+
+pruning = pruning.rename(columns={
         'running_time_ms': 'Running time [ms]',
         'num_queue_pushs_k': 'Queue pushs [$\\cdot 10^3$]',
-        'CH': 'CH-Pot.',
-        'CCH': 'CCH-Pot.',
     }) \
-    .rename(index={
-        'Average': 'Avg.',
-        'Symmetric': 'Sym.',
-    })
+    .reset_index()
 
-lines = pruning.to_latex(escape=False).split("\n")
+lines = pruning.to_latex(escape=False, index=False, multicolumn_format='c', column_format='c@{\hskip6pt}c@{\hskip3pt}crrrrrrrr').split("\n")
 
-lines = [R'\begin{tabular}{lllrrrrrrrr}'] + [lines[1]] + [
-    ' & '.join(["Low Deg.", "Bidir.", "Impr."] + lines[2].split(' & ')[3:]) + R" \cmidrule(lr){4-7} \cmidrule(lr){8-11}",
-    ' & '.join(["Opt.",     "Pot.",   "Pruning"] + lines[3].split(' & ')[3:])
[…]

### 2021-10-04 paper/table/graphs.tex
update graph table for journal

@@
-\begin{tabular}{lrrr}
+\begin{tabular}{lrrrrr}
 \toprule
- &          Nodes &          Edges & Preprocessing \\ & $[\cdot 10^6]$ & $[\cdot 10^6]$ &           [s] \\
+ &                &                & \multicolumn{3}{c}{Preprocessing [s]} \\ \cmidrule(lr){4-6} & Nodes          & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} \\ \cmidrule(lr){5-6} & $[\cdot 10^6]$ & $[\cdot 10^6]$ &                     & Phase 1 & Phase 2 \\
 \midrule
-OSM Ger &       16.2 &       35.4 &                         295.2 \\
-TDEur17 &       25.8 &       55.5 &                         292.6 \\
-TDGer06 &        4.7 &       10.8 &                          58.9 \\
+OSM Ger       &       16.2 &       35.4 &                          312.8 &        1\,466.7 &          10.1 \\
+DIMACs Europe &       18.0 &       42.2 &                          276.2 &        2\,480.9 &          12.4 \\
+TDGer06       &        4.7 &       10.8 &                           59.2 &         331.7 &           2.7 \\
+TDEur17       &       25.8 &       55.5 &                          293.9 &        2\,102.3 &          14.1 \\
+TDEur20       &       28.5 &       60.9 &                          311.9 &        2\,219.5 &          15.2 \\
 \bottomrule

### 2021-10-04 eval/graphs.py

@@ -8,7 +8,6 @@ import os
 import re
 
 base = "exp/"
-cwd = os.getcwd()
 paths = glob.glob(base + "preprocessing/*.json")
 data = [json.load(open(path)) for path in paths]
 
@@ -24,27 +23,27 @@ def path_to_graph(path):
     }[[x for x in path.split('/') if x != ''][-1]]
 
 graphs = pd.DataFrame.from_records([{
-    **run
+    **run,
     'graph': path_to_graph(run['args'][1]),
-    'num_nodes': run['graph'],
-    'num_edges': run['graph'],
+    'num_nodes': run['graph']['num_nodes'],
+    'num_edges': run['graph']['num_arcs'],
 } for run in data if 'graph' in run])
 
 runtime_pattern = re.compile(".*Needed (\\d+)musec\\..*")
 
 def parse_contraction_output(path):
-    stats = { 'ch_contraction_running_time_s': 0.0 }
+  stats = { 'ch_contraction_running_time_s': 0.0 }
 
-    with open(path, 'r') as f:
-        for line in f:
-            if not 'graph' in stats:
-                stats['graph'] = path_to_graph(line.strip())
-            else:
-                match = runtime_pattern.match(line)
-                if match:
-                    stats['ch_contraction_running_time_s'] += int(match[1]) / 1000000
+  with open(path, 'r') as f:
+    for line in f:
+      if not 'graph' in stats:
+        stats['graph'] = path_to_graph(line.strip())
+      else:
+        match = runtime_pattern.match(line)
+        if match:
+          stats['ch_contraction_running_time_s'] += int(match[1]) / 1000000
 
-    return stats
+  return stats
 
 ch_preprocessing = pd.DataFrame.from_records([parse_contraction_output(path) for path in glob.glob(base + "preprocessing/ch/*.out")])
 
@@ -64,22 +63,27 @@ def parse_flowcutter_partition_output(path):
 
   return stats
 
-cch_ordering = pd.DataFrame.from_records([parse_contraction_output(path) for path in glob.glob(base + "preprocessing/cch/*.out")])
+cch_ordering = pd.DataFrame.from_records([parse_flowcutter_partition_output(path) for path in glob.glob(base + "preprocessing/cch/*.out")])
 
-table = graphs.groupby(['graph'])[['num_nodes', 'num_edges']].mean()
[…]

### 2021-10-04 eval/building_blocks.py
update building blocks table

@@ -12,34 +12,22 @@ paths = glob.glob(base + "building_blocks/*.json")
 data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
-    'features': run['feature_flags'],
-    **algo }
-    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') in ['CH Potentials Query', 'Dijkstra Query']])
+    **algo,
+    'BCC': run['bcc_core'],
+    'Skip Deg. 2': run['skip_deg2'],
+    'Skip Deg. 3': run['skip_deg3'],
+    'Heuristic': run['potential'],
+    'num_queue_pops': algo['num_queue_pops'] + algo.get('core_search', {}).get('num_queue_pops', 0),
+    'num_queue_pushs': algo['num_queue_pushs'] + algo.get('core_search', {}).get('num_queue_pushs', 0),
+    'num_relaxed_arcs': algo['num_relaxed_arcs'] + algo.get('core_search', {}).get('num_relaxed_arcs', 0),
+} for run in data for algo in run['algo_runs'] if algo.get('algo') in ['Virtual Topocore Component Query', 'Dijkstra Query']])
 
-queries['BCC'] = (queries['algo'] == 'CH Potentials Query') & ~queries['features'].str.contains('CHPOT_NO_BCC')
-queries['Skip Deg. 2'] = (queries['algo'] == 'CH Potentials Query') & ~queries['features'].str.contains('CHPOT_NO_DEG2')
-queries['Skip Deg. 3'] = (queries['algo'] == 'CH Potentials Query') & ~queries['features'].str.contains('CHPOT_NO_DEG3')
-
-def heuristic_name(feature_flags):
-    if 'CHPOT_ONLY_TOPO' in feature_flags:
-        return 'Zero'
-    elif 'CHPOT_ORACLE' in feature_flags:
-        return 'Oracle'
-    elif 'CHPOT_ALT' in feature_flags:
-        return 'ALT'
-    else:
-        return 'CH'
-
-queries['Heuristic'] = queries['features'].apply(heuristic_name)
-
[…]

### 2021-10-04 Rakefile.rb
update engine, temporarily disable penalty app exp

@@ -280,8 +280,8 @@ namespace "exp" do
       sh "cargo run --release --bin chpot_blocked -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin bidir_chpot_blocked -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin bidir_chpot_turns -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
-      sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
-      sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{dimacs_graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
+      # sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
+      # sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{dimacs_graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
     end
   end
 end

### 2021-10-12 paper/ch_potentials.tex
lots of rough writing

@@
 
+\subsection{Low Degree A* Improvements}\label{sec:low-deg-improvment}
+
+Preliminary experiments showed, that a significant amount of query running time is spent in heuristic evaluations and queue operations.
+We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
+Avoiding pushing low degree nodes into the queue is the focus of this section.
+The techniques discussed here are a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
+
+We modify A* by processing low degree nodes consecutively without pushing them into the queue.
+Our algorithm uses the undirected degree $d(x)$ of a node $x$.
+Formally, $d(x)$ is the number of nodes $y$ such that $(xy)\in E$ or $(yx)\in E$.
+
+Analogous to A*, our algorithm stores for every node $x$ a tentative distance $D[x]$.
+Additionally, it maintains a minimum priority queue.
+Diverging from A*, not all nodes can be pushed but every node has a tentative distance.
+
+\subsubsection{Skip Degree Two Nodes}
+
+Our algorithm differs from A* when removing a node $x$ from the queue.
+A* iterates over the outgoing arcs $(xy)$ of $x$ and tries to reduce $D[y]$ by relaxing $(xy)$.
+If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
+Our algorithm, however, behaves differently, if $d(y)\le 2$.
+Our algorithm determines the longest degree two chain of nodes $x y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
+If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
+Instead, it iteratively tries to reduce all $D[y_i]$.
+If it does not reach $z$, then only $D$ is modified but no queue action is performed.
+If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h_t(z)$.
+
+As the target node $t$ might have degree two, our algorithm cannot rely on stopping, when $t$ is removed from the queue.
+Instead, our algorithm stops as soon as $D[t]$ is less than the minimum weight in the queue.
+
+\subsubsection{Skip Degree Three Nodes}
+
+We can also skip some degree three nodes.
+Denote by $x y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
+If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
+Otherwise, there exist up to two degree chains $z a_1,\ldots,a_p b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
+Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
+If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h_t(\beta)$.
+Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h_t(b)$.
+If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
+
+\subsubsection{Stay in Largest Biconnected Component}\label{sec:largested-biconnected-component}
+
+A lot of nodes in road networks lead to dead-ends.
+Unless the source or target is in this dead-end, it is unnecessary to explore these nodes.
+
+In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}.
+$G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
+We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
+For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
+For nodes in the core, $a_v=v$.
+We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
+
+The query phase is divided into two steps.
+In the first step, we apply A* with CH-Potentials to $G_C$ combined with the component that contains $s$.
+This can be achieved implicitly by removing edges from $G_C$ into other components during preprocessing.
+If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
+Otherwise, we find $a_t$.
+In that case, we continue by searching a path from $a_t$ towards $t$ restricted to $t$'s biconnected component.
+The final result is the concatenation of both paths.
+
 \subsection{Bidirectional A*}
@@
 
-\subsection{Low Degree A* Improvements}\label{sec:low-deg-improvment}
-
-Preliminary experiments showed, that a significant amount of query running time is spent in heuristic evaluations and queue operations.
-We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
-Avoiding pushing low degree nodes into the queue is the focus of this section.
-The techniques discussed here are a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
-
-We modify A* by processing low degree nodes consecutively without pushing them into the queue.
-Our algorithm uses the undirected degree $d(x)$ of a node $x$.
-Formally, $d(x)$ is the number of nodes $y$ such that $(xy)\in E$ or $(yx)\in E$.
-
-Analogous to A*, our algorithm stores for every node $x$ a tentative distance $D[x]$.
-Additionally, it maintains a minimum priority queue.
-Diverging from A*, not all nodes can be pushed but every node has a tentative distance.
-
-\subsubsection{Skip Degree Two Nodes}
-
-Our algorithm differs from A* when removing a node $x$ from the queue.
-A* iterates over the outgoing arcs $(xy)$ of $x$ and tries to reduce $D[y]$ by relaxing $(xy)$.
-If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
-Our algorithm, however, behaves differently, if $d(y)\le 2$.
-Our algorithm determines the longest degree two chain of nodes $x y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
-If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
-Instead, it iteratively tries to reduce all $D[y_i]$.
-If it does not reach $z$, then only $D$ is modified but no queue action is performed.
-If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h_t(z)$.
-
-As the target node $t$ might have degree two, our algorithm cannot rely on stopping, when $t$ is removed from the queue.
-Instead, our algorithm stops as soon as $D[t]$ is less than the minimum weight in the queue.
-
-\subsubsection{Skip Degree Three Nodes}
-
-We can also skip some degree three nodes.
-Denote by $x y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
-If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
-Otherwise, there exist up to two degree chains $z a_1,\ldots,a_p b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
-Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
-If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h_t(\beta)$.
-Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h_t(b)$.
-If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
-
-\subsubsection{Stay in Largest Biconnected Component}\label{sec:largested-biconnected-component}
-
-A lot of nodes in road networks lead to dead-ends.
-Unless the source or target is in this dead-end, it is unnecessary to explore these nodes.
-
-In the preprocessing phase, we compute the subgraph $G_C$, called \emph{core}.
-$G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
-We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
-For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
-For nodes in the core, $a_v=v$.
-We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
-
-The query phase is divided into two steps.
-In the first step, we apply A* with CH-Potentials to $G_C$ combined with the component that contains $s$.
-This can be achieved implicitly by removing edges from $G_C$ into other components during preprocessing.
-If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
-Otherwise, we find $a_t$.
-In that case, we continue by searching a path from $a_t$ towards $t$ restricted to $t$'s biconnected component.
-The final result is the concatenation of both paths.
+% TODO alternating directions
 
@@
 Our submitted paper is the finished version of the ArXiv preprint.
-% der punkt macht fuer mich keinen Sinn.
-% Dass wir keine der Contributions von kswz-erptd-p-20 beschreiben ist ja offensichtlich, sieht ja jeder in dem Abschnitt.
-% Wichtiger finde ich, dass kswz-erptd-p-20 nix von uns beschreibt.
-% Das meinte ich zumindest mit dem Satz den du geloescht hast.
-% Aber der jetzt folgende Satz sagt nur, dass unser paper hier nix von kswz-erptd-p-20 beschreibt.
 \cite{kswz-erptd-p-20} does not describe any of the contributions of this paper.
@@
 \centering
-\caption{Instances used in the evaluation.}\label{tab:graphs}
+\caption{Instances used in the evaluation with preprocessing running times to construct a CH or CCH-Potential.
+With CCH-Potentials $w_{\ell}$ can be updated by rerunning Phase 2 again.}\label{tab:graphs}
 \input{table/graphs.tex}
@@
 
-\paragraph{Inputs and Methodology}
+\subsection{Inputs and Methodology}
 Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
@@
 We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
-Additionally, we have two graphs with proprietary traffic predictions provided by PTV\footnote{\url{https://ptvgroup.com}}.
+We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACs implementation challenge~\cite{DemetrescuGJ09}.
+Additionally, we have three graphs with proprietary traffic predictions also provided by PTV.
 The PTV instances are not OSM-based.
 One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
+% TODO PTV20
 Table~\ref{tab:graphs} contains an overview over our instances.
-In this table, we further include the sequential running time necessary to construct the CH.
+In this table, we further include the sequential preprocessing running time to construct the (C)CH.
 We report preprocessing running times as averages over 10 runs.
@@
 
-\paragraph{Experiments}
+\subsection{Lazy RPHAST}
+
+% Methodology
+% time per source
+% selection times
+
+% what do we want to show
+% - for one source target set we are as fast as RPHAST
+%   - dimensions: phase (style), graph (color), CH Var (row)
+% - incremental is cool
 
@@
 \end{figure}
+
+\subsubsection{k-Nearest-Neighbor Queries with CCH}
+
+One-to-many queries appear in many extended applications as a subproblem.
+In some of these cases Lazy RPHAST can be used to achieve significant speed-ups with surprisingly little effort.
+One such example are k-Nearest-Neighbor Queries with CCH~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
+There, shortest distances are computed repeatedly from the same source to boundary nodes of a cell of nodes.
+On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code~\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
+We reran their improved experiments and include the results in our appendix.
+Briefly summarized, integrating Lazy RPHAST into this algorithm decreased running times by up to 15 times depending on the distribution of the target set for almost no effort.
+
+\subsection{(C)CH-Potentials Heuristic}
 
@@
 
-The performance of A* depends on the tightness of the heuristic.
+The performance of A* depends on the tightness of the heuristic and the overhead of evaluating the heuristic.
 CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
@@
 
-Table~\ref{tab:building_blocks} depicts the performance of A* with different heuristics and optimizations.
-We compare CH-Potentials to three other heuristics.
+Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
+We compare (C)CH-Potentials to three other heuristics.
 First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
@@
 We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
-Our ALT implementation is uni-directional.
-In this work, we do not consider bidirectional search as it creates problems for some settings, such as predicted traffic.
 Finally, we compare against a hypothetical \emph{Oracle-A*} heuristic.
@@
 In contrast, CH-Potentials already compute exact distances with respect to $w_\ell$.
-The number of popped nodes is the same for CH-Potentials and Oracle-A*.
-The only difference between CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
+The number of popped nodes is the same for CH-Potentials, CCH-Potentials and Oracle-A*.
+The only difference between (C)CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
 This overhead leads to a slowdown of around 1.6.
@@
 This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
+CCH-Potentials are slightly slower than CH-Potentials because their search space is denser.
+
+\subsection{Bidirectional A*}
 
@@
 \centering
-\caption{
-TODO
-OSM Ger, Weight Scaled uniformly by 1.05, alternating dirs
-}\label{tab:bidir_pruning}
+\caption{Performance of different variants of bidirectional A* on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:bidir_pruning}
 \input{table/bidir_pruning.tex}
@@
 
+Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A*.
+As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
+Symmetric bidirectional A* without our improved pruning has the worst performance.
+Enabling the improved pruning brings the performance symmetric bidirectional A* on par with the average potential for all heuristics except ALT.
+Apparently, the average potential gives ALT a better heuristic than the original one for for each direction individually, contrary to the statements made in~\cite{gh-cspas-05}.
+For all other heuristics, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
+Without the low degree improvements, the improved symmetric variant is marginally faster and with the low degree improvements the average potential remains slightly faster.
+This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
+Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.
+
 \begin{table}
@@
 \caption{
-TODO
-OSM Ger, Weight Scaled 1.5, sym with improved pruning, avg without, with topo opt
-}\label{tab:bidir_switching}
-\input{table/bidir_switching.tex}
-\end{table}
-
-\begin{table}
-\centering
-\caption{
-TODO
-OSM Ger
+Performance of bidirectional and unidirectional A* on OSM Ger with different query weights.
+The symmetric variant uses the improved pruning, the average variant does not.
+All variants use all low degree optimizations.
 }\label{tab:bidir}
@@
 \end{table}
+% TODO remove fourth scenario
+
+In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search for A* depending on the query weights.
+Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
+When the query weights are equal, CH-Potentials will already only traverse the shortest path with unidirectional search.
+In this case it is clear that bidirectional search will only introduce unnecessary overhead.
+When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
+This changes significantly when the query weight increases are applied non-uniformly as in the third scenario.
+Here only weights with a speed less than 80\,kph were scaled up.
+This touches only the beginning and end of most shortest paths between randomly chosen nodes.
+The middle part of the shortest paths will typically use faster edges like highways.
+In this case bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
+The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
+In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
+This is also the reason why ALT behaves like this for all query weights.
+By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
+This makes bidirectional search so critical for ALTs performance.
+We conclude that having a potential as tight as CH-Potentials usually makes bidirectional search unnecessary.
+Bidirectional search for A* only pays of when the reduced weights exhibit certain very specific patterns.
+
+\subsection{Applications}
 
@@
 
+\bibliographystyle{ACM-Reference-Format}
 \bibliography{references}
+
+\appendix
+
+\section{Improved Results for k-Nearest-Neighbor Queries on CCH}
+
+\begin{table}
+\centering
+\caption{
+Performance of different closest-POI algorithms for various POI distributions.
+For each distribution, we report the time to index a set of POIs (selection time), the space consumed by the
+index (selection space), and the time to find the $k = 1, 4, 8$ closest POIs (query time).
+}\label{tab:knn}
+\input{table/knn.tex}
+\end{table}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/knn_ball_size.pdf}
+\caption{
+Selection and query times of various closest-POI algorithms with $|P| = 2^14$ POIs picked at random from a ball of varying size $|B|$.
+Queries find the $k = 4$ closest POIs.
+}\label{fig:knn_ball_size}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/knn_num_pois.pdf}
+\caption{
+Selection and query times of various closest-POI algorithms with a varying number $|P|$ of POIs picked at random from a ball of size $|B| = |V|$.
+Queries find the $k = 4$ closest POIs.
+}\label{fig:knn_num_pois}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=.6363\linewidth]{fig/knn_demands.pdf}
+\caption{
+Time to generate a single trip with different demand generators for various values of $\lambda$.
+}\label{fig:knn_demands}
+\end{figure}
+
+\section{Extended Experimental Results on Bidrectional A*}
+
+\begin{table}
+\centering
+\caption{
+Performance of different direction selection criteria of bidirectional A* on OSM Ger with different query weights.
+The symmetric variant uses the improved pruning, the average variant does not.
+All variants use all low degree optimizations.
+}\label{tab:bidir_switching}
+\input{table/bidir_switching.tex}
+\end{table}
+
 \end{document}

### 2021-10-12 eval/bidir_features.py

@@ -118,7 +118,7 @@ bidir = queries.loc[lambda x: x['algo'].str.contains('Topocore')] \
     .groupby(['potential', 'experiment', 'factor', 'probability', 'bidir_pot'])[['running_time_ms', 'num_queue_pushs_k']].mean().unstack(0) \
     .reindex(columns=['Zero', 'ALT', 'CH', 'CCH', 'Oracle'], level=1) \
     .reindex(index=['weight_scale', 'probabilistic_scale_by_speed'], level=0) \
-    .reindex(index=[1.0, 0.0], level=2) \
+    .reindex(index=[1.0], level=2) \
     .reindex(index=['unidir', 'Average', 'Symmetric'], level=3)
 
 bidir[('num_queue_pushs_k', '(C)CH/Oracle')] = bidir[('num_queue_pushs_k', 'CH')]
@@ -129,7 +129,6 @@ bidir = bidir.rename(columns={
     'num_queue_pushs_k': 'Queue pushs [$\\cdot 10^3$]',
 }, index={'unidir': 'Unidirectional'})
 
-bidir.loc[('probabilistic_scale_by_speed', 1.5, 0.0), '$w_q$'] = '> 80kph'
 bidir.loc[('probabilistic_scale_by_speed', 1.5, 1.0), '$w_q$'] = '< 80kph'
 bidir.loc[('weight_scale', 1.0, 1.0), '$w_q$'] = 'unmodified'
 bidir.loc[('weight_scale', 1.05, 1.0), '$w_q$'] = '*1.05'
@@ -141,7 +140,7 @@ lines = lines[0:2] + [
     R' &  & \multicolumn{5}{c}{Running time [ms]} & \multicolumn{3}{c}{Queue pushs [$\cdot 10^3$]} \\ \cmidrule(lr){3-7} \cmidrule(lr){8-10}'
     R'\multirow{2}{*}{$w_q$} & & \multirow{2}{*}{Zero} & \multirow{2}{*}{ALT} & \multirow{2}{*}{CH} & \multirow{2}{*}{CCH} & \multirow{2}{*}{Oracle} & \multirow{2}{*}{Zero} & \multirow{2}{*}{ALT} & (C)CH/ \\',
     R' & & & & & & & & & Oracle \\'
-] + lines[5:9] + ['\\addlinespace'] + lines[9:12] + ['\\addlinespace'] + lines[12:15] + ['\\addlinespace'] + lines[15:]
[…]

### 2021-10-12 Rakefile.rb
implement incremental lazy rphast exp

@@ -188,7 +188,7 @@ end
 
 namespace "exp" do
   desc "Run all experiments"
-  task all: [:preprocessing, :scaled_weights, :building_blocks, :rphast, :bidir_features, :applications]
+  task all: [:preprocessing, :scaled_weights, :building_blocks, :rphast, :rphast_inc, :bidir_features, :applications]
 
   task queries: [:scaled_weights, :building_blocks, :applications]
 
@@ -198,6 +198,7 @@ namespace "exp" do
   directory "#{exp_dir}/scaled_weights"
   directory "#{exp_dir}/building_blocks"
   directory "#{exp_dir}/rphast"
+  directory "#{exp_dir}/rphast_inc"
   directory "#{exp_dir}/bidir_features"
   directory "#{exp_dir}/applications"
 
@@ -247,6 +248,14 @@ namespace "exp" do
     Dir.chdir "code/rust_road_router" do
       static_graphs.each do |graph|
         sh "cargo run --release --bin lazy_rphast -- #{graph} > #{exp_dir}/rphast/$(date --iso-8601=seconds).json"
+      end
+    end
+  end
+
+  task rphast_inc: static_graphs.map { |g| g + 'lower_bound_ch' } + static_graphs.map { |g| g + 'cch_perm' } + ["#{exp_dir}/rphast_inc"] do
+    Dir.chdir "code/rust_road_router" do
+      static_graphs.each do |graph|
+        sh "cargo run --release --bin lazy_rphast_incremental -- #{graph} > #{exp_dir}/rphast_inc/$(date --iso-8601=seconds).json"
       end
     end
   end

### 2021-10-14 paper/ch_potentials.tex
progress on lazy rphast exp

@@
 
-RPHAST, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously).
+RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously).
 Given a set of source nodes $S$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
@@
 We report preprocessing running times as averages over 10 runs.
-For queries, we perform 10\,000 point-to-point queries where both source and target are nodes drawn uniformly at random and report average results.
+To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are nodes drawn uniformly at random and report average results.
+For many-to-one and many-to-many algorithms, we use the methodology from~\cite{delling_et_al:OASIcs:2011:3266}.
+We generate a \emph{ball} $B$ of nodes around 100 uniformly at random chosen nodes by running a Dijkstra search until $|B|$ nodes are settled where $|B|$ is a parameter varied throughout the experiments.
+From each ball we uniformly at random draw a set of sources $S$ and targets $T$ with $|S| = |T| = 2^{14}$.
+For one-to-many experiments, we pick 100 sources of $S$ and for each source compute the distances to all nodes in the respective $T$ for a total of 10\,000 one-to-many queries.
+For many-to-many experiments, we compute the distance table between the 100 $S$, $T$ pairs.
 
@@
 
-% Methodology
-% time per source
-% selection times
-
-% what do we want to show
-% - for one source target set we are as fast as RPHAST
-%   - dimensions: phase (style), graph (color), CH Var (row)
-% - incremental is cool
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{fig/lazy_rphast_inc.pdf}
+\caption{
+Total and per distance running times of incremental Lazy RPHAST with up to $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$ on OSM Germany.
+Note the different y-axis scales and units.
+}\label{fig:lazy_rphast_inc}
+\end{figure}
+
+To evaluate Lazy RPHAST in the incremental setting, we measure the elapsed time after $2^i$ distances were computed.
+Figure~\ref{fig:lazy_rphast_inc} depicts the total elapsed time and the average running time to compute a single distance.
+The first few distances are the most expensive ones since a lot of the CH search space has not been explored yet.
+For later queries, only little work remains to be done and the overhead per distance becomes basically a constant depending on the ball size.
+
+The performance difference between the CH and CCH based variant is interesting.
+The CCH based variant can utilize the elimination tree for a more efficient implementation but the search space denser.
+This makes earlier queries more expensive and later queries cheaper.
+With this trade-off, it depends on the ball size which variant is faster in terms of total running time.
 
@@
 \caption{
-Many-to-One CH
+Running times of CH based Lazy RPHAST with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
 }\label{fig:many_to_one}
@@
 \caption{
-Many-to-One CCH
+Running times of CCH based Lazy RPHAST with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
 }\label{fig:many_to_one_cch}
@@
 
+Despite the laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental one-to-many queries.
+Figures~\ref{fig:many_to_one} and~\ref{fig:many_to_one_cch} depict average running times to compute distances between one node and $2^{14}$ targets for different ball sizes.
+This experiment uses the methodology exactly as described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to relate our results to the performance of competing algorithms (set aside the performance differences caused by using a different test machine).
+When looking at total running times for one source target set pair, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}
+For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACs Europe.
+Lazy RPHAST (on a CH) takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
+The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
+This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster query phases.
+When computing distances from different source towards the same target set, RPHAST will outperform Lazy RPHAST.
+However, whenever the target set may change between queries, Lazy RPHAST is a simpler and faster alternative.
+
 \begin{figure}
@@
 \caption{
-Many-to-Many
+Average running times of Lazy RPHAST for many-to-many queries with $|S| = |T| = 2^{14}$ sources and targets picked from a ball of varying size $|B|$ with standard deviation indicated as error bars.
+The lower part of the bars indicates the running time of the Bucket CH subphase.
 }\label{fig:many_to_many}
 \end{figure}
+
+Applying Lazy RPHAST to the many-to-many problem results in running times as depicted in Figure~\ref{fig:many_to_many}.
+The results are again at least competitive to the performance reported for SSE RPHAST in~\cite{delling_et_al:OASIcs:2011:3266} which to the best of our knowledge is the fastest known approach.
+The Bucket CH based first query phase takes between half of the time for $|B| = 2^{14}$ and a tenth of the time for $|B| = 2^{24}$.
+It is worth noting that our approach does not require any explicit utilization of SSE instructions -- contrary to SSE RPHAST.
+Nevertheless, an inspection of the generated assembly reveals that the code for Lazy RPHAST on an array of distances was automatically vectorized by the compiler.
 
@@
 \caption{
-Selection and query times of various closest-POI algorithms with $|P| = 2^14$ POIs picked at random from a ball of varying size $|B|$.
+Selection and query times of various closest-POI algorithms with $|P| = 2^{14}$ POIs picked at random from a ball of varying size $|B|$.
 Queries find the $k = 4$ closest POIs.

### 2021-10-14 Rakefile.rb

@@ -9,6 +9,7 @@ file "paper/ch_potentials.pdf" => [
   "paper/fig/lazy_rphast_many_to_one.pdf",
   "paper/fig/lazy_rphast_cch_many_to_one.pdf",
   "paper/fig/lazy_rphast_many_to_many.pdf",
+  "paper/fig/lazy_rphast_inc.pdf",
   "paper/table/applications.tex",
   "paper/table/bidir_pruning.tex",
   "paper/table/bidir_switching.tex",
@@ -38,6 +39,10 @@ namespace "fig" do
 
   file "paper/fig/lazy_rphast_many_to_many.pdf" => FileList["#{exp_dir}/rphast/*.json"] + ["eval/rphast_comparison.py"] do
     sh "eval/rphast_comparison.py"
+  end
+
+  file "paper/fig/lazy_rphast_inc.pdf" => FileList["#{exp_dir}/rphast_inc/*.json"] + ["eval/rphast_inc.py"] do
+    sh "eval/rphast_inc.py"
   end
 end
 

### 2021-10-14 eval/rphast_comparison.py

@@ -26,6 +26,7 @@ queries = pd.DataFrame.from_records([{
     for run in data for exp in run['experiments'] for algo in exp['algo_runs']])
 
 queries['running_time_s'] = queries['running_time_ms'] / 1000.0
+queries['selection_running_time_s'] = queries['selection_running_time_ms'] / 1000.0
 
 plt.figure(figsize=(11,4))
 g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_one'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_ms', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
@@ -50,9 +51,15 @@ plt.tight_layout()
 g.get_figure().savefig('paper/fig/lazy_rphast_cch_many_to_one.pdf')
 
 plt.figure(figsize=(11,4))
-g = sns.boxplot(data=queries.loc[lambda x: x['algo'] == 'lazy_rphast_many_to_many'].loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_s', hue='graph', showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
+g = sns.barplot(data=queries
+                 .loc[lambda x: x['algo'] == 'lazy_rphast_many_to_many']
+                 .loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='running_time_s', hue='graph')
+g = sns.barplot(data=queries
+                 .loc[lambda x: x['algo'] == 'lazy_rphast_many_to_many']
+                 .loc[lambda x: x['target_set_size_exp'] == 14], x='ball_size_exp', y='selection_running_time_s', hue='graph', palette=sns.color_palette("pastel"))
 g.set_xlabel('Ball Size $|B|$')
 g.set_ylabel('Running Time [s]')
-g.legend().set_title('')
+handles, labels = g.get_legend_handles_labels()
+g.legend(handles[0:2], labels[0:2])
 plt.tight_layout()
 g.get_figure().savefig('paper/fig/lazy_rphast_many_to_many.pdf')

### 2021-10-14 eval/rphast_inc.py

@@ -1,0 +1,44 @@
+#!/usr/bin/env python3
+
+import numpy as np
+import matplotlib as mpl
+from matplotlib import pyplot as plt
+import seaborn as sns
+sns.set()
+
+import pandas as pd
+
+import json
+import glob
+import os
+
+base = "exp/"
+paths = glob.glob(base + "rphast_inc/*.json")
+data = [json.load(open(path)) for path in paths]
+
+queries = pd.DataFrame.from_records([{
+    **progress,
+    'algo': algo['algo'],
+    'experiment': exp['experiment'],
+    'ball_size_exp': exp['ball_size_exp'],
+    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
+    }
+    for run in data for exp in run['experiments'] for algo in exp['algo_runs'] for progress in algo['progress_states']])
+
+queries['single_query_time_mus'] = queries['passed_time_mus'] / (2 ** queries['terminal_rank'])
+queries['terminal_idx'] = (2 ** queries['terminal_rank'])
+queries['passed_time_ms'] = queries['passed_time_mus'] / 1000.0
+queries['Algorithm'] = queries['algo'].map({ 'lazy_rphast_many_to_one': 'CH', 'lazy_rphast_cch_many_to_one': 'CCH' })
+queries['Ball Size $|B|$'] = queries['ball_size_exp'].map('$2^{{{}}}$'.format)
+
+fig, (ax1, ax2) = plt.subplots(2, sharex=True, figsize=(11,8))
+g = sns.lineplot(data=queries.query('graph == "OSM Germany" & ball_size_exp in [14,17,21,24]'), x='terminal_idx', y='passed_time_ms', hue='Ball Size $|B|$', style='Algorithm', ci=None, marker='X', ax=ax1)
+g.set_xlabel('Number of Computed Distances')
+g.set_ylabel('Running Time [ms]')
+g = sns.lineplot(data=queries.query('graph == "OSM Germany" & ball_size_exp in [14,17,21,24]'), x='terminal_idx', y='single_query_time_mus', hue='Ball Size $|B|$', style='Algorithm', ci=None, marker='X', ax=ax2, legend=False)
+g.set_yscale('log')
+g.set_xlabel('Number of Computed Distances')
[…]

### 2021-10-14 eval/building_blocks.py
continue with application exp

@@ -35,7 +35,7 @@ table = table.unstack(0).stack(0) \
 
 lines = table.to_latex(escape=False).split("\n")
 
-lines = [R'\begin{tabular}{clllrrrrr}'] + [lines[1]] + [
+lines = [R'\begin{tabular}{ccccrrrrr}'] + [lines[1]] + [
     R"       & BCC & Deg2 & Deg3 & Zero & ALT & CH & CCH & Oracle \\"
 ] + lines[4:]
 

### 2021-10-15 paper/ch_potentials.tex
move bidir switching exp into main body

@@
 
-In this section, we discuss a very specific setting arises naturally from A* heuristics.
+In this section, we discuss a very specific setting which naturally arises from A* heuristics.
 Here, the target node is known in the selection step but the source nodes are only queried one after another.
@@
 By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
-We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{todo}.
+We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{sec:exp_lazy_rphast}.
 
@@
 
-% TODO alternating directions
+For both average and symmetric bidirectional A*, we always alternate between settling a node form the forward and the backward search.
+In our experiments, we also evaluate choosing the direction by the smaller minimal queue key.
+Even though this appears to better fit together with the stopping criterion for the symmetric approach, our experiments show that it is never beneficial in terms of running time (see Table~\ref{tab:bidir_switching}).
 
@@
 % A* is a flexible algorithm.
-% Many extended problems can therefore be solved using the CH-Potentials framework.
+Many extended problems can be solved using CH-Potentials.
 We start by establishing a common formal framework for the use of CH-Potentials.
@@
 
-\subsection{Lazy RPHAST}
+\subsection{Lazy RPHAST}\label{sec:exp_lazy_rphast}
 
@@
 \caption{
+Performance of different direction selection criteria of bidirectional A* on OSM Ger with different query weights.
+The symmetric variant uses the improved pruning, the average variant does not.
+All variants use all low degree optimizations.
+}\label{tab:bidir_switching}
+\input{table/bidir_switching.tex}
+\end{table}
+
+Table~\ref{tab:bidir_switching} shows the performance of bidirectional A* with different strategies to decide whether to advance the forward or the backward search next.
+The results clearly show that alternating the directions is always superior.
+Selecting the direction by minimal queue key may lead to huge imbalances in the progress of the searches.
+This causes the searches to meet later and the total search space to grow significantly.
+
+\begin{table}
+\centering
+\caption{
 Performance of bidirectional and unidirectional A* on OSM Ger with different query weights.
@@
 \end{table}
-% TODO remove fourth scenario
-
-In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search for A* depending on the query weights.
+
+In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
 Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
@@
 
-\section{Extended Experimental Results on Bidrectional A*}
-
-\begin{table}
-\centering
-\caption{
-Performance of different direction selection criteria of bidirectional A* on OSM Ger with different query weights.
-The symmetric variant uses the improved pruning, the average variant does not.
-All variants use all low degree optimizations.
-}\label{tab:bidir_switching}
-\input{table/bidir_switching.tex}
-\end{table}
-
 \end{document}

### 2021-10-15 eval/bidir_features.py

@@ -73,8 +73,9 @@ bidir_switch_strat = queries.loc[lambda x: x['bidir_pot'] != 'unidir'] \
     .loc[lambda x: x['algo'].str.contains('Topocore')] \
     .loc[lambda x: ~(((x['bidir_pot'] == 'Average') & (x['improved_pruning'] == True)) | ((x['bidir_pot'] == 'Symmetric') & (x['improved_pruning'] == False)))] \
     .groupby(['potential', 'experiment', 'factor', 'probability', 'bidir_pot', 'choose_direction_strategy'])[['running_time_ms', 'num_queue_pushs_k']].mean().unstack(0) \
-    .loc[('probabilistic_scale_by_speed', 1.50)] \
-    .reindex(columns=['Zero', 'ALT', 'CH', 'CCH', 'Oracle'], level=1)
+    .reindex(columns=['Zero', 'ALT', 'CH', 'CCH', 'Oracle'], level=1) \
+    .reindex(index=['weight_scale', 'probabilistic_scale_by_speed'], level=0) \
+    .reindex(index=[1.0], level=2)
 
 bidir_switch_strat[('num_queue_pushs_k', '(C)CH/Oracle')] = bidir_switch_strat[('num_queue_pushs_k', 'CH')]
 bidir_switch_strat = bidir_switch_strat.drop([('num_queue_pushs_k', 'CH'), ('num_queue_pushs_k', 'CCH'), ('num_queue_pushs_k', 'Oracle')], axis = 'columns')
@@ -84,15 +85,19 @@ bidir_switch_strat = bidir_switch_strat.rename(columns={
         'num_queue_pushs_k': 'Queue pushs [$\\cdot 10^3$]',
     }) \
     .rename(index={
-        0.00: '> 80kph',
-        1.00: '< 80kph',
         'alternating': 'Alternating',
         'min_key': 'Min. Key',
     })
 
+bidir_switch_strat.loc[('probabilistic_scale_by_speed', 1.5, 1.0), '$w_q$'] = '< 80kph'
+bidir_switch_strat.loc[('weight_scale', 1.0, 1.0), '$w_q$'] = 'unmodified'
+bidir_switch_strat.loc[('weight_scale', 1.05, 1.0), '$w_q$'] = '*1.05'
[…]

### 2021-10-15 paper/ch_potentials.tex
some discussion of valentins results

@@
 
+In this section, we report the results of the k-Nearest-Neighbor Queries on CCH with Lazy RPHAST.
+The authors of~\cite{buchhold_et_al:LIPIcs.SEA.2021.18} provide the code for their algorithm and their experiments on Github\footnote{\url{https://github.com/vbuchhold/routing-framework}}.
+This includes the code for both the old version and the new one improved with Lazy RPHAST.
+We reran the experiments performed for~\cite{buchhold_et_al:LIPIcs.SEA.2021.18} in both the old and the new configuration on our machine.
+In the results, we denote the old version as \emph{CCH/CRAD} and the new one as \emph{CCH*/CRAD*}.
+Our results exhibit the same characteristics as observed in the original publication.
+Absolute running times of course vary slightly.
+
+Lazy RPHAST yields the greatest improvements of more than an order of magnitude when the points of interest (\emph{POIs}) are concentrated in a smaller region (see Table~\ref{tab:knn} and Figure~\ref{fig:knn_ball_size}).
+When the POIs are distributed over the whole graph, the speed-up is only around factor of three.
+Lazy RPHAST also achieves greater speed-ups for larger numbers of POIs (see Figure~\ref{fig:knn_num_pois}).
+For the demand generation application, this translates to a speed-up factor of around 2.5 (see Figure~\ref{fig:knn_demands}).
+For a more detailed discussion of these results and a comparison to competing algorithms, we refer to~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
+
 \begin{table}

### 2021-10-15 paper/ch_potentials.tex
improve writing

@@
 
-\title{A Fast and Tight Heuristic for A* in Road Networks}
+\title{Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks}
 
@@
 \begin{abstract}
-We study exact, efficient and practical algorithms for route planning in large road networks.
+We study exact, efficient and practical algorithms for route planning applications in large road networks.
 Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
@@
 Extending them to certain practical applications can be hard.
-In this paper, we present an algorithm to efficiently extract distance estimates for A* from Contraction Hierarchies (CH), a hierarchical technique.
-We call our heuristic CH-Potentials.
+In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as a A* heuristic.
+Since it utilizes Contraction Hierarchies (CH), a hierarchical technique, we call our heuristic CH-Potentials.
 Our approach allows decoupling the supported extensions from the hierarchical speed-up technique.
-Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks.
+Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present an improved pruning criterion for symmetrical bidirectional A*.
+An extensive experimental study confirms the practicality of our approach for many practical problems.
 \end{abstract}
@@
 However, the algorithm development is vastly more complex.
-For every feature, dedicated research paper(s) exist that extend CH.
+For every feature, dedicated research papers exist that extend CH and other approaches.
 Supporting the combination of several features is even harder.
@@
 Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
+% TODO contribution vs conference
 
@@
 
-\subsection{Contraction Hierarchy (CH)}
+\subsection{Contraction Hierarchies (CH)}
 
@@
 
-\subsection{Customizable Contraction Hierarchy (CCH)}
+\subsection{Customizable Contraction Hierarchies (CCH)}
 
@@
 They use a node importance order independent of the weight function.
+% TODO so far, we only have levels but not ranks
 This allows splitting the preprocessing in two subphases.
 The first one is slow but independent of the weight function.
-The second one, called \emph{customization}, is fast (less than a second) and applies a weight function to the preprocessed data.
+The second one, called \emph{customization}, is fast (a couple of seconds or less) and applies a weight function to the preprocessed data.
 This allows regular updates to the weight function, for example to incorporate live traffic.
@@
 The root of the elimination tree is the most important node.
+% TODO so far, we only have levels but not ranks
 For every other node its parent is its least important upward neighbor.
@@
 This makes it possible to explore the CCH search space with a more efficient shortest path algorithm for DAGs:
-Traverse elimination tree from source and target to root node and for each node relax all up-edges (reversed down-edges for the backward search).
+Traverse elimination tree from source and target to the root node and for each node relax all up-edges (reversed down-edges for the backward search).
 Because this algorithm does not require a queue, it is faster than a Dijkstra based CH query.
@@
 
-PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the one-to-all case works analogously).
+PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the reverse case works analogously).
+This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
 The preprocessing phase remains the same as for CH.
@@
 
-RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously).
+RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously), solving the \emph{many-to-one} problem.
 Given a set of source nodes $S$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
 This step is called \emph{selection}.
-Then, in the query step a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
+Then, in the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
 RPHAST is particularly effective when many targets are queried for the same source set $S$.
 If the source set changes often, selection times may become problematic.
+RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
+Combined with SSE extension it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.
 
@@
 
-Most route planning algorithms focus on the point-to-point shortest path problem.
-However, certain applications need distances between sets $S$ and $T$ of nodes.
-This is variant is known as the \emph{many-to-many} problem variant.
-In some cases one of the sets is a single node.
-This is then called a \emph{one-to-many} or \emph{many-to-one} problem.
-RPHAST, as discussed in the previous paragraph, can be applied to these problems.
-
-In this section, we discuss a very specific setting which naturally arises from A* heuristics.
+In this section, we discuss another very specific problem setting which naturally arises from A* heuristics.
 Here, the target node is known in the selection step but the source nodes are only queried one after another.
@@
 The first step is the \emph{target selection} where the target node is provided.
-Then, an arbitrary number of source nodes $s_i$ are given one after another and for each source the distance to the target has to be computed.
+Then, an arbitrary number of source nodes $s_i$ are given one after another.
+For each source the distance to the target has to be computed before the next source will be provided.
 
@@
 However, we can also exploit the elimination tree for the Lazy RPHAST algorithm.
-Computing the distance $D[x]$ requires knowing the final distance of all upward neighbors.
-In Algorithm~\ref{algo:pot} we have to compute these distances recursively.
+To compute the distance $D[x]$ of a node $x$ the distances of all upward neighbors must be final.
+In Algorithm~\ref{algo:pot} we have to compute all these distances recursively.
+However, on CCH it is sufficient to ensure that the distance of the elimination tree parent of $x$ is final.
 All upward neighbors of $x$ are by definition included in the search space, thus they are also ancestors in the elimination tree.
-If the distance of the parent in the elimination tree has already been computed, the distances of all other upward neighbors also has to have been computed.
+Thus, if the distance of the parent in the elimination tree is final, the distances of all other upward neighbors must already be final, too.
 Thus, it is sufficient to recursively (or iteratively) ensure that the distance of the parent in the elimination tree has been computed.
@@
 
-The CH-Potentials heuristic is a straightforward application of the Lazy RPHAST algorithm.
+The \emph{CH-Potentials} heuristic is a straightforward application of the Lazy RPHAST algorithm.
 In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.

### 2021-10-15 paper/ch_potentials.tex
lots of writing

@@
 
-In this section, we discuss how the Lazy RPHAST algorithm can be applied in a many-to-many context.
-To the best of our knowledge, the fastest way to compute a full distance table between a set of sources and a set of targets is RPHAST utilizing SSE instructions to process several sources at once.
-First, the selection phase is performed for the full target set.
+Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investing how well Lazy RPHAST performs in this setting appears to be an interesting question.
+SSE RPHAST starts by performing the selection phase is performed for the full target set.
 Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
 This second step can be improved by processing $k$ (typically 16) sources at once using SIMD vector instructions (SSE RPHAST).
-
-Since RPHAST is the fastest known algorithm for many-to-many computations, investing how well Lazy RPHAST performs in this setting appears to be an interesting question.
-The naive approach would be to select one target after another and for each one doing the full lazy exploration from all sources.
+Naively translating this approach to Lazy RPHAST would mean selecting one target after another and for each doing the full lazy exploration from all sources.
 However, this approach would explore the upward search space of all the sources over and over again.
@@
 
-Luckily, we can do better by using a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
+Luckily, we can do better by integrating a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
 The Bucket CH selection phase works as follows:
 For each node $x$, we maintain an initially empty list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
-These buckets can be populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
-Now we can utilize Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) to compute distances from a given source to \emph{all} targets at once.
+These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
+Now Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
 We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
 After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target node.
-This way, we explore the search space of the sources only once.
+This way, the search space of the sources is explored only once.
 
 The Bucket-CH selection phase still explores the shared parts of the search space of the targets several times.
-This can also be improved with the following modified CH backward search:
+This can be improved with the following modified CH backward search:
 All target nodes are inserted into the queue at once.
@@
 The buckets of each target node are initialized with a single entry for the respective node with distance zero.
-When relaxing an edge, the bucket of the head nodes is merged with the bucket of the tail node with each distance increased by the weight of the edge.
+When relaxing an edge, the entries in the current node's bucket are increased by the edge's weight and merged with the bucket entries of the head node of the edge.
 By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
@@
 The problem is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
-Thus, in the equivalent bidirectional Dijkstra search each direction would use different graph.
-This would break the bidirectional Dijkstra stopping criterion.
+Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
+This breaks the bidirectional Dijkstra stopping criterion.
 To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until their minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
 But this is the same stopping criterion that a unidirectional A* search has.
-Thus, this straightforward bidirectional A* performs about twice the work than a unidirectional A*.
+Thus, this straightforward bidirectional A* performs twice the work than a unidirectional A*.
 This can be slightly improved by introducing pruning.
-Nodes, for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e.
+Nodes for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e.
 their distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, can not contribute to a shorter path and need not to be added to the queue.
@@
 
-To overcome this problem the authors of propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
-The reduced graph of the average potential is the graph with the average weights of the reduced graphs of the forward and backward potentials.
+To overcome this problem the authors of~\cite{gh-cspas-05} propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
+The reduced graph of the average potential is the graph with the average weights of the both reduced graphs.
 The potential function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$ (and $-h_{\overline{st}}(x)$ for the backward search).
@@
 It takes the search progress (in form of the growing minimal queue keys) into account to tighten the pruning with increasing search progress and compensates for the worse stopping criterion.
-As our experiments show, this makes the simpler symmetric approach at least competitive and sometimes even faster than the average potential.
+As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
 Recall that the queue is ordered by $D[x] + h(x)$.
@@
 The pruning for the backwards search works analogously.
+Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
 
@@
 One could also apply live traffic updates to $w_q$ immediately and only update $w_\ell$ once the difference between $w_\ell$ and $w_q$ grows too big or when some edge weight update would set $w_q < w_\ell$.
-% For applications which allow for a three-phase setup, CCH-Potentials are likely a better choice for supporting live traffic.
-% However, evaluating CCH-Potentials is beyond the scope of this paper.
-% We focus on evaluating CH-Potentials as a simple building block in the two-phase setup.
 
@@
 For sake of keeping this section self-contained, we now briefly reiterate the details of our configuration.
-The stretch limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \cdot \dist(u,v)$.
-Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was $k$ times on a shortest path its weight will be $w_q = w_{\ell}\cdot\psi^k$.
+The stretch is limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \cdot \dist(u,v)$.
+Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was on the shortest path in $k$ iterations its weight will be $w_q = w_{\ell}\cdot\psi^k$.
 We also penalize edges incident to shortest paths with an asymmetric additive rejoin penalty.
@@
 
+In~\cite{adgw-arrn-13} Abraham et al. describe a variation of the penalty method to directly obtain alternative routes rather than an alternative graph:
+Just return the first path which shares less than 80\% with the original path and has stretch less than $1+\epsilon$.
+However, the approach in~\cite{adgw-arrn-13} uses a simpler penalization schema.
+We also implement this approach with CH-Potentials but with the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative}.
+This leads to some qualitative improvements compared to the results reported in~\cite{adgw-arrn-13}.
+
 \subsubsection{Temporary Driving Bans}
@@
 In this section, we present our experimental evaluation.
-Our benchmark machine runs openSUSE Leap 15.1 (kernel 4.12.14), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
+Our benchmark machine runs openSUSE Leap 15.3 (kernel 5.3.18), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
 All experiments were performed sequentially.
-Our code is written in Rust and compiled with rustc 1.47.0-nightly in the release profile with the target-cpu=native option.
+Our code is written in Rust and compiled with rustc 1.57.0-nightly in the release profile with the target-cpu=native option.
 The source code of our implementation and the experimental evaluation can be found on Github\footnote{\url{https://github.com/kit-algo/ch_potentials}}.
@@
 The PTV instances are not OSM-based.
-One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges and the other one a newer instance of Europe with predictions for 27\% of the edges.
-% TODO PTV20
-Table~\ref{tab:graphs} contains an overview over our instances.
+One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges which was used for evaluation of many competing time-dependent algorithms.
+The other two are newer instances of Europe with predictions for 27\% (TDEur17) and 76\% (TDEur20) of the edges.
+Table~\ref{tab:graphs} contains an overview over the instances.
 In this table, we further include the sequential preprocessing running time to construct the (C)CH.
@@
 The first few distances are the most expensive ones since a lot of the CH search space has not been explored yet.
-For later queries, only little work remains to be done and the overhead per distance becomes basically a constant depending on the ball size.
+With around 100\,$\mu$s, the running times are comparable to standard CH queries.
+For later queries, only little work remains to be done and the overhead per distance becomes almost constant depending on the ball size.
 
@@
 Figures~\ref{fig:many_to_one} and~\ref{fig:many_to_one_cch} depict average running times to compute distances between one node and $2^{14}$ targets for different ball sizes.
-This experiment uses the methodology exactly as described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to relate our results to the performance of competing algorithms (set aside the performance differences caused by using a different test machine).
+This experiment uses the exact methodology described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to roughly relate our results to the performance of competing algorithms.
 When looking at total running times for one source target set pair, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}
@@
 When computing distances from different source towards the same target set, RPHAST will outperform Lazy RPHAST.
-However, whenever the target set may change between queries, Lazy RPHAST is a simpler and faster alternative.
+However, whenever the target set may change between queries, Lazy RPHAST offers a simple and fast alternative.
 
@@
 Applying Lazy RPHAST to the many-to-many problem results in running times as depicted in Figure~\ref{fig:many_to_many}.
-The results are again at least competitive to the performance reported for SSE RPHAST in~\cite{delling_et_al:OASIcs:2011:3266} which to the best of our knowledge is the fastest known approach.
+The results are again at least competitive to the performance reported for SSE RPHAST in~\cite{delling_et_al:OASIcs:2011:3266}.
 The Bucket CH based first query phase takes between half of the time for $|B| = 2^{14}$ and a tenth of the time for $|B| = 2^{24}$.
-It is worth noting that our approach does not require any explicit utilization of SSE instructions -- contrary to SSE RPHAST.
-Nevertheless, an inspection of the generated assembly reveals that the code for Lazy RPHAST on an array of distances was automatically vectorized by the compiler.
+It is worth noting that contrary to SSE RPHAST, we did not explicitly utilize SSE instructions.
+However, the code for Lazy RPHAST on an arrays of distances can be trivially vectorized by the compiler.
+We confirmed that the code was indeed vectorized by inspecting the generated assembly.
 
@@
 
-Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A*.
+In this section, we closely investigate the performance of bidirectional A*.
+Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A* with and without our improved pruning.
 As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
@@
 This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
-Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.
+Enabling the improved pruning for the average potential, marginally reduces the search space size and slightly increases running times.
 
@@
 Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
-When the query weights are equal, CH-Potentials will already only traverse the shortest path with unidirectional search.
-In this case it is clear that bidirectional search will only introduce unnecessary overhead.
+With $w_q = w_{\ell}$, unidirectional CH-Potentials are already optimal and only traverse the shortest path.
+Here, bidirectional search will only introduce unnecessary overhead.
 When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
-This changes significantly when the query weight increases are applied non-uniformly as in the third scenario.
+This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
 Here only weights with a speed less than 80\,kph were scaled up.
@@
 The middle part of the shortest paths will typically use faster edges like highways.
-In this case bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
+In this case, bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
 The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
@@
 By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
+Conversely, unimportant edges have bad reduced weights.
 This makes bidirectional search so critical for ALTs performance.
-We conclude that having a potential as tight as CH-Potentials usually makes bidirectional search unnecessary.
-Bidirectional search for A* only pays of when the reduced weights exhibit certain very specific patterns.
+In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
+Bidirectional search for CH-Potentials only pays of when the reduced weights exhibit certain very specific patterns.
 
@@
 
-In this paper, we introduced CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and CH for finding shortest paths in road networks.
+In this article, we introduced Lazy RPHAST, an efficient algorithm for the incremental many-to-one problem.
+We also showed that it delivers competitive performance for one-to-many and many-to-many problems.
+Based on Lazy RPHAST, we developed (C)CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and (C)CH for finding shortest paths in road networks.
 The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
@@
 Achieving significantly faster running times could still be possible in variations of the problem setting.
-% This leads to multiple avenues for future research.
+This leads to multiple avenues for future research.
 
@@
 Hierarchical techniques have been shown to work well on these graphs~\cite{DBLP:conf/aaai/UrasK14}.
+It might also be worthwhile to apply CH-Potentials to even more routing applications, for example to the shortest $\epsilon$-smooth path problem~\cite{dss-tarrn-18}.
 

### 2021-10-18 paper/ch_potentials.tex
more application exps

@@
 Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present an improved pruning criterion for symmetrical bidirectional A*.
-An extensive experimental study confirms the practicality of our approach for many practical problems.
+An extensive experimental study confirms the practicality of our approach for many applications.
 \end{abstract}
@@
 For example, we are not aware of any work combining all features mentioned above.
-In this paper, we describe an algorithmic building block, that allows handling the combination of all above mentioned features -- and probably more.
+In this article, we describe an algorithmic building block, that allows handling the combination of all above mentioned features -- and probably more.
 
@@
 % However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
-In this paper, we describe CH-Potentials, a fast heuristic with tight estimates.
+In this article, we introduce Lazy RPHAST and build on it CH-Potentials, a fast heuristic with tight estimates.
 Internally, the heuristic uses a CH.
@@
 This enables us to design algorithms for a multitude of features.
-In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
-These low degree optimizations are applicable to Dijkstra's algorithm and A*.
-
-The rest of the paper is organized as follows.
+
+This article is the extended version of our conference paper ``A Fast and Tight Heuristic for A* in Road Networks''~\cite{strasser_et_al:LIPIcs.SEA.2021.6}.
+The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree nodes, common in road networks, and some applications of CH-Potentials.
+In this article we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in much greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
+Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
+The original paper only discussed unidirectional CH-Potentials.
+Here, we also discuss a bidirectional variant and also introduce an improved pruning criterion for symmetrical bidirectional A* which can also be utilized for other heuristic.
+Finally, we include additional applications of CH-Potentials, for example alternative routes, and provide a significantly expanded evaluation.
+
+The rest of the article is organized as follows.
 In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
+After establishing preliminaries in~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and our algorithm to solve it efficiently in Section~\ref{sec:lazy-rphast}.
 CH-Potentials, our new distance estimation function is introduced in Section~\ref{sec:main-algo}.
-Section~\ref{sec:low-deg-improvment} discusses our improvements for the handling of low-degree nodes.
+There we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
+Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
 In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
 Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
-% TODO contribution vs conference
 
@@
 
-\subsection{Bidirectional A*}
+\subsection{Bidirectional A*}\label{sec:bidir_astar}
 
@@
 Just return the first path which shares less than 80\% with the original path and has stretch less than $1+\epsilon$.
-However, the approach in~\cite{adgw-arrn-13} uses a simpler penalization schema.
-We also implement this approach with CH-Potentials but with the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative}.
+However, their approach uses a simpler penalization schema.
+We also implement this variant with CH-Potentials but stick to the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative}.
 This leads to some qualitative improvements compared to the results reported in~\cite{adgw-arrn-13}.
@@
 CH-Potentials performance for different route planning applications.
+Depending on the problem we apply unidirectional or bidirectional CH-Potentials (CH U or CH B) or CCH-Potentials (CCH U/B).
 We report average running times and number of queue pushes.
@@
 
-Table~\ref{tab:applications} depicts the running times of CH-Potentials in various applications, such as those described in Section~\ref{sec:extensions}.
+Table~\ref{tab:applications} depicts the running times of (C)CH-Potentials in various applications, such as those described in Section~\ref{sec:extensions}.
 We report speedups compared to extensions of Dijkstra's algorithm for each application respectively.
@@
 CH achieves average query running times of 0.16\,ms on OSM Ger.
-CH-Potentials are roughly four times slower but still achieve a huge speedup of 3243 over Dijkstra.
+CH-Potentials are roughly four times slower but still achieve a huge speedup of 3795 over Dijkstra.
 Such large speedups are typical for CH.
 This shows that CH-Potentials gracefully converges toward a CH in the $w_q = w_\ell$ special case.
+On DIMACs Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.
 
@@
 The achieved speedups are therefore comparable to CH speedups.
-%
-Mapbox live traffic has a length increase of around 15\%, which yields running times of 127\,ms.
-The length increase of Mapbox traffic predictions are about 18\%, and results in a running time of 200\,ms.
+Since turn costs an restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speed-ups.
+Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.
+Applying bidirectional CH-Potentials in this case is actually detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
+The same is the case for forbidden tunnels or highways.
+Applying CCH-Potentials customized to the current traffic (or the metric with forbidden edges) to these problems yields running times similar to the unmodified case and speed-ups of more than three orders of magnitude.
+With bidirectional CCH-Potentials, one can easily additionally support turn costs and still have running times of a few milliseconds without any adjustments to the customization.
+
+The length increase of Mapbox traffic predictions are about 12.3\%, and results in a running time of 120\,ms.
 The speedup in the predicted scenario is larger than in the live setting, as the travel time function evaluations slow down Dijkstra's algorithm.
-Combining predicted and live traffic results in a running time only slightly higher than for the predicted scenario.
-Further adding turn restrictions, increases the running times.
+Combining predicted and live traffic results in a running time of around 200\,ms.
+Further adding turn restrictions, significantly increases the running times.
 This increase is mostly due to the BCC optimization of Section~\ref{sec:largested-biconnected-component} becoming ineffective when considering turns.
 It is not due to the length increase of using turns.
-With everything activated, our algorithm still has a speedup of 12.2 over the baseline.
+With everything activated, our algorithm still has a speedup of 13.5 over the baseline.
 Interestingly, the PTV traffic predictions have a much smaller length increase than the Mapbox predictions.
-This results in smaller running times of our algorithm.
+This results in somewhat smaller running times of our algorithm.
 

### 2021-10-18 eval/applications.py

@@ -12,6 +12,17 @@ cwd = os.getcwd()
 paths = glob.glob(base + "applications/*.json")
 data = [json.load(open(path)) for path in paths]
 
+def path_to_graph(path):
+    return {
+        'ger06': 'TDGer06',
+        'ptv17': 'TDEur17',
+        'ptv20': 'TDEur20',
+        'osm_europe': 'OSM Europe',
+        'osm_ger': 'OSM Ger',
+        'osm_ger_td': 'OSM Ger',
+        'europe': 'DIMACs Eur',
+    }[[x for x in path.split('/') if x != ''][-1]]
+
 only_public = 'ONLY_PUBLIC' in os.environ
 
 queries = pd.DataFrame.from_records([{

### 2021-10-18 eval/graphs.py

@@ -17,8 +17,8 @@ def path_to_graph(path):
         'ptv17': 'TDEur17',
         'ptv20': 'TDEur20',
         'osm_europe': 'OSM Europe',
-        'osm_ger': 'OSM Ger',
-        'osm_ger_td': 'TD OSM Ger',
+        'osm_ger': 'OSM Germany',
+        'osm_ger_td': 'OSM Germany',
         'europe': 'DIMACs Europe',
     }[[x for x in path.split('/') if x != ''][-1]]
 

### 2021-10-18 paper/ch_potentials.tex
introduce both node ranks and levels

@@
 
-A CH places nodes into levels.
-No edge must connect two nodes within one level.
-Levels are ordered by ``importance''.
+In a CH, each node gets a unique rank representing its ``importance''.
 The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
-An edge goes \emph{up} when it goes from a node in a lower level to higher level.
+An edge goes \emph{up} when its tail node has lower rank than its head node.
 \emph{Down} edges are defined analogously.
@@
 Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
-This is done by repeatedly contracting unimportant nodes and adding shortcuts between its neighbors.
+This is done by repeatedly contracting the least important node and adding shortcuts between its neighbors.
 %See~\cite{gssv-erlrn-12} for the details.
@@
 For every other node its parent is its least important upward neighbor.
-In~\cite{bcrw-s-16}, it was proven that set of ancestors of a node in the elimination tree is equal to the nodes reachable in a forward/backward CH search.
+In~\cite{bcrw-s-16}, it was proven that set of ancestors of a node in the elimination tree is equal to the nodes reachable in a CH search.
 This makes it possible to explore the CCH search space with a more efficient shortest path algorithm for DAGs:
@@
 Algorithm~\ref{algo:ch-backward} shows this first step.
-The second step iterates over all CH levels from top to bottom.
+For the second step, PHAST organizes nodes into \emph{levels}.
+These levels correspond to the importance ordering but allow multiple nodes in the same level.
+No edge must connect two nodes within the same level.
+PHAST then iterates over all CH levels from top to bottom.
 In each iteration, all up-edges starting within the current level are followed in reverse.

### 2021-10-18 Rakefile.rb
penalty exp

@@ -195,7 +195,7 @@ namespace "exp" do
   desc "Run all experiments"
   task all: [:preprocessing, :scaled_weights, :building_blocks, :rphast, :rphast_inc, :bidir_features, :applications]
 
-  task queries: [:scaled_weights, :building_blocks, :applications]
+  task queries: [:scaled_weights, :building_blocks, :applications, :alternatives]
 
   directory "#{exp_dir}/preprocessing"
   directory "#{exp_dir}/preprocessing/ch"
@@ -206,6 +206,7 @@ namespace "exp" do
   directory "#{exp_dir}/rphast_inc"
   directory "#{exp_dir}/bidir_features"
   directory "#{exp_dir}/applications"
+  directory "#{exp_dir}/alternatives"
 
   task preprocessing: ["code/compute_ch/build/compute_ch", "#{exp_dir}/preprocessing/ch", "#{exp_dir}/preprocessing/cch", "code/rust_road_router/lib/InertialFlowCutter/build/console"] + graphs.map { |g| g + 'lower_bound' } do
     graphs.each do |graph|
@@ -300,8 +301,15 @@ namespace "exp" do
       sh "cargo run --release --bin chpot_blocked -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin bidir_chpot_blocked -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
       sh "cargo run --release --bin bidir_chpot_turns -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
-      # sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{osm_ger} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
-      # sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_penalty -- #{dimacs_graph} > #{exp_dir}/applications/$(date --iso-8601=seconds).json"
+    end
+  end
+
+  task alternatives: static_graphs.map { |g| g + 'lower_bound_ch' } + static_graphs.map { |g| g + 'cch_perm' } + ["#{exp_dir}/alternatives"] do
+    Dir.chdir "code/rust_road_router" do
+      static_graphs.each do |graph|
[…]

### 2021-10-19 paper/ch_potentials.tex
add alternative results

@@
 However, their approach uses a simpler penalization schema.
-We also implement this variant with CH-Potentials but stick to the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative}.
+We also implement this variant with CH-Potentials but stick to the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative} and denote this approach as \emph{penalty routes}.
 This leads to some qualitative improvements compared to the results reported in~\cite{adgw-arrn-13}.
@@
 
+\subsubsection{Alternative Routes}
+
+Our implementation of the penalty method reproduces the qualitative results reported in~\cite{kobitzsch2015alternative} (for random queries on DIMACs Europe we on average achieve a total distance of 2.93, an average distance of 1.04, 5.57 decision edges and a success rate of 98.7\% with 6.7 iterations).
+The average running times to compute an alternative graph are 1.37\,s on DIMACs Europe and 0.85\,s on OSM Germany.
+These running times are roughly one order of magnitude slower than the results achieved by Kobitzsch.
+The simpler penalty routes approach is much faster and takes only 156\,ms on DIMACs Europe and 66\,ms on OSM Germany.
+This is two orders of magnitude faster than the running times reported in~\cite{adgw-arrn-13} which is not surprising since their implementation uses an unaccelerated Dijkstra.
+Due to the advanced penalization schema, our average qualitative results differ from~\cite{adgw-arrn-13}: the success rate is with 96.8\% slightly higher, the UBS with around 9\% slightly worse, the sharing is with 42\% somewhat higher but the local optimality, which was the weakest point in the original approach, with 21\% significantly better.
+Finding these alternatives takes only 1.44 iterations on average.
+This makes CH-Potential-based penalty routes a simple and feasible practical approach to compute alternative routes.
+
 \section{Conclusion}

### 2021-10-21 paper/ch_potentials.tex
integrate some comments from ben

@@
 Since it utilizes Contraction Hierarchies (CH), a hierarchical technique, we call our heuristic CH-Potentials.
-Our approach allows decoupling the supported extensions from the hierarchical speed-up technique.
+Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
 Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present an improved pruning criterion for symmetrical bidirectional A*.
@@
 Extending Dijkstra's algorithm to support these features is usually easy.
-Extending hierarchical speed-up techniques is also possible.
+Extending hierarchical speed-up techniques is often possible, too.
 However, the algorithm development is vastly more complex.
@@
 
-Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
+Our approach decouples the development of extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
 A* is a goal-directed variant of Dijkstra's algorithm.
@@
 A* uses a \emph{heuristic} to guide the search towards the goal.
-A heuristic is function that maps a node to an estimate of the remaining distance to the goal.
+A heuristic is a function that maps a node to an estimate of the remaining distance to the goal.
 A*'s running time crucially depends on how tight this estimate is.
@@
 The original paper only discussed unidirectional CH-Potentials.
-Here, we also discuss a bidirectional variant and also introduce an improved pruning criterion for symmetrical bidirectional A* which can also be utilized for other heuristic.
-Finally, we include additional applications of CH-Potentials, for example alternative routes, and provide a significantly expanded evaluation.
+In this journal version, we also discuss a bidirectional variant and also introduce an improved pruning criterion for symmetrical bidirectional A* which can also be utilized for other heuristic.
+Finally, we describe additional applications of CH-Potentials, for example alternative routes, and provide a significantly expanded evaluation.
 
@@
 In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
-After establishing preliminaries in~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and our algorithm to solve it efficiently in Section~\ref{sec:lazy-rphast}.
+After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and our algorithm to solve it efficiently in Section~\ref{sec:lazy-rphast}.
 CH-Potentials, our new distance estimation function is introduced in Section~\ref{sec:main-algo}.
@@
 We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
-Given nodes $s$ and $t$ we want to obtain the path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
+Given nodes $s$ and $t$ we want to obtain a path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
 We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
@@
 
-Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this point to point shortest path problem.
-It maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
+Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this shortest path problem.
+In this section, we recall the classical algorithm and establish some notation that we will employ in the remainder of this paper.
+Dijkstra's algorithm maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
 Initially, all distances are set to $\infty$.
@@
 Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
-Outgoing edges $uv$ are relaxed, i.e. $d[v] = \min(d[v], d[u] + w(uv))$.
+Outgoing edges $uv$ are relaxed, i.e. $d[v]$ is set to $\min(d[v], d[u] + w(uv))$.
 If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
@@
 
-\subsection{A*}\label{sec:a_star}
+\subsection{A* Algorithm}\label{sec:a_star}
 
@@
 A heuristic function is called \emph{feasible} if $w_{h_t} \geq 0$ for all edges.
+A* is guaranteed to compute an optimal path, if the heuristic is feasible.
+In this article, we therefore limit our discussion to feasible heuristics.
 
@@
 \includegraphics{fig/ch}
+% TODO nicer fig
 \caption{
@@
 In a CH, each node gets a unique rank representing its ``importance''.
-The intuition is that dead-ends are unimportant and at the bottom while highway bridges are very important and at the top.
+By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
+The intuition is that dead-ends are unimportant while highways or bridges are important.
 An edge goes \emph{up} when its tail node has lower rank than its head node.
@@
 See Figure~\ref{fig:ch} for a proof sketch.
-From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) exists.
+From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
 Thus, we can restrict our search to up-down paths in $G^+$.
@@
 In~\cite{bcrw-s-16}, it was proven that set of ancestors of a node in the elimination tree is equal to the nodes reachable in a CH search.
-This makes it possible to explore the CCH search space with a more efficient shortest path algorithm for DAGs:
-Traverse elimination tree from source and target to the root node and for each node relax all up-edges (reversed down-edges for the backward search).
-Because this algorithm does not require a queue, it is faster than a Dijkstra based CH query.
+This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
+For the forward search, start at the source and follow the elimination tree upward until the root is reached.
+All outgoing edges of encountered nodes are relaxed.
+The backward search works analogously but starts at the target.
+Because this algorithm does not require a queue, it is faster than a Dijkstra-based CH query.
 

### 2021-11-08 paper/ch_potentials.tex
wordsmithing

@@
 RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
-Combined with SSE extension it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.
+Implemented with SIMD instructions it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.
 
@@
 \KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
+\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
@@
 Algorithm~\ref{algo:pot} can be applied to CCH without any modifications.
-However, we can also exploit the elimination tree for the Lazy RPHAST algorithm.
+However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
 To compute the distance $D[x]$ of a node $x$ the distances of all upward neighbors must be final.
@@
 \KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[x]$: memoized distance final from $x$ to $t$, $\bot$ initially}
+\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
 \KwData{$P[x]$: parent node of $x$ in the elimination tree}
@@
 \end{algorithm2e}
+% TODO iterative
 
@@
 
-Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investing how well Lazy RPHAST performs in this setting appears to be an interesting question.
-SSE RPHAST starts by performing the selection phase is performed for the full target set.
+Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investigating the Lazy RPHAST performance in this setting seems worthwhile.
+SSE RPHAST starts by performing the selection phase on the full target set.
+% TODO flip directions?
 Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
@@
 The Bucket CH selection phase works as follows:
+% TODO better structure or local outline
+% TODO those are not shortest distances
+% TODO besser herleiten, wieso kein 2d array
 For each node $x$, we maintain an initially empty list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
@@
 The priority queue is ordered by importance instead of distance.
-Distance labels are buckets instead if scalar values.
+Distance labels are buckets instead of scalar values.
 The buckets of each target node are initialized with a single entry for the respective node with distance zero.
@@
 We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{sec:exp_lazy_rphast}.
+% TODO describe improved^2 Bucket CH
 
@@
 We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
-We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACs implementation challenge~\cite{DemetrescuGJ09}.
+We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACS implementation challenge~\cite{DemetrescuGJ09}.
 Additionally, we have three graphs with proprietary traffic predictions also provided by PTV.
@@
 
+% better wrap story branch
+
 \begin{figure}
@@
 
-To evaluate Lazy RPHAST in the incremental setting, we measure the elapsed time after $2^i$ distances were computed.
+To evaluate Lazy RPHAST in the incremental setting, we measure the elapsed running time after $2^i$ result distances were computed. % TODO word smithing "result distances"
+% TODO selection not included
 Figure~\ref{fig:lazy_rphast_inc} depicts the total elapsed time and the average running time to compute a single distance.
@@
 \caption{
-Running times of CH based Lazy RPHAST with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
+Running times of CH based Lazy RPHAST for many-to-one queries with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
 }\label{fig:many_to_one}
@@
 \caption{
-Running times of CCH based Lazy RPHAST with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
+Running times of CCH based Lazy RPHAST for many-to-one queries with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
 }\label{fig:many_to_one_cch}
@@
 This experiment uses the exact methodology described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to roughly relate our results to the performance of competing algorithms.
-When looking at total running times for one source target set pair, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}
-For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACs Europe.
+When looking at total running times for a single many-to-one query, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
+For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
 Lazy RPHAST (on a CH) takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
 The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
-This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster query phases.
-When computing distances from different source towards the same target set, RPHAST will outperform Lazy RPHAST.
+This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
+When computing distances from different sources towards the same target set, RPHAST will outperform Lazy RPHAST.
 However, whenever the target set may change between queries, Lazy RPHAST offers a simple and fast alternative.
@@
 The results are again at least competitive to the performance reported for SSE RPHAST in~\cite{delling_et_al:OASIcs:2011:3266}.
-The Bucket CH based first query phase takes between half of the time for $|B| = 2^{14}$ and a tenth of the time for $|B| = 2^{24}$.
-It is worth noting that contrary to SSE RPHAST, we did not explicitly utilize SSE instructions.
+% TODO konkrete zahlen?
+The Bucket CH based first query phase of our implementation takes between half of the total query time for $|B| = 2^{14}$ and a tenth of the time for $|B| = 2^{24}$.
+It is worth noting that contrary to SSE RPHAST, we did not explicitly utilize SIMD instructions.
 However, the code for Lazy RPHAST on an arrays of distances can be trivially vectorized by the compiler.
@@
 There, shortest distances are computed repeatedly from the same source to boundary nodes of a cell of nodes.
-On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code~\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
+On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
 We reran their improved experiments and include the results in our appendix.
 Briefly summarized, integrating Lazy RPHAST into this algorithm decreased running times by up to 15 times depending on the distribution of the target set for almost no effort.
+This demonstrates that Lazy RPHAST is not only useful as an A* potential but also applicable in other contexts.
 
@@
 This shows that CH-Potentials gracefully converges toward a CH in the $w_q = w_\ell$ special case.
-On DIMACs Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.
+On DIMACS Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.
 
@@
 
-Our implementation of the penalty method reproduces the qualitative results reported in~\cite{kobitzsch2015alternative} (for random queries on DIMACs Europe we on average achieve a total distance of 2.93, an average distance of 1.04, 5.57 decision edges and a success rate of 98.7\% with 6.7 iterations).
-The average running times to compute an alternative graph are 1.37\,s on DIMACs Europe and 0.85\,s on OSM Germany.
+Our implementation of the penalty method reproduces the qualitative results reported in~\cite{kobitzsch2015alternative} (for random queries on DIMACS Europe we on average achieve a total distance of 2.93, an average distance of 1.04, 5.57 decision edges and a success rate of 98.7\% with 6.7 iterations).
+The average running times to compute an alternative graph are 1.37\,s on DIMACS Europe and 0.85\,s on OSM Germany.
 These running times are roughly one order of magnitude slower than the results achieved by Kobitzsch.
-The simpler penalty routes approach is much faster and takes only 156\,ms on DIMACs Europe and 66\,ms on OSM Germany.
+The simpler penalty routes approach is much faster and takes only 156\,ms on DIMACS Europe and 66\,ms on OSM Germany.
 This is two orders of magnitude faster than the running times reported in~\cite{adgw-arrn-13} which is not surprising since their implementation uses an unaccelerated Dijkstra.

### 2021-11-08 eval/applications.py

@@ -20,7 +20,7 @@ def path_to_graph(path):
         'osm_europe': 'OSM Europe',
         'osm_ger': 'OSM Ger',
         'osm_ger_td': 'OSM Ger',
-        'europe': 'DIMACs Eur',
+        'europe': 'DIMACS Eur',
     }[[x for x in path.split('/') if x != ''][-1]]
 
 only_public = 'ONLY_PUBLIC' in os.environ
@@ -101,7 +101,7 @@ table = table.rename(index={
     cwd + '/data/osm_europe/': 'OSM Europe',
     cwd + '/data/osm_ger/': ger_name,
     cwd + '/data/osm_ger_td/': ger_name,
-    cwd + '/data/europe/': 'DIMACs Europe',
+    cwd + '/data/europe/': 'DIMACS Europe',
     'perfect': 'Unmodified ($w_q=w_\\ell$)',
     'no_highways': 'No Highways',
     'no_tunnels': 'No Tunnels',

### 2021-11-08 eval/bidir_features.py

@@ -14,7 +14,7 @@ data = [json.load(open(path)) for path in paths]
 
 queries = pd.DataFrame.from_records([{
     **algo,
-    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
+    'graph': { 'europe': 'DIMACS Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
     'experiment': exp['experiment'],
     'factor': exp['factor'],
     'probability': exp.get('probability', 1),

### 2021-11-08 eval/graphs.py

@@ -19,7 +19,7 @@ def path_to_graph(path):
         'osm_europe': 'OSM Europe',
         'osm_ger': 'OSM Germany',
         'osm_ger_td': 'OSM Germany',
-        'europe': 'DIMACs Europe',
+        'europe': 'DIMACS Europe',
     }[[x for x in path.split('/') if x != ''][-1]]
 
 graphs = pd.DataFrame.from_records([{
@@ -68,7 +68,7 @@ cch_ordering = pd.DataFrame.from_records([parse_flowcutter_partition_output(path
 table = graphs.groupby(['graph'])[['basic_customization_running_time_ms', 'contraction_running_time_ms', 'graph_build_running_time_ms',
     'perfect_customization_running_time_ms', 'respecting_running_time_ms', 'num_nodes', 'num_edges']].mean()
 
-table = table.reindex(['OSM Ger', 'DIMACs Europe', 'TDGer06', 'TDEur17', 'TDEur20'])
+table = table.reindex(['OSM Ger', 'DIMACS Europe', 'TDGer06', 'TDEur17', 'TDEur20'])
 
 table = table.join(ch_preprocessing.groupby('graph').mean()).join(cch_ordering.groupby('graph').mean())
 table['num_nodes'] = table['num_nodes'] / 1000000.0

### 2021-11-08 eval/rphast_comparison.py

@@ -21,7 +21,7 @@ queries = pd.DataFrame.from_records([{
     'experiment': exp['experiment'],
     'target_set_size_exp': exp['target_set_size_exp'],
     'ball_size_exp': exp['ball_size_exp'],
-    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
+    'graph': { 'europe': 'DIMACS Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
     }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs']])
 

### 2021-11-08 eval/rphast_inc.py

@@ -21,7 +21,7 @@ queries = pd.DataFrame.from_records([{
     'algo': algo['algo'],
     'experiment': exp['experiment'],
     'ball_size_exp': exp['ball_size_exp'],
-    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
+    'graph': { 'europe': 'DIMACS Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
     }
     for run in data for exp in run['experiments'] for algo in exp['algo_runs'] for progress in algo['progress_states']])
 

### 2021-11-08 eval/scaled_weights.py

@@ -19,14 +19,14 @@ data = [json.load(open(path)) for path in paths]
 queries = pd.DataFrame.from_records([{
     **algo,
     'weight_factor': exp['factor'],
-    'graph': { 'europe': 'DIMACs Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
+    'graph': { 'europe': 'DIMACS Europe', 'osm_ger': 'OSM Germany', 'osm_europe': 'OSM Europe' }[[x for x in run['args'][1].split('/') if x != ''][-1]],
     'num_queue_pops': algo['num_queue_pops'] + algo.get('core_search', {}).get('num_queue_pops', 0),
     'num_queue_pushs': algo['num_queue_pushs'] + algo.get('core_search', {}).get('num_queue_pushs', 0),
     'num_relaxed_arcs': algo['num_relaxed_arcs'] + algo.get('core_search', {}).get('num_relaxed_arcs', 0)
 } for run in data for exp in run['experiments'] for algo in exp['algo_runs'] if algo.get('algo') == 'Virtual Topocore Component Query'])
 
 plt.figure(figsize=(11,4))
-g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', hue_order=['OSM Germany', 'DIMACs Europe'], showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
+g = sns.boxplot(data=queries, x='weight_factor', y='running_time_ms', hue='graph', hue_order=['OSM Germany', 'DIMACS Europe'], showmeans=True, linewidth=0.8, flierprops=dict(marker='o', markerfacecolor='none', markeredgewidth=0.3))
 g.xaxis.set_major_locator(mpl.ticker.MultipleLocator(5))
 # g.xaxis.set_major_formatter(mpl.ticker.ScalarFormatter())
 g.set_yscale('log')

### 2021-11-08 paper/ch_potentials.tex
Small changes to text, added some questions

@@
 \usepackage[algo2e,vlined]{algorithm2e}
+\usepackage{todonotes}
 
@@
 Naively translating this approach to Lazy RPHAST would mean selecting one target after another and for each doing the full lazy exploration from all sources.
-However, this approach would explore the upward search space of all the sources over and over again.
-This would not be competitive.
+However, this approach would explore the upward search space of all the sources multiple times.
+This work seems redundant.
 
@@
 % TODO better structure or local outline
-% TODO those are not shortest distances
 % TODO besser herleiten, wieso kein 2d array
-For each node $x$, we maintain an initially empty list $B[x]$ of tuples $(t_i, dist(x,t_i))$, also called a \emph{bucket}.
+For each node $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$. 
+These pairs are called \emph{buckets}.
 These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
@@
 The running time of the penalty method is usually dominated by shortest path computations.
-Since this boils down to computing shortest paths on $G_{\ell}$ with a few increased edge weights this seems like a natural fit for A* with CH-Potentials.
-While the schema outlined above is conceptually simple, it leaves open quite a few details which may have a severe impact on performance.
-In this work, we do not develop our own variant of the penalty method but aim to reproduce the configuration reported by Kobitzsch in his dissertation~\cite{kobitzsch2015alternative}.
-To the best of our knowledge, this is the latest, most thoroughly engineered and evaluated iteration of the penalty method which also achieves the qualitatively best results.
-While some details remain unclear from the description in~\cite{kobitzsch2015alternative}, luckily we could obtain the original implementation from the authors to fill in the remaining details.
-Since we reproduce this exact configuration except for the shortest path algorithm, qualitative results should carry over and we can focus only on running times in our work.
-
-For sake of keeping this section self-contained, we now briefly reiterate the details of our configuration.
+The computations are performed on $G_{\ell}$ with a few increased edge weights.
+This setting seems like a natural fit for A* with CH-Potentials.
+While the alternative route schema outlined above is conceptually simple, it leaves open quite a few details which may have a strong impact on performance.
+In this work, we do not develop our own variant of the penalty method.
+Instead we aim to reproduce the configuration used by Kobitzsch in his dissertation~\cite{kobitzsch2015alternative}.
+To the best of our knowledge, this is the latest, most thoroughly engineered and evaluated iteration of the penalty method.
+It also achieves the qualitatively best results.
+Some details remain unclear from the description in~\cite{kobitzsch2015alternative}.
+Luckily, we could obtain the original implementation from the authors to resolve all open questions.
+As we use the same configuration, qualitative results are directly comparable.
+We can focus on comparing running times.
+
+To ease our exposition, we briefly reiterate the details of Kobitzsch's configuration.
 The stretch is limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \cdot \dist(u,v)$.
+\todo{Das bedeutet, dass der Alternativgraph z.B. Zyklenfrei sein muss. Sobald es einen Zyklus gibt kann ich $w(P)$ beliebig groß machen.}
 Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was on the shortest path in $k$ iterations its weight will be $w_q = w_{\ell}\cdot\psi^k$.
 We also penalize edges incident to shortest paths with an asymmetric additive rejoin penalty.
+\todo{incident = nur incomming edges oder auch outgoing? rejoin klingt nur nach incomming, incident klingt nach beiden.}
 An edge $uv$ where $u$ is on a shortest path but $v$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(s,u)}{\dist_{w_q}(s,t)}$.
-An edge $uv$ where $v$ is on a shortest path but $u$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(v,t)}{\dist_{w_q}(s,t)}$.
+Analogously, an edge $uv$ where $v$ is on a shortest path but $u$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(v,t)}{\dist_{w_q}(s,t)}$.
+\todo{Was ist $w_q$ in diesem Kontext?}
+\todo{Ich habe nicht verstanden was diese $\psi_r$ Formeln tun.}
 $\psi_r$ is set to 0.01.
@@
 
-In~\cite{adgw-arrn-13} Abraham et al. describe a variation of the penalty method to directly obtain alternative routes rather than an alternative graph:
-Just return the first path which shares less than 80\% with the original path and has stretch less than $1+\epsilon$.
+In~\cite{adgw-arrn-13} Abraham et al. describe a variation of the penalty method to directly obtain alternative routes rather than an alternative graph.
+They return the first path which shares less than 80\% with the original path and has stretch less than $1+\epsilon$.
 However, their approach uses a simpler penalization schema.

### 2021-11-08 paper/ch_potentials.tex
rework A* description

@@
 
-Since bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm, a bidirectional variant of A* also seems desirable.
-However, as pointed out in~\cite{gh-cspas-05}, this is not as straightforward as one might expect.
-One would like to use a forward potential $\overrightarrow{h}_t$ in the forward search and a backward potential $\overleftarrow{h}_s$ (computing distances in the reversed graph) in the reverse search.
-The problem is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
-Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
-This breaks the bidirectional Dijkstra stopping criterion.
-To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until their minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
-But this is the same stopping criterion that a unidirectional A* search has.
-Thus, this straightforward bidirectional A* performs twice the work than a unidirectional A*.
-This can be slightly improved by introducing pruning.
-Nodes for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e.
-their distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, can not contribute to a shorter path and need not to be added to the queue.
-This approach is denoted as the \emph{symmetric} bidirectional A*.
-
-To overcome this problem the authors of~\cite{gh-cspas-05} propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
+\todo{potential und heuristic geht hier bunt durcheinander.}
+
+On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
+A bidirectional variant of A* is therefore also desirable.
+However, as shown in~\cite{gh-cspas-05}, this is not straightforward.
+
+The straightforward idea is to use two potentials.
+$\overrightarrow{h}_t(x)$ is a lower bound of the distance from $x$ to $t$ and is used in the forward search.
+Similarly, $\overleftarrow{h}_s(x)$ is a lower bound of the distance from $s$ to $x$.
+It is used in the backward search.
+We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$.
+Similarly, $\overleftarrow{D}[x]$ is the tentative distance of the backward search at $x$.
+Further, we denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the key of the minimum elements in the forward respectively backward queue.
+Finally, we denote by $\mu$ the length of the shortest known path from $s$ to $t$.
+
+The problem is that these two potentials induce different reduced graphs\footnote{See Section~\ref{sec:a_star} for a discussion of reduced graphs.}.
+Thus, in the equivalent bidirectional search each direction would run on a different graph.
+This breaks the usual bidirectional stopping criterion.
+We know of no way better way than to run both searches until one of the two unidirectional stopping criteria is met.
+With such a weak stopping criterion, bidirectional search seems pointless.
+Luckily, having two potentials allows us to perform additional pruning.
+With this additional pruning the algorithm can overall be faster.
+
+The idea is to use information of the backward search during the pruning in the forward search and vice versa.
+Once $x$ is removed from the forward queue $\overrightarrow{D}[x]$ is the shortest path length from $s$ to $x$.
+Further, by construction $\overrightarrow{h}_t(x)$ is a lower bound on the shortest path length from $x$ to $t$.
+We therefore know that after $x$ is removed from the forward queue, that $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$ is a lower bound on the shortest path length from $s$ to $t$ via $x$.
+If this is above the shortest known path length from $s$ to $t$, we can prune $x$.
+As the forward queue orders elements by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$, this pruning rule is not useful in the forward search.
+However, it can be used in the backward search.
+The backward search can be pruned at $x$, if $x$ was removed from the forward queue, and $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)\ge \mu$.
+Similarly, the forward search can be pruned, if $\overleftarrow{h}_s(x) + \overleftarrow{D}[x]\ge \mu$ and $x$ was removed from the backward queue.
+Using two searches and these two pruning rules is called the \emph{symmetric bidirectional A*}.
+
+To obtain a better stopping criterion, the authors of~\cite{gh-cspas-05} propose the \emph{average potential}.
+This is a combined heuristic, which is good for both directions.
+Both searches run on the same reduced graph which enables an efficient stopping criterion.
 The reduced graph of the average potential is the graph with the average weights of the both reduced graphs.
-The potential function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$ (and $-h_{\overline{st}}(x)$ for the backward search).
+The potential function used in the forward search is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$.
+The backward search uses $-h_{\overline{st}}(x)$.
 This potential allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
@@
 To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
-Still, it has two downsides.
-First, evaluating the average potential requires evaluating the original heuristics for both directions.
-This can cause some overhead depending on the heuristics.
-Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
-
-We propose an improved pruning criterion for the symmetric approach.
-It takes the search progress (in form of the growing minimal queue keys) into account to tighten the pruning with increasing search progress and compensates for the worse stopping criterion.
-As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
-Recall that the queue is ordered by $D[x] + h(x)$.
+Unfortunately, it has two downsides.
+The first downside is that the better stopping criterion comes at the cost of worse potentials.
+The average potential is a worse estimate than $\overrightarrow{h}_t$ in the forward search.
+It is also worse than $\overleftarrow{h}_s$ in the backward search.
+The second downside is that evaluating the average potential requires evaluating $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
+This means that the running time to evaluate both potentials is necessary
+
+To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and improve upon it.
+We do away with the average potential.
+Our insight consists of improving the pruning rules.
+In the backward search, we use $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$ to lower bound the minimum length of a path via $x$.
+If this is above $\mu$, we can prune the search at $x$.
+We improve this estimate and obtain a tighter pruning rule.
+\todo{Hier wurde ich abgehängt. Ich verstehe die neue Lower Bound nicht.}
 We use the minimal queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$.
@@
 
-For both average and symmetric bidirectional A*, we always alternate between settling a node form the forward and the backward search.
-In our experiments, we also evaluate choosing the direction by the smaller minimal queue key.
-Even though this appears to better fit together with the stopping criterion for the symmetric approach, our experiments show that it is never beneficial in terms of running time (see Table~\ref{tab:bidir_switching}).
+Unless stated differently, for all bidirectional A* variants, we always alternate between removing a node from the forward and the backward queues.
+In our experiments, we also evaluate expanding the search with the smaller queue element.
+This sounds sensible but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
 

### 2021-11-10 paper/ch_potentials.tex
rewrite bidir dijkstra section

--- 2021-11-08 paper/ch_potentials.tex
@@
 We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
-Given nodes $s$ and $t$ we want to obtain a path $P=(s=v_0,\dots,t=v_k)$ of minimal weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
+Given nodes $s$ and $t$ we want to obtain a path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
 We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
@@
 \subsection{Dijkstra's Algorithm}
+
+% TODO Capital letters for arrays?
 
@@
 Typically, the searches are interleaved by alternating settling a node from each direction.
-Another common approach is to always settle a node from the direction with the smaller minimal queue key.
+Another common approach is to always settle a node from the direction with the smaller minimum queue key.
 Theoretically any direction interleaving strategy is possible.
@@
 
-Since bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm, a bidirectional variant of A* also seems desirable.
-However, as pointed out in~\cite{gh-cspas-05}, this is not as straightforward as one might expect.
-One would like to use a forward potential $\overrightarrow{h}_t$ in the forward search and a backward potential $\overleftarrow{h}_s$ (computing distances in the reversed graph) in the reverse search.
-The problem is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
+On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
+A bidirectional variant of A* therefore appears also desirable.
+However, as shown in~\cite{gh-cspas-05}, this is not as simple as it seems.
+
+The straightforward idea would be to use two heuristics.
+The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$ where $\overrightarrow{h}_t(x)$ estimates the distance from $x$ to $t$.
+Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$ where $\overleftarrow{h}_s(x)$ is an estimate of the distance from $s$ to $x$.
+% We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
+We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.
+
+The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
 Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
-This breaks the bidirectional Dijkstra stopping criterion.
-To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until their minimum queue key $k$ becomes greater than the best known total tentative distance $\mu$.
-But this is the same stopping criterion that a unidirectional A* search has.
-Thus, this straightforward bidirectional A* performs twice the work than a unidirectional A*.
-This can be slightly improved by introducing pruning.
-Nodes for which $\overrightarrow{D}[x] + h(x) > \mu$ holds, i.e.
-their distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, can not contribute to a shorter path and need not to be added to the queue.
-This approach is denoted as the \emph{symmetric} bidirectional A*.
-
-To overcome this problem the authors of~\cite{gh-cspas-05} propose the \emph{average} potential, a combined heuristic, which is good for both directions but also makes both searches run on the same reduced graph and thus enables an efficient stopping criterion.
-The reduced graph of the average potential is the graph with the average weights of the both reduced graphs.
-The potential function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$ (and $-h_{\overline{st}}(x)$ for the backward search).
-This potential allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
-Additionally, the pruning from the symmetric approach can be utilized with the original potentials.
+This breaks the usual bidirectional Dijkstra stopping criterion.
+To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
+Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
+The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.
+
+To obtain a better stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
+It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined heuristic $h_{\overline{st}}$, which is good for both directions.
+Both searches run on the same reduced graph which enables an efficient stopping criterion.
+The reduced graph of the average potential is the graph with the average weights of the reduced graphs of $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
+The heuristic function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$.
+The backward search uses $-h_{\overline{st}}(x)$.
+This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
 To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
-Still, it has two downsides.
-First, evaluating the average potential requires evaluating the original heuristics for both directions.
-This can cause some overhead depending on the heuristics.
-Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
-
-We propose an improved pruning criterion for the symmetric approach.
-It takes the search progress (in form of the growing minimal queue keys) into account to tighten the pruning with increasing search progress and compensates for the worse stopping criterion.
-As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
-Recall that the queue is ordered by $D[x] + h(x)$.
-We use the minimal queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$.
-If a node $x$ was not yet settled by backward search then $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
+Unfortunately, it has two downsides.
+First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
+This makes it more expensive in terms of running time.
+% Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
+Second, the better stopping criterion comes at the cost of worse estimates for each direction on its own.
+The average potential estimates $\dist(x,t)$ less accurately than $\overrightarrow{h}_t(x)$ and $\dist(s,x)$ less accurately than $\overleftarrow{h}_s(x)$.
+The second downside can be partially mitigated through pruning with the original heuristics.
+When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(xy) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
+the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.
+The pruning for the backward search works analogously.
+
+To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
+We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.
+Recall that queues are ordered by $D[x] + h(x)$.
+If a node $x$ was not yet settled by backward search then $\dist(x,t) + \overleftarrow{h}_s(x) \geq \overleftarrow{k}$, and equivalently $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
 Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
 The pruning for the backwards search works analogously.
+As the pruning becomes tighter as the searches progress, it compensates for the worse stopping criterion.
 Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
-
-For both average and symmetric bidirectional A*, we always alternate between settling a node form the forward and the backward search.
-In our experiments, we also evaluate choosing the direction by the smaller minimal queue key.
-Even though this appears to better fit together with the stopping criterion for the symmetric approach, our experiments show that it is never beneficial in terms of running time (see Table~\ref{tab:bidir_switching}).
+As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
+
+Unless stated differently, for all bidirectional A* variants, we always alternate between removing a node from the forward and the backward queues.
+In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
+This sounds sensible but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
 
@@
 The results clearly show that alternating the directions is always superior.
-Selecting the direction by minimal queue key may lead to huge imbalances in the progress of the searches.
+Selecting the direction by minimum queue key may lead to huge imbalances in the progress of the searches.
 This causes the searches to meet later and the total search space to grow significantly.

### 2021-11-10 paper/ch_potentials.tex
add utf8 inputenc
queue besser in ch query besser beschreiben

--- 2021-11-08 paper/ch_potentials.tex
@@
 \documentclass[manuscript,review]{acmart}
+
+\usepackage[utf8]{inputenc}
 
@@
 \KwData{$B[x]$: tentative distance from $x$ to target $t$}
-\KwData{Min. priority queue $Q$, also called open list}
+\KwData{Minimum priority queue $Q$, also called open list, ordered by tentative distances}
 $B[x] \leftarrow +\infty$ for all $x\neq t$;

### 2021-11-10 paper/ch_potentials.tex
schleifen über CH Kanten vereinheitlichen
add Many-to-Many description for small k

@@
 	$y\leftarrow$ pop minimum element from $Q$\;
-	\For{$xy$ is down-edge in $G^+_\ell$}{
+	\For{all down-edges $xy$ in $G^+_\ell$}{
 		\If{$B[x] > w_\ell(xy) + B[y]$}{
@@
 \For{all CH levels $L$ from most to least important}{
-	\For{all up edges $xy$ in $G^+_\ell$ with $x$ in $L$}{
+	\For{all up-edges $xy$ in $G^+_\ell$ with $x$ in $L$}{
 		\If{$P[x] < P[y] + w_\ell(xy)$}{
@@
 		$D[x]\leftarrow B[x]$\;
-		\For{all up edges $xy$ in $G^+_\ell$}{
+		\For{all up-edges $xy$ in $G^+_\ell$}{
                         $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
@@
 		$D[x]\leftarrow B[x]$\;
-		\For{all up edges $xy$ in $G^+_\ell$}{
+		\For{all up-edges $xy$ in $G^+_\ell$}{
                         $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
@@
 
+RPHAST was originally developed to answer many-to-many queries.
+It was not designed to be used in conjunction with A*.
+Investigating how Lazy RPHAST performs in the many-to-many setting is therefore interesting.
+In this setting, a set of source and target nodes are given.
+The objective is to compute the distances between all pairs of source and target nodes.
+In this section, we described how Lazy RPHAST can be used to solve this many-to-many problem.
+Our experiments will show, that the achieved performance is comparable to the original RPHAST algorithm.
+
+To solve the many-to-many problem, RPHAST uses SIMD vector instructions.
+Let $k$ denote the number of elements in one vector.
+RPHAST computes the distances from all source nodes to $k$ target nodes in one run.
+If the target set contains more than $k$ nodes, then the target set is split into smaller sets of $k$ elements\footnote{If the number of targets is not divisible by $k$, the set is increased by adding additional target nodes.}.
+The algorithm is then run for each of these smaller sets.
+Each RPHAST run performs vectorized variants of the selection and query phases.\todo{Is query phase definiert?}
+
+We translate this setup to the Lazy RPHAST setting as follows.
+First, a set of $k$ target nodes need to be fixed.
+A vectorized selection phase is then performed for this fixed target set.
+After the selection, the source nodes are iteratively revealed.
+Once a source node is revealed, the algorithm must compute the distances to all $k$ targets.
+Only after the distances are computed, the next source node is revealed.
+
+The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:pot}.
+We describe vectorized variants of these algorithms.
+We replace the $B$ and $D$ arrays of scalars by arrays of $k$-dimensional vectors of scalars.
+To clarify which entities are SIMD vectors, we denote the vectorized arrays by $\hat{B}$ and $\hat{D}$.
+$\hat{D}[s]$ is an SIMD vector $\hat{v}$ of length $k$.
+The $i$-th element $\hat{v}[i]$ of $\hat{v}$ will contain the distance from $s$ to $i$-th selected target.
+We can also denote this distance using $\hat{D}[s][i]$.
+All operations on SIMD vectors are component-wise.
+Adding a scalar to an SIMD vector consists of adding the scalar to every element of the SIMD vector.
+
+\todo{Rechtschreibung: down edge vs down-edge und up edge vs up-edge. Geht beides aber bitte konsistent.}
+
+\begin{algorithm2e}
+\KwIn{$\hat{T}$: SIMD vector of $k$ target nodes}
+\KwData{$\hat{B}[x][i]$: tentative distance from $x$ to target $\hat{T}[i]$}
+\KwData{Minimum priority queue $Q$, also called open list, ordered by node rank}
+Make $Q$ an empty queue\;
+$\hat{B}[x][i] \leftarrow +\infty$ for all $x$ and $i$\;
+\For{all $i\in \{0\ldots k-1\}$}{
+  $\hat{B}[\hat{T}[i]][i] \leftarrow 0$\;
+  Add $\hat{T}[i]$ to $Q$\;
+}
+\While{not $Q$ empty}{
+  $y\leftarrow$ pop minimum element from $Q$\;
+  \For{all down-edges $xy$  in $G^+_\ell$}{
+    $\hat{B}[x] \leftarrow \min\{\hat{B}[x], \hat{B}[y] + w_\ell(xy)\}$\;
+    Add $x$ to $Q$ if not in $Q$\;
+  }
+}
+\caption{Vectorized Lazy RPHAST selection}
+\label{algo:simd-ch-backward}
+\end{algorithm2e}
+
+Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
+In the original algorithm, the nodes are ordered by distance.
+In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
+To resolve this problem, we ordered the nodes in the queue not by distance but by CH rank.
+The CH rank is the position of a node in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
+All other modifications are straight-forward.
+The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
+Algorithm~\ref{algo:pot} can be vectorized by just replacing the arrays by their SIMD variants.
+The result is illustrated in Algorithm~\ref{algo:simd-pot}.
+
+\begin{algorithm2e}
+\KwData{$\hat{B}[x][i]$: tentative distance from $x$ to $\hat{T}[i]$ as computed by Algorithm~\ref{algo:simd-ch-backward}}
+\KwData{$\hat{D}[x][i]$: memoized distance from $x$ to $\hat{T}[i]$, $\bot$ initially}
+\SetKwFunction{VecDist}{SIMDComputeAndMemoizeDist}
+\SetKwProg{Fn}{Function}{:}{}
+\Fn{\VecDist{$x$}}{
+  \If{$\hat{D}[x] = \bot$}{
+    $\hat{D}[x]\leftarrow \hat{B}[x]$\;
+    \For{all up-edges $xy$ in $G^+_\ell$}{
+      $\hat{D}[x]\leftarrow\min\{\hat{D}[x],w_\ell(xy)+\VecDist(y)\}$\;
+    }
+  }
+  \Return{$\hat{D}[x]$}\;
+}
+\caption{Vectorized Lazy RPHAST algorithm}
+\label{algo:simd-pot}
+\end{algorithm2e}
+
+\todo{Hier beginnt der ältere Text}
 Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investigating the Lazy RPHAST performance in this setting seems worthwhile.

### 2021-11-10 paper/ch_potentials.tex
changes to A* text

@@
 On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
-A bidirectional variant of A* therefore appears also desirable.
-However, as shown in~\cite{gh-cspas-05}, this is not as simple as it seems.
-
-The straightforward idea would be to use two heuristics.
-The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$ where $\overrightarrow{h}_t(x)$ estimates the distance from $x$ to $t$.
-Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$ where $\overleftarrow{h}_s(x)$ is an estimate of the distance from $s$ to $x$.
+A bidirectional variant of A* therefore is therefore also desirable.
+However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straight forward.
+We revisit bidirectional A* and propose an alternative approach.
+Our experiments show that it is competitive with the solution described by~\cite{gh-cspas-05}.
+
+The straightforward idea is to use two heuristics $\overrightarrow{h}_t(x)$ and $\overleftarrow{h}_s(x)$.
+The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$, where $\overrightarrow{h}_t(x)$ estimates the distance $\dist(x,t)$ from $x$ to $t$.
+Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$, where $\overleftarrow{h}_s(x)$ is an estimate of the distance $\dist(s,x)$ from $s$ to $x$.
 % We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
@@
 Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
-The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.
-
-To obtain a better stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
-It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined heuristic $h_{\overline{st}}$, which is good for both directions.
-Both searches run on the same reduced graph which enables an efficient stopping criterion.
-The reduced graph of the average potential is the graph with the average weights of the reduced graphs of $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
-The heuristic function which induces this average reduced graph is $h_{\overline{st}}(x) := \frac{\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x)}{2}$.
+This is not a useful algorithm.
+However, it can serve as basis for further algorithmic refinements.
+The authors of~\cite{gh-cspas-05} therefore refer to this approach as \emph{symmetric} bidirectional A*.
+
+To obtain a bidirectional stopping criterion, the \emph{average potential} is proposed~\cite{gh-cspas-05}.
+It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
+Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
+The forward search uses $h_{\overline{st}}(x)$ as heuristic.
 The backward search uses $-h_{\overline{st}}(x)$.
+This results in both searches running on the same reduced graph.
+The weights of the combined reduced graph corresponds to the averaged weights of the individual reduced graphs.\todo{Ist diese Aussage korrekt?}
 This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
-To the best of our knowledge, this method is the fastest known variant of bidirectional A*.
+To the best of our knowledge, using this stopping criterion yields the fastest known bidirectional A* variant.
 Unfortunately, it has two downsides.
 First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
-This makes it more expensive in terms of running time.
-% Second, for each direction on its own, the average potential is a worse heuristic than the original heuristic.
-Second, the better stopping criterion comes at the cost of worse estimates for each direction on its own.
-The average potential estimates $\dist(x,t)$ less accurately than $\overrightarrow{h}_t(x)$ and $\dist(s,x)$ less accurately than $\overleftarrow{h}_s(x)$.
-The second downside can be partially mitigated through pruning with the original heuristics.
-When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(xy) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
-the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.
-The pruning for the backward search works analogously.
-
+Evaluating the average heuristic is therefore slower than evaluating just one the composing heuristics.
+Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
+$h_{\overline{st}}(x)$ is a worse estimate for $\dist(x,t)$ than $\overrightarrow{h}_t$.
+Similarly, $-h_{\overline{st}}(x)$ is a worse estimate for $\dist(s,x)$ than $\overleftarrow{h}_s$.
+The second downside can be partially mitigated through pruning with the composing heuristics.
+When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(x,y) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
+the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.\todo{does not have to be inserted?
+needs not klingt komisch...
+aber English ist nicht meine Stärke.
+:) }
+The pruning rule for the backward search is analogous.
+
+\todo{$w(xy)$ vs $w(x,y)$}
 To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
+We do not use the average heuristic and use the unidirectional stopping criteria.
+We describe the idea for the forward search.
+The pruning rule for the backward search is analogous.
+The central idea consists of using information from the backward search to prune edges in the forward search.
+Let $xy$ be an edge that we relax in the forward search.
+Before pushing $y$ into the queue, we apply the pruning rule.
+If we can prove that every path using $xy$ is at least as long as the shortest known path length $\mu$, then we do not have to push $y$.
+We therefore want to obtain a lower bound for $\dist(s,x)+w(x,y)+\dist(y,t)$.
+As $x$ was settled, $\overrightarrow{D}[x]$ contains the shortest path length $\dist(s,x)$, i.e., $\dist(s,x)=\overrightarrow{D}[x]$.
+$w(x,y)$ is also exactly known as it's just an edge weight.
+It remains to lower bound $\dist(y,t)$.
+Nodes are removed from the backward queue ordered by $\overrightarrow{h}(y) + \dist(y,t)$.
+\todo{$\ge$ vs $\geq$ und $\le$ vs $\leq$, macht das einen Unterschied?}
+If $y$ was not yet removed from the backward queue, we know that $\overrightarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
+This gives us the required lower bound $\dist(y,t) \geq \overleftarrow{k}-\overrightarrow{h}(y)$.
+$y$ does not have to be pushed, if it was not settled by the backward search and $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{k}-\overrightarrow{h}(y)\ge \mu$.
+This does not mean that $y$ will never be pushed into the forward queue.
+It is possible that another edge $zy$ exists for which the pruning rule cannot be applied.
+
+\todo{Latex mag overrightarrow nicht in todos... Ich packe die Frage deswegen inline in den Text.}
+Kann man sinnvoll prunen wenn $y$ von der Rückwärtssuche gesettled wurde?
+Z.b.
+Kante $xy$ prunen wenn $x$ forward gestelled ist und $y$ backward gesettled ist und $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{D}[y]\ge \mu$ erfüllt ist.
+
+\todo{Der folgende Paragraph ist die Vorgängerversion vom Paragraph davor. Wenn der neue Paragraph besser ist, dann kann dieser weg.}
 We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.

### 2021-11-11 paper/ch_potentials.tex
changes to A* text

@@
 On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
-A bidirectional variant of A* therefore is therefore also desirable.
+A bidirectional variant of A* therefore also seems desirable.
 However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straight forward.
@@
 It combines a forward and a backward heuristic $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$ into a combined \emph{average heuristic} $h_{\overline{st}}$.
+The idea is to use a common reduced graph, whose weights are the average weights of the individual reduced graphs.
+In this case, both searches run on the same graph.
+This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
 Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
@@
 The backward search uses $-h_{\overline{st}}(x)$.
-This results in both searches running on the same reduced graph.
-The weights of the combined reduced graph corresponds to the averaged weights of the individual reduced graphs.\todo{Ist diese Aussage korrekt?}
-This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
-To the best of our knowledge, using this stopping criterion yields the fastest known bidirectional A* variant.
-Unfortunately, it has two downsides.
+Unfortunately, average potentials have two downsides.
 First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
@@
 To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
-We do not use the average heuristic and use the unidirectional stopping criteria.
 We describe the idea for the forward search.
@@
 The central idea consists of using information from the backward search to prune edges in the forward search.
+We do not use the average heuristic.
+Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
+Such a pruning rule will eventually prune all remaining branches and stop the search.
+Formally, we use the undirectional stopping criteria.
+
 Let $xy$ be an edge that we relax in the forward search.

### 2021-11-11 paper/ch_potentials.tex
A* text changes
wordsmithing on bidir A*

@@
 A bidirectional variant of A* therefore also seems desirable.
-However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straight forward.
+However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straightforward.
 We revisit bidirectional A* and propose an alternative approach.
@@
 Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$, where $\overleftarrow{h}_s(x)$ is an estimate of the distance $\dist(s,x)$ from $s$ to $x$.
-% We denote by $\overrightarrow{D}[x]$ the tentative distance of the forward search at $x$ and similarly, by $\overleftarrow{D}[x]$ the tentative distance in the backward search.
 We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.
@@
 To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
-Thus, this straightforward bidirectional A* performs roughly twice the work of a unidirectional A*.
-This is not a useful algorithm.
+The forward search can skip nodes already settled by the backward search and vice versa.
+However, this straightforward bidirectional A* still performs more work than a unidirectional A*.
+Thus, on its own it is not a useful algorithm.
 However, it can serve as basis for further algorithmic refinements.
-The authors of~\cite{gh-cspas-05} therefore refer to this approach as \emph{symmetric} bidirectional A*.
+The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.
 
@@
 The idea is to use a common reduced graph, whose weights are the average weights of the individual reduced graphs.
-In this case, both searches run on the same graph.
+Both searches run on the same common reduced graph.
 This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
 Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
-The forward search uses $h_{\overline{st}}(x)$ as heuristic.
+The forward search uses $h_{\overline{st}}(x)$ as its heuristic.
 The backward search uses $-h_{\overline{st}}(x)$.
@@
 First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
-Evaluating the average heuristic is therefore slower than evaluating just one the composing heuristics.
+Evaluating the average heuristic is therefore slower than evaluating just one of the composing heuristics.
 Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
@@
 When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(x,y) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
-the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, $y$ needs not to be inserted into the queue.\todo{does not have to be inserted?
-needs not klingt komisch...
-aber English ist nicht meine Stärke.
-:) }
+the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.
 The pruning rule for the backward search is analogous.
 
-\todo{$w(xy)$ vs $w(x,y)$}
 To avoid the downsides of the average potential, we revisit symmetric bidirectional A* and propose a new pruning criterion.
@@
 Such a pruning rule will eventually prune all remaining branches and stop the search.
-Formally, we use the undirectional stopping criteria.
+The stopping criteria for each direction remain the same (unidirectional) as before.
+However, often the search stops early because the queues are empty.
 
 Let $xy$ be an edge that we relax in the forward search.
-Before pushing $y$ into the queue, we apply the pruning rule.
+Before pushing $y$ into the queue, we apply the new pruning rule.
 If we can prove that every path using $xy$ is at least as long as the shortest known path length $\mu$, then we do not have to push $y$.
-We therefore want to obtain a lower bound for $\dist(s,x)+w(x,y)+\dist(y,t)$.
+We therefore want to obtain a lower bound for $\dist(s,x)+w(xy)+\dist(y,t)$.
 As $x$ was settled, $\overrightarrow{D}[x]$ contains the shortest path length $\dist(s,x)$, i.e., $\dist(s,x)=\overrightarrow{D}[x]$.
-$w(x,y)$ is also exactly known as it's just an edge weight.
+$w(xy)$ is also exactly known as it's just an edge weight.
 It remains to lower bound $\dist(y,t)$.
-Nodes are removed from the backward queue ordered by $\overrightarrow{h}(y) + \dist(y,t)$.
-\todo{$\ge$ vs $\geq$ und $\le$ vs $\leq$, macht das einen Unterschied?}
-If $y$ was not yet removed from the backward queue, we know that $\overrightarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
-This gives us the required lower bound $\dist(y,t) \geq \overleftarrow{k}-\overrightarrow{h}(y)$.
-$y$ does not have to be pushed, if it was not settled by the backward search and $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{k}-\overrightarrow{h}(y)\ge \mu$.
-This does not mean that $y$ will never be pushed into the forward queue.
-It is possible that another edge $zy$ exists for which the pruning rule cannot be applied.
-
-\todo{Latex mag overrightarrow nicht in todos... Ich packe die Frage deswegen inline in den Text.}
-Kann man sinnvoll prunen wenn $y$ von der Rückwärtssuche gesettled wurde?
-Z.b.
-Kante $xy$ prunen wenn $x$ forward gestelled ist und $y$ backward gesettled ist und $\overrightarrow{D}[x]+w(x,y)+\overleftarrow{D}[y]\ge \mu$ erfüllt ist.
-
-\todo{Der folgende Paragraph ist die Vorgängerversion vom Paragraph davor. Wenn der neue Paragraph besser ist, dann kann dieser weg.}
-We use the minimum queue key $\overleftarrow{k}$ of the backward search to obtain a tighter estimate of $\dist(x,t)$ for pruning in the forward search.
-Recall that queues are ordered by $D[x] + h(x)$.
-If a node $x$ was not yet settled by backward search then $\dist(x,t) + \overleftarrow{h}_s(x) \geq \overleftarrow{k}$, and equivalently $\dist(x,t) \geq \overleftarrow{k} - \overleftarrow{h}_s(x)$.
-Thus, in the forward search if $\overrightarrow{D}[x] + \overleftarrow{k} - \overleftarrow{h}_s(x) > \mu$, we do not need to push $x$ into the queue.
-The pruning for the backwards search works analogously.
-As the pruning becomes tighter as the searches progress, it compensates for the worse stopping criterion.
-Since this pruning is only used when $\mu > \infty$, for the greatest part of the search each direction only has to evaluate its own heuristic.
-As our experiments show, this makes the simpler symmetric approach competitive to the average potential.
+Nodes are removed from the backward queue ordered by $\overleftarrow{h}(y) + \dist(y,t)$.
+If $y$ was not yet removed from the backward queue, we know that $\overleftarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
+This gives us the required lower bound, i.e. $\dist(y,t) \geq \overleftarrow{k}-\overleftarrow{h}(y)$.
+Thus, $y$ does not have to be pushed if $\overrightarrow{D}[x]+w(xy)+\overleftarrow{k}-\overleftarrow{h}(y)\ge \mu$.
+The node $y$ might still be pushed into the queue when there is another edge $zy$ for which the pruning is not possible.
+Checking the pruning rule requires evaluating the backward heuristic.
+However, pruning is only possible once the searches have met, i.e. $\mu < \infty$.
+Before that, each direction only has to evaluate its own heuristic.
+Thus, our pruning improves on both downsides of the average potential.
 

### 2021-11-11 paper/ch_potentials.tex
more wordsmithing
iterative cch pot and minor notation adjustments

@@
 
-% TODO Capital letters for arrays?
-
 Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this shortest path problem.
 In this section, we recall the classical algorithm and establish some notation that we will employ in the remainder of this paper.
-Dijkstra's algorithm maintains an array of tentative distances $d$ and a priority queue of nodes ordered by increasing distance from $s$.
+Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue of nodes ordered by increasing distance from $s$.
 Initially, all distances are set to $\infty$.
-The start node distance $d[s]$ is set to zero and the queue initialized with $s$.
+The start node distance $D[s]$ is set to zero and the queue initialized with $s$.
 Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
-Outgoing edges $uv$ are relaxed, i.e. $d[v]$ is set to $\min(d[v], d[u] + w(uv))$.
+Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
 If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
-Once $t$ was settled, $d[t]$ contains the distance between $s$ and $t$.
+Once $t$ was settled, $D[t]$ contains the distance between $s$ and $t$.
 
@@
 \KwData{$P[x]$: parent node of $x$ in the elimination tree}
+\KwData{$S$: stack with nodes to compute distances}
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
+\SetKwFunction{Push}{push}
+\SetKwFunction{Pop}{pop}
+\SetKw{Break}{break}
 \SetKwProg{Fn}{Function}{:}{}
 \Fn{\Dist{$x$}}{
-	\If{$D[x] = \bot$}{
-		\If{$P[x] \neq \bot$ and $D[P[x]] = \bot$}{
-			$\Dist(P[x])$\;
-		}
-		$D[x]\leftarrow B[x]$\;
-		\For{all up-edges $xy$ in $G^+_\ell$}{
-                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+D[y]\}$\;
+  $y \gets x$\;
+	\While{$D[y] = \bot$}{
+    $S$.\Push{$y$}\;
+		\If{$P[y] = \bot$}{
+      \Break\;
+		}
+    $y \gets P[y]$
+  }
+  \While{$S$ not empty}{
+    $y \gets S$.\Pop{}\;
+		$D[y]\gets B[y]$\;
+		\For{all up-edges $yz$ in $G^+_\ell$}{
+                        $D[y]\gets\min\{D[y],w_\ell(yz)+D[z]\}$\;
 		}
@@
 \end{algorithm2e}
-% TODO iterative
 
@@
 The forward search can skip nodes already settled by the backward search and vice versa.
-However, this straightforward bidirectional A* still performs more work than a unidirectional A*.
+However, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
 Thus, on its own it is not a useful algorithm.
@@
 The stopping criteria for each direction remain the same (unidirectional) as before.
-However, often the search stops early because the queues are empty.
+However, usually the search stops early because the queues are empty.
 

### 2021-11-12 paper/ch_potentials.tex
remove many-to-many, kNN CCH and alternative routes. Shorten, clean up and improve Lazy RPHAST eval

@@
 Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
-Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present an improved pruning criterion for symmetrical bidirectional A*.
+Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
 An extensive experimental study confirms the practicality of our approach for many applications.
@@
 In this article we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in much greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
-Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
+% Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
 The original paper only discussed unidirectional CH-Potentials.
 In this journal version, we also discuss a bidirectional variant and also introduce an improved pruning criterion for symmetrical bidirectional A* which can also be utilized for other heuristic.
-Finally, we describe additional applications of CH-Potentials, for example alternative routes, and provide a significantly expanded evaluation.
+We also include a significantly expanded evaluation.
 
@@
 CH-Potentials, our new distance estimation function is introduced in Section~\ref{sec:main-algo}.
-There we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
-Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
+There, we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
+% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
 In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
@@
 
-Another feature which has received quite some attention is the computation alternative routes~\cite{bdgs-argrn-11,adgw-arrn-13}.
-Even formally defining what makes a good alternative route is non-trivial~\cite{bdgs-argrn-11}.
-A variety of paradigms exists.
-Competitive approaches require careful adjustment of their respective hierarchical speedup technique for the targeted paradigm.
-In~\cite{krs-eepma-13}, the CRP customization is utilized to repeatedly compute shortest paths on graphs with slightly modified weights.
-In~\cite{k-hdara-13}, partial shortest path trees are extracted from CH search spaces to find via-node candidates for alternative routes.
-In~\cite{barth2019alternative}, a multi-criteria CH is extended to compute alternative routes with regards to multiple edge weight functions.
+% Another feature which has received quite some attention is the computation alternative routes~\cite{bdgs-argrn-11,adgw-arrn-13}.
+% Even formally defining what makes a good alternative route is non-trivial~\cite{bdgs-argrn-11}.
+% A variety of paradigms exists.
+% Competitive approaches require careful adjustment of their respective hierarchical speedup technique for the targeted paradigm.
+% In~\cite{krs-eepma-13}, the CRP customization is utilized to repeatedly compute shortest paths on graphs with slightly modified weights.
+% In~\cite{k-hdara-13}, partial shortest path trees are extracted from CH search spaces to find via-node candidates for alternative routes.
+% In~\cite{barth2019alternative}, a multi-criteria CH is extended to compute alternative routes with regards to multiple edge weight functions.
 
@@
 
-\subsection{Many-to-Many with Lazy RPHAST}
-
-RPHAST was originally developed to answer many-to-many queries.
-It was not designed to be used in conjunction with A*.
-Investigating how Lazy RPHAST performs in the many-to-many setting is therefore interesting.
-In this setting, a set of source and target nodes are given.
-The objective is to compute the distances between all pairs of source and target nodes.
-In this section, we described how Lazy RPHAST can be used to solve this many-to-many problem.
-Our experiments will show, that the achieved performance is comparable to the original RPHAST algorithm.
-
-To solve the many-to-many problem, RPHAST uses SIMD vector instructions.
-Let $k$ denote the number of elements in one vector.
-RPHAST computes the distances from all source nodes to $k$ target nodes in one run.
-If the target set contains more than $k$ nodes, then the target set is split into smaller sets of $k$ elements\footnote{If the number of targets is not divisible by $k$, the set is increased by adding additional target nodes.}.
-The algorithm is then run for each of these smaller sets.
-Each RPHAST run performs vectorized variants of the selection and query phases.\todo{Is query phase definiert?}
-
-We translate this setup to the Lazy RPHAST setting as follows.
-First, a set of $k$ target nodes need to be fixed.
-A vectorized selection phase is then performed for this fixed target set.
-After the selection, the source nodes are iteratively revealed.
-Once a source node is revealed, the algorithm must compute the distances to all $k$ targets.
-Only after the distances are computed, the next source node is revealed.
-
-The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:pot}.
-We describe vectorized variants of these algorithms.
-We replace the $B$ and $D$ arrays of scalars by arrays of $k$-dimensional vectors of scalars.
-To clarify which entities are SIMD vectors, we denote the vectorized arrays by $\hat{B}$ and $\hat{D}$.
-$\hat{D}[s]$ is an SIMD vector $\hat{v}$ of length $k$.
-The $i$-th element $\hat{v}[i]$ of $\hat{v}$ will contain the distance from $s$ to $i$-th selected target.
-We can also denote this distance using $\hat{D}[s][i]$.
-All operations on SIMD vectors are component-wise.
-Adding a scalar to an SIMD vector consists of adding the scalar to every element of the SIMD vector.
-
-\todo{Rechtschreibung: down edge vs down-edge und up edge vs up-edge. Geht beides aber bitte konsistent.}
-
-\begin{algorithm2e}
-\KwIn{$\hat{T}$: SIMD vector of $k$ target nodes}
-\KwData{$\hat{B}[x][i]$: tentative distance from $x$ to target $\hat{T}[i]$}
-\KwData{Minimum priority queue $Q$, also called open list, ordered by node rank}
-Make $Q$ an empty queue\;
-$\hat{B}[x][i] \leftarrow +\infty$ for all $x$ and $i$\;
-\For{all $i\in \{0\ldots k-1\}$}{
-  $\hat{B}[\hat{T}[i]][i] \leftarrow 0$\;
-  Add $\hat{T}[i]$ to $Q$\;
-}
-\While{not $Q$ empty}{
-  $y\leftarrow$ pop minimum element from $Q$\;
-  \For{all down-edges $xy$  in $G^+_\ell$}{
-    $\hat{B}[x] \leftarrow \min\{\hat{B}[x], \hat{B}[y] + w_\ell(xy)\}$\;
-    Add $x$ to $Q$ if not in $Q$\;
-  }
-}
-\caption{Vectorized Lazy RPHAST selection}
-\label{algo:simd-ch-backward}
-\end{algorithm2e}
-
-Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
-In the original algorithm, the nodes are ordered by distance.
-In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
-To resolve this problem, we ordered the nodes in the queue not by distance but by CH rank.
-The CH rank is the position of a node in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
-All other modifications are straight-forward.
-The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
-Algorithm~\ref{algo:pot} can be vectorized by just replacing the arrays by their SIMD variants.
-The result is illustrated in Algorithm~\ref{algo:simd-pot}.
-
-\begin{algorithm2e}
-\KwData{$\hat{B}[x][i]$: tentative distance from $x$ to $\hat{T}[i]$ as computed by Algorithm~\ref{algo:simd-ch-backward}}
-\KwData{$\hat{D}[x][i]$: memoized distance from $x$ to $\hat{T}[i]$, $\bot$ initially}
-\SetKwFunction{VecDist}{SIMDComputeAndMemoizeDist}
-\SetKwProg{Fn}{Function}{:}{}
-\Fn{\VecDist{$x$}}{
-  \If{$\hat{D}[x] = \bot$}{
-    $\hat{D}[x]\leftarrow \hat{B}[x]$\;
-    \For{all up-edges $xy$ in $G^+_\ell$}{
-      $\hat{D}[x]\leftarrow\min\{\hat{D}[x],w_\ell(xy)+\VecDist(y)\}$\;
-    }
-  }
-  \Return{$\hat{D}[x]$}\;
-}
-\caption{Vectorized Lazy RPHAST algorithm}
-\label{algo:simd-pot}
-\end{algorithm2e}
-
-\todo{Hier beginnt der ältere Text}
-Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investigating the Lazy RPHAST performance in this setting seems worthwhile.
-SSE RPHAST starts by performing the selection phase on the full target set.
-% TODO flip directions?
-Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
-This second step can be improved by processing $k$ (typically 16) sources at once using SIMD vector instructions (SSE RPHAST).
-Naively translating this approach to Lazy RPHAST would mean selecting one target after another and for each doing the full lazy exploration from all sources.
-However, this approach would explore the upward search space of all the sources multiple times.
-This work seems redundant.
-
-Luckily, we can do better by integrating a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
-The Bucket CH selection phase works as follows:
-% TODO better structure or local outline
-% TODO besser herleiten, wieso kein 2d array
-For each node $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$. 
-These pairs are called \emph{buckets}.
-These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
-Now Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
-We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
-After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target node.
-This way, the search space of the sources is explored only once.
-
-The Bucket-CH selection phase still explores the shared parts of the search space of the targets several times.
-This can be improved with the following modified CH backward search:
-All target nodes are inserted into the queue at once.
-The priority queue is ordered by importance instead of distance.
-Distance labels are buckets instead of scalar values.
-The buckets of each target node are initialized with a single entry for the respective node with distance zero.
-When relaxing an edge, the entries in the current node's bucket are increased by the edge's weight and merged with the bucket entries of the head node of the edge.
-By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
-We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{sec:exp_lazy_rphast}.
-% TODO describe improved^2 Bucket CH
+% \subsection{Many-to-Many with Lazy RPHAST}
+
+% RPHAST was originally developed to answer many-to-many queries.
+% It was not designed to be used in conjunction with A*.
+% Investigating how Lazy RPHAST performs in the many-to-many setting is therefore interesting.
+% In this setting, a set of source and target nodes are given.
+% The objective is to compute the distances between all pairs of source and target nodes.
+% In this section, we described how Lazy RPHAST can be used to solve this many-to-many problem.
+% Our experiments will show, that the achieved performance is comparable to the original RPHAST algorithm.
+
+% To solve the many-to-many problem, RPHAST uses SIMD vector instructions.
+% Let $k$ denote the number of elements in one vector.
+% RPHAST computes the distances from all source nodes to $k$ target nodes in one run.
+% If the target set contains more than $k$ nodes, then the target set is split into smaller sets of $k$ elements\footnote{If the number of targets is not divisible by $k$, the set is increased by adding additional target nodes.}.
+% The algorithm is then run for each of these smaller sets.
+% Each RPHAST run performs vectorized variants of the selection and query phases.\todo{Is query phase definiert?}
+
+% We translate this setup to the Lazy RPHAST setting as follows.
+% First, a set of $k$ target nodes need to be fixed.
+% A vectorized selection phase is then performed for this fixed target set.
+% After the selection, the source nodes are iteratively revealed.
+% Once a source node is revealed, the algorithm must compute the distances to all $k$ targets.
+% Only after the distances are computed, the next source node is revealed.
+
+% The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:pot}.
+% We describe vectorized variants of these algorithms.
+% We replace the $B$ and $D$ arrays of scalars by arrays of $k$-dimensional vectors of scalars.
+% To clarify which entities are SIMD vectors, we denote the vectorized arrays by $\hat{B}$ and $\hat{D}$.
+% $\hat{D}[s]$ is an SIMD vector $\hat{v}$ of length $k$.
+% The $i$-th element $\hat{v}[i]$ of $\hat{v}$ will contain the distance from $s$ to $i$-th selected target.
+% We can also denote this distance using $\hat{D}[s][i]$.
+% All operations on SIMD vectors are component-wise.
+% Adding a scalar to an SIMD vector consists of adding the scalar to every element of the SIMD vector.
+
+% \todo{Rechtschreibung: down edge vs down-edge und up edge vs up-edge. Geht beides aber bitte konsistent.}
+
+% \begin{algorithm2e}
+% \KwIn{$\hat{T}$: SIMD vector of $k$ target nodes}
+% \KwData{$\hat{B}[x][i]$: tentative distance from $x$ to target $\hat{T}[i]$}
+% \KwData{Minimum priority queue $Q$, also called open list, ordered by node rank}
+% Make $Q$ an empty queue\;
+% $\hat{B}[x][i] \leftarrow +\infty$ for all $x$ and $i$\;
+% \For{all $i\in \{0\ldots k-1\}$}{
+%   $\hat{B}[\hat{T}[i]][i] \leftarrow 0$\;
+%   Add $\hat{T}[i]$ to $Q$\;
+% }
+% \While{not $Q$ empty}{
+%   $y\leftarrow$ pop minimum element from $Q$\;
+%   \For{all down-edges $xy$  in $G^+_\ell$}{
+%     $\hat{B}[x] \leftarrow \min\{\hat{B}[x], \hat{B}[y] + w_\ell(xy)\}$\;
+%     Add $x$ to $Q$ if not in $Q$\;
+%   }
+% }
+% \caption{Vectorized Lazy RPHAST selection}
+% \label{algo:simd-ch-backward}
+% \end{algorithm2e}
+
+% Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
+% In the original algorithm, the nodes are ordered by distance.
+% In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
+% To resolve this problem, we ordered the nodes in the queue not by distance but by CH rank.
+% The CH rank is the position of a node in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
+% All other modifications are straight-forward.
+% The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
+% Algorithm~\ref{algo:pot} can be vectorized by just replacing the arrays by their SIMD variants.
+% The result is illustrated in Algorithm~\ref{algo:simd-pot}.
+
+% \begin{algorithm2e}
+% \KwData{$\hat{B}[x][i]$: tentative distance from $x$ to $\hat{T}[i]$ as computed by Algorithm~\ref{algo:simd-ch-backward}}
+% \KwData{$\hat{D}[x][i]$: memoized distance from $x$ to $\hat{T}[i]$, $\bot$ initially}
+% \SetKwFunction{VecDist}{SIMDComputeAndMemoizeDist}
+% \SetKwProg{Fn}{Function}{:}{}
+% \Fn{\VecDist{$x$}}{
+%   \If{$\hat{D}[x] = \bot$}{
+%     $\hat{D}[x]\leftarrow \hat{B}[x]$\;
+%     \For{all up-edges $xy$ in $G^+_\ell$}{
+%       $\hat{D}[x]\leftarrow\min\{\hat{D}[x],w_\ell(xy)+\VecDist(y)\}$\;
+%     }
+%   }
+%   \Return{$\hat{D}[x]$}\;
+% }
+% \caption{Vectorized Lazy RPHAST algorithm}
+% \label{algo:simd-pot}
+% \end{algorithm2e}
+
+% \todo{Hier beginnt der ältere Text}
+% Since SSE RPHAST is the fastest known algorithm for many-to-many computations, investigating the Lazy RPHAST performance in this setting seems worthwhile.
+% SSE RPHAST starts by performing the selection phase on the full target set.
+% % TODO flip directions?
+% Second, a query is performed for each source which yields the distances from this source to all targets (RPHAST).
+% This second step can be improved by processing $k$ (typically 16) sources at once using SIMD vector instructions (SSE RPHAST).
+% Naively translating this approach to Lazy RPHAST would mean selecting one target after another and for each doing the full lazy exploration from all sources.
+% However, this approach would explore the upward search space of all the sources multiple times.
+% This work seems redundant.
+
+% Luckily, we can do better by integrating a different One-to-Many CH variant -- the Bucket-CH algorithm~\cite{gssv-erlrn-12} -- to select all targets at once.
+% The Bucket CH selection phase works as follows:
+% % TODO better structure or local outline
+% % TODO besser herleiten, wieso kein 2d array
+% For each node $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$.
+% These pairs are called \emph{buckets}.
+% These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
+% Now Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
+% We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
+% After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target node.
+% This way, the search space of the sources is explored only once.
+
+% The Bucket-CH selection phase still explores the shared parts of the search space of the targets several times.
+% This can be improved with the following modified CH backward search:
+% All target nodes are inserted into the queue at once.
+% The priority queue is ordered by importance instead of distance.
+% Distance labels are buckets instead of scalar values.
+% The buckets of each target node are initialized with a single entry for the respective node with distance zero.
+% When relaxing an edge, the entries in the current node's bucket are increased by the edge's weight and merged with the bucket entries of the head node of the edge.
+% By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
+% We evaluate our Lazy RPHAST many-to-many algorithm with this improved Bucket-CH selection in Section~\ref{sec:exp_lazy_rphast}.
+% % TODO describe improved^2 Bucket CH
 
@@
 
-\subsubsection{Alternative Routes with the Penalty Method}
-
-Providing users with multiple alternative good routes to choose from is another relevant feature for practical routing applications.
-The penalty method is an established framework for computing such alternative routes~\cite{bdgs-argrn-11,krs-eepma-13,pz-iarp-13,kobitzsch2015alternative}.
-It computes an alternative graph instead of a set of alternative routes.
-This graph is constructed iteratively.
-In each iteration, a shortest path between $s$ and $t$ is computed.
-This path (or certain parts of it) is then added to the alternative graph depending on certain quality criteria.
-Finally, all edge weights of the path and possibly adjoined edges are penalized.
-This is repeated until the current path is more than $1 + \epsilon$ times longer than the original distance or until certain other stopping criteria are fulfilled.
-
-The running time of the penalty method is usually dominated by shortest path computations.
-The computations are performed on $G_{\ell}$ with a few increased edge weights.
-This setting seems like a natural fit for A* with CH-Potentials.
-While the alternative route schema outlined above is conceptually simple, it leaves open quite a few details which may have a strong impact on performance.
-In this work, we do not develop our own variant of the penalty method.
-Instead we aim to reproduce the configuration used by Kobitzsch in his dissertation~\cite{kobitzsch2015alternative}.
-To the best of our knowledge, this is the latest, most thoroughly engineered and evaluated iteration of the penalty method.
-It also achieves the qualitatively best results.
-Some details remain unclear from the description in~\cite{kobitzsch2015alternative}.
-Luckily, we could obtain the original implementation from the authors to resolve all open questions.
-As we use the same configuration, qualitative results are directly comparable.
-We can focus on comparing running times.
-
-To ease our exposition, we briefly reiterate the details of Kobitzsch's configuration.
-The stretch is limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \cdot \dist(u,v)$.
-\todo{Das bedeutet, dass der Alternativgraph z.B. Zyklenfrei sein muss. Sobald es einen Zyklus gibt kann ich $w(P)$ beliebig groß machen.}
-Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was on the shortest path in $k$ iterations its weight will be $w_q = w_{\ell}\cdot\psi^k$.
-We also penalize edges incident to shortest paths with an asymmetric additive rejoin penalty.
-\todo{incident = nur incomming edges oder auch outgoing? rejoin klingt nur nach incomming, incident klingt nach beiden.}
-An edge $uv$ where $u$ is on a shortest path but $v$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(s,u)}{\dist_{w_q}(s,t)}$.
-Analogously, an edge $uv$ where $v$ is on a shortest path but $u$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(v,t)}{\dist_{w_q}(s,t)}$.
-\todo{Was ist $w_q$ in diesem Kontext?}
-\todo{Ich habe nicht verstanden was diese $\psi_r$ Formeln tun.}
-$\psi_r$ is set to 0.01.
-To avoid over-penalization, we count the times a \emph{node} was on a shortest path and do not apply penalties, when this number exceeds $k_{\max} := \lceil\log_{\psi}(1+\epsilon)\rceil + 2$.
-Shortest paths are split into segments not yet contained in the alternative graph.
-A segment is only added if it is long enough (at least $0.05 \cdot \dist_{w_{\ell}}(s,t)$) and does not violate the maximum stretch condition.
-We terminate the algorithm when any of the following conditions becomes true:
-\begin{itemize}
-	\item All nodes on the current shortest path have been penalized $k_{\max}$ times.
-	\item 10 ore more segments have been added to the alternative graph.
-	\item The shortest path is longer than $(1+\epsilon) \cdot \dist_{w_{\ell}}(s,t)$ with respect to the original metric.
-	\item The shortest path is longer than $(1+\epsilon) \cdot \psi + 2\psi_r \cdot \dist_{w_{\ell}}(s,t)$ with respect to the penalized metric.
-\end{itemize}
-The last two stopping criteria can both be utilized for additional pruning in the A*.
-As a final optimization, we run the forward and backward search in parallel in two threads.
-
-In~\cite{adgw-arrn-13} Abraham et al. describe a variation of the penalty method to directly obtain alternative routes rather than an alternative graph.
-They return the first path which shares less than 80\% with the original path and has stretch less than $1+\epsilon$.
-However, their approach uses a simpler penalization schema.
-We also implement this variant with CH-Potentials but stick to the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative} and denote this approach as \emph{penalty routes}.
-This leads to some qualitative improvements compared to the results reported in~\cite{adgw-arrn-13}.
+% \subsubsection{Alternative Routes with the Penalty Method}
+
+% Providing users with multiple alternative good routes to choose from is another relevant feature for practical routing applications.
+% The penalty method is an established framework for computing such alternative routes~\cite{bdgs-argrn-11,krs-eepma-13,pz-iarp-13,kobitzsch2015alternative}.
+% It computes an alternative graph instead of a set of alternative routes.
+% This graph is constructed iteratively.
+% In each iteration, a shortest path between $s$ and $t$ is computed.
+% This path (or certain parts of it) is then added to the alternative graph depending on certain quality criteria.
+% Finally, all edge weights of the path and possibly adjoined edges are penalized.
+% This is repeated until the current path is more than $1 + \epsilon$ times longer than the original distance or until certain other stopping criteria are fulfilled.
+
+% The running time of the penalty method is usually dominated by shortest path computations.
+% The computations are performed on $G_{\ell}$ with a few increased edge weights.
+% This setting seems like a natural fit for A* with CH-Potentials.
+% While the alternative route schema outlined above is conceptually simple, it leaves open quite a few details which may have a strong impact on performance.
+% In this work, we do not develop our own variant of the penalty method.
+% Instead we aim to reproduce the configuration used by Kobitzsch in his dissertation~\cite{kobitzsch2015alternative}.
+% To the best of our knowledge, this is the latest, most thoroughly engineered and evaluated iteration of the penalty method.
+% It also achieves the qualitatively best results.
+% Some details remain unclear from the description in~\cite{kobitzsch2015alternative}.
+% Luckily, we could obtain the original implementation from the authors to resolve all open questions.
+% As we use the same configuration, qualitative results are directly comparable.
+% We can focus on comparing running times.
+
+% To ease our exposition, we briefly reiterate the details of Kobitzsch's configuration.
+% The stretch is limited to $\epsilon = 0.25$, i.e. for any path $P = (u,\dots,v)$ in the alternative graph $w(P) \leq (1+\epsilon) \cdot \dist(u,v)$.
+% \todo{Das bedeutet, dass der Alternativgraph z.B. Zyklenfrei sein muss. Sobald es einen Zyklus gibt kann ich $w(P)$ beliebig groß machen.}
+% Shortest path edges are penalized multiplicative with a factor of $\psi = 1.1$, i.e. when an edges was on the shortest path in $k$ iterations its weight will be $w_q = w_{\ell}\cdot\psi^k$.
+% We also penalize edges incident to shortest paths with an asymmetric additive rejoin penalty.
+% \todo{incident = nur incomming edges oder auch outgoing? rejoin klingt nur nach incomming, incident klingt nach beiden.}
+% An edge $uv$ where $u$ is on a shortest path but $v$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(s,u)}{\dist_{w_q}(s,t)}$.
+% Analogously, an edge $uv$ where $v$ is on a shortest path but $u$ is not will get its weight increased by $\psi_r \cdot \dist_{w_{\ell}}(s,t) \cdot \frac{\dist_{w_q}(v,t)}{\dist_{w_q}(s,t)}$.
+% \todo{Was ist $w_q$ in diesem Kontext?}
+% \todo{Ich habe nicht verstanden was diese $\psi_r$ Formeln tun.}
+% $\psi_r$ is set to 0.01.
+% To avoid over-penalization, we count the times a \emph{node} was on a shortest path and do not apply penalties, when this number exceeds $k_{\max} := \lceil\log_{\psi}(1+\epsilon)\rceil + 2$.
+% Shortest paths are split into segments not yet contained in the alternative graph.
+% A segment is only added if it is long enough (at least $0.05 \cdot \dist_{w_{\ell}}(s,t)$) and does not violate the maximum stretch condition.
+% We terminate the algorithm when any of the following conditions becomes true:
+% \begin{itemize}
+% 	\item All nodes on the current shortest path have been penalized $k_{\max}$ times.
+% 	\item 10 ore more segments have been added to the alternative graph.
+% 	\item The shortest path is longer than $(1+\epsilon) \cdot \dist_{w_{\ell}}(s,t)$ with respect to the original metric.
+% 	\item The shortest path is longer than $(1+\epsilon) \cdot \psi + 2\psi_r \cdot \dist_{w_{\ell}}(s,t)$ with respect to the penalized metric.
+% \end{itemize}
+% The last two stopping criteria can both be utilized for additional pruning in the A*.
+% As a final optimization, we run the forward and backward search in parallel in two threads.
+
+% In~\cite{adgw-arrn-13} Abraham et al. describe a variation of the penalty method to directly obtain alternative routes rather than an alternative graph.
+% They return the first path which shares less than 80\% with the original path and has stretch less than $1+\epsilon$.
+% However, their approach uses a simpler penalization schema.
+% We also implement this variant with CH-Potentials but stick to the penalization schema from Kobitzsch~\cite{kobitzsch2015alternative} and denote this approach as \emph{penalty routes}.
+% This leads to some qualitative improvements compared to the results reported in~\cite{adgw-arrn-13}.
 
@@
 To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are nodes drawn uniformly at random and report average results.
-For many-to-one and many-to-many algorithms, we use the methodology from~\cite{delling_et_al:OASIcs:2011:3266}.
+For many-to-one algorithms, we use the methodology from~\cite{delling_et_al:OASIcs:2011:3266}.
 We generate a \emph{ball} $B$ of nodes around 100 uniformly at random chosen nodes by running a Dijkstra search until $|B|$ nodes are settled where $|B|$ is a parameter varied throughout the experiments.
-From each ball we uniformly at random draw a set of sources $S$ and targets $T$ with $|S| = |T| = 2^{14}$.
-For one-to-many experiments, we pick 100 sources of $S$ and for each source compute the distances to all nodes in the respective $T$ for a total of 10\,000 one-to-many queries.
-For many-to-many experiments, we compute the distance table between the 100 $S$, $T$ pairs.
+From each ball we uniformly at random pick 100 targets and a set of sources $S$ with $|S| = 2^{14}$.
+We perform a many-to-one query for each target computing distances from all sources to it.
+This gives us 100 many-to-one queries per ball and a total of 10\,000 one-to-many queries per ball size.
 
 \subsection{Lazy RPHAST}\label{sec:exp_lazy_rphast}
-
-% better wrap story branch
 
@@
 \caption{
-Total and per distance running times of incremental Lazy RPHAST with up to $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$ on OSM Germany.
+Running times of incremental Lazy RPHAST with up to $|S| = 2^{14}$ sources picked from a ball of varying size $|B|$ on OSM Germany excluding selection times.
+The upper figure contains the total running time.
+The lower figure contains the averaged running time per query, i.e. $y/x$ from the upper figure.
 Note the different y-axis scales and units.
@@
 
-To evaluate Lazy RPHAST in the incremental setting, we measure the elapsed running time after $2^i$ result distances were computed. % TODO word smithing "result distances"
-% TODO selection not included
+To evaluate Lazy RPHAST in the incremental setting, we measure the elapsed running time after distances from $2^i$ sources were queried.
 Figure~\ref{fig:lazy_rphast_inc} depicts the total elapsed time and the average running time to compute a single distance.
@@
 \centering
-\includegraphics[width=\linewidth]{fig/lazy_rphast_many_to_one.pdf}
+\includegraphics[width=\linewidth]{fig/lazy_rphast_many_to_one_both.pdf}
 \caption{
-Running times of CH based Lazy RPHAST for many-to-one queries with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
+Running times of (C)CH-based Lazy RPHAST for many-to-one queries with $|S| = 2^{14}$ sources picked from a ball of varying size $|B|$.
+The running time includes the selection and the time to compute all distances.
 }\label{fig:many_to_one}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{fig/lazy_rphast_cch_many_to_one.pdf}
-\caption{
-Running times of CCH based Lazy RPHAST for many-to-one queries with $|T| = 2^{14}$ targets picked from a ball of varying size $|B|$.
-}\label{fig:many_to_one_cch}
-\end{figure}
-
-Despite the laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental one-to-many queries.
-Figures~\ref{fig:many_to_one} and~\ref{fig:many_to_one_cch} depict average running times to compute distances between one node and $2^{14}$ targets for different ball sizes.
-This experiment uses the exact methodology described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to roughly relate our results to the performance of competing algorithms.
-When looking at total running times for a single many-to-one query, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
+Despite the laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental many-to-one queries.
+Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target node and $2^{14}$ sources for different ball sizes.
+This experiment uses the exact methodology described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to roughly relate our results to the performance of other one-to-many algorithms.
+Keep in mind that algorithms like RPHAST optimize for a different setting than we do.
+We can compare the performance for fixed $S$-$t$ terminal sets.
+The difference is that with RPHAST one can efficiently compute distances from a different $t'$ to the same $S$ set while with Lazy RPHAST one can efficiently extend $S$ while $t$ stays the same.
+
+When looking at total running times for a single many-to-one query including selection times, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
 For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
@@
 The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
-This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
-When computing distances from different sources towards the same target set, RPHAST will outperform Lazy RPHAST.
-However, whenever the target set may change between queries, Lazy RPHAST offers a simple and fast alternative.
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{fig/lazy_rphast_many_to_many.pdf}
-\caption{
-Average running times of Lazy RPHAST for many-to-many queries with $|S| = |T| = 2^{14}$ sources and targets picked from a ball of varying size $|B|$ with standard deviation indicated as error bars.
-The lower part of the bars indicates the running time of the Bucket CH subphase.
-}\label{fig:many_to_many}
-\end{figure}
-
-Applying Lazy RPHAST to the many-to-many problem results in running times as depicted in Figure~\ref{fig:many_to_many}.
-The results are again at least competitive to the performance reported for SSE RPHAST in~\cite{delling_et_al:OASIcs:2011:3266}.
-% TODO konkrete zahlen?
-The Bucket CH based first query phase of our implementation takes between half of the total query time for $|B| = 2^{14}$ and a tenth of the time for $|B| = 2^{24}$.
-It is worth noting that contrary to SSE RPHAST, we did not explicitly utilize SIMD instructions.
-However, the code for Lazy RPHAST on an arrays of distances can be trivially vectorized by the compiler.
-We confirmed that the code was indeed vectorized by inspecting the generated assembly.
-
-\subsubsection{k-Nearest-Neighbor Queries with CCH}
-
-One-to-many queries appear in many extended applications as a subproblem.
-In some of these cases Lazy RPHAST can be used to achieve significant speed-ups with surprisingly little effort.
-One such example are k-Nearest-Neighbor Queries with CCH~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
-There, shortest distances are computed repeatedly from the same source to boundary nodes of a cell of nodes.
-On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
-We reran their improved experiments and include the results in our appendix.
-Briefly summarized, integrating Lazy RPHAST into this algorithm decreased running times by up to 15 times depending on the distribution of the target set for almost no effort.
-This demonstrates that Lazy RPHAST is not only useful as an A* potential but also applicable in other contexts.
+We conclude that that Lazy RPHAST is a very useful extension to RPHAST.
+It allows efficiently handling dynamic source sets and is even faster than RPHAST in the setting where both the target and the source set change between queries.
+% This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
+% When computing distances from different sources towards the same target set, RPHAST will outperform Lazy RPHAST.
+% However, whenever the target set may change between queries, Lazy RPHAST offers a simple and fast alternative.
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=\linewidth]{fig/lazy_rphast_many_to_many.pdf}
+% \caption{
+% Average running times of Lazy RPHAST for many-to-many queries with $|S| = |T| = 2^{14}$ sources and targets picked from a ball of varying size $|B|$ with standard deviation indicated as error bars.
+% The lower part of the bars indicates the running time of the Bucket CH subphase.
+% }\label{fig:many_to_many}
+% \end{figure}
+
+% Applying Lazy RPHAST to the many-to-many problem results in running times as depicted in Figure~\ref{fig:many_to_many}.
+% The results are again at least competitive to the performance reported for SSE RPHAST in~\cite{delling_et_al:OASIcs:2011:3266}.
+% % TODO konkrete zahlen?
+% The Bucket CH based first query phase of our implementation takes between half of the total query time for $|B| = 2^{14}$ and a tenth of the time for $|B| = 2^{24}$.
+% It is worth noting that contrary to SSE RPHAST, we did not explicitly utilize SIMD instructions.
+% However, the code for Lazy RPHAST on an arrays of distances can be trivially vectorized by the compiler.
+% We confirmed that the code was indeed vectorized by inspecting the generated assembly.
+
+% \subsubsection{k-Nearest-Neighbor Queries with CCH}
+
+% One-to-many queries appear in many extended applications as a subproblem.
+% In some of these cases Lazy RPHAST can be used to achieve significant speed-ups with surprisingly little effort.
+% One such example are k-Nearest-Neighbor Queries with CCH~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
+% There, shortest distances are computed repeatedly from the same source to boundary nodes of a cell of nodes.
+% On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
+% We reran their improved experiments and include the results in our appendix.
+% Briefly summarized, integrating Lazy RPHAST into this algorithm decreased running times by up to 15 times depending on the distribution of the target set for almost no effort.
+% This demonstrates that Lazy RPHAST is not only useful as an A* potential but also applicable in other contexts.
 
@@
 
-\subsubsection{Alternative Routes}
-
-Our implementation of the penalty method reproduces the qualitative results reported in~\cite{kobitzsch2015alternative} (for random queries on DIMACS Europe we on average achieve a total distance of 2.93, an average distance of 1.04, 5.57 decision edges and a success rate of 98.7\% with 6.7 iterations).
-The average running times to compute an alternative graph are 1.37\,s on DIMACS Europe and 0.85\,s on OSM Germany.
-These running times are roughly one order of magnitude slower than the results achieved by Kobitzsch.
-The simpler penalty routes approach is much faster and takes only 156\,ms on DIMACS Europe and 66\,ms on OSM Germany.
-This is two orders of magnitude faster than the running times reported in~\cite{adgw-arrn-13} which is not surprising since their implementation uses an unaccelerated Dijkstra.
-Due to the advanced penalization schema, our average qualitative results differ from~\cite{adgw-arrn-13}: the success rate is with 96.8\% slightly higher, the UBS with around 9\% slightly worse, the sharing is with 42\% somewhat higher but the local optimality, which was the weakest point in the original approach, with 21\% significantly better.
-Finding these alternatives takes only 1.44 iterations on average.
-This makes CH-Potential-based penalty routes a simple and feasible practical approach to compute alternative routes.
+% \subsubsection{Alternative Routes}
+
+% Our implementation of the penalty method reproduces the qualitative results reported in~\cite{kobitzsch2015alternative} (for random queries on DIMACS Europe we on average achieve a total distance of 2.93, an average distance of 1.04, 5.57 decision edges and a success rate of 98.7\% with 6.7 iterations).
+% The average running times to compute an alternative graph are 1.37\,s on DIMACS Europe and 0.85\,s on OSM Germany.
+% These running times are roughly one order of magnitude slower than the results achieved by Kobitzsch.
+% The simpler penalty routes approach is much faster and takes only 156\,ms on DIMACS Europe and 66\,ms on OSM Germany.
+% This is two orders of magnitude faster than the running times reported in~\cite{adgw-arrn-13} which is not surprising since their implementation uses an unaccelerated Dijkstra.
+% Due to the advanced penalization schema, our average qualitative results differ from~\cite{adgw-arrn-13}: the success rate is with 96.8\% slightly higher, the UBS with around 9\% slightly worse, the sharing is with 42\% somewhat higher but the local optimality, which was the weakest point in the original approach, with 21\% significantly better.
+% Finding these alternatives takes only 1.44 iterations on average.
+% This makes CH-Potential-based penalty routes a simple and feasible practical approach to compute alternative routes.
 
@@
 In this article, we introduced Lazy RPHAST, an efficient algorithm for the incremental many-to-one problem.
-We also showed that it delivers competitive performance for one-to-many and many-to-many problems.
+We also showed that it delivers competitive performance for many-to-one problems.
 Based on Lazy RPHAST, we developed (C)CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and (C)CH for finding shortest paths in road networks.
@@
 Hierarchical techniques have been shown to work well on these graphs~\cite{DBLP:conf/aaai/UrasK14}.
-It might also be worthwhile to apply CH-Potentials to even more routing applications, for example to the shortest $\epsilon$-smooth path problem~\cite{dss-tarrn-18}.
+It might also be worthwhile to apply CH-Potentials to even more routing applications, for example to the shortest $\epsilon$-smooth path problem~\cite{dss-tarrn-18} or the problem of computing alternative routes~\cite{adgw-arrn-13,bdgs-argrn-11,kobitzsch2015alternative}.
+Many-to-one problems also appear as subproblems in other routing algorithms, for example in nearest neighbor computations~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
+Investigating whether Lazy RPHAST can be used to improve these algorithms appears to be a worthwhile direction for future research.
+Studying the performance of Lazy RPHAST in a many-to-many context would also be interesting.
 
@@
 
-\appendix
-
-\section{Improved Results for k-Nearest-Neighbor Queries on CCH}
-
-In this section, we report the results of the k-Nearest-Neighbor Queries on CCH with Lazy RPHAST.
-The authors of~\cite{buchhold_et_al:LIPIcs.SEA.2021.18} provide the code for their algorithm and their experiments on Github\footnote{\url{https://github.com/vbuchhold/routing-framework}}.
-This includes the code for both the old version and the new one improved with Lazy RPHAST.
-We reran the experiments performed for~\cite{buchhold_et_al:LIPIcs.SEA.2021.18} in both the old and the new configuration on our machine.
-In the results, we denote the old version as \emph{CCH/CRAD} and the new one as \emph{CCH*/CRAD*}.
-Our results exhibit the same characteristics as observed in the original publication.
-Absolute running times of course vary slightly.
-
-Lazy RPHAST yields the greatest improvements of more than an order of magnitude when the points of interest (\emph{POIs}) are concentrated in a smaller region (see Table~\ref{tab:knn} and Figure~\ref{fig:knn_ball_size}).
-When the POIs are distributed over the whole graph, the speed-up is only around factor of three.
-Lazy RPHAST also achieves greater speed-ups for larger numbers of POIs (see Figure~\ref{fig:knn_num_pois}).
-For the demand generation application, this translates to a speed-up factor of around 2.5 (see Figure~\ref{fig:knn_demands}).
-For a more detailed discussion of these results and a comparison to competing algorithms, we refer to~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
-
-\begin{table}
-\centering
-\caption{
-Performance of different closest-POI algorithms for various POI distributions.
-For each distribution, we report the time to index a set of POIs (selection time), the space consumed by the
-index (selection space), and the time to find the $k = 1, 4, 8$ closest POIs (query time).
-}\label{tab:knn}
-\input{table/knn.tex}
-\end{table}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{fig/knn_ball_size.pdf}
-\caption{
-Selection and query times of various closest-POI algorithms with $|P| = 2^{14}$ POIs picked at random from a ball of varying size $|B|$.
-Queries find the $k = 4$ closest POIs.
-}\label{fig:knn_ball_size}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{fig/knn_num_pois.pdf}
-\caption{
-Selection and query times of various closest-POI algorithms with a varying number $|P|$ of POIs picked at random from a ball of size $|B| = |V|$.
-Queries find the $k = 4$ closest POIs.
-}\label{fig:knn_num_pois}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=.6363\linewidth]{fig/knn_demands.pdf}
-\caption{
-Time to generate a single trip with different demand generators for various values of $\lambda$.
-}\label{fig:knn_demands}
-\end{figure}
+% \appendix
+
+% \section{Improved Results for k-Nearest-Neighbor Queries on CCH}
+
+% In this section, we report the results of the k-Nearest-Neighbor Queries on CCH with Lazy RPHAST.
+% The authors of~\cite{buchhold_et_al:LIPIcs.SEA.2021.18} provide the code for their algorithm and their experiments on Github\footnote{\url{https://github.com/vbuchhold/routing-framework}}.
+% This includes the code for both the old version and the new one improved with Lazy RPHAST.
+% We reran the experiments performed for~\cite{buchhold_et_al:LIPIcs.SEA.2021.18} in both the old and the new configuration on our machine.
+% In the results, we denote the old version as \emph{CCH/CRAD} and the new one as \emph{CCH*/CRAD*}.
+% Our results exhibit the same characteristics as observed in the original publication.
+% Absolute running times of course vary slightly.
+
+% Lazy RPHAST yields the greatest improvements of more than an order of magnitude when the points of interest (\emph{POIs}) are concentrated in a smaller region (see Table~\ref{tab:knn} and Figure~\ref{fig:knn_ball_size}).
+% When the POIs are distributed over the whole graph, the speed-up is only around factor of three.
+% Lazy RPHAST also achieves greater speed-ups for larger numbers of POIs (see Figure~\ref{fig:knn_num_pois}).
+% For the demand generation application, this translates to a speed-up factor of around 2.5 (see Figure~\ref{fig:knn_demands}).
+% For a more detailed discussion of these results and a comparison to competing algorithms, we refer to~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
+
+% \begin{table}
+% \centering
+% \caption{
+% Performance of different closest-POI algorithms for various POI distributions.
+% For each distribution, we report the time to index a set of POIs (selection time), the space consumed by the
+% index (selection space), and the time to find the $k = 1, 4, 8$ closest POIs (query time).
+% }\label{tab:knn}
+% \input{table/knn.tex}
+% \end{table}
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=\linewidth]{fig/knn_ball_size.pdf}
+% \caption{
+% Selection and query times of various closest-POI algorithms with $|P| = 2^{14}$ POIs picked at random from a ball of varying size $|B|$.
+% Queries find the $k = 4$ closest POIs.
+% }\label{fig:knn_ball_size}
+% \end{figure}
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=\linewidth]{fig/knn_num_pois.pdf}
+% \caption{
+% Selection and query times of various closest-POI algorithms with a varying number $|P|$ of POIs picked at random from a ball of size $|B| = |V|$.
+% Queries find the $k = 4$ closest POIs.
+% }\label{fig:knn_num_pois}
+% \end{figure}
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=.6363\linewidth]{fig/knn_demands.pdf}
+% \caption{
+% Time to generate a single trip with different demand generators for various values of $\lambda$.
+% }\label{fig:knn_demands}
+% \end{figure}
 

### 2021-11-12 eval/rphast_comparison.py

@@ -63,3 +63,22 @@ handles, labels = g.get_legend_handles_labels()
 g.legend(handles[0:2], labels[0:2])
 plt.tight_layout()
 g.get_figure().savefig('paper/fig/lazy_rphast_many_to_many.pdf')
+
+
+plt.figure(figsize=(11,5))
+g = sns.lineplot(data=queries
+                 .loc[lambda x: x['algo'] != 'lazy_rphast_many_to_many']
+                 .loc[lambda x: x['target_set_size_exp'] == 14]
+                 .groupby(['ball_size_exp', 'graph', 'algo'])['running_time_ms'].mean()
+                 .reset_index(), x='ball_size_exp', y='running_time_ms', hue='graph', style='algo', markers=True, ci='sd', hue_order=['OSM Germany', 'DIMACS Europe'], style_order=['lazy_rphast_many_to_one', 'lazy_rphast_cch_many_to_one'])
+g.set_xlabel('Ball Size $|B|$')
+g.set_ylabel('Running Time [ms]')
+g.set_yscale('log')
+g.yaxis.set_major_formatter(mpl.ticker.LogFormatter())
+g.xaxis.set_major_locator(mpl.ticker.MaxNLocator(11))
+g.xaxis.set_major_formatter(mpl.ticker.FuncFormatter(lambda val, pos: f"$2^{{{int(val)}}}$"))
+plt.grid(True, which="minor", linewidth=0.6)
+handles, labels = g.get_legend_handles_labels()
+g.legend(handles[1:3] + handles[4:6], labels[1:3] + ['CH', 'CCH'])
+plt.tight_layout()
+g.get_figure().savefig('paper/fig/lazy_rphast_many_to_one_both.pdf')

### 2021-11-12 eval/rphast_inc.py

@@ -33,11 +33,11 @@ queries['Ball Size $|B|$'] = queries['ball_size_exp'].map('$2^{{{}}}$'.format)
 
 fig, (ax1, ax2) = plt.subplots(2, sharex=True, figsize=(11,8))
 g = sns.lineplot(data=queries.query('graph == "OSM Germany" & ball_size_exp in [14,17,21,24]'), x='terminal_idx', y='passed_time_ms', hue='Ball Size $|B|$', style='Algorithm', ci=None, marker='X', ax=ax1)
-g.set_xlabel('Number of Computed Distances')
+g.set_xlabel('Number of Queried Distances')
 g.set_ylabel('Running Time [ms]')
 g = sns.lineplot(data=queries.query('graph == "OSM Germany" & ball_size_exp in [14,17,21,24]'), x='terminal_idx', y='single_query_time_mus', hue='Ball Size $|B|$', style='Algorithm', ci=None, marker='X', ax=ax2, legend=False)
 g.set_yscale('log')
-g.set_xlabel('Number of Computed Distances')
+g.set_xlabel('Number of Queried Distances')
 g.set_ylabel('Average Running Time per Distance [$\\mu$s]')
 g.yaxis.set_major_formatter(mpl.ticker.FuncFormatter(lambda val, pos: f"{val if val < 1.0 else int(val)}"))
 fig.tight_layout()

### 2021-11-14 paper/ch_potentials.tex
changes in many places

@@
 In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as a A* heuristic.
-Since it utilizes Contraction Hierarchies (CH), a hierarchical technique, we call our heuristic CH-Potentials.
+As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
 Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
@@
 The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree nodes, common in road networks, and some applications of CH-Potentials.
-In this article we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in much greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
+In this article, we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
 % Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
 The original paper only discussed unidirectional CH-Potentials.
-In this journal version, we also discuss a bidirectional variant and also introduce an improved pruning criterion for symmetrical bidirectional A* which can also be utilized for other heuristic.
-We also include a significantly expanded evaluation.
+In this journal version, we also discuss a bidirectional variant and introduce an improved pruning criterion for symmetrical bidirectional A*.
+The improved pruning is not specific to CH-Potentials and can also be used with other heuristics.
+Compared to the conference paper, the journal version contains a significantly expanded experimental evaluation.
 
@@
 We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
-Given nodes $s$ and $t$ we want to obtain a path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
+Given nodes $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
 We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
+If there is no path, we set $\dist_w(s,t)=\infty$.
 
@@
 
-Dijkstra's algorithm~\cite{d-ntpcg-59} can be used to solve this shortest path problem.
-In this section, we recall the classical algorithm and establish some notation that we will employ in the remainder of this paper.
-Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue of nodes ordered by increasing distance from $s$.
+In the this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
+Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of nodes ordered by increasing distance from $s$.
 Initially, all distances are set to $\infty$.
-The start node distance $D[s]$ is set to zero and the queue initialized with $s$.
-Then, in each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
+The start node distance $D[s]$ is set to zero and the queue is initialized with $s$.
+In each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
 Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
-If the distance at the head node $v$ was improved it will be inserted into the queue or the queue position will be updated if $v$ was already in the queue.
-Once $t$ was settled, $D[t]$ contains the distance between $s$ and $t$.
+If the tentative distance at the head node $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
+Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
+The search can stop when $t$ is settled.
+
+\todo{$w(uv)$ vs $w(u,v)$, egal welches aber bitte konsistent.}
 
@@
 We call this a \emph{backward search}.
-Running two Dijkstra searches, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.
+Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.\todo{Kann von mir aus bleiben, aber ich muss immer an Janus denken bei ``Bidirectional Dijkstra''.}
 Typically, the searches are interleaved by alternating settling a node from each direction.
@@
 Theoretically any direction interleaving strategy is possible.
-The searches can also be run in parallel in two threads.
+The searches can also be run in parallel in two threads.\todo{Bringt es uns etwas, dass hier Alternierungsstrategien aufgelistet werden, oder kann das weg?}
 When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.
@@
 % A* uses a \emph{heuristic} to guide the search towards the goal.
-It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target.
-The priority queue is ordered by $d[v] + h_t(v)$.
-Thus, nodes closer to the target a explored earlier.
-
+It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target $t$.
+A* orders the nodes in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as is done by Dijkstra's algorithm.
+This results in nodes closer to the target being explored earlier.
+Dijkstra's algorithm can be obtained as a special case of A* by setting $h_t(v)=0$.
 A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
 This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
-A heuristic function is called \emph{feasible} if $w_{h_t} \geq 0$ for all edges.
-A* is guaranteed to compute an optimal path, if the heuristic is feasible.
-In this article, we therefore limit our discussion to feasible heuristics.
+
+For A* to be optimal, the heuristic needs to be feasible.
+A heuristic function is called \emph{feasible}, if $w_{h_t}(e) \geq 0$ for all edges $e$.
+In this article, we limit our discussion to feasible heuristics.
 
@@
 
-Customizable Contraction Hierarchies~\cite{dsw-cch-15} are a special variant of Contraction Hierarchies.
-They use a node importance order independent of the weight function.
-% TODO so far, we only have levels but not ranks
-This allows splitting the preprocessing in two subphases.
-The first one is slow but independent of the weight function.
-The second one, called \emph{customization}, is fast (a couple of seconds or less) and applies a weight function to the preprocessed data.
-This allows regular updates to the weight function, for example to incorporate live traffic.
-
-As for CH, the result of the preprocessing is an augmented graph $G^+$ fulfilling the same properties as for CH.
-Thus, the CH query algorithms can be applied without modifications.
+\label{sec:cch-intro}
+
+Some hierarchical techniques were extended to support a three phase setup.
+The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
+The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
+The preprocessing phase is independent of the weights and only exploits the graph topology.
+Weights are introduces in the customization phase.
+Paths are computed in the query phase.
+This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
+The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}
+
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} are a variant of Contraction Hierarchies that uses this three phase setup.
+The key idea is to use a node importance order that is independent of the weight function.
+As a CCH is also a CH, the same concepts such as the augmented graph $G^+$ exists.
+The CH query algorithms can therefore be applied without modifications.
 However, the CCH preprocessing grants some additional useful properties.
+
 During preprocessing, an \emph{elimination tree} can be constructed.
@@
 The backward search works analogously but starts at the target.
-Because this algorithm does not require a queue, it is faster than a Dijkstra-based CH query.
+Because this elimination tree based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
+However, only weight-independent node orderings admit an elimination tree of low height.
+The elimination tree query can therefore not be used for a general CH.
+Using a CH with a weight-specific importance order is often faster than a CCH with elimination tree query.
 
@@
 
-\subsection{Lazy RPHAST}
+\subsection{Lazy RPHAST for Contraction Hierarchies}
 
@@
 
-\subsection{Lazy RPHAST on CCH}
+\subsection{Lazy RPHAST for Customizable Contraction Hierarchies}
 
@@
 To compute the distance $D[x]$ of a node $x$ the distances of all upward neighbors must be final.
-In Algorithm~\ref{algo:pot} we have to compute all these distances recursively.
+In Algorithm~\ref{algo:pot}, we have to compute all these distances recursively.
 However, on CCH it is sufficient to ensure that the distance of the elimination tree parent of $x$ is final.
@@
 \KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
-\KwData{$P[x]$: parent node of $x$ in the elimination tree}
-\KwData{$S$: stack with nodes to compute distances}
+\KwData{$P[x]$: parent node of $x$ in the elimination tree, parent of root is $\bot$}
+\KwData{$S$: stack with nodes to compute distances, empty initially, only used to store intermediate data}
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
@@
 \Fn{\Dist{$x$}}{
+  \tcp{Determine for what nodes, the $D$ needs to be filled}
   $y \gets x$\;
@@
 		}
-    $y \gets P[y]$
-  }
+    $y \gets P[y]$\;
+  }
+  \tcp{Fill $D$ for those nodes}
   \While{$S$ not empty}{
@@
 
-On road graphs, bidirectional search provides an easy way to halve the running time of Dijkstra's algorithm.
+On road graphs, bidirectional search provides a way to halve the running time of Dijkstra's algorithm.\todo{sag niemals etwas sei ``easy''.
+Es gibt immer jemand der es schwer fand.
+Den schlägst du ins Gesicht.}
 A bidirectional variant of A* therefore also seems desirable.
@@
 
-Supporting live traffic is also possible with a three-phase setup:
-A slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
-The customization phase is run regularly and incorporates updates to the weights into the auxiliary preprocessing data.
-CRP~\cite{dgpw-crprn-13} and CCH~\cite{dsw-cch-15} follow this setup.
-As discussed in Section~\ref{sec:lazy-rphast}, the Lazy RPHAST algorithm and thus CH-Potentials can also be used with a CCH.
-In terms of our framework this would mean having a second preprocessing phase which allows updating $w_\ell$ to the current traffic situation.
-Thus, with CCH-Potentials, we could also support a three-phase setup.
-This could, for example, be used to combine live traffic with turn costs and restrictions.
+Live traffic can alternatively be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
+We do this by exchanging the CH for a CCH.
+The preprocessing phase is identical as for a CCH and is independent of $w_\ell$.
+The $w_\ell$ is fixed during the customization phase.
+It can be changed arbitrarily by rerunning the customization.
+The A* algorithm is part of the query phase.
+The weights $w_q$ used in the query phase must not be smaller than the $w_\ell$ used during the last customization.
+
+This setup provides an alternative way of integrating live traffic compared to the approach described in~\ref{sec:live-predicted-traffic}.
+Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
+Vehicle specific restrictions are accounted for in the query by adjusting $w_q$.
+For severe traffic events, the three phase approach is likely faster.
+However, it comes at the expense of a more complex algorithm.
+Further, integrating predicted traffic becomes difficult, as $w_q$ may not drop below the $w_\ell$ with integrated congestion.
+
+\todo{Den Gedanken sollte man ausbauen oder streichen}
 One could also apply live traffic updates to $w_q$ immediately and only update $w_\ell$ once the difference between $w_\ell$ and $w_q$ grows too big or when some edge weight update would set $w_q < w_\ell$.
@@
 We compare (C)CH-Potentials to three other heuristics.
-First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
+The first heuristic is, the \emph{Zero} heuristic where $h(x)$ is $0$ for all nodes $x$.
 This corresponds to using Dijkstra's algorithm.
-Second, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
+Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
 We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
@@
 We fill this array before each query using a reverse Dijkstra search from the target node.
+In this experiment, we do not account for the running time of Dijkstra's algorithm.
 Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
-CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
+(C)CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
 Comparing against Oracle-A* allows us to measure this overhead.
-Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.
+No other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.
 
 We observe that the number of queue pushes roughly correlates with running time.
+As (C)CH-Potentials and Oracle-A* have the same heuristic values, the number of queue pushes are by construction equal.
 Each optimization reduces both queue pushes and running times.
@@
 % Skipping degree three nodes has the smallest impact.
-CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly fewer nodes.
+(C)CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly fewer nodes.
 This is not surprising, since ALT computes worse distance estimates.
@@
 This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
-CCH-Potentials are slightly slower than CH-Potentials because their search space is denser.
+CCH-Potentials are slightly slower than CH-Potentials because they use a weight independent node importance order.
+This enables CCH-Potentials to support a customization whereas CH-Potentials cannot.
 
@@
 Enabling the improved pruning brings the performance symmetric bidirectional A* on par with the average potential for all heuristics except ALT.
+\todo{den nächsten Satz verstehe ich nicht}
 Apparently, the average potential gives ALT a better heuristic than the original one for for each direction individually, contrary to the statements made in~\cite{gh-cspas-05}.
@@
 Enabling the improved pruning for the average potential, marginally reduces the search space size and slightly increases running times.
+
+\todo{Vergleich zwischen Tabelle 3 und 4? Warum ist uni directional langsamer? Hier sollte man schreiben, dass das Thema bei Tabelle 5 addressiert wird}
+\todo{was ist die Alternierungsstrategie von Tabelle 3?}
 

### 2021-11-15 paper/ch_potentials.tex
mostly integrate bens comments

@@
 We call this a \emph{backward search}.
-Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called a \emph{Bidirectional Dijkstra}.\todo{Kann von mir aus bleiben, aber ich muss immer an Janus denken bei ``Bidirectional Dijkstra''.}
+Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called \emph{bidirectional} search.
 Typically, the searches are interleaved by alternating settling a node from each direction.
 Another common approach is to always settle a node from the direction with the smaller minimum queue key.
-Theoretically any direction interleaving strategy is possible.
-The searches can also be run in parallel in two threads.\todo{Bringt es uns etwas, dass hier Alternierungsstrategien aufgelistet werden, oder kann das weg?}
+Theoretically, any direction interleaving strategy is possible.
 When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.
@@
 It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target $t$.
-A* orders the nodes in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as is done by Dijkstra's algorithm.
+A* orders the nodes in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as Dijkstra's algorithm does it.
 This results in nodes closer to the target being explored earlier.
-Dijkstra's algorithm can be obtained as a special case of A* by setting $h_t(v)=0$.
+Dijkstra's algorithm is a special case of A*  with $h_t(v)=0$.
 A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
 This reduced weight function is defined as $w_{h_t}(uv) := w(uv) - h_t(u) + h_t(v)$.
-
-For A* to be optimal, the heuristic needs to be feasible.
 A heuristic function is called \emph{feasible}, if $w_{h_t}(e) \geq 0$ for all edges $e$.
+If the employed heuristic is feasible, A* is guaranteed to find optimal distances.
 In this article, we limit our discussion to feasible heuristics.
+% TODO more explicit details on A* correctness?
 
@@
 \subsection{Customizable Contraction Hierarchies (CCH)}
-
 \label{sec:cch-intro}
@@
 
-Customizable Contraction Hierarchies~\cite{dsw-cch-15} are a variant of Contraction Hierarchies that uses this three phase setup.
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three phase setup.
 The key idea is to use a node importance order that is independent of the weight function.
-As a CCH is also a CH, the same concepts such as the augmented graph $G^+$ exists.
+A CCH is also a CH.
+The final result of the preprocessing is also an augmented graph $G^+$.
+It fulfills all necessary properties for CH query algorithms.
 The CH query algorithms can therefore be applied without modifications.
@@
 The root of the elimination tree is the most important node.
-% TODO so far, we only have levels but not ranks
 For every other node its parent is its least important upward neighbor.
@@
 The elimination tree query can therefore not be used for a general CH.
-Using a CH with a weight-specific importance order is often faster than a CCH with elimination tree query.
 
@@
 
-On road graphs, bidirectional search provides a way to halve the running time of Dijkstra's algorithm.\todo{sag niemals etwas sei ``easy''.
-Es gibt immer jemand der es schwer fand.
-Den schlägst du ins Gesicht.}
+On road graphs, bidirectional search provides a way to halve the running time of Dijkstra's algorithm.
 A bidirectional variant of A* therefore also seems desirable.
@@
 
-This setup provides an alternative way of integrating live traffic compared to the approach described in~\ref{sec:live-predicted-traffic}.
+This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
 Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
-Vehicle specific restrictions are accounted for in the query by adjusting $w_q$.
-For severe traffic events, the three phase approach is likely faster.
+Additional features such as turn restrictions can be implemented by adjusting $w_q$.
+For severe traffic events, the three phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
 However, it comes at the expense of a more complex algorithm.
@@
 
-\todo{Den Gedanken sollte man ausbauen oder streichen}
-One could also apply live traffic updates to $w_q$ immediately and only update $w_\ell$ once the difference between $w_\ell$ and $w_q$ grows too big or when some edge weight update would set $w_q < w_\ell$.
+% One could also apply live traffic updates to $w_q$ immediately and only update $w_\ell$ once the difference between $w_\ell$ and $w_q$ grows too big or when some edge weight update would set $w_q < w_\ell$.
 
@@
 \centering
-\caption{Performance of different variants of bidirectional A* on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:bidir_pruning}
+\caption{
+Performance of different variants of bidirectional A* on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
+All variants alternate between the forward and the backward search.}
+\label{tab:bidir_pruning}
 \input{table/bidir_pruning.tex}
@@
 This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
-Enabling the improved pruning for the average potential, marginally reduces the search space size and slightly increases running times.
+Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.
 
 \todo{Vergleich zwischen Tabelle 3 und 4? Warum ist uni directional langsamer? Hier sollte man schreiben, dass das Thema bei Tabelle 5 addressiert wird}
-\todo{was ist die Alternierungsstrategie von Tabelle 3?}
 

### 2021-11-15 paper/table/graphs.tex
some final DIMACs -> DIMACS

@@
 \midrule
-OSM Ger       &       16.2 &       35.4 &                          312.8 &        1\,466.7 &          10.1 \\
-DIMACs Europe &       18.0 &       42.2 &                          276.2 &        2\,480.9 &          12.4 \\
+OSM Germany   &       16.2 &       35.4 &                          298.7 &        1\,467.4 &          10.1 \\
+DIMACS Europe &       18.0 &       42.2 &                          276.2 &        2\,480.9 &          12.4 \\
 TDGer06       &        4.7 &       10.8 &                           59.2 &         331.7 &           2.7 \\

### 2021-11-15 eval/graphs.py

@@ -68,7 +68,7 @@ cch_ordering = pd.DataFrame.from_records([parse_flowcutter_partition_output(path
 table = graphs.groupby(['graph'])[['basic_customization_running_time_ms', 'contraction_running_time_ms', 'graph_build_running_time_ms',
     'perfect_customization_running_time_ms', 'respecting_running_time_ms', 'num_nodes', 'num_edges']].mean()
 
-table = table.reindex(['OSM Ger', 'DIMACS Europe', 'TDGer06', 'TDEur17', 'TDEur20'])
+table = table.reindex(['OSM Germany', 'DIMACS Europe', 'TDGer06', 'TDEur17', 'TDEur20'])
 
 table = table.join(ch_preprocessing.groupby('graph').mean()).join(cch_ordering.groupby('graph').mean())
 table['num_nodes'] = table['num_nodes'] / 1000000.0

### 2021-11-15 paper/ch_potentials.tex
improve CH visualization
pseudo code improvements

@@
 \usepackage{todonotes}
+
+\usepackage{tikz}
+\usetikzlibrary{automata,arrows,positioning,calc}
+\tikzstyle{node}=[circle,inner sep=0.5mm,minimum size=6.5mm,draw = black]
 
@@
 
+\begin{figure}
+\centering
+\begin{tikzpicture}[]
+\tikzset{auto rotate/.style={to path={let \p1=(\tikztostart),\p2=(\tikztotarget),
+\n1={atan2(\y2-\y1,\x2-\x1)},\n2={\n1-10},\n3={\n1+190}
+in (\tikztostart.\n2) -- (\tikztotarget.\n3) \tikztonodes}}}
+
+\node (s) at (0,0)     [node] {$s$};
+\node (2) at (1.5,1.5) [node] {};
+\node (3) at (3.7,1)   [node] {};
+\node (4) at (5,3)     [node] {$m$};
+\node (5) at (5.75,-0.1)  [node] {};
+\node (6) at (6.75,1.2)   [node] {};
+\node (7) at (7.75,-0.5)    [node] {};
+\node (8) at (8.5,2)   [node] {};
+\node (t) at (10,0.5)  [node] {$t$};
+
+\draw[->] (-1,-1) -- (-1,3.5) node [midway, above, sloped] { rank };
+
+\draw (s) edge[red, auto rotate] (2);
+\draw[red] (2) -- (3);
+\draw[red] (3) -- (4);
+\draw[red] (4) -- (5);
+\draw[red] (5) -- (6);
+\draw[red] (6) -- (7);
+\draw[red] (7) -- (8);
+\draw (8) edge[red, auto rotate] (t);
+
+\draw (2) edge[blue, auto rotate] (s);
+\draw[blue, dashed] (2) -- (4);
+\draw[blue, dashed] (4) -- (8);
+\draw (t) edge[blue, auto rotate] (8);
+
+\draw[dashed] (4) -- (6);
+\draw[dashed] (6) -- (8);
+\end{tikzpicture}
+\caption{
+Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is the mid node.
+}
+\label{fig:ch}
+\end{figure}
+
+A CH is a two phase technique to efficiently compute exact, shortest paths.
+For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
+In this section, we give an introduction.
+
 \begin{algorithm2e}
@@
 Make $Q$ only contain $t$ with weight $0$\;
-\While{not $Q$ empty}{
+\While{$Q$ not empty}{
   $y\leftarrow$ pop minimum element from $Q$\;
@@
 \end{algorithm2e}
-
-\begin{figure}
-\centering
-\includegraphics{fig/ch}
-% TODO nicer fig
-\caption{
-Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is the mid node.
-}
-\label{fig:ch}
-\end{figure}
-
-A CH is a two phase technique to efficiently compute exact, shortest paths.
-For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
-In this section, we give an introduction.
 
@@
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
-\SetKwFunction{Push}{push}
-\SetKwFunction{Pop}{pop}
 \SetKw{Break}{break}
@@
 	\While{$D[y] = \bot$}{
-    $S$.\Push{$y$}\;
+    Push $y$ onto $S$\;
 		\If{$P[y] = \bot$}{
@@
   \While{$S$ not empty}{
-    $y \gets S$.\Pop{}\;
+    $y\leftarrow$ pop top element from $S$\;
 		$D[y]\gets B[y]$\;

### 2021-11-15 paper/ch_potentials.tex
consistency uv vs u,v

@@
 The search can stop when $t$ is settled.
-
-\todo{$w(uv)$ vs $w(u,v)$, egal welches aber bitte konsistent.}
 
@@
 The second downside can be partially mitigated through pruning with the composing heuristics.
-When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(x,y) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
+When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(xy) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
 the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.

### 2021-11-15 paper/ch_potentials.tex
consistently use edges instead of arcs
wordsmithing, make use of notation more consitent

@@
 \begin{algorithm2e}
-\KwData{$B[x]$: tentative distance from $x$ to target $t$}
+\KwData{$B[v]$: tentative distance from $v$ to target $t$}
 \KwData{Minimum priority queue $Q$, also called open list, ordered by tentative distances}
-$B[x] \leftarrow +\infty$ for all $x\neq t$;
-$B[t] \leftarrow 0$\;
+$B[v] \gets +\infty$ for all $v\neq t$;
+$B[t] \gets 0$\;
 Make $Q$ only contain $t$ with weight $0$\;
 \While{$Q$ not empty}{
-  $y\leftarrow$ pop minimum element from $Q$\;
-  \For{all down-edges $xy$ in $G^+_\ell$}{
-    \If{$B[x] > w_\ell(xy) + B[y]$}{
-      $B[x]\leftarrow w_\ell(xy) + B[y]$\;
-                        Add $x$ or decrease $x$'s key in $Q$ to $B[x]$\;
+  $u\gets$ pop minimum element from $Q$\;
+  \For{all down-edges $vu$ in $G^+_\ell$}{
+    \If{$B[v] > w_\ell(vu) + B[u]$}{
+      $B[v]\gets w_\ell(vu) + B[u]$\;
+                        Add $v$ or decrease $v$'s key in $Q$ to $B[v]$\;
     }
@@
 \begin{algorithm2e}
-\KwData{$P[x]$: tentative distance from $x$ to $t$}
+\KwData{$P[v]$: tentative distance from $v$ to $t$}
 Execute Algorithm~\ref{algo:ch-backward}\;
 \For{all CH levels $L$ from most to least important}{
-	\For{all up-edges $xy$ in $G^+_\ell$ with $x$ in $L$}{
-		\If{$P[x] < P[y] + w_\ell(xy)$}{
-			$P[x] \leftarrow P[y] + w_\ell(xy)$\;
+	\For{all up-edges $uv$ in $G^+_\ell$ with $u$ in $L$}{
+		\If{$P[u] < P[v] + w_\ell(uv)$}{
+			$P[u] \leftarrow P[v] + w_\ell(uv)$\;
 		}
@@
 \begin{algorithm2e}
-\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
+\KwData{$B[v]$: tentative distance from $v$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$D[v]$: memoized distance from $v$ to $t$, $\bot$ initially}
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
 \SetKwProg{Fn}{Function}{:}{}
-\Fn{\Dist{$x$}}{
-	\If{$D[x] = \bot$}{
-		$D[x]\leftarrow B[x]$\;
-		\For{all up-edges $xy$ in $G^+_\ell$}{
-                        $D[x]\leftarrow\min\{D[x],w_\ell(xy)+\Dist(y)\}$\;
-		}
-	}
-	\Return{$D[x]$}\;
+\Fn{\Dist{$u$}}{
+	\If{$D[u] = \bot$}{
+		$D[u]\leftarrow B[u]$\;
+		\For{all up-edges $uv$ in $G^+_\ell$}{
+                        $D[u]\leftarrow\min\{D[u],w_\ell(uv)+\Dist(v)\}$\;
+		}
+	}
+	\Return{$D[u]$}\;
 }
@@
 In the target selection, we first run the backward CH search from $t$ to obtain an array $B$.
-$B[x]$ is the minimum down $xt$-path distance or $+\infty$, if there is no such path.
+$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
 $B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.
 
-To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i x)$ the distance $D[x]$.
-Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_x\{w_\ell(s_i x) + D[x]\}$.
+To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i v)$ the distance $D[v]$.
+Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_v\{w_\ell(s_i v) + D[v]\}$.
 As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
 This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
-% A* with this heuristic is the basic CH-Potentials algorithm.
 
@@
 However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
-To compute the distance $D[x]$ of a node $x$ the distances of all upward neighbors must be final.
+To compute the distance $D[v]$ of a node $v$ the distances of all upward neighbors must be final.
 In Algorithm~\ref{algo:pot}, we have to compute all these distances recursively.
-However, on CCH it is sufficient to ensure that the distance of the elimination tree parent of $x$ is final.
-All upward neighbors of $x$ are by definition included in the search space, thus they are also ancestors in the elimination tree.
-Thus, if the distance of the parent in the elimination tree is final, the distances of all other upward neighbors must already be final, too.
-Thus, it is sufficient to recursively (or iteratively) ensure that the distance of the parent in the elimination tree has been computed.
+However, on CCH it is sufficient to ensure that the distance of the elimination tree parent of $v$ is final.
+All upward neighbors of $v$ are by definition included in the search space, thus they are also ancestors in the elimination tree.
+If the distance of the parent in the elimination tree is final, the distances of all other upward neighbors must already be final, too.
+Thus, it is sufficient to iteratively ensure that the distance of the parent in the elimination tree has been computed.
 As soon as a node with final distance is encountered, all ancestors are known to have their final distance.
@@
 \begin{algorithm2e}
-\KwData{$B[x]$: tentative distance from $x$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[x]$: memoized distance from $x$ to $t$, $\bot$ initially}
-\KwData{$P[x]$: parent node of $x$ in the elimination tree, parent of root is $\bot$}
+\KwData{$B[v]$: tentative distance from $v$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$D[v]$: memoized distance from $v$ to $t$, $\bot$ initially}
+\KwData{$P[v]$: parent node of $v$ in the elimination tree, parent of root is $\bot$}
 \KwData{$S$: stack with nodes to compute distances, empty initially, only used to store intermediate data}
@@
 \SetKwProg{Fn}{Function}{:}{}
-\Fn{\Dist{$x$}}{
+\Fn{\Dist{$u$}}{
   \tcp{Determine for what nodes, the $D$ needs to be filled}
-  $y \gets x$\;
-	\While{$D[y] = \bot$}{
-    Push $y$ onto $S$\;
-		\If{$P[y] = \bot$}{
+  $v \gets u$\;
+	\While{$D[v] = \bot$}{
+    Push $v$ onto $S$\;
+		\If{$P[v] = \bot$}{
       \Break\;
 		}
-    $y \gets P[y]$\;
+    $v \gets P[v]$\;
   }
@@
   \While{$S$ not empty}{
-    $y\leftarrow$ pop top element from $S$\;
-		$D[y]\gets B[y]$\;
-		\For{all up-edges $yz$ in $G^+_\ell$}{
-                        $D[y]\gets\min\{D[y],w_\ell(yz)+D[z]\}$\;
-		}
-	}
-	\Return{$D[x]$}\;
+    $v\leftarrow$ pop top element from $S$\;
+		$D[v]\gets B[v]$\;
+		\For{all up-edges $vw$ in $G^+_\ell$}{
+                        $D[v]\gets\min\{D[v],w_\ell(vw)+D[w]\}$\;
+		}
+	}
+	\Return{$D[u]$}\;
 }
@@
 In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
-The heuristic function $h_t(x)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for node $x$.
+The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for node $v$.
 
@@
 We modify A* by processing low degree nodes consecutively without pushing them into the queue.
-Our algorithm uses the undirected degree $d(x)$ of a node $x$.
-Formally, $d(x)$ is the number of nodes $y$ such that $(xy)\in E$ or $(yx)\in E$.
-
-Analogous to A*, our algorithm stores for every node $x$ a tentative distance $D[x]$.
+Our algorithm uses the undirected degree $d(v)$ of a node $v$.
+Formally, $d(v)$ is the number of nodes $u$ such that $(vu)\in E$ or $(uv)\in E$.
+
+Analogous to A*, our algorithm stores for every node $v$ a tentative distance $D[v]$.
 Additionally, it maintains a minimum priority queue.
@@
 
-Our algorithm differs from A* when removing a node $x$ from the queue.
-A* iterates over the outgoing arcs $(xy)$ of $x$ and tries to reduce $D[y]$ by relaxing $(xy)$.
-If A* succeeds, $y$'s weight in the queue is set to $D[y]+h_t(y)$.
-Our algorithm, however, behaves differently, if $d(y)\le 2$.
-Our algorithm determines the longest degree two chain of nodes $x y_1,\ldots, y_k, z$ such that $d(y_i)=2$ and $d(z) > 2$.
-If our algorithm succeeds in reducing $D[y_1]$, it does not push $y_1$ into the queue.
-Instead, it iteratively tries to reduce all $D[y_i]$.
-If it does not reach $z$, then only $D$ is modified but no queue action is performed.
-If $D[z]$ is modified and $d(z)>2$, $z$'s weight in the queue is set to $D[z]+h_t(z)$.
+Our algorithm differs from A* when removing a node $u$ from the queue.
+A* iterates over the outgoing edges $(uv)$ of $u$ and tries to reduce $D[v]$ by relaxing $(uv)$.
+If A* succeeds, $v$'s weight in the queue is set to $D[v]+h_t(v)$.
+Our algorithm, however, behaves differently, if $d(v)\le 2$.
+Our algorithm determines the longest degree two chain of nodes $u, v_1,\ldots, v_k, w$ such that $d(v_i)=2$ and $d(w) > 2$.
+If our algorithm succeeds in reducing $D[v_1]$, it does not push $v_1$ into the queue.
+Instead, it iteratively tries to reduce all $D[v_i]$.
+If it does not reach $w$, then only $D$ is modified but no queue action is performed.
+If $D[w]$ is modified and $d(w)>2$, $w$'s weight in the queue is set to $D[w]+h_t(w)$.
 
@@
 We can also skip some degree three nodes.
-Denote by $x y_1,\ldots, y_k, z$ a degree two chain as described in the previous section.
-If $d(z) > 3$ or $z$ is in the queue, our algorithm proceeds as in the previous section.
-Otherwise, there exist up to two degree chains $z a_1,\ldots,a_p b$ and $z,\alpha_1,\ldots,\alpha_q,\beta$ such that $a_1\neq y_k \neq \alpha_1$.
-Our algorithm iteratively tries to reduce all $D[a_i]$ and $D[\alpha_i]$.
-If it reaches $\beta$, $\beta$'s weight in the queue is set to $D[\beta]+h_t(\beta)$.
-Analogously, if $b$ is reached, $b$'s weight is set to $D[b]+h_t(b)$.
-If $b$ respectively $\beta$ are not reached, our algorithm does nothing.
+Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
+If $d(w) > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
+Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq y_k \neq a_1$.
+Our algorithm iteratively tries to reduce all $D[x_i]$ and $D[a_i]$.
+If it reaches $b$, $b$'s weight in the queue is set to $D[b]+h_t(b)$.
+Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
+If $y$ respectively $b$ are not reached, our algorithm does nothing.
 
@@
 
-The straightforward idea is to use two heuristics $\overrightarrow{h}_t(x)$ and $\overleftarrow{h}_s(x)$.
-The forward search has its queue ordered by $\overrightarrow{D}[x] + \overrightarrow{h}_t(x)$, where $\overrightarrow{h}_t(x)$ estimates the distance $\dist(x,t)$ from $x$ to $t$.
-Similarly, the backward search has its queue ordered by $\overleftarrow{D}[x] + \overleftarrow{h}_s(x)$, where $\overleftarrow{h}_s(x)$ is an estimate of the distance $\dist(s,x)$ from $s$ to $x$.
+The straightforward idea is to use two heuristics $\overrightarrow{h}_t(v)$ and $\overleftarrow{h}_s(v)$.
+The forward search has its queue ordered by $\overrightarrow{D}[v] + \overrightarrow{h}_t(v)$, where $\overrightarrow{h}_t(v)$ estimates the distance $\dist(v,t)$ from $v$ to $t$.
+Similarly, the backward search has its queue ordered by $\overleftarrow{D}[v] + \overleftarrow{h}_s(v)$, where $\overleftarrow{h}_s(v)$ is an estimate of the distance $\dist(s,v)$ from $s$ to $v$.
 We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.
@@
 This allows stopping the searches when $\overrightarrow{k} + \overleftarrow{k} \geq \mu$.
-Formally, $h_{\overline{st}}(x)$ is defined as $(\overrightarrow{h}_t(x) - \overleftarrow{h}_s(x))/2$.
-The forward search uses $h_{\overline{st}}(x)$ as its heuristic.
-The backward search uses $-h_{\overline{st}}(x)$.
+Formally, $h_{\overline{st}}(v)$ is defined as $(\overrightarrow{h}_t(v) - \overleftarrow{h}_s(v))/2$.
+The forward search uses $h_{\overline{st}}(v)$ as its heuristic.
+The backward search uses $-h_{\overline{st}}(v)$.
 Unfortunately, average potentials have two downsides.
@@
 Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
-$h_{\overline{st}}(x)$ is a worse estimate for $\dist(x,t)$ than $\overrightarrow{h}_t$.
-Similarly, $-h_{\overline{st}}(x)$ is a worse estimate for $\dist(s,x)$ than $\overleftarrow{h}_s$.
+$h_{\overline{st}}(v)$ is a worse estimate for $\dist(v,t)$ than $\overrightarrow{h}_t$.
+Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$.
 The second downside can be partially mitigated through pruning with the composing heuristics.
-When the forward search scans an edge $xy$ where $\overrightarrow{D}[x] + w(xy) + \overrightarrow{h}_s(y) > \mu$ holds, i.e.
+When the forward search scans an edge $uv$ where $\overrightarrow{D}[u] + w(uv) + \overrightarrow{h}_s(v) > \mu$ holds, i.e.
 the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.
@@
 
-Let $xy$ be an edge that we relax in the forward search.
-Before pushing $y$ into the queue, we apply the new pruning rule.
-If we can prove that every path using $xy$ is at least as long as the shortest known path length $\mu$, then we do not have to push $y$.
-We therefore want to obtain a lower bound for $\dist(s,x)+w(xy)+\dist(y,t)$.
-As $x$ was settled, $\overrightarrow{D}[x]$ contains the shortest path length $\dist(s,x)$, i.e., $\dist(s,x)=\overrightarrow{D}[x]$.
-$w(xy)$ is also exactly known as it's just an edge weight.
-It remains to lower bound $\dist(y,t)$.
-Nodes are removed from the backward queue ordered by $\overleftarrow{h}(y) + \dist(y,t)$.
-If $y$ was not yet removed from the backward queue, we know that $\overleftarrow{h}(y) + \dist(y,t) \geq \overleftarrow{k}$.
-This gives us the required lower bound, i.e. $\dist(y,t) \geq \overleftarrow{k}-\overleftarrow{h}(y)$.
-Thus, $y$ does not have to be pushed if $\overrightarrow{D}[x]+w(xy)+\overleftarrow{k}-\overleftarrow{h}(y)\ge \mu$.
-The node $y$ might still be pushed into the queue when there is another edge $zy$ for which the pruning is not possible.
+Let $uv$ be an edge that we relax in the forward search.
+Before pushing $v$ into the queue, we apply the new pruning rule.
+If we can prove that every path using $uv$ is at least as long as the shortest known path length $\mu$, then we do not have to push $v$.
+We therefore want to obtain a lower bound for $\dist(s,u)+w(uv)+\dist(v,t)$.
+As $u$ was settled, $\overrightarrow{D}[u]$ contains the shortest path length $\dist(s,u)$, i.e., $\dist(s,u)=\overrightarrow{D}[u]$.
+$w(uv)$ is also exactly known as it's just an edge weight.
+It remains to lower bound $\dist(v,t)$.
+Nodes are removed from the backward queue ordered by $\overleftarrow{h}(v) + \dist(v,t)$.
+If $v$ was not yet removed from the backward queue, we know that $\overleftarrow{h}(v) + \dist(v,t) \geq \overleftarrow{k}$.
+This gives us the required lower bound, i.e. $\dist(v,t) \geq \overleftarrow{k}-\overleftarrow{h}(v)$.
+Thus, $v$ does not have to be pushed if $\overrightarrow{D}[u]+w(uv)+\overleftarrow{k}-\overleftarrow{h}(v)\ge \mu$.
+The node $v$ might still be pushed into the queue when there is another edge $wv$ for which the pruning is not possible.
 Checking the pruning rule requires evaluating the backward heuristic.
@@
 Our implementation stores edge travel times using piece-wise linear functions.
-The A* search uses the tentative distance $\tau$ at a node $x$ when to evaluating the travel time of outgoing edges $x y$.
+The A* search uses the tentative distance $\tau$ at a node $u$ when to evaluating the travel time of outgoing edges $u v$.
 This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
@@
 \caption{
-Running times of incremental Lazy RPHAST with up to $|S| = 2^{14}$ sources picked from a ball of varying size $|B|$ on OSM Germany excluding selection times.
-The upper figure contains the total running time.
-The lower figure contains the averaged running time per query, i.e. $y/x$ from the upper figure.
+Average running times of incremental Lazy RPHAST while querying $|S| = 2^{14}$ from a ball of varying size $|B|$ on OSM Germany excluding selection times.
+The upper figure contains the total elapsed running time.
+The lower figure contains the averaged running time per source, i.e. $y/x$ from the upper figure.
 Note the different y-axis scales and units.
@@
 We compare (C)CH-Potentials to three other heuristics.
-The first heuristic is, the \emph{Zero} heuristic where $h(x)$ is $0$ for all nodes $x$.
+The first heuristic is, the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
 This corresponds to using Dijkstra's algorithm.

### 2021-11-15 paper/ch_potentials.tex
Figure descriptions

@@
 \begin{document}
-
-\bibliographystyle{plainurl} % the mandatory bibstyle
 
@@
 \label{img:search-space}
+\Description[A map where nodes visited by A* are marked.]{
+A map of Karlsruhe where nodes visited by A* are marked.
+The shape of the area with the visited nodes is similar to a drop.
+The source node roughly lies in the center of the drop, the target node at the tip.
+}
 \end{figure}
@@
 \label{fig:ch}
+\Description[Nodes and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
+Nodes and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
+The y-position on the figure indicates the rank of each node.
+Where the original shortest path down and then up, a shortcut edge connects the two higher nodes.
+The path in the augmented graph goes only up and then down.
+}
 \end{figure}
@@
 }\label{fig:lazy_rphast_inc}
+\Description{
+Two line graphs with each eight lines for different ball sizes and algorithms.
+The x-axis in both graphs is the number of nodes that were queried up to that point in the incremental query.
+In each graph, all lines have a similar shape and appear as scaled variants of each other.
+The bigger the ball size the slower the whole algorithm.
+The CCH based variant is somewhat slower for the most part.
+Only for small ball sizes it is marginally faster when computing distances for more then 4096 sources.
+When the ball is the entire graph, the CH based variant takes around 2.5ms for the first 512 sources, 5ms for 2048 sources and 25ms for 16384 sources.
+Conversely, the average time per query is around 100mus for the first few queries, then drops to around 5mus after 512 sources, down to 2mus after 16384 sources.
+For the smallest ball size of 16384, everything is 1.5 orders of magnitude faster.
+}
 \end{figure}
@@
 }\label{fig:many_to_one}
+\Description{
+A double logarithmic line plot with four lines with similar shapes and absolute values.
+Running times for both algorithms and graphs with a ball size of 16384 is around 1ms.
+For a ball size of 2 to the 20 CH based variants need 10ms and CCH based variants slightly more.
+For a ball size of 2 to the 24 CH based variants are at 30ms and CCH based variants slightly above 40ms.
+}
 \end{figure}
@@
 }\label{fig:scaled_weights}
+\Description{Fully described in the text.}
 \end{figure}

### 2021-11-16 paper/ch_potentials.tex
more polishing

@@
 A heuristic function is called \emph{feasible}, if $w_{h_t}(e) \geq 0$ for all edges $e$.
-If the employed heuristic is feasible, A* is guaranteed to find optimal distances.
+If the employed heuristic is feasible, A* is guaranteed to have found the optimal distance after settling $t$.
 In this article, we limit our discussion to feasible heuristics.
@@
 
-In this section, we closely investigate the performance of bidirectional A*.
+In this section, we investigate the performance of bidirectional A*.
+We first evaluate different variants of bidirectional A* in Table~\ref{tab:bidir_pruning} and~\ref{tab:bidir_switching} then compare the best ones against unidirectional A* in Table~\ref{tab:bidir}.
 Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A* with and without our improved pruning.
@@
 Symmetric bidirectional A* without our improved pruning has the worst performance.
-Enabling the improved pruning brings the performance symmetric bidirectional A* on par with the average potential for all heuristics except ALT.
-\todo{den nächsten Satz verstehe ich nicht}
-Apparently, the average potential gives ALT a better heuristic than the original one for for each direction individually, contrary to the statements made in~\cite{gh-cspas-05}.
-For all other heuristics, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
+Enabling the improved pruning significantly improves the performance of symmetric bidirectional A*.
+For all heuristics except ALT, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
 Without the low degree improvements, the improved symmetric variant is marginally faster and with the low degree improvements the average potential remains slightly faster.
@@
 Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.
-
-\todo{Vergleich zwischen Tabelle 3 und 4? Warum ist uni directional langsamer? Hier sollte man schreiben, dass das Thema bei Tabelle 5 addressiert wird}
 

### 2021-11-16 eval/bidir_features.py
consistency in unidir results

@@ -118,6 +118,18 @@ with open("paper/table/bidir_switching.tex", 'w') as f:
   f.write(output)
 
 
+unidir_data = [json.load(open(path)) for path in glob.glob("exp/building_blocks/*.json")]
+unidir_queries = pd.DataFrame.from_records([{
+    **algo,
+    'BCC': run['bcc_core'],
+    'Skip Deg. 2': run['skip_deg2'],
+    'Skip Deg. 3': run['skip_deg3'],
+    'Heuristic': run['potential'],
+    'num_queue_pops': algo['num_queue_pops'] + algo.get('core_search', {}).get('num_queue_pops', 0),
+    'num_queue_pushs': algo['num_queue_pushs'] + algo.get('core_search', {}).get('num_queue_pushs', 0),
+    'num_relaxed_arcs': algo['num_relaxed_arcs'] + algo.get('core_search', {}).get('num_relaxed_arcs', 0),
+} for run in unidir_data for algo in run['algo_runs'] if algo.get('algo') in ['Virtual Topocore Component Query', 'Dijkstra Query']])
+unidir_queries['num_queue_pushs_k'] = unidir_queries['num_queue_pushs'] / 1000
 
 bidir = queries.loc[lambda x: x['algo'].str.contains('Topocore')] \
     .loc[lambda x: x['choose_direction_strategy'] != 'min_key'] \
@@ -127,6 +139,8 @@ bidir = queries.loc[lambda x: x['algo'].str.contains('Topocore')] \
     .reindex(index=['weight_scale', 'probabilistic_scale_by_speed'], level=0) \
     .reindex(index=[1.0], level=2) \
     .reindex(index=['unidir', 'Average', 'Symmetric'], level=3)
+
+bidir.loc[('weight_scale', 1.05, 1.00, 'unidir')] = unidir_queries.groupby(['Heuristic', 'BCC', 'Skip Deg. 2', 'Skip Deg. 3'])[['running_time_ms', 'num_queue_pushs_k']].mean().unstack(0).loc[(True, True, True)]
 
 bidir[('num_queue_pushs_k', '(C)CH/Oracle')] = bidir[('num_queue_pushs_k', 'CH')]
 bidir = bidir.drop([('num_queue_pushs_k', 'CH'), ('num_queue_pushs_k', 'CCH'), ('num_queue_pushs_k', 'Oracle')], axis = 'columns')

### 2021-11-16 eval/applications.py
style app table

@@ -8,7 +8,6 @@ import os
 import re
 
 base = "exp/"
-cwd = os.getcwd()
 paths = glob.glob(base + "applications/*.json")
 data = [json.load(open(path)) for path in paths]
 
@@ -20,113 +19,111 @@ def path_to_graph(path):
         'osm_europe': 'OSM Europe',
         'osm_ger': 'OSM Ger',
         'osm_ger_td': 'OSM Ger',
-        'europe': 'DIMACS Eur',
+        'europe': 'DIMACs Eur',
     }[[x for x in path.split('/') if x != ''][-1]]
 
-only_public = 'ONLY_PUBLIC' in os.environ
-
+algos = ['Virtual Topocore Component Query', 'Virtual Topocore Bidirectional Core Query']
 queries = pd.DataFrame.from_records([{
+    **algo,
     'exp': exp['experiment'],
-    'graph': run['args'][1],
-    **algo }
-    for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if algo.get('algo') == 'CH Potentials Query'])
+    'graph': path_to_graph(run['args'][1]),
+    'num_queue_pops': algo.get('num_queue_pops', 0) + algo.get('core_search', {}).get('num_queue_pops', 0),
+    'num_queue_pushs': algo.get('num_queue_pushs', 0) + algo.get('core_search', {}).get('num_queue_pushs', 0),
+    'num_relaxed_arcs': algo.get('num_relaxed_arcs', 0) + algo.get('core_search', {}).get('num_relaxed_arcs', 0)
+    } for run in data for exp in run.get('experiments', []) for algo in exp['algo_runs'] if (algo.get('algo') in algos or ('algo' not in algo and 'biggest_bcc_size' not in algo))])
 
 queries = queries.append(pd.DataFrame.from_records([{
+    **algo,
     'exp': run['program'],
-    'graph': run['args'][1],
-    **algo }
-    for run in data for algo in run.get('algo_runs', []) if algo.get('algo') == 'CH Potentials Query']))
+    'graph': path_to_graph(run['args'][1]),
+    'num_queue_pops': algo.get('num_queue_pops', 0) + algo.get('core_search', {}).get('num_queue_pops', 0),
[…]

### 2021-11-16 paper/ch_potentials.tex
grammar checking

@@
 Extending them to certain practical applications can be hard.
-In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as a A* heuristic.
+In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as an A* heuristic.
 As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
@@
 For example, we are not aware of any work combining all features mentioned above.
-In this article, we describe an algorithmic building block, that allows handling the combination of all above mentioned features -- and probably more.
+In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features -- and probably more.
 
@@
 For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
-The proposed approach achieves fast queries but preprocessing becomes an order of magnitude slower compared to classical CH.
+The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower compared to classical CH.
 
@@
 
-In the this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
+In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
 Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of nodes ordered by increasing distance from $s$.
@@
 The two searches meet at the mid node.
-Pseudo-code for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
+Pseudocode for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
 The forward search works analogously.
@@
 
-Some hierarchical techniques were extended to support a three phase setup.
+Some hierarchical techniques were extended to support a three-phase setup.
 The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
@@
 This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
-The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}
-
-Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three phase setup.
+The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}.
+
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three-phase setup.
 The key idea is to use a node importance order that is independent of the weight function.
@@
 After all levels are processed, the shortest distances from all nodes to $t$ were computed.
-Pseudo-code is provided in Algorithm~\ref{algo:phast}.
+Pseudocode is provided in Algorithm~\ref{algo:phast}.
 PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
@@
 If the source set changes often, selection times may become problematic.
-RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
-Implemented with SIMD instructions it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.
+% RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
+% Implemented with SIMD instructions it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.
 
@@
 In this section, we discuss another very specific problem setting which naturally arises from A* heuristics.
-Here, the target node is known in the selection step but the source nodes are only queried one after another.
+Here, the target node is known in the selection step, but the source nodes are only queried one after another.
 We denote this problem as the \emph{incremental many-to-one} problem.
@@
 Additionally, it maintains a minimum priority queue.
-Diverging from A*, not all nodes can be pushed but every node has a tentative distance.
+Diverging from A*, not all nodes need be pushed, but every node has a tentative distance.
 
@@
 Instead, it iteratively tries to reduce all $D[v_i]$.
-If it does not reach $w$, then only $D$ is modified but no queue action is performed.
+If it does not reach $w$, then only $D$ is modified, but no queue action is performed.
 If $D[w]$ is modified and $d(w)>2$, $w$'s weight in the queue is set to $D[w]+h_t(w)$.
@@
 We do not use the average heuristic.
-Instead of a strong stopping criteria, we use a pruning rule that gets stronger the longer the search runs.
+Instead of a strong stopping criterion, we use a pruning rule that gets stronger the longer the search runs.
 Such a pruning rule will eventually prune all remaining branches and stop the search.
@@
 In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
-This sounds sensible but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
+This sounds sensible, but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
 
@@
 
-% Ich bleibe bei v_i hier, damit keine Verwechselung mit den freien Variablen x,y,z gibt.
 A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
@@
 Our implementation stores edge travel times using piece-wise linear functions.
-The A* search uses the tentative distance $\tau$ at a node $u$ when to evaluating the travel time of outgoing edges $u v$.
+The A* search uses the tentative distance $\tau$ at a node $u$ when evaluating the travel time of outgoing edges $u v$.
 This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
@@
 Additional features such as turn restrictions can be implemented by adjusting $w_q$.
-For severe traffic events, the three phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
+For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
 However, it comes at the expense of a more complex algorithm.
@@
 The CCH based variant is somewhat slower for the most part.
-Only for small ball sizes it is marginally faster when computing distances for more then 4096 sources.
+Only for small ball sizes it is marginally faster when computing distances for more than 4096 sources.
 When the ball is the entire graph, the CH based variant takes around 2.5ms for the first 512 sources, 5ms for 2048 sources and 25ms for 16384 sources.
@@
 Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
-(C)CH-Potentials compute the same distance estimates but the heuristic evaluation has some overhead.
+(C)CH-Potentials compute the same distance estimates, but the heuristic evaluation has some overhead.
 Comparing against Oracle-A* allows us to measure this overhead.
@@
 Here, bidirectional search will only introduce unnecessary overhead.
-When query weights are scaled up uniformly, bidirectional search achieves some search space reduction but it is not enough to significantly reduce running times due to the overhead of running a second search.
+When query weights are scaled up uniformly, bidirectional search achieves some search space reduction, but it is not enough to significantly reduce running times due to the overhead of running a second search.
 This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
@@
 The achieved speedups are therefore comparable to CH speedups.
-Since turn costs an restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speed-ups.
+Since turn costs and restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speed-ups.
 Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.

### 2021-11-17 paper/ch_potentials.tex
polishing

@@
 Extending them to certain practical applications can be hard.
-In this paper, we present Lazy RPHAST, an algorithm to efficiently compute many distance estimates towards a single target and use it as an A* heuristic.
+In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many nodes towards a single target and use it as an A* heuristic.
 As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
@@
 In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
-After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and our algorithm to solve it efficiently in Section~\ref{sec:lazy-rphast}.
-CH-Potentials, our new distance estimation function is introduced in Section~\ref{sec:main-algo}.
+After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and Lazy RPHAST, our algorithm to solve it efficiently, in Section~\ref{sec:lazy-rphast}.
+CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
 There, we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
@@
 
-A CH is a two phase technique to efficiently compute exact, shortest paths.
+A CH is a two-phase technique to efficiently compute exact, shortest paths.
 For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
@@
 An \emph{up path} is a path where the last node is the mid node.
-Similarly, the first node is the mid node of a \emph{down path}.
+Similarly, for a \emph{down path}, the first node has to be the mid node.
 % Every up and down path is an up-down path.
@@
 Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
-This is done by repeatedly contracting the least important node and adding shortcuts between its neighbors.
+This is done by repeatedly contracting the least important node and adding shortcuts between its higher ranked neighbors.
 %See~\cite{gssv-erlrn-12} for the details.
@@
 The preprocessing phase is independent of the weights and only exploits the graph topology.
-Weights are introduces in the customization phase.
+Weights are introduced in the customization phase.
 Paths are computed in the query phase.
@@
 The CH query algorithms can therefore be applied without modifications.
-However, the CCH preprocessing grants some additional useful properties.
+However, the CCH preprocessing grants some useful additional properties.
 
@@
 If $d(w) > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
-Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq y_k \neq a_1$.
+Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq v_k \neq a_1$.
 Our algorithm iteratively tries to reduce all $D[x_i]$ and $D[a_i]$.
@@
 This breaks the usual bidirectional Dijkstra stopping criterion.
-To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
+To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for either direction.
 The forward search can skip nodes already settled by the backward search and vice versa.
@@
 $h_{\overline{st}}(v)$ is a worse estimate for $\dist(v,t)$ than $\overrightarrow{h}_t$.
-Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$.
+Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$\footnote{
+To obtain an actual lower bound from this average heuristic, one has to add $\overleftarrow{h}_s(t)/2$ in the forward case and $\overrightarrow{h}_t(s)/2$ in the backward.
+Adding any constant to a heuristic function does not change the reduced graph.
+}.
 The second downside can be partially mitigated through pruning with the composing heuristics.
@@
 
-To enable quick shortest path computations, we consider a two phase setup with an additional off-line preprocessing phase before the on-line query phase.
+To enable quick shortest path computations, we consider a two-phase setup with an additional off-line preprocessing phase before the on-line query phase.
 The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
@@
 In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
-We need to make sure that the modified travel time function fulfills the no-waiting property.
+We need to make sure that the modified travel time function fulfills the FIFO property.
 For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
@@
 % We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
-In our implementation, we to not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
-We set $w_\ell$ again to the freeflow travel time.
+In our implementation, we do not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
+We set $w_\ell$ to the freeflow travel time.
 
@@
 However, it comes at the expense of a more complex algorithm.
-Further, integrating predicted traffic becomes difficult, as $w_q$ may not drop below the $w_\ell$ with integrated congestion.
+Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.
 
@@
 The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
-We conclude that that Lazy RPHAST is a very useful extension to RPHAST.
-It allows efficiently handling dynamic source sets and is even faster than RPHAST in the setting where both the target and the source set change between queries.
+While the numbers are not perfectly comparable due to different benchmark machines, we can still conclude that Lazy RPHAST is a very useful extension to RPHAST.
+It allows efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
 % This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
@@
 We compare (C)CH-Potentials to three other heuristics.
-The first heuristic is, the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
+The first heuristic is the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
 This corresponds to using Dijkstra's algorithm.
@@
 We first evaluate different variants of bidirectional A* in Table~\ref{tab:bidir_pruning} and~\ref{tab:bidir_switching} then compare the best ones against unidirectional A* in Table~\ref{tab:bidir}.
-Table~\ref{tab:bidir_pruning} depicts the performance of different variants of bidirectional A* with and without our improved pruning.
+Table~\ref{tab:bidir_pruning} studies the impact of our improved pruning and the low degree optimizations.
 As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
@@
 In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
-Bidirectional search for CH-Potentials only pays of when the reduced weights exhibit certain very specific patterns.
+Bidirectional search for CH-Potentials only pays off when the reduced weights are bad around the terminals.
 
@@
 Such large speedups are typical for CH.
-This shows that CH-Potentials gracefully converges toward a CH in the $w_q = w_\ell$ special case.
+This shows that CH-Potentials gracefully converge toward CH in the $w_q = w_\ell$ special case.
 On DIMACS Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.

### 2021-11-17 paper/ch_potentials.tex
update CATCHUp Citation

@@
 Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
-Recently, a new approach has been published~\cite{swz-sfert-20} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
+Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
 
@@
 Recurring traffic can be predicted by observing the traffic in the past.
-It is common \cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-20} to represent these predictions as \emph{travel time functions}.
+It is common \cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-21} to represent these predictions as \emph{travel time functions}.
 An edge weight is no longer a scalar value but a function that maps the entry time onto the traversal time.
@@
 % With this setup, we extended CH-Potentials to support time-dependent routing.
-By keeping travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-20,dn-crdtd-12} which have to create shortcuts of travel time functions.
+By keeping travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-21,dn-crdtd-12} which have to create shortcuts of travel time functions.
 

### 2021-12-15 Rakefile.rb
clean up for public exps

@@ -93,7 +93,8 @@ osm_eur = "#{data_dir}/osm_europe/"
 dimacs_graph = "#{data_dir}/europe/"
 
 # static_graphs = [osm_ger, osm_eur, dimacs_graph]
-static_graphs = [osm_ger, dimacs_graph]
+static_graphs = [osm_ger]
+static_graphs += [dimacs_graph] unless only_public
 td_graphs = []
 unless only_public
   td_graphs += [
@@ -134,6 +135,9 @@ namespace "prep" do
         sh "#{wd}/code/osm_import/build/import_osm #{osm_ger_src_file} #{Dir[live_dir + '*'].join(' ')} #{typical_file}"
       end
     end
+    Dir.chdir "code/rust_road_router" do
+      sh "cargo run --release --bin write_unit_files -- #{osm_ger} 1000 1"
+    end
   end
 
   file "#{osm_ger}/live_travel_time" => osm_ger do
@@ -157,6 +161,9 @@ namespace "prep" do
     wd = Dir.pwd
     Dir.chdir osm_eur do
       sh "#{wd}/code/osm_import/build/import_osm #{osm_eur_src_file}"
+    end
+    Dir.chdir "code/rust_road_router" do
+      sh "cargo run --release --bin write_unit_files -- #{osm_eur} 1000 1"
     end
   end
 
@@ -271,7 +278,7 @@ namespace "exp" do
       [[], ['chpot-improved-pruning']].each do |pruning_feats|
         [[], ['chpot-alt'], ['chpot-cch'], ['chpot-oracle'], ['chpot-only-topo']].each do |pot_feats|
           if pot_feats == ['chpot-only-topo']
-            sh "CHPOT_NUM_QUERIES=1000 cargo run --release --bin chpot_bidir #{features(pruning_feats + pot_feats)} -- #{osm_ger} > #{exp_dir}/bidir_features/$(date --iso-8601=seconds).json"
+            sh "CHPOT_NUM_QUERIES=#{(ENV['NUM_DIJKSTRA_QUERIES'] || 1000).to_i} cargo run --release --bin chpot_bidir #{features(pruning_feats + pot_feats)} -- #{osm_ger} > #{exp_dir}/bidir_features/$(date --iso-8601=seconds).json"
           else
             sh "cargo run --release --bin chpot_bidir #{features(pruning_feats + pot_feats)} -- #{osm_ger} > #{exp_dir}/bidir_features/$(date --iso-8601=seconds).json"
           end
[…]

### 2021-12-15 eval/applications.py

@@ -79,7 +79,7 @@ table['algo'] = table['exp'].map({
     'bidir_no_tunnels': 'CH B',
 })
 table['app'] = table['exp'].map({
-    'perfect': 'Unmodified $w_q = w_{\ell}$',
+    'perfect': 'Unmodified $w_q = w_{\\ell}$',
     'chpot_td': 'TD',
     'chpot_live': 'Live',
     'chpot_turns': 'Turns',
@@ -96,15 +96,17 @@ table['app'] = table['exp'].map({
     'bidir_no_tunnels': 'No Tunnels',
 })
 table = table.set_index(['graph', 'app', 'algo']) \
-    .reindex(index=['Unmodified $w_q = w_{\ell}$', 'No Tunnels', 'No Highways', 'Live', 'Turns', 'Live + Turns', 'TD', 'TD + Live', 'TD + Live + Turns'], level=1) \
+    .reindex(index=['Unmodified $w_q = w_{\\ell}$', 'No Tunnels', 'No Highways', 'Live', 'Turns', 'Live + Turns', 'TD', 'TD + Live', 'TD + Live + Turns'], level=1) \
     .reindex(index=['CH U', 'CH B', 'CCH U', 'CCH B'], level=2) \
     .drop(columns=['exp'])
 table.at[('OSM Ger', 'No Tunnels', 'CH B'), 'dijkstra_running_time_ms'] = table.at[('OSM Ger', 'No Tunnels', 'CH U'), 'dijkstra_running_time_ms']
 table.at[('OSM Ger', 'No Highways', 'CH B'), 'dijkstra_running_time_ms'] = table.at[('OSM Ger', 'No Highways', 'CH U'), 'dijkstra_running_time_ms']
-table.at[('OSM Ger', 'Live', 'CH B'), 'dijkstra_running_time_ms'] = table.at[('OSM Ger', 'Live', 'CH U'), 'dijkstra_running_time_ms']
-table.at[('OSM Ger', 'Live', 'CCH U'), 'dijkstra_running_time_ms'] = table.at[('OSM Ger', 'Live', 'CH U'), 'dijkstra_running_time_ms']
+if 'ONLY_PUBLIC' not in os.environ:
+    table.at[('OSM Ger', 'Live', 'CH B'), 'dijkstra_running_time_ms'] = table.at[('OSM Ger', 'Live', 'CH U'), 'dijkstra_running_time_ms']
[…]

### 2022-07-21 paper/ch_potentials.tex
update affiliation
relaxation fix
integrate a lot of reviewer feedback

@@
 
-\newcommand*{\dist}{\mathcal{D}}
+\newcommand*{\dist}{\operatorname{dist}}
+\newcommand*{\gchu}{G^{\uparrow}}
+\newcommand*{\gchd}{G^{\downarrow}}
+\newcommand*{\rgchd}{\overleftarrow{G^{\downarrow}}}
+\newcommand*{\rgchu}{\overleftarrow{G^{\uparrow}}}
+\newcommand*{\echu}{E^{\uparrow}}
+\newcommand*{\echd}{E^{\downarrow}}
+\newcommand*{\rechd}{\overleftarrow{E^{\downarrow}}}
+\newcommand*{\rechu}{\overleftarrow{E^{\uparrow}}}
 
@@
 \email{academia@ben-strasser.net}
-\affiliation{}
+\affiliation{%
+  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
+  \city{Karlsruhe}
+  \country{Germany}
+}
 
@@
 We study exact, efficient and practical algorithms for route planning applications in large road networks.
-Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the exact application.
+Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the specific application.
 While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
@@
 Extending them to certain practical applications can be hard.
-In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many nodes towards a single target and use it as an A* heuristic.
+In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many vertices towards a single target and use it as an A* heuristic.
 As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
 Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
-Additionally, we describe A* optimizations to accelerate the processing of low degree nodes, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
-An extensive experimental study confirms the practicality of our approach for many applications.
+Additionally, we describe A* optimizations to accelerate the processing of low degree vertices, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
+An extensive experimental study confirms the practical applicability of our approach for many applications.
 \end{abstract}
@@
 \label{sec:intro}
-The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
+The past decade has seen a plethora of research works on route planning in large road networks~\cite{bdgmpsww-rptn-16}.
 Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
-Nodes represent intersections.
+Vertices represent intersections.
 Roads are modeled using edges.
@@
 \includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
-\caption{Nodes explored by A*. Color indicates the node removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
+\caption{Vertices explored by A*. Color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
 \label{img:search-space}
-\Description[A map where nodes visited by A* are marked.]{
-A map of Karlsruhe where nodes visited by A* are marked.
-The shape of the area with the visited nodes is similar to a drop.
-The source node roughly lies in the center of the drop, the target node at the tip.
+\Description[A map where vertices visited by A* are marked.]{
+A map of Karlsruhe where vertices visited by A* are marked.
+The shape of the area with the visited vertices is similar to a drop.
+The source vertex roughly lies in the center of the drop, the target vertex at the tip.
 }
@@
 Extending hierarchical speed-up techniques is often possible, too.
-However, the algorithm development is vastly more complex.
+However, the algorithm development is vastly more complex. % TODO cite some papers
 For every feature, dedicated research papers exist that extend CH and other approaches.
@@
 For example, we are not aware of any work combining all features mentioned above.
-In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features -- and probably more.
+In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features.
 
@@
 A* is a goal-directed variant of Dijkstra's algorithm.
-See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
+See Figure~\ref{img:search-space} for an example of vertices traversed during an A* search.
 A* uses a \emph{heuristic} to guide the search towards the goal.
-A heuristic is a function that maps a node to an estimate of the remaining distance to the goal.
+A heuristic is a function that maps a vertex to an estimate of the remaining distance to the goal.
 A*'s running time crucially depends on how tight this estimate is.
@@
 
-This article is the extended version of our conference paper ``A Fast and Tight Heuristic for A* in Road Networks''~\cite{strasser_et_al:LIPIcs.SEA.2021.6}.
-The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree nodes, common in road networks, and some applications of CH-Potentials.
+% TODO rewrite, make self contained
+The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree vertices, common in road networks, and some applications of CH-Potentials.
 In this article, we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
@@
 CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
-There, we also discuss our improvements to the handling of low-degree nodes (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
+There, we also discuss our improvements to the handling of low-degree vertices (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
 % Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
-In Section~\ref{sec:extensions}, we demonstrate CH-Potential's flexibility, by describing how to apply the approach to different practical applications.
+In Section~\ref{sec:extensions}, we demonstrate flexibility of our approach, by describing how to apply the approach to different practical applications.
 Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
 
-\section{Related Work}\label{sec:related_work}
+\subsection{Related Work}\label{sec:related_work}
 
@@
 Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
+% TODO argue what is complicated to implement?
 
@@
 
-In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH).
+In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH). % TODO paragraph looks ugly, somehow make more compact
 CCH also has a second preprocessing phase where weights can be altered.
 Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
-% Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
+Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
 
@@
 
-While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
-Further, in every extension the flexibility available at query time is fairly limited.
-Combining these hierarchical extensions is an unsolved problem.
-
-CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
+Clearly, extending hierarchical approaches is possible, but it requires significant research effort.
+% While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
+% TODO Reviewer1 similarly, discuss why "they also show it is non trivial".
+What does this mean?
+Further, extensions to these speedup techniques cannot easily be composed with each other.
+Combining these extensions is an unsolved problem.
+
+CH-Potentials is not the first technique to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
 However, previous works mostly focused on accelerating hierarchical approaches further rather than exploiting A*'s flexibility.
@@
 CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
-A CPD can quickly compute the first edge of a shortest path between any two nodes.
+A CPD can quickly compute the first edge of a shortest path between any two vertices.
 In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
 For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
-Unfortunately, the employed CPD's quadratic preprocessing running time is problematic on large street networks.
+Unfortunately, the employed CPD's quadratic preprocessing running time is problematic on large road networks.
 %
@@
 
+\subsection{Contribution and Outline}
+
+In this work, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem (Section~\ref{sec:lazy-rphast}).
+In Section~\ref{sec:main-algo}, we present CH-Potentials, a novel fast and tight heuristic for A* in road networks build on Lazy RPHAST.
+Further, we present several useful optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
+In Section~\ref{TODO}, we introduce a unified framework to apply CH-Potentials to a variety of practical route planning problems without any modifications to the preprocessing.
+Finally, an extensive experimental evaluation in Section~\ref{sec:experiments} confirms the applicability of our results.
+
 \section{Preliminaries}\label{sec:preliminaries}
 
-We consider directed graphs $G=(V,E)$ with $n=|V|$ nodes and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}$.
-Given nodes $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
+We consider directed graphs $G=(V,E)$ with $E \subseteq V \times V$ with $n=|V|$ vertices and $m=|E|$ edges with weight functions $w : E \to \mathbb{R}^{\geq 0}$.
+We use $uv$ as a short notation for the edge $(u,v)$.
+The \emph{reversed} graph $\overleftarrow{G} := (V, \{ vu \mid uv \in E \})$ contains all edges in reverse direction.
+The corresponding reversed weight function is $\overleftarrow{w}(vu) := w(uv)$.
+We denote by $N(u) = \{ v \mid uv \in E \vee vu \in E \}$ the \emph{undirected neighborhood} of a vertex $u$ and refer to the numbers of distinct neighbors $|N(u)|$ as the \emph{degree} of $u$.
+
+Given vertices $s$ and $t$ we want to obtain a $st$-path $P=(s=v_0,\dots,t=v_k)$ of minimum weight $w(P) := \sum_{i} w(v_{i-1}v_i)$.
 We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
-If there is no path, we set $\dist_w(s,t)=\infty$.
+If there is no path, we set $\dist_w(s,t)=\infty$. % TODO edge notation, degree (undirected)
 
@@
 In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
-Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of nodes ordered by increasing distance from $s$.
+Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of vertices ordered by increasing distance from $s$.
+We denote by $k$ the minimum key in $Q$, i.e.\ the tentative distance of the closest remaining vertex.
 Initially, all distances are set to $\infty$.
-The start node distance $D[s]$ is set to zero and the queue is initialized with $s$.
-In each iteration, the closest remaining node $u$ is popped from the queue and \emph{settled}.
+The start vertex distance $D[s]$ is set to zero and the queue is initialized with $s$.
+In each iteration, the closest remaining vertex $u$ is popped from the queue and \emph{settled}.
 Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
-If the tentative distance at the head node $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
+If the tentative distance at the head vertex $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
 Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
 The search can stop when $t$ is settled.
-
-Dijkstra's algorithm can also be run from the target $t$ on the reversed graph.
+The vertices visited by Dijkstra's algorithm throughout the search are denoted as the \emph{search space}.
+
+Dijkstra's algorithm can also be run from the target $t$ on the reversed graph $\overleftarrow{G}$. % TODO introduce reverse dist etc
 We call this a \emph{backward search}.
 Running two Dijkstra searches simultaneously, one from $s$ and a backward search from $t$, until the searches meet is called \emph{bidirectional} search.
-Typically, the searches are interleaved by alternating settling a node from each direction.
-Another common approach is to always settle a node from the direction with the smaller minimum queue key.
+In this case, we denote by $\overrightarrow{D}$, $\overrightarrow{Q}$ and $\overrightarrow{k}$ the distances, queue and minimum queue key of the forward search and by $\overleftarrow{D}$, $\overleftarrow{Q}$ and $\overleftarrow{k}$ the respective data of the backward search.
+Typically, the searches are interleaved by alternating settling a vertex from each direction.
+Another common approach is to always settle a vertex from the direction with the smaller minimum queue key.
 Theoretically, any direction interleaving strategy is possible.
@@
 % A* uses a \emph{heuristic} to guide the search towards the goal.
-It uses a heuristic function $h_t(v)$ that maps a node $v$ onto an estimate of the distance from $v$ to the target $t$.
-A* orders the nodes in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as Dijkstra's algorithm does it.
-This results in nodes closer to the target being explored earlier.
+It uses a heuristic function $h_t(v)$ that maps a vertex $v$ onto an estimate of the distance from $v$ to the target $t$.
+A* orders the vertices in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as Dijkstra's algorithm does it.
+This results in vertices closer to the target being explored earlier.
 Dijkstra's algorithm is a special case of A* with $h_t(v)=0$.
@@
 \node (3) at (3.7,1)   [node] {};
-\node (4) at (5,3)     [node] {$m$};
+\node (4) at (5,3)     [node] {$v^*$};
 \node (5) at (5.75,-0.1)  [node] {};
@@
 \caption{
-Solid lines are edges in $G$. Dotted lines are shortcuts. Red is shortest $st$-path in $G$. Blue is equaly long up-down $st$-path in $G^+$. $m$ is the mid node.
+Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$. $v^*$ is the mid vertex. % TODO different letter for m
 }
 \label{fig:ch}
-\Description[Nodes and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
-Nodes and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
-The y-position on the figure indicates the rank of each node.
-Where the original shortest path down and then up, a shortcut edge connects the two higher nodes.
+\Description[Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.]{
+Vertices and edges of a shortest path in the original graph and the corresponding path in the augmented graph.
+The y-position on the figure indicates the rank of each vertex.
+Where the original shortest path down and then up, a shortcut edge connects the two higher vertices.
 The path in the augmented graph goes only up and then down.
@@
 
-A CH is a two-phase technique to efficiently compute exact, shortest paths.
+A CH is a two-phase technique to efficiently compute exact, shortest paths on road networks.
 For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
@@
 
-\begin{algorithm2e}
-\KwData{$B[v]$: tentative distance from $v$ to target $t$}
-\KwData{Minimum priority queue $Q$, also called open list, ordered by tentative distances}
-$B[v] \gets +\infty$ for all $v\neq t$;
-$B[t] \gets 0$\;
-Make $Q$ only contain $t$ with weight $0$\;
+\begin{algorithm2e}[tb]
+\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to target $t$ in $\gchd$}
+\KwData{Minimum priority queue $Q$, ordered by tentative distances}
+$D^{\downarrow}[v] \gets +\infty$ for all $v\neq t$\;
+$D^{\downarrow}[t] \gets 0$\;
+Make $Q$ only contain $t$ with tentative distance $0$\;
 \While{$Q$ not empty}{
   $u\gets$ pop minimum element from $Q$\;
-  \For{all down-edges $vu$ in $G^+_\ell$}{
-    \If{$B[v] > w_\ell(vu) + B[u]$}{
-      $B[v]\gets w_\ell(vu) + B[u]$\;
-                        Add $v$ or decrease $v$'s key in $Q$ to $B[v]$\;
-    }
-  }
-}
-\caption{CH backward search}
+  \For{all down-edges $vu \in \echd$}{
+    \If{$D^{\downarrow}[v] > w(vu) + D^{\downarrow}[u]$}{
+      $D^{\downarrow}[v]\gets w(vu) + D^{\downarrow}[u]$\;
+      Add $v$ or decrease $v$'s key in $Q$ to $D^{\downarrow}[v]$\;
+    }
+  }
+}
+\caption{CH backward search.}
 \label{algo:ch-backward}
@@
 
-In a CH, each node gets a unique rank representing its ``importance''.
-By convention, important nodes are usually depicted at the ``top'' while unimportant nodes are positioned at the ``bottom''.
+In a CH, each vertex gets a unique rank representing its ``importance''.
+By convention, important vertices are usually depicted at the ``top'' while unimportant vertices are positioned at the ``bottom''.
 The intuition is that dead-ends are unimportant while highways or bridges are important.
-An edge goes \emph{up} when its tail node has lower rank than its head node.
+An edge goes \emph{up} when its tail vertex has lower rank than its head vertex.
 \emph{Down} edges are defined analogously.
-An \emph{up-down path} is a path where only one node $m$ is more important than both its neighbors.
-$m$ is called the \emph{mid} node.
-An \emph{up path} is a path where the last node is the mid node.
-Similarly, for a \emph{down path}, the first node has to be the mid node.
+An \emph{up-down path} is a path where only one vertex $v^*$ is more important than both its neighbors.
+$v^*$ is called the \emph{mid} vertex.
+An \emph{up path} is a path where the last vertex is the mid vertex.
+Similarly, for a \emph{down path}, the first vertex has to be the mid vertex.
 % Every up and down path is an up-down path.
 %
-In the preprocessing phase, first, a node ordering is computed heuristically.
+In the preprocessing phase, first, a vertex ordering is computed heuristically.
 Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
-This is done by repeatedly contracting the least important node and adding shortcuts between its higher ranked neighbors.
+This is done by repeatedly contracting the least important vertex and adding shortcuts between its higher ranked neighbors.
 %See~\cite{gssv-erlrn-12} for the details.
-After the preprocessing, for every pair of nodes $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
-See Figure~\ref{fig:ch} for a proof sketch.
+After the preprocessing, for every pair of vertices $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
+See Figure~\ref{fig:ch} for an illustration.
 From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
@@
 Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
-The two searches meet at the mid node.
-Pseudocode for the backward search, i.e., the path from $m$ to $t$, is presented in Algorithm~\ref{algo:ch-backward}.
+By construction, the mid vertex must be in the intersection of the forward and backward search spaces.
+Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
 The forward search works analogously.
-%
-A CH query is fast, if the number of nodes reachable via only up- or down-nodes is small.
+% TODO \gchu and \gchd
+
+A CH query is fast, if the number of vertices reachable via only up- or down-edges is small.
 On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
@@
 % This yields tight estimates but a high overhead for the heuristic evaluation.
-% While a single CH query is fast, answering one for every node explored in the $A^*$ search is slow.
+% While a single CH query is fast, answering one for every vertex explored in the $A^*$ search is slow.
 % Fortunately, we can do better.
@@
 Paths are computed in the query phase.
-This setup is useful to allow for weight changes that are common to all vehicles such as life traffic updates.
-The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}.
-
-Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that uses this three-phase setup.
-The key idea is to use a node importance order that is independent of the weight function.
-A CCH is also a CH.
-The final result of the preprocessing is also an augmented graph $G^+$.
-It fulfills all necessary properties for CH query algorithms.
+This setup is useful to allow for weight changes that are common to all vehicles such as live traffic updates.
+The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}. % TODO better clear up whats related work and whats prelim
+
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that follows this three-phase setup.
+The key idea is to use a vertex importance order that is independent of the weight function.
+CCH use \emph{nested dissection} orders for this, i.e.\ a small balanced vertex separator is computed and the separator vertices get the highest importance.
+The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
+This results in surprisingly good CH orders.
+Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
+
+Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
+Once the order was computed, the augmented graph $G^+$ is computed.
+The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
+However, discussing them is beyond the scope of this work.
+See~\cite{dsw-cch-15,BuchholdSW19} for details.
+A CCH augmented graph fulfills all necessary properties for CH query algorithms.
 The CH query algorithms can therefore be applied without modifications.
@@
 During preprocessing, an \emph{elimination tree} can be constructed.
-The root of the elimination tree is the most important node.
-For every other node its parent is its least important upward neighbor.
-In~\cite{bcrw-s-16}, it was proven that set of ancestors of a node in the elimination tree is equal to the nodes reachable in a CH search.
+The root of the elimination tree is the most important vertex.
+For every other vertex its parent is its least important upward neighbor.
+In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to the vertices reachable in a CH search.
 This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
 For the forward search, start at the source and follow the elimination tree upward until the root is reached.
-All outgoing edges of encountered nodes are relaxed.
+All outgoing edges of encountered vertices are relaxed.
 The backward search works analogously but starts at the target.
-Because this elimination tree based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
-However, only weight-independent node orderings admit an elimination tree of low height.
+Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
+With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a milliseconds.
+CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
+However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries are the faster variant.
+Note that only nested dissection orderings admit an elimination tree of low height.
 The elimination tree query can therefore not be used for a general CH.
@@
 
-\begin{algorithm2e}
-\KwData{$P[v]$: tentative distance from $v$ to $t$}
-Execute Algorithm~\ref{algo:ch-backward}\;
+\begin{algorithm2e}[tb]
+\KwData{$D[v]$: tentative distance from any $v \in V$ to $t$}
+Execute Algorithm~\ref{algo:ch-backward}, filling $D$\;
 \For{all CH levels $L$ from most to least important}{
-	\For{all up-edges $uv$ in $G^+_\ell$ with $u$ in $L$}{
-		\If{$P[u] < P[v] + w_\ell(uv)$}{
-			$P[u] \leftarrow P[v] + w_\ell(uv)$\;
-		}
-	}
-}
-\caption{PHAST basic all-to-one search}
+	\For{all up-edges $uv \in \echu$ with $u$ in $L$}{
+		\If{$D[u] > D[v] + w(uv)$}{
+			$D[u] \leftarrow D[v] + w(uv)$\;
+		}
+	}
+}
+\caption{PHAST basic all-to-one search.}
 \label{algo:phast}
@@
 
-PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all nodes to one target node (or vice versa, the reverse case works analogously).
+PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all vertices to one target vertex (or vice versa, the reverse case works analogously).
 This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
@@
 The first step is analogue to the CH query:
-From $t$, all reachable nodes via reversed down-edges are explored.
-Algorithm~\ref{algo:ch-backward} shows this first step.
-For the second step, PHAST organizes nodes into \emph{levels}.
-These levels correspond to the importance ordering but allow multiple nodes in the same level.
-No edge must connect two nodes within the same level.
+From $t$, all reachable vertices via reversed down-edges are explored, as shown in Algorithm~\ref{algo:ch-backward}.
+For the second step, PHAST organizes vertices into \emph{levels}.
+These levels correspond to the importance ordering but allow multiple vertices in the same level.
+However, no edge must connect two vertices within the same level.
+Such a level assignment can be obtained by assigning all vertices without lower ranked neighbors to the lowest level.
+All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.
 PHAST then iterates over all CH levels from top to bottom.
 In each iteration, all up-edges starting within the current level are followed in reverse.
-After all levels are processed, the shortest distances from all nodes to $t$ were computed.
+After all levels are processed, the shortest distances from all vertices to $t$ were computed.
 Pseudocode is provided in Algorithm~\ref{algo:phast}.
-PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures.
-It can also easily be parallelized.
+PHAST is faster than Dijkstra's algorithm on road graphs because it is a better fit for modern processor architectures, better at utilizing data locality and, most importantly, can be parallelized very effectively.
 % However, we will not consider parallelization in this paper.
-We refer to~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
+See~\cite{dgnw-phast-13} for an in-depth experimental performance analysis.
 % Using PHAST, we can also compute a tight A* heuristic.
-% In the query phase, we first run PHAST to compute the distances from every node to $t$ with respect to $w_\ell$ and store the result in an array $H$.
+% In the query phase, we first run PHAST to compute the distances from every vertex to $t$ with respect to $w$ and store the result in an array $H$.
 % Next, we run $A^*$ and implement the heuristic as a lookup in the array $H$.
@@
 % However, the PHAST step before the search is comparatively expensive.
-% The reason is that the distances towards $t$ are computed for \emph{all} nodes.
-% Ideally, we only want to compute the distances from the nodes explored in the $A^*$ search.
+% The reason is that the distances towards $t$ are computed for \emph{all} vertices.
+% Ideally, we only want to compute the distances from the vertices explored in the $A^*$ search.
 
@@
 
-RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source nodes to one target node (again, the reversed case works analogously), solving the \emph{many-to-one} problem.
-Given a set of source nodes $S$, the first step is to copy all up-edges reachable only through up-edges from the source nodes into a new graph.
-This step is called \emph{selection}.
+RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source vertices to one target vertex (again, the reversed case works analogously), solving the \emph{many-to-one} problem.
+Given a set of source vertices $S$, the first step is to copy the combined search space of all sources into a \emph{restricted subgraph}.
+Formally, let $V_{\operatorname{restr}}$ be the set of vertices reachable in $\gchu$ from any $s_i \in S$.
+The restricted subgraph $\gchu_{\operatorname{restr}}$ is subgraph of $\gchu$ induced by $V_{\operatorname{rest}}$.
+Copying the relevant edges into this restricted subgraph makes up the \emph{selection} step.
 Then, in the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
@@
 If the source set changes often, selection times may become problematic.
-% RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of nodes.
+% RPHAST can also be applied to the \emph{many-to-many} problem, computing a distance matrix between sets $S$ and $T$ of vertices.
 % Implemented with SIMD instructions it is, to the best of our knowledge, the fastest currently known algorithm for this problem setting.
@@
 
-In this section, we discuss another very specific problem setting which naturally arises from A* heuristics.
-Here, the target node is known in the selection step, but the source nodes are only queried one after another.
+In this section, we discuss a variant of the many-to-one shortest path problem which naturally arises from A* heuristics.
+Here, the target vertex $t$ is known in the selection step, but source vertices $s_1,\dots,s_k$ are queried one after another.
 We denote this problem as the \emph{incremental many-to-one} problem.
-The first step is the \emph{target selection} where the target node is provided.
-Then, an arbitrary number of source nodes $s_i$ are given one after another.
-For each source the distance to the target has to be computed before the next source will be provided.
+The first step is the \emph{target selection} where the target vertex $t$ is provided.
+Then, an arbitrary number of source vertices are given one after another.
+For each source $s_i$ the distance $\dist(s_i, t)$ has to be computed before the next source $s_{i+1}$ is provided.
 
 We consider the combined running time of the target selection and each incremental query as the total running time to answer an incremental many-to-one query.
-Since the target selection time is included, computing the distances to all nodes with Dijkstra or PHAST is too slow.
+Since the target selection time is included, computing the distances to all vertices with Dijkstra or PHAST is too slow.
 Also, since the source set is only provided incrementally, RPHAST in its basic form is not well suited to our problem.
@@
 
-\begin{algorithm2e}
-\KwData{$B[v]$: tentative distance from $v$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[v]$: memoized distance from $v$ to $t$, $\bot$ initially}
+\begin{algorithm2e}[tb]
+\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$D[v]$: memoized distance from vertex $v \in V$ to $t$, shared between invocations, initialized to $\bot$ during the selection}
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
 \SetKwProg{Fn}{Function}{:}{}
+
+Execute Algorithm~\ref{algo:ch-backward}, filling $D^{\downarrow}$\;
+$D[v] \gets \bot$ for all $v\neq t$\;
+
 \Fn{\Dist{$u$}}{
 	\If{$D[u] = \bot$}{
-		$D[u]\leftarrow B[u]$\;
-		\For{all up-edges $uv$ in $G^+_\ell$}{
-                        $D[u]\leftarrow\min\{D[u],w_\ell(uv)+\Dist(v)\}$\;
+		$D[u]\leftarrow D^{\downarrow}[u]$\;
+		\For{all up-edges $uv \in \echu$}{
+      $D[u]\leftarrow\min(D[u],w(uv)+\Dist(v))$\;
 		}
@@
 }
-\caption{Lazy RPHAST algorithm}
+\caption{Lazy RPHAST algorithm.}
 \label{algo:pot}
@@
 
-The core idea of our algorithm is to do the (R)PHAST computation lazily and using memoization as depicted in Algorithm~\ref{algo:pot}.
-In the target selection, we first run the backward CH search from $t$ to obtain an array $B$.
-$B[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
-$B$ is computed as shown in Algorithm~\ref{algo:ch-backward}.
-
-To compute the distance from a source $s_i$, we recursively compute for all up-edges $(s_i v)$ the distance $D[v]$.
-Next, we compute the minimum distance over all up-down paths that contain at least one up-edge using $d = \min_v\{w_\ell(s_i v) + D[v]\}$.
-As not all shortest up-down paths contain an up-edge, we set $D[s_i] = \min \{ B[s_i], d \}$.
-This calculation is correct, as it computes the minimum up-down $s_it$-path distance in $G^+_\ell$, which corresponds to the minimum $s_it$-path distance in a CH.
+The core idea of our algorithm is to do the RPHAST computation lazily and using memoization.
+In the target selection, we first run the backward CH search from $t$ to obtain an array $D^{\downarrow}$.
+$D^{\downarrow}[v]$ is the minimum down $vt$-path distance or $+\infty$, if there is no such path.
+$D^{\downarrow}$ is computed as shown in Algorithm~\ref{algo:ch-backward}.
+Further, the distances $D[v]$ are initialized to a sentinel value $\bot$ where $\bot$ is some special value distinct from any other valid distance (including $\infty$).
+This value indicates that the distance to from $v$ to $t$ has not yet been computed.
+
+Now, distances from many sources $s_i$ to $t$ can be computed incrementally with the \textsf{ComputeAndMemoizeDist} function as shown in Algorithm~\ref{algo:pot}.
+The key to doing this efficiently is reusing the distance information $D$ across invocations through memoization.
+Thus, the first step of the algorithm is always to check if the distance for the requested vertex has already been computed and immediately returning it if this is the case.
+If not, the distance $D[s_i]$ is initialized to the shortest down-path distance $D^{\downarrow}[s_i]$ obtained by the backward search.
+Then, the algorithm iterates over all up-edges $(s_i v)$ and checks if the up-down path through this neighbor can improve the distance.
+To obtain the shortest distance from a neighbor $v$ to $t$, the algorithm is invoked recursively.
+
+\paragraph{Correctness}
+Due to~\cite{gssv-erlrn-12}, an up-down path of shortest distance must exist in $G^+$.
+Further, $G^+$ can be decomposed into two directed acyclic graphs (DAG) $\gchu$ and $\gchd$ and the up path can be found in $\gchu$ and the down path in $\gchd$.
+Observe that the second part of Lazy RPHAST is, in fact, a recursive depth-first search (DFS) on $\gchu$ which relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
+Since DFS can be used to find shortest paths on DAGs, the search will find shortest up paths on $\gchu$.
+Concatenated with the down paths obtained by the backward search, this yields correct shortest distances.
 
@@
 However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
-To compute the distance $D[v]$ of a node $v$ the distances of all upward neighbors must be final.
-In Algorithm~\ref{algo:pot}, we have to compute all these distances recursively.
-However, on CCH it is sufficient to ensure that the distance of the elimination tree parent of $v$ is final.
-All upward neighbors of $v$ are by definition included in the search space, thus they are also ancestors in the elimination tree.
-If the distance of the parent in the elimination tree is final, the distances of all other upward neighbors must already be final, too.
-Thus, it is sufficient to iteratively ensure that the distance of the parent in the elimination tree has been computed.
-As soon as a node with final distance is encountered, all ancestors are known to have their final distance.
-This allows for the more efficient implementation depicted in Algorithm~\ref{algo:cch_pot}.
-
-\begin{algorithm2e}
-\KwData{$B[v]$: tentative distance from $v$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[v]$: memoized distance from $v$ to $t$, $\bot$ initially}
-\KwData{$P[v]$: parent node of $v$ in the elimination tree, parent of root is $\bot$}
-\KwData{$S$: stack with nodes to compute distances, empty initially, only used to store intermediate data}
+To compute the distance $D[v]$ of a vertex $v$, the distances of all upward neighbors must be final.
+In Algorithm~\ref{algo:pot}, these upward neighbor distances are computed recursively.
+Thus, the search space is explored in a DFS-like fashion and distances are finalized in DFS post order.
+However, the path from a vertex to the root in the elimination tree is a also topological order for the upward search space.
+This is because the ancestors in the elimination tree contain the entire upward search space~\cite{bcrw-s-16}.
+Therefore, iterating over the vertices on the elimination tree path from the root to $v$ while relaxing outgoing upward edges of each vertex also yields shortest distances.
+Further, with this approach, when a vertex $v$ already has a distance $D[v] \neq \bot$, all ancestors of $v$ must already have their final distance, too.
+Thus, as soon as the algorithm encounters a vertex which already has a final distance, the remaining search space is known to have final distances.
+
+\begin{algorithm2e}[tb]
+\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$D[v]$: memoized distance from any vertex $v \in V$ to $t$, shared between invocations, initialized to $\bot$ during the selection}
+\KwData{$P[v]$: parent of a vertex $v \in V$ in the elimination tree, a parent of $\bot$ indicates a root vertex}
+\KwData{$S$: stack with vertices to compute distances, empty initially, only used to store intermediate data}
 \SetKwFunction{Dist}{ComputeAndMemoizeDist}
@@
 \SetKwProg{Fn}{Function}{:}{}
+
+Execute Algorithm~\ref{algo:ch-backward}, filling $D^{\downarrow}$\;
+$D[v] \gets \bot$ for all $v\neq t$\;
+
 \Fn{\Dist{$u$}}{
-  \tcp{Determine for what nodes, the $D$ needs to be filled}
+  \tcp{Determine the vertices $v$ for which $D[v]$ needs to be computed}
   $v \gets u$\;
@@
   }
-  \tcp{Fill $D$ for those nodes}
+  \tcp{Compute $D$ for those vertices}
   \While{$S$ not empty}{
     $v\leftarrow$ pop top element from $S$\;
-		$D[v]\gets B[v]$\;
-		\For{all up-edges $vw$ in $G^+_\ell$}{
-                        $D[v]\gets\min\{D[v],w_\ell(vw)+D[w]\}$\;
+		$D[v]\gets D^{\downarrow}[v]$\;
+		\For{all up-edges $vw \in \echu$}{
+      $D[v]\gets\min(D[v],w(vw)+D[w])$\;
 		}
@@
 }
-\caption{Elimination tree based Lazy RPHAST algorithm}
+\caption{Elimination tree based Lazy RPHAST algorithm.}
 \label{algo:cch_pot}
 \end{algorithm2e}
+
+This allows for the more efficient implementation depicted in Algorithm~\ref{algo:cch_pot}.
+For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
+To compute missing distances, the elimination tree ancestors which also do not have a distance yet are collected and pushed onto a stack $S$.
+The elimination tree can be traversed through parent pointers $P[v]$.
+As soon as a vertex with final distance is encountered, all remaining ancestors are known to have their final distance.
+Now, the vertices are successively popped from the stack, i.e.\ the algorithm walks the elimination tree path back down and relaxes outgoing upward edges.
+This yields the shortest distances for all the vertices on the path.
 
@@
 % Investigating how Lazy RPHAST performs in the many-to-many setting is therefore interesting.
-% In this setting, a set of source and target nodes are given.
-% The objective is to compute the distances between all pairs of source and target nodes.
+% In this setting, a set of source and target vertices are given.
+% The objective is to compute the distances between all pairs of source and target vertices.
 % In this section, we described how Lazy RPHAST can be used to solve this many-to-many problem.
@@
 % Let $k$ denote the number of elements in one vector.
-% RPHAST computes the distances from all source nodes to $k$ target nodes in one run.
-% If the target set contains more than $k$ nodes, then the target set is split into smaller sets of $k$ elements\footnote{If the number of targets is not divisible by $k$, the set is increased by adding additional target nodes.}.
+% RPHAST computes the distances from all source vertices to $k$ target vertices in one run.
+% If the target set contains more than $k$ vertices, then the target set is split into smaller sets of $k$ elements\footnote{If the number of targets is not divisible by $k$, the set is increased by adding additional target vertices.}.
 % The algorithm is then run for each of these smaller sets.
@@
 % We translate this setup to the Lazy RPHAST setting as follows.
-% First, a set of $k$ target nodes need to be fixed.
+% First, a set of $k$ target vertices need to be fixed.
 % A vectorized selection phase is then performed for this fixed target set.
-% After the selection, the source nodes are iteratively revealed.
-% Once a source node is revealed, the algorithm must compute the distances to all $k$ targets.
-% Only after the distances are computed, the next source node is revealed.
+% After the selection, the source vertices are iteratively revealed.
+% Once a source vertex is revealed, the algorithm must compute the distances to all $k$ targets.
+% Only after the distances are computed, the next source vertex is revealed.
 
@@
 % \begin{algorithm2e}
-% \KwIn{$\hat{T}$: SIMD vector of $k$ target nodes}
+% \KwIn{$\hat{T}$: SIMD vector of $k$ target vertices}
 % \KwData{$\hat{B}[x][i]$: tentative distance from $x$ to target $\hat{T}[i]$}
-% \KwData{Minimum priority queue $Q$, also called open list, ordered by node rank}
+% \KwData{Minimum priority queue $Q$, also called open list, ordered by vertex rank}
 % Make $Q$ an empty queue\;
@@
 % Vectorizing Algorithm~\ref{algo:ch-backward} needs some adjustments.
-% In the original algorithm, the nodes are ordered by distance.
+% In the original algorithm, the vertices are ordered by distance.
 % In the vectorized variant, distances are SIMD vectors and can therefore not be ordered.
-% To resolve this problem, we ordered the nodes in the queue not by distance but by CH rank.
-% The CH rank is the position of a node in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
-% All other modifications are straight-forward.
+% To resolve this problem, we ordered the vertices in the queue not by distance but by CH rank.
+% The CH rank is the position of a vertex in the contraction order. \todo{wurde rank schon eingeführt? Falls ja, dann kann der Satz weg.}
+% All other modifications are straightforward.
 % The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
@@
 % % TODO besser herleiten, wieso kein 2d array
-% For each node $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$.
+% For each vertex $x$, we maintain an initially empty list $B[x]$ of pair $(t_i, \textrm{dist}_{\textrm{up}}(x,t_i))$.
 % These pairs are called \emph{buckets}.
@@
 % We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
-% After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target node.
+% After calling the adjusted \texttt{ComputeAndMemoizeDist} for each source, we know the shortest distances between each pair of source and target vertex.
 % This way, the search space of the sources is explored only once.
@@
 % This can be improved with the following modified CH backward search:
-% All target nodes are inserted into the queue at once.
+% All target vertices are inserted into the queue at once.
 % The priority queue is ordered by importance instead of distance.
 % Distance labels are buckets instead of scalar values.
-% The buckets of each target node are initialized with a single entry for the respective node with distance zero.
-% When relaxing an edge, the entries in the current node's bucket are increased by the edge's weight and merged with the bucket entries of the head node of the edge.
+% The buckets of each target vertex are initialized with a single entry for the respective vertex with distance zero.
+% When relaxing an edge, the entries in the current vertex's bucket are increased by the edge's weight and merged with the bucket entries of the head vertex of the edge.
 % By keeping the buckets sorted, this merging can be efficiently implemented with a coordinated linear sweep over both buckets.
@@
 
-\section{A* with CH-Potentials}\label{sec:main-algo}
+\section{CH-Potentials: A New A* Heuristic}\label{sec:main-algo}
 
@@
 We also discuss CH-Potentials for the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
-Further, we present several low degree node optimization which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
+Further, we present several low degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
 Finally, we introduce an algorithmic framework to apply A* with CH-Potentials to a variety of practical route planning problems.
@@
 In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
-The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for node $v$.
+The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $v$.
 
@@
 
-Preliminary experiments showed, that a significant amount of query running time is spent in heuristic evaluations and queue operations.
-We can reduce both by keeping some nodes out of the queue, as the heuristic needs to be evaluated when a node is pushed into the queue.
-Avoiding pushing low degree nodes into the queue is the focus of this section.
-The techniques discussed here are a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
-
-We modify A* by processing low degree nodes consecutively without pushing them into the queue.
-Our algorithm uses the undirected degree $d(v)$ of a node $v$.
-Formally, $d(v)$ is the number of nodes $u$ such that $(vu)\in E$ or $(uv)\in E$.
-
-Analogous to A*, our algorithm stores for every node $v$ a tentative distance $D[v]$.
-Additionally, it maintains a minimum priority queue.
-Diverging from A*, not all nodes need be pushed, but every node has a tentative distance.
-
-\subsubsection{Skip Degree Two Nodes}
-
-Our algorithm differs from A* when removing a node $u$ from the queue.
+Preliminary experiments showed that a significant amount of query running time is spent in heuristic evaluations and queue operations.
+We can reduce both by keeping some vertices out of the queue, as the heuristic needs to be evaluated when a vertex is pushed into the queue.
+Consider for example a chain of vertices with exactly two neighbors.
+Traversing this chain by successively pushing each vertex into the queue, computing the heuristic and popping the vertex again from the queue appears quite wasteful.
+Therefore, in this section, we discuss techniques to avoid pushing such low degree vertices into the queue and rather process them consecutively.
+The techniques discussed here are essentially a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
+
+% We modify A* by processing low degree vertices consecutively without pushing them into the queue.
+% Analogously to A*, our algorithm stores for every vertex $v$ a tentative distance $D[v]$ and maintains a minimum priority queue.
+% Diverging from A*, not all vertices need be pushed into the priority queue even though every vertex has a tentative distance.
+
+\subsubsection{Skip Degree Two Vertices}
+
+Recall that $|N(u)|$ is the number of neighbors $v$ such that $vu \in E$ or $uv \in E$.
+Our algorithm differs from A* when removing a vertex $u$ from the queue.
 A* iterates over the outgoing edges $(uv)$ of $u$ and tries to reduce $D[v]$ by relaxing $(uv)$.
 If A* succeeds, $v$'s weight in the queue is set to $D[v]+h_t(v)$.
-Our algorithm, however, behaves differently, if $d(v)\le 2$.
-Our algorithm determines the longest degree two chain of nodes $u, v_1,\ldots, v_k, w$ such that $d(v_i)=2$ and $d(w) > 2$.
+Our algorithm, however, behaves differently, if $|N(v)|\le 2$.
+Our algorithm determines the longest degree two chain of vertices $u, v_1,\ldots, v_k, w$ such that $|N(v_i)|=2$ and $|N(w)| > 2$.
 If our algorithm succeeds in reducing $D[v_1]$, it does not push $v_1$ into the queue.
@@
 If it does not reach $w$, then only $D$ is modified, but no queue action is performed.
-If $D[w]$ is modified and $d(w)>2$, $w$'s weight in the queue is set to $D[w]+h_t(w)$.
-
-As the target node $t$ might have degree two, our algorithm cannot rely on stopping, when $t$ is removed from the queue.
+If $D[w]$ is modified and $|N(w)|>2$, $w$'s weight in the queue is set to $D[w]+h_t(w)$.
+
+As the target vertex $t$ might have degree two, our algorithm cannot rely on stopping when $t$ is removed from the queue.
 Instead, our algorithm stops as soon as $D[t]$ is less than the minimum weight in the queue.
 
-\subsubsection{Skip Degree Three Nodes}
-
-We can also skip some degree three nodes.
+\subsubsection{Skip Degree Three Vertices}
+
+We can also skip some degree three vertices.
 Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
-If $d(w) > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
+If $|N(w)| > 3$ or $w$ is in the queue, our algorithm proceeds as in the previous section.
 Otherwise, there exist up to two degree chains $w, x_1,\ldots,x_p, y$ and $w,a_1,\ldots,a_q,b$ such that $x_1\neq v_k \neq a_1$.
@@
 
-A lot of nodes in road networks lead to dead-ends.
-Unless the source or target is in this dead-end, it is unnecessary to explore these nodes.
+A lot of vertices in road networks lead to dead-ends.
+Unless the source or target is in this dead-end, it is unnecessary to explore these vertices.
 
@@
 We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
-For every node $v$ in the input graph $G$, we store the attachment node $a_v$ to the core.
-For nodes in the core, $a_v=v$.
-We exploit that all attachment nodes are single node separators and the problem can be decomposed along them.
+For every vertex $v$ in the input graph $G$, we store the attachment vertex $a_v$ to the core.
+For vertices in the core, $a_v=v$.
+We exploit that all attachment vertices are single vertex separators and the problem can be decomposed along them.
 
@@
 
-On road graphs, bidirectional search provides a way to halve the running time of Dijkstra's algorithm.
+On road graphs, bidirectional search provides a simple way to halve the practical running time of Dijkstra's algorithm.
 A bidirectional variant of A* therefore also seems desirable.
@@
 Similarly, the backward search has its queue ordered by $\overleftarrow{D}[v] + \overleftarrow{h}_s(v)$, where $\overleftarrow{h}_s(v)$ is an estimate of the distance $\dist(s,v)$ from $s$ to $v$.
-We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.
+% We denote by $\overrightarrow{k}$ and $\overleftarrow{k}$ the minimum queue keys and by $\mu$ the length of the shortest $st$-path found so far.
 
@@
 To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for either direction.
-The forward search can skip nodes already settled by the backward search and vice versa.
+The forward search can skip vertices already settled by the backward search and vice versa.
 However, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
@@
 When the forward search scans an edge $uv$ where $\overrightarrow{D}[u] + w(uv) + \overrightarrow{h}_s(v) > \mu$ holds, i.e.
-the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $y$ into the queue.
+the distance plus the estimate of the remaining distance is already greater than the currently known tentative distance, it is not necessary to push $v$ into the queue.
 The pruning rule for the backward search is analogous.
@@
 It remains to lower bound $\dist(v,t)$.
-Nodes are removed from the backward queue ordered by $\overleftarrow{h}(v) + \dist(v,t)$.
+Vertices are removed from the backward queue ordered by $\overleftarrow{h}(v) + \dist(v,t)$.
 If $v$ was not yet removed from the backward queue, we know that $\overleftarrow{h}(v) + \dist(v,t) \geq \overleftarrow{k}$.
@@
 Thus, $v$ does not have to be pushed if $\overrightarrow{D}[u]+w(uv)+\overleftarrow{k}-\overleftarrow{h}(v)\ge \mu$.
-The node $v$ might still be pushed into the queue when there is another edge $wv$ for which the pruning is not possible.
+The vertex $v$ might still be pushed into the queue when there is another edge $wv$ for which the pruning is not possible.
 Checking the pruning rule requires evaluating the backward heuristic.
@@
 
-Unless stated differently, for all bidirectional A* variants, we always alternate between removing a node from the forward and the backward queues.
+Unless stated differently, for all bidirectional A* variants, we always alternate between removing a vertex from the forward and the backward queues.
 In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
@@
 
-We consider a variety different applications, with slightly different problem models.
+% TODO make sure to distinguish PROBLEM model from SOLUTION appraoch
+
+We consider a variety of different of applications, with slightly different problem models.
 The goal is always to quickly answer many shortest path queries.
 For the purpose of describing our framework, we establish a shared notation:
-Input to each query are nodes $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
+Input to each query are vertices $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
 However, the precise formal inputs of the query and what exactly $w_q$ represents depends on the application.
@@
 
-To enable quick shortest path computations, we consider a two-phase setup with an additional off-line preprocessing phase before the on-line query phase.
-The input to the preprocessing phase is a graph $G_\ell$ with lower bound weights $w_\ell$ and a node mapping function $\phi$.
+To enable quick shortest path computations, we consider a two-phase setup with an additional offline preprocessing phase before the online query phase.
+The input to the preprocessing phase is a graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$.
 $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
-We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$. % with respect to $w_\ell$.
-The output of the preprocessing is auxiliary data that enables an efficient heuristic function $h_t(x)$.
-$h_t(x)$ is the exact distance from $\phi(x)$ to $\phi(t)$ in $G_\ell$.
+The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
 In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
 
-The query phase uses this heuristic in an A* search between nodes $s$ and $t$ on $G_q$ and $w_q$.
-Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph and only $w_q$ changes for the queries.
-The exact implementation of this A* search depends on the application.
+The query phase uses this auxiliary data in an A* search between vertices $s$ and $t$ on $G_q = (V_q, E_q)$ with weights $w_q$.
+Let $\phi : V_q \to V_{\ell}$ denote a function mapping vertices in the query graph to vertices in the lower bound graph.
+We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$ with respect to $w_\ell$.
+Then, A* uses the heuristic function $h_t(x) = \dist_{w_{\ell}}(\phi(x), \phi(t))$ computed using Lazy RPHAST.
+Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
+The exact implementation of the A* search depends on the application.
 Our approach only provides the heuristic $h_t$ for the A* search.
@@
 
-The classical shortest path problem allows to freely change edges at nodes.
+The classical shortest path problem allows to freely change edges at vertices.
 However, in the real world, turn restrictions, such as a forbidden left or right turn, exist.
@@
 % Afterwards, we describe how it can be solved with CH-Potentials.
-A \emph{turn weight} $w_t$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns.
+A \emph{turn weight} $w_t : E \times E \to \mathbb{R}^{\geq 0}$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns. % TODO formally
 For CH-Potentials, we use zero as lower bound for every turn weight in the heuristic.
@@
 
-A path with nodes $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
-w_\ell(v_1 v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1},v_i,v_{i+1}) + w_\ell(v_i v_{i+1})
+A path with vertices $v_1, v_2,\ldots v_k$ has the following \emph{turn-aware weight}: \[
+w_\ell(v_1 v_2) + \sum_{i=2}^{k-1}  w_t(v_{i-1} v_i, v_i v_{i+1}) + w_\ell(v_i v_{i+1})
 \]
@@
 We solve this problem by constructing a \emph{turn-expanded} graph as $G_q$.
-Edges in the input graph $G_\ell$ correspond to \emph{expanded nodes} in $G_q$.
-For every pair of incident edges $x y$ and $y z$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(x,y,z) + w_\ell(y z)$.
-A sequence of expanded nodes in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
+Edges in the input graph $G_\ell$ correspond to \emph{expanded vertices} in $G_q$.
+For every pair of incident edges $x y$ and $y z$ in $G_\ell$, there is an \emph{expanded edge} in $G_q$ with expanded weight $w_t(xy,yz) + w_\ell(y z)$.
+A sequence of expanded vertices in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
 The weight of a path in $G_q$ is equal to the turn-aware weight of the corresponding path in $G_\ell$ minus the irrelevant $w_\ell(v_1,v_2)$ term.
@@
 In this scenario, preprocessing and query use different graphs $G_\ell$ and $G_q$.
-We define the node mapping function $\phi$ as $\phi(x y) = y$.
-Obviously, $w_q(xy, yz) = w_t(x,y,z) + w_\ell(y z) \geq \dist_\ell(\phi(x y), \phi(y z))$ and this approach yields a feasible heuristic.
+We define the vertex mapping function $\phi$ as $\phi(x y) = y$.
+Obviously, $w_q((xy, yz)) = w_t(xy,yz) + w_\ell(y z) \geq \dist_\ell(\phi(x y), \phi(y z))$ and this approach yields a feasible heuristic.
 Sadly, the undirected graph underlying $G_q$ is always biconnected, if the input graph is strongly connected.
 The optimization described in Section~\ref{sec:largested-biconnected-component} is therefore ineffective.
-With this setup, CH-Potentials support turns without requiring turn information in the CH.
+With this setup, CH-Potentials supports turn costs without requiring turn information in the CH.
 
@@
 $w_q(e, \tau)$ is the travel time through edge $e$ when entering it at moment $\tau$.
-The input to the extended problem consists of a source node $s$ and a target node $t$, as in the classical problem formulation.
+The input to the extended problem consists of a source vertex $s$ and a target vertex $t$, as in the classical problem formulation.
 Additionally, the input contains a source time $\tau_s$.
@@
 Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
-% The FIFO property states that it is never beneficial to wait at a node before entering an edge.
+% The FIFO property states that it is never beneficial to wait at a vertex before entering an edge.
 The FIFO property states that it is not possible to arrive earlier by departing later.
@@
 Our implementation stores edge travel times using piece-wise linear functions.
-The A* search uses the tentative distance $\tau$ at a node $u$ when evaluating the travel time of outgoing edges $u v$.
+The A* search uses the tentative distance $\tau$ at a vertex $u$ when evaluating the travel time of outgoing edges $u v$.
 This strategy is very similar to TD-ALT~\cite{ndls-bastd-12,dw-lbrdg-07}.
@@
 
-With this setup, CH-Potentials support a combination of live and predicted traffic.
+With this setup, CH-Potentials supports a combination of live and predicted traffic.
 We did not make any modification, that would hinder a combination with other extensions.
 Further adding tunnel and/or highway avoidance or turn-aware routing is simple.
-This straight-forward integration of complex routing problems is the strength of the CH-Potentials.
+This straightforward integration of complex routing problems is the strength of CH-Potentials.
 
@@
 % $\psi_r$ is set to 0.01.
-% To avoid over-penalization, we count the times a \emph{node} was on a shortest path and do not apply penalties, when this number exceeds $k_{\max} := \lceil\log_{\psi}(1+\epsilon)\rceil + 2$.
+% To avoid over-penalization, we count the times a \emph{vertex} was on a shortest path and do not apply penalties, when this number exceeds $k_{\max} := \lceil\log_{\psi}(1+\epsilon)\rceil + 2$.
 % Shortest paths are split into segments not yet contained in the alternative graph.
@@
 % \begin{itemize}
-% 	\item All nodes on the current shortest path have been penalized $k_{\max}$ times.
+% 	\item All vertices on the current shortest path have been penalized $k_{\max}$ times.
 % 	\item 10 ore more segments have been added to the alternative graph.
@@
 
-\subsubsection{Temporary Driving Bans}
-
-Truck routing differs from car routing due to night driving bans and other restrictions.
-In~\cite{kswz-erptd-p-20}, a preliminary version of CH-Potentials\footnote{
-In~\cite{kswz-erptd-p-20}, CH-Potentials are used as a blackbox referring to an early ArXiv preprint~\cite{strasser2019perfect} of ours.
-Our submitted paper is the finished version of the ArXiv preprint.
-\cite{kswz-erptd-p-20} does not describe any of the contributions of this paper.
-} is used for such a scenario.
-The work considers time-dependent blocked edges and waiting at parking locations.
-Further, a trade-off between arrival time and route quality is considered.
+% \subsubsection{Temporary Driving Bans}
+
+% Truck routing differs from car routing due to night driving bans and other restrictions.
+% In~\cite{kswz-erptd-p-20}, a preliminary version of CH-Potentials is used for such a scenario.
+% The work considers time-dependent blocked edges and waiting at parking locations.
+% Further, a trade-off between arrival time and route quality is considered.
 
@@
 \centering
-\caption{Instances used in the evaluation with preprocessing running times to construct a CH or CCH-Potential.
-With CCH-Potentials $w_{\ell}$ can be updated by rerunning Phase 2 again.}\label{tab:graphs}
+\caption{Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials. With CCH-Potentials $w_{\ell}$ can be updated by rerunning Phase 2 again.}\label{tab:graphs}
 \input{table/graphs.tex}
@@
 To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
-The graph has 16M nodes and 35M edges.
+The graph has 16M vertices and 35M edges.
 For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://mapbox.com}}.
-The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the nodes.
+The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the vertices.
 It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
@@
 In this table, we further include the sequential preprocessing running time to construct the (C)CH.
+CH are constructed with RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
+For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
 We report preprocessing running times as averages over 10 runs.
-To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are nodes drawn uniformly at random and report average results.
-For many-to-one algorithms, we use the methodology from~\cite{delling_et_al:OASIcs:2011:3266}.
-We generate a \emph{ball} $B$ of nodes around 100 uniformly at random chosen nodes by running a Dijkstra search until $|B|$ nodes are settled where $|B|$ is a parameter varied throughout the experiments.
-From each ball we uniformly at random pick 100 targets and a set of sources $S$ with $|S| = 2^{14}$.
-We perform a many-to-one query for each target computing distances from all sources to it.
-This gives us 100 many-to-one queries per ball and a total of 10\,000 one-to-many queries per ball size.
+To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are vertices drawn uniformly at random and report average results.
+
+The performance of Lazy RPHAST is evaluated with many-to-one queries where each query consists of $2^{14}$ sources and one target vertex.
+However, instead of picking sources and targets from the full vertex set $V$, we draw them from local subsets of vertices $B$ of varying size $|B|$ called \emph{balls}.
+A ball $B$ is generated by picking a center uniformly at random and running Dijkstra's algorithm from it until the desired number of vertices $|B|$ was settled.
+This allows us to evaluate the performance depending on the distribution of the vertices.
+Since we use a fixed number of sources per query, a small ball size means that the vertices are densely clustered in the same region while large ball sizes mean that the vertices are distributed over large parts of the network.
+For each ball size, we generate 100 balls.
+From each ball, we pick one set of sources to which we compute distances from 100 different targets selected uniformly at random from the same ball.
+Therefore, each reported running time is the mean over 10\,000 queries.
+With this, we follow the methodology from~\cite{delling_et_al:OASIcs:2011:3266}.
 
@@
 Two line graphs with each eight lines for different ball sizes and algorithms.
-The x-axis in both graphs is the number of nodes that were queried up to that point in the incremental query.
+The x-axis in both graphs is the number of vertices that were queried up to that point in the incremental query.
 In each graph, all lines have a similar shape and appear as scaled variants of each other.
@@
 The performance difference between the CH and CCH based variant is interesting.
-The CCH based variant can utilize the elimination tree for a more efficient implementation but the search space denser.
+The CCH based variant can utilize the elimination tree for a more efficient implementation but the search space is denser.
 This makes earlier queries more expensive and later queries cheaper.
@@
 Despite the laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental many-to-one queries.
-Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target node and $2^{14}$ sources for different ball sizes.
+Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target vertex and $2^{14}$ sources for different ball sizes.
 This experiment uses the exact methodology described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to roughly relate our results to the performance of other one-to-many algorithms.
@@
 The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
-While the numbers are not perfectly comparable due to different benchmark machines, we can still conclude that Lazy RPHAST is a very useful extension to RPHAST.
+While the numbers are not perfectly comparable due to different benchmark machines,\footnote{
+According to the comparison methodology from~\cite{BastDGMPSWW16} (see \url{https://i11www.iti.kit.edu/~pajor/survey/}), the machine used in~\cite{delling_et_al:OASIcs:2011:3266} (SPA-2) is about 30\% slower than ours (we obtained a score of 33159\,ms).
+However, these scaling factors have to be interpreted very carefully.
+They are obtained from one-to-all Dijkstra searches on continental-sized road networks.
+This heavily emphasizes memory bandwidth while neglecting other critical factors such as CPU frequency and cache size and speed which are likely much more important for algorithms carefully tuned to exploit data locality.
+} we can still conclude that Lazy RPHAST is a very useful extension to RPHAST.
 It allows efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
@@
 % One such example are k-Nearest-Neighbor Queries with CCH~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
-% There, shortest distances are computed repeatedly from the same source to boundary nodes of a cell of nodes.
+% There, shortest distances are computed repeatedly from the same source to boundary vertices of a cell of vertices.
 % On our suggestion, the authors integrated Lazy RPHAST into their algorithm which took only a couple of lines of code\footnote{\url{https://github.com/vbuchhold/routing-framework/commit/c739dc8e81}}.
@@
 This illustrates both the strengths and limits of our approach and goal directed search in general.
-CH-Potentials can only achieve competitive running times if the application allows for a sufficiently tight lower bounds at preprocessing time.
+CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.
 
@@
 The query distance is thus not the reason.
-After some investigation, we concluded that this is due to non-uniform road graph density.
+After some investigation, we concluded that this is due to non-uniform road network density.
 Some regions have more roads per area than others.
-The number explored A* nodes depends on the density of the search space area.
+The number of vertices explored by A* depends on the density of the search space area.
 As the density varies, the running times vary.
@@
 \centering
-\caption{Average query running times and number of queue pushs with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.}\label{tab:building_blocks}
+\caption{
+Average query running times and number of queue pushes with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
+The BCC, Deg2 and Deg3 columns indicate which optimizations from Section~\ref{sec:low-deg-improvment} were used.
+}\label{tab:building_blocks} % TODO explain labels and justify 1.05
 \input{table/building_blocks.tex}
@@
 
-Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations.
+Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
+The factor 1.05 was chosen to resemble realistic problem settings where goal directed search can achieve reasonable speedups (compare to Table~\ref{tab:applications}).
 We compare (C)CH-Potentials to three other heuristics.
-The first heuristic is the \emph{Zero} heuristic where $h(v)$ is $0$ for all nodes $v$.
+The first heuristic is the \emph{Zero} heuristic where $h(v)$ is $0$ for all vertices $v$.
 This corresponds to using Dijkstra's algorithm.
@@
 This heuristic has instant access to a shortest distance array with respect to $w_\ell$, i.e. it is faster than the fastest heuristic possible in our model.
-We fill this array before each query using a reverse Dijkstra search from the target node.
-In this experiment, we do not account for the running time of Dijkstra's algorithm.
+We fill this array before each query using a reverse Dijkstra search from the target vertex but do not include the running time for this step.
 Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
-(C)CH-Potentials compute the same distance estimates, but the heuristic evaluation has some overhead.
+(C)CH-Potentials computes the same distance estimates, but the heuristic evaluation has some overhead.
 Comparing against Oracle-A* allows us to measure this overhead.
@@
 All optimizations yield a combined speed-up of around 3.
-% Skipping degree three nodes has the smallest impact.
-(C)CH-Potentials outperform ALT by a factor of between six and seven and settle correspondingly fewer nodes.
+% Skipping degree three vertices has the smallest impact.
+(C)CH-Potentials outperforms ALT by a factor of between six and seven and settle correspondingly fewer vertices.
 This is not surprising, since ALT computes worse distance estimates.
-In contrast, CH-Potentials already compute exact distances with respect to $w_\ell$.
-The number of popped nodes is the same for CH-Potentials, CCH-Potentials and Oracle-A*.
+In contrast, CH-Potentials already computes exact distances with respect to $w_\ell$.
+The number of popped vertices is the same for CH-Potentials, CCH-Potentials and Oracle-A*.
 The only difference between (C)CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.
 This overhead leads to a slowdown of around 1.6.
-Thus, CH-Potentials are already very close to the best possible heuristic in this model.
+Thus, CH-Potentials is already very close to the best possible heuristic in this model.
 % Very interesting is the comparison with Oracle-A*.
@@
 This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
-CCH-Potentials are slightly slower than CH-Potentials because they use a weight independent node importance order.
+CCH-Potentials is slightly slower than CH-Potentials because they use a weight independent vertex importance order.
 This enables CCH-Potentials to support a customization whereas CH-Potentials cannot.
@@
 Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
-With $w_q = w_{\ell}$, unidirectional CH-Potentials are already optimal and only traverse the shortest path.
+With $w_q = w_{\ell}$, unidirectional CH-Potentials is already optimal and only traverses the shortest path.
 Here, bidirectional search will only introduce unnecessary overhead.
@@
 Here only weights with a speed less than 80\,kph were scaled up.
-This touches only the beginning and end of most shortest paths between randomly chosen nodes.
+This touches only the beginning and end of most shortest paths between randomly chosen vertices.
 The middle part of the shortest paths will typically use faster edges like highways.
-In this case, bidirectional (C)CH-Potentials are a factor of two faster than the unidirectional variant.
+In this case, bidirectional (C)CH-Potentials is a factor of two faster than the unidirectional variant.
 The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
@@
 CH achieves average query running times of 0.16\,ms on OSM Ger.
-CH-Potentials are roughly four times slower but still achieve a huge speedup of 3795 over Dijkstra.
+CH-Potentials is roughly four times slower but still achieve a huge speedup of 3795 over Dijkstra.
 Such large speedups are typical for CH.
-This shows that CH-Potentials gracefully converge toward CH in the $w_q = w_\ell$ special case.
+This shows that CH-Potentials gracefully converges toward CH in the $w_q = w_\ell$ special case.
 On DIMACS Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.
@@
 
-The length increase of Mapbox traffic predictions are about 12.3\%, and results in a running time of 120\,ms.
+The length increase of Mapbox traffic predictions is about 12.3\%, and results in a running time of 120\,ms.
 The speedup in the predicted scenario is larger than in the live setting, as the travel time function evaluations slow down Dijkstra's algorithm.
@@
 
+\section*{Acknowledgements}
+This article is the extended version of our conference paper ``A Fast and Tight Heuristic for A* in Road Networks''~\cite{strasser_et_al:LIPIcs.SEA.2021.6}.
+
 %%

### 2022-07-21 paper/table/graphs.tex
integrate a lot of reviewer feedback

@@
 \toprule
- &                &                & \multicolumn{3}{c}{Preprocessing [s]} \\ \cmidrule(lr){4-6} & Nodes          & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} \\ \cmidrule(lr){5-6} & $[\cdot 10^6]$ & $[\cdot 10^6]$ &                     & Phase 1 & Phase 2 \\
+ &                &                & \multicolumn{3}{c}{Preprocessing [s]} \\ \cmidrule(lr){4-6} & Vertices          & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} \\ \cmidrule(lr){5-6} & $[\cdot 10^6]$ & $[\cdot 10^6]$ &                     & Phase 1 & Phase 2 \\
 \midrule

### 2022-07-21 eval/bidir_features.py

@@ -52,7 +52,7 @@ pruning = pruning.rename(columns={
 lines = pruning.to_latex(escape=False, index=False, multicolumn_format='c', column_format='c@{\hskip6pt}c@{\hskip3pt}crrrrrrrr').split("\n")
 
 lines = lines[0:2] + [
-    R' &  &  & \multicolumn{5}{c}{Running time [ms]} & \multicolumn{3}{c}{Queue pushs [$\cdot 10^3$]} \\ \cmidrule(lr){4-8} \cmidrule(lr){9-11}',
+    R' &  &  & \multicolumn{5}{c}{Running time [ms]} & \multicolumn{3}{c}{Queue pushes [$\cdot 10^3$]} \\ \cmidrule(lr){4-8} \cmidrule(lr){9-11}',
     R'Low Deg.      & Bidirectional & New & \multirow{2}{*}{Zero} & \multirow{2}{*}{ALT} & \multirow{2}{*}{CH} & \multirow{2}{*}{CCH} & \multirow{2}{*}{Oracle} & \multirow{2}{*}{Zero} & \multirow{2}{*}{ALT} & (C)CH/ \\',
     R'Opt. & Potential     & Pruning  & & & & & & & & Oracle \\'
 ] + lines[4:9] + ['\\addlinespace'] + lines[9:]
@@ -82,7 +82,7 @@ bidir_switch_strat = bidir_switch_strat.drop([('num_queue_pushs_k', 'CH'), ('num
 
 bidir_switch_strat = bidir_switch_strat.rename(columns={
         'running_time_ms': 'Running time [ms]',
-        'num_queue_pushs_k': 'Queue pushs [$\\cdot 10^3$]',
+        'num_queue_pushs_k': 'Queue pushes [$\\cdot 10^3$]',
     }) \
     .rename(index={
         'alternating': 'Alternating',
@@ -102,14 +102,14 @@ table.loc[9:11, '$w_q$'] = ''
 lines = table.to_latex(escape=False, index=False, multicolumn_format='c', column_format='cccrrrrrrrr').split("\n")
 
 lines = lines[0:2] + [
-    R' &  &  & \multicolumn{5}{c}{Running time [ms]} & \multicolumn{3}{c}{Queue pushs [$\cdot 10^3$]} \\ \cmidrule(lr){4-8} \cmidrule(lr){9-11}'
[…]

### 2022-07-21 eval/building_blocks.py

@@ -30,7 +30,7 @@ table = table.unstack(0).stack(0) \
     .loc[['running_time_ms', 'num_queue_pushs']][['Zero', 'ALT', 'CH', 'CCH', 'Oracle']].reset_index(level=[2,3]) \
     .rename(index={
         'running_time_ms': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{Running\\time [ms]}}}',
-        'num_queue_pushs': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{Queue\\pushs [$\cdot 10^3$]}}}',
+        'num_queue_pushs': R'\multirow{4}{*}{\rotatebox[origin=c]{90}{\shortstack{Queue\\ {} [$\cdot 10^3$]}}}',
     })
 
 lines = table.to_latex(escape=False).split("\n")

### 2022-07-21 eval/graphs.py

@@ -82,7 +82,7 @@ lines = table.to_latex(escape=False).split("\n")
 
 lines = lines[:2] + [
   R" &                &                & \multicolumn{3}{c}{Preprocessing [s]} \\ \cmidrule(lr){4-6}"
-  R" & Nodes          & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} \\ \cmidrule(lr){5-6}"
+  R" & Vertices       & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} \\ \cmidrule(lr){5-6}"
   R" & $[\cdot 10^6]$ & $[\cdot 10^6]$ &                     & Phase 1 & Phase 2 \\"
 ] + lines[4:]
 

### 2022-07-22 paper/ch_potentials.tex
further text improvements

@@
 
-% TODO rewrite, make self contained
-The paper introduced the basic CH-Potentials heuristic, some query optimizations for handling of low-degree vertices, common in road networks, and some applications of CH-Potentials.
-In this article, we discuss and evaluate Lazy RPHAST, the central ingredient of CH-Potentials, in greater detail and introduce an improved variant for Customizable Contraction Hierarchies.
-% Furthermore, we evaluate Lazy RPHAST in the context of one-to-many and many-to-many problems.
-The original paper only discussed unidirectional CH-Potentials.
-In this journal version, we also discuss a bidirectional variant and introduce an improved pruning criterion for symmetrical bidirectional A*.
-The improved pruning is not specific to CH-Potentials and can also be used with other heuristics.
-Compared to the conference paper, the journal version contains a significantly expanded experimental evaluation.
-
-The rest of the article is organized as follows.
-In Section~\ref{sec:related_work}, we discuss related works on goal directed search and extensions for realistic applications for hierarchical techniques.
-After establishing preliminaries in Section~\ref{sec:preliminaries}, we introduce the incremental many-to-one problem and Lazy RPHAST, our algorithm to solve it efficiently, in Section~\ref{sec:lazy-rphast}.
-CH-Potentials, our new A* heuristic is introduced in Section~\ref{sec:main-algo}.
-There, we also discuss our improvements to the handling of low-degree vertices (Section~\ref{sec:low-deg-improvment}) and for symmetrical bidirectional A* (Section~\ref{sec:bidir_astar}).
-% Section~\ref{sec:extensions} presents a multitude of applications of CH-Potentials.
-In Section~\ref{sec:extensions}, we demonstrate flexibility of our approach, by describing how to apply the approach to different practical applications.
-Finally, in Section~\ref{sec:experiments}, we present an experimental evaluation of our approach.
-
 \subsection{Related Work}\label{sec:related_work}
+
+% classical
+% goal, hierarchy, labeling
+% extended query (m2o)
+% extended application
+
+% TODO RPHAST etc
 
@@
 
-In this work, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem (Section~\ref{sec:lazy-rphast}).
-In Section~\ref{sec:main-algo}, we present CH-Potentials, a novel fast and tight heuristic for A* in road networks build on Lazy RPHAST.
-Further, we present several useful optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
-In Section~\ref{TODO}, we introduce a unified framework to apply CH-Potentials to a variety of practical route planning problems without any modifications to the preprocessing.
-Finally, an extensive experimental evaluation in Section~\ref{sec:experiments} confirms the applicability of our results.
+After discussing notation and fundamental algorithms in Section~\ref{sec:preliminaries}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem in Section~\ref{sec:lazy-rphast}.
+Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
+Our main contribution, the CH-Potentials framework is presented in Section~\ref{sec:framework} where we show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
+Based on this heuristic, we provide a unified framework (Section~\ref{sec:formal_framework} and~\ref{sec:chpot}) for a variety of practical route planning problems (Section~\ref{sec:applications}).
+CH-Potentials can be applied to all of these problems without any modifications to the preprocessing.
+In Section~\ref{sec:experiments}, we provide an extensive experimental evaluation analyzing the performance characteristics of our algorithms.
+It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
+However, the strength of our approach is not in query running times but in its flexibility:
+Problem extensions can be supported at the cost of some slowdown to the queries but without any algorithmic adjustments.
 
@@
 We denote this shortest path weight as the distance $\dist_w(s,t)$ between $s$ and $t$.
-If there is no path, we set $\dist_w(s,t)=\infty$. % TODO edge notation, degree (undirected)
+If there is no path, we set $\dist_w(s,t)=\infty$.
 
@@
 
-Dijkstra's algorithm can also be run from the target $t$ on the reversed graph $\overleftarrow{G}$. % TODO introduce reverse dist etc
+Dijkstra's algorithm can also be run from the target $t$ on the reversed graph $\overleftarrow{G}$.
 We call this a \emph{backward search}.
@@
 
-\subsection{Contraction Hierarchies (CH)}
+\subsection{Contraction Hierarchies}
 
@@
 \caption{
-Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$. $v^*$ is the mid vertex. % TODO different letter for m
+Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$.
 }
@@
 \end{figure}
-
-A CH is a two-phase technique to efficiently compute exact, shortest paths on road networks.
-For details, we refer to \cite{gssv-erlrn-12,dsw-cch-15}.
-In this section, we give an introduction.
 
@@
   $u\gets$ pop minimum element from $Q$\;
-  \For{all down-edges $vu \in \echd$}{
-    \If{$D^{\downarrow}[v] > w(vu) + D^{\downarrow}[u]$}{
-      $D^{\downarrow}[v]\gets w(vu) + D^{\downarrow}[u]$\;
+  \For{all reversed downward edges $uv \in \rechd$}{
+    \If{$D^{\downarrow}[u] + \overleftarrow{w}(uv) < D^{\downarrow}[v]$}{
+      $D^{\downarrow}[v] \gets D^{\downarrow}[u] + \overleftarrow{w}(uv)$\;
       Add $v$ or decrease $v$'s key in $Q$ to $D^{\downarrow}[v]$\;
@@
 
-In a CH, each vertex gets a unique rank representing its ``importance''.
-By convention, important vertices are usually depicted at the ``top'' while unimportant vertices are positioned at the ``bottom''.
-The intuition is that dead-ends are unimportant while highways or bridges are important.
-An edge goes \emph{up} when its tail vertex has lower rank than its head vertex.
-\emph{Down} edges are defined analogously.
-An \emph{up-down path} is a path where only one vertex $v^*$ is more important than both its neighbors.
-$v^*$ is called the \emph{mid} vertex.
-An \emph{up path} is a path where the last vertex is the mid vertex.
-Similarly, for a \emph{down path}, the first vertex has to be the mid vertex.
-% Every up and down path is an up-down path.
-%
-In the preprocessing phase, first, a vertex ordering is computed heuristically.
-Then, \emph{shortcut} edges are added to the input graph $G$ to obtain $G^+$.
-This is done by repeatedly contracting the least important vertex and adding shortcuts between its higher ranked neighbors.
-%See~\cite{gssv-erlrn-12} for the details.
-After the preprocessing, for every pair of vertices $s$ and $t$ there exists a shortest up-down $st$-path in $G^+$ with the same length as a shortest path in $G$.
+\emph{Contraction Hierarchies} (CH) is a two-phase speed-up technique to accelerate shortest path computations on road networks through precomputation.
+For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
+Here, we only briefly introduce necessary notation and algorithms used in this paper.
+In a preprocessing phase, vertices are ordered totally by ``importance'' where more important vertices should lie on more shortest paths.
+Intuitively, vertices on highways are more important than vertices on rural streets.
+For CH, such an ordering is obtained heuristically.
+The position of a vertex in the order is also denoted as its \emph{rank}.
+Vertices of higher rank are informally referred to and visualized as ``higher up'' in the hierarchy.
+Therefore, an edge where the tail has lower rank than the head is an \emph{upward} edge.
+Analogously, when the head vertex has the lower rank, the edge is said to go \emph{downward}.
+Once such an importance ordering was obtained, all vertices are contracted successively by ascending importance.
+To \emph{contract} a vertex means to temporarily remove it from the graph while inserting \emph{shortcut edges} between more important neighbors to preserve shortest distances among them.
+The result is an \emph{augmented graph} $G^+$ with original edges and shortcuts.
+We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echu)$ containing only downward edges.
+
+The augmented graph has the property that for any two vertices $s$ and $t$, there always exists an \emph{up-down} $st$-path, i.e.\ a path  which first uses only edges from $\gchu$ and then only edges from $\gchd$, with the same length as a shortest path in $G$.
 See Figure~\ref{fig:ch} for an illustration.
 From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
-Thus, we can restrict our search to up-down paths in $G^+$.
-The search is bidirectional.
-The forward search starts from $s$ and only follows up-edges.
-Similarly, the backward search starts at $t$ and only follows down-edges in reversed direction.
-By construction, the mid vertex must be in the intersection of the forward and backward search spaces.
+Such an up-down path can be found by running Dijkstra's algorithm from $s$ on $\gchu$ and from $t$ on the reversed downward graph $\rgchd$ graph.
+By construction, at least the highest ranked vertex on the up-down path must be in the intersection of the forward and backward search spaces.
 Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
-The forward search works analogously.
-% TODO \gchu and \gchd
-
-A CH query is fast, if the number of vertices reachable via only up- or down-edges is small.
-On road networks, this is the case~\cite{gssv-erlrn-12,dgpw-crprn-13}.
-On graphs with low treewidth, this is also the case~\cite{dsw-cch-15,hs-gbpo-18}.
-
-% Using the CH query algorithm, we can already give a simple heuristic.
-% The heuristic evaluation $h_t(x)$ performs a CH-query from $x$ to $t$.
-% This yields tight estimates but a high overhead for the heuristic evaluation.
-% While a single CH query is fast, answering one for every vertex explored in the $A^*$ search is slow.
-% Fortunately, we can do better.
-% For this we need another component called PHAST~\cite{dgnw-phast-13}.
-
-\subsection{Customizable Contraction Hierarchies (CCH)}
+The set of vertices reachable in $\gchu$ and $\rgchd$ is called the \emph{CH search space} of a vertex.
+The CH query performance strongly correlates with the size of this CH search space and the number of edges between the vertices in the search space.
+Luckily, on road networks, the CH search space is small~\cite{gssv-erlrn-12,dgpw-crprn-13}.
+
+\subsection{Customizable Contraction Hierarchies}
 \label{sec:cch-intro}
 
-Some hierarchical techniques were extended to support a three-phase setup.
+Some hierarchical techniques were extended to support a three-phase setup~\cite{dgpw-crprn-13}.
 The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
 The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
-The preprocessing phase is independent of the weights and only exploits the graph topology.
+The preprocessing phase is independent of the weights and only uses the graph topology.
 Weights are introduced in the customization phase.
 Paths are computed in the query phase.
-This setup is useful to allow for weight changes that are common to all vehicles such as live traffic updates.
-The first technique to follow this setup is CRP~\cite{dgpw-crprn-13}. % TODO better clear up whats related work and whats prelim
-
-Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) are a variant of Contraction Hierarchies that follows this three-phase setup.
+
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) is a variant of Contraction Hierarchies that follows this three-phase setup.
 The key idea is to use a vertex importance order that is independent of the weight function.
-CCH use \emph{nested dissection} orders for this, i.e.\ a small balanced vertex separator is computed and the separator vertices get the highest importance.
+CCH use \emph{nested dissection} orders for this.
+A small balanced vertex separator is computed and the separator vertices get the highest importance.
 The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
@@
 Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
-
 Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
@@
 The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
-However, discussing them is beyond the scope of this work.
+However, discussing them in detail is beyond the scope of this work.
 See~\cite{dsw-cch-15,BuchholdSW19} for details.
@@
 For every other vertex its parent is its least important upward neighbor.
-In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to the vertices reachable in a CH search.
+In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to its CH search space.
 This makes it possible to explore the CCH search space with a more efficient shortest path algorithm:
@@
 Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
-With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a milliseconds.
+With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a millisecond.
 CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
-However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries are the faster variant.
+However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out the faster variant.
 Note that only nested dissection orderings admit an elimination tree of low height.
-The elimination tree query can therefore not be used for a general CH.
+The elimination tree query can therefore not be used with a classical CH.
 
@@
 
-\section{CH-Potentials: A New A* Heuristic}\label{sec:main-algo}
-
-In this section, we describe how to use the Lazy RPHAST algorithm to build a fast and tight heuristic for A*.
-We also discuss CH-Potentials for the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
-Further, we present several low degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
-Finally, we introduce an algorithmic framework to apply A* with CH-Potentials to a variety of practical route planning problems.
-
-The \emph{CH-Potentials} heuristic is a straightforward application of the Lazy RPHAST algorithm.
-In the beginning of each query, we perform the target selection, a backward CH search, from the target $t$.
-The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $v$.
-
-This yields a \emph{perfect} heuristic, at least as long as the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function.
-This case is of course not particularly interesting.
-One could just answer the shortest path query directly with a CH query.
-This approach becomes useful when the query runs on a \emph{different but related} graph or weight function than the preprocessing.
-We will discuss such applications in Section~\ref{sec:extensions}.
-But before that, we discuss two extensions/optimizations to A* which we will utilize.
+\section{Optimizations for A* in Road Networks}\label{sec:astar_opts}
+
+In this section, we propose optimizations for A* in road networks.
+First, we present several low degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
+Second, we discuss the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
+These optimizations can be used with \emph{any} A* heuristic.
 
@@
 
-\subsection{Applications}\label{sec:extensions}
-
-% A* is a flexible algorithm.
-Many extended problems can be solved using CH-Potentials.
-We start by establishing a common formal framework for the use of CH-Potentials.
-Then, we exemplary describe some extended routing problems and how to apply CH-Potentials to them.
-
-\subsubsection{Formal Setup: Inputs, Outputs, and Phases}
-
-% TODO make sure to distinguish PROBLEM model from SOLUTION appraoch
+\section{The CH-Potentials Framework}\label{sec:framework}
+
+% , we describe how to use the Lazy RPHAST algorithm to build a fast and tight heuristic for A*.
+In this section, we introduce an algorithmic framework to apply A* with a Lazy RPHAST-based heuristic to a variety of practical route planning problems.
+We call our approach \emph{CH-Potentials}.
+The core idea is to compute a CH augmented graph during preprocessing and use A* with a straightforward application of Lazy RPHAST as the heuristic to answer queries.
+However, when the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function, this yields a \emph{perfect} heuristic
+This case is, of course, not particularly interesting.
+One could just answer the shortest path query directly with a CH query.
+The approach becomes useful when the query runs on a \emph{different but related} graph or weight function than the preprocessing.
+Therefore, we start by establishing a common formal framework for the use of CH-Potentials.
+Then, we exemplarily describe some extended routing problems and how to apply the CH-Potentials framework to them.
+
+\subsection{Formal Setup: Inputs, Outputs, and Phases}\label{sec:formal_framework}
 
@@
 In the simplest case, $w_q$ will be scalar edge weights.
-However, this is not a requirement.
+Live traffic is an example for this.
+The challenge in this scenario is that values of $w_q$ might change between queries.
+However, $w_q$ can also represent something more complex than scalar numbers.
 It can be any function that computes a weight for an edge.
 This function can also take additional parameters from the state of the search.
-For example, in the case of live-traffic, $w_q$ represents scalar edge weights.
-However, values of $w_q$ might change between queries.
 In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
@@
 To enable quick shortest path computations, we consider a two-phase setup with an additional offline preprocessing phase before the online query phase.
-The input to the preprocessing phase is a graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$.
-$w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
+The input to the preprocessing phase is a lower bound graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$ where $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
 The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
-In the applications considered in this paper, $w_\ell$ is always the freeflow travel time.
-
-The query phase uses this auxiliary data in an A* search between vertices $s$ and $t$ on $G_q = (V_q, E_q)$ with weights $w_q$.
+In the applications considered in this paper, $w_\ell$ is always the \emph{freeflow} travel time.
+
+The query phase may use this auxiliary data to answer shortest path queries between vertices $s$ and $t$ on $G_q = (V_q, E_q)$ with weights $w_q$.
 Let $\phi : V_q \to V_{\ell}$ denote a function mapping vertices in the query graph to vertices in the lower bound graph.
-We require that $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ from $\phi(u)$ to $\phi(v)$ in $G_\ell$ with respect to $w_\ell$.
-Then, A* uses the heuristic function $h_t(x) = \dist_{w_{\ell}}(\phi(x), \phi(t))$ computed using Lazy RPHAST.
-Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
-The exact implementation of the A* search depends on the application.
+The crucial requirement for a routing problem to fit into our framework is that the query weight of an edge $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ between the corresponding vertices $\phi(u)$ and $\phi(v)$ in the lower bound graph $G_\ell$ with respect to $w_\ell$.
+Typically, and unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
+
+\subsection{CH-Potentials}\label{sec:chpot}
+CH-Potentials can be used to solve any problem in this framework.
+The preprocessing is always the computation of the CH augmented graph and remains the same regardless of the specific routing problem.
+The query consists of A* with the heuristic function $h_t(v) = \dist_{w_{\ell}}(\phi(v), \phi(t))$ computed using Lazy RPHAST.
+In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
+The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $v$ (see Algorithm~\ref{algo:pot}).
+Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
 Our approach only provides the heuristic $h_t$ for the A* search.
-In contrast, the preprocessing phase remains the same for all applications.
-
+As the performance of A* depends crucially on the accuracy of the heuristic estimates, the smaller the difference between query weights and lower bound distances, the better CH-Potentials will perform.
+
+\paragraph{Correctness}
 Our heuristic is always \emph{feasible}, i.e. $w_q(u v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
@@
 
+\subsection{Applications}\label{sec:applications}
 \subsubsection{Avoiding Tunnels and/or Highways}
@@
 % Afterwards, we describe how it can be solved with CH-Potentials.
-A \emph{turn weight} $w_t : E \times E \to \mathbb{R}^{\geq 0}$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns. % TODO formally
+A \emph{turn weight} $w_t : E \times E \to \mathbb{R}^{\geq 0}$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns.
 For CH-Potentials, we use zero as lower bound for every turn weight in the heuristic.
@@
 The FIFO property states that it is not possible to arrive earlier by departing later.
-Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\delta\in \mathbb{R}^+: w_q(e, \tau) \le w_q(e, \tau+\delta) + \delta$.
-Our implementation stores edge travel times using piece-wise linear functions.
+Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\varepsilon\in \mathbb{R}^{>0}: w_q(e, \tau) \le w_q(e, \tau+\varepsilon) + \varepsilon$.
+Our implementation stores edge travel times using piecewise linear functions.
 The A* search uses the tentative distance $\tau$ at a vertex $u$ when evaluating the travel time of outgoing edges $u v$.
@@
 % With this setup, we extended CH-Potentials to support time-dependent routing.
-By keeping travel time functions out of the CH, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-21,dn-crdtd-12} which have to create shortcuts of travel time functions.
+By keeping travel time functions out of the CH preprocessing, we avoid a lot of algorithmic complications compared to~\cite{bgsv-mtdtt-13,bdpw-dtdrp-16,swz-sfert-21,dn-crdtd-12} which have to create shortcuts of travel time functions.
 
@@
 
-\subsubsection{CCH Potentials}
+\subsubsection{Live Traffic with CCH-Potentials}
 
 Live traffic can alternatively be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
-We do this by exchanging the CH for a CCH.
-The preprocessing phase is identical as for a CCH and is independent of $w_\ell$.
-The $w_\ell$ is fixed during the customization phase.
-It can be changed arbitrarily by rerunning the customization.
-The A* algorithm is part of the query phase.
-The weights $w_q$ used in the query phase must not be smaller than the $w_\ell$ used during the last customization.
-
-This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic}.
-Changes to the live traffic situation are handled by changing $w_\ell$ via a customization.
+For this, we use CCH instead of CH preprocessing to build the augmented graph.
+This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization algorithm.
+This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic} where live traffic was represented only in $w_q$.
+In contrast, with CCH-Potentials, changes to the live traffic situation are handled by also changing $w_\ell$ via a customization.
 Additional features such as turn restrictions can be implemented by adjusting $w_q$.
 For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
-However, it comes at the expense of a more complex algorithm.
+However, it comes at the expense of a more complex setup.
+Also, the customization is not fast enough to rerun it for every query.
 Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.
@@
 
-% \subsubsection{Alternative Routes with the Penalty Method}
+% \subsection{Alternative Routes with the Penalty Method}
 
@@
 \section{Evaluation}
-
 \label{sec:experiments}
@@
 \centering
-\caption{Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials. With CCH-Potentials $w_{\ell}$ can be updated by rerunning Phase 2 again.}\label{tab:graphs}
+\caption{
+Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials.
+With CCH-Potentials, $w_{\ell}$ can be updated by rerunning Phase 2 again.
+}\label{tab:graphs}
 \input{table/graphs.tex}
@@
 Table~\ref{tab:graphs} contains an overview over the instances.
-In this table, we further include the sequential preprocessing running time to construct the (C)CH.
-CH are constructed with RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
+In this table, we further include the sequential preprocessing running time to construct the augmented graphs and the space consumption.
+We use the CH preprocessing from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
 For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
@@
 Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target vertex and $2^{14}$ sources for different ball sizes.
-This experiment uses the exact methodology described in~\cite{delling_et_al:OASIcs:2011:3266} which allows to roughly relate our results to the performance of other one-to-many algorithms.
+The query generation methodology is the same as in the previous experiment.
+As this is the same methodology also used in~\cite{delling_et_al:OASIcs:2011:3266}, we can relate our results to the performance of other one-to-many algorithms.
 Keep in mind that algorithms like RPHAST optimize for a different setting than we do.
@@
 While the numbers are not perfectly comparable due to different benchmark machines,\footnote{
-According to the comparison methodology from~\cite{BastDGMPSWW16} (see \url{https://i11www.iti.kit.edu/~pajor/survey/}), the machine used in~\cite{delling_et_al:OASIcs:2011:3266} (SPA-2) is about 30\% slower than ours (we obtained a score of 33159\,ms).
+According to the comparison methodology from~\cite{bdgmpsww-rptn-16} (see \url{https://i11www.iti.kit.edu/~pajor/survey/}), the machine used in~\cite{delling_et_al:OASIcs:2011:3266} (SPA-2) is about 30\% slower than ours (we obtained a score of 33\,159\,ms).
 However, these scaling factors have to be interpreted very carefully.
@@
 The BCC, Deg2 and Deg3 columns indicate which optimizations from Section~\ref{sec:low-deg-improvment} were used.
-}\label{tab:building_blocks} % TODO explain labels and justify 1.05
+}\label{tab:building_blocks}
 \input{table/building_blocks.tex}
@@
 
-Table~\ref{tab:applications} depicts the running times of (C)CH-Potentials in various applications, such as those described in Section~\ref{sec:extensions}.
+Table~\ref{tab:applications} depicts the running times of (C)CH-Potentials in various applications, such as those described in Section~\ref{sec:applications}.
 We report speedups compared to extensions of Dijkstra's algorithm for each application respectively.
@@
 
+\subsubsection{Comparison with Related Work}
+
+While the query running times reported in Table~\ref{tab:applications} are decent in many settings, they are not competitive to techniques tailored to specific applications.
+Again, these numbers are not perfectly comparable due to different benchmark machines.
+Nevertheless, the overall picture is clear enough.
+In the simple $w_q = w_{\ell}$ setting, Hub Labels can be used to answer queries in less than a microsecond~\cite{TODO}, more than three orders of magnitude faster than with CH-Potentials.
+Live traffic and arbitrary weight functions can be handled with CCH resulting in query times of around a 0.1\,ms~\cite{BuchholdSW19}.
+CCH can also handle turn costs~\cite{bwzz-cchtc-20}.
+With some algorithmic adjustments to the preprocessing, query times around an order of magnitude faster than CCH-Potentials are possible (0.3\,ms compared to 2.1\,ms with CCH-Potentials).
+For time-dependent routing, CATCHUp~\cite{swz-sfert-21} achieves query times more than an order of magnitude faster than CH-Potentials (6.3\,ms compared to our 97.7\,ms).
+However, the advantage of the CH-Potentials framework over these fine-tuned techniques is that it is unified and flexible approach which can handle all these applications without any adjustments to the preprocessing.
+Further, CH-Potentials preprocessing times are at least an order of magnitude faster than approaches like CATCHUp (sequentially more than four hours on TDEur20 compared to five minutes with CH-Potentials) and require significantly less memory (Hub Labels, for example, needs 20\,GiB on DIMACS Europe, compared to around 750\,MiB with CH-Potentials, see Table~\ref{tab:graphs}).
+Finally, to the best of our knowledge, for problem settings such as the combination of predicted and live traffic, there simply does not exist any exact technique able to handle this setting, let alone additionally integrating turn costs.
+The key achievement of CH-Potentials is that problem extensions can be integrated by trading query performance rather than developing new algorithms.
+
 % \subsubsection{Alternative Routes}
@@
 In this article, we introduced Lazy RPHAST, an efficient algorithm for the incremental many-to-one problem.
-We also showed that it delivers competitive performance for many-to-one problems.
-Based on Lazy RPHAST, we developed (C)CH-Potentials, a fast, exact, and flexible two-phase algorithm based on A* and (C)CH for finding shortest paths in road networks.
-The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity.
+We showed that it also delivers competitive performance for many-to-one problems.
+Based on Lazy RPHAST, we developed CH-Potentials, a fast, exact, and flexible two-phase technique based on A* and CH for finding shortest paths in road networks.
+The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity and without any changes to the preprocessing algorithms.
 CH-Potentials provides \emph{exact} distances with respect to lower bound weights known at preprocessing time as an A* heuristic.

### 2022-07-22 paper/table/graphs.tex
further text improvements

--- 2021-10-04 paper/table/graphs.tex
@@
-\begin{tabular}{lrrrrr}
+\begin{tabular}{lrrrrrrr}
 \toprule
- &                &                & \multicolumn{3}{c}{Preprocessing [s]} \\ \cmidrule(lr){4-6} & Nodes          & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} \\ \cmidrule(lr){5-6} & $[\cdot 10^6]$ & $[\cdot 10^6]$ &                     & Phase 1 & Phase 2 \\
+ &                &                & \multicolumn{3}{c}{Preprocessing [s]} & \multicolumn{2}{c}{Aux. Data [MiB]} \\ \cmidrule(lr){4-6} \cmidrule(lr){7-8}
+ & Vertices          & Edges          & \multirow{2}{*}{CH} & \multicolumn{2}{c}{CCH} & \multirow{2}{*}{CH} & \multirow{2}{*}{CCH} \\ \cmidrule(lr){5-6} & $[\cdot 10^6]$ & $[\cdot 10^6]$ &                     & Phase 1 & Phase 2 \\
 \midrule
-OSM Ger       &       16.2 &       35.4 &                          312.8 &        1\,466.7 &          10.1 \\
-DIMACs Europe &       18.0 &       42.2 &                          276.2 &        2\,480.9 &          12.4 \\
-TDGer06       &        4.7 &       10.8 &                           59.2 &         331.7 &           2.7 \\
-TDEur17       &       25.8 &       55.5 &                          293.9 &        2\,102.3 &          14.1 \\
-TDEur20       &       28.5 &       60.9 &                          311.9 &        2\,219.5 &          15.2 \\
+OSM Germany   &       16.2 &       35.4 &                          298.7 &        1\,467.4 &          10.1 &  645 & 616 \\
+DIMACS Europe &       18.0 &       42.2 &                          276.2 &        2\,480.9 &          12.4 &  742 & 765 \\
+TDGer06       &        4.7 &       10.8 &                           59.2 &           331.7 &           2.7 &  196 & 169 \\
+TDEur17       &       25.8 &       55.5 &                          293.9 &        2\,102.3 &          14.1 & 1030 & 881 \\
+TDEur20       &       28.5 &       60.9 &                          311.9 &        2\,219.5 &          15.2 & 1130 & 959 \\
 \bottomrule

### 2022-07-22 eval/applications.py

@@ -19,7 +19,7 @@ def path_to_graph(path):
         'osm_europe': 'OSM Europe',
         'osm_ger': 'OSM Ger',
         'osm_ger_td': 'OSM Ger',
-        'europe': 'DIMACs Eur',
+        'europe': 'DIMACS Eur',
     }[[x for x in path.split('/') if x != ''][-1]]
 
 algos = ['Virtual Topocore Component Query', 'Virtual Topocore Bidirectional Core Query']

### 2022-08-02 paper/ch_potentials.tex
rewrite intro

@@
 \affiliation{%
-  \institution{Institute of Theoretical Informatics, Algorithmics I, Karlsruhe Institute of Technology}
+  \institution{Institute of Theoretical Informatics, Algorithmics, Karlsruhe Institute of Technology}
   \city{Karlsruhe}
@@
 As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
-Our approach allows decoupling the supported applications from the hierarchical speed-up technique.
+Our approach allows decoupling the supported applications from the hierarchical speedup technique.
 Additionally, we describe A* optimizations to accelerate the processing of low degree vertices, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
@@
 The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
-Unfortunately, on continental sized networks, it is too slow for many applications.
-Thus, speed-up techniques have been developed.
-One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
-They have been used successfully in many real world applications.
-A CH exploits the inherent hierarchy of road networks.
-In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
-Another popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
-It is also used in practice~\cite{bingblog}.
-MLD also uses shortcut edges.
-Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.
-
-\begin{figure}
-\centering
-\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
-\caption{Vertices explored by A*. Color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
-\label{img:search-space}
-\Description[A map where vertices visited by A* are marked.]{
-A map of Karlsruhe where vertices visited by A* are marked.
-The shape of the area with the visited vertices is similar to a drop.
-The source vertex roughly lies in the center of the drop, the target vertex at the tip.
-}
-\end{figure}
-
-Unfortunately, for many real world applications, this basic graph model is too simplistic.
+Unfortunately, on continental-sized networks, it is too slow for many applications.
+Therefore, speedup techniques have been developed.
+These techniques employ an offline \emph{preprocessing} phase where auxiliary data is precomputed.
+This auxiliary data is then utilized to accelerate shortest path computations in the online \emph{query} phase.
+With this approach, speedups of more than three orders of magnitude have been achieved.
+This allows interactive query times of milliseconds or less on continental-sized road networks.
+
+However, for many real world applications, this basic model is too simplistic.
 For realistic routing, many additional features need to be considered.
-This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions. %, temporary driving bans for certain types of vehicles.
+This includes turn costs and restrictions, live traffic, user preferences, and traffic predictions.
 Some applications may have additional application-specific requirements.
 Extending Dijkstra's algorithm to support these features is usually easy.
-Extending hierarchical speed-up techniques is often possible, too.
-However, the algorithm development is vastly more complex. % TODO cite some papers
-For every feature, dedicated research papers exist that extend CH and other approaches.
-Supporting the combination of several features is even harder.
-For example, we are not aware of any work combining all features mentioned above.
-In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features.
-
-Our approach decouples the development of extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
-A* is a goal-directed variant of Dijkstra's algorithm.
-See Figure~\ref{img:search-space} for an example of vertices traversed during an A* search.
-A* uses a \emph{heuristic} to guide the search towards the goal.
-A heuristic is a function that maps a vertex to an estimate of the remaining distance to the goal.
-A*'s running time crucially depends on how tight this estimate is.
-Further, evaluating the heuristic must be fast.
-% The free flow travel time to the goal is a valid heuristic for most features.
-% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
-% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
-In this article, we introduce Lazy RPHAST and build on it CH-Potentials, a fast heuristic with tight estimates.
-Internally, the heuristic uses a CH.
-Fortunately, this is an implementation detail from the perspective of the A*.
-To support a new feature, we only need to modify the A* algorithm.
-The heuristic core containing the CH remains untouched.
-Extending A* is vastly easier than extending a CH.
-This enables us to design algorithms for a multitude of features.
+In contrast, extending speedup techniques is vastly more complex.
+The results on extending specific speedup techniques to specific extended problem fill many research papers~\cite{dgpw-crprn-13} and sometimes entire dissertations~\cite{b-tdrpc-14}.
+% TODO cite some papers
+We will present a brief overview in Section~\ref{sec:related_work}.
+While techniques achieving fast query times have been successfully developed for many extended scenarios, we observe two problems:
+The resulting techniques are complex to implement and hard to further extend and adjust.
+For example, while there are efficient speedup techniques for each of the above mentioned features, we are not aware of any work supporting the combination of these features.
+We believe that for \emph{practical} algorithms, a different trade-off is necessary.
+Therefore, in this work, we prioritize a simple and extensible approach over the fastest possible query times.
+
+% In this article, we describe an algorithmic building block, that allows handling the combination of all above-mentioned features.
+% Our approach decouples the development of extensions from the hierarchical speedup technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
+% A* is a goal-directed variant of Dijkstra's algorithm.
+% See Figure~\ref{img:search-space} for an example of vertices traversed during an A* search.
+% A* uses a \emph{heuristic} to guide the search towards the goal.
+% A heuristic is a function that maps a vertex to an estimate of the remaining distance to the goal.
+% A*'s running time crucially depends on how tight this estimate is.
+% Further, evaluating the heuristic must be fast.
+% % The free flow travel time to the goal is a valid heuristic for most features.
+% % In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
+% % However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
+% In this article, we introduce Lazy RPHAST and build on it CH-Potentials, a fast heuristic with tight estimates.
+% Internally, the heuristic uses a CH.
+% Fortunately, this is an implementation detail from the perspective of the A*.
+% To support a new feature, we only need to modify the A* algorithm.
+% The heuristic core containing the CH remains untouched.
+% Extending A* is vastly easier than extending a CH.
+% This enables us to design algorithms for a multitude of features.
 
@@
 
-% classical
-% goal, hierarchy, labeling
-% extended query (m2o)
+In the past two decades, a considerable amount of research effort has been put into accelerating shortest path computations in transportation networks.
+For a comprehensive overview, we refer to~\cite{bdgmpsww-rptn-16}.
+Here, we first highlight a few key results for the classical shortest path problem on road networks specifically relevant to our work.
+In the second part of this section, we will discuss adaptions of these techniques to extended problem models.
+
+\paragraph{Speedup Techniques for Shortest Paths in Road Networks}
+A*~\cite{hnr-afbhd-68} is a well-known approach to accelerate shortest path queries by directing the search towards the target through heuristic distance estimates.
+It has been successfully employed in many problems beyond route planning in transportation networks~\cite{DBLP:conf/socs/StrasserHB14,DBLP:conf/ijcai/BonoGHS19,DBLP:conf/ijcai/0002UJAKK18}.
+It is one of the algorithms fundamental to our work.
+The performance of A* crucially depends on the quality of the heuristic estimates.
+A natural but rather ineffective heuristic can be obtained through euclidian distances.
+On road networks, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
+A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
+ALT was one of the early speedup technique for routing in road networks.
+It uses precomputed distances to certain landmark vertices combined with the triangle inequality to compute distance estimates.
+On road networks, this achieves speedups of around two orders of magnitude over Dijkstra's algorithm.
+However, the speedups obtained by ALT are still limited by the quality of the distance estimates.
+To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
+Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal directed search on.
+This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
+Since then, there has been little development on A*-based techniques for routing in road networks.
+This is because hierarchical techniques have proven much more effective.
+
+Hierarchical speedup techniques utilize the inherent hierarchy in road networks, usually in combination with the \emph{contraction} of less important vertices.
+One popular example is Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
+In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network at query time.
+The ``importance'' of vertices is determined heuristically during preprocessing.
+This results in queries taking less than a tenth of a millisecond which is a speedup of four or more orders of magnitude over Dijkstra's algorithm.
+The preprocessing takes only a few minutes.
+CH has been used successfully in many real world applications and quite a few open-source implementations exist\footnote{
+\url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
+While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
+}.
+CH is another fundamental algorithm for our work.
+More specifically, we extend on PHAST~\cite{dgnw-phast-13} and RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
+These algorithms reuse the CH preprocessing but modify the query phase to efficiently support \emph{one-to-all} queries from one source vertex to \emph{all} other vertices (PHAST), or \emph{one-to-many} queries to a subset of vertices known in advance (RPHAST).
+%
+Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
+It is based on a multi-level partition of the road network computed during preprocessing.
+Queries are slightly slower than CH queries but still three to four orders of magnitude faster than Dijkstra's algorithm.
+Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
+However, these works also primarily focused on obtaining faster queries rather than simple and flexible approaches.
+
+The speedup technique known to currently achieve the fastest queries is Hub Labels (HL)~\cite{TODO}.
+With HL, query times in less than a microseconds are possible.
+This is only a couple of times the time of an uncached memory access.
+This comes at the cost of an expensive preprocessing phase and a tremendous memory footprint, more than an order of magnitude larger than the input graph.
+Despite the extremely fast query times, it appears that the trade-off offered by HL is, in practice, often not very attractive.
+Simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (network latency, etc.).
+Also, we are not aware of any works extending HL to extended problem models.
+
 % extended application
-
-% TODO RPHAST etc
-
-There is a lot of work that extends hierarchical speed-up techniques to more complex settings~\cite{bdgmpsww-rptn-16}.
-For example, in~\cite{gv-errnt-11} turn information is integrated into CH.
+\paragraph{Speedup Technique Adaptions for Extended Problems}
+One of the earliest approaches to adapt speedup techniques to extended problems was based on A*.
+In~\cite{dw-lbrdg-07}, ALT is used for routing on dynamic and time-dependent graphs.
+This approach is conceptually similar to ours.
+However, because the heuristic is ALT-based, the accuracy of the estimates is limited and the resulting query performance not competitive.
+Further, the evaluation was performed only with synthetic traffic data.
+With production-grade real-world traffic data, the problem becomes significantly harder~\cite{swz-sfert-21}.
+Similar to the development for the classical problem, the approach was also combined with contraction~\cite{ndls-bastd-12,dn-crdtd-12}.
+Even though this improves query running times, they obtained speedups are not competitive with purely hierarchical techniques.
+
+CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
+For this, a second, faster, preprocessing phase is introduced which takes at most a few seconds.
+This phase is called the \emph{customization}.
+It can be run regularly to update weights.
+This enables the integration of live traffic and user preferences.
+CRP was designed to support turn costs without any additional modifications.
+Queries in CRP take at most a few milliseconds.
+CRP is one of the few examples of a speedup technique designed for flexibility rather than maximum query performance.
+However, its flexibility is also limited.
+Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
+Achieving reasonable memory consumption and query times was only possible by giving up exactness.
+Further, TD-CRP can only compute approximate shortest distances (rather than paths).
+
+There is also lot of work that extends CH to more complex problem models.
+In~\cite{gv-errnt-11}, turn information is integrated into CH.
 The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower compared to classical CH.
-
+In~\cite{dsw-cch-15}, CH is extended to Customizable Contraction Hierarchies (CCH).
+CCH also has a second preprocessing (customization) phase where weights can be altered.
+This allows to support user preferences and live traffic with queries an order of magnitude faster than CRP.
+Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
+Even with all improvements proposed in~\cite{bwzz-cchtc-20}, turn integration still causes a slowdown of at least factor three to the customization phase.
 A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
@@
 
-CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
-For this, a faster, second preprocessing phase is introduced.
-It can be run regularly to update weights.
-In theory, this enables the integration of live traffic and user preferences.
-In practice, live traffic feed data is imperfect.
-Computing ``good'' routes without undesired detours due to artifacts in the data requires additional algorithmic extensions~\cite{dss-tarrn-18}.
-CRP also supports turn costs.
-Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
-On continental sized networks, TD-CRP can only compute approximate shortest distances (rather than paths).
-
-In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH). % TODO paragraph looks ugly, somehow make more compact
-CCH also has a second preprocessing phase where weights can be altered.
-Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
-Turn integration still causes a slowdown of at least factor three to the second preprocessing phase.
-
-Other extensions studied include electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11} or multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10}.
-
-% Another feature which has received quite some attention is the computation alternative routes~\cite{bdgs-argrn-11,adgw-arrn-13}.
+Finally, note that there are many other extended problem models.
+While they are beyond the scope of this work, they highlight the need for extensible speedup techniques.
+Examples for this are electric vehicle routing~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11},
+multi-criteria optimization~\cite{fns-opca-14,gks-rpfof-10},
+computing alternative routes~\cite{bdgs-argrn-11,adgw-arrn-13},
+routing with incomplete and noisy traffic data~\cite{dss-tarrn-18},
+and routing for trucks~\cite{kswz-erptd-p-20}.
+
 % Even formally defining what makes a good alternative route is non-trivial~\cite{bdgs-argrn-11}.
@@
 
-Clearly, extending hierarchical approaches is possible, but it requires significant research effort.
+% Clearly, extending hierarchical approaches is possible, but it requires significant research effort.
 % While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
@@
 What does this mean?
-Further, extensions to these speedup techniques cannot easily be composed with each other.
-Combining these extensions is an unsolved problem.
-
-CH-Potentials is not the first technique to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, previous works mostly focused on accelerating hierarchical approaches further rather than exploiting A*'s flexibility.
-
-ALT~\cite{gh-cspas-05,gw-cppsp-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
-% Just as our approach, ALT is A*-based.
-ALT has been combined with shortcuts~\cite{bdsssw-chgds-10} and also extended for dynamic graphs~\cite{dw-lbrdg-07} and time-dependent routing~\cite{ndls-bastd-12,dn-crdtd-12}.
-% However, ALT's heuristic is not preprocessing-tight.
-%
-CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
-A CPD can quickly compute the first edge of a shortest path between any two vertices.
-In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
-For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
-Unfortunately, the employed CPD's quadratic preprocessing running time is problematic on large road networks.
-%
-% MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
-%
-In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap. % such that distances in space and distances in the graph roughly correspond.
-The Euclidean distance is then used as a distance estimate for A*.
-% This results in a good but not a preprocessing-tight heuristic.
+% Further, extensions to these speedup techniques cannot easily be composed with each other.
+% Combining these extensions is an unsolved problem.
 
@@
 
-After discussing notation and fundamental algorithms in Section~\ref{sec:preliminaries}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem in Section~\ref{sec:lazy-rphast}.
+Clearly, there is a plethora of efficient speedup techniques for routing in road networks.
+Further, for many extended problems there are dedicated research results on extensions of some techniques.
+While many problems can be solved very efficiently, we still believe that this situation is not satisfactory in practice.
+For practical applications, a simple, unified, and extensible approach is often more important than the fastest query performance.
+Therefore, in this work, we revisit the A* algorithm and build a flexible, unified framework for a multitude of routing problems.
+The core of our approach is a new CH-based A* heuristic.
+It allows for much tighter estimates than previous A* heuristics and thus significantly faster queries.
+While the query running times of our approach are for the most part not competitive with approaches significantly tailored to specific problems, they are still sufficient for practical applications.
+The remainder of this work is organized as follows:
+
+We discuss notation and fundamental algorithms in Section~\ref{sec:preliminaries}.
+In Section~\ref{sec:lazy-rphast}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem.
 Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
-Our main contribution, the CH-Potentials framework is presented in Section~\ref{sec:framework} where we show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
+Our main contribution, the CH-Potentials framework, is presented in Section~\ref{sec:framework}.
+There, we first show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
 Based on this heuristic, we provide a unified framework (Section~\ref{sec:formal_framework} and~\ref{sec:chpot}) for a variety of practical route planning problems (Section~\ref{sec:applications}).
@@
 However, the strength of our approach is not in query running times but in its flexibility:
-Problem extensions can be supported at the cost of some slowdown to the queries but without any algorithmic adjustments.
+Problem extensions can be supported without complex algorithmic adjustments.
+The price for each extension is a certain query slowdown depending on how tight the lower bounds used during preprocessing remain.
 
@@
 A* is a goal-directed variant of Dijkstra's algorithm.
-% A* uses a \emph{heuristic} to guide the search towards the goal.
 It uses a heuristic function $h_t(v)$ that maps a vertex $v$ onto an estimate of the distance from $v$ to the target $t$.
@@
 This results in vertices closer to the target being explored earlier.
+Figure~\ref{img:search-space} depicts an example of the search space of the A* algorithm.
+With Dijkstra's algorithm, all vertices closer to $s$ than the target would have been explored.
+Thus the search space would be something resembling a circle around $s$.
+With A*, the search is directed towards the target and fewer vertices are explored.
 Dijkstra's algorithm is a special case of A* with $h_t(v)=0$.
+
 A* is equivalent to running Dijkstra's algorithm on the graph with \emph{reduced weights}~\cite{hnr-afbhd-68}.
@@
 % TODO more explicit details on A* correctness?
+
+\begin{figure}
+\centering
+\includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
+\caption{Vertices explored by A*. Color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
+\label{img:search-space}
+\Description[A map where vertices visited by A* are marked.]{
+A map of Karlsruhe where vertices visited by A* are marked.
+The shape of the area with the visited vertices is similar to a drop.
+The source vertex roughly lies in the center of the drop, the target vertex at the tip.
+}
+\end{figure}
 
@@
 
-\emph{Contraction Hierarchies} (CH) is a two-phase speed-up technique to accelerate shortest path computations on road networks through precomputation.
+\emph{Contraction Hierarchies} (CH) is a two-phase speedup technique to accelerate shortest path computations on road networks through precomputation.
 For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
@@
 From $t$, all reachable vertices via reversed down-edges are explored, as shown in Algorithm~\ref{algo:ch-backward}.
-For the second step, PHAST organizes vertices into \emph{levels}.
+For the second step, PHAST utilizes an assignment of vertices into \emph{levels}.
 These levels correspond to the importance ordering but allow multiple vertices in the same level.
@@
 All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.
-PHAST then iterates over all CH levels from top to bottom.
+
+The main work of the second step consists in iterating over all CH levels from top to bottom.
 In each iteration, all up-edges starting within the current level are followed in reverse.
@@
 Given a set of source vertices $S$, the first step is to copy the combined search space of all sources into a \emph{restricted subgraph}.
-Formally, let $V_{\operatorname{restr}}$ be the set of vertices reachable in $\gchu$ from any $s_i \in S$.
+Let $V_{\operatorname{restr}}$ be the set of vertices reachable in $\gchu$ from any $s_i \in S$.
 The restricted subgraph $\gchu_{\operatorname{restr}}$ is subgraph of $\gchu$ induced by $V_{\operatorname{rest}}$.
-Copying the relevant edges into this restricted subgraph makes up the \emph{selection} step.
-Then, in the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
+Finding and copying the relevant edges into this restricted subgraph makes up the \emph{selection} step.
+In the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
 RPHAST is particularly effective when many targets are queried for the same source set $S$.
@@
 Each optimization reduces both queue pushes and running times.
-All optimizations yield a combined speed-up of around 3.
+All optimizations yield a combined speedup of around 3.
 % Skipping degree three vertices has the smallest impact.
@@
 Table~\ref{tab:bidir_pruning} studies the impact of our improved pruning and the low degree optimizations.
-As observed in the previous section, enabling the low degree optimization achieves a speed-up by roughly a factor of three.
+As observed in the previous section, enabling the low degree optimization achieves a speedup by roughly a factor of three.
 Symmetric bidirectional A* without our improved pruning has the worst performance.
@@
 In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
-Interestingly, only the zero heuristic and ALT consistently achieve speed-ups through bidirectional search on all query weights.
+Interestingly, only the zero heuristic and ALT consistently achieve speedups through bidirectional search on all query weights.
 With $w_q = w_{\ell}$, unidirectional CH-Potentials is already optimal and only traverses the shortest path.
@@
 The achieved speedups are therefore comparable to CH speedups.
-Since turn costs and restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speed-ups.
+Since turn costs and restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speedups.
 Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.
@@
 The same is the case for forbidden tunnels or highways.
-Applying CCH-Potentials customized to the current traffic (or the metric with forbidden edges) to these problems yields running times similar to the unmodified case and speed-ups of more than three orders of magnitude.
+Applying CCH-Potentials customized to the current traffic (or the metric with forbidden edges) to these problems yields running times similar to the unmodified case and speedups of more than three orders of magnitude.
 With bidirectional CCH-Potentials, one can easily additionally support turn costs and still have running times of a few milliseconds without any adjustments to the customization.

### 2022-08-02 paper/ch_potentials.tex
polishing

@@
 We study exact, efficient and practical algorithms for route planning applications in large road networks.
-Routing applications often require integrating the current traffic situation, planning ahead with traffic predictions for the future, respecting forbidden turns, and many other features depending on the specific application.
-While Dijkstra's algorithm can be used to solve these problems, it is too slow for many applications.
-A* is a classical approach to accelerate Dijkstra's algorithm.
-A* can support many extended scenarios without much additional implementation complexity.
-However, A*'s performance depends on the availability of a good heuristic that estimates distances.
-Computing tight distance estimates is a challenge on its own.
-On road networks, shortest paths can also be quickly computed using hierarchical speedup techniques.
-They achieve speed and exactness but sacrifice A*'s flexibility.
-Extending them to certain practical applications can be hard.
-In this paper, we present Lazy RPHAST, an algorithm to efficiently compute distances from many vertices towards a single target and use it as an A* heuristic.
-As it uses Contraction Hierarchies (CH), a hierarchical technique, we call our A* heuristic CH-Potentials.
-Our approach allows decoupling the supported applications from the hierarchical speedup technique.
-Additionally, we describe A* optimizations to accelerate the processing of low degree vertices, which often occur in road networks and present a new pruning criterion for symmetrical bidirectional A*.
-An extensive experimental study confirms the practical applicability of our approach for many applications.
+On the one hand, such algorithms should be able to answer shortest path queries within milliseconds.
+On the other hand, routing applications often require integrating the current traffic situation, planning ahead with predictions for future traffic, respecting forbidden turns, and many other features depending on the specific application.
+Therefore, such algorithms must be flexible and able to support a variety of problem variants.
+% There exists a lot of research focused on techniques to achieve the fastest possible query times.
+% Further, there are a lot of results on extending certain techniques to specific scenarios and achieving fast queries in that specific extended problem.
+% However, there is little research focused on designing extensible algorithms which can be adapted to new problem variants with minimal effort.
+In this work, we revisit the A* algorithm to build a simple, extensible and unified algorithmic framework applicable to a variety of route planning problems.
+% without adjustments to the preprocessing.
+A* has been previously used for routing in road networks.
+However, its performance was not competitive because no sufficiently fast and tight distance estimation function was available.
+We present a novel, efficient and accurate A* heuristic by utilizing Contraction Hierarchies (CH), another popular speedup technique.
+The core of our heuristic is a new CH query algorithm called Lazy RPHAST which can efficiently compute shortest distances from many incrementally provided sources towards a common target.
+Additionally, we describe A* optimizations to accelerate the processing of low-degree vertices, common in road networks, and present a new pruning criterion for symmetrical bidirectional A*.
+An extensive experimental study confirms the practicality of our approach for many applications.
 \end{abstract}
@@
 \label{sec:intro}
-The past decade has seen a plethora of research works on route planning in large road networks~\cite{bdgmpsww-rptn-16}.
+The past two decades have seen a plethora of research works on route planning in large road networks~\cite{bdgmpsww-rptn-16}.
 Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
@@
 In contrast, extending speedup techniques is vastly more complex.
-The results on extending specific speedup techniques to specific extended problem fill many research papers~\cite{dgpw-crprn-13} and sometimes entire dissertations~\cite{b-tdrpc-14}.
-% TODO cite some papers
+The results on extending specific speedup techniques to specific extended problem fill many research papers~\cite{gv-errnt-11,dgpw-crprn-13,dsw-cch-15,bwzz-cchtc-20,bgsv-mtdtt-13,swz-sfert-21} and sometimes entire dissertations~\cite{Delling2009_1000011046,b-tdrpc-14,Baum2018_1000082225}.
 We will present a brief overview in Section~\ref{sec:related_work}.
@@
 The resulting techniques are complex to implement and hard to further extend and adjust.
-For example, while there are efficient speedup techniques for each of the above mentioned features, we are not aware of any work supporting the combination of these features.
-We believe that for \emph{practical} algorithms, a different trade-off is necessary.
+For example, while there are speedup techniques achieving fast queries for each of the above mentioned features, we are not aware of any work supporting the \emph{combination} of these features.
+We believe that for practical algorithms, a different trade-off is necessary.
 Therefore, in this work, we prioritize a simple and extensible approach over the fastest possible query times.
@@
 
-In the past two decades, a considerable amount of research effort has been put into accelerating shortest path computations in transportation networks.
+A considerable amount of research effort has been put into accelerating shortest path computations in transportation networks.
 For a comprehensive overview, we refer to~\cite{bdgmpsww-rptn-16}.
@@
 The performance of A* crucially depends on the quality of the heuristic estimates.
-A natural but rather ineffective heuristic can be obtained through euclidian distances.
-On road networks, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
+For example, while geographic distances may seem like a natural choice for distance estimates in road networks, the resulting heuristic is rather ineffective.
+In fact, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
 A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
@@
 The ``importance'' of vertices is determined heuristically during preprocessing.
-This results in queries taking less than a tenth of a millisecond which is a speedup of four or more orders of magnitude over Dijkstra's algorithm.
-The preprocessing takes only a few minutes.
+This results in queries taking less than a tenth of a millisecond which is a speedup of more than four orders of magnitude over Dijkstra's algorithm.
+The preprocessing takes only a few minutes and produces little more auxiliary data than the input graph.
 CH has been used successfully in many real world applications and quite a few open-source implementations exist\footnote{
@@
 More specifically, we extend on PHAST~\cite{dgnw-phast-13} and RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
-These algorithms reuse the CH preprocessing but modify the query phase to efficiently support \emph{one-to-all} queries from one source vertex to \emph{all} other vertices (PHAST), or \emph{one-to-many} queries to a subset of vertices known in advance (RPHAST).
+These algorithms reuse the CH preprocessing but modify the query phase.
+PHAST supports \emph{one-to-all} queries from one source vertex to \emph{all} other vertices.
+RPHAST is a variant for \emph{one-to-many} queries to a subset of vertices known in advance.
 %
-Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
+Another popular hierarchical speedup technique also used in practice~\cite{bingblog} is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02}.
 It is based on a multi-level partition of the road network computed during preprocessing.
 Queries are slightly slower than CH queries but still three to four orders of magnitude faster than Dijkstra's algorithm.
+The memory consumption is even smaller than with CH.
 Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
@@
 
-The speedup technique known to currently achieve the fastest queries is Hub Labels (HL)~\cite{TODO}.
-With HL, query times in less than a microseconds are possible.
+The speedup technique known to currently achieve the fastest queries is Hub Labels (HL)~\cite{DBLP:conf/esa/AbrahamDGW12,DBLP:conf/wea/DellingGW13}.
+With HL, query times in less than a microsecond are possible.
 This is only a couple of times the time of an uncached memory access.
@@
 In~\cite{dw-lbrdg-07}, ALT is used for routing on dynamic and time-dependent graphs.
-This approach is conceptually similar to ours.
+The approach is conceptually similar to ours.
 However, because the heuristic is ALT-based, the accuracy of the estimates is limited and the resulting query performance not competitive.
-Further, the evaluation was performed only with synthetic traffic data.
+Further, the evaluation was performed with only synthetic traffic data.
 With production-grade real-world traffic data, the problem becomes significantly harder~\cite{swz-sfert-21}.
@@
 CRP is one of the few examples of a speedup technique designed for flexibility rather than maximum query performance.
-However, its flexibility is also limited.
-Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
-Achieving reasonable memory consumption and query times was only possible by giving up exactness.
-Further, TD-CRP can only compute approximate shortest distances (rather than paths).
+However, its flexibility also has limits.
+For example, integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
+However, achieving reasonable memory consumption and query times was only possible by giving up exactness.
+Further, TD-CRP can only compute approximate shortest distances rather than paths.
 
@@
 Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
-Even with all improvements proposed in~\cite{bwzz-cchtc-20}, turn integration still causes a slowdown of at least factor three to the customization phase.
+However, even with all improvements proposed in~\cite{bwzz-cchtc-20}, turn integration still causes a slowdown of at least factor three to the customization phase.
 A considerable amount of research and engineering effort has been put into studying the combination of traffic predictions with CH.
@@
 Different variants with trade-offs regarding exactness, query speed and space consumption were proposed~\cite{bgsv-mtdtt-13}.
-Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.
-% TODO argue what is complicated to implement?
+Recently, a new approach has been published~\cite{swz-sfert-21} which simultaneously achieves competitive results in all three aspects but only at the cost of considerable implementation complexity.\footnote{In \url{https://github.com/kit-algo/rust_road_router}, the preprocessing alone has more than 10\,000 lines of code.
+In contrast, the classical CH can typically be realized within a few 100 lines of code.}
 
@@
 
-Clearly, there is a plethora of efficient speedup techniques for routing in road networks.
-Further, for many extended problems there are dedicated research results on extensions of some techniques.
-While many problems can be solved very efficiently, we still believe that this situation is not satisfactory in practice.
-For practical applications, a simple, unified, and extensible approach is often more important than the fastest query performance.
-Therefore, in this work, we revisit the A* algorithm and build a flexible, unified framework for a multitude of routing problems.
+Clearly, there is an enormous amount of efficient speedup techniques for routing in road networks.
+Further, for many extended problems there are dedicated research results on extensions of certain techniques.
+While many problems can be solved quite efficiently, we still believe that this situation is not satisfactory in practice.
+For practical applications, a unified and extensible approach with manageable implementation complexity is often more important than the fastest query performance.
+Therefore, we revisit the A* algorithm and build a flexible, unified framework for a multitude of routing problems.
 The core of our approach is a new CH-based A* heuristic.
@@
 It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
-However, the strength of our approach is not in query running times but in its flexibility:
+However, the strength of our approach is not primarily in query running times but in its flexibility:
 Problem extensions can be supported without complex algorithmic adjustments.
@@
 
-The augmented graph has the property that for any two vertices $s$ and $t$, there always exists an \emph{up-down} $st$-path, i.e.\ a path  which first uses only edges from $\gchu$ and then only edges from $\gchd$, with the same length as a shortest path in $G$.
+The augmented graph has the property that for any two vertices $s$ and $t$, there always exists an \emph{up-down} $st$-path, i.e.\ a path  which first uses only edges from $\echu$ and then only edges from $\echd$, with the same length as a shortest path in $G$.
 See Figure~\ref{fig:ch} for an illustration.
@@
 
-Some hierarchical techniques were extended to support a three-phase setup~\cite{dgpw-crprn-13}.
-The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and fast queries.
-The preprocessing phase typically takes hours, the customization seconds, and the query milliseconds.
+Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) is a variant of Contraction Hierarchies extended to a three-phase setup.
+The phases consist of a slow preprocessing phase, a faster \emph{customization} phase, and the accelerated \emph{query} phase.
+The preprocessing phase typically takes hours, the customization seconds, and the query fractions of a millisecond.
 The preprocessing phase is independent of the weights and only uses the graph topology.
 Weights are introduced in the customization phase.
-Paths are computed in the query phase.
-
-Customizable Contraction Hierarchies~\cite{dsw-cch-15} (CCH) is a variant of Contraction Hierarchies that follows this three-phase setup.
-The key idea is to use a vertex importance order that is independent of the weight function.
+Shortest paths are computed in the query phase.
+
+The key idea for CCH is to use a vertex importance order that is independent of the weight function.
 CCH use \emph{nested dissection} orders for this.
@@
 However, discussing them in detail is beyond the scope of this work.
-See~\cite{dsw-cch-15,BuchholdSW19} for details.
+See~\cite{dsw-cch-15,DBLP:journals/jea/Buchhold0W19} for details.
 A CCH augmented graph fulfills all necessary properties for CH query algorithms.
@@
 The root of the elimination tree is the most important vertex.
-For every other vertex its parent is its least important upward neighbor.
+For every other vertex, its parent is its least important upward neighbor.
 In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to its CH search space.
@@
 Note that only nested dissection orderings admit an elimination tree of low height.
-The elimination tree query can therefore not be used with a classical CH.
+Thus, the elimination tree query can not be used with classical CH.
 
@@
 The first step is analogue to the CH query:
-From $t$, all reachable vertices via reversed down-edges are explored, as shown in Algorithm~\ref{algo:ch-backward}.
+From $t$, all reachable vertices via reversed down-edges are explored as shown in Algorithm~\ref{algo:ch-backward}.
 For the second step, PHAST utilizes an assignment of vertices into \emph{levels}.
@@
 The main work of the second step consists in iterating over all CH levels from top to bottom.
-In each iteration, all up-edges starting within the current level are followed in reverse.
+In each iteration, all up-edges starting within the current level are relaxed in reverse.
 After all levels are processed, the shortest distances from all vertices to $t$ were computed.
@@
 
-RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source vertices to one target vertex (again, the reversed case works analogously), solving the \emph{many-to-one} problem.
+RPHAST~\cite{delling_et_al:OASIcs:2011:3266}, short for Restricted PHAST, is a PHAST extension for efficiently computing distances from a smaller set of source vertices to one target vertex (again, the reverse case works analogously), solving the \emph{many-to-one} problem.
 Given a set of source vertices $S$, the first step is to copy the combined search space of all sources into a \emph{restricted subgraph}.
@@
 Since the target selection time is included, computing the distances to all vertices with Dijkstra or PHAST is too slow.
-Also, since the source set is only provided incrementally, RPHAST in its basic form is not well suited to our problem.
+Also, since the source set is provided incrementally, RPHAST in its basic form is not well suited to our problem.
 Fortunately, we can do better.
 
-\subsection{Lazy RPHAST for Contraction Hierarchies}
+\subsection{Lazy RPHAST on Contraction Hierarchies}
 
@@
 Execute Algorithm~\ref{algo:ch-backward}, filling $D^{\downarrow}$\;
-$D[v] \gets \bot$ for all $v\neq t$\;
+$D[v] \gets \bot$ for all $v \in V$\;
 
@@
 
-\subsection{Lazy RPHAST for Customizable Contraction Hierarchies}
+\subsection{Lazy RPHAST on Customizable Contraction Hierarchies}
 
@@
 \KwData{$D[v]$: memoized distance from any vertex $v \in V$ to $t$, shared between invocations, initialized to $\bot$ during the selection}
-\KwData{$P[v]$: parent of a vertex $v \in V$ in the elimination tree, a parent of $\bot$ indicates a root vertex}
+\KwData{$P[v]$: parent of a vertex $v \in V$ in the elimination tree, a parent of $\bot$ indicates the root vertex}
 \KwData{$S$: stack with vertices to compute distances, empty initially, only used to store intermediate data}
@@
 For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
-To compute missing distances, the elimination tree ancestors which also do not have a distance yet are collected and pushed onto a stack $S$.
+To compute missing distances, the elimination tree ancestors which also do not have a distance yet are collected onto a stack $S$.
 The elimination tree can be traversed through parent pointers $P[v]$.
@@
 Traversing this chain by successively pushing each vertex into the queue, computing the heuristic and popping the vertex again from the queue appears quite wasteful.
-Therefore, in this section, we discuss techniques to avoid pushing such low degree vertices into the queue and rather process them consecutively.
+Therefore, we now discuss techniques to process such vertices consecutively without the queue.
 The techniques discussed here are essentially a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
@@
 Recall that $|N(u)|$ is the number of neighbors $v$ such that $vu \in E$ or $uv \in E$.
-Our algorithm differs from A* when removing a vertex $u$ from the queue.
-A* iterates over the outgoing edges $(uv)$ of $u$ and tries to reduce $D[v]$ by relaxing $(uv)$.
+Our algorithm differs from classical A* when removing a vertex $u$ from the queue.
+A* iterates over the outgoing edges $uv$ of $u$ and tries to reduce $D[v]$ by relaxing $uv$.
 If A* succeeds, $v$'s weight in the queue is set to $D[v]+h_t(v)$.
@@
 Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
-If $y$ respectively $b$ are not reached, our algorithm does nothing.
+If neither $y$ nor $b$ are not reached, no queue operation is performed.
 
@@
 The query phase is divided into two steps.
-In the first step, we apply A* with CH-Potentials to $G_C$ combined with the component that contains $s$.
-This can be achieved implicitly by removing edges from $G_C$ into other components during preprocessing.
+In the first step, we apply A* to $G_C$ combined with the component that contains $s$.
+This can be achieved implicitly by removing edges from $G_C$ into other components.
 If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
@@
 However, it can serve as basis for further algorithmic refinements.
-The authors of~\cite{gh-cspas-05} refer to this approach as \emph{symmetric} bidirectional A*.
+The authors of~\cite{gh-cspas-05} refer to this as \emph{symmetric} bidirectional A*.
 
@@
 $h_{\overline{st}}(v)$ is a worse estimate for $\dist(v,t)$ than $\overrightarrow{h}_t$.
-Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$\footnote{
+Similarly, $-h_{\overline{st}}(v)$ is a worse estimate for $\dist(s,v)$ than $\overleftarrow{h}_s$.\footnote{
 To obtain an actual lower bound from this average heuristic, one has to add $\overleftarrow{h}_s(t)/2$ in the forward case and $\overrightarrow{h}_t(s)/2$ in the backward.
 Adding any constant to a heuristic function does not change the reduced graph.
-}.
+}
 The second downside can be partially mitigated through pruning with the composing heuristics.
@@
 In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
-This sounds sensible, but our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
+While this may sound sensible, our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
 
@@
 The core idea is to compute a CH augmented graph during preprocessing and use A* with a straightforward application of Lazy RPHAST as the heuristic to answer queries.
-However, when the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function, this yields a \emph{perfect} heuristic
-This case is, of course, not particularly interesting.
+When the CH preprocessing and the A* algorithm are performed on the same graph with the same weight function, this yields a \emph{perfect} heuristic.
+However, this case is, of course, not particularly interesting.
 One could just answer the shortest path query directly with a CH query.
@@
 
-\subsection{Formal Setup: Inputs, Outputs, and Phases}\label{sec:formal_framework}
+\subsection{Formal Problem Setup: Inputs, Outputs, and Phases}\label{sec:formal_framework}
 
@@
 Let $\phi : V_q \to V_{\ell}$ denote a function mapping vertices in the query graph to vertices in the lower bound graph.
-The crucial requirement for a routing problem to fit into our framework is that the query weight of an edge $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ between the corresponding vertices $\phi(u)$ and $\phi(v)$ in the lower bound graph $G_\ell$ with respect to $w_\ell$.
-Typically, and unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
+The only requirement for a routing problem to fit into our problem framework is that the query weight of an edge $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ between the corresponding vertices $\phi(u)$ and $\phi(v)$ in the lower bound graph $G_\ell$ with respect to $w_\ell$.
+Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
 
 \subsection{CH-Potentials}\label{sec:chpot}
-CH-Potentials can be used to solve any problem in this framework.
-The preprocessing is always the computation of the CH augmented graph and remains the same regardless of the specific routing problem.
+CH-Potentials can be used to solve any problem in this setup.
+The preprocessing is always the computation of the CH augmented graph $G^+_{\ell}$ and remains the same regardless of the specific routing problem.
 The query consists of A* with the heuristic function $h_t(v) = \dist_{w_{\ell}}(\phi(v), \phi(t))$ computed using Lazy RPHAST.
 In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
-The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $v$ (see Algorithm~\ref{algo:pot}).
+The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $\phi(v)$ (see Algorithm~\ref{algo:pot}).
 Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
@@
 A sequence of expanded vertices in the expanded graph $G_q$ corresponds to a sequence of edges in the input graph $G_\ell$.
-The weight of a path in $G_q$ is equal to the turn-aware weight of the corresponding path in $G_\ell$ minus the irrelevant $w_\ell(v_1,v_2)$ term.
+The weight of a path in $G_q$ is equal to the turn-aware weight of the corresponding path in $G_\ell$ minus the irrelevant $w_\ell(v_1 v_2)$ term.
 Thus, the turn-aware routing problem can be solved by searching for shortest paths in $G_q$.
@@
 The necessary modification to A* is analogous.
-Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89}.
+Without the FIFO property the problem becomes NP-hard \cite{or-tnp-89,z-nphsp-22}.
 % The FIFO property states that it is never beneficial to wait at a vertex before entering an edge.
@@
 Live traffic can alternatively be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
-For this, we use CCH instead of CH preprocessing to build the augmented graph.
-This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization algorithm.
+For this, we use the CCH instead of the CH preprocessing to build the augmented graph.
+This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization phase.
 This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic} where live traffic was represented only in $w_q$.
@@
 
-\begin{table}
-\centering
-\caption{
-Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials.
-With CCH-Potentials, $w_{\ell}$ can be updated by rerunning Phase 2 again.
-}\label{tab:graphs}
-\input{table/graphs.tex}
-\end{table}
-
 In this section, we present our experimental evaluation.
@@
 All experiments were performed sequentially.
-Our code is written in Rust and compiled with rustc 1.57.0-nightly in the release profile with the target-cpu=native option.
+Our code is written in Rust and compiled with \texttt{rustc 1.57.0-nightly} in the release profile with the \texttt{target-cpu=native} option.
 The source code of our implementation and the experimental evaluation can be found on Github\footnote{\url{https://github.com/kit-algo/ch_potentials}}.
@@
 \subsection{Inputs and Methodology}
-Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}.
-To obtain the routing graph, we use the import from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
+Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map.\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}
+To obtain the routing graph, we use the import from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
 The graph has 16M vertices and 35M edges.
-For this instance, we have proprietary traffic data provided by Mapbox\footnote{\url{https://mapbox.com}}.
+For this instance, we have proprietary traffic data provided by Mapbox.\footnote{\url{https://mapbox.com}}
 The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the vertices.
@@
 Table~\ref{tab:graphs} contains an overview over the instances.
-In this table, we further include the sequential preprocessing running time to construct the augmented graphs and the space consumption.
-We use the CH preprocessing from RoutingKit\footnote{\url{https://github.com/RoutingKit/RoutingKit}}.
+In this table, we further report the sequential preprocessing running time to construct the augmented graphs and the space consumption.
+We use the CH preprocessing from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
 For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
@@
 To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are vertices drawn uniformly at random and report average results.
+
+\begin{table}
+\centering
+\caption{
+Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials.
+With CCH-Potentials, $w_{\ell}$ can be updated by rerunning Phase 2 again.
+}\label{tab:graphs}
+\input{table/graphs.tex}
+\end{table}
 
@@
 This heavily emphasizes memory bandwidth while neglecting other critical factors such as CPU frequency and cache size and speed which are likely much more important for algorithms carefully tuned to exploit data locality.
-} we can still conclude that Lazy RPHAST is a very useful extension to RPHAST.
+} we can still conclude that Lazy RPHAST is a useful extension to RPHAST.
 It allows efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
@@
 Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.
-Applying bidirectional CH-Potentials in this case is actually detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
+Applying bidirectional CH-Potentials in this case is, in fact, detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
 The same is the case for forbidden tunnels or highways.
@@
 
-\subsubsection{Comparison with Related Work}
+\paragraph{Comparison with Related Work}
 
 While the query running times reported in Table~\ref{tab:applications} are decent in many settings, they are not competitive to techniques tailored to specific applications.
+In the simple $w_q = w_{\ell}$ setting, Hub Labels can be used to answer queries in less than a microsecond~\cite{DBLP:conf/wea/DellingGW13}, more than three orders of magnitude faster than with CH-Potentials.
+Live traffic and arbitrary weight functions can be handled with CCH resulting in query times of around a 0.1\,ms~\cite{DBLP:journals/jea/Buchhold0W19}.
+CCH can also be adjusted to turn costs.
+With some algorithmic adjustments to the preprocessing~\cite{bwzz-cchtc-20}, query times around an order of magnitude faster than CCH-Potentials are possible (0.3\,ms compared to 2.1\,ms with CCH-Potentials).
+For time-dependent routing, CATCHUp~\cite{swz-sfert-21} achieves query times more than an order of magnitude faster than CH-Potentials (6.3\,ms compared to our 97.7\,ms).
 Again, these numbers are not perfectly comparable due to different benchmark machines.
 Nevertheless, the overall picture is clear enough.
-In the simple $w_q = w_{\ell}$ setting, Hub Labels can be used to answer queries in less than a microsecond~\cite{TODO}, more than three orders of magnitude faster than with CH-Potentials.
-Live traffic and arbitrary weight functions can be handled with CCH resulting in query times of around a 0.1\,ms~\cite{BuchholdSW19}.
-CCH can also handle turn costs~\cite{bwzz-cchtc-20}.
-With some algorithmic adjustments to the preprocessing, query times around an order of magnitude faster than CCH-Potentials are possible (0.3\,ms compared to 2.1\,ms with CCH-Potentials).
-For time-dependent routing, CATCHUp~\cite{swz-sfert-21} achieves query times more than an order of magnitude faster than CH-Potentials (6.3\,ms compared to our 97.7\,ms).
 However, the advantage of the CH-Potentials framework over these fine-tuned techniques is that it is unified and flexible approach which can handle all these applications without any adjustments to the preprocessing.
@@
 
-In this article, we introduced Lazy RPHAST, an efficient algorithm for the incremental many-to-one problem.
-We showed that it also delivers competitive performance for many-to-one problems.
-Based on Lazy RPHAST, we developed CH-Potentials, a fast, exact, and flexible two-phase technique based on A* and CH for finding shortest paths in road networks.
+In this article, we introduced CH-Potentials, a fast, exact, and flexible two-phase routing framework based on A* and CH.
 The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity and without any changes to the preprocessing algorithms.
@@
 Achieving significantly faster running times could still be possible in variations of the problem setting.
+The core building block of our approach is Lazy RPHAST, a new CH query variant for the incremental many-to-one problem.
+We showed that it also delivers competitive performance for many-to-one problems.
 This leads to multiple avenues for future research.

### 2022-08-22 paper/ch_potentials.tex
typos and wordsmithing and labels from diss

@@
 To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
-Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal directed search on.
+Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal-directed search on.
 This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
 Since then, there has been little development on A*-based techniques for routing in road networks.
-This is because hierarchical techniques have proven much more effective.
+This is because hierarchical techniques have proven more effective.
 
@@
 The preprocessing takes only a few minutes and produces little more auxiliary data than the input graph.
-CH has been used successfully in many real world applications and quite a few open-source implementations exist\footnote{
+CH has been used successfully in many real world applications and quite a few open-source implementations exist.\footnote{
 \url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
 While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
-}.
+}
 CH is another fundamental algorithm for our work.
@@
 \node (3) at (3.7,1)   [node] {};
-\node (4) at (5,3)     [node] {$v^*$};
+\node (4) at (5,3)     [node] {};
 \node (5) at (5.75,-0.1)  [node] {};
@@
 \caption{Lazy RPHAST algorithm.}
-\label{algo:pot}
+\label{algo:lazy_rphast_ch}
 \end{algorithm2e}
@@
 
-Now, distances from many sources $s_i$ to $t$ can be computed incrementally with the \textsf{ComputeAndMemoizeDist} function as shown in Algorithm~\ref{algo:pot}.
+Now, distances from many sources $s_i$ to $t$ can be computed incrementally with the \textsf{ComputeAndMemoizeDist} function as shown in Algorithm~\ref{algo:lazy_rphast_ch}.
 The key to doing this efficiently is reusing the distance information $D$ across invocations through memoization.
@@
 
-Algorithm~\ref{algo:pot} can be applied to CCH without any modifications.
+Algorithm~\ref{algo:lazy_rphast_ch} can be applied to CCH without any modifications.
 However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
 To compute the distance $D[v]$ of a vertex $v$, the distances of all upward neighbors must be final.
-In Algorithm~\ref{algo:pot}, these upward neighbor distances are computed recursively.
+In Algorithm~\ref{algo:lazy_rphast_ch}, these upward neighbor distances are computed recursively.
 Thus, the search space is explored in a DFS-like fashion and distances are finalized in DFS post order.
@@
 
-% The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:pot}.
+% The core of Lazy RPHAST are the Algorithms~\ref{algo:ch-backward} and~\ref{algo:lazy_rphast_ch}.
 % We describe vectorized variants of these algorithms.
@@
 % The vectorized version is presented in Algorithm~\ref{algo:simd-ch-backward}.
-% Algorithm~\ref{algo:pot} can be vectorized by just replacing the arrays by their SIMD variants.
+% Algorithm~\ref{algo:lazy_rphast_ch} can be vectorized by just replacing the arrays by their SIMD variants.
 % The result is illustrated in Algorithm~\ref{algo:simd-pot}.
@@
 % These buckets are populated by running a backward CH search (Algorithm~\ref{algo:ch-backward}) from each $t_i$ and inserting the resulting distances into the buckets.
-% Now Algorithm~\ref{algo:pot} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
+% Now Algorithm~\ref{algo:lazy_rphast_ch} (or~ Algorithm~\ref{algo:cch_pot}) can compute distances from a given source to \emph{all} targets at once.
 % We only have to change $D[x]$ so that it is an array of distances to each $t_i$ instead of a single scalar value.
@@
 In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
-The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $\phi(v)$ (see Algorithm~\ref{algo:pot}).
+The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $\phi(v)$ (see Algorithm~\ref{algo:lazy_rphast_ch}).
 Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
@@
 For $\alpha > 1.1$, the growth slows down.
-This illustrates both the strengths and limits of our approach and goal directed search in general.
+This illustrates both the strengths and limits of our approach and goal-directed search in general.
 CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.
@@
 Table~\ref{tab:building_blocks} depicts the performance of unidirectional A* with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
-The factor 1.05 was chosen to resemble realistic problem settings where goal directed search can achieve reasonable speedups (compare to Table~\ref{tab:applications}).
+The factor 1.05 was chosen to resemble realistic problem settings where goal-directed search can achieve reasonable speedups (compare to Table~\ref{tab:applications}).
 We compare (C)CH-Potentials to three other heuristics.

### 2022-08-30 paper/ch_potentials.tex
wordsmithing

@@
 However, no edge must connect two vertices within the same level.
-Such a level assignment can be obtained by assigning all vertices without lower ranked neighbors to the lowest level.
+Such a level assignment can be obtained by assigning all vertices without lower-ranked neighbors to the lowest level.
 All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.
@@
 The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
-In the applications considered in this paper, $w_\ell$ is always the \emph{freeflow} travel time.
+In the applications considered in this paper, $w_\ell$ is always the \emph{free-flow} travel time.
 
@@
 Implementing this feature with CH-Potentials is easy.
-We set $w_\ell$ to the freeflow travel time.
+We set $w_\ell$ to the free-flow travel time.
 If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
-Otherwise, $w_q$ is set to the freeflow travel time.
+Otherwise, $w_q$ is set to the free-flow travel time.
 
@@
 
-To support only live traffic, we set $w_\ell$ to the freeflow travel time.
+To support only live traffic, we set $w_\ell$ to the free-flow travel time.
 $w_q$ is set to the travel time accounting for current traffic.
@@
 Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
-Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
+Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$.
 If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
@@
 In our implementation, we do not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
-We set $w_\ell$ to the freeflow travel time.
+We set $w_\ell$ to the free-flow travel time.
 
@@
 It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
-We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
+We exclude speed values which are faster than the free-flow speed computed by RoutingKit.
 We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACS implementation challenge~\cite{DemetrescuGJ09}.
@@
 We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
-The lower bound $w_\ell$ is set to the freeflow travel time.
+The lower bound $w_\ell$ is set to the free-flow travel time.
 The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.

### 2022-08-30 paper/ch_potentials.tex
wordsmithing and merge bens feedback

@@
 Some applications may have additional application-specific requirements.
-Extending Dijkstra's algorithm to support these features is usually easy.
+Further, it is not sufficient to support each of these additional features individually.
+We need to support their combination.
+
+Extending Dijkstra's algorithm to support these features is comparatively easy.
 In contrast, extending speedup techniques is vastly more complex.
@@
 For example, while there are speedup techniques achieving fast queries for each of the above mentioned features, we are not aware of any work supporting the \emph{combination} of these features.
-We believe that for practical algorithms, a different trade-off is necessary.
+We believe that for practical applications, a different trade-off between running time and extensibility is necessary.
 Therefore, in this work, we prioritize a simple and extensible approach over the fastest possible query times.
@@
 Further, for many extended problems there are dedicated research results on extensions of certain techniques.
-While many problems can be solved quite efficiently, we still believe that this situation is not satisfactory in practice.
+While many problems can be solved quite efficiently, we still believe that this situation is not satisfactory.
 For practical applications, a unified and extensible approach with manageable implementation complexity is often more important than the fastest query performance.
-Therefore, we revisit the A* algorithm and build a flexible, unified framework for a multitude of routing problems.
+Therefore, we revisit the A* algorithm and propose a flexible, unified framework for a multitude of routing problems.
+It can be applied to any extended routing problem where tight lower bounds are available at preprocessing time.
 The core of our approach is a new CH-based A* heuristic.
 It allows for much tighter estimates than previous A* heuristics and thus significantly faster queries.
-While the query running times of our approach are for the most part not competitive with approaches significantly tailored to specific problems, they are still sufficient for practical applications.
+While the query running times of our approach are not competitive with approaches tailored to specific problems, they are still sufficient for practical applications.
+
 The remainder of this work is organized as follows:
-
 We discuss notation and fundamental algorithms in Section~\ref{sec:preliminaries}.
@@
 It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
-However, the strength of our approach is not primarily in query running times but in its flexibility:
-Problem extensions can be supported without complex algorithmic adjustments.
-The price for each extension is a certain query slowdown depending on how tight the lower bounds used during preprocessing remain.
+However, the strength of our approach is not in query running times but in its flexibility:
+Problem extensions can be supported without adjustments to the preprocessing.
+The price for each extension is a query slowdown depending on how tight the lower bounds used during preprocessing remain.
 
@@
 The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
-This results in surprisingly good CH orders.
+This results in orders that can be used as CH orders.
 Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
-Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19} in less than an hour even for continental-sized road networks.
-Once the order was computed, the augmented graph $G^+$ is computed.
+Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19,hs-gbpo-18} in less than an hour even for continental-sized road networks.
+After obtaining the order, the augmented graph $G^+$ is computed.
 The standard CH preprocessing algorithms could be used for this but specialized, significantly faster CCH variants exist.
@@
 CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
-However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out the faster variant.
+However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out to be the faster variant.
 Note that only nested dissection orderings admit an elimination tree of low height.
@@
 Let $V_{\operatorname{restr}}$ be the set of vertices reachable in $\gchu$ from any $s_i \in S$.
-The restricted subgraph $\gchu_{\operatorname{restr}}$ is subgraph of $\gchu$ induced by $V_{\operatorname{rest}}$.
-Finding and copying the relevant edges into this restricted subgraph makes up the \emph{selection} step.
+The restricted subgraph $\gchu_{\operatorname{restr}}$ is a subgraph of $\gchu$ induced by $V_{\operatorname{rest}}$.
+Finding and copying the relevant edges into this restricted subgraph is called the \emph{selection} step.
 In the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
@@
 In this section, we discuss a variant of the many-to-one shortest path problem which naturally arises from A* heuristics.
-Here, the target vertex $t$ is known in the selection step, but source vertices $s_1,\dots,s_k$ are queried one after another.
+Here, the target vertex $t$ is known in the selection step, but the source vertices $s_1,\dots,s_k$ are queried one after another.
 We denote this problem as the \emph{incremental many-to-one} problem.
@@
 Further, the distances $D[v]$ are initialized to a sentinel value $\bot$ where $\bot$ is some special value distinct from any other valid distance (including $\infty$).
-This value indicates that the distance to from $v$ to $t$ has not yet been computed.
+This value indicates that the distance from $v$ to $t$ has not yet been computed.
 
@@
 Thus, the search space is explored in a DFS-like fashion and distances are finalized in DFS post order.
-However, the path from a vertex to the root in the elimination tree is a also topological order for the upward search space.
+However, the path from a vertex to the root in the elimination tree also is a topological order for the upward search space.
 This is because the ancestors in the elimination tree contain the entire upward search space~\cite{bcrw-s-16}.
@@
 
-This allows for the more efficient implementation depicted in Algorithm~\ref{algo:cch_pot}.
+Using the elimination tree, we obtain the algorithm in Algorithm~\ref{algo:cch_pot}.
 For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
-To compute missing distances, the elimination tree ancestors which also do not have a distance yet are collected onto a stack $S$.
-The elimination tree can be traversed through parent pointers $P[v]$.
-As soon as a vertex with final distance is encountered, all remaining ancestors are known to have their final distance.
-Now, the vertices are successively popped from the stack, i.e.\ the algorithm walks the elimination tree path back down and relaxes outgoing upward edges.
-This yields the shortest distances for all the vertices on the path.
+Otherwise, the algorithm follows the elimination tree upward until a node with a finalized distance in encountered.
+The elimination tree is represented as parent pointer array $P$.
+The visited nodes are pushed onto a stack $S$.
+This enables the algorithm to enumerate the vertices in revered order by popping them from the stack.
+While popping the nodes, all outgoing upward edges are relaxed.
+This finalizes the shortest distances for all vertices on the elimination tree path, including the desired query vertex $v$.
 
@@
 Preliminary experiments showed that a significant amount of query running time is spent in heuristic evaluations and queue operations.
-We can reduce both by keeping some vertices out of the queue, as the heuristic needs to be evaluated when a vertex is pushed into the queue.
-Consider for example a chain of vertices with exactly two neighbors.
-Traversing this chain by successively pushing each vertex into the queue, computing the heuristic and popping the vertex again from the queue appears quite wasteful.
-Therefore, we now discuss techniques to process such vertices consecutively without the queue.
+We can reduce both by keeping some vertices out of the queue, as the heuristic only needs to be evaluated when a vertex is pushed into the queue.
+Consider for example a chain of vertices with exactly two neighbours.
+Traversing this chain by successively pushing each vertex into the queue, evaluating the heuristic and popping the vertex again from the queue appears quite wasteful.
+Therefore, we now explore techniques to process such vertices consecutively without using the queue.
 The techniques discussed here are essentially a lazy variant of the ideas used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}.
@@
 
-Recall that $|N(u)|$ is the number of neighbors $v$ such that $vu \in E$ or $uv \in E$.
+Recall that $|N(u)|$ is the number of neighbours $v$ such that $vu \in E$ or $uv \in E$.
 Our algorithm differs from classical A* when removing a vertex $u$ from the queue.
@@
 Instead, it iteratively tries to reduce all $D[v_i]$.
+It stops, if a $D[v_i]$ cannot be reduced.
 If it does not reach $w$, then only $D$ is modified, but no queue action is performed.
@@
 If neither $y$ nor $b$ are not reached, no queue operation is performed.
+Using this method, we avoid pushing every other degree three vertex into the queue.
 
@@
 $G_C$ is induced by the largest biconnected component of the undirected graph underlying $G$.
-We do this using Tarjan's algorithm \cite{t-dfslg2-72}.
-For every vertex $v$ in the input graph $G$, we store the attachment vertex $a_v$ to the core.
+We compute the core using Tarjan's algorithm \cite{t-dfslg2-72}.
+For every vertex $v$ in the input graph $G$, we store an \emph{attachment vertex} $a_v$ to the core.
 For vertices in the core, $a_v=v$.
-We exploit that all attachment vertices are single vertex separators and the problem can be decomposed along them.
+For every vertex $v$ outside of the core, the attachment vertex $a_v$, is the cut vertex in the core that separates $v$'s component from the core.
 
 The query phase is divided into two steps.
-In the first step, we apply A* to $G_C$ combined with the component that contains $s$.
-This can be achieved implicitly by removing edges from $G_C$ into other components.
+In the first step, we run A* on the subgraph induced by the core and $s$'s component.
+Formulated differently, we only consider nodes that are in the core or have the same attachment vertex as $s$.
+This can be achieved implicitly by not following edges to nodes without this property.
 If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
@@
 This breaks the usual bidirectional Dijkstra stopping criterion.
-To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for either direction.
+To the best of our knowledge, no better stopping criterion exists than running \emph{both} searches until the unidirectional stopping criterion is met for one direction.
 The forward search can skip vertices already settled by the backward search and vice versa.
-However, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
+Unfortunately, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
 Thus, on its own it is not a useful algorithm.
@@
 \]
-Thus, A* will always determine the correct shortest distances.
+Thus, A* will always determine optimal shortest distances.
 
@@
 Additional features such as turn restrictions can be implemented by adjusting $w_q$.
-For severe traffic events, the three-phase approach will be significantly faster than the one described in Section~\ref{sec:live-predicted-traffic}.
-However, it comes at the expense of a more complex setup.
-Also, the customization is not fast enough to rerun it for every query.
-Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.
+For severe traffic events, the three-phase approach will have significantly faster query running time than the one described in Section~\ref{sec:live-predicted-traffic}.
+However, it comes at the expense of a more complex setup and updating $w_\ell$ takes a few seconds.
+% Also, the customization is not fast enough to rerun it for every query.
+% Further, when integrating predicted traffic, CCH-Potentials will not have much advantage over CH-Potentials because the heuristic must be a lower bound for the entire time considered and not only the current situation.
 

### 2022-08-30 paper/ch_potentials.tex
wordsmithing

@@
 % However, there is little research focused on designing extensible algorithms which can be adapted to new problem variants with minimal effort.
-In this work, we revisit the A* algorithm to build a simple, extensible and unified algorithmic framework applicable to a variety of route planning problems.
-% without adjustments to the preprocessing.
+In this work, we revisit the A* algorithm to build a simple, extensible and unified algorithmic framework applicable to many route planning problems. % without adjustments to the preprocessing.
 A* has been previously used for routing in road networks.
 However, its performance was not competitive because no sufficiently fast and tight distance estimation function was available.
-We present a novel, efficient and accurate A* heuristic by utilizing Contraction Hierarchies (CH), another popular speedup technique.
-The core of our heuristic is a new CH query algorithm called Lazy RPHAST which can efficiently compute shortest distances from many incrementally provided sources towards a common target.
-Additionally, we describe A* optimizations to accelerate the processing of low-degree vertices, common in road networks, and present a new pruning criterion for symmetrical bidirectional A*.
+We present a novel, efficient and accurate A* heuristic using Contraction Hierarchies (CH), another popular speedup technique.
+The core of our heuristic is a new CH query algorithm called Lazy RPHAST which can efficiently compute shortest distances from many incrementally provided sources toward a common target.
+Additionally, we describe A* optimizations to accelerate the processing of low-degree vertices, typical in road networks and present a new pruning criterion for symmetrical bidirectional A*.
 An extensive experimental study confirms the practicality of our approach for many applications.
@@
 Vertices represent intersections.
-Roads are modeled using edges.
+Roads are modelled using edges.
 Edges are weighted by their traversal times.
@@
 
-However, for many real world applications, this basic model is too simplistic.
+However, for many real-world applications, this basic model is too simplistic.
 For realistic routing, many additional features need to be considered.
@@
 Some applications may have additional application-specific requirements.
-Further, it is not sufficient to support each of these additional features individually.
-We need to support their combination.
+Further, it is insufficient to support each of these features independently.
+Instead, we need to support their combination.
 
@@
 While techniques achieving fast query times have been successfully developed for many extended scenarios, we observe two problems:
-The resulting techniques are complex to implement and hard to further extend and adjust.
-For example, while there are speedup techniques achieving fast queries for each of the above mentioned features, we are not aware of any work supporting the \emph{combination} of these features.
-We believe that for practical applications, a different trade-off between running time and extensibility is necessary.
-Therefore, in this work, we prioritize a simple and extensible approach over the fastest possible query times.
+The resulting techniques are complex to implement and hard to extend further.
+For example, while there exist speedup techniques achieving fast queries for each of the features mentioned above, we are unaware of any work supporting the combination of these features.
+We believe that a different trade-off between running time and extensibility is necessary for practical applications.
+Therefore, we prioritize a simple and extensible approach over the fastest possible query times in this work.
 
@@
 For a comprehensive overview, we refer to~\cite{bdgmpsww-rptn-16}.
-Here, we first highlight a few key results for the classical shortest path problem on road networks specifically relevant to our work.
-In the second part of this section, we will discuss adaptions of these techniques to extended problem models.
+Here, we first highlight a few key results for the classical shortest path problem on road networks that are specifically relevant to our work.
+In the second part of this section, we will discuss adaptations of these techniques to extended problem models.
 
@@
 It is one of the algorithms fundamental to our work.
-The performance of A* crucially depends on the quality of the heuristic estimates.
-For example, while geographic distances may seem like a natural choice for distance estimates in road networks, the resulting heuristic is rather ineffective.
-In fact, it performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
+The performance of A* depends on the quality of the heuristic estimates.
+For example, while geographic distances may seem like a natural choice for distance estimates in road networks, the resulting heuristic is relatively ineffective: It performs worse than Dijkstra's algorithm~\cite{gh-cspas-05}.
 A much more accurate heuristic can be obtained with ALT~\cite{gh-cspas-05,gw-cppsp-05}.
-ALT was one of the early speedup technique for routing in road networks.
-It uses precomputed distances to certain landmark vertices combined with the triangle inequality to compute distance estimates.
+ALT was one of the early speedup techniques for routing in road networks.
+It uses precomputed distances to specific landmark vertices combined with the triangle inequality to compute distance estimates.
 On road networks, this achieves speedups of around two orders of magnitude over Dijkstra's algorithm.
@@
 To achieve faster queries, Core-ALT has been developed~\cite{bdsssw-chgds-10}.
-Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal-directed search on.
-This yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
+Here, parts of the network are \emph{contracted} during the preprocessing, leaving a much smaller remaining core graph to run the goal-directed search.
+Core-ALT yields speedups of about three orders of magnitude compared to Dijkstra's algorithm.
 Since then, there has been little development on A*-based techniques for routing in road networks.
-This is because hierarchical techniques have proven more effective.
+That is because hierarchical techniques have proven more effective.
 
@@
 The ``importance'' of vertices is determined heuristically during preprocessing.
-This results in queries taking less than a tenth of a millisecond which is a speedup of more than four orders of magnitude over Dijkstra's algorithm.
+CH queries take less than a tenth of a millisecond which is a speedup of more than four orders of magnitude over Dijkstra's algorithm.
 The preprocessing takes only a few minutes and produces little more auxiliary data than the input graph.
-CH has been used successfully in many real world applications and quite a few open-source implementations exist.\footnote{
+CH has been used successfully in many real-world applications, and quite a few open-source implementations exist.\footnote{
 \url{https://gist.github.com/PayasR/bc46af938195a827e42006c3f5544e4a}.
-While this list is certainly not exhaustive and possibly not representative for what algorithms are used in cooperate contexts, it still paints a fairly clear picture on which techniques have reached a certain popularity in practice.
+While this list is certainly not exhaustive and possibly not representative of what algorithms are used in practice, it still paints a relatively clear picture of which techniques have reached popularity.
 }
@@
 Hierarchical speedup techniques have been combined with goal-directed search and A* in~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
-However, these works also primarily focused on obtaining faster queries rather than simple and flexible approaches.
-
-The speedup technique known to currently achieve the fastest queries is Hub Labels (HL)~\cite{DBLP:conf/esa/AbrahamDGW12,DBLP:conf/wea/DellingGW13}.
+However, these works primarily focused on obtaining faster queries rather than simple and flexible approaches.
+
+The speedup technique known to achieve the fastest queries is Hub Labels (HL)~\cite{DBLP:conf/esa/AbrahamDGW12,DBLP:conf/wea/DellingGW13}.
 With HL, query times in less than a microsecond are possible.
-This is only a couple of times the time of an uncached memory access.
-This comes at the cost of an expensive preprocessing phase and a tremendous memory footprint, more than an order of magnitude larger than the input graph.
+This is only a few times as much as an uncached memory access.
+This speed comes at the cost of an expensive preprocessing phase and a tremendous memory footprint, more than an order of magnitude larger than the input graph.
 Despite the extremely fast query times, it appears that the trade-off offered by HL is, in practice, often not very attractive.
-Simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (network latency, etc.).
+Simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (such as the network latency).
 Also, we are not aware of any works extending HL to extended problem models.
@@
 \paragraph{Speedup Technique Adaptions for Extended Problems}
-One of the earliest approaches to adapt speedup techniques to extended problems was based on A*.
+One of the earliest approaches adapting speedup techniques to extended problems was built on A*.
 In~\cite{dw-lbrdg-07}, ALT is used for routing on dynamic and time-dependent graphs.
 The approach is conceptually similar to ours.
-However, because the heuristic is ALT-based, the accuracy of the estimates is limited and the resulting query performance not competitive.
+However, because the heuristic is ALT-based, the accuracy of the estimates is limited, and the resulting query performance is not competitive.
 Further, the evaluation was performed with only synthetic traffic data.
 With production-grade real-world traffic data, the problem becomes significantly harder~\cite{swz-sfert-21}.
-Similar to the development for the classical problem, the approach was also combined with contraction~\cite{ndls-bastd-12,dn-crdtd-12}.
-Even though this improves query running times, they obtained speedups are not competitive with purely hierarchical techniques.
+Similar to the development for the classical problem, the approach was combined with contraction~\cite{ndls-bastd-12,dn-crdtd-12}.
+Even though this improves query running times, the obtained speedups are not competitive with purely hierarchical techniques.
 
 CRP~(Customizable Route Planning)~\cite{dgpw-crprn-13} is an engineered variant of MLD~\cite{swz-umlgt-02} which was developed to allow updating weights without invalidating the entire preprocessing.
-For this, a second, faster, preprocessing phase is introduced which takes at most a few seconds.
+For this, a second, faster preprocessing phase is introduced, which takes at most a few seconds.
 This phase is called the \emph{customization}.
@@
 
-There is also lot of work that extends CH to more complex problem models.
+There are many works extending CH to more complex problem models.
 In~\cite{gv-errnt-11}, turn information is integrated into CH.
-The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower compared to classical CH.
+The proposed approach achieves fast queries, but preprocessing becomes an order of magnitude slower than classical CH.
 In~\cite{dsw-cch-15}, CH is extended to Customizable Contraction Hierarchies (CCH).
 CCH also has a second preprocessing (customization) phase where weights can be altered.
-This allows to support user preferences and live traffic with queries an order of magnitude faster than CRP.
+This allows supporting user preferences and live traffic with queries an order of magnitude faster than CRP.
 Supporting turn costs in CCH was studied in~\cite{bwzz-cchtc-20}.
@@
 Clearly, there is an enormous amount of efficient speedup techniques for routing in road networks.
-Further, for many extended problems there are dedicated research results on extensions of certain techniques.
-While many problems can be solved quite efficiently, we still believe that this situation is not satisfactory.
+Further, for many extended problems, there are dedicated research results on extensions of specific techniques.
+While many problems can be solved quite efficiently, we still believe this situation is unsatisfactory.
 For practical applications, a unified and extensible approach with manageable implementation complexity is often more important than the fastest query performance.
-Therefore, we revisit the A* algorithm and propose a flexible, unified framework for a multitude of routing problems.
+Therefore, we revisit the A* algorithm and propose a flexible, unified framework for many routing problems.
 It can be applied to any extended routing problem where tight lower bounds are available at preprocessing time.
@@
 In Section~\ref{sec:lazy-rphast}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem.
-Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
+Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low-degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
 Our main contribution, the CH-Potentials framework, is presented in Section~\ref{sec:framework}.
-There, we first show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
+We first show how to use Lazy RPHAST to build a fast and tight heuristic for A* in road networks.
 Based on this heuristic, we provide a unified framework (Section~\ref{sec:formal_framework} and~\ref{sec:chpot}) for a variety of practical route planning problems (Section~\ref{sec:applications}).
@@
 It shows that CH-Potentials achieve decent running times when tight lower bounds are available at preprocessing time.
-However, the strength of our approach is not in query running times but in its flexibility:
+However, the strength of our approach is not in query running times but its flexibility:
 Problem extensions can be supported without adjustments to the preprocessing.
@@
 
-In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce notation used throughout this paper.
+In this section, we recall the central aspects of Dijkstra's classical shortest path algorithm~\cite{d-ntpcg-59} and introduce the notation used throughout this paper.
 Dijkstra's algorithm maintains an array of tentative distances $D$ and a priority queue $Q$ of vertices ordered by increasing distance from $s$.
@@
 Initially, all distances are set to $\infty$.
-The start vertex distance $D[s]$ is set to zero and the queue is initialized with $s$.
+The start vertex distance $D[s]$ is set to zero, and the queue is initialized with $s$.
 In each iteration, the closest remaining vertex $u$ is popped from the queue and \emph{settled}.
 Outgoing edges $uv$ are relaxed, i.e. $D[v]$ is set to $\min(D[v], D[u] + w(uv))$.
-If the tentative distance at the head vertex $D[v]$ was improved, it's added to the queue or its position in the queue is adjusted.
+If the tentative distance at the head vertex $D[v]$ is improved, it is added to the queue or its position in the queue is adjusted.
 Once $t$ was settled, $D[t]$ contains the shortest path distance between $s$ and $t$.
@@
 Typically, the searches are interleaved by alternating settling a vertex from each direction.
-Another common approach is to always settle a vertex from the direction with the smaller minimum queue key.
+Another common approach is to settle a vertex from the direction with the smaller minimum queue key.
 Theoretically, any direction interleaving strategy is possible.
-When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the so far best known tentative total distance $\mu$ the algorithm can terminate.
+When the sum of the minimum keys in both queues $\overrightarrow{k} + \overleftarrow{k}$ is greater than the best-known tentative total distance $\mu$, the algorithm can terminate.
 
@@
 A* orders the vertices in the priority queue by $D[v] + h_t(v)$ instead of $D[v]$ as Dijkstra's algorithm does it.
-This results in vertices closer to the target being explored earlier.
+Thus, vertices closer to the target are explored earlier.
 Figure~\ref{img:search-space} depicts an example of the search space of the A* algorithm.
@@
 \includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
-\caption{Vertices explored by A*. Color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
+\caption{Vertices explored by A*. The color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
 \label{img:search-space}
@@
 For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
-Here, we only briefly introduce necessary notation and algorithms used in this paper.
+Here, we only briefly introduce the necessary notation and algorithms used in this paper.
 In a preprocessing phase, vertices are ordered totally by ``importance'' where more important vertices should lie on more shortest paths.
@@
 Vertices of higher rank are informally referred to and visualized as ``higher up'' in the hierarchy.
-Therefore, an edge where the tail has lower rank than the head is an \emph{upward} edge.
+Therefore, an edge where the tail has a lower rank than the head is an \emph{upward} edge.
 Analogously, when the head vertex has the lower rank, the edge is said to go \emph{downward}.
-Once such an importance ordering was obtained, all vertices are contracted successively by ascending importance.
-To \emph{contract} a vertex means to temporarily remove it from the graph while inserting \emph{shortcut edges} between more important neighbors to preserve shortest distances among them.
+Once such an importance order was obtained, all vertices are contracted successively by ascending importance.
+To \emph{contract} a vertex means temporarily removing it from the graph while inserting \emph{shortcut edges} between more important neighbours to preserve shortest distances among them.
 The result is an \emph{augmented graph} $G^+$ with original edges and shortcuts.
@@
 From every shortest path (red) in $G$, an up-down path of equal length in $G^+$ (blue) can be constructed.
-Such an up-down path can be found by running Dijkstra's algorithm from $s$ on $\gchu$ and from $t$ on the reversed downward graph $\rgchd$ graph.
-By construction, at least the highest ranked vertex on the up-down path must be in the intersection of the forward and backward search spaces.
+Such an up-down path can be found by running Dijkstra's algorithm from $s$ on $\gchu$ and $t$ on the reversed downward graph $\rgchd$.
+By construction, at least the highest-ranked vertex on the up-down path must be in the intersection of the forward and backward search spaces.
 Pseudocode for the backward search is presented in Algorithm~\ref{algo:ch-backward}.
@@
 CCH use \emph{nested dissection} orders for this.
-A small balanced vertex separator is computed and the separator vertices get the highest importance.
+A small balanced vertex separator is computed, and the separator vertices get the highest importance.
 The remaining vertices are ordered by recursing on the independent cells obtained by removing the separator vertices.
 This results in orders that can be used as CH orders.
-Regardless of the weight function, any shortest path between the cells has to use some of the separator vertices, therefore these vertices will always be ``important''.
-Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19,hs-gbpo-18} in less than an hour even for continental-sized road networks.
+Regardless of the weight function, any shortest path between cells has to use some of the separator vertices.
+Therefore these vertices will always be ``important''.
+Such orders can be obtained with partitioning algorithms tailored to road networks~\cite{GottesburenHUW19,hs-gbpo-18} in less than an hour, even for continental-sized road networks.
 After obtaining the order, the augmented graph $G^+$ is computed.
@@
 See~\cite{dsw-cch-15,DBLP:journals/jea/Buchhold0W19} for details.
-A CCH augmented graph fulfills all necessary properties for CH query algorithms.
+A CCH augmented graph fulfils all necessary properties for CH query algorithms.
 The CH query algorithms can therefore be applied without modifications.
@@
 The root of the elimination tree is the most important vertex.
-For every other vertex, its parent is its least important upward neighbor.
+For every other vertex, its parent is its least important upward neighbour.
 In~\cite{bcrw-s-16}, it was proven that the set of ancestors of a vertex in the elimination tree is equal to its CH search space.
@@
 Because this elimination tree-based algorithm does not require a queue, it is faster than a Dijkstra-based CCH query.
-With this algorithm, CCH can consistently answer shortest path queries on continental sized road networks in fractions of a millisecond.
-CH queries are still marginally faster for weight functions with a strong hierarchy such as travel times.
-However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly and CCH queries turn out to be the faster variant.
+With this algorithm, CCH can consistently answer shortest path queries on continental-sized road networks in fractions of a millisecond.
+CH queries are still marginally faster for weight functions with a strong hierarchy, such as travel times.
+However, on weight functions with a less pronounced hierarchy, CH performance degrades significantly, and CCH queries turn out to be the faster variant.
 Note that only nested dissection orderings admit an elimination tree of low height.
@@
 However, no edge must connect two vertices within the same level.
-Such a level assignment can be obtained by assigning all vertices without lower-ranked neighbors to the lowest level.
-All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.
+Such a level assignment can be obtained by assigning all vertices without lower-ranked neighbours to the lowest level.
+All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbour.
 
@@
 In each iteration, all up-edges starting within the current level are relaxed in reverse.
-After all levels are processed, the shortest distances from all vertices to $t$ were computed.
+Once all levels were processed, the shortest distances from all vertices to $t$ were computed.
 Pseudocode is provided in Algorithm~\ref{algo:phast}.
@@
 Finding and copying the relevant edges into this restricted subgraph is called the \emph{selection} step.
-In the query step, a target $t$ is given and the PHAST algorithm is applied but the downward sweep (second step of the PHAST algorithm) is performed only on the restricted subgraph.
+In the query step, a target $t$ is given, and the PHAST algorithm is applied, but the downward sweep (the second step of the PHAST algorithm) is performed only on the restricted subgraph.
 RPHAST is particularly effective when many targets are queried for the same source set $S$.
@@
 
-In this section, we discuss a variant of the many-to-one shortest path problem which naturally arises from A* heuristics.
+This section discusses a variant of the many-to-one shortest path problem, which naturally arises from A* heuristics.
 Here, the target vertex $t$ is known in the selection step, but the source vertices $s_1,\dots,s_k$ are queried one after another.
@@
 
-The core idea of our algorithm is to do the RPHAST computation lazily and using memoization.
+The core idea of our algorithm is to do the RPHAST computation lazily using memoization.
 In the target selection, we first run the backward CH search from $t$ to obtain an array $D^{\downarrow}$.
@@
 $D^{\downarrow}$ is computed as shown in Algorithm~\ref{algo:ch-backward}.
-Further, the distances $D[v]$ are initialized to a sentinel value $\bot$ where $\bot$ is some special value distinct from any other valid distance (including $\infty$).
+Further, the distances $D[v]$ are initialized to a sentinel value $\bot$ where $\bot$ is some particular value distinct from any other valid distance (including $\infty$).
 This value indicates that the distance from $v$ to $t$ has not yet been computed.
@@
 The key to doing this efficiently is reusing the distance information $D$ across invocations through memoization.
-Thus, the first step of the algorithm is always to check if the distance for the requested vertex has already been computed and immediately returning it if this is the case.
+Thus, the first step of the algorithm is always to check if the distance for the requested vertex has already been computed.
+If this is the case, it immediately returns this distance.
 If not, the distance $D[s_i]$ is initialized to the shortest down-path distance $D^{\downarrow}[s_i]$ obtained by the backward search.
-Then, the algorithm iterates over all up-edges $(s_i v)$ and checks if the up-down path through this neighbor can improve the distance.
-To obtain the shortest distance from a neighbor $v$ to $t$, the algorithm is invoked recursively.
+Then, the algorithm iterates over all up-edges $(s_i v)$ and checks if the up-down path through this neighbour can improve the distance.
+The algorithm is invoked recursively to obtain the shortest distance from a neighbour $v$ to $t$.
 
@@
 Further, $G^+$ can be decomposed into two directed acyclic graphs (DAG) $\gchu$ and $\gchd$ and the up path can be found in $\gchu$ and the down path in $\gchd$.
-Observe that the second part of Lazy RPHAST is, in fact, a recursive depth-first search (DFS) on $\gchu$ which relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
+Observe that the second part of Lazy RPHAST is, in fact, a recursive depth-first search (DFS) on $\gchu$, which relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
 Since DFS can be used to find shortest paths on DAGs, the search will find shortest up paths on $\gchu$.
-Concatenated with the down paths obtained by the backward search, this yields correct shortest distances.
+Concatenated with the down paths obtained by the backward search, this yields optimal shortest distances.
 
@@
 Algorithm~\ref{algo:lazy_rphast_ch} can be applied to CCH without any modifications.
-However, we can also exploit the elimination tree in the Lazy RPHAST algorithm.
-To compute the distance $D[v]$ of a vertex $v$, the distances of all upward neighbors must be final.
-In Algorithm~\ref{algo:lazy_rphast_ch}, these upward neighbor distances are computed recursively.
+However, we can also utilize the elimination tree in the Lazy RPHAST algorithm.
+To compute the distance $D[v]$ of a vertex $v$, the distances of all upward neighbours must be final.
+In Algorithm~\ref{algo:lazy_rphast_ch}, these upward neighbour distances are computed recursively.
 Thus, the search space is explored in a DFS-like fashion and distances are finalized in DFS post order.
@@
 Therefore, iterating over the vertices on the elimination tree path from the root to $v$ while relaxing outgoing upward edges of each vertex also yields shortest distances.
-Further, with this approach, when a vertex $v$ already has a distance $D[v] \neq \bot$, all ancestors of $v$ must already have their final distance, too.
-Thus, as soon as the algorithm encounters a vertex which already has a final distance, the remaining search space is known to have final distances.
+Further, with this approach, when a vertex $v$ already has a distance $D[v] \neq \bot$, all ancestors of $v$ must already have their final distance.
+Thus, as soon as the algorithm encounters a vertex with a final distance, the remaining search space is known to have final distances.
 
@@
 }
-\caption{Elimination tree based Lazy RPHAST algorithm.}
+\caption{Elimination tree-based Lazy RPHAST algorithm.}
 \label{algo:cch_pot}
@@
 
-Using the elimination tree, we obtain the algorithm in Algorithm~\ref{algo:cch_pot}.
+We obtain the procedure described Algorithm~\ref{algo:cch_pot}, which utilizes the elimination tree.
 For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
-Otherwise, the algorithm follows the elimination tree upward until a node with a finalized distance in encountered.
+Otherwise, the algorithm follows the elimination tree upward until a node with a finalized distance is encountered.
 The elimination tree is represented as parent pointer array $P$.
 The visited nodes are pushed onto a stack $S$.
-This enables the algorithm to enumerate the vertices in revered order by popping them from the stack.
+This enables the algorithm to enumerate the vertices in reversed order by popping them from the stack.
 While popping the nodes, all outgoing upward edges are relaxed.
@@
 In this section, we propose optimizations for A* in road networks.
-First, we present several low degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
+First, we present several low-degree vertex optimizations which exploit road network characteristics to reduce the overhead of queue operations and heuristic evaluations.
 Second, we discuss the bidirectional A* algorithm and propose an improved pruning criterion for symmetric bidirectional potentials.
@@
 
-\subsection{Low Degree A* Improvements}\label{sec:low-deg-improvment}
+\subsection{Low-Degree A* Improvements}\label{sec:low-deg-improvment}
 
@@
 We can reduce both by keeping some vertices out of the queue, as the heuristic only needs to be evaluated when a vertex is pushed into the queue.
-Consider for example a chain of vertices with exactly two neighbours.
+For example, consider a chain of vertices with precisely two neighbours.
 Traversing this chain by successively pushing each vertex into the queue, evaluating the heuristic and popping the vertex again from the queue appears quite wasteful.
@@
 
-% We modify A* by processing low degree vertices consecutively without pushing them into the queue.
+% We modify A* by processing low-degree vertices consecutively without pushing them into the queue.
 % Analogously to A*, our algorithm stores for every vertex $v$ a tentative distance $D[v]$ and maintains a minimum priority queue.
@@
 
-\subsubsection{Skip Degree Two Vertices}
+\subsubsection{Skip Degree-Two Vertices}
 
@@
 Instead, it iteratively tries to reduce all $D[v_i]$.
-It stops, if a $D[v_i]$ cannot be reduced.
+It stops if a $D[v_i]$ cannot be reduced.
 If it does not reach $w$, then only $D$ is modified, but no queue action is performed.
@@
 
-\subsubsection{Skip Degree Three Vertices}
-
-We can also skip some degree three vertices.
+\subsubsection{Skip Degree-Three Vertices}
+
+We can also skip some degree-three vertices.
 Denote by $u, v_1,\ldots, v_k, w$ a degree two chain as described in the previous section.
@@
 If neither $y$ nor $b$ are not reached, no queue operation is performed.
-Using this method, we avoid pushing every other degree three vertex into the queue.
+Using this method, we avoid pushing every other degree-three vertex into the queue.
 
@@
 
-A lot of vertices in road networks lead to dead-ends.
+Many vertices in road networks lead to dead ends.
 Unless the source or target is in this dead-end, it is unnecessary to explore these vertices.
@@
 For vertices in the core, $a_v=v$.
-For every vertex $v$ outside of the core, the attachment vertex $a_v$, is the cut vertex in the core that separates $v$'s component from the core.
+For every vertex $v$ outside of the core, the attachment vertex $a_v$ is the cut vertex in the core that separates $v$'s component from the core.
 
@@
 In the first step, we run A* on the subgraph induced by the core and $s$'s component.
-Formulated differently, we only consider nodes that are in the core or have the same attachment vertex as $s$.
+Formulated differently, we only consider nodes which are in the core or have the same attachment vertex as $s$.
 This can be achieved implicitly by not following edges to nodes without this property.
@@
 On road graphs, bidirectional search provides a simple way to halve the practical running time of Dijkstra's algorithm.
-A bidirectional variant of A* therefore also seems desirable.
+Thus, a bidirectional variant of A*  also seems desirable.
 However, as shown in~\cite{gh-cspas-05}, the necessary modifications are not straightforward.
@@
 The problem with this straightforward approach is that these two potentials induce different reduced graphs (see Section~\ref{sec:a_star}).
-Thus, in the equivalent bidirectional Dijkstra search each direction would run on a different graph.
+Thus, each direction would run on a different graph in the equivalent bidirectional Dijkstra search.
 This breaks the usual bidirectional Dijkstra stopping criterion.
@@
 Unfortunately, this straightforward bidirectional A* still performs more work than a unidirectional A*~\cite{gh-cspas-05}.
-Thus, on its own it is not a useful algorithm.
-However, it can serve as basis for further algorithmic refinements.
+Thus, on its own, it is not a useful algorithm.
+However, it can serve as a basis for further algorithmic refinements.
 The authors of~\cite{gh-cspas-05} refer to this as \emph{symmetric} bidirectional A*.
@@
 First, evaluating the average potential requires evaluating both $\overrightarrow{h}_t$ and $\overleftarrow{h}_s$.
-Evaluating the average heuristic is therefore slower than evaluating just one of the composing heuristics.
+Evaluating the average heuristic is, therefore, slower than evaluating just one of the composing heuristics.
 Second, the bidirectional stopping criterion comes at the cost of worse estimates for each direction on its own.
@@
 The stopping criteria for each direction remain the same (unidirectional) as before.
-However, usually the search stops early because the queues are empty.
+However, usually, the search stops early because the queues are empty.
 
@@
 As $u$ was settled, $\overrightarrow{D}[u]$ contains the shortest path length $\dist(s,u)$, i.e., $\dist(s,u)=\overrightarrow{D}[u]$.
-$w(uv)$ is also exactly known as it's just an edge weight.
+$w(uv)$ is also known as it is just an edge weight.
 It remains to lower bound $\dist(v,t)$.
@@
 Thus, $v$ does not have to be pushed if $\overrightarrow{D}[u]+w(uv)+\overleftarrow{k}-\overleftarrow{h}(v)\ge \mu$.
-The vertex $v$ might still be pushed into the queue when there is another edge $wv$ for which the pruning is not possible.
+The vertex $v$ might still be pushed into the queue when there is another edge $wv$ for which pruning is impossible.
 Checking the pruning rule requires evaluating the backward heuristic.
@@
 Unless stated differently, for all bidirectional A* variants, we always alternate between removing a vertex from the forward and the backward queues.
-In our experiments, we also evaluate expanding the search with the smaller minimum queue key.
+We also evaluate expanding the search with the smaller minimum queue key in our experiments.
 While this may sound sensible, our experiments show in Table~\ref{tab:bidir_switching} that it is never beneficial in terms of running time.
@@
 % , we describe how to use the Lazy RPHAST algorithm to build a fast and tight heuristic for A*.
-In this section, we introduce an algorithmic framework to apply A* with a Lazy RPHAST-based heuristic to a variety of practical route planning problems.
+In this section, we introduce an algorithmic framework to apply A* with a Lazy RPHAST-based heuristic to various practical route planning problems.
 We call our approach \emph{CH-Potentials}.
@@
 
-We consider a variety of different of applications, with slightly different problem models.
-The goal is always to quickly answer many shortest path queries.
-For the purpose of describing our framework, we establish a shared notation:
+We consider a variety of different applications with slightly different problem models.
+The goal is always to answer many shortest path queries quickly.
+To describe our framework, we establish a shared notation:
 Input to each query are vertices $s$ and $t$, and a graph $G_q$ with query weights $w_q$.
@@
 In the simplest case, $w_q$ will be scalar edge weights.
-Live traffic is an example for this.
+Live traffic is an example of this.
 The challenge in this scenario is that values of $w_q$ might change between queries.
@@
 This function can also take additional parameters from the state of the search.
-In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time and the query takes an additional departure time parameter.
+In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time, and the query takes an additional departure time parameter.
 
@@
 The input to the preprocessing phase is a lower bound graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$ where $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
-The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
+The preprocessing output is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
 In the applications considered in this paper, $w_\ell$ is always the \emph{free-flow} travel time.
@@
 The only requirement for a routing problem to fit into our problem framework is that the query weight of an edge $w_q(u v)$ is greater or equal to the shortest distance $\dist_\ell(\phi(u), \phi(v))$ between the corresponding vertices $\phi(u)$ and $\phi(v)$ in the lower bound graph $G_\ell$ with respect to $w_\ell$.
-Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function, and only $w_q$ changes for the queries.
+Unless stated otherwise, $G_q$ and $G_\ell$ are the same graph, $\phi$ is the identity function and only $w_q$ changes for the queries.
 
@@
 The query consists of A* with the heuristic function $h_t(v) = \dist_{w_{\ell}}(\phi(v), \phi(t))$ computed using Lazy RPHAST.
-In the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
+At the beginning of each query, we perform the target selection, i.e.\ a backward CH search, from the target $t$.
 The heuristic function $h_t(v)$ is implemented by a call to the \texttt{ComputeAndMemoizeDist} for vertex $\phi(v)$ (see Algorithm~\ref{algo:lazy_rphast_ch}).
-Note that, in contrast to the preprocessing, the exact implementation of the A* search depends on the application.
+In contrast to the preprocessing phase, the exact implementation of the A* search depends on the application.
 Our approach only provides the heuristic $h_t$ for the A* search.
-As the performance of A* depends crucially on the accuracy of the heuristic estimates, the smaller the difference between query weights and lower bound distances, the better CH-Potentials will perform.
+As the performance of A* depends on the accuracy of the heuristic estimates, the smaller the difference between query weights and lower bound distances, the better CH-Potentials will perform.
 
@@
 Our heuristic is always \emph{feasible}, i.e. $w_q(u v) - h_t(u) + h_t(v) \geq 0$ holds for all edges.
-By requirement and because of the triangle inequality the following holds:
+By requirement and because of the triangle inequality, the following must hold:
 \[
@@
 \]
-Thus, A* will always determine optimal shortest distances.
+Thus, A* will always determine the optimal shortest distance.
 
@@
 
-Avoiding tunnels and/or highways is a common feature of navigation devices.
+Avoiding tunnels or highways is a common feature of navigation devices.
 Implementing this feature with CH-Potentials is easy.
 We set $w_\ell$ to the free-flow travel time.
-If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
+If an edge is a tunnel or a highway, we set $w_q$ to $+\infty$.
 Otherwise, $w_q$ is set to the free-flow travel time.
@@
 
-The classical shortest path problem allows to freely change edges at vertices.
+The classical shortest path problem allows changing edges at vertices freely.
 However, in the real world, turn restrictions, such as a forbidden left or right turn, exist.
@@
 A \emph{turn weight} $w_t : E \times E \to \mathbb{R}^{\geq 0}$ maps a pair of incident edges onto the turning time or $+\infty$ for forbidden turns.
-For CH-Potentials, we use zero as lower bound for every turn weight in the heuristic.
+For CH-Potentials, we use zero as the lower bound for every turn weight in the heuristic.
 Thus, the graph $G_\ell$ and weights $w_\ell$ for preprocessing is the unmodified input graph without turn weights.
@@
 The classical shortest path problem assumes that edge weights are scalars.
-However, in practice, travel times vary along an edge due to the traffic situation.
+However, in practice, travel times vary along an edge due to traffic.
 % The primary reason is traffic.
@@
 
-If all travel time functions fulfill the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
+If all travel time functions fulfil the \emph{FIFO property}, this problem can be solved using a straightforward extension of Dijkstra's algorithm \cite{d-aassp-69}.
 The necessary modification to A* is analogous.
@@
 % The FIFO property states that it is never beneficial to wait at a vertex before entering an edge.
-The FIFO property states that it is not possible to arrive earlier by departing later.
+The FIFO property states that arriving earlier by departing later is impossible.
 Formally stated, the following must hold $\forall e\in E,\tau\in \mathbb{R},\varepsilon\in \mathbb{R}^{>0}: w_q(e, \tau) \le w_q(e, \tau+\varepsilon) + \varepsilon$.
@@
 
-Beside predicted traffic, we also consider live traffic.
+Besides predicted traffic, we also consider live traffic.
 Live traffic refers to the current traffic situation.
-It is important to distinguish between predicted and live traffic.
-Live traffic data is more accurate for the current moment than predicted data.
-It is possible that it differs significantly from predicted traffic, if unexpected events like accidents happen.
+It is essential to distinguish between predicted and live traffic.
+Live traffic data is more accurate for the current moment than predicted data and may differ significantly from predicted traffic if unexpected events like accidents happen.
 % The predicted travel time along an edge was estimated in the past.
 % Accidents are examples of such unexpected events.
-However, just using live traffic data is problematic for long routes as traffic changes while driving.
-At some point, one wants to switch from live traffic to the predicted traffic.
+However, only using live traffic data is problematic for long routes as traffic changes while driving.
+At some point, one wants to switch from live traffic to predicted traffic.
 In this section, we first describe a setup with only live traffic and then combine it with predicted traffic.
@@
 As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
-In a real world application, values from $w_q$ could be updated between queries.
+In a real-world application, values from $w_q$ could be updated between queries.
 This is all that is necessary to apply CH-Potentials in a live traffic scenario.
@@
 Further, $w_c(e)$ is the travel time according to current live traffic.
-Finally, we denote by $\tau_{\mathrm{soon}}$ the moment when we switch to predicted traffic.
+Finally, we denote by $\tau_{\mathrm{soon}}$ the moment we switch to predicted traffic.
 In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
-We need to make sure that the modified travel time function fulfills the FIFO property.
+We must ensure that the modified travel time function fulfils the FIFO property.
 For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
@@
 
-With this setup, CH-Potentials supports a combination of live and predicted traffic.
-We did not make any modification, that would hinder a combination with other extensions.
-Further adding tunnel and/or highway avoidance or turn-aware routing is simple.
+With this setup, CH-Potentials supports a combination of real-time and predicted traffic.
+We did not make any modifications that would hinder a combination with other extensions.
+Further adding tunnel or highway avoidance or turn-aware routing is simple.
 This straightforward integration of complex routing problems is the strength of CH-Potentials.
@@
 
-Live traffic can alternatively be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
-For this, we use the CCH instead of the CH preprocessing to build the augmented graph.
+Alternatively, live traffic can be supported using a three-phase setup as described in Section~\ref{sec:cch-intro}.
+In this case, we use the CCH instead of the CH preprocessing to build the augmented graph.
 This allows us to introduce updates to $w_\ell$ and the augmented graph within a few seconds using the CCH customization phase.
 This setup provides an alternative way of integrating live traffic compared to the approach described in Section~\ref{sec:live-predicted-traffic} where live traffic was represented only in $w_q$.
-In contrast, with CCH-Potentials, changes to the live traffic situation are handled by also changing $w_\ell$ via a customization.
-Additional features such as turn restrictions can be implemented by adjusting $w_q$.
-For severe traffic events, the three-phase approach will have significantly faster query running time than the one described in Section~\ref{sec:live-predicted-traffic}.
+In contrast, with CCH-Potentials, changes to the live traffic situation are handled by changing $w_\ell$ via customization.
+Additional features, such as turn restrictions, can be implemented by adjusting $w_q$.
+For severe traffic events, the three-phase approach will have significantly faster query running times than the one described in Section~\ref{sec:live-predicted-traffic}.
 However, it comes at the expense of a more complex setup and updating $w_\ell$ takes a few seconds.
@@
 In this section, we present our experimental evaluation.
-Our benchmark machine runs openSUSE Leap 15.3 (kernel 5.3.18), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
+Our benchmark machine runs openSUSE Leap 15.3 (kernel 5.3.18), and has 128\,GiB of DDR4-2133 RAM and an Intel Xeon E5-1630 v3 CPUs, which has four cores clocked at 3.7\,Ghz and 4~$\times$~32\,KiB of L1, 8~$\times$~256\,KiB of L2, and 10\,MiB of shared L3 cache.
 All experiments were performed sequentially.
 Our code is written in Rust and compiled with \texttt{rustc 1.57.0-nightly} in the release profile with the \texttt{target-cpu=native} option.
-The source code of our implementation and the experimental evaluation can be found on Github\footnote{\url{https://github.com/kit-algo/ch_potentials}}.
+The source code of our implementation and the experimental evaluation can be found on Github.\footnote{\url{https://github.com/kit-algo/ch_potentials}}
 
 \subsection{Inputs and Methodology}
-Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map.\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}
+Our primary benchmark instance is a graph of the road network of Germany obtained from Open Street Map.\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}
 To obtain the routing graph, we use the import from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
@@
 The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the vertices.
-It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
-We exclude speed values which are faster than the free-flow speed computed by RoutingKit.
+It also includes traffic predictions for 38\% of the edges as predicted speeds for all five-minute periods over a week.
+We exclude speed values faster than the free-flow speed computed by RoutingKit.
 We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACS implementation challenge~\cite{DemetrescuGJ09}.
@@
 The PTV instances are not OSM-based.
-One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges which was used for evaluation of many competing time-dependent algorithms.
+One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges, which was used to evaluate many competing time-dependent algorithms.
 The other two are newer instances of Europe with predictions for 27\% (TDEur17) and 76\% (TDEur20) of the edges.
@@
 For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
-We report preprocessing running times as averages over 10 runs.
+We report preprocessing running times as averages over ten runs.
 To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are vertices drawn uniformly at random and report average results.
@@
 Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials.
-With CCH-Potentials, $w_{\ell}$ can be updated by rerunning Phase 2 again.
+With CCH-Potentials, $w_{\ell}$ can be updated by rerunning Phase 2.
 }\label{tab:graphs}
@@
 
-The performance of Lazy RPHAST is evaluated with many-to-one queries where each query consists of $2^{14}$ sources and one target vertex.
+Lazy RPHAST is evaluated with many-to-one queries where each query consists of $2^{14}$ sources and one target vertex.
 However, instead of picking sources and targets from the full vertex set $V$, we draw them from local subsets of vertices $B$ of varying size $|B|$ called \emph{balls}.
-A ball $B$ is generated by picking a center uniformly at random and running Dijkstra's algorithm from it until the desired number of vertices $|B|$ was settled.
+A ball $B$ is generated by picking a centre uniformly at random and running Dijkstra's algorithm from it until the desired number of vertices $|B|$ is settled.
 This allows us to evaluate the performance depending on the distribution of the vertices.
-Since we use a fixed number of sources per query, a small ball size means that the vertices are densely clustered in the same region while large ball sizes mean that the vertices are distributed over large parts of the network.
+Since we use a fixed number of sources per query, a small ball size means that the vertices are densely clustered in the same region, while large ball sizes mean that the vertices are distributed over large parts of the network.
 For each ball size, we generate 100 balls.
-From each ball, we pick one set of sources to which we compute distances from 100 different targets selected uniformly at random from the same ball.
+We pick one set of sources from each ball to which we compute distances from 100 different targets selected uniformly at random from the same ball.
 Therefore, each reported running time is the mean over 10\,000 queries.
@@
 Figure~\ref{fig:lazy_rphast_inc} depicts the total elapsed time and the average running time to compute a single distance.
-The first few distances are the most expensive ones since a lot of the CH search space has not been explored yet.
+The first few distances are the most expensive since much of the CH search space has not been explored yet.
 With around 100\,$\mu$s, the running times are comparable to standard CH queries.
-For later queries, only little work remains to be done and the overhead per distance becomes almost constant depending on the ball size.
-
-The performance difference between the CH and CCH based variant is interesting.
-The CCH based variant can utilize the elimination tree for a more efficient implementation but the search space is denser.
+For later queries, little work remains to be done, and the overhead per distance becomes almost constant depending on the ball size.
+
+The performance difference between the CH and CCH-based variants is interesting.
+The CCH-based variant can utilize the elimination tree for a more efficient implementation, but the search space is denser.
 This makes earlier queries more expensive and later queries cheaper.
-With this trade-off, it depends on the ball size which variant is faster in terms of total running time.
+Therefore, it depends on the ball size which variant is faster in terms of total running time.
 
@@
 
-Despite the laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental many-to-one queries.
+Despite laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental many-to-one queries.
 Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target vertex and $2^{14}$ sources for different ball sizes.
@@
 We can compare the performance for fixed $S$-$t$ terminal sets.
-The difference is that with RPHAST one can efficiently compute distances from a different $t'$ to the same $S$ set while with Lazy RPHAST one can efficiently extend $S$ while $t$ stays the same.
-
-When looking at total running times for a single many-to-one query including selection times, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
+The difference is that with RPHAST, one can efficiently compute distances from a different $t'$ to the same $S$ set, while with Lazy RPHAST, one can efficiently extend $S$ while $t$ stays the same.
+
+When looking at total running times for a single many-to-one query, including selection times, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
 For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
@@
 They are obtained from one-to-all Dijkstra searches on continental-sized road networks.
-This heavily emphasizes memory bandwidth while neglecting other critical factors such as CPU frequency and cache size and speed which are likely much more important for algorithms carefully tuned to exploit data locality.
-} we can still conclude that Lazy RPHAST is a useful extension to RPHAST.
-It allows efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
+This experiment heavily emphasizes memory bandwidth while neglecting other critical factors such as CPU frequency and cache size and speed, which are likely much more critical for algorithms carefully tuned to exploit data locality.
+} we can still conclude that Lazy RPHAST is a valuable extension to RPHAST.
+It allows for efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
 % This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
@@
 CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
-However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
+However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise, one could use CH without A*).
 We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
@@
 Figure~\ref{fig:scaled_weights} depicts the results.
-Clearly, $\alpha$ has significant influence on the running time.
+Clearly, $\alpha$ has a significant influence on the running time.
 Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
@@
 For $\alpha > 1.1$, the growth slows down.
-This illustrates both the strengths and limits of our approach and goal-directed search in general.
-CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.
-
-We observe that the running times for a fixed $\alpha$ vary strongly.
+This illustrates the strengths and limits of our approach and goal-directed search in general.
+CH-Potentials can only achieve competitive running times if the application allows tight lower bounds at preprocessing time.
+
+We observe that the running times for a fixed $\alpha$ fluctuate heavily.
 This is an interesting observation, as with uniform source and target sampling, nearly all queries are long-distance.
@@
 This corresponds to using Dijkstra's algorithm.
-Secondly, we compare against our own implementation of ALT~\cite{gw-cppsp-05}.
+Secondly, we compare against our implementation of ALT~\cite{gw-cppsp-05}.
 We use 16 landmarks generated with the avoid strategy~\cite{gw-cppsp-05} and activate all during every query.
@@
 (C)CH-Potentials outperforms ALT by a factor of between six and seven and settle correspondingly fewer vertices.
-This is not surprising, since ALT computes worse distance estimates.
+This is not surprising since ALT computes worse distance estimates.
 In contrast, CH-Potentials already computes exact distances with respect to $w_\ell$.
@@
 % Our algorithm is only 1.6 times slower than accessing the exact $w_\ell$ distances instantly from a precomputed array.
-This means that no competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
-CCH-Potentials is slightly slower than CH-Potentials because they use a weight independent vertex importance order.
-This enables CCH-Potentials to support a customization whereas CH-Potentials cannot.
+No competing algorithm such as ALT or CPD-Heuristics can be significantly faster.
+CCH-Potentials is slightly slower than CH-Potentials because they use a weight-independent vertex importance order.
+% This enables CCH-Potentials to support a customization phase, whereas CH-Potentials cannot.
 
@@
 We first evaluate different variants of bidirectional A* in Table~\ref{tab:bidir_pruning} and~\ref{tab:bidir_switching} then compare the best ones against unidirectional A* in Table~\ref{tab:bidir}.
-Table~\ref{tab:bidir_pruning} studies the impact of our improved pruning and the low degree optimizations.
-As observed in the previous section, enabling the low degree optimization achieves a speedup by roughly a factor of three.
+Table~\ref{tab:bidir_pruning} studies the impact of our improved pruning and the low-degree optimizations.
+As observed in the previous section, enabling the low-degree optimizations achieves a speedup of roughly three.
 Symmetric bidirectional A* without our improved pruning has the worst performance.
-Enabling the improved pruning significantly improves the performance of symmetric bidirectional A*.
-For all heuristics except ALT, symmetric A* with the improved pruning has smaller search spaces than the average potential and very similar running times.
-Without the low degree improvements, the improved symmetric variant is marginally faster and with the low degree improvements the average potential remains slightly faster.
-This is due to the reduced impact of the heuristic evaluation overhead with the low degree optimizations.
-Enabling the improved pruning for the average potential marginally reduces the search space size and slightly increases running times.
+Enabling the improved pruning improves the performance of symmetric bidirectional A* significantly.
+For all heuristics except ALT, symmetric A* with improved pruning has smaller search spaces than the average potential and similar running times.
+Without the low-degree improvements, the improved symmetric variant is marginally faster.
+With the low-degree improvements, the average potential remains slightly faster.
+This is due to the reduced impact of the heuristic evaluation overhead with the low-degree optimizations.
+Enabling the improved pruning for the average potential reduces the search space size marginally and slightly increases running times.
 
@@
 The symmetric variant uses the improved pruning, the average variant does not.
-All variants use all low degree optimizations.
+All variants use all low-degree optimizations.
 }\label{tab:bidir_switching}
@@
 The symmetric variant uses the improved pruning, the average variant does not.
-All variants use all low degree optimizations.
+All variants use all low-degree optimizations.
 }\label{tab:bidir}
@@
 
-In Table~\ref{tab:bidir} we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
-Interestingly, only the zero heuristic and ALT consistently achieve speedups through bidirectional search on all query weights.
+In Table~\ref{tab:bidir}, we investigate the effectiveness of bidirectional search compared to unidirectional search depending on the query weights.
+Interestingly, only the zero heuristic and ALT consistently achieve speedups through bidirectional search.
 With $w_q = w_{\ell}$, unidirectional CH-Potentials is already optimal and only traverses the shortest path.
-Here, bidirectional search will only introduce unnecessary overhead.
-When query weights are scaled up uniformly, bidirectional search achieves some search space reduction, but it is not enough to significantly reduce running times due to the overhead of running a second search.
+Here, the bidirectional search will only introduce unnecessary overhead.
+When query weights are scaled up uniformly, bidirectional search achieves some search space reduction.
+However, it is not enough to significantly reduce running times due to the overhead of running a second search.
 This changes drastically when the query weight increases are applied non-uniformly in the third scenario.
-Here only weights with a speed less than 80\,kph were scaled up.
+Here only weights with speed less than 80\,kph were scaled up.
 This touches only the beginning and end of most shortest paths between randomly chosen vertices.
@@
 In this case, bidirectional (C)CH-Potentials is a factor of two faster than the unidirectional variant.
-The reason for this is that the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
-In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero and thus avoid this expansion.
-This is also the reason why ALT behaves like this for all query weights.
-By construction, the ALT heuristic has better reduced weights for edges which are on many shortest paths like highways.
+This is because the search space of a unidirectional search expands greatly while exploring the end of the path to the target where the reduced weights are bad.
+In contrast, the bidirectional searches meet in the middle of the shortest path where the reduced weights are close to zero, thus avoiding this expansion.
+This is also why ALT behaves like this for all query weights.
+By construction, the ALT heuristic has better reduced weights for edges which lie on many shortest paths like highways.
 Conversely, unimportant edges have bad reduced weights.
-This makes bidirectional search so critical for ALTs performance.
-In contrast, having a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
+This makes bidirectional search so critical for the ALT performance.
+In contrast, a potential as tight as CH-Potentials makes bidirectional search in many scenarios unnecessary.
 Bidirectional search for CH-Potentials only pays off when the reduced weights are bad around the terminals.
@@
 CH-Potentials performance for different route planning applications.
-Depending on the problem we apply unidirectional or bidirectional CH-Potentials (CH U or CH B) or CCH-Potentials (CCH U/B).
-We report average running times and number of queue pushes.
-We also report the average length increase, that is how much longer the final shortest distance is compared to the lower bound.
+Depending on the problem, we apply unidirectional or bidirectional CH-Potentials (CH U or CH B) or CCH-Potentials (CCH U/B).
+We report average running times and the number of queue pushes.
+We also report the average length increase, that is, how much longer the final shortest distance is compared to the lower bound.
 Finally, we report the average running time of Dijkstra's algorithm as a baseline and the speedup over this baseline.
@@
 Table~\ref{tab:applications} depicts the running times of (C)CH-Potentials in various applications, such as those described in Section~\ref{sec:applications}.
-We report speedups compared to extensions of Dijkstra's algorithm for each application respectively.
+We report speedups compared to extensions of Dijkstra's algorithm for each application.
 We start with the base case where $w_q = w_\ell$.
@@
 CH achieves average query running times of 0.16\,ms on OSM Ger.
-CH-Potentials is roughly four times slower but still achieve a huge speedup of 3795 over Dijkstra.
-Such large speedups are typical for CH.
+CH-Potentials is roughly four times slower but still achieves a speedup of 3795 over Dijkstra.
+Such significant speedups are typical for CH.
 This shows that CH-Potentials gracefully converges toward CH in the $w_q = w_\ell$ special case.
-On DIMACS Europe, the average degree is somewhat higher which makes the low-degree optimizations less effective which leads to more queue operations and a slightly slower running time.
+On DIMACS Europe, the average degree is somewhat higher, making the low-degree optimizations less effective and leading to more queue operations and a slightly slower running time.
 
@@
 We measure this quality using the length increase of $w_q$ compared to $w_\ell$.
-Forbidding highways results in the largest length increase and in the smallest speedup.
-The other extreme are turn restrictions.
-They have only a small impact on the length increase.
-The achieved speedups are therefore comparable to CH speedups.
-Since turn costs and restrictions appear mostly in the beginning and end of shortest-paths and not in the middle on highways, utilizing bidirectional search results in even better speedups.
+Avoiding highways results in the most significant length increase and the smallest speedup.
+The other extreme is turn restrictions.
+They have little impact on the length increase.
+The achieved speedups are, therefore, comparable to CH speedups.
+Since turn costs and restrictions appear primarily in the beginning and end of shortest paths and not in the middle on highways, utilizing bidirectional search results in even better speedups.
 Mapbox live traffic has a length increase of around 15\%, which yields running times of 130\,ms.
-Applying bidirectional CH-Potentials in this case is, in fact, detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
+Applying bidirectional CH-Potentials, in this case, is, in fact, detrimental to the performance because the bad reduced edge weights appear in the middle of shortest paths.
 The same is the case for forbidden tunnels or highways.
@@
 Combining predicted and live traffic results in a running time of around 200\,ms.
-Further adding turn restrictions, significantly increases the running times.
-This increase is mostly due to the BCC optimization of Section~\ref{sec:largested-biconnected-component} becoming ineffective when considering turns.
+Adding turn restrictions additionally increases the running times significantly.
+This increase is primarily due to the BCC optimization of Section~\ref{sec:largested-biconnected-component} becoming ineffective when considering turns.
 It is not due to the length increase of using turns.
@@
 Interestingly, the PTV traffic predictions have a much smaller length increase than the Mapbox predictions.
-This results in somewhat smaller running times of our algorithm.
+This results in somewhat faster running times of our algorithm.
 
@@
 
-While the query running times reported in Table~\ref{tab:applications} are decent in many settings, they are not competitive to techniques tailored to specific applications.
+While the query running times reported in Table~\ref{tab:applications} are decent in many settings, they are not competitive with techniques tailored to specific applications.
 In the simple $w_q = w_{\ell}$ setting, Hub Labels can be used to answer queries in less than a microsecond~\cite{DBLP:conf/wea/DellingGW13}, more than three orders of magnitude faster than with CH-Potentials.
-Live traffic and arbitrary weight functions can be handled with CCH resulting in query times of around a 0.1\,ms~\cite{DBLP:journals/jea/Buchhold0W19}.
+Live traffic and arbitrary weight functions can be handled with CCH resulting in query times of around 0.1\,ms~\cite{DBLP:journals/jea/Buchhold0W19}.
 CCH can also be adjusted to turn costs.
@@
 Nevertheless, the overall picture is clear enough.
-However, the advantage of the CH-Potentials framework over these fine-tuned techniques is that it is unified and flexible approach which can handle all these applications without any adjustments to the preprocessing.
+However, the advantage of the CH-Potentials framework over these fine-tuned techniques is that it is a unified and flexible approach which can handle all these applications without any adjustments to the preprocessing.
 Further, CH-Potentials preprocessing times are at least an order of magnitude faster than approaches like CATCHUp (sequentially more than four hours on TDEur20 compared to five minutes with CH-Potentials) and require significantly less memory (Hub Labels, for example, needs 20\,GiB on DIMACS Europe, compared to around 750\,MiB with CH-Potentials, see Table~\ref{tab:graphs}).
-Finally, to the best of our knowledge, for problem settings such as the combination of predicted and live traffic, there simply does not exist any exact technique able to handle this setting, let alone additionally integrating turn costs.
+Finally, to the best of our knowledge, for problem settings such as the combination of predicted and live traffic, there does not exist any exact technique to handle this setting, let alone to integrate turn costs additionally.
 The key achievement of CH-Potentials is that problem extensions can be integrated by trading query performance rather than developing new algorithms.
@@
 In this article, we introduced CH-Potentials, a fast, exact, and flexible two-phase routing framework based on A* and CH.
-The approach can handle a multitude of complex, integrated routing scenarios with very little implementation complexity and without any changes to the preprocessing algorithms.
-CH-Potentials provides \emph{exact} distances with respect to lower bound weights known at preprocessing time as an A* heuristic.
-Thus, the query performance of CH-Potentials crucially depends on the availability of good lower bounds in the preprocessing phase.
-Our experiments show, that this availability highly depends on the application.
+The approach can handle complex, integrated routing scenarios with little implementation complexity and no changes to the preprocessing algorithms.
+CH-Potentials provides \emph{exact} distances for lower bound weights known at preprocessing time as an A* heuristic.
+Thus, the query performance of CH-Potentials crucially depends on the availability of reasonable lower bounds in the preprocessing phase.
+Our experiments show that this availability highly depends on the application.
 We also show that the overhead of our heuristic is within a factor 1.6 of a hypothetical A*-heuristic that can instantly access lower bound distances.
@@
 Another promising research avenue would be to investigate graphs other than road networks.
-A lot of research into grid maps exists including a series of competitions called GPPC~\cite{DBLP:conf/socs/SturtevantTTUKS15}.
+Much research for grid maps exists, including a series of competitions called GPPC~\cite{DBLP:conf/socs/SturtevantTTUKS15}.
 Hierarchical techniques have been shown to work well on these graphs~\cite{DBLP:conf/aaai/UrasK14}.
-It might also be worthwhile to apply CH-Potentials to even more routing applications, for example to the shortest $\epsilon$-smooth path problem~\cite{dss-tarrn-18} or the problem of computing alternative routes~\cite{adgw-arrn-13,bdgs-argrn-11,kobitzsch2015alternative}.
-Many-to-one problems also appear as subproblems in other routing algorithms, for example in nearest neighbor computations~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
+It might also be worthwhile to apply CH-Potentials to even more routing applications, for example, to the shortest $\epsilon$-smooth path problem~\cite{dss-tarrn-18} or the problem of computing alternative routes~\cite{adgw-arrn-13,bdgs-argrn-11,kobitzsch2015alternative}.
+Many-to-one problems also appear as subproblems in other routing algorithms, for example, in nearest neighbour computations~\cite{buchhold_et_al:LIPIcs.SEA.2021.18}.
 Investigating whether Lazy RPHAST can be used to improve these algorithms appears to be a worthwhile direction for future research.

### 2022-08-30 paper/ch_potentials.tex
augmented weights

@@
   \For{all reversed downward edges $uv \in \rechd$}{
-    \If{$D^{\downarrow}[u] + \overleftarrow{w}(uv) < D^{\downarrow}[v]$}{
-      $D^{\downarrow}[v] \gets D^{\downarrow}[u] + \overleftarrow{w}(uv)$\;
+    \If{$D^{\downarrow}[u] + \overleftarrow{w}^+(uv) < D^{\downarrow}[v]$}{
+      $D^{\downarrow}[v] \gets D^{\downarrow}[u] + \overleftarrow{w}^+(uv)$\;
       Add $v$ or decrease $v$'s key in $Q$ to $D^{\downarrow}[v]$\;
@@
 To \emph{contract} a vertex means temporarily removing it from the graph while inserting \emph{shortcut edges} between more important neighbours to preserve shortest distances among them.
-The result is an \emph{augmented graph} $G^+$ with original edges and shortcuts.
+The result is an \emph{augmented graph} $G^+$ of original edges and shortcuts with weights $w^+$.
 We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echu)$ containing only downward edges.
@@
 	\For{all up-edges $uv \in \echu$ with $u$ in $L$}{
-		\If{$D[u] > D[v] + w(uv)$}{
-			$D[u] \leftarrow D[v] + w(uv)$\;
+		\If{$D[u] > D[v] + w^+(uv)$}{
+			$D[u] \leftarrow D[v] + w^+(uv)$\;
 		}
@@
 		\For{all up-edges $uv \in \echu$}{
-      $D[u]\leftarrow\min(D[u],w(uv)+\Dist(v))$\;
+      $D[u]\leftarrow\min(D[u],w^+(uv)+\Dist(v))$\;
 		}
@@
 		$D[v]\gets D^{\downarrow}[v]$\;
-		\For{all up-edges $vw \in \echu$}{
-      $D[v]\gets\min(D[v],w(vw)+D[w])$\;
+		\For{all up-edges $vx \in \echu$}{
+      $D[v]\gets\min(D[v],w^+(vx)+D[x])$\;
 		}

### 2022-08-31 paper/ch_potentials.tex
more final wordsmithing

@@
 Some applications may have additional application-specific requirements.
-Further, it is insufficient to support each of these features independently.
-Instead, we need to support their combination.
+Further, it is insufficient to handle each of these features independently.
+Instead, all requirements must be supported in combination.
 
@@
 In contrast, extending speedup techniques is vastly more complex.
-The results on extending specific speedup techniques to specific extended problem fill many research papers~\cite{gv-errnt-11,dgpw-crprn-13,dsw-cch-15,bwzz-cchtc-20,bgsv-mtdtt-13,swz-sfert-21} and sometimes entire dissertations~\cite{Delling2009_1000011046,b-tdrpc-14,Baum2018_1000082225}.
+The results on adopting specific speedup techniques to specific extended problem fill many research papers~\cite{gv-errnt-11,dgpw-crprn-13,dsw-cch-15,bwzz-cchtc-20,bgsv-mtdtt-13,swz-sfert-21} and sometimes entire dissertations~\cite{Delling2009_1000011046,b-tdrpc-14,Baum2018_1000082225}.
 We will present a brief overview in Section~\ref{sec:related_work}.
@@
 For example, while there exist speedup techniques achieving fast queries for each of the features mentioned above, we are unaware of any work supporting the combination of these features.
-We believe that a different trade-off between running time and extensibility is necessary for practical applications.
+We believe a different trade-off between running time and extensibility is necessary for practical applications.
 Therefore, we prioritize a simple and extensible approach over the fastest possible query times in this work.
@@
 Despite the extremely fast query times, it appears that the trade-off offered by HL is, in practice, often not very attractive.
-Simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (such as the network latency).
-Also, we are not aware of any works extending HL to extended problem models.
+We are not aware of any works adapting HL to extended problem models.
+Also, simpler techniques like CH already offer queries fast enough to completely disappear behind other parts of practical applications (such as the network latency).
 
@@
 For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
-We report preprocessing running times as averages over ten runs.
+We report sequential preprocessing running times as averages over ten runs.
+Note that CCH preprocessing can be parallelized efficiently.
+Thus, practical running times can be even faster.
 To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are vertices drawn uniformly at random and report average results.
@@
 \caption{
-Instances used in the evaluation with preprocessing running times to construct (C)CH-Potentials.
+Instances used in the evaluation with sequential preprocessing running times to construct (C)CH-Potentials.
 With CCH-Potentials, $w_{\ell}$ can be updated by rerunning Phase 2.

### 2022-10-07 paper/ch_potentials.tex
typo fixes

@@
 The result is an \emph{augmented graph} $G^+$ of original edges and shortcuts with weights $w^+$.
-We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echu)$ containing only downward edges.
+We often refer to the augmented graph split into an upward graph $\gchu = (V, \echu)$ containing only upward edges and a downward graph $\gchd = (V, \echd)$ containing only downward edges.
 
@@
 For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
-Lazy RPHAST (on a CH) takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
+Lazy RPHAST on CH takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
 The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.

### 2022-10-10 paper/ch_potentials.tex
integrate reviewer feedback

@@
 Further, for many extended problems, there are dedicated research results on extensions of specific techniques.
-While many problems can be solved quite efficiently, we still believe this situation is unsatisfactory.
+While many problems can be solved efficiently, we still believe this situation is unsatisfactory.
 For practical applications, a unified and extensible approach with manageable implementation complexity is often more important than the fastest query performance.
@@
 The core of our approach is a new CH-based A* heuristic.
-It allows for much tighter estimates than previous A* heuristics and thus significantly faster queries.
-While the query running times of our approach are not competitive with approaches tailored to specific problems, they are still sufficient for practical applications.
+It allows for much tighter estimates than previous A* heuristics and, thus, significantly faster queries.
+While the query running times of our technique are not competitive with approaches tailored to specific problems, i.e.
+typically an order of magnitude slower, they are often sufficient for practical applications.
+Further, our approach only requires the classical CH preprocessing regardless of the specific extended routing problem.
+Therefore, preprocessing time and memory consumption are typically significantly better than approaches aiming for competitive query times in specific problems.
 
@@
 In Section~\ref{sec:lazy-rphast}, we introduce Lazy RPHAST, a simple and efficient CH-based algorithm for the incremental many-to-one shortest path problem.
+It is the first algorithm to efficiently support accelerated shortest path computations from \emph{dynamic} source sets to a fixed target.
 Section~\ref{sec:astar_opts} contains several optimizations for A* in road networks accelerating the processing of low-degree vertices (Section~\ref{sec:low-deg-improvment}) and an improved variant of bidirectional A* (Section~\ref{sec:bidir_astar}).
@@
 
+\begin{algorithm2e}[tb]
+\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
+\KwData{$D[v]$: memoized distance from any vertex $v \in V$ to $t$, shared between invocations, initialized to $\bot$ during the selection}
+\KwData{$P[v]$: parent of a vertex $v \in V$ in the elimination tree, a parent of $\bot$ indicates the root vertex}
+\KwData{$S$: stack with vertices to compute distances, empty initially, only used to store intermediate data}
+\SetKwFunction{Dist}{ComputeAndMemoizeDist}
+\SetKw{Break}{break}
+\SetKwProg{Fn}{Function}{:}{}
+
+Execute Algorithm~\ref{algo:ch-backward}, filling $D^{\downarrow}$\;
+$D[v] \gets \bot$ for all $v \in V$\;
+
+\Fn{\Dist{$u$}}{
+  \tcp{Determine the vertices $v$ for which $D[v]$ needs to be computed}
+  $v \gets u$\;
+	\While{$D[v] = \bot$}{
+    Push $v$ onto $S$\;
+		\If{$P[v] = \bot$}{
+      \Break\;
+		}
+    $v \gets P[v]$\;
+  }
+  \tcp{Compute $D$ for those vertices}
+  \While{$S$ not empty}{
+    $v\leftarrow$ pop top element from $S$\;
+		$D[v]\gets D^{\downarrow}[v]$\;
+		\For{all up-edges $vx \in \echu$}{
+      $D[v]\gets\min(D[v],w^+(vx)+D[x])$\;
+		}
+	}
+	\Return{$D[u]$}\;
+}
+\caption{Elimination tree-based Lazy RPHAST algorithm.}
+\label{algo:cch_pot}
+\end{algorithm2e}
+
 Algorithm~\ref{algo:lazy_rphast_ch} can be applied to CCH without any modifications.
@@
 Thus, as soon as the algorithm encounters a vertex with a final distance, the remaining search space is known to have final distances.
-
-\begin{algorithm2e}[tb]
-\KwData{$D^{\downarrow}[v]$: tentative distance from any vertex $v \in V$ to $t$ as computed by Algorithm~\ref{algo:ch-backward}}
-\KwData{$D[v]$: memoized distance from any vertex $v \in V$ to $t$, shared between invocations, initialized to $\bot$ during the selection}
-\KwData{$P[v]$: parent of a vertex $v \in V$ in the elimination tree, a parent of $\bot$ indicates the root vertex}
-\KwData{$S$: stack with vertices to compute distances, empty initially, only used to store intermediate data}
-\SetKwFunction{Dist}{ComputeAndMemoizeDist}
-\SetKw{Break}{break}
-\SetKwProg{Fn}{Function}{:}{}
-
-Execute Algorithm~\ref{algo:ch-backward}, filling $D^{\downarrow}$\;
-$D[v] \gets \bot$ for all $v \in V$\;
-
-\Fn{\Dist{$u$}}{
-  \tcp{Determine the vertices $v$ for which $D[v]$ needs to be computed}
-  $v \gets u$\;
-	\While{$D[v] = \bot$}{
-    Push $v$ onto $S$\;
-		\If{$P[v] = \bot$}{
-      \Break\;
-		}
-    $v \gets P[v]$\;
-  }
-  \tcp{Compute $D$ for those vertices}
-  \While{$S$ not empty}{
-    $v\leftarrow$ pop top element from $S$\;
-		$D[v]\gets D^{\downarrow}[v]$\;
-		\For{all up-edges $vx \in \echu$}{
-      $D[v]\gets\min(D[v],w^+(vx)+D[x])$\;
-		}
-	}
-	\Return{$D[u]$}\;
-}
-\caption{Elimination tree-based Lazy RPHAST algorithm.}
-\label{algo:cch_pot}
-\end{algorithm2e}
 
@@
 
-\subsubsection{Skip Degree-Two Vertices}
+\subsubsection{Skip Degree-Two Vertices}\label{sec:deg2}
 
@@
 
-\subsubsection{Skip Degree-Three Vertices}
+\subsubsection{Skip Degree-Three Vertices}\label{sec:deg3}
 
@@
 
-Lazy RPHAST is evaluated with many-to-one queries where each query consists of $2^{14}$ sources and one target vertex.
+Lazy RPHAST is evaluated with many-to-one queries where each query consists of a source set $S$ with $2^{14}$ sources and one target vertex $t$.
 However, instead of picking sources and targets from the full vertex set $V$, we draw them from local subsets of vertices $B$ of varying size $|B|$ called \emph{balls}.
@@
 \caption{
-Running times of (C)CH-based Lazy RPHAST for many-to-one queries with $|S| = 2^{14}$ sources picked from a ball of varying size $|B|$.
+Running times of (C)CH-based Lazy RPHAST and CH-based RPHAST for many-to-one queries with $|S| = 2^{14}$ sources picked from a ball of varying size $|B|$.
 The running time includes the selection and the time to compute all distances.
@@
 \Description{
-A double logarithmic line plot with four lines with similar shapes and absolute values.
-Running times for both algorithms and graphs with a ball size of 16384 is around 1ms.
-For a ball size of 2 to the 20 CH based variants need 10ms and CCH based variants slightly more.
-For a ball size of 2 to the 24 CH based variants are at 30ms and CCH based variants slightly above 40ms.
+A double logarithmic line plot with six lines with similar shapes and absolute values.
+Running times for both Lazy RPHAST variants on both graphs graphs with a ball size of 16384 is around 1ms.
+Classical RPHAST takes up to 2ms.
+For a ball size of 2 to the 20 CH-based Lazy RPHAST variants need 10ms and CCH based variants slightly more.
+For a ball size of 2 to the 24 CH-based Lazy RPHAST variants are at 30ms and CCH based variants slightly above 40ms.
 }
@@
 Despite laziness being the distinguishing feature in Lazy RPHAST, the algorithm can still be used for non-incremental many-to-one queries.
-Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target vertex and $2^{14}$ sources for different ball sizes.
+Figure~\ref{fig:many_to_one} depicts average running times to compute distances between one target vertex and $2^{14}$ sources for different ball sizes for Lazy RPHAST and our own implementation of RPHAST~\cite{delling_et_al:OASIcs:2011:3266}.
 The query generation methodology is the same as in the previous experiment.
-As this is the same methodology also used in~\cite{delling_et_al:OASIcs:2011:3266}, we can relate our results to the performance of other one-to-many algorithms.
+As this is the same methodology also used in~\cite{delling_et_al:OASIcs:2011:3266}, we can also roughly relate our results to the performance of other one-to-many algorithms.
 Keep in mind that algorithms like RPHAST optimize for a different setting than we do.
@@
 
-When looking at total running times for a single many-to-one query, including selection times, Lazy RPHAST delivers very competitive performance and is as fast if not faster than all algorithms evaluated in~\cite{delling_et_al:OASIcs:2011:3266}.
-For example, Delling et al. report an average RPHAST running time of 1.97\,ms for selection and query combined for $|B| = 2^{14}$ and 28.52\,ms for $|B| = 2^{20}$ on DIMACS Europe.
-Lazy RPHAST on CH takes 0.96\,ms and 10.76\,ms respectively, to compute the same distances.
-The CCH variant is slightly slower (1.00\,ms and 14.41\,ms) but still significantly faster than RPHAST.
-While the numbers are not perfectly comparable due to different benchmark machines,\footnote{
-According to the comparison methodology from~\cite{bdgmpsww-rptn-16} (see \url{https://i11www.iti.kit.edu/~pajor/survey/}), the machine used in~\cite{delling_et_al:OASIcs:2011:3266} (SPA-2) is about 30\% slower than ours (we obtained a score of 33\,159\,ms).
-However, these scaling factors have to be interpreted very carefully.
-They are obtained from one-to-all Dijkstra searches on continental-sized road networks.
-This experiment heavily emphasizes memory bandwidth while neglecting other critical factors such as CPU frequency and cache size and speed, which are likely much more critical for algorithms carefully tuned to exploit data locality.
-} we can still conclude that Lazy RPHAST is a valuable extension to RPHAST.
-It allows for efficiently handling dynamic source sets and is competitive in the setting where both the target and the source set change between queries.
+When looking at total running times for a single many-to-one query, including selection times, Lazy RPHAST yields performance competitive to RPHAST.
+The CH-based variant is consistently faster than RPHAST.
+Further, even the Lazy RPHAST CCH-based variant is faster than CH-based RPHAST on ball sizes up to $|B| = 2^{19}$ despite the larger search space.
+Unsurprisingly, the absolute running times of RPHAST and Lazy RPHAST are similar overall:
+Both algorithms run on the same CH search space.
+% On the one hand, Lazy RPHAST needs to do slightly less work and looks at each edge in the CH search space only once.
+% On the other, RPHAST looks at edges in the search space both during the selection and the query but has a more pipeline and cache-friendly organization.
+Our RPHAST running times roughly reproduce the results reported in~\cite{delling_et_al:OASIcs:2011:3266}: Our running times are between 20\% (on small ball sizes) and 60\% (on larger ball sizes) faster, which is likely due to differences in benchmark machine performance.
+% We can, therefore, even expect competitive performance compared to other many-to-one algorithms.
+We conclude that Lazy RPHAST is a very valuable extension of RPHAST.
+% It is the first algorithm to efficiently support accelerated shortest path computations with dynamic source sets.
+% Further, it even features competitive running times in the setting where both the target and the source set change between queries.
+It allows for efficiently handling dynamic source sets and is even competitive in the setting where both the target and the source set change between queries.
 % This is of course not a fair comparison since the approaches in~\cite{delling_et_al:OASIcs:2011:3266} are geared towards having one expensive selection phase and than many faster queries.
@@
 \caption{
-Average query running times and number of queue pushes with different heuristics and optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
-The BCC, Deg2 and Deg3 columns indicate which optimizations from Section~\ref{sec:low-deg-improvment} were used.
+Average query running times and number of queue pushes with different heuristics and low-degree optimizations on OSM Ger with $w_q = 1.05 \cdot w_\ell$.
+The BCC column indicates if the search stays within the largest biconnected component (see Section~\ref{sec:largested-biconnected-component}), Deg2 if vertices of degree two are skipped (Section~\ref{sec:deg2}) and Deg3 if vertices of degree three are skipped (Section~\ref{sec:deg3}).
 }\label{tab:building_blocks}
@@
 For time-dependent routing, CATCHUp~\cite{swz-sfert-21} achieves query times more than an order of magnitude faster than CH-Potentials (6.3\,ms compared to our 97.7\,ms).
-Again, these numbers are not perfectly comparable due to different benchmark machines.
-Nevertheless, the overall picture is clear enough.
+These numbers are not perfectly comparable due to different benchmark machines but the overall picture is clear enough.
 However, the advantage of the CH-Potentials framework over these fine-tuned techniques is that it is a unified and flexible approach which can handle all these applications without any adjustments to the preprocessing.

### 2022-11-08 paper/ch_potentials.tex
update with improvements from diss

@@
 Further, $G^+$ can be decomposed into two directed acyclic graphs (DAG) $\gchu$ and $\gchd$ and the up path can be found in $\gchu$ and the down path in $\gchd$.
-Observe that the second part of Lazy RPHAST is, in fact, a recursive depth-first search (DFS) on $\gchu$, which relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
-Since DFS can be used to find shortest paths on DAGs, the search will find shortest up paths on $\gchu$.
+The Lazy RPHAST selection finds the shortest down path in $\gchd$.
+Now observe that the \textsf{ComputeAndMemoizeDist} function is, in fact, a recursive depth-first search (DFS) on $\gchu$.
+Edges $uv$ are relaxed once all upward neighbors of $u$ have been processed, i.e.\ in DFS-post order.
+A DFS-post order also is a reverse topological order for $\gchu$.
+Therefore, the algorithm relaxes edges $uv$ in reverse topological order of the tail vertices $u$.
+Since $\gchu$ is a DAG, this yields shortest up distances in $\gchu$.
 Concatenated with the down paths obtained by the backward search, this yields optimal shortest distances.
@@
 For any vertex which already has a memoized distance $D[v] \neq \bot$, the algorithm immediately returns this distance.
-Otherwise, the algorithm follows the elimination tree upward until a node with a finalized distance is encountered.
+Otherwise, the algorithm follows the elimination tree upward until a vertex with a finalized distance is encountered.
 The elimination tree is represented as parent pointer array $P$.
-The visited nodes are pushed onto a stack $S$.
+The visited vertices are pushed onto a stack $S$.
 This enables the algorithm to enumerate the vertices in reversed order by popping them from the stack.
-While popping the nodes, all outgoing upward edges are relaxed.
+While popping the vertices, all outgoing upward edges are relaxed.
 This finalizes the shortest distances for all vertices on the elimination tree path, including the desired query vertex $v$.
@@
 Analogously, if $y$ is reached, $y$'s weight is set to $D[y]+h_t(y)$.
-If neither $y$ nor $b$ are not reached, no queue operation is performed.
+If neither $y$ nor $b$ are reached, no queue operation is performed.
 Using this method, we avoid pushing every other degree-three vertex into the queue.
@@
 For vertices in the core, $a_v=v$.
-For every vertex $v$ outside of the core, the attachment vertex $a_v$ is the cut vertex in the core that separates $v$'s component from the core.
+For every vertex $v$ outside of the core, the attachment vertex $a_v$ is the cut vertex in the core that separates $v$'s component from the core (or a sentinel value $\bot$ for vertices in components not connected to the core).
 
 The query phase is divided into two steps.
-In the first step, we run A* on the subgraph induced by the core and $s$'s component.
-Formulated differently, we only consider nodes which are in the core or have the same attachment vertex as $s$.
-This can be achieved implicitly by not following edges to nodes without this property.
+First, we run A* on the subgraph induced by the core and $s$'s component.
+Formulated differently, we only consider vertices which are in the core or have the same attachment vertex as $s$.
+We achieve this implicitly by not following edges to vertices without this property.
 If $t$ is part of $G_C$ or in the same component as $s$, this A* search finds it.
@@
 The final result is the concatenation of both paths.
+When $t$ is not connected to the core ($a_t = \bot$) but $s$ is ($a_s \neq \bot$), we immediately return a distance of $\infty$.
 
@@
 We observe that the number of queue pushes roughly correlates with running time.
-As (C)CH-Potentials and Oracle-A* have the same heuristic values, the number of queue pushes are by construction equal.
 Each optimization reduces both queue pushes and running times.
@@
 In contrast, CH-Potentials already computes exact distances with respect to $w_\ell$.
-The number of popped vertices is the same for CH-Potentials, CCH-Potentials and Oracle-A*.
+As (C)CH-Potentials and Oracle-A* have the same heuristic values, the number of queue pushes are by construction equal.
 The only difference between (C)CH-Potentials and Oracle-A* is the overhead of the heuristic evaluation.

### 2022-11-08 paper/ch_potentials.tex
prep for final submission

@@
 % \documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
-\documentclass[manuscript,review]{acmart}
+\documentclass[manuscript]{acmart}
 
@@
 
-\usepackage{todonotes}
 % \usepackage{booktabs}
@@
 \usepackage[algo2e,vlined]{algorithm2e}
-\usepackage{todonotes}
 
@@
 % However, there is little research focused on designing extensible algorithms which can be adapted to new problem variants with minimal effort.
-In this work, we revisit the A* algorithm to build a simple, extensible and unified algorithmic framework applicable to many route planning problems. % without adjustments to the preprocessing.
+In this work, we revisit the A* algorithm to build a simple, extensible and unified algorithmic framework applicable to many route planning problems.
+% without adjustments to the preprocessing.
 A* has been previously used for routing in road networks.
@@
 \includegraphics[width=.6\columnwidth]{fig/searchspace_st.png}
-\caption{Vertices explored by A*. The color indicates the vertex removal order from the queue. Blue was removed first. Next is green. Red was removed last.}
+\caption{Vertices explored by A*.
+The color indicates the vertex removal order from the queue.
+Blue was removed first.
+Next is green.
+Red was removed last.
+Background by \href{https://www.openstreetmap.org/copyright}{OpenStreeMap}.}
 \label{img:search-space}
@@
 \caption{
-Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down $st$-path in $G^+$.
+Solid lines are edges in $G$. Dotted lines are shortcuts. Red is a shortest $st$-path in $G$. Blue is equally long up-down path in $G^+$.
 }
@@
 For a detailed discussion we refer to~\cite{gssv-erlrn-12}.
-Here, we only briefly introduce the necessary notation and algorithms used in this paper.
+% Here, we only briefly introduce the necessary notation and algorithms used in this paper.
 In a preprocessing phase, vertices are ordered totally by ``importance'' where more important vertices should lie on more shortest paths.
@@
 
-Unless stated differently, for all bidirectional A* variants, we always alternate between removing a vertex from the forward and the backward queues.
+Unless stated differently, for bidirectional A*, we always alternate between removing a vertex from the forward and the backward queues.
 We also evaluate expanding the search with the smaller minimum queue key in our experiments.
@@
 This function can also take additional parameters from the state of the search.
-In the case of traffic predictions, $w_q$ is a function which maps the edge entry time to the traversal time, and the query takes an additional departure time parameter.
+In the case of traffic predictions, $w_q$ is a function mapping the edge entry time to the traversal time, and the query takes an additional departure time parameter.
 
@@
 $w_q$ is set to the travel time accounting for current traffic.
-As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
+As traffic only increases travel times, $w_\ell$ is a valid lower bound for $w_q$.
 In a real-world application, values from $w_q$ could be updated between queries.
@@
 For severe traffic events, the three-phase approach will have significantly faster query running times than the one described in Section~\ref{sec:live-predicted-traffic}.
-However, it comes at the expense of a more complex setup and updating $w_\ell$ takes a few seconds.
+However, it leads to a more complex setup and updating $w_\ell$ takes a few seconds.
 % Also, the customization is not fast enough to rerun it for every query.

### 2022-11-08 paper/export_acm.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+if [[ $(git diff --shortstat 2> /dev/null | tail -n1) != "" ]] ; then
+  echo "Cant export - would possibly override stuff"
+  exit 0
+fi
+
+sed -i '/^%/ d' ch_potentials.tex
+
+zip jea.zip \
+ch_potentials.tex \
+references.bib \
+fig/* \
+table/*
+
+git checkout ch_potentials.tex
